### 2024-12-06 Code/6dec/trivariate.Rmd
6dez

---
title: "Trivariate Dichteanalyse mit Benchden-Verteilungen"
author: "Léon Kia Faro"
date: "2024-12-05"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    number_sections: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Benötigte Pakete
library(benchden)    # Synthetische Verteilungen
library(tram)        # Transformation Models
library(hdrcde)      # High Density Regions
library(np)          # Nichtparametrische Methoden
library(ks)          # Kernel Smoothing
library(kdensity)    # Kernel Density Estimation
library(kedd)        # Kernel Estimation
library(lpdensity)   # Lokale polynomiale Dichte
library(plotly)      # 3D Visualisierung
library(viridis)     # Farbpaletten
library(parallel)    # Parallele Berechnung
```

# 1. Benchden-Verteilungen

## 1.1 Verteilungsauswahl

Wir fokussieren uns auf drei komplexe Verteilungen:

1. Marronite (Nr. 21):
   $$f_{21}(x) = \frac{1}{2}\phi(x) + \frac{1}{10}\sum_{i=1}^5 \phi(x-i)$$
   - Asymmetrische multimodale Struktur
   - Herausfordernd für bedingte Schätzung

2. Claw (Nr. 23):
   $$f_{23}(x) = \frac{1}{2}\phi(x) + \sum_{i=0}^4 \frac{1}{10}\phi(x-2+i)$$
   - Multiple diskrete Peaks
   - Test für Bandbreitenauswahl

3. Sawtooth (Nr. 27):
   $$f_{27}(x) = \sum_{i=1}^5 \frac{1}{5}\phi(x-i)$$
   - Periodische Struktur
   - Gleichmässige Peaks

```{r data_generation}

# Grundeinstellungen
set.seed(42)
n_samples <- 5000
selected_dist <- c(21, 23, 27)

# Generiere Daten für jede Verteilung
generate_samples <- function(dist_id, n) {
  data <- rberdev(n, dist_id)
  attr(data, "dist_name") <- nberdev(dist_id)
  return(data)
}

# Datengenerierung (vereinfacht ohne mclapply für Kompatibilität)
data_list <- lapply(selected_dist, function(d) {
  generate_samples(d, n_samples)
})

# Erstelle kombinierten Datensatz
data_all <- data.frame(
  x1 = data_list[[1]],
  x2 = data_list[[2]],
  x3 = data_list[[3]])

# Grid für wahre Dichten
x_grid <- seq(-4, 4, length.out = 200)
true_densities <- lapply(selected_dist, function(d) {
  dberdev(x_grid, d)
})
```

## 1.2 Explorative Analyse

```{r exploratory}
# Grundlegende Statistiken
basic_stats <- sapply(data_list, function(x) {
  c(Mean = mean(x),
    SD = sd(x),
    Skewness = mean((x-mean(x))^3)/sd(x)^3,
    Kurtosis = mean((x-mean(x))^4)/sd(x)^4,
    Modes = length(density(x)$y[which(diff(sign(diff(density(x)$y)))==-2)]+1))
})
rownames(basic_stats) <- c("Mittelwert", "Std.Abw.", "Schiefe", "Kurtosis", "Modalitäten")
colnames(basic_stats) <- sapply(selected_dist, nberdev)
knitr::kable(basic_stats, digits = 3)

# Visualisiere Marginalverteilungen
par(mfrow=c(2,2))
for(i in 1:3) {
  hist(data_list[[i]], freq=FALSE, breaks=50,
       main=paste("Verteilung:", nberdev(selected_dist[i])),
       xlab="x", ylab="Dichte")
  lines(x_grid, true_densities[[i]], col="red", lwd=2)
  lines(density(data_list[[i]]), col="blue", lty=2)
  legend("topright", c("Wahr", "KDE"), col=c("red", "blue"), lty=1:2)
}

# 3D Visualisierung
plot_ly(as.data.frame(data_all), 
        x = ~x1, 
        y = ~x2, 
        z = ~x3,
        type = "scatter3d", 
        mode = "markers",
        marker = list(
          size = 3,
          color = ~(x1+x2+x3),
          colorscale = "Viridis"
        )) %>%
  layout(scene = list(
    xaxis = list(title = "Marronite"),
    yaxis = list(title = "Claw"),
    zaxis = list(title = "Sawtooth")
  ))
```

# 2. Implementierung der Schätzmethoden

## 2.1 Transformation Forest

```{r transformation_forest}
# Funktion für bedingte Dichteschätzung
fit_transformation_forest <- function(data) {
  # Ensure data is data.frame
  data <- as.data.frame(data)

  models <- list(
    m1 = tram::BoxCox(x1 ~ 1, data = data),
    m2 = tram::BoxCox(x2 ~ x1, data = data),
    m3 = tram::BoxCox(x3 ~ x1 + x2, data = data)
  )

  predict_joint <- function(newdata, models) {
    d1 <- predict(models$m1, newdata = newdata, type = "density")
    d2 <- predict(models$m2, newdata = newdata, type = "density")
    d3 <- predict(models$m3, newdata = newdata, type = "density")
    return(d1 * d2 * d3)
  }

  return(list(models = models, predict_joint = predict_joint))
}

# Fitte TF Modell
tf_fit <- fit_transformation_forest(data_all)

# Evaluiere marginale Dichten
tf_marginals <- list(
  x1 = predict(tf_fit$models$m1, 
              newdata = data.frame(x1 = x_grid), 
              type = "density"),
  x2 = predict(tf_fit$models$m2, 
              newdata = data.frame(x1 = median(data_all$x1), 
                                 x2 = x_grid), 
              type = "density"),
  x3 = predict(tf_fit$models$m3, 
              newdata = data.frame(x1 = median(data_all$x1), 
                                 x2 = median(data_all$x2), 
                                 x3 = x_grid), 
              type = "density")
)
```

## 2.2 HDR-basierte Schätzung

```{r hdr_estimation}
# Funktion für HDR-Schätzung
fit_hdr <- function(data) {
  # Univariate HDRs
  hdr_univariate <- lapply(1:3, function(i) {
    hdr.den(data[[i]], prob = c(50, 90, 95))
  })

  # Multivariate HDR
  hdr_multivariate <- hdr(as.matrix(data))

  return(list(
    univariate = hdr_univariate,
    multivariate = hdr_multivariate
  ))
}

# Fitte HDR
hdr_fit <- fit_hdr(data_all)

# Extrahiere Dichten
hdr_densities <- lapply(1:3, function(i) {
  den <- density(data_all[[i]])
  approx(den$x, den$y, xout = x_grid)$y
})
```

## 2.3 NP Kernel Schätzung
```{r np_estimation}
# Funktion für NP-Schätzung
fit_np <- function(data) {
  # Stelle sicher, dass die Daten im richtigen Format sind
  data_matrix <- as.matrix(data)

  # Univariate Schätzungen
  np_univariate <- lapply(1:ncol(data_matrix), function(i) {
    npudens(data_matrix[,i])
  })

  # Multivariate Schätzung
  bw <- npudensbw(data_matrix, bwmethod = "normal-reference")
  np_multivariate <- npudens(bws = bw)

  return(list(
    univariate = np_univariate,
    multivariate = np_multivariate,
    bw = bw
  ))
}

# Fitte NP
np_fit <- fit_np(data_all)

# Extrahiere Dichten
np_densities <- lapply(np_fit$univariate, function(fit) {
  predict(fit, newdata = matrix(x_grid, ncol = 1))
})
```

## 2.4 KS Kernel Smoothing
```{r ks_estimation}
# Funktion für KS-Schätzung
fit_ks <- function(data) {
  # Konvertiere Daten in Matrix-Format
  data_matrix <- as.matrix(data)

  # Univariate Schätzungen mit optimaler Bandbreite
  ks_univariate <- lapply(1:ncol(data_matrix), function(i) {
    h <- hpi(data_matrix[,i])
    kde(x = data_matrix[,i, drop = FALSE], h = h)
  })

  # Multivariate Schätzung
  H <- Hpi(data_matrix)
  ks_multivariate <- kde(x = data_matrix, H = H)

  return(list(
    univariate = ks_univariate,
    multivariate = ks_multivariate,
    H = H
  ))
}

# Fitte KS
ks_fit <- fit_ks(data_all)

# Extrahiere Dichten
ks_densities <- lapply(1:length(ks_fit$univariate), function(i) {
  predict(ks_fit$univariate[[i]], x = matrix(x_grid, ncol = 1))
})
```
## 2.5 Adaptive Kerndichte
```{r kdensity_estimation}
# Funktion für adaptive Kerndichteschätzung
fit_kdensity <- function(data) {
  # Univariate Schätzungen
  kd_univariate <- lapply(1:ncol(as.data.frame(data)), function(i) {
    estimate <- try({
      kdensity(x = as.numeric(data[[i]]), 
               kernel = "gaussian",
               bw = "silverman")  # Verwende Silverman's Regel statt adjust/start
    }, silent = TRUE)

    if(inherits(estimate, "try-error")) {
      warning(paste("Fehler bei Schätzung für Variable", i))
      return(NULL)
    }
    return(estimate)
  })

  return(list(univariate = kd_univariate))
}

# Fitte kdensity
kd_fit <- fit_kdensity(data_all)

# Extrahiere Dichten (mit Fehlerprüfung)
kd_densities <- lapply(kd_fit$univariate, function(fit) {
  if(!is.null(fit)) {
    predict(fit, x = x_grid)
  } else {
    rep(NA, length(x_grid))
  }
})
```

## 2.6 KEDD Schätzung
```{r kedd_estimation}
# Funktion für KEDD-Schätzung
fit_kedd <- function(data) {
  data_matrix <- as.matrix(data)

  # Univariate Schätzungen
  kedd_univariate <- lapply(1:ncol(data_matrix), function(i) {
    x <- as.numeric(data_matrix[,i])

    # Berechne Bandbreite mit Silverman's Regel
    n <- length(x)
    sigma <- sd(x)
    h <- 1.06 * sigma * n^(-1/5)

    # Verwende kde statt dkde
    tryCatch({
      kde(x = x, h = h, kernel = "gaussian")
    }, error = function(e) {
      warning(paste("Fehler bei KEDD-Schätzung für Variable", i, ":", e$message))
      NULL
    })
  })

  return(list(
    univariate = kedd_univariate,
    bandwidths = sapply(1:ncol(data_matrix), function(i) {
      x <- as.numeric(data_matrix[,i])
      1.06 * sd(x) * length(x)^(-1/5)
    })
  ))
}

# Fitte KEDD
kedd_fit <- fit_kedd(data_all)

# Extrahiere Dichten (mit Fehlerprüfung)
kedd_densities <- lapply(1:length(kedd_fit$univariate), function(i) {
  fit <- kedd_fit$univariate[[i]]
  if(!is.null(fit)) {
    evaluate(fit, x_grid)
  } else {
    rep(NA, length(x_grid))
  }
})
```

# 3. Vergleichende Analyse

## 3.1 Marginale Dichtevergleiche

```{r marginal_comparison}
# Vergleichsplot-Funktion für marginale Dichten
plot_density_comparison <- function(dim_idx) {
  # Sammle alle Schätzungen und stelle sicher, dass sie die gleiche Länge haben
  estimates <- list(
    "Wahr" = true_densities[[dim_idx]]
  )

  # Füge weitere Schätzungen nur hinzu, wenn sie die richtige Länge haben
  if(length(tf_marginals[[dim_idx]]) == length(x_grid)) {
    estimates[["TF"]] <- tf_marginals[[dim_idx]]
  }
  if(length(hdr_densities[[dim_idx]]) == length(x_grid)) {
    estimates[["HDR"]] <- hdr_densities[[dim_idx]]
  }
  if(length(np_densities[[dim_idx]]) == length(x_grid)) {
    estimates[["NP"]] <- np_densities[[dim_idx]]
  }
  if(length(ks_densities[[dim_idx]]) == length(x_grid)) {
    estimates[["KS"]] <- ks_densities[[dim_idx]]
  }
  if(length(kd_densities[[dim_idx]]) == length(x_grid)) {
    estimates[["KDensity"]] <- kd_densities[[dim_idx]]
  }
  if(length(kedd_densities[[dim_idx]]) == length(x_grid)) {
    estimates[["KEDD"]] <- kedd_densities[[dim_idx]]
  }

  # Überprüfe die Längen und entferne NAs
  estimates <- lapply(estimates, function(x) {
    if(length(x) != length(x_grid) || any(is.na(x))) {
      return(NULL)
    }
    return(x)
  })
  estimates <- estimates[!sapply(estimates, is.null)]

  # Berechne Fehlermetriken für verfügbare Schätzungen
  if(length(estimates) > 1) {
    metrics <- sapply(estimates[-1], function(est) {
      if(!is.null(est)) {
        c(
          RMSE = sqrt(mean((est - estimates$Wahr)^2, na.rm = TRUE)),
          MAE = mean(abs(est - estimates$Wahr), na.rm = TRUE),
          KL = mean(estimates$Wahr * log(pmax(estimates$Wahr/pmax(est, 1e-10), 1e-10)), na.rm = TRUE)
        )
      } else {
        c(RMSE = NA, MAE = NA, KL = NA)
      }
    })
  } else {
    metrics <- matrix(NA, nrow = 3, ncol = 0)
    rownames(metrics) <- c("RMSE", "MAE", "KL")
  }

  # Plot 1: Dichtevergleich
  par(mfrow = c(2,2))
  matplot(x_grid, do.call(cbind, estimates), type = "l",
          col = viridis(length(estimates)), lty = 1:length(estimates),
          main = paste("Marginale Dichte", c("Marronite", "Claw", "Sawtooth")[dim_idx]),
          xlab = "x", ylab = "Dichte")
  legend("topright", names(estimates), col = viridis(length(estimates)),
         lty = 1:length(estimates), cex = 0.7)

  # Plot 2: Fehlermetriken (nur wenn Metriken verfügbar)
  if(ncol(metrics) > 0) {
    barplot(metrics["RMSE",], main = "RMSE", las = 2,
            col = viridis(ncol(metrics)))
  }

  # Plot 3: Detail der Peaks
  peak_region <- x_grid > -2 & x_grid < 2
  matplot(x_grid[peak_region], 
          do.call(cbind, lapply(estimates, function(d) d[peak_region])),
          type = "l", col = viridis(length(estimates)), lty = 1:length(estimates),
          main = "Detailansicht der Peaks",
          xlab = "x", ylab = "Dichte")

  # Plot 4: Absolute Fehler (nur wenn mehr als eine Schätzung verfügbar)
  if(length(estimates) > 1) {
    abs_errors <- sapply(estimates[-1], function(est) abs(est - estimates$Wahr))
    matplot(x_grid, abs_errors, type = "l",
            col = viridis(ncol(abs_errors)), lty = 1:ncol(abs_errors),
            main = "Absolute Fehler",
            xlab = "x", ylab = "Absoluter Fehler")
  }

  return(metrics)
}

# Erstelle Vergleiche für alle drei Dimensionen
metrics_list <- lapply(1:3, plot_density_comparison)
```

## 3.2 Trivariate Analyse

```{r trivariate_analysis}
# Berechne trivariate Dichten auf einem Gitter
grid_3d <- expand.grid(
  x1 = seq(-4, 4, length.out = 20),
  x2 = seq(-4, 4, length.out = 20),
  x3 = seq(-4, 4, length.out = 20)
)

# Schätze trivariate Dichten
densities_3d <- list(
  TF = apply(grid_3d, 1, function(p) {
    tf_fit$predict_joint(data.frame(t(p)), tf_fit$models)
  }),
  NP = predict(np_fit$multivariate, newdata = grid_3d),
  KS = predict(ks_fit$multivariate, x = as.matrix(grid_3d))
)

# Visualisiere 3D-Dichten auf verschiedenen Schnitten
plot_3d_slices <- function(densities_3d, grid_3d) {
  # Wähle einen mittleren Schnitt für jede Dimension
  mid_idx <- round(length(unique(grid_3d$x1))/2)

  for(method in names(densities_3d)) {
    # Reshape Dichten für Plotting
    density_matrix <- array(densities_3d[[method]], 
                          dim = c(20, 20, 20))

    # Erstelle Heatmaps für verschiedene Schnitte
    par(mfrow = c(2,2))
    # xy-Schnitt
    image(density_matrix[,,mid_idx], 
          main = paste(method, "- xy-Schnitt"),
          col = viridis(100))
    # xz-Schnitt
    image(density_matrix[,mid_idx,], 
          main = paste(method, "- xz-Schnitt"),
          col = viridis(100))
    # yz-Schnitt
    image(density_matrix[mid_idx,,], 
          main = paste(method, "- yz-Schnitt"),
          col = viridis(100))
  }
}

plot_3d_slices(densities_3d, grid_3d)
```
## 3.3 Vergleich der bedingten Dichten
```{r conditional_densities}
# Berechne bedingte Dichten für verschiedene Werte der Bedingung
calculate_conditional_densities <- function(data, x1_val, x2_val) {
  # TF bedingte Dichten
  tf_cond <- predict(tf_fit$models$m3, 
                    newdata = data.frame(x1 = x1_val, 
                                       x2 = x2_val, 
                                       x3 = x_grid),
                    type = "density")

  # KS bedingte Dichte mit kde
  data_matrix <- cbind(data$x3, data$x1, data$x2)
  H <- Hpi(data_matrix)
  ks_fit <- kde(x = data_matrix, H = H)

  eval_points <- cbind(x_grid,
                      matrix(rep(c(x1_val, x2_val), 
                               each = length(x_grid)), 
                           ncol = 2))

  ks_cond <- predict(ks_fit, x = eval_points) /
             predict(kde(x = cbind(data$x1, data$x2), H = H[2:3, 2:3]),
                    x = matrix(c(x1_val, x2_val), ncol = 2))

  # NP bedingte Dichte
  np_cond <- npcdens(ydat = data$x3,
                     cdat = cbind(data$x1, data$x2),
                     ceval = matrix(c(x1_val, x2_val), ncol = 2))

  return(list(
    TF = tf_cond,
    KS = ks_cond,
    NP = fitted(np_cond)
  ))
}

# Berechne bedingte Dichten für verschiedene Bedingungen
conditional_points <- expand.grid(
  x1 = quantile(data_all$x1, probs = c(0.25, 0.5, 0.75)),
  x2 = quantile(data_all$x2, probs = c(0.25, 0.5, 0.75))
)

# Berechne die bedingten Dichten mit Fehlerbehandlung
conditional_densities <- lapply(1:nrow(conditional_points), function(i) {
  p <- conditional_points[i,]
  tryCatch({
    calculate_conditional_densities(data_all, p$x1, p$x2)
  }, error = function(e) {
    warning(paste("Fehler bei Punkt", i, ":", e$message))
    return(NULL)
  })
})

# Visualisiere die bedingten Dichten
plot_conditional_densities <- function(densities, point_idx) {
  if(is.null(densities[[point_idx]])) return()

  dens <- densities[[point_idx]]
  point <- conditional_points[point_idx,]

  plot(x_grid, dens$TF, type = "l", col = "blue",
       main = sprintf("Bedingte Dichte bei x1=%.2f, x2=%.2f", 
                     point$x1, point$x2),
       xlab = "x3", ylab = "Dichte")
  lines(x_grid, dens$KS, col = "red")
  lines(x_grid, dens$NP, col = "green")
  legend("topright", c("TF", "KS", "NP"),
         col = c("blue", "red", "green"), lty = 1)
}

# Erstelle Plot-Grid für alle bedingten Dichten
par(mfrow = c(3,3))
for(i in 1:nrow(conditional_points)) {
  plot_conditional_densities(conditional_densities, i)
}
```

# 4. Performance-Analyse

## 4.1 Quantitative Performance-Metriken

```{r performance_metrics}
# Berechne Performance-Metriken für alle Methoden
calculate_performance <- function(data, true_densities, estimates) {
  # Sammle alle Metriken
  metrics <- data.frame(
    Methode = character(),
    RMSE = numeric(),
    LogLik = numeric(),
    KL_Div = numeric(),
    Runtime = numeric(),
    stringsAsFactors = FALSE
  )

  # Zeitmessung und Metrikberechnung für jede Methode
  for(method in names(estimates)) {
    start_time <- Sys.time()

    # Berechne Metriken
    rmse <- sqrt(mean((estimates[[method]] - true_densities)^2))
    loglik <- sum(log(estimates[[method]] + 1e-10))
    kl_div <- mean(true_densities * log(true_densities/(estimates[[method]] + 1e-10)))
    runtime <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

    metrics <- rbind(metrics, data.frame(
      Methode = method,
      RMSE = rmse,
      LogLik = loglik,
      KL_Div = kl_div,
      Runtime = runtime
    ))
  }

  return(metrics)
}

# Berechne Performance für alle Dimensionen
performance_results <- lapply(1:3, function(i) {
  estimates <- list(
    TF = tf_marginals[[i]],
    HDR = hdr_densities[[i]],
    NP = np_densities[[i]],
    KS = ks_densities[[i]],
    KDensity = kd_densities[[i]],
    KEDD = kedd_densities[[i]]
  )

  perf <- calculate_performance(data_all[[i]], true_densities[[i]], estimates)
  perf$Dimension <- c("Marronite", "Claw", "Sawtooth")[i]
  return(perf)
})

# Kombiniere Ergebnisse
all_results <- do.call(rbind, performance_results)

# Visualisiere Performance
plots <- list()
metrics <- c("RMSE", "LogLik", "KL_Div", "Runtime")

par(mfrow=c(2,2))
for(metric in metrics) {
  boxplot(as.formula(paste0(metric, " ~ Methode")), data = all_results,
          main = metric, las = 2, col = viridis(length(unique(all_results$Methode))))
}
```

## 4.2 Peak-Detection-Analyse

```{r peak_detection}
# Analyse der Peak-Erkennung
analyze_peaks <- function(density_values, threshold = 0.1) {
  # Finde lokale Maxima
  peaks <- which(diff(sign(diff(density_values))) == -2) + 1
  peak_heights <- density_values[peaks]

  # Filtere signifikante Peaks
  significant_peaks <- peaks[peak_heights > threshold * max(peak_heights)]

  return(list(
    n_peaks = length(significant_peaks),
    peak_locations = x_grid[significant_peaks],
    peak_heights = peak_heights[peak_heights > threshold * max(peak_heights)]
  ))
}

# Vergleiche Peak-Erkennung für alle Methoden
peak_comparison <- lapply(1:3, function(i) {
  estimates <- list(
    True = true_densities[[i]],
    TF = tf_marginals[[i]],
    HDR = hdr_densities[[i]],
    NP = np_densities[[i]],
    KS = ks_densities[[i]],
    KDensity = kd_densities[[i]],
    KEDD = kedd_densities[[i]]
  )

  peaks <- lapply(estimates, analyze_peaks)

  # Erstelle Vergleichsplot
  plot(x_grid, estimates$True, type = "l", lwd = 2,
       main = paste("Peak-Vergleich -", c("Marronite", "Claw", "Sawtooth")[i]),
       xlab = "x", ylab = "Dichte")

  for(method in names(peaks)) {
    points(peaks[[method]]$peak_locations, 
           peaks[[method]]$peak_heights,
           col = which(names(peaks) == method),
           pch = 19)
  }

  legend("topright", names(peaks),
         col = 1:length(peaks),
         pch = 19)

  return(sapply(peaks, function(p) p$n_peaks))
})

# Tabellarische Zusammenfassung der Peak-Erkennung
peak_summary <- data.frame(
  Methode = names(peak_comparison[[1]]),
  Marronite = peak_comparison[[1]],
  Claw = peak_comparison[[2]],
  Sawtooth = peak_comparison[[3]]
)

knitr::kable(peak_summary, caption = "Anzahl erkannter Peaks pro Methode")
```

## 4.3 Bandbreitenanalyse

```{r bandwidth_analysis}
# Vergleiche Bandbreitenauswahl der verschiedenen Methoden
bandwidth_comparison <- data.frame(
  Dimension = rep(c("Marronite", "Claw", "Sawtooth"), each = 4),
  Methode = rep(c("NP", "KS", "KDensity", "KEDD"), 3),
  Bandbreite = c(
    np_fit$bw$xbw[1], ks_fit$univariate[[1]]$h, 
    kd_fit$univariate[[1]]$bw, kedd_fit$univariate[[1]]$h,
    np_fit$bw$xbw[2], ks_fit$univariate[[2]]$h,
    kd_fit$univariate[[2]]$bw, kedd_fit$univariate[[2]]$h,
    np_fit$bw$xbw[3], ks_fit$univariate[[3]]$h,
    kd_fit$univariate[[3]]$bw, kedd_fit$univariate[[3]]$h
  )
)

# Visualisiere Bandbreitenvergleich
ggplot(bandwidth_comparison, 
       aes(x = Methode, y = Bandbreite, fill = Dimension)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Vergleich der gewählten Bandbreiten",
       x = "Methode", y = "Bandbreite") +
  scale_fill_viridis_d()
```

### 2025-03-07 Code/2024_12_06/trivariate2 and halfmoon.Rmd
07_03_2025

```yaml
---
title: "Trivariate und Bivariate Dichteanalyse mit Benchden-Verteilungen"
author: "Léon Kia Faro"
date: "2024-12-05"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    number_sections: true
    theme: united
    highlight: tango
---
```

# Einleitung

- Ziel: Multivariate Dichteschätzung durch Faktorisierung in univariate bedingte Dichten.  
- Vergleich mit klassischen Schätzern (np, ks) und Normalizing Flows.  
- Fokus:  
  - *Trivariate Daten* aus Benchden-Verteilungen (für einfache Demonstration).  
  - *Bivariate Daten* (Half-Moon), um komplexe Strukturen zu testen.  
- Verwendung von Transformation Forests (tram) und klassischen Kernel-Methoden.  
- Schrittweiser Aufbau: Erst klassische Methoden, dann Fortgeschrittenes (Normalizing Flows).

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

# Pakete
library(benchden)      # Synthetische Verteilungen
library(tram)          # Transformation Models (für bedingte Dichten)
library(np)            # Nichtparametrische Dichteschätzung
library(ks)            # Kernel Smoothing für mehrdimensionale Dichten
library(kdensity)      # Univariate Kernel Dichte
library(kedd)          # Univariate Kernel Dichtebandweite
library(lpdensity)     # Lokale polynomiale Dichte (univariat)
library(hdrcde)        # High-Density Regions
library(viridis)       # Farbpalette
library(torch)         # Für Normalizing Flows (Deep Learning)
library(corrplot)      # Korrelationen visualisieren
```

# 1. Benchden-Verteilungen

## 1.1 Verteilungsauswahl

- Benchmark-Verteilungen für Experimente.  
- Erzeugung von synthetischen, teils komplexen Dichten.  
- Beispiel: Skewed Bimodal, Claw, Smooth Comb.

```{r}
# Zeige alle verfügbaren Verteilungen aus dem benchden-Package
cat("Alle verfügbaren Verteilungen aus benchden:\n")
for(i in 1:28) {
  cat(sprintf("%2d: %s\n", i, nberdev(i)))
}
```

# 2. Datengenerierung (Trivariates Beispiel)

- Generiere drei Variablen (y1,y2,y3) aus drei verschiedenen Verteilungen.
- Idee: p(y1,y2,y3) faktorisieren.
- Später: Vergleich mit np, ks und Transformation Forests.

```{r}
set.seed(123)
n <- 1000
verteilungen <- c(22, 23, 24) # Skewed bimodal, Claw, Smooth Comb
namen <- sapply(verteilungen, nberdev)
daten <- list()

# Daten generieren
for(i in seq_along(verteilungen)) {
  daten[[i]] <- rberdev(n, verteilungen[i])
}

data_all <- data.frame(
  y1 = daten[[1]],
  y2 = daten[[2]],
  y3 = daten[[3]]
)
```

# 3. Explorative Analyse (Trivariat)

- Univariate Statistiken  
- Visualisierung wahre vs. geschätzte Dichte (Kerne)  
- HDR-Analyse prüfen (Achtung: Manche Verteilungen liefern evtl. nicht mehrere HDR-Intervalle)

```{r}
stats <- data.frame(
  Verteilung = namen,
  Minimum = sapply(daten, min),
  Maximum = sapply(daten, max),
  Mittelwert = sapply(daten, mean),
  Median = sapply(daten, median),
  SD = sapply(daten, sd),
  Schiefe = sapply(daten, function(x) mean((x-mean(x))^3)/sd(x)^3),
  Kurtosis = sapply(daten, function(x) mean((x-mean(x))^4)/sd(x)^4)
)
stats
```

```{r}
# Vergleich wahre Dichte vs. Kernel Estimate

x_seq <- seq(-4,4,length=200)
for(i in seq_along(verteilungen)) {
  hist(daten[[i]], breaks=30, freq=FALSE,
       main=paste("Verteilung:", namen[i]),
       xlab="Wert", ylab="Dichte")
  # Wahre Dichte
  lines(x_seq, dberdev(x_seq, verteilungen[i]), col="red", lwd=2)
  # Kernel Estimate
  lines(density(daten[[i]]), col="blue", lty=2)
  legend("topright", c("True", "Estimate"), col=c("red","blue"), lty=c(1,2))
}
```

```{r}
# HDR-Analyse (High Density Regions)
hdr_test <- hdr.den(daten[[1]], prob=c(50, 90, 95))
# Prüfen ob hdr vorhanden
if(!is.null(hdr_test$hdr)) {
  hdr_test$hdr
} else {
  cat("Keine HDR-Intervalle gefunden.\n")
}
```

## Korrelationen (Trivariat)

- Korrelationen zwischen y1,y2,y3 untersuchen.
- 2D-Heatmap per corrplot.
- 3D-Darstellung als Höhenfeld.

```{r}
cor_matrix <- cor(data_all)
cor_matrix
```

```{r}
# 2D-Korrelationsheatmap
corrplot(cor_matrix, method="color", addCoef.col="black",
         tl.col="black", tl.srt=45, main="2D Korrelationsübersicht")
```

```{r}
# 3D-Darstellung der Korrelationsmatrix
x_grid <- 1:nrow(cor_matrix)
y_grid <- 1:ncol(cor_matrix)
z_mat <- cor_matrix
persp(x_grid, y_grid, z_mat, theta=30, phi=30, expand=0.5, col="lightblue",
      ltheta=120, shade=0.75, ticktype="detailed", main="3D Darstellung der Korrelationsmatrix")
```

# 4. Transformation Forest Faktorisierung

- Modellierung: p(y1), p(y2|y1), p(y3|y1,y2) mit tram.  
- Jede bedingte Dichte separat schätzbar.  
- Anschließend Multiplikation zur Gesamtverteilung.

```{r}
tf_model1 <- tram::BoxCox(y1 ~ 1, data = data_all)
tf_model2 <- tram::BoxCox(y2 ~ y1, data = data_all)
tf_model3 <- tram::BoxCox(y3 ~ y1 + y2, data = data_all)

get_conditional_density <- function(model, newdata, grid_size=100) {
  # Modellierte bedingte Dichte via grid von y-Werten
  response_var <- names(model$data)[1]
  y_range <- range(model$data[[response_var]])
  y_grid <- seq(y_range[1], y_range[2], length.out=grid_size)
  newdata <- as.data.frame(newdata)
  densities <- matrix(NA, nrow=nrow(newdata), ncol=length(y_grid))
  for(i in 1:nrow(newdata)) {
    extended_data <- newdata[rep(i, grid_size), , drop=FALSE]
    extended_data[[response_var]] <- y_grid
    dens_vals <- predict(model, newdata=extended_data, type="density")
    densities[i, ] <- dens_vals
  }
  rowMeans(densities)
}
```

# 5. Vergleich Faktorisierung vs. Direkte Schätzung (np, ks)

- np-Paket: npudens für multivariate Kernel-Schätzung.  
- ks-Paket: kde für multivariate Kernel-Schätzung.  
- Faktorisiertes Modell: p(y1,y2,y3) = p(y1)*p(y2|y1)*p(y3|y1,y2).  
- Log-Likelihood Vergleich als Performance-Indikator.

```{r}
eval_data <- data_all

ll_factorized <- sum(
  log(get_conditional_density(tf_model1, data.frame(y1=eval_data$y1))) +
  log(get_conditional_density(tf_model2, eval_data["y1"])) +
  log(get_conditional_density(tf_model3, eval_data[c("y1","y2")]))
)

# Direct (NP)
np_bw <- npudensbw(eval_data)
np_est <- npudens(np_bw)
ll_np <- sum(log(fitted(np_est)))

# Direct (KS)
H <- Hpi(as.matrix(eval_data))
ks_est <- kde(as.matrix(eval_data), H=H)
ll_ks <- sum(log(predict(ks_est, x=as.matrix(eval_data))))

results <- data.frame(
  Method = c("Factorized (TF)", "Direct (NP)", "Direct (KS)"),
  LogLik = c(ll_factorized, ll_np, ll_ks)
)
results
```

```{r}
par(mfrow=c(2,2))
plot(tf_model1, main="p(y1)", xlab="y1", ylab="Dichte")
plot(tf_model2, main="p(y2|y1=median)", xlab="y2", ylab="Dichte")
plot(tf_model3, main="p(y3|y1=y2=median)", xlab="y3", ylab="Dichte")

barplot(results$LogLik,
        names.arg=results$Method,
        main="Log-Likelihood Vergleich",
        ylab="Log-Likelihood",
        las=2, cex.names=0.9, cex.axis=0.9)
```

# 6. Normalizing Flows (simple)

- Nutzung von torch, einfacher Flow ohne vollständige Korrektur des Jacobians.  
- Nur Demonstration.

```{r}
NormalizingFlow <- nn_module(
  "NormalizingFlow",
  initialize = function(dim = 3, hidden = 32) {
    self$net <- nn_sequential(
      nn_linear(dim, hidden),
      nn_relu(),
      nn_linear(hidden, hidden),
      nn_relu(),
      nn_linear(hidden, dim)
    )
  },
  forward = function(x) {
    transformed <- self$net(x)
    log_det <- torch_zeros(x$size(1))
    list(transformed, log_det)
  }
)

flow <- NormalizingFlow(dim=3)
optimizer <- optim_adam(flow$parameters, lr=0.01)
data_tensor <- torch_tensor(as.matrix(eval_data))

# Training NF (vereinfachtes Training, wenige Epochen)
for(epoch in 1:100) {
  optimizer$zero_grad()
  out <- flow(data_tensor)
  z <- out[[1]]
  log_det <- out[[2]]
  loss <- -torch_mean(torch_sum(-0.5 * z^2 - 0.5 * log(2*pi), dim=2) + log_det)
  loss$backward()
  optimizer$step()
}

nf_log_density <- function(x_mat) {
  x_tensor <- torch_tensor(as.matrix(x_mat))
  out <- flow(x_tensor)
  z <- out[[1]]
  logpdf_z <- -0.5*torch_sum(z^2, dim=2)-0.5*ncol(x_mat)*log(2*pi)
  as.numeric(logpdf_z)
}

ll_nf <- sum(nf_log_density(eval_data))
results_nf <- rbind(results, data.frame(Method="Normalizing Flow (NF)", LogLik=ll_nf))
results_nf
```

```{r}
barplot(results_nf$LogLik,
        names.arg=results_nf$Method,
        main="Log-Likelihood Vergleich (inkl. NF)",
        ylab="Log-Likelihood",
        las=2, cex.names=0.9, cex.axis=0.9)
```

# 7. Bivariates Half-Moon Problem

- Half-Moon Daten für komplexere Strukturen.  
- Vergleich von TF, NP, KS, NF.  
- Marginale Dichten mit kdensity, kedd, lpdensity.

```{r, error=FALSE}

set.seed(42)
n <- 1000
sigma <- 0.1

# Erster Halbmond
t1 <- seq(-pi/2, pi/2, length.out = n/2)
x1 <- cbind(cos(t1), sin(t1))*1.2
x1[,2] <- x1[,2] + 0.8
x1 <- x1 + matrix(rnorm(n,0,sigma), ncol=2)

# Zweiter Halbmond
t2 <- seq(pi/2, 3*pi/2, length.out = n/2)
x2 <- cbind(cos(t2), sin(t2))*1.2
x2[,2] <- x2[,2] - 0.8
x2 <- x2 + matrix(rnorm(n,0,sigma), ncol=2)

# Rotation der Halbmonde
rotation <- function(x, angle) {
  theta <- angle*pi/180
  R <- matrix(c(cos(theta),-sin(theta),
                sin(theta),cos(theta)), 2,2)
  t(apply(x,1,function(p) R%*%p))
}

angle <- 45
x1_rot <- rotation(x1, angle)
x2_rot <- rotation(x2, angle)
x1_rot[,1] <- x1_rot[,1] + 0.5
x2_rot[,1] <- x2_rot[,1] + 0.5

data_hm <- data.frame(
  x = c(x1_rot[,1], x2_rot[,1]),
  y = c(x1_rot[,2], x2_rot[,2]),
  class = factor(rep(c("C1","C2"), each=n/2))
)

grid_size <- 50
x_range <- range(data_hm$x)
y_range <- range(data_hm$y)

grid_hm <- expand.grid(
  x = seq(x_range[1], x_range[2], length.out=grid_size),
  y = seq(y_range[1], y_range[2], length.out=grid_size)
)

# Transformation Forest
tf_fit_hm <- tram::BoxCox(y ~ x, data = data_hm)
tf_density_hm_vals <- predict(tf_fit_hm, newdata=grid_hm, type="density")
tf_density_hm <- matrix(tf_density_hm_vals, grid_size, grid_size)

# NP Schätzung
np_fit_hm <- npudens(bws=npudensbw(data_hm[,c("x","y")]))
np_pred_hm <- predict(np_fit_hm, newdata=grid_hm)
np_density_hm <- matrix(np_pred_hm, grid_size, grid_size)

# KS Schätzung
H_hm <- Hpi(as.matrix(data_hm[,c("x","y")]))
ks_fit_hm <- kde(x=as.matrix(data_hm[,c("x","y")]), H=H_hm)
ks_density_hm <- matrix(predict(ks_fit_hm, x=as.matrix(grid_hm[,c("x","y")])), grid_size, grid_size)

# Univariate Dichten für x
kd_fit_x <- kdensity(data_hm$x)

# Bandweite mit h.ucv aus kedd
bw_hm_obj <- h.ucv(data_hm$x, deriv.order=0, kernel="gaussian")
bw_hm <- bw_hm_obj$h

# dkde statt dkernel aus kedd
res_kedd <- dkde(x = data_hm$x, deriv.order=0, kernel="gaussian",
                 eval.points = seq(x_range[1], x_range[2], length.out=100))
kedd_dens <- res_kedd$est.fx

# lpdensity
lp_res <- lpdensity(data_hm$x, grid=seq(x_range[1], x_range[2], length.out=100), bwselect="imse-dpi")

# Normalizing Flow
NormalizingFlow2D <- nn_module(
  "NormalizingFlow2D",
  initialize = function(dim=2, hidden=32) {
    self$net <- nn_sequential(
      nn_linear(dim,hidden),
      nn_relu(),
      nn_linear(hidden,hidden),
      nn_relu(),
      nn_linear(hidden,dim)
    )
  },
  forward = function(x) {
    transformed <- self$net(x)
    log_det <- torch_zeros(x$size(1))
    list(transformed, log_det)
  }
)

flow_hm <- NormalizingFlow2D(dim=2)
optimizer_hm <- optim_adam(flow_hm$parameters, lr=0.01)
data_tensor_hm <- torch_tensor(as.matrix(data_hm[,c("x","y")]))

for(epoch in 1:100) {
  optimizer_hm$zero_grad()
  out_hm <- flow_hm(data_tensor_hm)
  z_hm <- out_hm[[1]]
  log_det_hm <- out_hm[[2]]
  loss_hm <- -torch_mean(torch_sum(-0.5*z_hm^2 -0.5*log(2*pi), dim=2)+log_det_hm)
  loss_hm$backward()
  optimizer_hm$step()
}

nf_log_density_hm <- function(x_mat) {
  x_tensor <- torch_tensor(as.matrix(x_mat))
  out <- flow_hm(x_tensor)
  z <- out[[1]]
  logpdf_z <- -0.5*torch_sum(z^2, dim=2)-0.5*ncol(x_mat)*log(2*pi)
  as.numeric(logpdf_z)
}

# Log-Likelihood Berechnungen
ll_tf_hm <- sum(log(tf_density_hm_vals+1e-12))
ll_np_hm <- sum(log(np_pred_hm+1e-12))
# Wichtig: predict(ks_fit_hm,...) nur mit passender Dimension:
ll_ks_hm <- sum(log(predict(ks_fit_hm, x=as.matrix(data_hm[,c("x","y")]))+1e-12))
ll_nf_hm <- sum(nf_log_density_hm(data_hm[,c("x","y")]))

results_hm <- data.frame(
  Method=c("TF","NP","KS","NF"),
  LogLik=c(ll_tf_hm, ll_np_hm, ll_ks_hm, ll_nf_hm)
)

par(mfrow=c(2,3))
plot(data_hm$x, data_hm$y,
     pch=ifelse(data_hm$class=="C1",4,16),
     col=ifelse(data_hm$class=="C1","blue","red"),
     main="1. Half-Moon Data",
     xlab="x", ylab="y", xlim=c(-1,2), ylim=c(-1.5,1.5))

plot_density_hm <- function(density, title) {
  image(seq(x_range[1], x_range[2], length.out=grid_size),
        seq(y_range[1], y_range[2], length.out=grid_size),
        density,
        col=viridis(50),
        main=title, xlab="x", ylab="y")
  points(data_hm$x, data_hm$y,
         pch=ifelse(data_hm$class=="C1",4,16),
         col="white", cex=0.5)
}

plot_density_hm(tf_density_hm, "2. TF Density")
plot_density_hm(np_density_hm, "3. NP Density")
plot_density_hm(ks_density_hm, "4. KS Density")

nf_grid_vals <- nf_log_density_hm(grid_hm[,c("x","y")])
nf_density_hm <- matrix(exp(nf_grid_vals), grid_size, grid_size)
plot_density_hm(nf_density_hm, "5. NF Density")

par(mfrow=c(1,1), mar=c(7,4,3,2))
barplot(results_hm$LogLik, names.arg=results_hm$Method,
        las=2, main="Log-Likelihood Vergleich (Half-Moon)",
        ylab="Log-Likelihood", cex.names=0.9)

x_seq_hm <- seq(x_range[1], x_range[2], length.out=100)
kd_x_vals <- kd_fit_x(x_seq_hm)
plot(x_seq_hm, kd_x_vals, type="l", col="red",
     main="Marginale Dichten x (Half-Moon)",
     xlab="x", ylab="Dichte")

```

# 8. Visualisierung Half-Moon

- Original Daten plotten.  
- Dichten von TF, NP, KS, NF.  
- Marginale Dichte von x: kdensity, kedd, lpdensity.

```{r, error=FALSE}
par(mfrow=c(2,3))
plot(data_hm$x, data_hm$y,
     pch=ifelse(data_hm$class=="C1",4,16),
     col=ifelse(data_hm$class=="C1","blue","red"),
     main="1. Half-Moon Data",
     xlab="x", ylab="y", xlim=c(-1,2), ylim=c(-1.5,1.5))

plot_density_hm <- function(density, title) {
  image(seq(x_range[1], x_range[2], length.out=grid_size),
        seq(y_range[1], y_range[2], length.out=grid_size),
        density,
        col=viridis(50),
        main=title, xlab="x", ylab="y")
  points(data_hm$x, data_hm$y,
         pch=ifelse(data_hm$class=="C1",4,16),
         col="white", cex=0.5)
}

plot_density_hm(tf_density_hm, "2. TF Density")
plot_density_hm(np_density_hm, "3. NP Density")
plot_density_hm(ks_density_hm, "4. KS Density")

nf_grid_vals <- nf_log_density_hm(grid_hm)
nf_density_hm <- matrix(exp(nf_grid_vals), grid_size, grid_size)
plot_density_hm(nf_density_hm, "5. NF Density")

par(mfrow=c(1,1), mar=c(7,4,3,2))
barplot(results_hm$LogLik, names.arg=results_hm$Method,
        las=2, main="Log-Likelihood Vergleich (Half-Moon)",
        ylab="Log-Likelihood", cex.names=0.9)

x_seq_hm <- seq(x_range[1], x_range[2], length.out=100)
kd_x_vals <- kd_fit_x(x_seq_hm)
plot(x_seq_hm, kd_x_vals, type="l", col="red",
     main="Marginale Dichten x (Half-Moon)",
     xlab="x", ylab="Dichte")

```

### 2025-03-07 Code/2025_02_21/04_analyze.R

@@ -26,9 +26,9 @@ analyze_models <- function(dtrain, dtest=NULL) {
         "\n[Test ] PM  LogLik =", logLik_on_data(pm_dtest), "\n")
   }
   
-  kl_tf <- calc_kl_divergence(500, function(n) my_dgp_d(N=n, d=3, models=models, cond=cond), tf$predict_joint)
-  kl_kd <- calc_kl_divergence(500, function(n) my_dgp_d(N=n, d=3, models=models, cond=cond), function(dd) kd$predict_kde(dd))
-  kl_pm <- calc_kl_divergence(500, function(n) my_dgp_d(N=n, d=3, models=models, cond=cond), pm$predict_joint)
+  kl_tf <- calc_kl_divergence(100, function(n) my_dgp_d(N=n, d=3, models=models, cond=cond), tf$predict_joint)
+  kl_kd <- calc_kl_divergence(100, function(n) my_dgp_d(N=n, d=3, models=models, cond=cond), function(dd) kd$predict_kde(dd))
+  kl_pm <- calc_kl_divergence(100, function(n) my_dgp_d(N=n, d=3, models=models, cond=cond), pm$predict_joint)
   
   cat("\nApprox KL Divergence:\n",
       "TF =", kl_tf, "\n",
@@ -40,7 +40,6 @@ analyze_models <- function(dtrain, dtest=NULL) {
     LLtrain = c(TF = tf_ll, KDE = kd_ll, PM = pm_ll)
   ))
 }
-
 
 
 

### 2025-03-07 Code/2025_03_07/00_setup.R

@@ -1,0 +1,22 @@
+# 00_setup.R
+
+library(tram)       # diverse transformations
+library(trtf)       # transformation forests
+library(ks)         # kernel smoothing
+library(partykit)   # partykit
+library(party)      # party
+library(foreach)    # parallel
+library(doParallel) # parallel
+
+set.seed(123)       # fix seed
+options(digits=4)   # 4 decimals
+
+init_parallel <- function(ncores=2) {
+  cl <- makeCluster(ncores)    # cluster creation
+  registerDoParallel(cl)       # register
+  cl                           # return cluster
+}
+
+
+
+

### 2025-03-07 Code/2025_03_07/04_analyze.R

@@ -1,0 +1,83 @@
+# 04_analyze.R
+# -------------------------------------------------------------------
+# Funktionen zur Modellbewertung: Log-Likelihood, KL-Divergenz etc.
+# -------------------------------------------------------------------
+
+logLik_on_data <- function(dvals) {
+  sum(log(pmax(dvals, 1e-15)))
+}
+
+calc_kl_divergence <- function(N, dgp_fun, model_predict, true_logdens_fun) {
+  samp <- dgp_fun(N)
+  log_ftrue <- true_logdens_fun(samp)
+  fhat      <- model_predict(samp)
+  log_fhat  <- log(pmax(fhat, 1e-15))
+  mean(log_ftrue - log_fhat)
+}
+
+analyze_models <- function(dtrain, dtest = NULL,
+                           config, distLibrary,
+                           showTest = TRUE) {
+  # Nur Demo für d=3-Fall:
+  tf <- fit_transformation_forest(dtrain)
+  kd <- fit_kernel_density_3d(dtrain)
+  pm <- fit_simple_param(dtrain)
+  
+  tf_dtrain <- tf$predict_joint(dtrain)
+  kd_dtrain <- kd$predict_kde(dtrain)
+  pm_dtrain <- pm$predict_joint(dtrain)
+  
+  tf_ll <- logLik_on_data(tf_dtrain)
+  kd_ll <- logLik_on_data(kd_dtrain)
+  pm_ll <- logLik_on_data(pm_dtrain)
+  
+  cat("\n[Train] TRFT LogLik =", tf_ll,
+      "\n[Train] KDE  LogLik =", kd_ll,
+      "\n[Train] PM   LogLik =", pm_ll, "\n")
+  
+  if(!is.null(dtest) && showTest) {
+    tf_dtest <- tf$predict_joint(dtest)
+    kd_dtest <- kd$predict_kde(dtest)
+    pm_dtest <- pm$predict_joint(dtest)
+    cat("\n[Test ] TRFT LogLik =", logLik_on_data(tf_dtest),
+        "\n[Test ] KDE  LogLik =", logLik_on_data(kd_dtest),
+        "\n[Test ] PM   LogLik =", logLik_on_data(pm_dtest), "\n")
+  }
+  
+  true_logdens_fun <- function(df) {
+    compute_logdensity(df, d=3, config=config, distLibrary=distLibrary)
+  }
+  
+  kl_tf <- calc_kl_divergence(
[…]

### 2025-03-07 Code/2025_03_07/06_KLDiv_Viz.R

@@ -1,0 +1,109 @@
+
+# 06_KLDiv_Viz.R
+# Erweiterte Visualisierung der KL-Divergenz und Dichtevergleiche
+
+
+# 1) Balkendiagramm mit optionalen Fehlerbalken
+#    Barplot der KL-Divergenz
+viz_kl_divergence_bar <- function(kl_list, kl_se=NULL) {
+  methods <- names(kl_list)
+  bp <- barplot(height = kl_list,
+                main   = "KL-Divergenz (niedriger = besser)",
+                xlab   = "Methoden",
+                ylab   = "KL Divergence",
+                col    = c("steelblue", "tomato", "forestgreen"),
+                ylim   = c(0, 1.1 * max(kl_list + if(is.null(kl_se)) 0 else kl_se, na.rm=TRUE))
+  )
+  if(!is.null(kl_se)) {
+    stopifnot(length(kl_se) == length(kl_list))
+    arrows(x0 = bp, y0 = kl_list - kl_se,
+           x1 = bp, y1 = kl_list + kl_se,
+           angle = 90, code = 3, length = 0.05, col = "black")
+  }
+  axis(1, at = bp, labels = methods, tick = FALSE)
+}
+
+# 2) Linienplot der KL-Divergenz über verschiedene Stichprobengrößen
+#    kl_values: Matrix, Zeilen=Methoden, Spalten=verschiedene N
+viz_kl_vs_n <- function(sample_sizes, kl_values) {
+  matplot(x = sample_sizes, y = t(kl_values), type = "b", pch = 19,
+          lty = 1, col = c("steelblue", "tomato", "forestgreen"),
+          xlab = "Stichprobengröße (N)", ylab = "KL Divergence",
+          main = "KL-Divergenz vs. Sample-Size")
+  legend("topright", legend = rownames(kl_values),
+         col = c("steelblue", "tomato", "forestgreen"), pch = 19, lty = 1)
+}
+
+# 3) Overlay von true vs. estimated Dichte (1D)
+#    Zeigt wahre vs. geschätzte Dichteverläufe
+viz_density_overlay <- function(xvals, dens_true, dens_est, main_title = "True vs. Estimated") {
+  plot(xvals, dens_true, type = "l", col = "red", lwd = 2,
[…]

### 2025-03-07 Code/2025_02_21/00_setup.R
07_03_2025

@@ -11,3 +11,4 @@ library(doParallel)
 
 set.seed(123)
 options(digits=4)
+

### 2025-03-07 Code/2025_02_21/01_distribution.R

@@ -51,3 +51,5 @@ cond <- list(
     list(alpha = a_, beta = b_)
   }
 )
+
+

### 2025-03-07 Code/2025_03_07/02_dgp.R

@@ -1,0 +1,148 @@
+# 02_dgp.R
+
+clamp_parameters <- function(paramList, lower_bound=1e-8, upper_bound=1e6, debug=FALSE) {
+  for(nm in names(paramList)) {
+    vec <- paramList[[nm]]
+    if(is.numeric(vec)) {
+      bad <- !is.finite(vec)
+      if(any(bad)&&debug) cat("clamp: NA/NaN/Inf in", nm,"\n")
+      vec[bad] <- lower_bound
+      lo <- vec<lower_bound
+      if(any(lo)&&debug) cat("clamp:",nm,"<",lower_bound,"\n")
+      vec[lo] <- lower_bound
+      hi <- vec>upper_bound
+      if(any(hi)&&debug) cat("clamp:",nm,">",upper_bound,"\n")
+      vec[hi] <- upper_bound
+      paramList[[nm]] <- vec
+    }
+  }
+  paramList
+}
+
+generate_data <- function(N,d,config,distLibrary,debug=FALSE) {
+  if(debug) cat("generate_data\n")
+  Y <- matrix(NA,nrow=N,ncol=d)
+  for(j in seq_len(d)) {
+    distName <- config[[j]]$distName
+    paramFun <- config[[j]]$paramFun
+    if(debug) {
+      cat("\nj=",j,"distName=",distName,"\n")
+      cat("paramFun formals:",names(formals(paramFun)),"\n")
+    }
+    if(j==1) {
+      pList <- paramFun()
+      if(debug) {
+        cat("paramFun() ->\n")
+        str(pList)
+      }
+    } else {
+      prev <- as.data.frame(Y[,1:(j-1),drop=FALSE])
+      frmls <- names(formals(paramFun))
+      if(debug) {
+        cat("colnames before:",colnames(prev),"\n")
+        cat("frmls:",frmls,"\n")
+      }
+      n1 <- ncol(prev)
+      n2 <- length(frmls)
+      if(n2>0) colnames(prev)[1:min(n1,n2)] <- frmls[1:min(n1,n2)]
+      if(debug) {
+        cat("colnames after:",colnames(prev),"\n")
+        cat("calling paramFun\n")
+      }
+      pList <- do.call(paramFun,prev)
+      if(debug) {
+        cat("paramFun ->\n")
+        str(pList)
+      }
+    }
+    pList <- clamp_parameters(pList,debug=debug)
[…]

### 2025-03-07 Code/2025_03_07/03_models.R

@@ -1,0 +1,99 @@
+# 03_models.R
+
+fit_transformation_forest <- function(data) {
+  # Beispiel: Y2>0 => log(Y2)
+  data$Y2_log <- log(pmax(data$Y2, 1e-15))
+  
+  # Univariates Transformationsmodell
+  d1 <- BoxCox(Y1 ~ 1, data = data)
+  d2 <- BoxCox(Y2_log ~ 1, data = data)
+  d3 <- Colr(Y3 ~ 1, data = data, bounds = c(0, 1))
+  
+  # Forest-Fits
+  f2 <- traforest(d2, formula = Y2_log ~ Y1, data = data,
+                  ntree = 100, mtry = 1)
+  f3 <- traforest(d3, formula = Y3 ~ Y1 + Y2, data = data,
+                  ntree = 100, mtry = 2)
+  
+  mods <- list(m1 = d1, m2 = d2, m3 = d3, f2 = f2, f3 = f3)
+  
+  get_density_value <- function(obj) {
+    val <- if(is.list(obj)) obj[[1]] else obj
+    max(val, 1e-15)
+  }
+  
+  predict_joint <- function(newdata) {
+    n <- nrow(newdata)
+    out1 <- numeric(n)
+    out2 <- numeric(n)
+    out3 <- numeric(n)
+    
+    for(i in seq_len(n)) {
+      d1o <- predict(mods$m1, newdata[i, , drop=FALSE],
+                     type="density", q=newdata$Y1[i],
+                     simplify=FALSE)
+      out1[i] <- get_density_value(d1o)
+      
+      if(newdata$Y2[i] <= 0) {
+        out2[i] <- 1e-15
+      } else {
+        z2  <- log(newdata$Y2[i])
+        tmp <- newdata[i, , drop=FALSE]
+        tmp$Y2_log <- z2
+        d2o <- predict(mods$f2, newdata=tmp,
+                       type="density", q=z2,
+                       simplify=FALSE)
+        out2[i] <- get_density_value(d2o) * (1 / newdata$Y2[i])
+      }
+      
+      d3o <- predict(mods$f3, newdata[i, , drop=FALSE],
+                     type="density", q=newdata$Y3[i],
+                     simplify=FALSE)
+      out3[i] <- get_density_value(d3o)
+    }
+    
[…]

### 2025-03-07 Code/2025_03_07/05_viz.R

@@ -1,0 +1,95 @@
+# 05_viz.R
+
+# Diese erweiterte Funktion zeigt einerseits einen 1D-Schnitt p(y3 | y1=cval1, y2=cval2)
+# und andererseits einen 2D-Konturplot p(y3, y2 | y1=cval1).
+# So erhält man sowohl die direkte Form von p(y3|...) in 1D als auch den Überblick
+# über die gemeinsame Abhängigkeit von (y3, y2) bei fixem y1 in 2D.
+
+visualize_1D_and_2D <- function(dtrain,
+                                tf_predict,
+                                kd_predict,
+                                pm_predict,
+                                fix_y1 = NULL,
+                                fix_y2 = NULL) {
+  # Falls fix_y1/fix_y2 nicht gegeben, nehmen wir Mediane:
+  if(is.null(fix_y1)) fix_y1 <- median(dtrain$Y1)
+  if(is.null(fix_y2)) fix_y2 <- median(dtrain$Y2)
+  
+  # Panel 1 (links): 1D-Schnitt p(y3 | Y1=fix_y1, Y2=fix_y2)
+  seqY3 <- seq(min(dtrain$Y3), max(dtrain$Y3), length.out=200)
+  df_1d <- data.frame(Y1=fix_y1, Y2=fix_y2, Y3=seqY3)
+  
+  # Modell-Dichten:
+  tf_val_1d <- tf_predict(df_1d)
+  kd_val_1d <- kd_predict(df_1d)
+  pm_val_1d <- pm_predict(df_1d)
+  
+  # Rand p(Y1=fix_y1, Y2=fix_y2) => Errechnung bedingter Dichte p(y3|y1,y2)
+  # d.h. p(y1=fix_y1, y2=fix_y2)
+  df_marg_1d <- data.frame(Y1=fix_y1, Y2=fix_y2, Y3=0)
+  tf_marg_1d <- tf_predict(df_marg_1d)[1]
+  kd_marg_1d <- kd_predict(df_marg_1d)[1]
+  pm_marg_1d <- pm_predict(df_marg_1d)[1]
+  
+  tf_cond_1d <- tf_val_1d / pmax(tf_marg_1d, 1e-15)
+  kd_cond_1d <- kd_val_1d / pmax(kd_marg_1d, 1e-15)
[…]

### 2025-05-17 Archive/triangular_transport_per_step.Rmd
17_05

---
title: "Data generation and Map Transport (Block 1)"
author: "Léon Kia Faro"
date: "06_May_2025"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE, warning = FALSE, message = FALSE)

```

# Einleitung

*   R Skript: Block 1 (Daten) & Block 2 (Modell).
*   **Block 1**: Wahres $K$-dim Ziel $\pi(x)$. Wahre Maps $S$ (Ziel $\to$ Ref) & $S^{-1}$ (Ref $\to$ Ziel). Ref $\eta(z)$ ist $N(0,I_K)$.
    *   Dichte: $\pi(x) = \prod_{k=1}^{K} f_{d_k}(x_k | x_{<k-1})$ (Paper, K2, S4).
    *   Map $S$: $X \sim \pi \to Z \sim \eta$. $U_k = F_{d_k}(X_k | X_{<k-1})$, $Z_k = \Phi^{-1}(U_k)$ (Paper, K3, S9).
    *   $F_{d_k}$ wahre bed. CDF. $\Phi$ Std.-Normal-CDF.
    *   Map $S^{-1}$ erzeugt $X \sim \pi$ aus $Z \sim \eta$.
*   **Block 2**: Modelle aus Daten. Ohne Info aus Block 1 (ausser Eval.).
*   Dokument: Block 1 R Code. Überblick Block 2 & Info-Fluss.

# Block 1: Transport-Mapping

Tabelle: Komponenten Block 1. Definieren Datengenerierung and Map Transport.

| Schritt | Abschnitt                     | Ziel                                                                 | Code                       | Notizen                                                                 |
|------|-------------------------------------|----------------------------------------------------------------------|----------------------------|-------------------------------------------------------------------------|
| **1** | Komponenten-Spez.             | Wahre Struktur $K$-dim Ziel $\pi(x)$ durch bed. Komp. | `config`, `lam()`, `alpha()` | `config` = Wahrheit. Wichtig für Daten $\pi(x)$. Block 2: kein Zugriff. |
| **2** | Verteilungshelfer                | PDF, CDF, Quantilfkt. Std.-Vert.                   | `pdf_k()`, `cdf_k()`, `qtf_k()` | R Basis. Wahrheits-abh. mit Param. aus `config`.                       |
| **3** | Zieldichte                      | Exakte Dichte $\pi(x)$ für $x$.                                       | `pi_density()`             | Wahre bed. Spez. Nur Eval./Bench. Nicht für Training Block 2. (Paper) |
| **4** | Referenzverteilung              | Stichproben aus $K$-dim Gauss $N(0, I_K)$.                           | `eta_sample()`             | Zufall $\to$ Stichproben $\pi(x)$.                                         |
| **5** | Vorwärts-Map $S$                   | Daten $X \sim \pi \to Z \sim N(0, I_K)$ mit wahren bed. CDFs.       | `S_map()`                  | $Z=S(X)$. Braucht wahre $F_{d_k}$. (Paper)                                 |
| **6** | Inverse Map $S^{-1}$                | $Z \sim N(0, I_K) \to X \sim \pi$ mit wahren bed. Quantilfkt.       | `S_inv()`                  | $X=S^{-1}(Z)$. Kern Daten $\pi(x)$. Braucht wahre $F^{-1}_{d_k}$. (Paper) |
| **7** | Jakobi + Likelihood             | $\log|\det\nabla S|$ & $\ell(x)$ unter wahrer Map $S$.                | `det_J()`, `loglik()`      | Wichtig für wahre Likelihood. $\ell(x) = \dots$ (Paper)                   |
| **8** | Diagn. Round-Trip             | Prüft $S^{-1}(S(X)) \approx X$ & $S(S^{-1}(U)) \approx U$.               | `round_trip_test()`        | Plausibilität $S, S^{-1}$.                                              |
| **9** | Wahre Log-Dichte (Wrap)        | Für $\log \pi(x)$.                                                     | `log_pi()`                 | Abkürzung `pi_density(..., log=TRUE)`.                                  |
| **10**| Sampling $\pi$ via $S^{-1}$    | i.i.d. $X \sim \pi$ mit wahrer $S^{-1}$.                             | `pi_sample()`              | Hauptweg für Trainings-/Testdaten aus $\pi$.                           |
| **11**| Sampling-Diagnostik              | $X \sim \pi$ mit wahren $\ell(x)$.                                    | `sample_and_loglik()`      | Debugging / Datensätze mit Wahrheits-Likelihood.                      |
| **12**| Bedingtes Sampling              | $\pi(X_{\text{frei}} | X_{\text{fix}})$ durch partielle Inv. $S$.       | `cond_sample()`            | Bedingte Generierung (wahres Modell). (Paper)                         |

```{r start, include=FALSE}
N_demo <- 3 # Demo Stichproben
eps <- 1e-10 # Num. Stabilität

lam_coeff   <- c(1, 0.1) # Koeffs lam(x1)
alpha_coeff <- c(1, 1)   # Koeffs alpha(x2)

lam <- function(x1){ # Fkt Lambda
  base <- lam_coeff[1] + lam_coeff[2] * x1 # Basiswert
  pmin(20, pmax(0.05, base)) # Klemmen
}

alpha <- function(x2){ # Fkt Alpha
  base <- alpha_coeff[1] + alpha_coeff[2] * x2 # Basiswert
  pmax(1e-3, base) # Klemmen
}

config <- list( # Ziel-Vert. Konfig
  list(distr = "norm",  parm = NULL), # X1 ~ N(0,1)
  list(distr = "exp",   parm = function(d) list(rate  = lam(d$X1))), # X2|X1
  list(distr = "gamma", parm = function(d) list(shape = alpha(d$X2), rate  = 1)) # X3|X2
)

# Step 2 distribution helpers
dist_fun <- function(pref, d) get(paste0(pref, d))

get_pars <- function(k, x_prev, cfg){
  ck <- cfg[[k]]
  if (is.null(ck$parm)) return(list())
  names(x_prev) <- paste0("X", seq_along(x_prev))
  ck$parm(as.data.frame(as.list(x_prev)))
}

pdf_k <- function(k, xk, x_prev, cfg, log = FALSE)
  do.call(dist_fun("d", cfg[[k]]$distr),
          c(list(xk), get_pars(k, x_prev, cfg), list(log = log)))

cdf_k <- function(k, xk, x_prev, cfg)
  do.call(dist_fun("p", cfg[[k]]$distr),
          c(list(xk), get_pars(k, x_prev, cfg)))

qtf_k <- function(k, uk, x_prev, cfg)
  do.call(dist_fun("q", cfg[[k]]$distr),
          c(list(uk), get_pars(k, x_prev, cfg)))

# Step 3 target density
pi_density <- function(x, cfg = config, log = FALSE){
  if (is.null(dim(x))) x <- matrix(x, nrow = 1)
  lp <- apply(x, 1, function(row)
    sum(vapply(seq_along(row),
               function(k) pdf_k(k, row[k], row[seq_len(k - 1)], cfg, log = TRUE),
               numeric(1))))
  if (log) lp else exp(lp)
}

# Step 4 reference distribution
eta_sample <- function(N = 1L, K = length(config)){
  Z <- matrix(rnorm(N * K), nrow = N)
  U <- pnorm(Z)
  list(U = U, Z = Z)
}

# Step 5 forward map S
S_map <- function(X, cfg = config, eps = 1e-10){
  if (is.null(dim(X))) X <- matrix(X, nrow = 1)
  clamp <- function(u) pmin(1 - eps, pmax(eps, u))
  U <- t(apply(X, 1, function(row)
    vapply(seq_along(row),
           function(k) clamp(cdf_k(k, row[k], row[seq_len(k - 1)], cfg)),
           numeric(1))))
  Z <- qnorm(U)
  list(U = U, Z = Z)
}

# Step 6 inverse map S^{-1}
S_inv <- function(U, cfg = config){
  if (is.null(dim(U))) U <- matrix(U, nrow = 1)
  Z <- qnorm(U); n <- nrow(U); K <- ncol(U)
  X <- matrix(NA_real_, n, K)
  for (i in seq_len(n)){
    x_prev <- numeric(0)
    for (k in seq_len(K)){
      x_prev[k] <- qtf_k(k, U[i, k], x_prev[seq_len(k - 1)], cfg)
    }
    X[i, ] <- x_prev
  }
  logd <- matrix(NA_real_, n, K)
  for (i in seq_len(n))
    for (k in seq_len(K))
      logd[i, k] <- pdf_k(k, X[i, k], X[i, seq_len(k - 1)], cfg, log = TRUE) -
    dnorm(Z[i, k], log = TRUE)
  list(X = X, Z = Z, logd = logd)
}

# Step 7 Jacobian and likelihood
det_J <- function(logd) rowSums(logd)

loglik <- function(Z, logdet) -0.5 * rowSums(Z ^ 2) -
  (ncol(Z) / 2) * log(2 * pi) + logdet

# Step 8 diagnostic round‑trip
round_trip_test <- function(U, cfg = config, tol = 1e-10){
  inv  <- S_inv(U, cfg)
  fwd  <- S_map(inv$X, cfg)
  maxZ <- max(abs(fwd$Z - inv$Z))
  inv2 <- S_inv(fwd$U, cfg)
  maxX <- max(abs(inv2$X - inv$X))
  c(maxZ = maxZ, maxX = maxX) <= tol
}

# Step 9 true log density
log_pi <- function(x, cfg = config) pi_density(x, cfg, log = TRUE)

# Step 10 sampling from π via S^{-1}
pi_sample <- function(N = 1L, cfg = config){
  ref <- eta_sample(N, length(cfg))
  inv <- S_inv(ref$U, cfg)
  list(X = inv$X, U = ref$U, Z = inv$Z, logd = inv$logd)
}

# Step 11 sampling diagnostics helper (optional)
sample_and_loglik <- function(N = 1L, cfg = config){
  samp <- pi_sample(N, cfg)
  detJ <- det_J(samp$logd)
  ll   <- loglik(samp$Z, detJ)
  cbind(samp$X, loglik = ll)
}

# Step 12 conditional sampling
cond_sample <- function(x_fixed, k, N = 1L, cfg = config){
  K <- length(cfg)
  stopifnot(k >= 1, k < K, length(x_fixed) == k)
  # CDF values for fixed coordinates
  U_fixed <- numeric(k)
  for (j in seq_len(k))
    U_fixed[j] <- cdf_k(j, x_fixed[j], x_fixed[seq_len(j - 1)], cfg)
  Z_fixed <- qnorm(U_fixed)
  X_out <- matrix(NA_real_, N, K)
  for (i in seq_len(N)){
    z <- rnorm(K); z[1:k] <- Z_fixed
    u <- pnorm(z)
    X_out[i, ] <- S_inv(matrix(u, nrow = 1), cfg)$X
  }
  colnames(X_out) <- paste0("X", 1:K)
  X_out
}

```

```{r start2, include=FALSE, echo=FALSE}
set.seed(42)

## Demo-Parameter & Hilfsfunktionen (für alle Schritte)
N_demo <- 3                 # Anzahl Demo-Stichproben
eps     <- 1e-10            # numerische Stabilität

lam_coeff   <- c(1, 0.1)    # λ-Koeffizienten
alpha_coeff <- c(1, 1)      # α-Koeffizienten

lam <- function(x1){        # λ(x1) – geklemmt
  base <- lam_coeff[1] + lam_coeff[2] * x1
  pmin(20, pmax(0.05, base))
}
alpha <- function(x2){      # α(x2) – geklemmt
  base <- alpha_coeff[1] + alpha_coeff[2] * x2
  pmax(1e-3, base)
}

```
# Block 1: Code-Durchlauf

## Schritt 1: Komponenten-Spez.
Section 2.3, Eq. (5), p. 9 — Definition der triangulären Map $S$ mit Komponenten $S_k(x_{1:k})$.

*   **Prozess**:
    *   `lam(x1)`: Exp Rate $\lambda(x_1)$. Geklemmt.
    *   `alpha(x2)`: Gamma Shape $\alpha(x_2)$. Geklemmt.
    *   `config`: Liste $K$-dim Vert. $D$. $K=3$: $X_1 \sim N(0,1)$, $X_2|X_1 \sim Exp(\lambda(X_1))$, $X_3|X_2 \sim Gamma(\alpha(X_2),1)$.
*   **Out**: `lam`, `alpha`, `config`. Demo: $K$, Vert.-Namen.
*   **Info**: Def. wahre bed. Vert. für $\pi(x)$. (Paper)

```{r step1_chunk}
N_demo <- 3 # Demo Stichproben
eps <- 1e-10 # Num. Stabilität

lam_coeff   <- c(1, 0.1) # Koeffs lam(x1)
alpha_coeff <- c(1, 1)   # Koeffs alpha(x2)

lam <- function(x1){ # Fkt Lambda
  base <- lam_coeff[1] + lam_coeff[2] * x1 # Basiswert
  pmin(20, pmax(0.05, base)) # Klemmen
}

alpha <- function(x2){ # Fkt Alpha
  base <- alpha_coeff[1] + alpha_coeff[2] * x2 # Basiswert
  pmax(1e-3, base) # Klemmen
}

config <- list( # Ziel-Vert. Konfig
  list(distr = "norm",  parm = NULL), # X1 ~ N(0,1)
  list(distr = "exp",   parm = function(d) list(rate  = lam(d$X1))), # X2|X1
  list(distr = "gamma", parm = function(d) list(shape = alpha(d$X2), rate  = 1)) # X3|X2
)

## Demo
cat("K =", length(config),
    "; distributions =", paste(vapply(config, `[[`, "", "distr"),
                               collapse = ", "), "\n")

```

## Schritt 2: Verteilungshelfer

Section 2.2, Eq. (3), p. 6 — Change-of-variables-Formel, die PDF, CDF $\bigl(F_{d_k}\bigr)$ und Quantil­funktionen verbindet.

* **Inputs**

  * `pref`, `d` → R-Präfix („d/p/q“) & Verteilungsname
  * `k`, `x_prev` → Index der Komponente & vorherige $X_{<k}$-Werte
  * `cfg` → Konfiguration der wahren Bedingungen
  * `xk`, `uk`, `log` → aktueller Wert, Uniform-Wert, Log-Flag

* **Ablauf**

  1. `dist_fun()` baut Funktionsnamen (z. B. `dnorm`) und holt ihn.
  2. `get_pars()` berechnet Parameter für Komponente *k* aus `x_prev`.
  3. `pdf_k()`, `cdf_k()`, `qtf_k()` rufen die passende R-Funktion mit diesen Parametern auf.

* **Outputs**

  * Gewählte R-Verteilungsfunktion
  * Parameter-Liste
  * Gewünschter Wert: PDF, CDF oder Quantil

```{r step2_chunk}
dist_fun <- function(pref, d) get(paste0(pref, d)) # R Vert. Fkt holen

get_pars <- function(k, x_prev, cfg){ # Param. k-te Komp.
  ck <- cfg[[k]] # k-te Konfig
  if (is.null(ck$parm)) return(list()) # Keine Param -> leer
  names(x_prev) <- paste0("X", seq_along(x_prev)) # X-Werte benennen
  ck$parm(as.data.frame(as.list(x_prev))) # Param. Fkt aufrufen
}

pdf_k <- function(k, xk, x_prev, cfg, log = FALSE) # PDF k-te Komp.
  do.call(dist_fun("d", cfg[[k]]$distr), # Dichte Fkt
          c(list(xk), get_pars(k, x_prev, cfg), list(log = log))) # Aufruf

cdf_k <- function(k, xk, x_prev, cfg) # CDF k-te Komp.
  do.call(dist_fun("p", cfg[[k]]$distr), # Vert. Fkt
          c(list(xk), get_pars(k, x_prev, cfg))) # Aufruf

qtf_k <- function(k, uk, x_prev, cfg) # Quantil k-te Komp.
  do.call(dist_fun("q", cfg[[k]]$distr), # Quantil Fkt
          c(list(uk), get_pars(k, x_prev, cfg))) # Aufruf

## Demo
densities <- vapply(seq_along(config), function(k){
  pars <- get_pars(k, numeric(k - 1), config)
  par_str <- if (length(pars))
    paste0(" (", paste(names(pars), round(unlist(pars), 3),
                       sep = "=", collapse = ", "), ")") else ""
  paste0(k, ": ", config[[k]]$distr, par_str)
}, character(1))
print(densities)
```

## Schritt 3: Zieldichte
Section 2.3.2, Eq. (14), p. 12 — Faktorisierung der Zieldichte $π(x)=\prod_{k}π(x_k\mid x_{<k})$.

* **Input**

  * `x` … Beobachtungsmatrix (Zeilen = Samples, Spalten = Dimensionen)
  * `cfg` … Ziel-Konfiguration aus Schritt 1
  * `log` … TRUE ⇒ log-Dichte, FALSE ⇒ Dichte
* **Berechnung**

  * pro Zeile: log PDF jeder Komponente via `pdf_k`
  * Summe über *k* (Komponenten)
  * falls `log = FALSE` ⇒ Exponentiale für echte Dichte
* **Output**

  * Vektor der Länge `nrow(x)` mit π(xᵢ) bzw. log π(xᵢ)

```{r step3_chunk, include=TRUE}
pi_density <- function(x, cfg = config, log = FALSE){        # Zieldichte π(x)

  lp <- apply(x, 1, function(row)                            # Zeilenweise
    sum(vapply(seq_along(row),                               # über k
               function(k) pdf_k(k, row[k], row[seq_len(k-1)],
                                 cfg, log = TRUE), numeric(1))))
  if (log) lp else exp(lp)                                   # Log/Dichte
}

## Demo Drei π-Samples + Dichte
x_demo <- pi_sample(N_demo)$X
print(round(cbind(x_demo, dens = pi_density(x_demo)), 5))

```

## Schritt 4: Referenzverteilung
Section 2.3, Figure 5, p. 8 — Standard-Normal-Referenz $η(z)$ und Beziehung $z=S(x)$.

* **Input (N, K)** – wie viele Stichproben *N* in welcher Dimension *K* erzeugt werden sollen
* **Ziehe Z** – erzeuge ein *N × K*-Gitter Standardnormal-Rauschen Z ∼ 𝒩(0, 1) (unabhängige Spalten)
* **Transformiere zu U** – wende die Normal-CDF Φ elementweise auf Z an ⇒ gleichverteilte Matrix U = Φ(Z) in \[0, 1]
* **Output** – gib `list(U = …, Z = …)` zurück; beide Matrizen tragen exakt dieselbe Zufalls­information in zwei Skalen
* **Zweck** – (U, Z) bilden die Basisverteilung η(z); Z ist der Startpunkt für die inverse Transport-Map S⁻¹, die daraus echte π-Samples X erzeugt

```{r step4_chunk}
eta_sample <- function(N = 1L, K = length(config)){ # Samples Ref.Vert N(0,I)
  Z <- matrix(rnorm(N * K), nrow = N) # N*K N(0,1) -> Matrix Z
  U <- pnorm(Z) # Z -> U (Uniform) via Normal-CDF
  list(U = U, Z = Z) # Liste U, Z
}

## Demo – Referenz-Samples η(z)
demo_ref <- eta_sample(N_demo)
cat("U (uniform):\n"); print(round(demo_ref$U, 4))
cat("Z (normal):\n");  print(round(demo_ref$Z, 4))

```

## Schritt 5 – Vorwärts-Map S
Section 2.3, Eq. (5), p. 9 — Vorwärtsauswertung $S(x)\to z$ über bedingte CDFs.

* **Input:** Matrix `X`, Konfiguration `cfg`, kleines `eps` zum Klemmen
* wandle jede Zeile: `U_k = F_{d_k}(X_k | X_{<k})`, dann `U_k ← clamp(U_k, eps)`
* Probit-Schritt: `Z_k = Φ⁻¹(U_k)` (Normalisierung)
* verarbeite alle *k* nacheinander, nutzt echte bedingte CDFs
* **Output:** Liste `list(U = …, Z = …)` – uniform + normal Darstellung derselben Daten

```{r step5_chunk}
S_map <- function(X, cfg = config, eps = 1e-10){ # Vorwärts-Map S: X -> Z
  if (is.null(dim(X))) X <- matrix(X, nrow = 1) # X zu Matrix
  clamp <- function(u) pmin(1 - eps, pmax(eps, u)) # Hilfsfkt: Klemmen
  U <- t(apply(X, 1, function(row) # Pro Zeile X (transp. U)
    vapply(seq_along(row), # Pro Komp k
           function(k) clamp(cdf_k(k, row[k], row[seq_len(k - 1)], cfg)), # U_k, klemmen
           numeric(1)))) # Output numerisch
  Z <- qnorm(U) # U -> Z via Inv. Normal-CDF
  list(U = U, Z = Z) # Liste U, Z
}

```

## Schritt 6 – Inverse-Map S⁻¹
Section 2.3.1, Eq. (7), p. 9 — Sequenzielle Inversion $S^{-1}(z)$ zur Rücktransformation $z\mapsto x$.

* **Input:** Matrix `U`, Konfiguration `cfg`
* Probit zurück: `Z = Φ⁻¹(U)`
* autoregressiv: `X_k = F_{d_k}^{-1}(U_k | X_{<k})` über *k*
* log-Jacobi-Diag: `logd_{ik} = log f_{d_k}(X_k|…) − log φ(Z_k)`
* **Output:** Liste `list(X = …, Z = …, logd = …)` für spätere Likelihoods

```{r step6_chunk}
S_inv <- function(U, cfg = config){ # Inverse Map S^-1: U -> X
  if (is.null(dim(U))) U <- matrix(U, nrow = 1) # U zu Matrix
  Z <- qnorm(U); n <- nrow(U); K <- ncol(U) # U -> Z; Dims
  X <- matrix(NA_real_, n, K) # X Matrix init
  for (i in seq_len(n)){ # Pro Sample i
    x_prev <- numeric(0) # X_1..X_{k-1} init
    for (k in seq_len(K)){ # Pro Dim k
      x_prev[k] <- qtf_k(k, U[i, k], x_prev[seq_len(k - 1)], cfg) # X_k berechnen
    }
    X[i, ] <- x_prev # Sample X_i speichern
  }
  logd <- matrix(NA_real_, n, K) # logd Matrix init
  for (i in seq_len(n)) # Pro Sample i
    for (k in seq_len(K)) # Pro Dim k
      logd[i, k] <- pdf_k(k, X[i, k], X[i, seq_len(k - 1)], cfg, log = TRUE) - # log f_k
                    dnorm(Z[i, k], log = TRUE) # - log phi(Z_k)
  list(X = X, Z = Z, logd = logd) # Liste X, Z, logd
}

## Demo – Inverse Map S⁻¹ ⇒ X
demo_inv <- S_inv(demo_ref$U)
cat("X = S_inv(U):\n"); print(round(demo_inv$X, 4))

```

## Schritt 7 – Jakobidet & Log-Likelihood
Section 2.3, Eq. (6), p. 9 — Berechnung von $|\det\nabla S|$ als Produkt der Diagonalelemente (Jacobian & Log-Likelihood).

* **Input:** Matrix `logd`, Matrix `Z`
* summiere Spalten → `log|det J|` pro Zeile
* Log-Likelihood: `ℓ(x) = −½‖Z‖² − (K/2) log(2π) + log|det J|`
* getrennt berechenbar dank Dreiecks-Jacobi
* **Output:** Vektoren `logdet`, `loglik`

```{r step7_chunk}
det_J <- function(logd) rowSums(logd) # Log-Det Jakobi

loglik <- function(Z, logdet_val) -0.5 * rowSums(Z ^ 2) - # Log-Likelihood X
  (ncol(Z) / 2) * log(2 * pi) + # Norm. Log-Gauss
  logdet_val # + log|det J|

## Demo – log|det J| & log-Likelihood
logdet     <- det_J(demo_inv$logd)
loglik_vec <- loglik(demo_ref$Z, logdet)
cat("log|detJ|:\n"); print(round(logdet, 4))
cat("loglik:\n");    print(round(loglik_vec, 4))

```

## Schritt 8 – Round-Trip-Diagnose
Section 2.3, Figure 5, p. 8 — Invertierbarkeit von $S$ ermöglicht Round-Trip-Diagnose $x\to z\to x$.

* **Input:** Matrix `U`, Konfiguration `cfg`, Toleranz `tol`
* erzeuge `X = S⁻¹(U)` → zurück  `U_rec, Z_rec = S(X)`
* erneut invertieren: `X_rec = S⁻¹(U_rec)`
* prüfe Max-Abweichungen < `tol` für Z und X
* **Output:** logisches Paar `(maxZ_ok, maxX_ok)` bestätigt Konsistenz

```{r step8_chunk}
round_trip_test <- function(U, cfg = config, tol = 1e-10){ # Round-Trip Test S, S_inv
  inv  <- S_inv(U, cfg) # X_gen = S_inv(U)
  fwd  <- S_map(inv$X, cfg, eps=eps) # (U_rec,Z_rec)=S(X_gen)
  maxZ <- max(abs(fwd$Z - inv$Z)) # Max Diff Z
  inv2 <- S_inv(fwd$U, cfg) # X_rec = S_inv(U_rec)
  maxX <- max(abs(inv2$X - inv$X)) # Max Diff X
  c(maxZ = maxZ, maxX = maxX) <= tol # Diff <= Tol?
}

## Demo – Round-Trip-Test
deltaZ <- max(abs(S_map(demo_inv$X)$Z - demo_ref$Z))
cat("max|ΔZ| =", signif(deltaZ, 3),
    "; round_trip =", round_trip_test(demo_ref$U), "\n")

```

## Schritt 9 – Wahre Log-Dichte (Wrapper)
Section 2.2, Eq. (4), p. 6 — Log-Dichte-Formel $ \log π(x)=\log η(S(x))+ \log|\det\nabla S(x)|$.

* **Input:** Beobachtungen `x`, Konfiguration `cfg`
* Aufruf `pi_density(x, cfg, log = TRUE)`
* vermeidet Code-Duplikation in späteren Analysen
* **Output:** Vektor `log π(x)` für alle Zeilen
* praktisch für Vergleich mit geschätzten Modellen

```{r step9_chunk}
log_pi <- function(x, cfg = config) pi_density(x, cfg, log = TRUE) # Wrapper wahre Log-Dichte

## Demo – Wahre log π(X)
cat("log_pi:\n"); print(round(log_pi(demo_inv$X), 4))

```

## Schritt 10 – π-Sampling via S⁻¹
 Section 2.3, p. 8 — Sampling von $x\sim π$ durch Ziehen $z\sim η$ und Anwenden von $S^{-1}$.

* **Input:** Anzahl `N`, Konfiguration `cfg`
* ziehe `(U, Z)` aus Referenz η mit `eta_sample`
* wandle mit `S_inv` → erhält `X`, `logd`
* liefert i.i.d. Trainings-/Test-Paare aus der wahren Dichte
* **Output:** Liste `X, U, Z, logd`

```{r step10_chunk}
pi_sample <- function(N = 1L, cfg = config){ # Samples Ziel pi via S_inv
  ref <- eta_sample(N, length(cfg)) # Samples (U,Z) Ref.Vert
  inv <- S_inv(ref$U, cfg) # U -> X via S_inv
  list(X = inv$X, U = ref$U, Z = inv$Z, logd = inv$logd) # Liste X,U,Z,logd
}

## Demo – Neue π-Stichprobe
samp <- pi_sample(N_demo)
cat("sample X:\n"); print(round(samp$X, 4))

```

## Schritt 11 – Sampling-Diagnose
Section 2.3, Eq. (6), p. 9 — Log-Likelihood $ℓ(x)= -\tfrac12\|S(x)\|^2 + \log|\det\nabla S|$ für Sample-Diagnostik.

* **Input:** Anzahl `N`, Konfiguration `cfg`
* erstelle Samples `X` via `pi_sample`
* berechne `log|det J|` und `ℓ(X)` mit Funktionen aus Schritt 7
* gebe beides neben `X` zurück – schnelle Qualitätskontrolle
* **Output:** Matrix `[X | loglik]` (Spalten K + 1)

```{r step11_chunk}
sample_and_loglik <- function(N = 1L, cfg = config){ # Diagnose: Samples pi + wahre LogL
  samp_res <- pi_sample(N, cfg) # N Samples X (U,Z,logd)
  detJ_val <- det_J(samp_res$logd) # log|det J|
  ll   <- loglik(samp_res$Z, detJ_val) # Log-Likelihoods
  cbind(samp_res$X, loglik = ll) # X + LogL
}

```

## Schritt 12 – Bedingtes Sampling
Section 2.3.2, Figure 6, p. 11 — Bedingtes Sampling durch Manipulation der Zwischenverteilung im inversen Pfad.

* **Input:** feste Koordinaten `x_fixed` (Länge k), Index `k`, Anzahl `N`, `cfg`
* bestimme zugehörige `U_fixed`, `Z_fixed` über CDF/Probit
* ziehe neue Normal-Zufallszahlen für freie Achsen, kombiniere zu `Z_cond`
* transformiere komplett mit `S_inv` → erhält bedingte Samples
* **Output:** Matrix `X_out (N × K)` – Draws aus π(X\_{k+1\:K} | X\_{1\:k}^\*)\`

```{r step12_chunk}
cond_sample <- function(x_fixed, k, N = 1L, cfg = config){ # Bedingtes Sampling
  K_total <- length(cfg) # Gesamt-Dim X
  stopifnot(k >= 1, k < K_total, length(x_fixed) == k) # Input Validierung

  U_fixed <- numeric(k) # U_1..U_k init
  for (j in seq_len(k)) { # Pro feste Komp.
    U_fixed[j] <- cdf_k(j, x_fixed[j], x_fixed[seq_len(j - 1)], cfg) # U_j berechnen
  }
  U_fixed_clamped <- pmin(1 - eps, pmax(eps, U_fixed)) # U_fixed klemmen
  Z_fixed <- qnorm(U_fixed_clamped) # Z_j für feste Komp.

  X_output <- matrix(NA_real_, N, K_total) # Output Matrix init

  for (i in seq_len(N)){ # Pro bed. Sample
    z_sample <- rnorm(K_total) # Frischer Z Vektor N(0,1)
    z_sample[1:k] <- Z_fixed # Feste Z-Werte setzen
    u_sample <- pnorm(z_sample) # Z -> U
    X_output[i, ] <- S_inv(matrix(u_sample, nrow = 1), cfg)$X # S_inv -> X
  }
  colnames(X_output) <- paste0("X", 1:K_total) # Spalten benennen
  X_output # N bed. Samples
}

## Demo – Bedingtes Sampling (gegeben X₁)
cond <- cond_sample(samp$X[1, 1], k = 1, N = N_demo)
cat("conditional sample given X1 =",
    round(samp$X[1, 1], 3), "\n")
print(round(cond, 4))

```

# Notationsübersicht

Tabelle: Symbole Paper vs. R.

| Symbol | Bedeutung (Paper) | R Name |
|---|---|---|
| **S** | Ziel $\to$ Ref Map $S:\mathbb{R}^K\to\mathbb{R}^K$ | `S` / `S_map()` |
| $S_k$ | $k$-te Map Komp. $u_k = S_k(x_k | x_{<k})$ | `S_k()` / `S_list[[k]]` |
| $S^{-1}$ | Inv. Map (Ref $\to$ Ziel) $S^{-1}: \mathbb{R}^K\to\mathbb{R}^K$ | `S_inv()` |
| $S^{-1}_k$ | Seq. Inv. $x_k = S^{-1}_k(u_k | x_{<k})$ | `S_inv_k()` |
| $\nabla S$ | Jakobi-Matrix $S$ | `J_S` |
| $\det\nabla S$ | Det. Jakobi $\nabla S$ | `det_J` |
| $\partial_{x_k}S_k$ | Diag. Jakobi $\frac{\partial u_k}{\partial x_k} = f_{d_k}(x_k|x_{<k})$ | `dSdx_k` / (in `logd`) |
| **$\pi$** | Zielvert. $\mathbb{R}^K$ | `pi_dist` |
| $\pi(x)$ | Ziel-PDF | `pi_density` |
| **$\eta$** | Ref.-Vert. $\mathbb{R}^K$, oft $\mathcal{N}(0,I_K)$ | `eta_dist` |
| $\eta(z)$ | Ref.-PDF | `eta_density` (`dnorm`) |
| **x** | $X \sim \pi$ | `X` (Matrix $N \times K$) |
| **z** | $Z \sim \eta$ | `Z` (Matrix $N \times K$) |
| $K$ | Dim. Ziel-/Ref.-Raum | `K` |
| $N$ | Anz. Samples | `N` |
| $L(x)$ | Likelihood $L(x) = \eta(S(x))\,|\det\nabla S(x)|$ | `likelihood` |
| $\ell(x)$ | Log-Likelihood $\log L(x)$ | `loglik` |
| $\Phi(\cdot)$ | Std.-Normal CDF | `pnorm()` |
| $\varphi(\cdot)$ | Std.-Normal PDF | `dnorm()` |
| $U_k$ | Unif. Var. $U_k = F_{d_k}(X_k|X_{<k})$ | `U` (Matrix $N \times K$) |
| $Z_k = \Phi^{-1}(U_k)$ | Probit Transf. | `Z_probit` / `qnorm(U)` |
| $\text{condSample}(\cdot)$ | Ziehe von $\pi(X_{k^*+1:K}\mid X_{1:k^*}^*)$ | `cond_sample()` |

# Block 2: Modellschätzung

Block 2: Modellschätzung aus Daten. Ohne Zugriff auf Block 1.

| Schritt   | Beschreibung                                          | Code (Konzept)          | Notizen                                                                 |
|--------|-------------------------------------------------------|--------------------------------------|-------------------------------------------------------------------------|
| **17** | Transf. Forest Schätzung                            | `train_TF()` (pro $k$)               | $S_k$ mit TF anpassen $\to \hat{F}_k(x_k | x_{<k})$. Nonparam. bed. CDFs. (Paper) |
| **18** | TF-basierte Map-Konstr.                               | `S_hat_TF()`                         | $\hat{S}$ aus Forests: $\hat{S}_k(x) = \Phi^{-1}(\hat{F}_k(x_k | x_{<k}))$. (Paper) |
| **19** | TF Log-Likelihood Eval.                               | `loglik_TF()`                        | Modell $\ell_{TF}(x)$ mit $\hat{f}_k$ & Jakobi $\hat{S}$. Modellbewertung.    |
| **20** | Empir. Marg. Transf. (Copula Vorv.)                   | `to_uniforms()`                      | $X_j \to U_j = \hat{F}_j(X_j)$. Copula auf unif. Marginalen.                 |
| **21** | Copula Anpassung                                      | `fit_copula()`                       | Multivariate Copula $\hat{C}(u_1, \dots, u_K)$ an Daten. (Paper)             |
| **22** | Copula Dichte Zusammenbau                             | `pi_hat_C()`                         | Volle Dichte $\hat{\pi}_C(x) = \hat{c}(\hat{u}) \prod \hat{f}_k(x_k)$. (Paper) |
| **23** | Normalizing Flow Arch.                                | `flow_model` (Spez.)         | Invert. neur. Netz $f_{\psi}: \mathbb{R}^K \to \mathbb{R}^K$. (Paper)              |
| **24** | Flow Training                                         | `train_NF()`                         | Parameter $\psi$ von $f_{\psi}$ anpassen. (Paper)                          |
| **25** | Flow Dichte Formel                                    | `density_NF()`                       | $\hat{\pi}_{NF}(x) = \eta(f_{\psi}(x)) \cdot |\det (\dots)|$. (Paper)         |
| **26** | KL-Divergenz Eval.                                    | `eval_KL()`                          | $D_{KL}(\pi \| \hat{\pi}_m)$. Braucht Wahrheit Block 1. (Paper)           |

# Informationsfluss Blöcke

Wichtig: Block 2 unabhängig von Block 1 Wahrheit.

| Schritt | Funktion(en)                      | Wahrheits-Info?                                                       | OK Block 1? | Risiko Block 2?      |
|------|----------------------------------|---------------------------------------------------------------------------|-----------------|-------------------------------------------------|
| 1-2  | `config`, `lam()`, `alpha()`     | Def. *wahre* bed. Param. $\lambda(x_1), \alpha(x_2)$, Vert. $d_k$.      | Ja (DGP Def.)   | **Ja**: Def. $\pi$; Block 2 darf nicht kennen. |
| 3    | `pi_density()`                   | Nutzt wahre $f_{d_k}(x_k|x_{<k})$ aus `config`.                           | Ja (Eval.)      | **Ja**: Falls Ziel für Training.  |
| 4    | `eta_sample()`                   | Nur Samples Ref. $\eta = N(0,I_K)$.                                       | Ja (Std.logik)  | Nein.                                           |
| 5    | `S_map()`                        | Nutzt wahre bed. CDFs $F_{d_k}$ aus `config`.                            | Ja (Map Def.)   | **Ja**: Falls Map Modell beeinflusst. |
| 6    | `S_inv()`                        | Nutzt wahre bed. Quantilfkt. $F_{d_k}^{-1}$ aus `config`.                   | Ja (Map Def.)   | **Ja**: Falls Map Modell beeinflusst. |
| 7    | `det_J()`, `loglik()`            | Wahre $f_{d_k}$, $S(x)$.                                                   | Ja (Wahr.Eval.) | **Ja**: Falls wahre Likelihood Training leitet. |
| 8    | `round_trip_test()`              | Kombiniert `S_map()`, `S_inv()`.                                          | Ja (Map QS)     | Nein (Interne Diagnose B1).                     |
| 9    | `log_pi()`                       | Wrapper `pi_density(..., log=TRUE)`.                                      | Ja (Wahr.Eval.) | **Ja**: Wie `pi_density()`.                  |
| 10   | `pi_sample()`                    | `eta_sample()` dann `S_inv()`.                                           | Ja (Daten Gen.) | Nein (Daten ok für B2).                         |
| 11   | `sample_and_loglik()`            | `pi_sample()` mit wahrer `loglik()`.                                       | Ja (Diag. Gen.) | Nein (wie oben, mit wahren Labels).             |
| 12   | `cond_sample()`                  | Wahre $F_{d_k}$, $F_{d_k}^{-1}$ für part. Inv.                             | Ja (Demo Wahr.) | **Ja**: Falls für bed. Modell genutzt.         |

---

### 2025-05-17 Archive/13_05_2025/main4.R

@@ -1,0 +1,169 @@
+suppressPackageStartupMessages({
+  library(extraDistr)  # für Laplace
+  library(tram)        # für Lm, Colr, BoxCox, mlt
+  library(trtf)        # optional
+  library(ggplot2)     # für Plots
+})
+
+# --- Block A: Konfiguration für K=4 ---
+clip <- function(x, lo = 1e-6, hi = 1e6) pmin(hi, pmax(lo, x))
+EPS <- 1e-10
+config4 <- list(
+  list(distr = "norm",    parm = NULL),
+  list(distr = "t",       parm = function(d) list(df = 3 + 0.5 * d$X1)),
+  list(distr = "laplace", parm = function(d) list(m = 0.3 * d$X2, s = clip(1 + 0.1 * d$X2))),
+  list(distr = "logis",   parm = function(d) list(location = 0.2 * d$X3, scale = clip(1 + 0.05 * d$X3)))
+)
+config <- config4; K <- length(config)
+
+# Numerisch stabile Hilfsfunktionen
+dist_fun <- function(pref, name) get(paste0(pref, name))
+safe_logpdf <- function(val, eps = EPS) log(pmax(val, eps))
+safe_cdf    <- function(val, eps = EPS) pmax(eps, pmin(1 - eps, val))
+get_pars <- function(k, x_prev, cfg) {
+  ck <- cfg[[k]]
+  if (is.null(ck$parm)) return(list())
+  names(x_prev) <- paste0("X", seq_along(x_prev))
+  ck$parm(as.data.frame(as.list(x_prev)))
+}
+pdf_k <- function(k, xk, x_prev, cfg, log = TRUE) {
+  pars <- get_pars(k, x_prev, cfg)
+  dens <- do.call(dist_fun("d", cfg[[k]]$distr), c(list(x = xk), pars, list(log = FALSE)))
+  if (log) safe_logpdf(dens) else dens
+}
+cdf_k <- function(k, xk, x_prev, cfg, log = TRUE) {
+  dname <- cfg[[k]]$distr; pars <- get_pars(k, x_prev, cfg)
+  args <- c(if(dname=="sn") list(p = xk) else list(q = xk), pars, list(log.p = FALSE))
+  cdfv <- do.call(dist_fun("p", dname), args)
[…]

### 2025-05-17 Code/2025_01_31/Splines.R

@@ -1,243 +1,0 @@
-###########################################################
-# PAKETE
-###########################################################
-library(copula)
-library(benchden)
-library(mlt)       # ctm(), numeric_var(), Bernstein_basis()
-library(trtf)      # z.B. traforest()
-library(np)        # optional Kernel
-library(ks)        # optional Kernel
-
-###########################################################
-# 1) Datengenerierung (3D- oder 2D-Fall)
-#    Normal-Copula + benchden
-###########################################################
-generate_data_copula_benchden <- function(n = 1000,
-                                          rho = 0.5,
-                                          dist_indices = c(11, 12, 5)) {
-  d <- length(dist_indices)
-  stopifnot(d %in% c(2, 3))
-  
-  # Copula definieren
-  if (d == 2) {
-    myCop <- normalCopula(param = rho, dim = 2)
-  } else {
-    corMat <- matrix(rho, nrow = 3, ncol = 3)
-    diag(corMat) <- 1
-    myCop <- normalCopula(param = corMat[upper.tri(corMat)], 
-                          dim = 3, dispstr = "un")
-  }
-  
-  set.seed(123)
-  U <- rCopula(n, myCop)
-  
-  data_mat <- matrix(NA, nrow = n, ncol = d)
-  for (j in seq_len(d)) {
-    data_mat[, j] <- qberdev(U[, j], dnum = dist_indices[j])
-  }
-  
-  df <- as.data.frame(data_mat)
-  names(df) <- paste0("Y", seq_len(d))
-  df
-}
-
-
-###########################################################
-# 2) "Splines"-basiertes Faktorisiertes Modell:
-#    p(Y1), p(Y2|Y1), p(Y3|Y1,Y2) via Bernstein-Basis
-###########################################################
-fit_factorized_splines <- function(df) {
-  d <- ncol(df)
-  stopifnot(d %in% c(2, 3))
-  names(df) <- paste0("Y", seq_len(d))
-  
-  # 1) p(Y1) univariat
-  respY1  <- numeric_var("Y1", support = range(df$Y1))
-  basisY1 <- Bernstein_basis(respY1, order = 5, ui = "increasing")
-  # ctm + fit
[…]

### 2025-05-17 Code/2025_02_21/06_KLDiv_Viz.R

@@ -1,109 +1,0 @@
-
-# 06_KLDiv_Viz.R
-# Erweiterte Visualisierung der KL-Divergenz und Dichtevergleiche
-
-
-# 1) Balkendiagramm mit optionalen Fehlerbalken
-#    Barplot der KL-Divergenz
-viz_kl_divergence_bar <- function(kl_list, kl_se=NULL) {
-  methods <- names(kl_list)
-  bp <- barplot(height = kl_list,
-                main   = "KL-Divergenz (niedriger = besser)",
-                xlab   = "Methoden",
-                ylab   = "KL Divergence",
-                col    = c("steelblue", "tomato", "forestgreen"),
-                ylim   = c(0, 1.1 * max(kl_list + if(is.null(kl_se)) 0 else kl_se, na.rm=TRUE))
-  )
-  if(!is.null(kl_se)) {
-    stopifnot(length(kl_se) == length(kl_list))
-    arrows(x0 = bp, y0 = kl_list - kl_se,
-           x1 = bp, y1 = kl_list + kl_se,
-           angle = 90, code = 3, length = 0.05, col = "black")
-  }
-  axis(1, at = bp, labels = methods, tick = FALSE)
-}
-
-# 2) Linienplot der KL-Divergenz über verschiedene Stichprobengrößen
-#    kl_values: Matrix, Zeilen=Methoden, Spalten=verschiedene N
-viz_kl_vs_n <- function(sample_sizes, kl_values) {
-  matplot(x = sample_sizes, y = t(kl_values), type = "b", pch = 19,
-          lty = 1, col = c("steelblue", "tomato", "forestgreen"),
-          xlab = "Stichprobengröße (N)", ylab = "KL Divergence",
-          main = "KL-Divergenz vs. Sample-Size")
-  legend("topright", legend = rownames(kl_values),
-         col = c("steelblue", "tomato", "forestgreen"), pch = 19, lty = 1)
-}
-
-# 3) Overlay von true vs. estimated Dichte (1D)
-#    Zeigt wahre vs. geschätzte Dichteverläufe
-viz_density_overlay <- function(xvals, dens_true, dens_est, main_title = "True vs. Estimated") {
-  plot(xvals, dens_true, type = "l", col = "red", lwd = 2,
[…]

### 2025-05-17 Code/2025_03_07/00_setup.R

@@ -1,22 +1,0 @@
-# 00_setup.R
-
-library(tram)       # diverse transformations
-library(trtf)       # transformation forests
-library(ks)         # kernel smoothing
-library(partykit)   # partykit
-library(party)      # party
-library(foreach)    # parallel
-library(doParallel) # parallel
-
-set.seed(123)       # fix seed
-options(digits=4)   # 4 decimals
-
-init_parallel <- function(ncores=2) {
-  cl <- makeCluster(ncores)    # cluster creation
-  registerDoParallel(cl)       # register
-  cl                           # return cluster
-}
-
-
-
-

### 2025-05-17 Code/2025_03_07/02_dgp.R

@@ -1,148 +1,0 @@
-# 02_dgp.R
-
-clamp_parameters <- function(paramList, lower_bound=1e-8, upper_bound=1e6, debug=FALSE) {
-  for(nm in names(paramList)) {
-    vec <- paramList[[nm]]
-    if(is.numeric(vec)) {
-      bad <- !is.finite(vec)
-      if(any(bad)&&debug) cat("clamp: NA/NaN/Inf in", nm,"\n")
-      vec[bad] <- lower_bound
-      lo <- vec<lower_bound
-      if(any(lo)&&debug) cat("clamp:",nm,"<",lower_bound,"\n")
-      vec[lo] <- lower_bound
-      hi <- vec>upper_bound
-      if(any(hi)&&debug) cat("clamp:",nm,">",upper_bound,"\n")
-      vec[hi] <- upper_bound
-      paramList[[nm]] <- vec
-    }
-  }
-  paramList
-}
-
-generate_data <- function(N,d,config,distLibrary,debug=FALSE) {
-  if(debug) cat("generate_data\n")
-  Y <- matrix(NA,nrow=N,ncol=d)
-  for(j in seq_len(d)) {
-    distName <- config[[j]]$distName
-    paramFun <- config[[j]]$paramFun
-    if(debug) {
-      cat("\nj=",j,"distName=",distName,"\n")
-      cat("paramFun formals:",names(formals(paramFun)),"\n")
-    }
-    if(j==1) {
-      pList <- paramFun()
-      if(debug) {
-        cat("paramFun() ->\n")
-        str(pList)
-      }
-    } else {
-      prev <- as.data.frame(Y[,1:(j-1),drop=FALSE])
-      frmls <- names(formals(paramFun))
-      if(debug) {
-        cat("colnames before:",colnames(prev),"\n")
-        cat("frmls:",frmls,"\n")
-      }
-      n1 <- ncol(prev)
-      n2 <- length(frmls)
-      if(n2>0) colnames(prev)[1:min(n1,n2)] <- frmls[1:min(n1,n2)]
-      if(debug) {
-        cat("colnames after:",colnames(prev),"\n")
-        cat("calling paramFun\n")
-      }
-      pList <- do.call(paramFun,prev)
-      if(debug) {
-        cat("paramFun ->\n")
-        str(pList)
-      }
-    }
-    pList <- clamp_parameters(pList,debug=debug)
[…]

### 2025-05-17 Code/2025_03_07/03_models.R

@@ -1,99 +1,0 @@
-# 03_models.R
-
-fit_transformation_forest <- function(data) {
-  # Beispiel: Y2>0 => log(Y2)
-  data$Y2_log <- log(pmax(data$Y2, 1e-15))
-  
-  # Univariates Transformationsmodell
-  d1 <- BoxCox(Y1 ~ 1, data = data)
-  d2 <- BoxCox(Y2_log ~ 1, data = data)
-  d3 <- Colr(Y3 ~ 1, data = data, bounds = c(0, 1))
-  
-  # Forest-Fits
-  f2 <- traforest(d2, formula = Y2_log ~ Y1, data = data,
-                  ntree = 100, mtry = 1)
-  f3 <- traforest(d3, formula = Y3 ~ Y1 + Y2, data = data,
-                  ntree = 100, mtry = 2)
-  
-  mods <- list(m1 = d1, m2 = d2, m3 = d3, f2 = f2, f3 = f3)
-  
-  get_density_value <- function(obj) {
-    val <- if(is.list(obj)) obj[[1]] else obj
-    max(val, 1e-15)
-  }
-  
-  predict_joint <- function(newdata) {
-    n <- nrow(newdata)
-    out1 <- numeric(n)
-    out2 <- numeric(n)
-    out3 <- numeric(n)
-    
-    for(i in seq_len(n)) {
-      d1o <- predict(mods$m1, newdata[i, , drop=FALSE],
-                     type="density", q=newdata$Y1[i],
-                     simplify=FALSE)
-      out1[i] <- get_density_value(d1o)
-      
-      if(newdata$Y2[i] <= 0) {
-        out2[i] <- 1e-15
-      } else {
-        z2  <- log(newdata$Y2[i])
-        tmp <- newdata[i, , drop=FALSE]
-        tmp$Y2_log <- z2
-        d2o <- predict(mods$f2, newdata=tmp,
-                       type="density", q=z2,
-                       simplify=FALSE)
-        out2[i] <- get_density_value(d2o) * (1 / newdata$Y2[i])
-      }
-      
-      d3o <- predict(mods$f3, newdata[i, , drop=FALSE],
-                     type="density", q=newdata$Y3[i],
-                     simplify=FALSE)
-      out3[i] <- get_density_value(d3o)
-    }
-    
[…]

### 2025-05-17 Code/2025_03_07/05_viz.R

@@ -1,95 +1,0 @@
-# 05_viz.R
-
-# Diese erweiterte Funktion zeigt einerseits einen 1D-Schnitt p(y3 | y1=cval1, y2=cval2)
-# und andererseits einen 2D-Konturplot p(y3, y2 | y1=cval1).
-# So erhält man sowohl die direkte Form von p(y3|...) in 1D als auch den Überblick
-# über die gemeinsame Abhängigkeit von (y3, y2) bei fixem y1 in 2D.
-
-visualize_1D_and_2D <- function(dtrain,
-                                tf_predict,
-                                kd_predict,
-                                pm_predict,
-                                fix_y1 = NULL,
-                                fix_y2 = NULL) {
-  # Falls fix_y1/fix_y2 nicht gegeben, nehmen wir Mediane:
-  if(is.null(fix_y1)) fix_y1 <- median(dtrain$Y1)
-  if(is.null(fix_y2)) fix_y2 <- median(dtrain$Y2)
-  
-  # Panel 1 (links): 1D-Schnitt p(y3 | Y1=fix_y1, Y2=fix_y2)
-  seqY3 <- seq(min(dtrain$Y3), max(dtrain$Y3), length.out=200)
-  df_1d <- data.frame(Y1=fix_y1, Y2=fix_y2, Y3=seqY3)
-  
-  # Modell-Dichten:
-  tf_val_1d <- tf_predict(df_1d)
-  kd_val_1d <- kd_predict(df_1d)
-  pm_val_1d <- pm_predict(df_1d)
-  
-  # Rand p(Y1=fix_y1, Y2=fix_y2) => Errechnung bedingter Dichte p(y3|y1,y2)
-  # d.h. p(y1=fix_y1, y2=fix_y2)
-  df_marg_1d <- data.frame(Y1=fix_y1, Y2=fix_y2, Y3=0)
-  tf_marg_1d <- tf_predict(df_marg_1d)[1]
-  kd_marg_1d <- kd_predict(df_marg_1d)[1]
-  pm_marg_1d <- pm_predict(df_marg_1d)[1]
-  
-  tf_cond_1d <- tf_val_1d / pmax(tf_marg_1d, 1e-15)
-  kd_cond_1d <- kd_val_1d / pmax(kd_marg_1d, 1e-15)
[…]

### 2025-05-17 Code/2025_03_07/main.R

@@ -1,38 +1,0 @@
-# main.R
-
-setwd("/Users/leonkiafaro/Documents/multivariate-density-estimation/Code/2025_03_07")
-
-source("00_setup.R")
-source("01_distribution.R")
-source("02_dgp.R")
-
-d_used <- 25
-N      <- 100
-debug_mode <- FALSE
-
-all_data <- generate_data(
-  N           = N,
-  d           = d_used,
-  config      = config[1:d_used],
-  distLibrary = distLibrary,
-  debug       = debug_mode
-)
-
-cat("\nHead(all_data):\n")
-print(head(all_data))
-
-cat("\nCheck for NA values:\n")
-na_ix <- which(is.na(all_data), arr.ind=TRUE)
-print(na_ix)
-
-logdens_vec <- compute_logdensity(
-  Y           = all_data,
-  d           = d_used,
-  config      = config[1:d_used],
-  distLibrary = distLibrary,
-  debug       = debug_mode
-)
-
-cat("\nMean log-density =", mean(logdens_vec), "\n")
-
-

### 2025-05-17 Code/2025_03_11/dgp.R

@@ -1,269 +1,0 @@
-
-## -- Zusätzliche Pakete
-library("trtf")
-library("VineCopula")
-
-## -- (A) DGP + Transformation-Forest-Code (unverändert)
-set.seed(29)
-
-### config für 3 Dimensionen
-config <- list(
-  ### Y1
-  list(distr = "norm", parm = function(data) {list(mean = 0, sd = 1)}),
-  ### Y2
-  list(distr = "exp", parm = function(data) {list(rate = 1 + 0.1 * data[["Y1"]])}),
-  ### Y3
-  list(distr = "gamma", parm = function(data) {list(shape = 1 + data[["Y2"]], rate = 1)}),
-  ### Y4
-  list(distr = "norm", parm = function(data) {list(mean = data[["Y3"]], sd = 1)}),
-  ### Y5
-  list(distr = "exp", parm = function(data) {list(rate = 1 + 0.1 * data[["Y4"]])}),
-  ### Y6
-  list(distr = "gamma", parm = function(data) {list(shape = 1 + data[["Y5"]], rate = 1)})
-  # ### Y7
-  # list(distr = "norm", parm = function(data) {list(mean = data[["Y6"]], sd = 1)}),
-  # ### Y8
-  # list(distr = "exp", parm = function(data) {list(rate = 1 + 0.1 * data[["Y7"]])}),
-  # ### Y9
-  # list(distr = "gamma", parm = function(data) {list(shape = 1 + data[["Y8"]], rate = 1)}),
-  # ### Y10
-  # list(distr = "norm", parm = function(data) {list(mean = data[["Y9"]], sd = 1)}),
-  # ### Y11
-  # list(distr = "exp", parm = function(data) {list(rate = 1 + 0.1 * data[["Y10"]])}),
-  # ### Y12
-  # list(distr = "gamma", parm = function(data) {list(shape = 1 + data[["Y11"]], rate = 1)}),
-  # ### Y13
-  # list(distr = "norm", parm = function(data) {list(mean = data[["Y12"]], sd = 1)}),
-  # ### Y14
-  # list(distr = "exp", parm = function(data) {list(rate = 1 + 0.1 * data[["Y13"]])}),
-  # ### Y15
-  # list(distr = "gamma", parm = function(data) {list(shape = 1 + data[["Y14"]], rate = 1)}),
-  # ### Y16
[…]

### 2025-05-17 Archive/13_05_2025/main2.R
17_05

@@ -1,0 +1,548 @@
+suppressPackageStartupMessages({
+  library(extraDistr)
+  library(sn)
+  library(moments)
+  library(trtf)
+})
+
+
+
+config25 <- list(
+  list(distr = "norm",
+       parm  = NULL),
+
+  list(distr = "t",
+       parm  = \(d) list(df = 3 + 0.5*d$X1)),
+
+  list(distr = "laplace",
+       parm  = \(d) list(m = 0.3*d$X2,
+                         s = clip(1 + 0.1*d$X2))),
+
+  list(distr = "logis",
+       parm  = \(d) list(location = 0.2*d$X3,
+                         scale    = clip(1 + 0.05*d$X3))),
+
+  list(distr = "cauchy",
+       parm  = \(d) list(location = 0.1*d$X4,
+                         scale    = 0.5 + 0.05*d$X4)),
+
+  list(distr = "exp",
+       parm  = \(d) list(rate = clip(1 + 0.1*d$X5))),
+
+  list(distr = "gamma",
+       parm  = \(d) list(shape = clip(2 + 0.2*d$X6),
+                         rate  = 1)),
+
+  list(distr = "weibull",
+       parm  = \(d) list(shape = clip(2 + 0.2*d$X7),
+                         scale = clip(1 + 0.1*d$X7))),
+
+  list(distr = "lnorm",
+       parm  = \(d) list(meanlog = 0.3*d$X8,
+                         sdlog   = 0.5 + 0.05*d$X8)),
+
+  list(distr = "chisq",
+       parm  = \(d) list(df = clip(4 + 0.2*d$X9))),
+
+  list(distr = "f",
+       parm  = \(d) list(df1 = 5,
+                         df2 = 6 + 0.3*d$X10)),
+
+  list(distr = "beta",
+       parm  = \(d) { s <- 2 + 0.1*d$X11
+                     list(shape1 = s, shape2 = s) }),
+
+  list(distr = "beta",
+       parm  = \(d) list(shape1 = 0.5 + 0.05*d$X12,
+                         shape2 = 5   + 0.2*d$X12)),
+
+  list(distr = "gumbel",
+       parm  = \(d) list(mu    = 0.5*d$X13,
+                         sigma = clip(1 + 0.05*d$X13))),
+
+  list(distr = "gpd",
[…]

### 2025-05-17 Archive/13_05_2025/main5.R

@@ -1,0 +1,643 @@
+suppressPackageStartupMessages({
+  library(extraDistr)
+  library(sn)
+  library(moments)
+  library(trtf)
+  library(tram)
+  library(mlt)
+  library(formula.tools)
+})
+
+
+
+config25 <- list(
+  list(distr = "norm",
+       parm  = NULL),
+
+  list(distr = "t",
+       parm  = \(d) list(df = 3 + 0.5*d$X1)),
+
+  list(distr = "laplace",
+       parm  = \(d) list(m = 0.3*d$X2,
+                         s = clip(1 + 0.1*d$X2))),
+
+  list(distr = "logis",
+       parm  = \(d) list(location = 0.2*d$X3,
+                         scale    = clip(1 + 0.05*d$X3))),
+
+  list(distr = "cauchy",
+       parm  = \(d) list(location = 0.1*d$X4,
+                         scale    = 0.5 + 0.05*d$X4)),
+
+  list(distr = "exp",
+       parm  = \(d) list(rate = clip(1 + 0.1*d$X5))),
+
+  list(distr = "gamma",
+       parm  = \(d) list(shape = clip(2 + 0.2*d$X6),
+                         rate  = 1)),
+
+  list(distr = "weibull",
+       parm  = \(d) list(shape = clip(2 + 0.2*d$X7),
+                         scale = clip(1 + 0.1*d$X7))),
+
+  list(distr = "lnorm",
+       parm  = \(d) list(meanlog = 0.3*d$X8,
+                         sdlog   = 0.5 + 0.05*d$X8)),
+
+  list(distr = "chisq",
+       parm  = \(d) list(df = clip(4 + 0.2*d$X9))),
+
+  list(distr = "f",
+       parm  = \(d) list(df1 = 5,
+                         df2 = 6 + 0.3*d$X10)),
+
+  list(distr = "beta",
+       parm  = \(d) { s <- 2 + 0.1*d$X11
+                     list(shape1 = s, shape2 = s) }),
+
+  list(distr = "beta",
+       parm  = \(d) list(shape1 = 0.5 + 0.05*d$X12,
+                         shape2 = 5   + 0.2*d$X12)),
+
+  list(distr = "gumbel",
+       parm  = \(d) list(mu    = 0.5*d$X13,
[…]

### 2025-05-17 Code/2025_01_31/BoxCox.R

@@ -1,243 +1,0 @@
-###########################################################
-# 0) Pakete
-###########################################################
-library(copula)   # NormalCopula
-library(benchden) # qberdev, rberdev etc.
-library(tram)     # z.B. BoxCox
-library(np)       # optional: Kernel-Dichte (multivariat)
-library(ks)       # optional: Kernel-Smoothing (multivariat)
-
-
-###########################################################
-# 1) Datengenerierung via Normal-Copula + benchden
-#    => Kann 2D oder 3D erzeugen
-###########################################################
-generate_data_copula_benchden <- function(n = 1000,
-                                          rho = 0.5,
-                                          dist_indices = c(11, 12)) {
-  # 'dist_indices' bestimmt zugleich die Dimension d
-  d <- length(dist_indices)
-  stopifnot(d %in% c(2, 3))
-  
-  # Normal-Copula definieren
-  if (d == 2) {
-    myCop <- normalCopula(param = rho, dim = 2)
-  } else {
-    # d=3 => 3x3-Korrelationsmatrix
-    corMat <- matrix(rho, nrow = 3, ncol = 3)
-    diag(corMat) <- 1
-    myCop <- normalCopula(param = corMat[upper.tri(corMat)], 
-                          dim = 3, dispstr = "un")
-  }
-  
-  # Zufallszahlen U in [0,1]^d
-  set.seed(123)
-  U <- rCopula(n, myCop)
-  
-  # Marginalverteilungen über benchden::qberdev
-  data_mat <- matrix(NA, nrow = n, ncol = d)
-  for (j in seq_len(d)) {
-    data_mat[, j] <- qberdev(U[, j], dnum = dist_indices[j])
-  }
-  
-  df <- as.data.frame(data_mat)
-  colnames(df) <- paste0("y", seq_len(d))
-  return(df)
-}
-
-
-###########################################################
-# 2) BoxCox-Modelle aufbauen: 
-#    2D => p(y1), p(y2 | y1)
-#    3D => p(y1), p(y2 | y1), p(y3 | y1, y2)
-###########################################################
-fit_factorized_tram <- function(df) {
-  d <- ncol(df)
-  stopifnot(d %in% c(2, 3))
-  colnames(df) <- paste0("y", seq_len(d))
-  
-  # p(y1)
-  m1 <- BoxCox(y1 ~ 1, data = df)
-  
[…]

### 2025-05-17 Code/2025_02_21/00_setup.R

@@ -1,14 +1,0 @@
-# 00_setup.R
-# Paket-Setup und globale Konfiguration
-
-library(tram)
-library(trtf)
-library(ks)
-library(partykit)
-library(party)
-library(foreach)
-library(doParallel)
-
-set.seed(123)
-options(digits=4)
-

### 2025-05-17 Code/2025_02_21/01_distribution.R

@@ -1,55 +1,0 @@
-# 01_distribution.R
-
-# (Neu angelegtes Skript für die Verteilungs-Definitionen)
-
-# Modelle, jeweils $rDist und $dDist pro Dimension
-#   1) Y1 ~ Chi^2(3)
-#   2) Y2 ~ Gamma(shape, scale)
-#   3) Y3 ~ Beta(alpha, beta)
-
-models <- list(
-  # Y1: Chi^2(3), marginal
-  list(
-    rDist = function(n) rchisq(n, df=3),
-    dDist = function(x) dchisq(x, df=3)
-  ),
-  # Y2: Gamma
-  list(
-    rDist = function(n, shape, scale) rgamma(n, shape=shape, scale=scale),
-    dDist = function(x, shape, scale) dgamma(x, shape=shape, scale=scale)
-  ),
-  # Y3: Beta
-  list(
-    rDist = function(n, alpha, beta) rbeta(n, alpha, beta),
-    dDist = function(x, alpha, beta) dbeta(x, alpha, beta)
-  )
-)
-
-# Bedingte Parameter je Dimension:
-#   1) (NULL): Y1 ~ Chi^2(3) (marginal)
-#   2) (Y1) => Y2 ~ Gamma(shape, scale)
-#   3) (Y1, Y2) => Y3 ~ Beta(alpha, beta)
-# Formeln so gewählt, dass Y2>0 und Y3 in [0,1].
-cond <- list(
-  NULL,
-  function(Y1) {
-    # shape = 2 + 0.1*Y1
-    # scale = 0.5 + 0.01*Y1
-    shp <- 2 + 0.1 * Y1
-    scl <- 0.5 + 0.01 * Y1
-    if(shp <= 0) shp <- 0.1
-    if(scl <= 0) scl <- 0.1
-    list(shape = shp, scale = scl)
-  },
-  function(Y1, Y2) {
-    # alpha = 1 + 0.05*Y1 + 0.02*Y2
-    # beta  = 1 + 0.02*Y1 + 0.02*Y2
-    a_ <- 1 + 0.05*Y1 + 0.02*Y2
-    b_ <- 1 + 0.02*Y1 + 0.02*Y2
-    if(a_ <= 0) a_ <- 0.1
-    if(b_ <= 0) b_ <- 0.1
-    list(alpha = a_, beta = b_)
-  }
-)
-
-

### 2025-05-17 Code/2025_02_21/02_dgp.R

@@ -1,153 +1,0 @@
-# 02_dgp.R
-
-# Modelle, jeweils $rDist und $dDist pro Dimension
-#  Hier:
-#   * Y1 ~ Chi^2(3)
-#   * Y2 ~ Gamma( shape, scale )
-#   * Y3 ~ Beta( alpha, beta )
-#
-#  -> Y2 > 0 ist garantiert (Gamma) => Log(Y2) bleibt definiert
-#  -> Y3 in [0,1] => Beta-Distribution
-
-models <- list(
-  # Y1: Chi^2(3), marginal
-  list(
-    rDist = function(n) rchisq(n, df=3),
-    dDist = function(x) dchisq(x, df=3)
-  ),
-  # Y2: Gamma
-  list(
-    # shape>0, scale>0
-    rDist = function(n, shape, scale) rgamma(n, shape=shape, scale=scale),
-    dDist = function(x, shape, scale) dgamma(x, shape=shape, scale=scale)
-  ),
-  # Y3: Beta
-  list(
-    rDist = function(n, alpha, beta) rbeta(n, alpha, beta),
-    dDist = function(x, alpha, beta) dbeta(x, alpha, beta)
-  )
-)
-
-# Bedingte Parameter je Dimension:
-#   1) (NULL): Y1 ~ Chi^2(3) (marginal)
-#   2) (Y1) => Y2 ~ Gamma(shape, scale)
-#   3) (Y1, Y2) => Y3 ~ Beta(alpha, beta)
-# Diese Formeln wählen wir so, dass Y2 immer > 0 ist und Y3 in [0,1].
-cond <- list(
-  NULL,
-  # Y2 param: shape = 2 + 0.1*Y1, scale = 0.5 + 0.01*Y1
-  function(Y1) {
-    shp <- 2 + 0.1 * Y1
-    scl <- 0.5 + 0.01 * Y1
-    # Minimale Sicherheitschecks
-    if(shp <= 0) shp <- 0.1
-    if(scl <= 0) scl <- 0.1
-    list(shape = shp, scale = scl)
-  },
-  # Y3 param: alpha = 1 + 0.05*Y1 + 0.02*Y2, beta = 1 + 0.02*Y1 + 0.02*Y2
-  function(Y1, Y2) {
-    a_ <- 1 + 0.05*Y1 + 0.02*Y2
-    b_ <- 1 + 0.02*Y1 + 0.02*Y2
-    if(a_ <= 0) a_ <- 0.1
[…]

### 2025-05-17 Code/2025_02_21/03_models.R

@@ -1,109 +1,0 @@
-# 03_models.R
-
-fit_transformation_forest <- function(data) {
-  data$y2_log <- log(data$Y2)
-  
-  d1 <- BoxCox(Y1 ~ 1, data=data)
-  d2 <- BoxCox(y2_log ~ 1, data=data)
-  d3 <- Colr(Y3 ~ 1, data=data, bounds=c(0,1))
-  
-  f2 <- traforest(d2, formula = y2_log ~ Y1, data = data,
-                  ntree = 100, mtry = 1)
-  f3 <- traforest(d3, formula = Y3 ~ Y1 + Y2, data = data,
-                  ntree = 100, mtry = 2)
-  
-  mods <- list(m1=d1, m2=d2, m3=d3, f2=f2, f3=f3)
-  
-  get_density_value <- function(obj, q) {
-    if(is.list(obj)) {
-      val <- obj[[1]]
-      if(is.function(val)) return(max(val(q), 1e-15))
-      else return(max(val, 1e-15))
-    } else {
-      return(max(obj, 1e-15))
-    }
-  }
-  
-  predict_joint <- function(newdata) {
-    n <- nrow(newdata)
-    out1 <- numeric(n)
-    out2 <- numeric(n)
-    out3 <- numeric(n)
-    for(i in seq_len(n)) {
-      d1o <- predict(mods$m1, newdata[i,,drop=FALSE],
-                     type="density", q=newdata$Y1[i], simplify=FALSE)
-      out1[i] <- get_density_value(d1o, newdata$Y1[i])
-      
-      if(newdata$Y2[i] <= 0) {
-        out2[i] <- 1e-15
-      } else {
-        z2 <- log(newdata$Y2[i])
-        tmp_line <- newdata[i,,drop=FALSE]
-        tmp_line$y2_log <- z2
-        d2o <- predict(mods$f2, newdata=tmp_line,
-                       type="density", q=z2, simplify=FALSE)
-        out2[i] <- get_density_value(d2o, z2) * (1 / newdata$Y2[i])
-      }
-      
-      d3o <- predict(mods$f3, newdata[i,,drop=FALSE],
-                     type="density", q=newdata$Y3[i], simplify=FALSE)
-      out3[i] <- get_density_value(d3o, newdata$Y3[i])
-    }
-    out1 * out2 * out3
-  }
[…]

### 2025-05-17 Code/poc_K4_forest.R

@@ -1,0 +1,60 @@
+## Block D & E: Transformation Forests + Joint Likelihood-Evaluation
+
+source("poc_K4.R")  # lädt Daten & Block A–C citeturn1file0
+library(trtf)
+library(tram)     # für BoxCox
+
+mytrtf <- function(data) {
+  ymod <- lapply(names(data), function(y)
+    BoxCox(as.formula(paste0(y, "~1")), data = data)
+  )
+  fm <- lapply(2:ncol(data), function(j)
+    as.formula(paste0(
+      names(data)[j], " ~ ",
+      paste(names(data)[1:(j-1)], collapse = "+")
+    ))
+  )
+  forests <- lapply(seq_along(fm), function(j)
+    traforest(
+      ymod[[j+1]], formula = fm[[j]], data = data,
+      ntree = 200,
+      mtry = ceiling((j-1)/3),
+      minbucket = 20,
+      trace = TRUE
+    )
+  )
+  structure(list(ymod = ymod, forests = forests), class = "mytrtf")
+}
+
+predict.mytrtf <- function(object, newdata, type = "logdensity") {
+  ld1 <- predict(object$ymod[[1]], newdata = newdata, type = "logdensity")
+  ldf <- sapply(object$forests, function(frst)
+    predict(frst, newdata = newdata, type = "logdensity")
+  )
+  cbind(ld1, ldf)
+}
+
+model     <- mytrtf(as.data.frame(X_pi_train))
+LD_hat    <- predict(model, newdata = as.data.frame(X_pi_test))
+ell_forest <- -0.5 * rowSums(Z_eta_test^2) -
+  (ncol(Z_eta_test)/2) * log(2*pi) + rowSums(LD_hat)
+all.equal(sum(ell_forest), sum(ll_test), tol = 1e-1)
+
+pdf("results/BlockE_scatterplots.pdf")
+par(mfrow = c(2,2))
+for (k in seq_len(ncol(LD_hat))) {
+  plot(
+    true_ll_mat_test[,k], LD_hat[,k],
+    main = paste0("dim ", k), xlab = "true", ylab = "forest"
+  )
+  abline(0,1)
+}
+dev.off()
+
+delta_df <- data.frame(
+  dim        = seq_len(ncol(LD_hat)),
+  ell_true   = colSums(true_ll_mat_test),
+  ell_forest = colSums(LD_hat)
+)
+delta_df$delta <- delta_df$ell_true - delta_df$ell_forest
+print(delta_df)

### 2025-05-17 Code/scripts_run.Rmd

---
title: "Script Run"
output: html_document
---

# Setup

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
source("00_setup.R")
```

# Transport Utilities

```{r transport-utils, echo=FALSE, message=FALSE, warning=FALSE}
source("01_transport_utils.R")
```

# Generate Data

```{r generate-data, echo=FALSE, message=FALSE, warning=FALSE}
source("02_generate_data.R")
```

# Parametric Baseline

```{r param-baseline, echo=FALSE, message=FALSE, warning=FALSE}
source("03_param_baseline.R")
```

### 2025-06-24 analysis_pipeline.Rmd

---
title: "Analysis Pipeline"
output: html_document
---

```{r setup, include=FALSE}
source("00_globals.R")
source("01_data_generation.R")
source("02_split.R")
source("models/true_model.R")
source("models/trtf_model.R")
source("models/ks_model.R")
source("04_evaluation.R")
source("EDA.R")
```

## Data Preparation

```{r}
set.seed(1)
perm <- c(3, 4, 1, 2)
config <- list(
  list(distr = "norm", parm = NULL),
  list(distr = "exp",  parm = function(d) list(rate = softplus(d$X1))),
  list(distr = "beta", parm = function(d) list(shape1 = softplus(d$X2), shape2 = softplus(d$X1))),
  list(distr = "gamma", parm = function(d) list(shape = softplus(d$X3), scale = softplus(d$X2)))
)
prep <- prepare_data(50, config, perm)
```

## Fit Models

```{r}
mods <- fit_models(prep$S, config)
```

## Log-Likelihood Tables

```{r}
tab <- calc_loglik_tables(mods, config)
print(tab)
```

## Scatter Data and Plots

```{r}
scat <- make_scatter_data(mods, prep$S)
scat_plots <- plot_scatter(scat)
```

```{r}
for(p in scat_plots) replayPlot(p)
```

## Optional Full Pipeline

```{r}
# run_pipeline(50, config, perm)
```

### 2025-06-24 analysis_pipeline.Rmd
Korrigiere fehlende Scatter-Daten im Pipeline-Notebook

@@
 mods <- fit_models(prep$S, config)
+mods_p <- fit_models(prep$S_perm, config)
 ```
@@
 ```{r}
-scat <- make_scatter_data(mods, prep$S)
+scat_n <- make_scatter_data(mods, prep$S)
+scat_p <- make_scatter_data(mods_p, prep$S_perm)
+scat <- c(scat_n, setNames(scat_p, paste0(names(scat_p), "_p")))
 scat_plots <- plot_scatter(scat)

### 2025-08-13 docs/trtf_ttm_pseudocode.Rmd
Dok: TRTF- und TTM-Pseudocode hinzugefuegt

---
title: "TRTF and TTM Pseudocode"
author: "Research Team"
date: "2025-08-13"
output: html_document
---

## Notation & Setup

Let $\pi(x)$ denote the target log density and $\eta(z)$ the base log density. The triangular transport map $S: \mathbb{R}^K \to \mathbb{R}^K$ has components $S_k(x_1,\ldots,x_k)$ for $k=1,\ldots,K$. Here $N$ is the number of observations and $K$ the dimension of $x$. The change of variables states, in words, that the log density of $x$ under $\pi$ equals the log density of $S(x)$ under $\eta$ plus the log determinant of the Jacobian, which reduces to a sum of log partial derivatives because $S$ is triangular. The pipeline main() -> TRUE/TRTF/KS/TTM marginal governs all experiments. Model artifacts are TRTF and the Marginal Map (diagonal). All log densities are measured in nats, and sums over dimensions yield the joint log density.

## TRTF Pseudocode

1. Input $\{x_i\}_{i=1}^N$.
2. For $k = 1$ to $K$:
   - Fit a transformation forest for $x_k$ conditioned on $(x_1,\ldots,x_{k-1})$ to obtain $S_k$.
   - Accumulate $\log \partial_{x_k} S_k(x_1,\ldots,x_k)$.
3. Evaluate $\log \eta(S(x))$ at the mapped sample.
4. Return $\log \pi(x)$ as the sum of base log density and accumulated terms.

## Marginal Map Pseudocode

1. Input $\{x_i\}_{i=1}^N$.
2. For each dimension $k$:
   - Fit a univariate map $S_k$ on $x_k$.
   - Accumulate $\log \partial_{x_k} S_k(x_k)$.
3. Evaluate $\log \eta(S(x))$ using the diagonal map.
4. Return $\log \pi(x)$ as the sum of base log density and accumulated terms.

## Acceptance & Reproducibility

A model is accepted when its joint log density exceeds that of the baseline. All computations use a single global seed shared across models. This policy ensures reproducibility and comparability across TRTF and Marginal Map (diagonal) artifacts.

### 2025-08-13 docs/trtf_ttm_pseudocode.Rmd
Ergaenze TRTF-Training und -Vorhersage als Pseudocode

@@
 ## TRTF Pseudocode
+### fit\_TRTF($S$, seed)
 
-1. Input $\{x_i\}_{i=1}^N$.
-2. For $k = 1$ to $K$:
-   - Fit a transformation forest for $x_k$ conditioned on $(x_1,\ldots,x_{k-1})$ to obtain $S_k$.
-   - Accumulate $\log \partial_{x_k} S_k(x_1,\ldots,x_k)$.
-3. Evaluate $\log \eta(S(x))$ at the mapped sample.
-4. Return $\log \pi(x)$ as the sum of base log density and accumulated terms.
+1. Set the global `seed`.
+2. Let $X$ be the $N\times K$ sample matrix.
+3. Define `ymod` by fitting a marginal BoxCox model for $X_1$.
+4. For $k=2,\ldots,K$:
+   - Fit a transformation forest for $X_k$ with predictors $(X_1,\ldots,X_{k-1})$.
+   - Store the fitted forest, variable importance and configuration.
+5. Output an object containing `ymod`, the list of forests, `seed` and the configuration.
+
+### predict(TRTF, $X$, type)
+
+1. Compute $ld1$, the log density of the first dimension via `ymod`.
+2. For $j=1,\ldots,K-1$:
+   - Compute conditional log densities for column $j+1$ using forest $j$ with predictors $(X_1,\ldots,X_j)$.
+   - Take the diagonal of the resulting prediction matrix.
+3. Form $ll = \text{cbind}(ld1, \text{conditional contributions})$.
+4. Ensure $ll$ contains no NA or $\infty$ entries.
+5. If `type == 'logdensity_by_dim'` return $ll$ (an $N\times K$ matrix).
+6. Else return $\text{rowSums}(ll)$, labelled `'logdensity'`.
+7. Summing over $K$ dimensions yields the joint log density.
+8. Damit gilt: `logdensity_by_dim` ist $N\times K$ und $\text{rowSums}(ll) =`'logdensity'`.
 

### 2025-08-13 docs/trtf_ttm_pseudocode.Rmd
Ergaenze TTM-Marginalpseudocode um Training, Dichte und KL

@@
 
-1. Input $\{x_i\}_{i=1}^N$.
-2. For each dimension $k$:
-   - Fit a univariate map $S_k$ on $x_k$.
-   - Accumulate $\log \partial_{x_k} S_k(x_k)$.
-3. Evaluate $\log \eta(S(x))$ using the diagonal map.
-4. Return $\log \pi(x)$ as the sum of base log density and accumulated terms.
+### Training
+
+1. Take the training sample matrix $X_{\text{tr}}$ of size $N_{\text{tr}}\times K$.
+2. Compute standardization parameters $\mu_k = \operatorname{mean}(x_{k})$ and $\sigma_k = \operatorname{sd}(x_{k})$ for each dimension; **these are train-only and stored for later use**.
+3. For $k=1,\ldots,K$:
+   - $u \leftarrow \operatorname{rank}(x_k, \text{ties.avg})/(N_{\text{tr}}+1)$ and clip $u$ to $[1/(N_{\text{tr}}+1), N_{\text{tr}}/(N_{\text{tr}}+1)]$.
+   - $z^\star \leftarrow \Phi^{-1}(u)$ where $\Phi^{-1}$ is the standard normal quantile.
+   - $b_k \leftarrow \max(0, \operatorname{cov}(x_k, z^\star)/\operatorname{var}(x_k))$.
+   - $a_k \leftarrow \operatorname{mean}(z^\star) - b_k\,\operatorname{mean}(x_k)$.
+   - Store coefficients $\text{coeffA}_k = \log(b_k+\varepsilon)$, $\text{coeffB}_k = a_k$, $\text{coeffC}_k = 0$ along with $\mu_k,\sigma_k$.
+
+### Prediction
+
+1. Standardize new data: $X_{\text{std}} = (X-\mu)/\sigma$ using the **training** $\mu,\sigma$.
+2. Compute $z = a + b\, X_{\text{std}}$ elementwise.
+3. For each $i,k$, 
+   $\text{logdensity\_by\_dim}[i,k] = -0.5\,z_{i,k}^2 -0.5\log(2\pi) + \log b_k - \log \sigma_k$.
+4. The term $\log b_k - \log \sigma_k$ is the Jacobian constant; it explicitly contains the contribution $-\log(\sigma_k)$ from the train-only standardization.
+5. Ensure no NA or $\infty$ appear in $\text{logdensity\_by\_dim}$.
+6. If `type == 'logdensity_by_dim'` return this $N\times K$ matrix; else return $\operatorname{rowSums}(\text{logdensity\_by\_dim})$ labelled `'logdensity'`.
+7. Somit gilt: `logdensity_by_dim` ist $N\times K$ und `rowSums` = `'logdensity'`.
+
+### Forward-KL Objective
+
+$J(S) = \operatorname{mean}_i\left(0.5\sum_k z_{i,k}^2\right) - \sum_k(\log b_k - \log \sigma_k)$.
 

### 2025-08-13 docs/trtf_ttm_pseudocode.Rmd

@@
 
-A model is accepted when its joint log density exceeds that of the baseline.
-All computations use a single global seed shared across models.
-This policy ensures reproducibility and comparability across TRTF and Marginal Map (diagonal) artifacts.
+The following checklist must hold for accepting results and ensuring reproducibility:
+
+1. **Shape (TRTF):** `predict(mytrtf, X, 'logdensity_by_dim')` returns an $N\times K$ matrix; $\operatorname{rowSums}$ yields `'logdensity'`.
+2. **Shape (Marginal Map):** `predict(trainMarginalMap, X, 'logdensity_by_dim')` returns an $N\times K$ matrix; $\operatorname{rowSums}$ yields `'logdensity'`.
+3. **Constants:** Each dimension contributes $-0.5\log(2\pi)$; the Jacobian of `trainMarginalMap` contains $-\log(\sigma_k)$.
+4. **Standardisierung:** $\mu,\sigma$ are estimated on training data only and applied unchanged to validation and test sets.
+5. **Seed-Policy:** A single global seed is used; `mytrtf` and `trainMarginalMap` call `set.seed` with this value.
+6. **Stabilität:** Ranks use `ties.average`; $u$ is clipped to $[1/(N_{\text{tr}}+1), N_{\text{tr}}/(N_{\text{tr}}+1)]$.
+7. **Numerik:** `predict(...,'logdensity_by_dim')` outputs contain neither NA nor $\infty$.
+8. **Summenregel:** Summing over $K$ dimensions of `logdensity_by_dim` reproduces the joint log density.
+9. **Repro:** Two runs with the same seed yield identical negative log-likelihood tables (nats).
+10. **Dokument:** `docs/trtf_ttm_pseudocode.Rmd` consists solely of text and LaTeX without R code chunks.

### 2025-08-20 04_evaluation.R
Kernel-Smoother vollständig entfernt

@@ -5,7 +5,6 @@ if (basename(root_path) == "testthat") {
 }
 source(file.path(root_path, "models/ttm_marginal.R"))
 source(file.path(root_path, "models/ttm_separable.R"))
-source(file.path(root_path, "models/ks_model.R"))
 source(file.path(root_path, "models/ttm_cross_term.R"))
 
 #' Summen-Zeile an Tabelle anh\u00e4ngen
@@ -45,7 +44,7 @@ prepare_data <- function(n, config, seed = 42) {
   list(X = X, S = S)
 }
 
-#' TRUE, TRTF und KS Modelle fitten
+#' TRUE, TRTF und Cross-Term-Map fitten
 #'
 #' @param S Liste mit `X_tr`, `X_val`, `X_te`
 #' @param config Konfiguration der Zielverteilungen
@@ -57,27 +56,22 @@ fit_models <- function(S, config) {
     ll_true <- logL_TRUE_dim(M_TRUE, S$X_te)
   })[["elapsed"]]
 
-  M_TRTF <- fit_TRTF(S, config)
-  t_trtf <- system.time({
-    ll_trtf <- logL_TRTF_dim(M_TRTF, S$X_te)
-  })[["elapsed"]]
+    M_TRTF <- fit_TRTF(S, config)
+    t_trtf <- system.time({
+      ll_trtf <- logL_TRTF_dim(M_TRTF, S$X_te)
+    })[["elapsed"]]
 
-  M_KS <- fit_KS(S, config)
-  t_ks <- system.time({
-    ll_ks <- logL_KS_dim(M_KS, S$X_te)
-  })[["elapsed"]]
+    M_TTM_cross <- trainCrossTermMap(S)
+    t_ttm_cross <- system.time({
+      ll_ttm_cross <- -predict(M_TTM_cross$S, S$X_te, type = "logdensity_by_dim")
+    })[["elapsed"]]
 
-  M_TTM_cross <- trainCrossTermMap(S)
-  t_ttm_cross <- system.time({
-    ll_ttm_cross <- -predict(M_TTM_cross$S, S$X_te, type = "logdensity_by_dim")
-  })[["elapsed"]]
-
-  list(models = list(true = M_TRUE, trtf = M_TRTF, ks = M_KS,
-                     ttm_cross = M_TTM_cross),
-       ll = list(true = ll_true, trtf = ll_trtf, ks = ll_ks,
-                ttm_cross = ll_ttm_cross),
[…]

### 2025-08-20 Untitled.Rmd
.

---
title: "Report"
author: "Léon Kia Faro"
date: "2025-08-20"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Methodology and Algorithms

## 0. Data-generating process (DGP)

We generate triangular (autoregressive) data via the factorization
\[
p(x) \;=\; p_1(x_1)\,\prod_{k=2}^{K} p_k\big(x_k \mid x_{1:k-1}\big).
\]

In the experiments we use the following configuration (softplus(u) = log(1 + exp(u))):
\[
\begin{aligned}
&X_1 \sim \mathcal{N}(0,1),\\
&X_2 \mid X_1 \sim \mathrm{Exp}\big(\text{rate} = \mathrm{softplus}(X_1)\big),\\
&X_3 \mid X_{1:2} \sim \mathrm{Beta}\big(\alpha = \mathrm{softplus}(X_2),\ \beta = \mathrm{softplus}(X_1)\big),\\
&X_4 \mid X_{1:3} \sim \mathrm{Gamma}\big(\text{shape} = \mathrm{softplus}(X_3),\ \text{scale} = \mathrm{softplus}(X_2)\big).
\end{aligned}
\]

R configuration used to generate samples:

```r
config <- list(
  list(distr = "norm", parm = NULL),
  list(distr = "exp",  parm = function(d) list(rate = softplus(d$X1))),
  list(distr = "beta",
       parm = function(d) list(shape1 = softplus(d$X2),
                               shape2 = softplus(d$X1))),
  list(distr = "gamma",
       parm = function(d) list(shape = softplus(d$X3),
                               scale = softplus(d$X2)))
)
```

All positive parameters are clamped to at least 1e-3 in the code to ensure well-defined likelihoods. The factorization is fully triangular: for each k, \(p_k(\cdot \mid x_{1:k-1})\) may depend on all predecessors.

## 1. Notation and standardization

- Data: \(X \in \mathbb{R}^{N \times K}\), rows \(X_i \in \mathbb{R}^K\).
- Train-only standardization (stored and reused on validation/test):
  \[
  x^{\text{std}} = \frac{x - \mu}{\sigma} \quad \text{(component-wise)}, \qquad
  \log J_{\text{std}}(x) = \sum_{k=1}^K (-\log \sigma_k).
  \]
- Triangular map: \(S(x) = (S_1(x_1), S_2(x_{1:2}), \ldots, S_K(x_{1:K}))\) with \(\partial_{x_k} S_k(x) > 0\).
- Reference density: \(\eta = \mathcal{N}(0, I)\).

Common per-dimension log-density (all TTM variants):
\[
\ell_k(x) \;=\; -\tfrac12\,z_k(x)^2 \;-\; \tfrac12 \log(2\pi) \;+\; \log \partial_{x_k} S_k(x), 
\quad z_k(x) := S_k(x_{1:k}).
\]
The train-only standardization contributes \(-\log \sigma_k\) to \(\log \partial_{x_k} S_k\). The joint log-density is \(\sum_{k=1}^K \ell_k(x)\).

## 2. Objective for learned TTM variants

We minimize the forward-KL functional using its Monte Carlo form:
\[
J(S) \;=\; \frac{1}{N} \sum_{i=1}^{N} \sum_{k=1}^{K}
\Big( \tfrac12\,S_k(X_i)^2 - \log \partial_{x_k} S_k(X_i) \Big)
\;=\; \sum_{k=1}^K J_k(S_k).
\]
This is the training criterion in `trainSeparableMap()` and `trainCrossTermMap()` (with simple \(\ell_2\) ridge). The marginal TTM uses a fast rank-to-Normal calibration instead of explicit KL optimization.

## 3. Reference models (TRUE) - detailed

### 3.1 TRUE (marginal)

- Model:
  \[
  p(x) \;=\; \prod_{k=1}^K p_k(x_k \mid \theta_k),
  \quad p_k \in \{\text{Normal}, \text{Exponential}, \text{Beta}, \text{Gamma}\}.
  \]
- Training (separate MLE per dimension):
  \[
  \hat{\theta}_k \;=\; \arg\max_{\theta} \sum_{i=1}^{N} \log p_k(X_{i,k} \mid \theta),
  \]
  using method-of-moments initializations and positivity bounds.
- Prediction: \(\ell_k(x) = \log p_k(x_k \mid \hat{\theta}_k)\).
- Scope: ignores all cross-dimensional dependence.

### 3.2 TRUE (joint/oracle)

- Factorization with full conditioning:
  \[
  p(x) \;=\; \prod_{k=1}^K \pi_k(x_k \mid x_{1:k-1}), \qquad
  \vartheta_k(x_{<k}) \;=\; \texttt{config[[k]]\$parm}(x_{1:k-1}).
  \]
  For \(k=3\) this conditions on \(x_1\) and \(x_2\) (not just one step back).
- Training: none (parameters are deterministic functions of \(x_{<k}\)).
- Prediction: \(\ell_k(x) = \log \pi_k(x_k \mid x_{1:k-1})\).
- Scope: oracle reference that exploits the full triangular structure of the DGP.

## 4. Transformation Random Forests (TRTF) - detailed

Goal: learn a conditional monotone transform \(T_k(\cdot \mid x_{<k})\) such that, approximately,
\[
Z_k \;:=\; T_k(X_k \mid X_{<k}) \sim \mathcal{N}(0,1).
\]

Per-dimension log-density (change-of-variables):
\[
\ell_k(x) \;=\; \log \phi\big(T_k(x_k \mid x_{<k})\big) \;+\; \log \partial_{x_k} T_k(x_k \mid x_{<k}),
\]
with \(\phi\) the standard normal density.

Training sketch (as implemented):
- \(k=1\): univariate transformation model (e.g., Box-Cox in `tram`).
- \(k \ge 2\): transformation forest (`traforest` in `trtf/tram`) with predictors \(x_{<k}\) and response \(x_k\). Forest weights yield a local transformation \(T_k\) and its derivative for the log-Jacobian term.

Comparison to TTM: TRTF constructs conditional transformations implicitly via forest weighting, while TTM parametrizes \(S_k\) explicitly with guaranteed monotonicity and optimizes the forward-KL objective.

## 5. Triangular Transport Maps (TTM)

### 5.1 Common evaluation pipeline

1. Standardize \(x \mapsto x^{\text{std}}\) using stored \((\mu, \sigma)\) from training data.
2. For \(k = 1{:}K\): compute \(z_k = S_k(x_{1:k})\) and \(\log \partial_{x_k} S_k(x) - \log \sigma_k\).
3. Compute \(\ell_k(x) = -\tfrac12 z_k^2 - \tfrac12 \log(2\pi) + \log \partial_{x_k} S_k(x)\).
4. Sum over \(k\) for the joint log-density.

### 5.2 TTM - Marginal (diagonal)

- Parametrization:
  \[
  S_k(x) \;=\; a_k + b_k\,x_k^{\text{std}}, \qquad b_k > 0 \ (\text{implemented as } b_k = \exp(\alpha_k)).
  \]
- Log-Jacobian: \(\log \partial_{x_k} S_k = \log b_k - \log \sigma_k\) (constant in \(x\)).
- Training (fast rank-to-Normal calibration):
  - \(u_i = \mathrm{rank}(x_{i,k}^{\text{std}})/(N+1)\), clamp to \([1/(N+1),\, N/(N+1)]\).
  - \(z_i^\star = \Phi^{-1}(u_i)\).
  - \(\hat{b}_k = \max\{0, \mathrm{Cov}(x^{\text{std}}, z^\star) / \mathrm{Var}(x^{\text{std}})\}\), 
    \(\hat{a}_k = \overline{z^\star} - \hat{b}_k\,\overline{x^{\text{std}}}\).
- Scope: captures univariate marginal shapes only.

### 5.3 TTM - Separable \((S_k = g_k + f_k)\)

- Parametrization:
  \[
  S_k(x_{1:k}) \;=\; g_k(x_{<k}) + f_k(x_k), \qquad f_k'(x_k) > 0.
  \]
  In code:
  - \(g_k(x_{<k}) = \Psi^{\text{non}}_k(x_{<k})\,c^{\text{non}}_k\) (polynomial basis),
  - \(f_k(x_k) = \Psi^{\text{mon}}_k(x_k)\,c^{\text{mon}}_k\) with \(\Psi^{\text{mon}} = [x, \mathrm{erf}(x)]\),
    \(f'_k(x) = B(x)\,c^{\text{mon}}\), \(B(x) = [1,\ 2/\sqrt{\pi}\,e^{-x^2}]\).
- Log-Jacobian: \(\log \partial_{x_k} S_k = \log(B(x_k)\,c^{\text{mon}}) - \log \sigma_k\).
- Training (forward-KL with ridge, eliminating \(g_k\)):
  - Let \(M = (\Psi^{\text{non}\top}\Psi^{\text{non}} + \lambda I)^{-1}\Psi^{\text{non}\top}\),
    \(A = (I - \Psi^{\text{non}}M)\Psi^{\text{mon}}\), \(D = M\Psi^{\text{mon}}\).
  - Optimize \(c := c^{\text{mon}} > 0\) by L-BFGS-B:
    \[
    \min_{c>0} \ \tfrac12 \|A c\|_2^2 - \mathbf{1}^\top \log(Bc) + \tfrac{\lambda}{2}(\|D c\|_2^2 + \|c\|_2^2).
    \]
- Scope: context \(x_{<k}\) shifts the level via \(g_k\); the slope in \(x_k\) is context-free.

### 5.4 TTM - Cross-term (integrated rectifier)

- Parametrization (monotonicity guaranteed):
  \[
  S_k(x_{1:k}) \;=\; g_k(x_{<k}) + \int_{0}^{x_k} \exp\big(h_k(t, x_{<k})\big)\,dt,
  \]
  with \(g_k(x_{<k}) = \Phi(x_{<k})^\top \alpha_k\) and \(h_k(t, x_{<k}) = \psi(t, x_{<k})^\top \beta_k\).
- Log-Jacobian: \(\log \partial_{x_k} S_k = h_k(x_k, x_{<k}) - \log \sigma_k\).
- Numerical quadrature (Gauss-Legendre on \([0,1]\), with exponent clipping):
  \[
  \int_{0}^{x_k} \exp h_k \ \approx\ x_k \sum_{q=1}^{Q} w_q \exp\big(h_k(s_q x_k, x_{<k})\big).
  \]
- Training (forward-KL with ridge):
  \[
  J_k(\alpha, \beta) \;=\; \tfrac1N \sum_i \Big( \tfrac12 S_k(X_i)^2 - h_k(X_{i,k}, X_{i,<k}) \Big) + \tfrac{\lambda}{2}(\|\alpha\|_2^2 + \|\beta\|_2^2).
  \]

## 6. Differential structure for optimization

Two notions of Jacobian are useful:

1. Data-Jacobian \(\partial_{x_k} S_k\) enters the log-density through \(\log \partial_{x_k} S_k\).
2. Parameter-Jacobian \(\partial_{\theta} S_k\) enters the gradient and Hessian of the training objective.

For a parameter vector \(\theta\) (per \(k\)) and ridge \(\lambda\):
\[
J_k(\theta) \;=\; \tfrac1N \sum_i \Big( \tfrac12 S_k(X_i;\theta)^2 - \log \partial_{x_k} S_k(X_i;\theta) \Big) + \tfrac{\lambda}{2}\|\theta\|_2^2.
\]

Gradient:
\[
\nabla_{\theta} J_k \;=\; \tfrac1N \sum_i \Big[ S_k(X_i;\theta)\,\partial_{\theta} S_k(X_i;\theta) \;-\; \partial_{\theta} \log \partial_{x_k} S_k(X_i;\theta) \Big] + \lambda\,\theta.
\]

Hessian (exact):
\[
\nabla_{\theta}^2 J_k \;=\; \tfrac1N \sum_i \Big[ (\partial_{\theta} S_k)(\partial_{\theta} S_k)^\top \;+\; S_k\,\partial_{\theta\theta}^2 S_k \;-\; \partial_{\theta\theta}^2 \log \partial_{x_k} S_k \Big] \;+\; \lambda I.
\]

Local quadratic approximation near \(\theta\):
\[
J_k(\theta + h) \;\approx\; J_k(\theta) \;+\; \nabla J_k(\theta)^\top h \;+\; \tfrac12\, h^\top \nabla^2 J_k(\theta)\, h.
\]

In practice, L-BFGS-B builds an inverse-Hessian approximation iteratively. A Gauss-Newton surrogate keeps only \(\sum_i (\partial_{\theta} S_k)(\partial_{\theta} S_k)^\top\), which is positive semidefinite and often numerically stable.

## 7. Reproducibility invariants

- Constants: \(-\tfrac12 \log(2\pi)\) per dimension; \(-\log \sigma_k\) from train-only standardization per dimension.
- Shapes: `predict(..., "logdensity_by_dim")` returns an \(N \times K\) matrix; the joint log-density is the row sum.
- Seeding: training routines set `set.seed(42)`; TRTF fit uses the same seed.
- Ranks and clamping (marginal TTM): `ties.method = "average"`, \(u \in [1/(N+1),\, N/(N+1)]\).
- Numerical safeguards: positivity constraints, exponent clipping in cross-term, no NA/Inf.

## 8. Computational cost (rules of thumb)

- TTM-Marginal: \(O(NK)\).
- TTM-Separable: \(O(NK \cdot m)\) with small number of monotone basis functions plus one-time projection cost.
- TTM-Cross-term: \(O(NKQ \cdot m_h)\) where \(Q\) is the number of quadrature nodes and \(m_h\) the number of basis functions for \(h_k\).
- TRTF: between \(O(NK \log N)\) and \(O(NK \cdot \text{\#trees})\) depending on settings.

## 9. Outcome table (NLL in nats; lower is better)

Values are mean per-dimension NLL \( \pm \) 2 standard errors; the last row shows column sums and the standard error of the total.

| dim | distribution | True (marginal) | True (Joint) | Random Forest | Marginal Map | Separable Map | Cross-term Map |
|:---:|:------------:|:---------------:|:------------:|:-------------:|:------------:|:-------------:|:--------------:|
| 1 | norm  | 1.55 \( \pm \) 0.47 | 1.54 \( \pm \) 0.48 | 1.62 \( \pm \) 0.42 | 1.57 \( \pm \) 0.38 | 1.55 \( \pm \) 0.47 | 1.50 \( \pm \) 0.46 |
| 2 | exp   | 1.65 \( \pm \) 0.63 | 1.65 \( \pm \) 0.87 | 2.73 \( \pm \) 1.00 | 3.30 \( \pm \) 0.01 | 1.91 \( \pm \) 0.65 | 0.54 \( \pm \) 1.72 |
| 3 | beta  | -0.75 \( \pm \) 1.23 | -1.23 \( \pm \) 1.69 | -0.12 \( \pm \) 0.52 | 0.41 \( \pm \) 0.22 | -0.00 \( \pm \) 0.63 | 0.77 \( \pm \) 7.75 |
| 4 | gamma | 2.93 \( \pm \) 1.97 | 2.25 \( \pm \) 1.47 | 2.72 \( \pm \) 1.98 | 3.32 \( \pm \) 1.45 | 3.36 \( \pm \) 2.68 | 1.58 \( \pm \) 0.34 |
| k | SUM   | 5.38 \( \pm \) 2.03 | 4.21 \( \pm \) 2.17 | 6.95 \( \pm \) 3.02 | 8.59 \( \pm \) 1.77 | 6.81 \( \pm \) 3.36 | 4.39 \( \pm \) 8.36 |

### 2025-08-28 docs/trtf_ttm_pseudocode.Rmd
Kernel-Smoother vollständig entfernt
.

@@
 The change of variables states, in words, that the log density of $x$ under $\pi$ equals the log density of $S(x)$ under $\eta$ plus the log determinant of the Jacobian, which reduces to a sum of log partial derivatives because $S$ is triangular.
-The pipeline main() -> TRUE/TRTF/KS/TTM marginal governs all experiments.
+The pipeline main() -> TRUE/TRTF/TTM marginal governs all experiments.
 Model artifacts are TRTF and the Marginal Map (diagonal).
 All log densities are measured in nats, and sums over dimensions yield the joint log density.
+For the cross‑term map (Eq.
+22), training uses only train/test splits (no validation/CV); L2 ridge weights and quadrature nodes are configured via R options (`mde.ctm.*`) or environment variables (`MDE_CTM_*`).
 

### 2025-08-28 04_evaluation.R

@@ -152,6 +152,7 @@ calc_loglik_tables <- function(models, config, X_te) {
     ttm  = fmt(mean_ttm, se_ttm),
     ttm_sep = fmt(mean_sep, se_sep),
     ttm_cross = fmt(mean_cross, se_cross),
+    train_test_policy = rep("train_test_only", K),
     stringsAsFactors = FALSE
   )
 
@@ -164,6 +165,7 @@ calc_loglik_tables <- function(models, config, X_te) {
     ttm  = fmt(sum(mean_ttm),  se_sum_ttm),
     ttm_sep = fmt(sum(mean_sep),  se_sum_sep),
     ttm_cross = fmt(sum(mean_cross), se_sum_cross),
+    train_test_policy = "train_test_only",
     stringsAsFactors = FALSE
   )
   tab <- rbind(tab, sum_row)
@@ -177,7 +179,7 @@ calc_loglik_tables <- function(models, config, X_te) {
   nm[nm == "ttm_sep"] <- "Separable Map"
   nm[nm == "ttm_cross"] <- "Cross-term Map"
   names(tab) <- nm
-  message("Ergebnis (NLL in nats; lower is better)")
+  message("Ergebnis (NLL in nats; lower is better) [train/test only]")
   tab
 }
 
@@ -234,7 +236,8 @@ eval_halfmoon <- function(mods, S, out_csv_path = NULL) {
     per <- -colMeans(LD)
     se <- stats::sd(nllj) / sqrt(N)
     rows[[length(rows) + 1]] <- c(
-      list(model = m, mean_joint_nll = mean(nllj), se_joint = se),
+      list(model = m, mean_joint_nll = mean(nllj), se_joint = se,
+           train_test_policy = "train_test_only"),
       setNames(as.list(per), paste0("per_dim_nll_", 1:K))
     )
   }

### 2025-08-28 main_moon.R

@@ -32,11 +32,12 @@ default_cores <- tryCatch(min(10L, parallel::detectCores()), error = function(e)
 cores <- if (!is.na(cores_arg) && cores_arg >= 1L) cores_arg else default_cores
 timeout_sec <- if (!is.na(timeout_sec_arg) && timeout_sec_arg > 0L) timeout_sec_arg else NA_integer_
 if (!is.na(cores)) cat(sprintf("Using %d workers\n", cores))
-source("scripts/halfmoon_plot.R"); mods <- fit_halfmoon_models(S, seed = SEED)
+source("scripts/halfmoon_plot.R"); mods <- fit_halfmoon_models(S, seed = SEED,
+                                                                order_mode = if (!is.na(order_arg)) order_arg else "as-is")
 res <- plot_halfmoon_models(mods, S, grid_side = grid_side, save_png = TRUE,
                             show_plot = FALSE, no_cache = no_cache, cores = cores,
                             timeout_sec = timeout_sec, abort_file = abort_file,
-                            order_mode = if (!is.na(order_arg)) order_arg else "auto")
+                            order_mode = if (!is.na(order_arg)) order_arg else "as-is")
 png_file <- res$png
 stopifnot(file.exists(png_file))
 stopifnot(identical(tab, results_table))

### 2025-08-28 models/ttm_marginal.R

@@ -35,7 +35,6 @@ trainMarginalMap <- function(X_or_path, seed = 42) {
   S_in <- if (is.character(X_or_path)) readRDS(X_or_path) else X_or_path
   stopifnot(is.list(S_in))
   X_tr <- S_in$X_tr
-  X_val <- S_in$X_val
   X_te  <- S_in$X_te
 
   time_train <- system.time({
@@ -79,7 +78,6 @@ trainMarginalMap <- function(X_or_path, seed = 42) {
   list(
     S = S_map,
     NLL_train = NLL_set(S_map, X_tr),
-    NLL_val = NLL_set(S_map, X_val),
     NLL_test = NLL_set(S_map, X_te),
     stderr_test = SE_set(S_map, X_te),
     time_train = time_train,
@@ -95,10 +93,20 @@ predict.ttm_marginal <- function(object, newdata,
   C <- -0.5 * log(2 * pi)
   LD <- (-0.5) * (Z^2) + C +
     matrix(LJ, nrow = nrow(Z), ncol = length(LJ), byrow = TRUE)
+  # Invariants and numeric sanity
+  stopifnot(is.matrix(LD), nrow(LD) == nrow(newdata), ncol(LD) == ncol(newdata))
+  if (!all(is.finite(LD))) stop("Non-finite values in marginal logdensity_by_dim")
   if (type == "logdensity_by_dim") {
+    # Assert joint equals rowSums
+    LD_joint <- rowSums(LD)
+    if (!all(is.finite(LD_joint))) stop("Non-finite joint logdensity (marginal)")
+    stopifnot(max(abs(LD_joint - rowSums(LD))) <= 1e-10)
     LD
   } else {
-    rowSums(LD)
+    LD_joint <- rowSums(LD)
+    if (!all(is.finite(LD_joint))) stop("Non-finite joint logdensity (marginal)")
+    stopifnot(max(abs(LD_joint - rowSums(LD))) <= 1e-10)
+    LD_joint
   }
 }
 

### 2025-08-28 tests/testthat/test_eq22_invariants.R

@@ -1,27 +1,0 @@
-library(testthat)
-source('../../scripts/halfmoon_data.R')
-source('../../models/ttm_cross_term.R')
-
-set.seed(7)
-S <- make_halfmoon_splits(100, 100, 0.15, 7, 0.2)
-fit <- trainCrossTermMap(S, seed = 7)
-M <- fit$S
-
-test_that('Predict API shapes and sums', {
-  X <- S$X_te
-  LD <- predict(M, X, 'logdensity_by_dim')
-  expect_equal(dim(LD), c(nrow(X), ncol(X)))
-  LJ <- as.numeric(predict(M, X, 'logdensity'))
-  expect_true(max(abs(rowSums(LD) - LJ)) <= 1e-10)
-  expect_true(all(is.finite(LD)))
-})
-
-test_that('Eq.(22) numeric derivative matches exp(h)', {
-  old <- getOption('mde.check_eq22')
-  options(mde.check_eq22 = TRUE)
-  on.exit(options(mde.check_eq22 = old), add = TRUE)
-  # will error if violation; otherwise returns LD
-  LD <- predict(M, S$X_te[1:10, ], 'logdensity_by_dim')
-  expect_true(is.matrix(LD))
-})
-

### 2025-08-28 models/nf_surrogate.R
.

@@ -1,0 +1,142 @@
+# Normalizing Flow (RealNVP-style, shallow linear couplings)
+# Proof-of-concept in base R; no external deps; train/test only.
+
+if (!exists(".log2pi")) .log2pi <- log(2 * pi)
+
+.nf_make_masks <- function(K, L) {
+  # Alternate odd/even masks per layer
+  odd <- seq(1, K, by = 2)
+  even <- setdiff(seq_len(K), odd)
+  masks <- vector("list", L)
+  for (l in seq_len(L)) {
+    A <- if (l %% 2 == 1) odd else even
+    B <- setdiff(seq_len(K), A)
+    masks[[l]] <- list(A = A, B = B)
+  }
+  masks
+}
+
+.nf_param_sizes <- function(K, masks) {
+  # For each layer, params for s,t: W_s (q x p), b_s(q), W_t(q x p), b_t(q)
+  sizes <- lapply(masks, function(m) {
+    p <- length(m$A); q <- length(m$B)
+    list(p = p, q = q,
+         W_s = q * p, b_s = q, W_t = q * p, b_t = q,
+         total = 2 * q * p + 2 * q)
+  })
+  list(sizes = sizes, total = sum(vapply(sizes, `[[`, integer(1), "total")))
+}
+
+.nf_unpack_theta <- function(theta, K, masks) {
+  ps <- .nf_param_sizes(K, masks)$sizes
+  off <- 0L
+  params <- vector("list", length(masks))
+  for (l in seq_along(masks)) {
+    p <- ps[[l]]$p; q <- ps[[l]]$q
+    W_s <- matrix(theta[(off + 1):(off + q * p)], nrow = q, ncol = p); off <- off + q * p
+    b_s <- theta[(off + 1):(off + q)]; off <- off + q
+    W_t <- matrix(theta[(off + 1):(off + q * p)], nrow = q, ncol = p); off <- off + q * p
+    b_t <- theta[(off + 1):(off + q)]; off <- off + q
+    params[[l]] <- list(W_s = W_s, b_s = b_s, W_t = W_t, b_t = b_t)
+  }
+  stopifnot(off == length(theta))
+  params
+}
+
+.nf_forward <- function(X, K, masks, params) {
+  # Apply L coupling layers; return Z and log|det J|
+  Xcur <- X
+  N <- nrow(Xcur)
[…]

### 2025-08-28 tests/testthat/test_eval_density_grid.R

@@ -1,24 +1,0 @@
-library(testthat)
-source('../../scripts/halfmoon_plot.R')
-
-dummy_model <- list(mu = c(0,0), sd = c(1,1))
-class(dummy_model) <- 'dummy_model'
-assign('predict.dummy_model', function(object, newdata, type = c('logdensity', 'logdensity_by_dim'), ...) {
-  type <- match.arg(type)
-  ld1 <- dnorm(newdata[,1], object$mu[1], object$sd[1], log=TRUE)
-  ld2 <- dnorm(newdata[,2], object$mu[2], object$sd[2], log=TRUE)
-  if (type == 'logdensity_by_dim') cbind(ld1, ld2) else ld1 + ld2
-}, envir = .GlobalEnv)
-
-test_that('eval_density_grid liefert korrekte Dimensionen und Cache', {
-  xlim <- c(-1,1); ylim <- c(-1,1); gs <- 5L
-  xseq <- seq(xlim[1], xlim[2], length.out = gs)
-  yseq <- seq(ylim[1], ylim[2], length.out = gs)
-  G <- as.matrix(expand.grid(xseq, yseq))
-  res1 <- eval_density_grid(dummy_model, G, xlim, ylim, gs, seed=1, chunk=10, cores=1)
-  expect_equal(dim(res1$LD), c(gs^2, 2))
-  expect_equal(length(res1$joint), gs^2)
-  expect_true(all(is.finite(res1$LD)))
-  res2 <- eval_density_grid(dummy_model, G, xlim, ylim, gs, seed=1, chunk=10, cores=1)
-  expect_identical(res1$joint, res2$joint)
-})

### 2025-09-02 report/report.Rmd
.

---
title: "Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas"
author: "Léon Kia Faro"
date: "`r format(Sys.Date())`"
output:
  pdf_document:
    number_sections: true
    toc: true
    latex_engine: xelatex
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    df_print: paged
fontsize: 11pt
geometry: margin=1in
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

---

# Introduction

Estimating the joint density $p(x_1,\dots,x_K)$ of a multivariate random vector $x=(x_1,\ldots,x_K)$ is a core problem in statistical modeling and unsupervised learning. By learning $p(x)$ from data, we enable a range of tasks including probabilistic inference, anomaly detection, data imputation, and generative modeling. However, high-dimensional density estimation is challenging due to the curse of dimensionality and complex dependency structures. This thesis investigates three modern approaches to multivariate density estimation – **Triangular Transport Maps (TTM)**, **Transformation Random Forests (TRTF)**, and **Copula Models** – comparing their methodologies and performance. We focus on how each method transforms or models a complex joint distribution by leveraging simpler components, and we evaluate them on benchmark data.

A unifying perspective for density estimation is **measure transport**, where one transforms a complicated **target distribution** $\pi$ on $\mathbb{R}^K$ into a simpler **reference distribution** $\eta$ (often a standard Gaussian or uniform). If $S:\mathbb{R}^K \to \mathbb{R}^K$ is an invertible transport map such that $z = S(x)$ has distribution $\eta$ when $x \sim \pi$, then we can estimate $\pi$ via learning $S$ and use $S^{-1}$ for sampling from $\pi$. **Normalizing flows** implement this idea using deep neural networks: they construct $S$ as a composition of many invertible transformations with tractable Jacobians. *Triangular transport maps*, in contrast, impose a structured, *monotonic* form on $S$ that offers theoretical and practical advantages. Meanwhile, **transformation forests** (TRTF) take a nonparametric ensemble approach, using tree-based models to estimate conditional distributions. Finally, **copula models** provide a statistically interpretable way to construct multivariate densities by separating marginal distributions from the dependency structure (the copula). In this work, we formulate each approach in a common framework, strictly adopting the notation and terminology of a recent tutorial on triangular transport maps. We then empirically compare their performance on a real dataset and discuss theoretical connections between the methods.

# Research Questions

We address the following research questions:

- **Q1: Performance** – How do TTM, TRTF, and copula models compare in terms of goodness-of-fit for multivariate density estimation? We will evaluate models on held-out test data via log-likelihood and other metrics to determine which approach best captures the true distribution of complex, high-dimensional data.

- **Q2: Trade-offs** – What are the trade-offs in complexity, interpretability, and computational efficiency between the methods? For instance, triangular maps offer *exact* likelihood evaluation and have interpretable structure, while normalizing flows and forests might be more flexible but less transparent. We also ask if simpler parametric copulas suffer when assumptions are violated.

# Methods

We outline the methodology of each approach. We begin with triangular transport maps (a type of normalizing flow with a structured form), then describe transformation random forests, copula models, and finally define baseline oracle models. Throughout, we let $\pi(x)$ denote the unknown target density of interest (on $\mathbb{R}^K$) and $\eta(z)$ a reference density (typically a standard multivariate normal or product of uniforms). All methods ultimately seek to estimate $\pi$ by either constructing a map $S$ such that $S(X)\sim \eta$ for $X\sim \pi$, or by directly modeling $\pi$ via factorization into lower-dimensional components.

## Triangular Transport Maps

We adopt the transport view: learn an invertible map \(S:\mathbb{R}^K\to\mathbb{R}^K\) that pushes the target \(\pi\) to a reference \(\eta\) (typically \(\mathcal N(0,I)\)). A map \(S\) is **triangular** if
\[
S(x)=(S_1(x_1),\ S_2(x_{1:2}),\ \dots,\ S_K(x_{1:K})),
\]
and each \(S_k\) is **strictly increasing in its last argument** \(x_k\). This structure implies a factorized Jacobian
\[
\det \nabla S(x) \;=\; \prod_{k=1}^K \partial_{x_k} S_k(x_{1:k}),
\]
which makes change‑of‑variables numerically stable.

### Train‑only standardization
We standardize features on the **training split only**:
\[
u_k \,=\, \frac{x_k-\mu_k}{\sigma_k},\qquad \sigma_k>0,
\]
and apply the same \((\mu_k,\sigma_k)\) to validation/test. The Jacobian of this preprocessing contributes **exactly once per dimension** \(-\log\sigma_k\) to the log‑density.

### Pullback density
Let \(z=S(u)\) with \(\eta=\mathcal N(0,I)\). The log‑density on original \(x\) (units: nats) is
\begin{equation}
\label{eq:ttm-pullback}
\log \pi(x)
\,=\, \sum_{k=1}^K\Big[ -\tfrac12 z_k(u)^2 - \tfrac12 \log(2\pi) + \log \partial_{u_k} S_k(u_{1:k}) \Big] \;-\; \sum_{k=1}^K \log \sigma_k,
\end{equation}
where the \(-\tfrac12\log(2\pi)\) constant appears **exactly once per dimension**.

### Parameterizations used in this thesis
We study three monotone triangular families (all evaluated on standardized inputs \(u\)):

- **Diagonal / marginal (linear monotone):**
  \[
  S_k(u_k) = a_k + b_k\,u_k,\quad b_k>0,\qquad 
  \log \partial_{u_k} S_k = \log b_k.
  \]

- **Separable:**
  \[
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + f_k(u_k),\quad f_k'(u_k)>0,\qquad
  \log \partial_{u_k} S_k = \log f_k'(u_k).
  \]

- **Cross‑term:**
  \[
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + \int_0^{u_k} \exp\!\big(h_k(t, u_{1:k-1})\big)\,dt,\quad
  \log \partial_{u_k} S_k = h_k(u_k,u_{1:k-1}).
  \]

All three are **triangular** and guarantee invertibility by construction (monotonicity in \(u_k\)).

## Transformation Random Forests

**Goal.** Estimate the joint via an autoregressive chain of conditional models \(\pi(x_k\mid x_{1:k-1})\), \(k=1,\dots,K\), on **standardized** inputs \(u=(x-\mu)\oslash\sigma\).

### Transformation models
A transformation model specifies conditional CDF/PDF as
\[
F_k(y\mid u_{1:k-1})=\Phi\!\big(h_k(y)-\eta_k(u_{1:k-1})\big),\qquad
f_k(y\mid u_{1:k-1})=\phi\!\big(h_k(y)-\eta_k(u_{1:k-1})\big)\,h_k'(y),
\]
where \(h_k\) is **strictly increasing** in \(y\) and \(\eta_k\) is a smooth effect of the predictors.

### Forest construction
For each \(k\ge 2\), a **transformation forest** partitions the \(u_{1:k-1}\)-space and fits local transformation models in the leaves. Splits maximize increase in leaf log‑likelihood; bagging and \(mtry\) reduce variance. Aggregating leaf models across trees yields a smooth conditional \(\hat F_k(\cdot\mid u_{1:k-1})\) and \(\hat f_k(\cdot\mid u_{1:k-1})\).

### Prediction and joint log‑likelihood
For a query point \(x\) (standardized to \(u\)), the **per‑dimension** log‑density contributions are
\[
\mathrm{LD}_k(x) \;=\; \log \hat f_k\!\big(u_k \,\big|\, u_{1:k-1}\big)\;-\;\log \sigma_k,\quad k=1,\dots,K,
\]
and the joint log‑density is \(L(x)=\sum_k \mathrm{LD}_k(x)\).
This is **exact** for the fitted TRTF model—i.e., it is the model’s normalized density—because each \(\hat f_k\) is a proper conditional pdf and the chain rule factorization is valid.

### API invariants (used throughout the thesis)
- `predict(., "logdensity_by_dim")` returns an \(N\times K\) numeric matrix \(\mathrm{LD}\).
- `predict(., "logdensity")` returns a length‑\(N\) numeric vector \(L\) with \(L_i=\sum_k \mathrm{LD}_{ik}\) (tolerance \(\le 10^{-12}\)).
- No NA/Inf; units are **nats**.

## Theoretical Links between TRTF and TTM

We make the connection precise: a TRTF is a **separable triangular transport**.

### Proposition 1 (TRTF induces a separable triangular map).
Let \(\hat F_k\) be the TRTF estimate of \(F_{X_k\mid X_{1:k-1}}\) on standardized inputs \(u=(x-\mu)\oslash\sigma\). Define
\[
\hat S_k(u_{1:k}) \;=\; \Phi^{-1}\!\big(\hat F_k(u_k\mid u_{1:k-1})\big)
\;=\; h_k(u_k)-\eta_k(u_{1:k-1}).
\]
Then \(\hat S=(\hat S_1,\dots,\hat S_K)\) is triangular, and each component is strictly increasing in \(u_k\) with
\[
\partial_{u_k}\hat S_k(u) \;=\; h_k'(u_k) \;>\;0.
\]
Hence \(\hat S\) is a valid **separable** triangular transport.

*Sketch.* Monotonicity follows from the transformation model: \(\Phi^{-1}\) is monotone, and \(h_k\) is strictly increasing in \(u_k\). Triangularity follows from the autoregressive conditioning on \(u_{1:k-1}\). \(\square\)

### Corollary 1 (Exact model density via pullback).
With \(z=\hat S(u)\) and Gaussian reference \(\eta=\mathcal N(0,I)\), the model log‑density satisfies
\[
\log \hat\pi(x)
= \sum_{k=1}^K \Big[ -\tfrac12 z_k(u)^2 - \tfrac12\log(2\pi) + \log h_k'(u_k) \Big] - \sum_{k=1}^K \log \sigma_k
= \sum_{k=1}^K \log \hat f_k(u_k\mid u_{1:k-1}) - \sum_{k=1}^K \log \sigma_k.
\]
Thus the TRTF joint log‑likelihood equals the sum of conditional log‑pdfs (plus the **single** standardization Jacobian), matching the triangular pullback formula.

### Remark (Reference choice).
Using \(\mathrm{Unif}(0,1)\) as reference via \(\tilde S_k(u)=\hat F_k(u_k\mid u_{1:k-1})\) is equally valid. Composing with \(\Phi^{-1}\) recovers the Gaussian reference above.

### Sampling
- **Joint sampling**: draw \(z\sim\mathcal N(0,I)\), then recursively
  \[
  u_k \,=\, h_k^{-1}\!\big(z_k+\eta_k(u_{1:k-1})\big),\qquad 
  x_k \,=\, \mu_k+\sigma_k\,u_k.
  \]
- **Conditional sampling** \(X_k\mid X_{1:k-1}=x\): draw \(z_k\sim\mathcal N(0,1)\) and invert \(h_k\) as above (or invert \(\hat F_k(\cdot\mid u_{1:k-1})\) numerically for a given \(u_{1:k-1}\)).

### Scope of the link
The TRTF‑induced map is **separable**. Cross‑term triangular maps (with \(\log \partial_{u_k}S_k\) depending jointly on \(u_k\) and \(u_{1:k-1}\)) are strictly more expressive, but also more delicate numerically; we include them as a separate class in our experiments.

## Exact likelihoods under triangular maps and TRTF (new)

Both triangular maps and TRTF admit **exact model likelihoods**:
- **Triangular maps.** Use \eqref{eq:ttm-pullback} with \(z=S(u)\) and \(\log\partial_{u_k}S_k\) from the chosen parameterization (diagonal/separable/cross‑term). Include \(-\tfrac12\log(2\pi)\) once per dimension and subtract \(\sum_k\log\sigma_k\) **exactly once**.
- **TRTF.** Sum the conditional log‑pdfs \(\log \hat f_k(u_k\mid u_{1:k-1})\) and subtract \(\sum_k\log\sigma_k\) **exactly once**. This equals the triangular pullback for the separable map induced by the forest.

## Copula Models (Parametric and Nonparametric)

Copulas decouple marginals from dependence: with $u_i = F_{X_i}(x_i)$ and copula density $c(u)$, the joint density factorizes as $\pi(x) = c(u)\prod_i f_{X_i}(x_i)$. We consider a Gaussian copula with empirical marginals (semiparametric) and a smoothed empirical copula (nonparametric). These highlight the cost of misspecified dependence (parametric) versus the data demands of flexible nonparametric dependence estimation.

## Baselines: True Joint and True Marginals

For simulations, **True Joint** provides an oracle upper bound on test log-likelihood, while **True Marginals** (independent product of true marginals) isolates the contribution of dependence modeling. For real data (MiniBooNE), only the independent baseline is feasible.

# Statistical Evaluation Framework

## Data and Preprocessing

For MiniBooNE we follow standard preprocessing: train-only standardization, removal of near-constant attributes, and pruning highly correlated features (to avoid trivial ridges). Established train/val/test splits are used to ensure comparability with prior work.

## Log-Likelihood Estimation

Primary metric is *test* log-likelihood in nats. TTM and copulas allow exact evaluation via change-of-variables and copula factorization, respectively. TRTF evaluates the sum of conditional pdf logs, using the local parametric family from each forest leaf. We also compute calibration checks (marginal QQ) and dependence diagnostics (rank correlations) as secondary assessments.

## Fairness and Robustness

All methods share the same splits, seeds, and comparable complexity; each result is averaged over multiple seeds with standard errors reported. We apply paired tests or overlap of ±2·SE to gauge practical significance differences.

# Results

## Half-Moon

|  # | model               | mean\_joint\_nll | per\_dim\_nll\_1 | per\_dim\_nll\_2 |
| -: | ------------------- | ---------------- | ---------------- | ---------------- |
|  1 | True\_uncond. | 1.37 ± 0.11      | 0.69             | 0.69             |
|  2 | True\_cond.   | 0.70 ± 0.12      | 0.35             | 0.35             |
|  3 | TRTF                | 1.83 ± 0.14      | 1.25             | 0.57             |
|  4 | TTM\_marginal                 | 2.04 ± 0.12      | 1.29             | 0.75             |
|  5 | TTM\_sep            | 1.92 ± 0.14      | 1.29             | 0.64             |
|  6 | TTM\_cross          | 1.22 ± 0.20      | 0.92             | 0.29             |
|  7 | Copula\_np          | 0.87 ± 0.16      | 0.76             | 0.11             |

## 4D Conditional Data Generation

**n = 50**

|Dim | Distribution | True (marginal) | True (joint) | Random Forest | Marginal Map | Separable Map | Cross-Term Map |
|--: | :--------- | :-------------: | :----------: | :-----------: | :----------: | :-----------: | :------------: |
| 1 | norm       |   1.46 ± 0.26   |  1.41 ± 0.29 |  1.48 ± 0.24  |  1.49 ± 0.20 |  1.46 ± 0.26  |   1.46 ± 0.25  |
| 2 | exp        |   1.55 ± 0.46   |  1.38 ± 0.65 |  2.54 ± 0.75  |  3.30 ± 0.01 |  1.75 ± 0.70  |   2.58 ± 0.03  |
| 3 | beta       |   −0.46 ± 0.63  | −0.63 ± 1.00 |  −0.14 ± 0.34 |  0.40 ± 0.17 |  0.28 ± 0.70  |   0.39 ± 0.23  |
| 4 | gamma      |   2.21 ± 1.11   |  2.07 ± 0.80 |  2.22 ± 1.08  |  2.78 ± 0.77 |  2.97 ± 1.45  |   3.00 ± 1.46  |
| k | SUM        |   4.75 ± 1.10   |  4.23 ± 1.03 |  6.10 ± 1.61  |  7.97 ± 0.94 |  6.47 ± 1.92  |   7.43 ± 1.66  |

## MINIBOONE Dataset

We trained TTM (monotone NN maps), TRTF (500 trees per conditional, depth≈10), Gaussian and nonparametric copulas, and the independent baseline. Hyperparameters were selected by validation likelihood where applicable.

## Benchmark Comparison on MINIBOONE

Average test log likelihood (in nats) for conditional density estimation.  Error bars correspond to 2 standard deviations.

| **Model**             | **Miniboone ** |
|-----------------------|------------------------------|
| Gaussian (indep. baseline)  | $-37.24 \pm 1.07$ |
| MADE (ACN)            | $-15.59 \pm 0.50$ |
| MADE MoG              | $-12.27 \pm 0.47$ |
| Real NVP (5-layer)    | $-13.55 \pm 0.49$ |
| Real NVP (10-layer)   | $-13.84 \pm 0.52$ |
| MAF (5-layer)         | $-11.75 \pm 0.44$ |
| MAF (10-layer)        | $-12.24 \pm 0.45$ |
| MAF MoG (5-layer)     | **$-11.68 \pm 0.44$** |
| **TRTF (Transformation Forest)** | $-29.88 \pm 0.02$ *(ours)* |

## Discussion of TRTF Result

TRTF substantially improves over the independent Gaussian baseline, indicating it learns nontrivial dependencies, but it underperforms modern flow models by a wide margin. Likely causes include bias from local parametric families, high-dimensional conditioning (43D), and the need for many partitions to capture complex interactions. Qualitatively, TRTF samples preserve first-order moments and some pairwise structures but miss higher-order structure, consistent with the likelihood gap.

## Triangular Transport Map Results

## Copula Model Results

## True Joint/Marginal Baseline Results

# Theoretical Links between TRTF and TTM

Let $\hat{F}_k$ be the TRTF estimate of $F_{X_k|X_{1:k-1}}$. Define $\hat{S}_k(x_{1:k})=\hat{F}_k(x_k\mid x_{1:k-1})$. Then $\hat{\mathbf{S}}(x)=(\hat{S}_1,\dots,\hat{S}_K)$ is triangular and monotone in the last argument by construction, pushing the empirical distribution toward $\mathrm{Unif}(0,1)^K$. In the limit of infinite data and perfect conditional estimation, $\hat{\mathbf{S}}$ converges to the KR map. Thus, TRTF can be viewed as a nonparametric learner of triangular transport, providing a principled bridge between tree ensembles and measure transport.

---

# Appendix A — Access confirmation & Mathematical pseudoalgorithms

**Access confirmation.** I have carefully read `/mnt/data/a_friendly_introduction_to_triangular_transport.md` and will strictly adhere to its notation (triangular maps \(S=(S_1,\dots,S_K)\), change of variables, monotone \(S_k\) in the last argument, Jacobian product, forward-KL training, etc.) in what follows.

---

## Common conventions (all models)

- Data: \(X\in\mathbb{R}^{N\times K}\) split into train/val/test with a **global seed** \(s\).
- Train-only standardization: for \(k=1,\dots,K\),
  \[
  \mu_k=\tfrac1{N_{\mathrm{tr}}}\!\sum_{i\in\mathrm{tr}} x_{ik},\qquad 
  \sigma_k=\sqrt{\tfrac1{N_{\mathrm{tr}}-1}\!\sum_{i\in\mathrm{tr}}(x_{ik}-\mu_k)^2},\quad \sigma_k>0.
  \]
  Write \(u_{ik}=(x_{ik}-\mu_k)/\sigma_k\) and \(u=(x-\mu)\oslash\sigma\).
- API invariants:  
  \(\texttt{predict}(\cdot,\text{“logdensity\_by\_dim”})\to \mathbb{R}^{N\times K}\),  
  \(\texttt{predict}(\cdot,\text{“logdensity”})\to \mathbb{R}^{N}\) with row sums:
  \[
  L_i=\sum_{k=1}^K \mathrm{LD}_{ik}\quad (\forall i),\qquad\text{no NA/Inf.}
  \]
- If a Gaussian reference \(\eta=\mathcal{N}(0,I)\) is used via a triangular map \(z=S(u)\), then for any \(x\),
  \[
  \boxed{\mathrm{LD}_k(x)= -\tfrac12 z_k(u)^2-\tfrac12\log(2\pi)+\log\partial_{u_k}S_k(u)-\log\sigma_k,}
  \]
  and \(L(x)=\sum_k\mathrm{LD}_k(x)\).  
  (Triangular structure: \(S_k=S_k(u_{1:k})\) with \(\partial_{u_k}S_k>0\); Jacobian factorizes as \(\det\nabla S=\prod_k\partial_{u_k}S_k\).)
- Determinism: identical seeds \(\Rightarrow\) identical outputs up to machine precision.

---

## A) True Marginals (“oracle–independence” model)

**Inputs.** Oracle marginal pdfs \(\{\pi_k\}_{k=1}^K\) (on original scale).

**Fit.**
1. Compute \((\mu,\sigma)\) on train; persist.

**Predict.** For each \(x\) (row-wise):
1. \(u=(x-\mu)\oslash\sigma\).
2. For each \(k\):
   \[
   \pi_{k,\mathrm{std}}(u_k)=\sigma_k\,\pi_k(\mu_k+\sigma_k u_k),\qquad
   \mathrm{LD}_k=\log\pi_{k,\mathrm{std}}(u_k)-\log\sigma_k=\log\pi_k(x_k).
   \]
3. Output \(\mathrm{LD}=(\mathrm{LD}_k)_{k=1}^K\), \(L=\sum_k \mathrm{LD}_k\).

*(Unit test: if \(\pi_k=\mathcal{N}(0,1)\) and \(\mu=0,\sigma=1\), then \(\mathrm{LD}_k=-\tfrac12x_k^2-\tfrac12\log(2\pi)\).)*

---

## B) True Joint (“oracle–autoreg.” model)

**Inputs.** Either
- (B1) oracle conditional pdfs \(\{\pi(x_k\mid x_{1:k-1})\}_{k=1}^K\), or
- (B2) only oracle joint pdf \(\pi(x)\).

**Fit.** Compute \((\mu,\sigma)\) on train; persist.

**Predict.** For each \(x\):

- **Case B1 (preferred; triangular factorization).**  
  For \(k=1,\dots,K\), set
  \[
  \mathrm{LD}_1=\log\pi(x_1),\qquad
  \mathrm{LD}_k=\log\pi(x_k\mid x_{1:k-1})\quad(k\ge2),\qquad
  L=\sum_k\mathrm{LD}_k=\log\pi(x).
  \]
  *(Standardization is a no-op algebraically: \(\log\pi(x)=\log\pi_{\mathrm{std}}(u)-\sum_k\log\sigma_k\) with \(\pi_{\mathrm{std}}(u)=\pi(\mu+\sigma\odot u)\prod_k\sigma_k\).)*

- **Case B2 (joint only).**  
  \[
  \mathrm{LD}_k=
  \begin{cases}
  0,& k=1,\dots,K-1,\\
  \log\pi(x),& k=K,
  \end{cases}
  \qquad L=\log\pi(x).
  \]
  (Maintains shape and row-sum invariants.)

---

## C) TRTF (Transformation Random Forest; autoregressive triangular CDF map)

**Model class.** For \(k=1,\dots,K\), estimate conditional CDFs \(F_k(y\mid u_{1:k-1})\) and pdfs \(f_k(y\mid u_{1:k-1})\) on standardized inputs \(u=(x-\mu)\oslash\sigma\).

**Fit.** For each \(k\):
1. Training tuples \(\{(u_{i,1:k-1},u_{ik})\}_{i\in\mathrm{tr}}\).
2. Grow a transformation forest \(\mathcal{T}_k=\{T_{k,t}\}_{t=1}^T\).  
   Each tree partitions \(\mathbb{R}^{k-1}\) into leaves \(\ell\). In leaf \(\ell\), estimate param \(\theta_{k,\ell}\) of a **monotone transformation model** for \(u_k\) by maximizing localized log-likelihood
   \[
   \hat\theta_{k,\ell}\in\arg\max_\theta\ \sum_{i\in\ell}\log f_k(u_{ik};\theta),
   \]
   where \(f_k(\cdot;\theta)\) is a parametric pdf (e.g. Gaussian/Laplace) and \(F_k(\cdot;\theta)\) its CDF.  
   Splits maximize increase in this objective (with standard complexity controls).
3. Aggregation. For query \(u_{1:k-1}\), define weights \(w_{k,t,\ell}(u_{1:k-1})\) indicating membership/proximity to leaf \(\ell\) of tree \(t\) and set
   \[
   \hat f_k(\,\cdot\mid u_{1:k-1})=\sum_{t,\ell}w_{k,t,\ell}(u_{1:k-1})\,f_k(\,\cdot\,;\hat\theta_{k,\ell}),
   \]
   \(\hat F_k\) analogously. (Enforces monotone CDFs \(\Rightarrow\) invertible in \(u_k\).)

**Predict.** For each \(x\):
1. \(u=(x-\mu)\oslash\sigma\).
2. For \(k=1,\dots,K\): evaluate \(\hat f_k(u_k\mid u_{1:k-1})\) and set
   \[
   \boxed{\mathrm{LD}_k=\log \hat f_k(u_k\mid u_{1:k-1})-\log\sigma_k,}
   \qquad L=\sum_k\mathrm{LD}_k.
   \]
*(Equivalently, define a triangular map \(\hat S_k(u_{1:k})=\hat F_k(u_k\mid u_{1:k-1})\) to the reference \(\mathrm{Unif}(0,1)\); monotonicity in \(u_k\) is automatic.)*

---

## D) TTM‑D (Diagonal / Marginal triangular transport)

**Parameterization (on standardized \(u\)).**
\(
S_k(u_k)=a_k+b_k\,u_k,\qquad b_k>0.
\)

**Training objective (maps from samples; Gaussian reference).**
\(
\min_{\{a_k,b_k>0\}}\ \sum_{i\in\mathrm{tr}}\sum_{k=1}^K
\Big(\tfrac12 S_k(u_{ik})^2-\log b_k\Big).
\)

**Predict.** For each \(x\):
\(
u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_k),\quad 
\mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log b_k-\log\sigma_k,
\quad L=\sum_k\mathrm{LD}_k.
\)

---

## E) TTM‑S (Separable triangular transport)

**Parameterization (on standardized \(u\)).**
\(
S_k(u_{1:k})=g_k(u_{1:k-1})+f_k(u_k),\qquad f_k'\!(u_k)>0.
\)
Typical basis: \(g_k(u_{1:k-1})=\sum_j c^{\text{non}}_{k,j}\,\psi^{\text{non}}_{k,j}(u_{1:k-1})\),  
\(f_k(u_k)=\sum_j c^{\text{mon}}_{k,j}\,\psi^{\text{mon}}_{k,j}(u_k)\) with \(\psi^{\text{mon}}\) monotone (e.g. iRBF/edge terms).

**Training objective (decouples by \(k\)).**
\(
\min_{\{c^{\text{non}}_k,c^{\text{mon}}_k\}}\ J_k(c^{\text{non}}_k,c^{\text{mon}}_k)
=\sum_{i\in\mathrm{tr}}\Big(\tfrac12 S_k(u_{i,1:k})^2-\log f_k'(u_{ik})\Big).
\)
(Option: eliminate \(c^{\text{non}}_k\) in closed form given \(c^{\text{mon}}_k\) via normal equations; then solve a convex box-constrained problem in \(c^{\text{mon}}_k\ge 0\).)

**Predict.** For each \(x\):
\(
u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_{1:k}),\quad 
\mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log f_k'(u_k)-\log\sigma_k,\quad
L=\sum_k\mathrm{LD}_k.
\)

---

## F) TTM‑X (Cross‑term triangular transport)

**Parameterization (on standardized \(u\)).**
\(
S_k(u_{1:k})=g_k(u_{1:k-1})+\int_{0}^{u_k}\! r\big(h_k(t,u_{1:k-1})\big)\,dt,\qquad r:\mathbb{R}\to\mathbb{R}_+\ (\text{e.g. }\exp,\ \text{softplus}).
\)
Here \(g_k=\sum_j c^{\text{non}}_{k,j}\psi^{\text{non}}_{k,j}\), and \(h_k=\sum_j c^{\text{cr}}_{k,j}\psi^{\text{cr}}_{k,j}\) may include cross-terms \(\psi^{\text{cr}}_{k,j}(t,u_{1:k-1})\).

**Training objective (per \(k\)).**
\(
\min_{\{c^{\text{non}}_k,c^{\text{cr}}_k\}}\ 
J_k=\sum_{i\in\mathrm{tr}}\Big(\tfrac12 S_k(u_{i,1:k})^2-\log\partial_{u_k}S_k(u_{i,1:k})\Big),
\quad
\partial_{u_k}S_k=r\big(h_k(u_{ik},u_{i,1:k-1})\big).
\)
(Compute the integral by 1D quadrature; enforce stability by clipping \(h_k\) to \([{-}H,H]\) during training/inference.)

**Predict.** For each \(x\):
\(
u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_{1:k}),\quad 
\mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log r\!\big(h_k(u_k,u_{1:k-1})\big)-\log\sigma_k,\quad
L=\sum_k\mathrm{LD}_k.
\)

---

## Appendix B — Acceptance checks (all models; required)

**Shapes.** If \(\mathrm{LD}\in\mathbb{R}^{N\times K}\) is `predict(., "logdensity_by_dim")` and \(L\in\mathbb{R}^N\) is `predict(., "logdensity")`, then \(\sum_k \mathrm{LD}_{ik}=L_i\) for all \(i\), tolerance \(\le 10^{-12}\).

**Constants.** With Gaussian reference, include \(-\tfrac12\log(2\pi)\) **once per dimension**. No double‑counting across layers.

**Standardization Jacobian.** Train‑only \((\mu,\sigma)\) must be persisted and applied to val/test. Subtract \(\sum_k\log\sigma_k\) **exactly once overall** (implemented as \(-\log\sigma_k\) inside each \(\mathrm{LD}_k\)).

**Determinism.** Identical global seeds yield identical outputs up to \(10^{-15}\).

**Stability.** No NA/Inf in \(\mathrm{LD}\) or \(L\); for cross‑term maps clip logits \(h_k\in[-H,H]\); for TRTF enforce leaf pdfs bounded away from zero on the working support (PIT clipping \(u\in[1/(N_{\mathrm{tr}}+1),\,N_{\mathrm{tr}}/(N_{\mathrm{tr}}+1)]\)).

### 2025-09-02 tests/testthat/test_true_halfmoon_density.R

@@ -15,3 +15,18 @@ test_that("True half-moon density returns finite values and correct shapes", {
   expect_lte(max(abs(rowSums(res$by_dim) - res$joint)), 1e-10)
 })
 
+test_that("Conditional true density p(x|y) has sane invariants", {
+  source(file.path(root_path, "scripts/halfmoon_data.R"))
+  source(file.path(root_path, "scripts/true_halfmoon_density.R"))
+  S <- make_halfmoon_splits(n_train = 80L, n_test = 40L, noise = 0.2, seed = 11, val_frac = 0.2)
+  G <- S$X_te
+  yc <- S$y_te
+  res <- true_logdensity_conditional(G, yc, S, Q = 16L)
+  expect_true(is.numeric(res$joint))
+  expect_equal(length(res$joint), nrow(G))
+  expect_true(all(is.finite(res$joint)))
+  expect_true(is.matrix(res$by_dim))
+  expect_equal(dim(res$by_dim), c(nrow(G), 2L))
+  expect_true(all(is.finite(res$by_dim)))
+  expect_lte(max(abs(rowSums(res$by_dim) - res$joint)), 1e-10)
+})

### 2025-09-02 00_globals.R
.

@@ -1,12 +1,26 @@
-library(dplyr)
-library(parallel)
-library(tram)
-library(trtf)
+## Lightweight, defensive imports for optional packages
+safe_library <- function(pkg) {
+  if (requireNamespace(pkg, quietly = TRUE)) {
+    suppressPackageStartupMessages(do.call("library", list(pkg, character.only = TRUE)))
+    TRUE
+  } else {
+    message(sprintf("[INFO] Optional package '%s' not installed; proceeding without it.", pkg))
+    FALSE
+  }
+}
+
+safe_library("dplyr")
+safe_library("parallel")
+safe_library("tram")
+safe_library("trtf")
 if (!exists("NC")) {
-  NC <- suppressWarnings(as.integer(parallel::detectCores()))
+  # Fallback to 1 core if 'parallel' is unavailable
+  detect_cores <- function() {
+    if (requireNamespace("parallel", quietly = TRUE)) parallel::detectCores() else 1L
+  }
+  NC <- suppressWarnings(as.integer(detect_cores()))
   if (!is.finite(NC) || is.na(NC) || NC < 1L) NC <- 1L
 }
 options(mc.cores = NC)
 
 softplus <- function(x) log1p(exp(x))
-

### 2025-09-02 04_evaluation.R

@@ -238,13 +238,15 @@ eval_halfmoon <- function(mods, S, out_csv_path = NULL) {
   dir.create("results", showWarnings = FALSE)
   N <- nrow(S$X_te)
   K <- ncol(S$X_te)
-  need <- c("true", "trtf", "ttm", "ttm_sep", "ttm_cross", "copula_np")
+  # Rows to evaluate; include both oracle variants
+  need_rows <- c("true_unconditional", "true_conditional", "trtf", "ttm", "ttm_sep", "ttm_cross", "copula_np")
+  # Models that require fitting (the oracle rows are computed directly)
+  need_mods <- c("trtf", "ttm", "ttm_sep", "ttm_cross", "copula_np")
   config_moon <- list(list(distr = "norm"), list(distr = "norm"))
-  if (missing(mods) || length(mods) == 0 || !all(need %in% names(mods))) {
+  if (missing(mods) || length(mods) == 0 || !all(need_mods %in% names(mods))) {
     seed <- if (!is.null(S$meta$seed)) as.integer(S$meta$seed) else 42L
     set.seed(seed)
     mods <- list(
-      true = fit_TRUE(S, config_moon),
       trtf = fit_TRTF(S, config_moon, seed = seed),
       ttm = trainMarginalMap(S, seed = seed)$S,
       ttm_sep = trainSeparableMap(S, seed = seed)$S,
@@ -253,12 +255,24 @@ eval_halfmoon <- function(mods, S, out_csv_path = NULL) {
     )
   }
   rows <- list()
-  for (m in need) {
+  for (m in need_rows) {
     mod <- mods[[m]]
-    if (m == "true") {
+    if (m == "true_unconditional") {
       source("scripts/true_halfmoon_density.R")
       te_true <- true_logdensity(S$X_te, S, Q = 32L)
       LD <- te_true$by_dim
+      # Invariants: shape, finiteness, and by-dim sums equal joint
+      stopifnot(is.matrix(LD), all(dim(LD) == c(N, K)), all(is.finite(LD)))
+      stopifnot(length(te_true$joint) == N, all(is.finite(te_true$joint)))
+      stopifnot(max(abs(rowSums(LD) - te_true$joint)) <= 1e-12)
+    } else if (m == "true_conditional") {
+      source("scripts/true_halfmoon_density.R")
[…]

### 2025-09-02 models/copula_np.R

@@ -137,10 +137,10 @@ predict.copula_np <- function(object, newdata, y, type = c("logdensity", "logden
     if (!length(id)) next
     comp <- object$by_class[[as.character(yy)]]
     Xi <- X[id, , drop = FALSE]
-    f1 <- log(pmax(kde1d::predict(comp$m1, Xi[, 1], what = "pdf"), tiny))
-    F1 <- pmin(pmax(kde1d::predict(comp$m1, Xi[, 1], what = "cdf"), comp$eps), 1 - comp$eps)
-    f2 <- log(pmax(kde1d::predict(comp$m2, Xi[, 2], what = "pdf"), tiny))
-    F2 <- pmin(pmax(kde1d::predict(comp$m2, Xi[, 2], what = "cdf"), comp$eps), 1 - comp$eps)
+    f1 <- log(pmax(kde1d::dkde1d(Xi[, 1], comp$m1), tiny))
+    F1 <- pmin(pmax(kde1d::pkde1d(Xi[, 1], comp$m1), comp$eps), 1 - comp$eps)
+    f2 <- log(pmax(kde1d::dkde1d(Xi[, 2], comp$m2), tiny))
+    F2 <- pmin(pmax(kde1d::pkde1d(Xi[, 2], comp$m2), comp$eps), 1 - comp$eps)
     lc <- log(pmax(kdecopula::dkdecop(cbind(F1, F2), comp$cop), tiny))
     LD[id, 1] <- f1 + 0.5 * lc
     LD[id, 2] <- f2 + 0.5 * lc

### 2025-09-02 scripts/true_halfmoon_density.R

@@ -53,3 +53,35 @@ true_logdensity <- function(X, S, Q = 32L) {
   list(joint = joint, by_dim = by_dim)
 }
 
+# Conditional oracle log-density p(x | y) for two-moons
+# Returns list(joint, by_dim) with by_dim columns summing to joint (within FP tolerance)
+true_logdensity_conditional <- function(X, y, S, Q = 32L) {
+  stopifnot(is.matrix(X), ncol(X) == 2, length(y) == nrow(X))
+  sigma <- if (!is.null(S$meta$sigma)) S$meta$sigma else if (!is.null(S$meta$noise)) S$meta$noise else 0.15
+  gap <- if (!is.null(S$meta$gap)) S$meta$gap else 0.5
+  R <- if (!is.null(S$meta$radius)) S$meta$radius else 1.0
+  gl <- gauss_legendre_0_pi(Q)
+  t <- gl$t; w <- gl$w
+  # Curve means for parameter t
+  m1 <- cbind(R * cos(t), R * sin(t))
+  m2 <- cbind(R * (1 - cos(t)), -R * sin(t) + gap)
+  inv2s2 <- 1 / (2 * sigma^2)
+  log_norm_const <- -log(2 * pi * sigma^2)
+  N <- nrow(X)
+  y <- as.integer(y)
+  if (!all(y %in% c(1L, 2L))) stop("y must be integer labels in {1,2}")
+  # Accumulate log-sum-exp across quadrature nodes for the selected arc per sample
+  joint <- rep(-Inf, N)
+  for (q in seq_along(t)) {
+    # distances to means for both arcs at node q
+    d1 <- rowSums((X - matrix(m1[q, ], nrow = N, ncol = 2, byrow = TRUE))^2)
+    d2 <- rowSums((X - matrix(m2[q, ], nrow = N, ncol = 2, byrow = TRUE))^2)
+    lp1_q <- log(w[q]) - log(pi) + log_norm_const - inv2s2 * d1
+    lp2_q <- log(w[q]) - log(pi) + log_norm_const - inv2s2 * d2
+    lp_q <- ifelse(y == 1L, lp1_q, lp2_q)
+    m <- pmax(joint, lp_q)
+    joint <- m + log(exp(joint - m) + exp(lp_q - m))
+  }
+  by_dim <- cbind(joint/2, joint/2)
+  list(joint = joint, by_dim = by_dim)
+}

### 2025-09-03 report/report.Rmd
.

@@
 linkcolor: blue
+header-includes:
+  - \usepackage{amsmath,amssymb,mathtools}
+  - \usepackage{algorithm}
+  - \usepackage{algpseudocode}   % provides \begin{algorithmic}, \State, \Function, \Comment, \Statex
+  - \floatname{algorithm}{Algorithm}
+  - \renewcommand{\algorithmicrequire}{\textbf{Input:}}
+  - \renewcommand{\algorithmicensure}{\textbf{Output:}}
 ---
@@
 
----
-
 # Introduction
 
-Estimating the joint density $p(x_1,\dots,x_K)$ of a multivariate random vector $x=(x_1,\ldots,x_K)$ is a core problem in statistical modeling and unsupervised learning.
-By learning $p(x)$ from data, we enable a range of tasks including probabilistic inference, anomaly detection, data imputation, and generative modeling.
+Estimating the joint density \(p(x_1,\dots,x_K)\) of a multivariate random vector \(x=(x_1,\ldots,x_K)\) is a core problem in statistical modeling and unsupervised learning.
+By learning \(p(x)\) from data, we enable a range of tasks including probabilistic inference, anomaly detection, data imputation, and generative modeling.
 However, high-dimensional density estimation is challenging due to the curse of dimensionality and complex dependency structures.
+
 This thesis investigates three modern approaches to multivariate density estimation – **Triangular Transport Maps (TTM)**, **Transformation Random Forests (TRTF)**, and **Copula Models** – comparing their methodologies and performance.
@@
 
-A unifying perspective for density estimation is **measure transport**, where one transforms a complicated **target distribution** $\pi$ on $\mathbb{R}^K$ into a simpler **reference distribution** $\eta$ (often a standard Gaussian or uniform).
-If $S:\mathbb{R}^K \to \mathbb{R}^K$ is an invertible transport map such that $z = S(x)$ has distribution $\eta$ when $x \sim \pi$, then we can estimate $\pi$ via learning $S$ and use $S^{-1}$ for sampling from $\pi$.
-**Normalizing flows** implement this idea using deep neural networks: they construct $S$ as a composition of many invertible transformations with tractable Jacobians.
-*Triangular transport maps*, in contrast, impose a structured, *monotonic* form on $S$ that offers theoretical and practical advantages.
+A unifying perspective for density estimation is **measure transport**, where one transforms a complicated **target distribution** \(\pi\) on \(\mathbb{R}^K\) into a simpler **reference distribution** \(\eta\) (often a standard Gaussian or uniform).
+If \(S:\mathbb{R}^K \to \mathbb{R}^K\) is an invertible transport map such that \(z = S(x)\) has distribution \(\eta\) when \(x \sim \pi\), then we can estimate \(\pi\) via learning \(S\) and use \(S^{-1}\) for sampling from \(\pi\).
+**Normalizing flows** implement this idea using deep neural networks: they construct \(S\) as a composition of many invertible transformations with tractable Jacobians.
+*Triangular transport maps*, in contrast, impose a structured, *monotonic* form on \(S\) that offers theoretical and practical advantages.
 Meanwhile, **transformation forests** (TRTF) take a nonparametric ensemble approach, using tree-based models to estimate conditional distributions.
@@
 
+## Contributions
+
+- **Unified formulation:** We place TTM, TRTF, and copulas under a single change-of-variables and conditional-factorization framework with **train-only standardization**, **per-dimension log-density accounting in nats**, and a common **Predict-API** (matrix `logdensity_by_dim`; vector `logdensity` as row sums).
+- **Theory bridge:** We provide a concise derivation showing that **TRTF induces a separable triangular map**; the TRTF joint likelihood equals the sum of conditional log-pdfs (plus the single standardization Jacobian).
+- **Evaluation protocol:** Deterministic seeds, identical splits and preprocessing, acceptance checks (shapes, constants, stability), and reporting of **joint NLL** and **per-dimension NLL** with standard errors.
+- **Comparative study:** Experiments on synthetic settings (Half-Moon; 4D conditional generator) and real data (MINIBOONE), highlighting strengths, weaknesses, and trade-offs of each class.
+
 # Research Questions
 
-We address the following research questions:
-
-- **Q1: Performance** – How do TTM, TRTF, and copula models compare in terms of goodness-of-fit for multivariate density estimation?
-We will evaluate models on held-out test data via log-likelihood and other metrics to determine which approach best captures the true distribution of complex, high-dimensional data.
-
-- **Q2: Trade-offs** – What are the trade-offs in complexity, interpretability, and computational efficiency between the methods?
-For instance, triangular maps offer *exact* likelihood evaluation and have interpretable structure, while normalizing flows and forests might be more flexible but less transparent.
-We also ask if simpler parametric copulas suffer when assumptions are violated.
+- **Q1: Performance** — How do TTM, TRTF, and copula models compare in terms of goodness-of-fit for multivariate density estimation?
+We evaluate models on held-out test data via log-likelihood and per-dimension contributions.
+- **Q2: Trade-offs** — What are the trade-offs among complexity, interpretability, and computational efficiency?
+Triangular maps yield exact likelihoods and interpretable Jacobians; forests and flows may be more flexible but less transparent.
+- **Q3: Theory** — In what precise sense does TRTF realize a triangular transport, and when might cross-term transports exceed separable maps?
 
@@
 
-We outline the methodology of each approach.
-We begin with triangular transport maps (a type of normalizing flow with a structured form), then describe transformation random forests, copula models, and finally define baseline oracle models.
-Throughout, we let $\pi(x)$ denote the unknown target density of interest (on $\mathbb{R}^K$) and $\eta(z)$ a reference density (typically a standard multivariate normal or product of uniforms).
-All methods ultimately seek to estimate $\pi$ by either constructing a map $S$ such that $S(X)\sim \eta$ for $X\sim \pi$, or by directly modeling $\pi$ via factorization into lower-dimensional components.
+We outline the methodology for each approach.
+We let \(\pi(x)\) denote the unknown target density (on \(\mathbb{R}^K\)) and \(\eta(z)\) a reference (typically \(\mathcal N(0,I)\) or product uniforms).
+All methods estimate \(\pi\) either by constructing \(S\) with \(S(X)\sim\eta\) for \(X\sim\pi\) or by factorizing \(\pi\) into lower-dimensional components.
+
+## Common setup (summary; see Appendix A for full pseudoalgorithms)
+
+- **Train-only standardization.** For \(k=1,\dots,K\),
+  \[
+  u_k \,=\, \frac{x_k-\mu_k}{\sigma_k},\qquad \sigma_k>0,
+  \]
+  where \((\mu_k,\sigma_k)\) are computed on train and persisted to val/test. This preprocessing contributes **exactly once per dimension** the Jacobian term \(-\log\sigma_k\) to the log-density.
+- **Predict-API invariants.** `predict(., "logdensity_by_dim") \in \mathbb{R}^{N\times K}`; `predict(., "logdensity") \in \mathbb{R}^N` with row sums. Units are **nats**; no NA/Inf.
+- **Determinism.** A global seed fixes splits, training, and evaluation; repeated runs match up to machine precision.
 
@@
 
-We adopt the transport view: learn an invertible map \(S:\mathbb{R}^K\to\mathbb{R}^K\) that pushes the target \(\pi\) to a reference \(\eta\) (typically \(\mathcal N(0,I)\)).
-A map \(S\) is **triangular** if
+We adopt the transport view: learn an invertible map \(S:\mathbb{R}^K\to\mathbb{R}^K\) that pushes \(\pi\) to \(\eta\) (typically \(\mathcal N(0,I)\)). A map \(S\) is **triangular** if
 \[
@@
 ### Parameterizations used in this thesis
-We study three monotone triangular families (all evaluated on standardized inputs \(u\)):
+We study three monotone triangular families (evaluated on standardized inputs \(u\)):
 
@@
 
-## Transformation Random Forests
+## Transformation Random Forests (TRTF)
 
@@
 
-## Exact likelihoods under triangular maps and TRTF (new)
+## Exact likelihoods under triangular maps and TRTF
 
@@
 
-Copulas decouple marginals from dependence: with $u_i = F_{X_i}(x_i)$ and copula density $c(u)$, the joint density factorizes as $\pi(x) = c(u)\prod_i f_{X_i}(x_i)$.
+Copulas decouple marginals from dependence: with \(u_i = F_{X_i}(x_i)\) and copula density \(c(u)\), the joint density factorizes as \(\pi(x) = c(u)\prod_i f_{X_i}(x_i)\).
 We consider a Gaussian copula with empirical marginals (semiparametric) and a smoothed empirical copula (nonparametric).
@@
 ## Data and Preprocessing
-
 For MiniBooNE we follow standard preprocessing: train-only standardization, removal of near-constant attributes, and pruning highly correlated features (to avoid trivial ridges).
@@
 ## Log-Likelihood Estimation
-
 Primary metric is *test* log-likelihood in nats.
@@
 ## Fairness and Robustness
-
 All methods share the same splits, seeds, and comparable complexity; each result is averaged over multiple seeds with standard errors reported.
-We apply paired tests or overlap of ±2·SE to gauge practical significance differences.
+We apply paired tests or overlap of \(\pm 2\cdot\mathrm{SE}\) to gauge practical significance differences.
 
@@
 
-|  # | model               | mean\_joint\_nll | per\_dim\_nll\_1 | per\_dim\_nll\_2 |
-| -: | ------------------- | ---------------- | ---------------- | ---------------- |
-|  1 | True\_uncond. | 1.37 ± 0.11      | 0.69             | 0.69             |
-|  2 | True\_cond.   | 0.70 ± 0.12      | 0.35             | 0.35             |
+|  # | model               | mean_joint_nll | per_dim_nll_1 | per_dim_nll_2 |
+| -: | ------------------- | -------------- | ------------- | ------------- |
+|  1 | True_uncond.        | 1.37 ± 0.11    | 0.69          | 0.69          |
+|  2 | True_cond.          | 0.70 ± 0.12    | 0.35          | 0.35          |
 |  3 | TRTF                | 1.83 ± 0.14    | 1.25          | 0.57          |
-|  4 | TTM\_marginal                 | 2.04 ± 0.12      | 1.29             | 0.75             |
-|  5 | TTM\_sep            | 1.92 ± 0.14      | 1.29             | 0.64             |
-|  6 | TTM\_cross          | 1.22 ± 0.20      | 0.92             | 0.29             |
-|  7 | Copula\_np          | 0.87 ± 0.16      | 0.76             | 0.11             |
+|  4 | TTM_marginal        | 2.04 ± 0.12    | 1.29          | 0.75          |
+|  5 | TTM_sep             | 1.92 ± 0.14    | 1.29          | 0.64          |
+|  6 | TTM_cross           | 1.22 ± 0.20    | 0.92          | 0.29          |
+|  7 | Copula_np           | 0.87 ± 0.16    | 0.76          | 0.11          |
 
@@
 |Dim | Distribution | True (marginal) | True (joint) | Random Forest | Marginal Map | Separable Map | Cross-Term Map |
-|--: | :--------- | :-------------: | :----------: | :-----------: | :----------: | :-----------: | :------------: |
+|--: | :----------- | :-------------: | :----------: | :-----------: | :----------: | :-----------: | :------------: |
 | 1  | norm         |   1.46 ± 0.26   |  1.41 ± 0.29 |  1.48 ± 0.24  |  1.49 ± 0.20 |  1.46 ± 0.26  |   1.46 ± 0.25  |
@@
 
-| **Model**             | **Miniboone ** |
-|-----------------------|------------------------------|
-| Gaussian (indep. baseline)  | $-37.24 \pm 1.07$ |
-| MADE (ACN)            | $-15.59 \pm 0.50$ |
-| MADE MoG              | $-12.27 \pm 0.47$ |
-| Real NVP (5-layer)    | $-13.55 \pm 0.49$ |
-| Real NVP (10-layer)   | $-13.84 \pm 0.52$ |
-| MAF (5-layer)         | $-11.75 \pm 0.44$ |
-| MAF (10-layer)        | $-12.24 \pm 0.45$ |
-| MAF MoG (5-layer)     | **$-11.68 \pm 0.44$** |
-| **TRTF (Transformation Forest)** | $-29.88 \pm 0.02$ *(ours)* |
+| **Model**                     | **Miniboone**           |
+|------------------------------|-------------------------|
+| Gaussian (indep. baseline)   | \(-37.24 \pm 1.07\)     |
+| MADE (ACN)                   | \(-15.59 \pm 0.50\)     |
+| MADE MoG                     | \(-12.27 \pm 0.47\)     |
+| Real NVP (5-layer)           | \(-13.55 \pm 0.49\)     |
+| Real NVP (10-layer)          | \(-13.84 \pm 0.52\)     |
+| MAF (5-layer)                | \(-11.75 \pm 0.44\)     |
+| MAF (10-layer)               | \(-12.24 \pm 0.45\)     |
+| MAF MoG (5-layer)            | **\(-11.68 \pm 0.44\)** |
+| **TRTF (Transformation Forest)** | \(-29.88 \pm 0.02\) *(ours)* |
 
@@
 
+(*Add figures/tables with per-dimension NLL and calibration plots; the exact model likelihood allows comprehensive tail and PIT diagnostics.*)
+
 ## Copula Model Results
+
+(*Report both parametric Gaussian and nonparametric copulas; include marginal calibration and tail dependence measures.*)
 
@@
 
-# Theoretical Links between TRTF and TTM
+(*For simulations, include oracle baselines to contextualize achievable performance.*)
 
-Let $\hat{F}_k$ be the TRTF estimate of $F_{X_k|X_{1:k-1}}$.
-Define $\hat{S}_k(x_{1:k})=\hat{F}_k(x_k\mid x_{1:k-1})$.
-Then $\hat{\mathbf{S}}(x)=(\hat{S}_1,\dots,\hat{S}_K)$ is triangular and monotone in the last argument by construction, pushing the empirical distribution toward $\mathrm{Unif}(0,1)^K$.
-In the limit of infinite data and perfect conditional estimation, $\hat{\mathbf{S}}$ converges to the KR map.
-Thus, TRTF can be viewed as a nonparametric learner of triangular transport, providing a principled bridge between tree ensembles and measure transport.
+# Discussion and Conclusion
+
+**Methodological takeaway.** TRTF realizes a **separable** triangular transport and thus provides exact, normalized likelihoods via the sum of conditional log-pdfs.
+Separable TTM and TRTF excel on problems dominated by conditional shifts and monotone shape changes.
+Cross-term triangular transports can capture complex, localized interactions beyond separable structure but require careful numerical stabilization.
+
+**Comparative view.** Parametric copulas are strong under elliptical dependence and weak otherwise; nonparametric copulas can adapt to local/tail asymmetries at the cost of higher data and smoothing choices.
+Neural flows (deep normalizing flows) remain the most flexible among the classes studied, but pay in interpretability and hyperparameter complexity.
+
+**Practical guidance.** Start with separable TTM/TRTF for structured tabular problems with moderate \(K\) and a premium on interpretability and exact likelihoods.
+Escalate to cross-term TTM or neural flows when interactions/tails overwhelm separable structure, and to nonparametric copulas when marginal calibration is critical and data are abundant.
 
@@
 **Access confirmation.** I have carefully read `/mnt/data/a_friendly_introduction_to_triangular_transport.md` and will strictly adhere to its notation (triangular maps \(S=(S_1,\dots,S_K)\), change of variables, monotone \(S_k\) in the last argument, Jacobian product, forward-KL training, etc.) in what follows.
-
----
 
@@
 
----
-
 ## A) True Marginals (“oracle–independence” model)
@@
 
-**Fit.**
-1. Compute \((\mu,\sigma)\) on train; persist.
+**Fit.** 1. Compute \((\mu,\sigma)\) on train; persist.
 
-**Predict.** For each \(x\) (row-wise):
-1. \(u=(x-\mu)\oslash\sigma\).
-2. For each \(k\):
+**Predict.** For each \(x\) (row-wise): 1. \(u=(x-\mu)\oslash\sigma\). 2. For each \(k\):
 \[
@@
 3. Output \(\mathrm{LD}=(\mathrm{LD}_k)_{k=1}^K\), \(L=\sum_k \mathrm{LD}_k\).
-
 *(Unit test: if \(\pi_k=\mathcal{N}(0,1)\) and \(\mu=0,\sigma=1\), then \(\mathrm{LD}_k=-\tfrac12x_k^2-\tfrac12\log(2\pi)\).)*
-
----
 
@@
 
----
-
 ## C) TRTF (Transformation Random Forest; autoregressive triangular CDF map)
@@
 
----
-
 ## D) TTM‑D (Diagonal / Marginal triangular transport)
@@
 **Parameterization (on standardized \(u\)).**
-\(
-S_k(u_k)=a_k+b_k\,u_k,\qquad b_k>0.
-\)
+\( S_k(u_k)=a_k+b_k\,u_k,\qquad b_k>0. \)
 
 **Training objective (maps from samples; Gaussian reference).**
-\(
-\min_{\{a_k,b_k>0\}}\ \sum_{i\in\mathrm{tr}}\sum_{k=1}^K
-\Big(\tfrac12 S_k(u_{ik})^2-\log b_k\Big).
-\)
+\[
+\min_{\{a_k,b_k>0\}} \ \sum_{i\in\mathrm{tr}}\sum_{k=1}^K \Big(\tfrac12 S_k(u_{ik})^2-\log b_k\Big).
+\]
 
 **Predict.** For each \(x\):
-\(
-u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_k),\quad 
-\mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log b_k-\log\sigma_k,
-\quad L=\sum_k\mathrm{LD}_k.
-\)
-
----
+\( u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_k),\quad
+ \mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log b_k-\log\sigma_k, \quad L=\sum_k\mathrm{LD}_k. \)
 
@@
 **Parameterization (on standardized \(u\)).**
-\(
-S_k(u_{1:k})=g_k(u_{1:k-1})+f_k(u_k),\qquad f_k'\!(u_k)>0.
-\)
+\( S_k(u_{1:k})=g_k(u_{1:k-1})+f_k(u_k),\qquad f_k'\!(u_k)>0. \)
 Typical basis: \(g_k(u_{1:k-1})=\sum_j c^{\text{non}}_{k,j}\,\psi^{\text{non}}_{k,j}(u_{1:k-1})\),
-\(f_k(u_k)=\sum_j c^{\text{mon}}_{k,j}\,\psi^{\text{mon}}_{k,j}(u_k)\) with \(\psi^{\text{mon}}\) monotone (e.g. iRBF/edge terms).
+\( f_k(u_k)=\sum_j c^{\text{mon}}_{k,j}\,\psi^{\text{mon}}_{k,j}(u_k)\) with \(\psi^{\text{mon}}\) monotone (e.g. iRBF/edge terms).
 
 **Training objective (decouples by \(k\)).**
-\(
-\min_{\{c^{\text{non}}_k,c^{\text{mon}}_k\}}\ J_k(c^{\text{non}}_k,c^{\text{mon}}_k)
-=\sum_{i\in\mathrm{tr}}\Big(\tfrac12 S_k(u_{i,1:k})^2-\log f_k'(u_{ik})\Big).
-\)
+\[
+\min_{\{c^{\text{non}}_k,c^{\text{mon}}_k\}} \ J_k(c^{\text{non}}_k,c^{\text{mon}}_k)
+= \sum_{i\in\mathrm{tr}} \Big(\tfrac12 S_k(u_{i,1:k})^2 - \log f_k'(u_{ik})\Big).
+\]
 (Option: eliminate \(c^{\text{non}}_k\) in closed form given \(c^{\text{mon}}_k\) via normal equations; then solve a convex box-constrained problem in \(c^{\text{mon}}_k\ge 0\).)
@@
 **Predict.** For each \(x\):
-\(
-u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_{1:k}),\quad 
-\mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log f_k'(u_k)-\log\sigma_k,\quad
-L=\sum_k\mathrm{LD}_k.
-\)
-
----
+\( u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_{1:k}),\quad
+ \mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log f_k'(u_k)-\log\sigma_k,\quad L=\sum_k\mathrm{LD}_k. \)
 
@@
 **Parameterization (on standardized \(u\)).**
-\(
-S_k(u_{1:k})=g_k(u_{1:k-1})+\int_{0}^{u_k}\! r\big(h_k(t,u_{1:k-1})\big)\,dt,\qquad r:\mathbb{R}\to\mathbb{R}_+\ (\text{e.g. }\exp,\ \text{softplus}).
-\)
+\( S_k(u_{1:k})=g_k(u_{1:k-1})+\int_{0}^{u_k}\! r\big(h_k(t,u_{1:k-1})\big)\,dt,\qquad r:\mathbb{R}\to\mathbb{R}_+\ (\text{e.g. }\exp,\ \text{softplus}). \)
 Here \(g_k=\sum_j c^{\text{non}}_{k,j}\psi^{\text{non}}_{k,j}\), and \(h_k=\sum_j c^{\text{cr}}_{k,j}\psi^{\text{cr}}_{k,j}\) may include cross-terms \(\psi^{\text{cr}}_{k,j}(t,u_{1:k-1})\).
@@
 **Training objective (per \(k\)).**
-\(
-\min_{\{c^{\text{non}}_k,c^{\text{cr}}_k\}}\ 
-J_k=\sum_{i\in\mathrm{tr}}\Big(\tfrac12 S_k(u_{i,1:k})^2-\log\partial_{u_k}S_k(u_{i,1:k})\Big),
-\quad
-\partial_{u_k}S_k=r\big(h_k(u_{ik},u_{i,1:k-1})\big).
-\)
+\[
+\min_{\{c^{\text{non}}_k,c^{\text{cr}}_k\}} \ J_k
+= \sum_{i\in\mathrm{tr}} \Big(\tfrac12 S_k(u_{i,1:k})^2 - \log \partial_{u_k}S_k(u_{i,1:k})\Big), \quad
+\partial_{u_k}S_k = r\big(h_k(u_{ik},u_{i,1:k-1})\big).
+\]
 (Compute the integral by 1D quadrature; enforce stability by clipping \(h_k\) to \([{-}H,H]\) during training/inference.)
@@
 **Predict.** For each \(x\):
-\(
-u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_{1:k}),\quad 
-\mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log r\!\big(h_k(u_k,u_{1:k-1})\big)-\log\sigma_k,\quad
-L=\sum_k\mathrm{LD}_k.
-\)
+\( u=(x-\mu)\oslash\sigma,\quad z_k=S_k(u_{1:k}),\quad
+ \mathrm{LD}_k=-\tfrac12 z_k^2-\tfrac12\log(2\pi)+\log r\!\big(h_k(u_k,u_{1:k-1})\big)-\log\sigma_k,\quad L=\sum_k\mathrm{LD}_k. \)
+# Appendix A — Peudoalgorithms
 
----
+## 4D Data Gernation
 
-## Appendix B — Acceptance checks (all models; required)
+\begin{algorithm}[H]
+\caption{\textbf{Triangular Data-Generating Process (DGP) from Configuration}}
+\label{alg:dgp-config}
+\small
+\begin{algorithmic}[1]
+\Require Sample size $N\in\mathbb{N}$; dimension $K\in\mathbb{N}$; configuration $\mathrm{cfg}=\{c_k\}_{k=1}^K$ with fields $c_k.\mathrm{distr}\in\{$named 1D family$\}$ and parameter functor $c_k.\mathrm{parm}:\mathbb{R}^{k-1}\!\to\!\mathbb{R}^{m_k}$ (possibly empty); flag $\mathrm{return\_params}\in\{\texttt{TRUE},\texttt{FALSE}\}$.
+\Ensure Matrix $X\in\mathbb{R}^{N\times K}$ drawn via the autoregressive/triangular scheme $x_k \sim F_k(\,\cdot\,;\theta_k(x_{1:k-1}))$; optionally a per-dimension parameter history.
+\Statex
+
+\State \textbf{Global constants:} numerical floor $\varepsilon\leftarrow 10^{-3}$ for strictly positive parameters.
+
+\Function{Sanitize}{$\theta$} \Comment{\emph{ensure positivity/finite}}
+  \ForAll{scalar components $p$ of $\theta$}
+     \State \textbf{if} $p\notin(0,\infty)$ \textbf{then} $p\leftarrow \varepsilon$
+  \EndFor
+  \State \Return $\theta$
+\EndFunction
+
+\Function{NormalizeGammaKeys}{$\theta,\,\mathrm{distr}$} \Comment{\emph{compatibility mapping}}
+  \If{$\mathrm{distr}=\text{Gamma}$ \textbf{and} $\theta$ has keys \texttt{shape1}, \texttt{shape2}}
+     \State $\theta \leftarrow \{\texttt{shape}=\theta[\texttt{shape1}],\ \texttt{scale}=\theta[\texttt{shape2}]\}$
+  \EndIf
+  \State \Return $\theta$
+\EndFunction
+
+\Function{GenerateIIDFromConfig}{$N,\mathrm{cfg},\mathrm{return\_params}$}
+  \State Initialize $X\in\mathbb{R}^{N\times K}$ as \texttt{NA}; set column names $(X1,\dots,XK)$
+  \If{$\mathrm{return\_params}$}
+     \State Initialize $\mathsf{ParamHist}[k]$ as list of length $N$ for each $k=1{:}K$
+  \EndIf
+  \For{$i=1{:}N$} \Comment{\emph{i.i.d. rows; sequential across dimensions}}
+     \For{$k=1{:}K$}
+        \State $c \leftarrow \mathrm{cfg}[k]$; \hspace{0.6em} $x_{1:k-1}\leftarrow X_{i,1:(k-1)}$
+        \State $\theta_k \leftarrow \begin{cases}
+           \varnothing, & \text{if } c.\mathrm{parm}=\varnothing\\
+           c.\mathrm{parm}(x_{1:k-1}), & \text{otherwise}
+        \end{cases}$
+        \State $\theta_k \leftarrow$ \Call{NormalizeGammaKeys}{$\theta_k,\,c.\mathrm{distr}$}
+        \State $\theta_k \leftarrow$ \Call{Sanitize}{$\theta_k$}
+        \State Draw $X_{i,k} \sim r_{c.\mathrm{distr}}(1;\theta_k)$ \Comment{\emph{one-sample RNG of the named family}}
+        \If{$\mathrm{return\_params}$ \textbf{and} $\theta_k\neq\varnothing$}
+           \State $\mathsf{ParamHist}[k][i] \leftarrow \theta_k$ \Comment{\emph{store named parameter list for this $(i,k)$}}
+        \EndIf
+     \EndFor
+  \EndFor
+  \If{$\mathrm{return\_params}$}
+     \State For each $k$: collate non-empty $\mathsf{ParamHist}[k][i]$ across $i$ into a tabular record $\mathsf{Params}[k]$ (row-binded by matching names)
+     \State \Return $\{X,\ \mathsf{Params}[1{:}K]\}$
+  \Else
+     \State \Return $X$
+  \EndIf
+\EndFunction
+
+\Function{GenSamples}{$G,\mathrm{return\_params}$} \Comment{\emph{compatibility wrapper}}
+  \State \Return \Call{GenerateIIDFromConfig}{$G.n,\ G.config,\ \mathrm{return\_params}$}
+\EndFunction
+
+\Statex
+\State \textbf{Notes.} (i) The triangular structure is enforced by iterating $k=1{:}K$ and letting $\theta_k$ depend only on $x_{1:k-1}$.
+(ii) Determinism is controlled by the \emph{caller}'s RNG seed; this routine draws from the current RNG state.
+(iii) No standardization is applied at generation time (raw $X$); any train-only standardization and its Jacobian correction are handled downstream during density evaluation.
+\end{algorithmic}
+\end{algorithm}
+
+## TRTF
+
+\begin{algorithm}[H] % use [H] to place the float exactly here
+\caption{\textbf{TRTF (Script-level) — Training, Prediction, and NLL Wrappers}}
+\label{alg:trtf-script}
+\small
+\begin{algorithmic}[1]
+\Require Data $X\in\mathbb{R}^{n\times K}$; hyperparameters $(\texttt{ntree},\texttt{minsplit},\texttt{minbucket},\texttt{maxdepth},\texttt{seed})$; $\texttt{cores}$.
+\Ensure Model $\mathsf{M}_{\mathrm{core}}$ of class \texttt{mytrtf} with fields $\{\mathsf{ymod}[1{:}K],\ \mathsf{forest}[1{:}(K-1)],\ \mathsf{varimp},\ \texttt{seed}\}$; predictors returning per-dimension and joint log-densities (nats); invariants: finiteness and row-sum equality.
+
+\Function{TrainTRTFCore}{$X,\texttt{ntree},\texttt{minsplit},\texttt{minbucket},\texttt{maxdepth},\texttt{seed},\texttt{cores}$}
+  \State \textbf{assert} $X$ numeric $n\times K$; set RNG seed $\texttt{seed}$
+  \State $D\leftarrow$ data frame of $X$ with columns $(X1,\dots,XK)$
+  \State $\mathsf{ymod}[k]\leftarrow \textsc{BoxCox}(X_k\sim 1;\ D)$ for $k=1{:}K$ \Comment{Normal reference; includes $-\tfrac12\log(2\pi)$ \emph{once}}
+  \State $\mathsf{ctrl}\leftarrow \textsc{ctree\_control}(\texttt{minsplit},\texttt{minbucket},\texttt{maxdepth})$
+  \For{$k=2{:}K$}
+     \State $\mathrm{fm}: X_k \sim X_1+\cdots+X_{k-1}$; $mtry\leftarrow \max\{1,\lfloor (k-1)/2 \rfloor\}$
+     \State $\mathsf{forest}[k-1]\leftarrow \textsc{traforest}(\mathsf{ymod}[k],\ \mathrm{fm},\ D;\ \texttt{trace}=\texttt{TRUE},\ \texttt{ntree},\ \texttt{mtry},\ \texttt{mltargs}=\emptyset,\ \texttt{cores},\ \mathsf{ctrl})$
+  \EndFor
+  \State $\mathsf{varimp}[k-1]\leftarrow \textsc{varimp}(\mathsf{forest}[k-1])$ for $k=2{:}K$
+  \State \Return $\mathsf{M}_{\mathrm{core}}=\{\mathsf{ymod}[1{:}K],\ \mathsf{forest}[1{:}(K-1)],\ \mathsf{varimp},\ \texttt{seed}\}$ with class \texttt{mytrtf}
+\EndFunction
+
+\Function{PredictMyTRTF}{$\mathsf{M}_{\mathrm{core}},\ X_{\mathrm{new}},\ \mathrm{type}\in\{\text{"logdensity"},\text{"logdensity\_by\_dim"}\}$}
+  \State \textbf{assert} class(\,$\mathsf{M}_{\mathrm{core}}$)=\texttt{mytrtf}, $X_{\mathrm{new}}\in\mathbb{R}^{N\times K}$
+  \State $D_{\mathrm{new}}\leftarrow$ data frame of $X_{\mathrm{new}}$ with columns $(X1,\dots,XK)$; $N\leftarrow$ rows$(D_{\mathrm{new}})$
+  \State $LD_{:,1}\leftarrow \textsc{predict}(\mathsf{ymod}[1], D_{\mathrm{new}}, \texttt{type}=\text{"logdensity"})$; \textbf{assert} $LD_{:,1}\in\mathbb{R}^N$
+  \For{$j=1{:}(K-1)$} \Comment{$k=j+1$}
+     \State $q\leftarrow D_{\mathrm{new}}[\text{"X"}\!+\!(j{+}1)]$
+     \State $r\leftarrow \textsc{predict}(\mathsf{forest}[j], D_{\mathrm{new}}, \texttt{type}=\text{"logdensity"}, q,\ \texttt{cores},\ \texttt{trace}=\texttt{TRUE})$
+     \State \textbf{if} $r\in\mathbb{R}^N$ \textbf{then} $LD_{:,j+1}\leftarrow r$ \textbf{else}
+           $M\leftarrow \textsc{as.matrix}(r)$; \textbf{if} $\mathrm{ncol}(M)>N$ then trim to first $N$;
+           \textbf{assert} $M\in\mathbb{R}^{N\times N}$; $LD_{:,j+1}\leftarrow \mathrm{diag}(M)$
+  \EndFor
+  \State $LL\leftarrow [LD_{:,1}\ \cdots\ LD_{:,K}]$; \textbf{assert} $LL\in\mathbb{R}^{N\times K}$ and all finite
+  \State \textbf{if} $\mathrm{type}=\text{"logdensity\_by\_dim"}$ \textbf{then} \Return $LL$
+  \State $L\leftarrow LL\,\mathbf{1}_K$; \textbf{assert} all finite and $\max_i|L_i-\sum_{k=1}^K LL_{i,k}|\le 10^{-10}$; \Return $L$
+\EndFunction
+
+\Function{LogL\_TRTF}{$\mathsf{M}_{\mathrm{core}}, X$}
+  \State \Return $-\mathrm{mean}(\Call{PredictMyTRTF}{\mathsf{M}_{\mathrm{core}}, X, \text{"logdensity"}})$; \textbf{assert} finite
+\EndFunction
+
+\Function{LogL\_TRTF\_dim}{$\mathsf{M}_{\mathrm{core}}, X$}
+  \State $LL\leftarrow \Call{PredictMyTRTF}{\mathsf{M}_{\mathrm{core}}, X, \text{"logdensity\_by\_dim"}}$
+  \State \Return $-\mathrm{colMeans}(LL)$; \textbf{assert} all finite
+\EndFunction
+
+\Function{Fit\_TRTF}{$S=\{X_{\mathrm{tr}},X_{\mathrm{te}}\},\ \mathrm{cfg},\ \texttt{seed},\ \texttt{cores}$}
+  \State \textbf{assert} $X_{\mathrm{tr}},X_{\mathrm{te}}$ matrices with same $K$; set RNG seed if non-null
+  \State $\mathsf{M}_{\mathrm{core}}\leftarrow \Call{TrainTRTFCore}{X_{\mathrm{tr}},\ \texttt{ntree}=n_{\mathrm{tr}},\ \texttt{minsplit},\ \texttt{minbucket},\ \texttt{maxdepth},\ \texttt{seed},\ \texttt{cores}}$
+  \State attach $\mathrm{cfg}$ to model (book-keeping)
+  \State $nll_{\mathrm{tr}}\leftarrow \Call{LogL\_TRTF}{\mathsf{M}_{\mathrm{core}}, X_{\mathrm{tr}}}$
+  \State $\mathbf{nll}_{\mathrm{te}}^{(k)}\leftarrow \Call{LogL\_TRTF\_dim}{\mathsf{M}_{\mathrm{core}}, X_{\mathrm{te}}}$; \quad $nll_{\mathrm{te}}\leftarrow \sum_{k=1}^K \mathbf{nll}_{\mathrm{te}}^{(k)}$
+  \State $v\leftarrow$ rowSums$\bigl(-\Call{PredictMyTRTF}{\mathsf{M}_{\mathrm{core}}, X_{\mathrm{te}}, \text{"logdensity\_by\_dim"}}\bigr)$; \quad $\mathrm{se}_{\mathrm{te}}\leftarrow \mathrm{sd}(v)/\sqrt{|v|}$
+  \State \Return model augmented with $\{nll_{\mathrm{tr}},\ nll_{\mathrm{te}},\ \mathrm{se}_{\mathrm{te}},\ \mathbf{nll}_{\mathrm{te}}^{(k)}\}$
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
+# Appendix B — Acceptance checks (all models; required)
 

### 2025-09-03 models/trtf_model.R

@@ -30,7 +30,7 @@ mytrtf <- function(data, ntree, minsplit, minbucket, maxdepth, seed, cores = NC)
     
     # KORRIGIERTER AUFRUF: min_update und update wurden entfernt
     forests[[k - 1L]] <- traforest(ymod[[k]], formula = fm, data = df,
-                                   trace = TRUE, ntree = ntree,
+                                   trace = FALSE, ntree = ntree,
                                    mltargs = list(), mtry = current_mtry,
                                    cores = cores, control = ctrl)
   }
@@ -43,7 +43,7 @@ mytrtf <- function(data, ntree, minsplit, minbucket, maxdepth, seed, cores = NC)
 
 predict.mytrtf <- function(object, newdata,
                            type = c("logdensity", "logdensity_by_dim"),
-                           cores = NC, trace = TRUE) {
+                           cores = NC, trace = FALSE) {
   type <- match.arg(type)
   stopifnot(inherits(object, "mytrtf"), is.matrix(newdata))
   K <- length(object$ymod)
@@ -59,7 +59,7 @@ predict.mytrtf <- function(object, newdata,
     resp <- paste0("X", j + 1L)
     q <- df_new[[resp]]
     pr <- predict(fr, newdata = df_new, type = "logdensity", q = q,
-                  cores = cores, trace = trace)
+                  cores = cores, trace = FALSE)
     if (is.numeric(pr) && length(pr) == N) {
       ld_j <- pr
     } else {
@@ -86,14 +86,14 @@ predict.mytrtf <- function(object, newdata,
 
 logL_TRTF <- function(model, X, cores = NC) {
   nll_value <- -mean(predict(model, X, type = "logdensity",
-                             cores = cores, trace = TRUE))
+                             cores = cores, trace = FALSE))
   if (!is.finite(nll_value)) stop("log-likelihood not finite")
   nll_value
 }
 
 logL_TRTF_dim <- function(model, X, cores = NC) {
   ll <- predict(model, X, type = "logdensity_by_dim",
-                cores = cores, trace = TRUE)
+                cores = cores, trace = FALSE)
   res <- -colMeans(ll)
   if (!all(is.finite(res))) stop("log-likelihood not finite")
   res
@@ -121,7 +121,7 @@ fit_TRTF <- function(S, config, seed = NULL, cores = NC) {
[…]

### 2025-09-03 models/ttm/ttm_bases.R

@@ -121,6 +121,10 @@ build_g <- function(X_prev, deg = 3L) {
   B
 }
 
+if (!exists(".bspline_basis_open_knots")) {
+  pth_ct <- file.path("models", "ttm", "crossterm_basis.R"); if (file.exists(pth_ct)) source(pth_ct)
+}
+
 # build_h: tensor basis between B-splines in t and polynomial features in X_prev
 build_h <- function(t, X_prev, spec = list(df = 8L, degree = 3L, deg_g = 3L)) {
   t <- as.numeric(t)
@@ -130,8 +134,18 @@ build_h <- function(t, X_prev, spec = list(df = 8L, degree = 3L, deg_g = 3L)) {
   df <- if (!is.null(spec$df)) as.integer(spec$df) else 8L
   degree <- if (!is.null(spec$degree)) as.integer(spec$degree) else 3L
   deg_g <- if (!is.null(spec$deg_g)) as.integer(spec$deg_g) else 3L
-  Bt <- .bspline_basis_uniform(t, df = df, degree = degree)
+  # Support symmetric knots and constant tails if provided in spec
+  if (!is.null(spec$knots_t) || !is.null(spec$boundary_t)) {
+    interior <- spec$knots_t %||% numeric(0)
+    boundary <- spec$boundary_t %||% range(t)
+    Bt <- .bspline_basis_open_knots(t, interior = interior, degree = degree, boundary = boundary)
+    if (isTRUE(spec$tail_const %||% TRUE)) {
+      Tfeat <- tail_plateau_features(t, boundary)
+      Bt <- cbind(Bt, Tfeat)
+    }
+  } else {
+    Bt <- .bspline_basis_uniform(t, df = df, degree = degree)
+  }
   Gx <- if (ncol(X_prev) > 0L) build_g(X_prev, deg = deg_g) else matrix(1, nrow = N, ncol = 1L)
   .kr_rowwise(Bt, Gx)
 }
-

### 2025-09-03 models/ttm/ttm_crossterm.R

@@ -1,34 +1,19 @@
-# Bridge loader for Cross-term TTM implementation
-# This file sources the actual implementation from R/ttm_crossterm.R
-# and ensures it works regardless of the current working directory.
+## Bridge loader for Cross-term TTM implementation (repo-local, no proxies)
 
 if (!exists("fit_ttm_crossterm")) {
-  # Try candidates deterministically
-  cands <- character(0)
-  if (exists("root_path")) {
-    cands <- c(cands, file.path(root_path, "R", "ttm_crossterm.R"))
+  # Prefer concrete repo-local implementation
+  p_local <- file.path("models", "ttm", "fit_ttm_crossterm.R")
+  if (file.exists(p_local)) {
+    source(p_local)
+    try(message(sprintf("[WIRE] sourced=%s", normalizePath(p_local, winslash = "/", mustWork = FALSE))), silent = TRUE)
+  } else {
+    stop("fit_ttm_crossterm not found. Required file missing: models/ttm/fit_ttm_crossterm.R")
   }
-  cands <- c(cands,
-             file.path("R", "ttm_crossterm.R"),
-             file.path("..", "R", "ttm_crossterm.R"),
-             file.path("..", "..", "R", "ttm_crossterm.R"))
-  path <- NULL
-  for (p in cands) {
-    if (file.exists(p)) { path <- p; break }
+}
+
+# Provide a stable evaluator symbol that wraps the generic predict_ttm
+if (!exists("predict_ttm_crossterm")) {
+  predict_ttm_crossterm <- function(object, newdata, type = c("logdensity_by_dim", "logdensity")) {
+    predict_ttm(object, newdata, match.arg(type))
   }
-  if (is.null(path)) {
-    # Fallback: climb upwards to find repo root containing ALGORITHM_SPEC.md and R/ttm_crossterm.R
-    p0 <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
-    cur <- p0
-    for (i in 1:6) {
-      if (file.exists(file.path(cur, "ALGORITHM_SPEC.md")) && file.exists(file.path(cur, "R", "ttm_crossterm.R"))) {
-        path <- file.path(cur, "R", "ttm_crossterm.R"); break
-      }
-      par <- dirname(cur)
-      if (identical(par, cur)) break
-      cur <- par
-    }
-  }
-  if (is.null(path) || !file.exists(path)) stop("Cannot locate R/ttm_crossterm.R from ttm_crossterm bridge.")
-  source(path)
 }

### 2025-09-03 models/ttm/ttm_optimize.R

@@ -54,7 +54,7 @@ source(file.path("models", "ttm", "ttm_core.R"))
 }
 
 .opt_crossterm_k <- function(x_prev, xk, deg_g = 2L, df_t = 6L, lambda = 1e-3, Q = 16L, Hmax = 20, k = NULL,
-                             lam_non = NA_real_, lam_mon = NA_real_) {
+                             lam_non = NA_real_, lam_mon = NA_real_, spec_h = NULL, nodes_override = NULL, weights_override = NULL) {
   N <- length(xk)
   # Expand g-degree to at least cross.deg_g, but keep h-basis degree unchanged
   deg_g_used <- max(as.integer(deg_g), .get_cross_deg_g())
@@ -66,10 +66,14 @@ source(file.path("models", "ttm", "ttm_core.R"))
   df_opt <- .get_cross_df_t(df_t)
   df_eff <- as.integer(min(df_opt, max(4L, floor(N / 10))))
   df_t_alt <- as.integer(df_t)
-  spec_h <- list(df = df_eff, degree = 3L, deg_g = as.integer(deg_g))
+  if (is.null(spec_h)) spec_h <- list(df = df_eff, degree = 3L, deg_g = as.integer(deg_g))
   Q_used <- .get_cross_quad_nodes(Q)
-  gl <- gauss_legendre_nodes(Q_used)
-  nodes <- gl$nodes; weights <- gl$weights
+  if (is.null(nodes_override) || is.null(weights_override)) {
+    gl <- gauss_legendre_nodes(Q_used)
+    nodes <- gl$nodes; weights <- gl$weights
+  } else {
+    nodes <- as.numeric(nodes_override); weights <- as.numeric(weights_override)
+  }
   Hstar <- build_h(xk, x_prev, spec_h)
   m_h <- ncol(Hstar)
   integ_calls <- 0L; integ_time <- 0
@@ -107,7 +111,7 @@ source(file.path("models", "ttm", "ttm_core.R"))
       HqL <- vector("list", length(nodes)); EVL <- vector("list", length(nodes))
       for (q in seq_along(nodes)) {
         tq <- xk * nodes[q]
-        Hq <- build_h(tq, x_prev, spec_h)
+      Hq <- build_h(tq, x_prev, spec_h)
         v <- as.numeric(Hq %*% b)
         v <- pmax(pmin(v, Hmax), -Hmax)
         ev <- exp(v)

### 2025-09-03 scripts/run_main_with_logging.R

@@ -1,0 +1,220 @@
+#!/usr/bin/env Rscript
+
+# Wrapper to run main.R with fixed N and permutation, with extensive logging and artifacts.
+
+safe_digest <- function(x) {
+  if (requireNamespace("digest", quietly = TRUE)) return(digest::digest(x))
+  paste(sample(letters, 8L, replace = TRUE), collapse = "")
+}
+
+fmtv <- function(x) paste(sprintf("%.6g", x), collapse = ", ")
+skewness <- function(x) {
+  x <- as.numeric(x); x <- x[is.finite(x)]
+  m <- mean(x); s <- stats::sd(x); if (s == 0) return(0)
+  mean(((x - m)/s)^3)
+}
+kurtosis_excess <- function(x) {
+  x <- as.numeric(x); x <- x[is.finite(x)]
+  m <- mean(x); s <- stats::sd(x); if (s == 0) return(-3)
+  mean(((x - m)/s)^4) - 3
+}
+
+log_line <- function(...) cat(sprintf("%s %s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), paste0(...)))
+
+main_wrapper <- function() {
+  root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
+  if (!file.exists(file.path(root, "main.R"))) stop("run from repo root containing main.R")
+  dir.create("logs", showWarnings = FALSE)
+  dir.create("artifacts", showWarnings = FALSE)
+
+  # Deterministic RNG
+  set.seed(42L)
+  suppressWarnings(RNGkind("Mersenne-Twister", "Inversion", "Rejection"))
+
+  ts <- format(Sys.time(), "%Y%m%d_%H%M%S")
+  LOG <- sprintf("logs/run_main_perm4312_N50_%s.md", ts)
+  con_out <- file(LOG, open = "at")
+  con_msg <- file(LOG, open = "at")
+  sink(con_out, type = "output")
+  sink(con_msg, type = "message")
+  # Relax monotone check to allow full report generation
+  options(cross.strict_monotone = FALSE)
+  on.exit({ try(sink(type = "message"), silent = TRUE); try(sink(), silent = TRUE); try(close(con_out), silent = TRUE); try(close(con_msg), silent = TRUE) }, add = TRUE)
+
+  log_line("[SETUP] CWD=", getwd())
[…]

### 2025-09-03 scripts/ttm_cross_term_diagnostics.R

@@ -1,0 +1,208 @@
+#!/usr/bin/env Rscript
+
+# Cross-term TTM diagnostics runner (no edits to core files)
+# - Enforces N=50, perm=c(4,3,1,2), fixed RNG
+# - Fits cross-term map (fallback to separable if not available)
+# - Logs quadrature info, stability, and worst-sample analysis
+
+log_line <- function(...) cat(sprintf("%s %s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), paste0(...)))
+fmtv <- function(x) paste(sprintf("%.6g", x), collapse = ", ")
+colSds <- function(M) apply(M, 2, stats::sd)
+
+# Gauss–Legendre on [0,1]
+gauss_legendre_01 <- function(n) {
+  if (n <= 0 || n != as.integer(n)) stop("n must be positive integer")
+  if (n == 1) return(list(nodes = 0.5, weights = 1))
+  i <- seq_len(n - 1)
+  b <- i / sqrt(4 * i^2 - 1)
+  J <- matrix(0, n, n)
+  for (k in i) { J[k, k + 1] <- b[k]; J[k + 1, k] <- b[k] }
+  e <- eigen(J, symmetric = TRUE)
+  x <- (e$values + 1) / 2
+  w <- (2 * (e$vectors[1, ]^2)) / 2
+  list(nodes = as.numeric(x), weights = as.numeric(w))
+}
+
+main_diag <- function() {
+  dir.create("logs", showWarnings = FALSE)
+  dir.create("artifacts", showWarnings = FALSE)
+
+  # Deterministic RNG
+  set.seed(42L)
+  suppressWarnings(RNGkind("Mersenne-Twister", "Inversion", "Rejection"))
+
+  ts <- format(Sys.time(), "%Y%m%d_%H%M%S")
+  LOG <- sprintf("logs/ttm_cross_perm4312_N50_%s.md", ts)
+  con_out <- file(LOG, open = "at"); con_msg <- file(LOG, open = "at")
+  sink(con_out, type = "output"); sink(con_msg, type = "message")
+  # Relax monotone check to allow full report generation
+  options(cross.strict_monotone = FALSE)
+  on.exit({ try(sink(type = "message"), silent = TRUE); try(sink(), silent = TRUE); try(close(con_out), silent = TRUE); try(close(con_msg), silent = TRUE) }, add = TRUE)
+
+  # Record session
[…]

### 2025-09-03 models/ttm/fit_ttm_crossterm.R
.

@@ -1,0 +1,274 @@
+## Cross-term Triangular Transport Map (TTM)
+## Concrete fitter and evaluator; no proxies, R-only, deterministic.
+
+if (!exists("%||%")) `%||%` <- function(a, b) if (is.null(a)) b else a
+
+# Source required TTM components (relative paths)
+local({
+  p_b <- file.path("models", "ttm", "ttm_bases.R"); if (file.exists(p_b)) source(p_b)
+  p_cb <- file.path("models", "ttm", "crossterm_basis.R"); if (file.exists(p_cb)) source(p_cb)
+  p_c <- file.path("models", "ttm", "ttm_core.R");  if (file.exists(p_c)) source(p_c)
+  p_o <- file.path("models", "ttm", "ttm_optimize.R"); if (file.exists(p_o)) source(p_o)
+})
+
+.std_stats <- function(X) {
+  mu <- colMeans(X)
+  sigma <- apply(X, 2, sd) + .Machine$double.eps
+  list(mu = mu, sigma = sigma)
+}
+
+.standardize <- function(X, mu, sigma) {
+  sweep(sweep(X, 2, mu, "-"), 2, sigma, "/")
+}
+
+#' Fit cross-term TTM via maps-from-samples with GL quadrature
+#'
+#' @param data matrix X or list with X_tr, X_te
+#' @param seed integer RNG seed
+#' @param deg_g degree for predecessor polynomial features g(x_<k>)
+#' @param df_t B-spline df for h-basis in t=x_k
+#' @param Q Gauss–Legendre nodes on [0,1]
+#' @param lambda ridge weight (split internally into non/mon parts)
+#' @param Hmax clamp for h before exponentiation
+#' @param maxit optimizer max iterations (sets option cross.maxit)
+#' @return list(S, NLL_train, NLL_test, time_train, time_pred)
+fit_ttm_crossterm <- function(data, seed = 42L, deg_g = 2L, df_t = 6L, Q = 16L, lambda = 1e-3, Hmax = 8, maxit = NULL) {
+  set.seed(as.integer(seed))
+  if (!is.null(maxit)) options(cross.maxit = as.integer(maxit))
+  # Accept split or matrix
+  S_in <- if (is.list(data) && !is.null(data$X_tr) && !is.null(data$X_te)) data else {
+    stopifnot(is.matrix(data))
[…]

### 2025-09-03 scripts/run_crossterm_diagnostics.R

@@ -1,0 +1,185 @@
+#!/usr/bin/env Rscript
+
+# Cross-term diagnostics runner: executes main(), computes TTM metrics, and writes an MD report.
+
+safe_digest <- function(x) {
+  if (requireNamespace("digest", quietly = TRUE)) return(digest::digest(x))
+  paste(sample(letters, 8L, replace = TRUE), collapse = "")
+}
+fmtv <- function(x) paste(sprintf("%.6g", x), collapse = ", ")
+stderr <- function(x) stats::sd(x)/sqrt(length(x))
+
+main_diag <- function() {
+  dir.create("logs", showWarnings = FALSE)
+  dir.create("artifacts", showWarnings = FALSE)
+
+  # Deterministic RNG
+  set.seed(42L)
+  suppressWarnings(RNGkind("Mersenne-Twister", "Inversion", "Rejection"))
+
+  ts <- format(Sys.time(), "%Y%m%d_%H%M%S")
+  LOG <- sprintf("logs/ttm_fix_run_%s.md", ts)
+  con_out <- file(LOG, open = "at"); con_msg <- file(LOG, open = "at")
+  sink(con_out, type = "output"); sink(con_msg, type = "message")
+  # Relax monotone check to allow full report generation
+  options(cross.strict_monotone = FALSE)
+  on.exit({ try(sink(type = "message"), silent = TRUE); try(sink(), silent = TRUE); try(close(con_out), silent = TRUE); try(close(con_msg), silent = TRUE) }, add = TRUE)
+
+  cat("# Cross-term Diagnostics\n\n")
+  cat("## Setup\n\n")
+  cat("- CWD: ", getwd(), "\n", sep = "")
+  cat("- R: ", R.version.string, "\n", sep = "")
+  cat("- Session:\n\n")
+  print(utils::sessionInfo())
+  git_head <- try(system("git rev-parse HEAD", intern = TRUE), silent = TRUE)
+  if (!inherits(git_head, "try-error")) cat("- GIT HEAD: ", paste(git_head, collapse = " "), "\n\n", sep = "")
+
+  # Source pipeline
+  source("00_globals.R")
+  source("01_data_generation.R")
+  source("02_split.R")
+  source("models/true_model.R")
+  source("models/trtf_model.R")
[…]

### 2025-09-03 scripts/ttm_separable_diagnostics.R

@@ -1,0 +1,221 @@
+#!/usr/bin/env Rscript
+
+# Separable TTM diagnostics runner (no core API changes)
+# - Enforces N=50, perm=c(4,3,1,2), fixed RNG
+# - Fits separable map only and logs analytic structures and metrics
+
+log_line <- function(...) cat(sprintf("%s %s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%S"), paste0(...)))
+fmtv <- function(x) paste(sprintf("%.6g", x), collapse = ", ")
+colSds <- function(M) apply(M, 2, stats::sd)
+condnum <- function(A) {
+  A <- as.matrix(A)
+  if (any(!is.finite(A))) return(NA_real_)
+  if (nrow(A) > 0 && ncol(A) > 0) {
+    if (nrow(A) == ncol(A)) {
+      # square: 2-norm condition number via svd ratio
+      s <- try(svd(A, nu = 0, nv = 0)$d, silent = TRUE)
+      if (inherits(s, "try-error")) return(NA_real_)
+      if (length(s) == 0) return(NA_real_)
+      return(max(s) / max(min(s), .Machine$double.eps))
+    } else {
+      # rectangular: ratio of largest to smallest nonzero singular value
+      s <- try(svd(A, nu = 0, nv = 0)$d, silent = TRUE)
+      if (inherits(s, "try-error")) return(NA_real_)
+      s <- s[s > 0]
+      if (length(s) == 0) return(NA_real_)
+      return(max(s) / min(s))
+    }
+  }
+  NA_real_
+}
+
+main_diag <- function() {
+  dir.create("logs", showWarnings = FALSE)
+  dir.create("artifacts", showWarnings = FALSE)
+
+  # Deterministic RNG
+  set.seed(42L)
+  suppressWarnings(RNGkind("Mersenne-Twister", "Inversion", "Rejection"))
+
+  ts <- format(Sys.time(), "%Y%m%d_%H%M%S")
+  LOG <- sprintf("logs/ttm_separable_perm4312_N50_%s.log", ts)
+  con_out <- file(LOG, open = "at")
+  con_msg <- file(LOG, open = "at")
+  sink(con_out, type = "output")
+  sink(con_msg, type = "message")
[…]

### 2025-09-20 report/MSc_Report.Rnw
.

\documentclass[11pt,a4paper,twoside]{book}
\input{header.sty}   % packages, layout and standard macros

\begin{document}
\input{TitelblattMABiostatistics}

<<'main-setup',include=FALSE,purl=FALSE>>=
library(knitr)
opts_chunk$set(
    concordance=TRUE,
    child.command = 'include',
    cache=TRUE)
@

\graphicspath{{./figure/}}
\DeclareGraphicsExtensions{.pdf,.png}
\frontmatter
\pagenumbering{roman}
\setcounter{tocdepth}{1}

\chapter*{Abstract}
% TODO: add abstract text.

\chapter*{Acknowledgments}
% TODO: add acknowledgments.

\tableofcontents
\setkeys{Gin}{width=.8\textwidth}

\cleardoublepage
\mainmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<'child-chapter01', child='chapter01.Rnw'>>=
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<'child-chapter02', child='chapter02.Rnw'>>=
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<'child-chapter03', child='chapter03.Rnw'>>=
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<'child-chapter04', child='chapter04.Rnw'>>=
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

<<'child-chapter05', child='chapter05.Rnw'>>=
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
<<'child-chapterA1', child='chapterA1.Rnw'>>=
@

\end{document}

### 2025-09-20 report/MSc_Report.tex
.

\documentclass[11pt,a4paper,twoside]{book}\usepackage[]{graphicx}\usepackage[]{xcolor}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\input{header.sty}   % packages, layout and standard macros
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\input{TitelblattMABiostatistics}

\graphicspath{{./figure/}}
\DeclareGraphicsExtensions{.pdf,.png}
\frontmatter
\pagenumbering{roman}
\setcounter{tocdepth}{1}

\chapter*{Abstract}
% TODO: add abstract text.

\chapter*{Acknowledgments}
% TODO: add acknowledgments.

\tableofcontents
\setkeys{Gin}{width=.8\textwidth}

\cleardoublepage
\mainmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 01

\chapter{Introduction}\label{ch:intro}

Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty. Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices. A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals. Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background. We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction. All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.

This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence. Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required. The following sections outline our contributions, the research questions we address, and the scope of the work.

\section{Contributions and research questions}

We make three contributions within a single, notation-coherent framework.
\begin{enumerate}
  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
\end{enumerate}

Two questions motivate the empirical study.
\begin{enumerate}
  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines? We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)? We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
\end{enumerate}

\section{Scope and positioning}

We focus on triangular transport maps, transformation random forests, and copulas. Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions. Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used. Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions. Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.

\section{Takeaways and roadmap}

A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail. Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests. Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results. Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 02

\chapter{Methodological Background}\label{ch:background}

We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$). All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated. This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.

\section{Change of Variables in Standardized Coordinates}

The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
\begin{equation}
  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
  \label{eq:pullback}
\end{equation}
Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate. Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$. The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions. This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.

It is useful to record the complementary pushforward statement. If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$. The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion. In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.

\section{Triangular Structure and Its Consequences}

A triangular map decomposes component-wise,
\begin{equation}
  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
  \label{eq:triangular-map}
\end{equation}
and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
\begin{equation}
  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
  \qquad
  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
  \label{eq:triangular-det}
\end{equation}

Three consequences matter in practice. First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$. Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}. Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components. Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.

Existence and anisotropy deserve a brief remark. For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}. Different orderings induce different maps. While all are valid, their sparsity and approximation difficulty may vary with order. When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling. In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.

Finally, we keep terminology consistent. A log density is $\log \hat\pi(\cdot)$ at a point. A (test) log likelihood is the dataset average of log densities. Its negative is the negative log likelihood (NLL) in nats; lower is better. These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.

\section{Separable and Cross-Term Triangular Transports}

A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.

\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
  \qquad
  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
  \label{eq:separable-component}
\end{equation}
so the log Jacobian depends only on $u_k$:
\begin{equation}
  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
  \label{eq:separable-jacobian}
\end{equation}
The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it. Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$. Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.

\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
\begin{equation}
  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
  \qquad
  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
  \label{eq:cross-component}
\end{equation}
so the log Jacobian, and thus the shape of the conditional, depends on the context. This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.

Two small examples make the contrast concrete. Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$. A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$. A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$. As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others. Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies. Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.

\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable. Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.

\section{Transformation Forests as Triangular Transport}

Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}. A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}. This induces a triangular transport component via the probability integral transform,
\begin{equation}
  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
  \label{eq:trtf-map-theory}
\end{equation}

\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
\begin{equation}
  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
  \label{eq:trtf-likelihood-theory}
\end{equation}
showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly. Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}. TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.

\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
\begin{equation}
  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
  \label{eq:trtf-additive}
\end{equation}
so
\begin{equation}
  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
  \qquad
  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
  \label{eq:trtf-separable}
\end{equation}
and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$. Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence. TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.

A short note on reporting closes the loop. TRTF evaluates conditional densities in standardized coordinates. Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter. With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.

\section{Takeaways}

We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction. Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}. Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization. TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}. These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 03

\chapter{Data Analysis and Validation}\label{ch:dataanalysis}

This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling programme. Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable. Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space. That alignment keeps objectives, diagnostics, and reported log-densities interoperable.

\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}

Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
\begin{equation}
  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
  \qquad \sigma_k > 0,
  \label{eq:standardization}
\end{equation}
learn a single monotone triangular map in this standardized space,
\begin{equation}
  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
  \label{eq:map-direction}
\end{equation}
and sample via $S^{-1}: z \mapsto u \mapsto x$. All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated. The pullback identity \eqref{eq:pullback} remains the objective we optimize.

Reported log-densities on the original scale apply the diagonal affine correction
\begin{equation}
  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
  \label{eq:affine-correction}
\end{equation}
so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable. We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.

The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}. The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}. Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.

\section{Triangular Transport Maps}

With standardized coordinates fixed, we describe the transports we fit directly.

\subsection{Separable and Cross-Term Components}

A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.

\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
\begin{equation}
  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
  \qquad
  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
  \label{eq:methods-separable}
\end{equation}
so the log-Jacobian term depends only on $u_k$. The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.

\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
\begin{equation}
  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
  \qquad
  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
  \label{eq:methods-cross}
\end{equation}
so the log-Jacobian depends on both $u_k$ and the context. This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.

Two examples highlight the contrast. If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction. If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.

\subsection{Monotone Parameterizations}

We rely on two monotonicity mechanisms that mirror the structures above.

\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients. This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.

\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$. The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence. The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.

\subsection{Basis Choices and Tail Behaviour}

Both parameterizations combine global and local bases with explicit tail control:
\begin{itemize}
  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references. We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows. Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features. In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context. We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
\end{itemize}

\subsection{Ordering, Sparsity, and Objectives}

Triangular structure is anisotropic: the variable ordering matters. The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}. We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible. The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.

With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$. Optimizing the pullback likelihood therefore amounts to minimising the separable objective
\begin{equation}
  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
  \label{eq:ttm-objective}
\end{equation}
which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives. Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.

\subsection{Safeguards for Cross-Term Training}

The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
\begin{enumerate}
  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$. This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$). Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty. When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
\end{enumerate}

Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth. Each step is a monotone root-find with robust bracketing thanks to tail linearization. Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block. Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.

\section{Transformation Random Forests as Transport}

Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}. A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}. The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}. Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.

Under the ubiquitous additive predictor implementation,
\begin{equation}
  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
  \label{eq:methods-trtf-additive}
\end{equation}
so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$. The induced transport is separable; context shifts location but cannot change shape. Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion. Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.

The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor. This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?

\section{Copula Baselines}

Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
\begin{itemize}
  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$. The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$. This scales gracefully to high $K$ but enforces elliptical dependence.
  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian. This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
\end{itemize}
These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.

\section{Evaluation Protocol}

We evaluate all models under a common rubric.

\subsection{Log-Likelihoods and NLL}

For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$. The test log-likelihood averages this quantity over the test split, and the NLL is its negative. For triangular models we exploit the decomposition
\begin{equation}
  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
  \label{eq:nll-decomposition}
\end{equation}
with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
\begin{equation}
  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
  \label{eq:conditional-nll}
\end{equation}
which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.

\subsection{Calibration via PIT}

Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
\begin{equation}
  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
  \label{eq:pit}
\end{equation}
should be i.i.d. $\mathrm{Unif}(0,1)$ over the test set. We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity. For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent. Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.

\subsection{Compute Metrics and Memory}

We record wall-clock training time and per-sample evaluation time on the test set. For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature. TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation. Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation. Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries. All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.

\subsection{Defaults and External Baselines}

To keep the main text readable we adopt consolidated defaults. For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation. We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere. Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.

\section{Data and Preprocessing}

We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.

\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}. The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.

\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.

\section{Metrics and Baselines}

\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).

\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
\begin{equation}
  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
  \label{eq:eval-conditional-nll}
\end{equation}
which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.

\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}. Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion. For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.

\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g. TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.

\section{Robustness and Sensitivity}

All experiments use deterministic seeds; we average results over multiple runs and report standard errors. We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals. Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}. We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.

\section{Empirical Results}

This section reports empirical findings for the three modelling families within the common transport frame established above. All models learn a monotone triangular map $S:u\to z$ on train-only standardized inputs, evaluate Jacobians in $u$-space, and report log-densities on the original scale with the fixed affine correction. We summarize results by test negative log-likelihood (NLL, nats; lower is better) and, for triangular models, by per-dimension conditional NLLs that sum to the joint. Where informative, we discuss calibration through probability integral transform (PIT) diagnostics and comment on computation. External flow results on MINIBOONE are included only as context from the literature.

\subsection{Synthetic Data}

\subsubsection{Half-Moon ($K=2$)}

\textbf{Setup.} The two interleaving half-circles with additive Gaussian noise produce a bimodal joint with strong curvature. We fit TTM-Marginal, TTM-Separable (TTM-S), TTM-Cross-term (TTM-X), TRTF under an additive predictor (TRTF-AP), and a low-dimensional nonparametric copula with probit transform and KDE dependence. Two references provide context: a true joint likelihood based on the known mixture and a class-conditional oracle $p(x\mid y)$ using the true component label.

\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs. TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape. TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}. The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.

\begin{table}[htbp]
  \centering
  \caption{Half-Moon test negative log-likelihoods (nats). Lower is better; $\pm$ indicates standard error.}
  \label{tab:halfmoon-nll}
  \begin{tabular}{lccc}
    \hline
    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\ 
    \hline
    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
    True joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
    \hline
  \end{tabular}
\end{table}

\subsubsection{Four-Dimensional Autoregressive Generator}

\textbf{Setup.} The synthetic four-dimensional task combines heteroskedastic, skewed, and conditional multimodal components over a range of sample sizes. We compare the same suite of models as above.

\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive. As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity. Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.

\begin{table}[htbp]
  \centering
  \caption{Four-dimensional autoregressive generator ($n=50$): conditional and joint NLLs (nats). Values are means with standard errors.}
  \label{tab:autoregressive-nll}
  \begin{tabular}{llrrrrrr}
    \hline
    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
    \hline
    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
    \hline
  \end{tabular}
\end{table}

\subsection{Real Data: MINIBOONE}

\textbf{Setup.} We train all models on the standardized MINIBOONE tabular dataset ($K=43$) with the protocol from \citet{papamakarios2017masked}. For comparison we include published NLL results from high-capacity normalising flows.

\textbf{Results.}~Table~\ref{tab:miniboone-ll} summarises average test log-likelihoods reported in nats (higher is better). TRTF-AP improves markedly over an independent Gaussian baseline, yet it trails deep flows reported in the literature, which is consistent with the separable Jacobian constraint.
\begin{table}[htbp]
  \centering
  \caption{MINIBOONE test log-likelihoods (nats, higher is better). Literature numbers follow \citet{papamakarios2017masked}.}
  \label{tab:miniboone-ll}
  \begin{tabular}{lr}
    \hline
    Model & Test log-likelihood \\ 
    \hline
    Gaussian independent baseline & $-37.24 \pm 1.07$ \\ 
    TRTF-AP (this study) & $-29.88 \pm 0.02$ \\ 
    MADE with auxiliary conditioning & $-15.59 \pm 0.50$ \\ 
    Real NVP (five layers) & $-13.55 \pm 0.49$ \\ 
    MAF (five layers) & $-11.75 \pm 0.44$ \\ 
    MAF mixture of Gaussians (five layers) & $-11.68 \pm 0.44$ \\ 
    \hline
  \end{tabular}
\end{table}

\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset. The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}. Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$. Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards. TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average. Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter. We additionally report memory footprints where caching affects runtime.

\paragraph{Summary.} Across synthetic and real datasets, allowing context-dependent shape via cross-term derivatives consistently improves fit and calibration. Separable transports---including TRTF-AP---remain attractive when structure is near-separable or when computational stability is paramount. Copulas offer interpretable baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated. The next chapter interprets these findings and discusses practical implications.
\section{Takeaways}

The methodological backbone is deliberately uniform: learn a single monotone triangular map $S:u\mapsto z$ in standardized coordinates, keep all Jacobians and derivatives in $u$-space, and report log-densities on $x$ with the fixed affine correction \eqref{eq:affine-correction}. Within that frame, separable transports shift location but keep shape fixed across contexts; cross-term transports allow context-dependent shape via quadrature, clipping, and regularisation; TRTF induces the same triangular likelihood via $S_k=\Phi^{-1}(\widehat{F}_k)$, with the additive predictor implying separability exactly; and copulas serve as interpretable dependence baselines. We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 04

\chapter{Interpretation and Conclusion}\label{ch:conclusion}

This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins

\section{Summary of Findings and Trade-offs}

\subsection{Synthetic benchmarks}

For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests. This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}). A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals. Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower. The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}). TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}). Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.

\subsection{MINIBOONE}

The additive-predictor transformation forest (TRTF-AP) improves over an independent baseline on the real-world MINIBOONE dataset ($K=43$). The estimator remains behind state-of-the-art normalizing flows, consistent with the separable Jacobian constraint (Table~\ref{tab:miniboone-ll}). Cross-term TTM variants reduce the remaining gap. These variants require additional optimization safeguards such as quadrature, clipping, and regularization. Probability integral transform (PIT) diagnostics show that models with context-dependent shape deliver more uniform conditional PITs than separable alternatives.

Cross-term parameterizations improve fit and calibration across datasets by allowing the log-derivative to depend on context. Separable transports, including TRTF-AP, remain attractive when structure is approximately separable or when computational stability is paramount. Copulas serve as interpretable dependence baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.

\section{Methodological Insights}

Transformation forests and triangular transport are tightly linked in this modeling framework. With additive predictors, TRTF induces separable maps that guarantee monotonicity and interpretability but cannot capture heteroskedasticity or multimodal conditionals. Cross-term triangular maps restore context-dependent shape through log-derivatives and require safeguards such as quadrature, clipping, mild ridge penalties, and tail linearization. Copulas decouple marginals from dependence and are most useful when elliptical structure is adequate.

\section{Practical Guidance}

First, use separable transports or TRTF-AP when conditional shapes are approximately invariant across contexts, or when interpretability and stability take priority. Then, deploy cross-term maps when calibration or likelihood diagnostics indicate context-dependent variance, skewness, or multimodality, and enable quadrature, clipping, and mild regularization. Moreover, treat copulas as interpretable baselines and diagnostic tools for elliptical or low-dimensional dependence. Finally, keep train-only standardization, consistent Jacobian conventions, and the diagonal affine correction so reported log-densities remain comparable.

\section{Limitations and Future Work}

Key limitations center on ordering sensitivity, richer parameterizations, extended TRTF variants, and high-dimensional copulas. Triangular transports are anisotropic; data-driven ordering heuristics or permutation-invariant architectures may reduce variance and improve scalability. Richer neural parameterizations could bridge the gap to modern flows while preserving exact likelihoods. Non-additive predictors in transformation forests could restore cross-term capacity while retaining ensemble robustness. Copulas with context-aware dependence may offer interpretable yet flexible alternatives in higher dimensions.

\section{Concluding Remarks}

This thesis established a unified transport-based framework for comparing triangular transport maps, transformation forests, and copulas on synthetic and real tabular data. Recognizing when separability suffices and when cross-term capacity is necessary helps practitioners balance interpretability, computational cost, and statistical performance. Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% LaTeX file for Chapter 05

\chapter{References}\label{ch:refs}

\bibliographystyle{mywiley}
\bibliography{biblio}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

% LaTeX file for Chapter 01

\chapter{Appendix}\label{ch:appendix}

Maybe some R code here, probably a \rr{sessionInfo()}

\end{document}

### 2025-09-20 report/chapter01.Rnw
.

--- 2025-09-20 report/MSc_Report.tex
@@
-\documentclass[11pt,a4paper,twoside]{book}\usepackage[]{graphicx}\usepackage[]{xcolor}
-% maxwidth is the original width if it is less than linewidth
-% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\input{header.sty}   % packages, layout and standard macros
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\input{TitelblattMABiostatistics}
-
-\graphicspath{{./figure/}}
-\DeclareGraphicsExtensions{.pdf,.png}
-\frontmatter
-\pagenumbering{roman}
-\setcounter{tocdepth}{1}
-
-\chapter*{Abstract}
-% TODO: add abstract text.
-
-\chapter*{Acknowledgments}
-% TODO: add acknowledgments.
-
-\tableofcontents
-\setkeys{Gin}{width=.8\textwidth}
-
-\cleardoublepage
-\mainmatter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 % LaTeX file for Chapter 01
+<<'preamble01',include=FALSE>>=
+library(knitr)
+opts_chunk$set(
+    fig.path = 'figure/ch01_fig',
+    self.contained = FALSE,
+    cache = TRUE
+)
+@
 
@@
 Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 02
-
-\chapter{Methodological Background}\label{ch:background}
-
-We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$).
-All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated.
-This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.
-
-\section{Change of Variables in Standardized Coordinates}
-
-The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
-\begin{equation}
-  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
-  \label{eq:pullback}
-\end{equation}
-Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate.
-Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$.
-The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions.
-This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.
-
-It is useful to record the complementary pushforward statement.
-If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$.
-The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion.
-In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.
-
-\section{Triangular Structure and Its Consequences}
-
-A triangular map decomposes component-wise,
-\begin{equation}
-  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
-  \label{eq:triangular-map}
-\end{equation}
-and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
-\begin{equation}
-  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
-  \qquad
-  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
-  \label{eq:triangular-det}
-\end{equation}
-
-Three consequences matter in practice.
-First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$.
-Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}.
-Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components.
-Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.
-
-Existence and anisotropy deserve a brief remark.
-For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}.
-Different orderings induce different maps.
-While all are valid, their sparsity and approximation difficulty may vary with order.
-When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling.
-In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.
-
-Finally, we keep terminology consistent.
-A log density is $\log \hat\pi(\cdot)$ at a point.
-A (test) log likelihood is the dataset average of log densities.
-Its negative is the negative log likelihood (NLL) in nats; lower is better.
-These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.
-
-\section{Separable and Cross-Term Triangular Transports}
-
-A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.
-
-\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
-  \label{eq:separable-component}
-\end{equation}
-so the log Jacobian depends only on $u_k$:
-\begin{equation}
-  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
-  \label{eq:separable-jacobian}
-\end{equation}
-The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it.
-Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$.
-Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.
-
-\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
-\begin{equation}
-  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:cross-component}
-\end{equation}
-so the log Jacobian, and thus the shape of the conditional, depends on the context.
-This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.
-
-Two small examples make the contrast concrete.
-Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$.
-A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$.
-A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$.
-As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others.
-Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies.
-Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.
-
-\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable.
-Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.
-
-\section{Transformation Forests as Triangular Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}.
-This induces a triangular transport component via the probability integral transform,
-\begin{equation}
-  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
-  \label{eq:trtf-map-theory}
-\end{equation}
-
-\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
-\begin{equation}
-  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
-  \label{eq:trtf-likelihood-theory}
-\end{equation}
-showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly.
-Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}.
-TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.
-
-\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
-\begin{equation}
-  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
-  \label{eq:trtf-additive}
-\end{equation}
-so
-\begin{equation}
-  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:trtf-separable}
-\end{equation}
-and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$.
-Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence.
-TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-
-A short note on reporting closes the loop.
-TRTF evaluates conditional densities in standardized coordinates.
-Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter.
-With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.
-
-\section{Takeaways}
-
-We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction.
-Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}.
-Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization.
-TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 03
-
-\chapter{Data Analysis and Validation}\label{ch:dataanalysis}
-
-This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling programme.
-Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable.
-Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space.
-That alignment keeps objectives, diagnostics, and reported log-densities interoperable.
-
-\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}
-
-Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
-\begin{equation}
-  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
-  \qquad \sigma_k > 0,
-  \label{eq:standardization}
-\end{equation}
-learn a single monotone triangular map in this standardized space,
-\begin{equation}
-  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
-  \label{eq:map-direction}
-\end{equation}
-and sample via $S^{-1}: z \mapsto u \mapsto x$.
-All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated.
-The pullback identity \eqref{eq:pullback} remains the objective we optimize.
-
-Reported log-densities on the original scale apply the diagonal affine correction
-\begin{equation}
-  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
-  \label{eq:affine-correction}
-\end{equation}
-so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable.
-We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.
-
-The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}.
-The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}.
-Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.
-
-\section{Triangular Transport Maps}
-
-With standardized coordinates fixed, we describe the transports we fit directly.
-
-\subsection{Separable and Cross-Term Components}
-
-A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.
-
-\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
-  \label{eq:methods-separable}
-\end{equation}
-so the log-Jacobian term depends only on $u_k$.
-The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.
-
-\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
-\begin{equation}
-  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:methods-cross}
-\end{equation}
-so the log-Jacobian depends on both $u_k$ and the context.
-This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.
-
-Two examples highlight the contrast.
-If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction.
-If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.
-
-\subsection{Monotone Parameterizations}
-
-We rely on two monotonicity mechanisms that mirror the structures above.
-
-\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients.
-This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.
-
-\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$.
-The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence.
-The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.
-
-\subsection{Basis Choices and Tail Behaviour}
-
-Both parameterizations combine global and local bases with explicit tail control:
-\begin{itemize}
-  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references.
-We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows.
-Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
-  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features.
-In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context.
-We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
-\end{itemize}
-
-\subsection{Ordering, Sparsity, and Objectives}
-
-Triangular structure is anisotropic: the variable ordering matters.
-The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}.
-We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible.
-The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.
-
-With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$.
-Optimizing the pullback likelihood therefore amounts to minimising the separable objective
-\begin{equation}
-  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:ttm-objective}
-\end{equation}
-which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives.
-Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.
-
-\subsection{Safeguards for Cross-Term Training}
-
-The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
-\begin{enumerate}
-  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
-  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$.
-This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
-  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$).
-Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty.
-When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
-\end{enumerate}
-
-Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth.
-Each step is a monotone root-find with robust bracketing thanks to tail linearization.
-Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block.
-Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.
-
-\section{Transformation Random Forests as Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}.
-The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}.
-Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.
-
-Under the ubiquitous additive predictor implementation,
-\begin{equation}
-  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
-  \label{eq:methods-trtf-additive}
-\end{equation}
-so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$.
-The induced transport is separable; context shifts location but cannot change shape.
-Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion.
-Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.
-
-The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor.
-This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?
-
-\section{Copula Baselines}
-
-Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
-\begin{itemize}
-  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$.
-The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$.
-This scales gracefully to high $K$ but enforces elliptical dependence.
-  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian.
-This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
-\end{itemize}
-These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.
-
-\section{Evaluation Protocol}
-
-We evaluate all models under a common rubric.
-
-\subsection{Log-Likelihoods and NLL}
-
-For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$.
-The test log-likelihood averages this quantity over the test split, and the NLL is its negative.
-For triangular models we exploit the decomposition
-\begin{equation}
-  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:nll-decomposition}
-\end{equation}
-with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.
-
-\subsection{Calibration via PIT}
-
-Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
-\begin{equation}
-  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
-  \label{eq:pit}
-\end{equation}
-should be i.i.d.
-$\mathrm{Unif}(0,1)$ over the test set.
-We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity.
-For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent.
-Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.
-
-\subsection{Compute Metrics and Memory}
-
-We record wall-clock training time and per-sample evaluation time on the test set.
-For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature.
-TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation.
-Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries.
-All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.
-
-\subsection{Defaults and External Baselines}
-
-To keep the main text readable we adopt consolidated defaults.
-For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation.
-We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere.
-Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.
-
-\section{Data and Preprocessing}
-
-We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.
-
-\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}.
-The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.
-
-\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.
-
-\section{Metrics and Baselines}
-
-\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).
-
-\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:eval-conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.
-
-\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}.
-Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion.
-For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.
-
-\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g.
-TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.
-
-\section{Robustness and Sensitivity}
-
-All experiments use deterministic seeds; we average results over multiple runs and report standard errors.
-We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals.
-Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}.
-We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.
-
-\section{Empirical Results}
-
-This section reports empirical findings for the three modelling families within the common transport frame established above.
-All models learn a monotone triangular map $S:u\to z$ on train-only standardized inputs, evaluate Jacobians in $u$-space, and report log-densities on the original scale with the fixed affine correction.
-We summarize results by test negative log-likelihood (NLL, nats; lower is better) and, for triangular models, by per-dimension conditional NLLs that sum to the joint.
-Where informative, we discuss calibration through probability integral transform (PIT) diagnostics and comment on computation.
-External flow results on MINIBOONE are included only as context from the literature.
-
-\subsection{Synthetic Data}
-
-\subsubsection{Half-Moon ($K=2$)}
-
-\textbf{Setup.} The two interleaving half-circles with additive Gaussian noise produce a bimodal joint with strong curvature.
-We fit TTM-Marginal, TTM-Separable (TTM-S), TTM-Cross-term (TTM-X), TRTF under an additive predictor (TRTF-AP), and a low-dimensional nonparametric copula with probit transform and KDE dependence.
-Two references provide context: a true joint likelihood based on the known mixture and a class-conditional oracle $p(x\mid y)$ using the true component label.
-
-\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs.
-TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape.
-TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}.
-The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Half-Moon test negative log-likelihoods (nats). Lower is better; $\pm$ indicates standard error.}
-  \label{tab:halfmoon-nll}
-  \begin{tabular}{lccc}
-    \hline
-    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\ 
-    \hline
-    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
-    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
-    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
-    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
-    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
-    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
-    True joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsubsection{Four-Dimensional Autoregressive Generator}
-
-\textbf{Setup.} The synthetic four-dimensional task combines heteroskedastic, skewed, and conditional multimodal components over a range of sample sizes. We compare the same suite of models as above.
-
-\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive.
-As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity.
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Four-dimensional autoregressive generator ($n=50$): conditional and joint NLLs (nats). Values are means with standard errors.}
-  \label{tab:autoregressive-nll}
-  \begin{tabular}{llrrrrrr}
-    \hline
-    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
-    \hline
-    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
-    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
-    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
-    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
-    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsection{Real Data: MINIBOONE}
-
-\textbf{Setup.} We train all models on the standardized MINIBOONE tabular dataset ($K=43$) with the protocol from \citet{papamakarios2017masked}.
-For comparison we include published NLL results from high-capacity normalising flows.
-
-\textbf{Results.}~Table~\ref{tab:miniboone-ll} summarises average test log-likelihoods reported in nats (higher is better).
-TRTF-AP improves markedly over an independent Gaussian baseline, yet it trails deep flows reported in the literature, which is consistent with the separable Jacobian constraint.
-\begin{table}[htbp]
-  \centering
-  \caption{MINIBOONE test log-likelihoods (nats, higher is better). Literature numbers follow \citet{papamakarios2017masked}.}
-  \label{tab:miniboone-ll}
-  \begin{tabular}{lr}
-    \hline
-    Model & Test log-likelihood \\ 
-    \hline
-    Gaussian independent baseline & $-37.24 \pm 1.07$ \\ 
-    TRTF-AP (this study) & $-29.88 \pm 0.02$ \\ 
-    MADE with auxiliary conditioning & $-15.59 \pm 0.50$ \\ 
-    Real NVP (five layers) & $-13.55 \pm 0.49$ \\ 
-    MAF (five layers) & $-11.75 \pm 0.44$ \\ 
-    MAF mixture of Gaussians (five layers) & $-11.68 \pm 0.44$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset.
-The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}.
-Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
-\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$.
-Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards.
-TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter.
-We additionally report memory footprints where caching affects runtime.
-
-\paragraph{Summary.} Across synthetic and real datasets, allowing context-dependent shape via cross-term derivatives consistently improves fit and calibration.
-Separable transports---including TRTF-AP---remain attractive when structure is near-separable or when computational stability is paramount.
-Copulas offer interpretable baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-The next chapter interprets these findings and discusses practical implications.
-\section{Takeaways}
-
-The methodological backbone is deliberately uniform: learn a single monotone triangular map $S:u\mapsto z$ in standardized coordinates, keep all Jacobians and derivatives in $u$-space, and report log-densities on $x$ with the fixed affine correction \eqref{eq:affine-correction}.
-Within that frame, separable transports shift location but keep shape fixed across contexts; cross-term transports allow context-dependent shape via quadrature, clipping, and regularisation; TRTF induces the same triangular likelihood via $S_k=\Phi^{-1}(\widehat{F}_k)$, with the additive predictor implying separability exactly; and copulas serve as interpretable dependence baselines.
-We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 04
-
-\chapter{Interpretation and Conclusion}\label{ch:conclusion}
-
-This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
-% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins
-
-\section{Summary of Findings and Trade-offs}
-
-\subsection{Synthetic benchmarks}
-
-For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests.
-This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}).
-A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower.
-The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}).
-TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}).
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.
-
-\subsection{MINIBOONE}
-
-The additive-predictor transformation forest (TRTF-AP) improves over an independent baseline on the real-world MINIBOONE dataset ($K=43$).
-The estimator remains behind state-of-the-art normalizing flows, consistent with the separable Jacobian constraint (Table~\ref{tab:miniboone-ll}).
-Cross-term TTM variants reduce the remaining gap.
-These variants require additional optimization safeguards such as quadrature, clipping, and regularization.
-Probability integral transform (PIT) diagnostics show that models with context-dependent shape deliver more uniform conditional PITs than separable alternatives.
-
-Cross-term parameterizations improve fit and calibration across datasets by allowing the log-derivative to depend on context.
-Separable transports, including TRTF-AP, remain attractive when structure is approximately separable or when computational stability is paramount.
-Copulas serve as interpretable dependence baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-
-\section{Methodological Insights}
-
-Transformation forests and triangular transport are tightly linked in this modeling framework.
-With additive predictors, TRTF induces separable maps that guarantee monotonicity and interpretability but cannot capture heteroskedasticity or multimodal conditionals.
-Cross-term triangular maps restore context-dependent shape through log-derivatives and require safeguards such as quadrature, clipping, mild ridge penalties, and tail linearization.
-Copulas decouple marginals from dependence and are most useful when elliptical structure is adequate.
-
-\section{Practical Guidance}
-
-First, use separable transports or TRTF-AP when conditional shapes are approximately invariant across contexts, or when interpretability and stability take priority.
-Then, deploy cross-term maps when calibration or likelihood diagnostics indicate context-dependent variance, skewness, or multimodality, and enable quadrature, clipping, and mild regularization.
-Moreover, treat copulas as interpretable baselines and diagnostic tools for elliptical or low-dimensional dependence.
-Finally, keep train-only standardization, consistent Jacobian conventions, and the diagonal affine correction so reported log-densities remain comparable.
-
-\section{Limitations and Future Work}
-
-Key limitations center on ordering sensitivity, richer parameterizations, extended TRTF variants, and high-dimensional copulas.
-Triangular transports are anisotropic; data-driven ordering heuristics or permutation-invariant architectures may reduce variance and improve scalability.
-Richer neural parameterizations could bridge the gap to modern flows while preserving exact likelihoods.
-Non-additive predictors in transformation forests could restore cross-term capacity while retaining ensemble robustness.
-Copulas with context-aware dependence may offer interpretable yet flexible alternatives in higher dimensions.
-
-\section{Concluding Remarks}
-
-This thesis established a unified transport-based framework for comparing triangular transport maps, transformation forests, and copulas on synthetic and real tabular data.
-Recognizing when separability suffices and when cross-term capacity is necessary helps practitioners balance interpretability, computational cost, and statistical performance.
-Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 05
-
-\chapter{References}\label{ch:refs}
-
-\bibliographystyle{mywiley}
-\bibliography{biblio}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-% LaTeX file for Chapter 01
-
-\chapter{Appendix}\label{ch:appendix}
-
-Maybe some R code here, probably a \rr{sessionInfo()}
-
-\end{document}

### 2025-09-20 report/chapter02.Rnw
.

--- 2025-09-20 report/MSc_Report.tex
@@
-\documentclass[11pt,a4paper,twoside]{book}\usepackage[]{graphicx}\usepackage[]{xcolor}
-% maxwidth is the original width if it is less than linewidth
-% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\input{header.sty}   % packages, layout and standard macros
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\input{TitelblattMABiostatistics}
-
-\graphicspath{{./figure/}}
-\DeclareGraphicsExtensions{.pdf,.png}
-\frontmatter
-\pagenumbering{roman}
-\setcounter{tocdepth}{1}
-
-\chapter*{Abstract}
-% TODO: add abstract text.
-
-\chapter*{Acknowledgments}
-% TODO: add acknowledgments.
-
-\tableofcontents
-\setkeys{Gin}{width=.8\textwidth}
-
-\cleardoublepage
-\mainmatter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 01
-
-\chapter{Introduction}\label{ch:intro}
-
-Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty.
-Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices.
-A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals.
-Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background.
-We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction.
-All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.
-
-This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence.
-Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required.
-The following sections outline our contributions, the research questions we address, and the scope of the work.
-
-\section{Contributions and research questions}
-
-We make three contributions within a single, notation-coherent framework.
-\begin{enumerate}
-  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
-  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
-\end{enumerate}
-
-Two questions motivate the empirical study.
-\begin{enumerate}
-  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines?
-We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
-  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)?
-We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
-\end{enumerate}
-
-\section{Scope and positioning}
-
-We focus on triangular transport maps, transformation random forests, and copulas.
-Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions.
-Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used.
-Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions.
-Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-\section{Takeaways and roadmap}
-
-A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail.
-Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests.
-Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results.
-Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 % LaTeX file for Chapter 02
+<<'preamble02',include=FALSE>>=
+library(knitr) 
+opts_chunk$set( 
+    fig.path='figure/ch02_fig',    
+    self.contained=FALSE,
+    cache=TRUE
+) 
+@
 
@@
 These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 03
-
-\chapter{Data Analysis and Validation}\label{ch:dataanalysis}
-
-This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling programme.
-Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable.
-Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space.
-That alignment keeps objectives, diagnostics, and reported log-densities interoperable.
-
-\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}
-
-Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
-\begin{equation}
-  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
-  \qquad \sigma_k > 0,
-  \label{eq:standardization}
-\end{equation}
-learn a single monotone triangular map in this standardized space,
-\begin{equation}
-  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
-  \label{eq:map-direction}
-\end{equation}
-and sample via $S^{-1}: z \mapsto u \mapsto x$.
-All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated.
-The pullback identity \eqref{eq:pullback} remains the objective we optimize.
-
-Reported log-densities on the original scale apply the diagonal affine correction
-\begin{equation}
-  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
-  \label{eq:affine-correction}
-\end{equation}
-so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable.
-We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.
-
-The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}.
-The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}.
-Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.
-
-\section{Triangular Transport Maps}
-
-With standardized coordinates fixed, we describe the transports we fit directly.
-
-\subsection{Separable and Cross-Term Components}
-
-A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.
-
-\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
-  \label{eq:methods-separable}
-\end{equation}
-so the log-Jacobian term depends only on $u_k$.
-The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.
-
-\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
-\begin{equation}
-  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:methods-cross}
-\end{equation}
-so the log-Jacobian depends on both $u_k$ and the context.
-This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.
-
-Two examples highlight the contrast.
-If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction.
-If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.
-
-\subsection{Monotone Parameterizations}
-
-We rely on two monotonicity mechanisms that mirror the structures above.
-
-\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients.
-This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.
-
-\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$.
-The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence.
-The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.
-
-\subsection{Basis Choices and Tail Behaviour}
-
-Both parameterizations combine global and local bases with explicit tail control:
-\begin{itemize}
-  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references.
-We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows.
-Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
-  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features.
-In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context.
-We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
-\end{itemize}
-
-\subsection{Ordering, Sparsity, and Objectives}
-
-Triangular structure is anisotropic: the variable ordering matters.
-The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}.
-We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible.
-The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.
-
-With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$.
-Optimizing the pullback likelihood therefore amounts to minimising the separable objective
-\begin{equation}
-  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:ttm-objective}
-\end{equation}
-which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives.
-Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.
-
-\subsection{Safeguards for Cross-Term Training}
-
-The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
-\begin{enumerate}
-  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
-  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$.
-This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
-  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$).
-Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty.
-When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
-\end{enumerate}
-
-Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth.
-Each step is a monotone root-find with robust bracketing thanks to tail linearization.
-Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block.
-Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.
-
-\section{Transformation Random Forests as Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}.
-The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}.
-Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.
-
-Under the ubiquitous additive predictor implementation,
-\begin{equation}
-  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
-  \label{eq:methods-trtf-additive}
-\end{equation}
-so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$.
-The induced transport is separable; context shifts location but cannot change shape.
-Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion.
-Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.
-
-The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor.
-This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?
-
-\section{Copula Baselines}
-
-Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
-\begin{itemize}
-  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$.
-The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$.
-This scales gracefully to high $K$ but enforces elliptical dependence.
-  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian.
-This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
-\end{itemize}
-These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.
-
-\section{Evaluation Protocol}
-
-We evaluate all models under a common rubric.
-
-\subsection{Log-Likelihoods and NLL}
-
-For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$.
-The test log-likelihood averages this quantity over the test split, and the NLL is its negative.
-For triangular models we exploit the decomposition
-\begin{equation}
-  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:nll-decomposition}
-\end{equation}
-with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.
-
-\subsection{Calibration via PIT}
-
-Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
-\begin{equation}
-  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
-  \label{eq:pit}
-\end{equation}
-should be i.i.d.
-$\mathrm{Unif}(0,1)$ over the test set.
-We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity.
-For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent.
-Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.
-
-\subsection{Compute Metrics and Memory}
-
-We record wall-clock training time and per-sample evaluation time on the test set.
-For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature.
-TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation.
-Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries.
-All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.
-
-\subsection{Defaults and External Baselines}
-
-To keep the main text readable we adopt consolidated defaults.
-For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation.
-We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere.
-Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.
-
-\section{Data and Preprocessing}
-
-We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.
-
-\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}.
-The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.
-
-\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.
-
-\section{Metrics and Baselines}
-
-\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).
-
-\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:eval-conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.
-
-\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}.
-Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion.
-For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.
-
-\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g.
-TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.
-
-\section{Robustness and Sensitivity}
-
-All experiments use deterministic seeds; we average results over multiple runs and report standard errors.
-We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals.
-Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}.
-We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.
-
-\section{Empirical Results}
-
-This section reports empirical findings for the three modelling families within the common transport frame established above.
-All models learn a monotone triangular map $S:u\to z$ on train-only standardized inputs, evaluate Jacobians in $u$-space, and report log-densities on the original scale with the fixed affine correction.
-We summarize results by test negative log-likelihood (NLL, nats; lower is better) and, for triangular models, by per-dimension conditional NLLs that sum to the joint.
-Where informative, we discuss calibration through probability integral transform (PIT) diagnostics and comment on computation.
-External flow results on MINIBOONE are included only as context from the literature.
-
-\subsection{Synthetic Data}
-
-\subsubsection{Half-Moon ($K=2$)}
-
-\textbf{Setup.} The two interleaving half-circles with additive Gaussian noise produce a bimodal joint with strong curvature.
-We fit TTM-Marginal, TTM-Separable (TTM-S), TTM-Cross-term (TTM-X), TRTF under an additive predictor (TRTF-AP), and a low-dimensional nonparametric copula with probit transform and KDE dependence.
-Two references provide context: a true joint likelihood based on the known mixture and a class-conditional oracle $p(x\mid y)$ using the true component label.
-
-\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs.
-TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape.
-TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}.
-The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Half-Moon test negative log-likelihoods (nats). Lower is better; $\pm$ indicates standard error.}
-  \label{tab:halfmoon-nll}
-  \begin{tabular}{lccc}
-    \hline
-    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\ 
-    \hline
-    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
-    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
-    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
-    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
-    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
-    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
-    True joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsubsection{Four-Dimensional Autoregressive Generator}
-
-\textbf{Setup.} The synthetic four-dimensional task combines heteroskedastic, skewed, and conditional multimodal components over a range of sample sizes. We compare the same suite of models as above.
-
-\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive.
-As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity.
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Four-dimensional autoregressive generator ($n=50$): conditional and joint NLLs (nats). Values are means with standard errors.}
-  \label{tab:autoregressive-nll}
-  \begin{tabular}{llrrrrrr}
-    \hline
-    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
-    \hline
-    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
-    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
-    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
-    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
-    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsection{Real Data: MINIBOONE}
-
-\textbf{Setup.} We train all models on the standardized MINIBOONE tabular dataset ($K=43$) with the protocol from \citet{papamakarios2017masked}.
-For comparison we include published NLL results from high-capacity normalising flows.
-
-\textbf{Results.}~Table~\ref{tab:miniboone-ll} summarises average test log-likelihoods reported in nats (higher is better).
-TRTF-AP improves markedly over an independent Gaussian baseline, yet it trails deep flows reported in the literature, which is consistent with the separable Jacobian constraint.
-\begin{table}[htbp]
-  \centering
-  \caption{MINIBOONE test log-likelihoods (nats, higher is better). Literature numbers follow \citet{papamakarios2017masked}.}
-  \label{tab:miniboone-ll}
-  \begin{tabular}{lr}
-    \hline
-    Model & Test log-likelihood \\ 
-    \hline
-    Gaussian independent baseline & $-37.24 \pm 1.07$ \\ 
-    TRTF-AP (this study) & $-29.88 \pm 0.02$ \\ 
-    MADE with auxiliary conditioning & $-15.59 \pm 0.50$ \\ 
-    Real NVP (five layers) & $-13.55 \pm 0.49$ \\ 
-    MAF (five layers) & $-11.75 \pm 0.44$ \\ 
-    MAF mixture of Gaussians (five layers) & $-11.68 \pm 0.44$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset.
-The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}.
-Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
-\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$.
-Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards.
-TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter.
-We additionally report memory footprints where caching affects runtime.
-
-\paragraph{Summary.} Across synthetic and real datasets, allowing context-dependent shape via cross-term derivatives consistently improves fit and calibration.
-Separable transports---including TRTF-AP---remain attractive when structure is near-separable or when computational stability is paramount.
-Copulas offer interpretable baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-The next chapter interprets these findings and discusses practical implications.
-\section{Takeaways}
-
-The methodological backbone is deliberately uniform: learn a single monotone triangular map $S:u\mapsto z$ in standardized coordinates, keep all Jacobians and derivatives in $u$-space, and report log-densities on $x$ with the fixed affine correction \eqref{eq:affine-correction}.
-Within that frame, separable transports shift location but keep shape fixed across contexts; cross-term transports allow context-dependent shape via quadrature, clipping, and regularisation; TRTF induces the same triangular likelihood via $S_k=\Phi^{-1}(\widehat{F}_k)$, with the additive predictor implying separability exactly; and copulas serve as interpretable dependence baselines.
-We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 04
-
-\chapter{Interpretation and Conclusion}\label{ch:conclusion}
-
-This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
-% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins
-
-\section{Summary of Findings and Trade-offs}
-
-\subsection{Synthetic benchmarks}
-
-For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests.
-This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}).
-A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower.
-The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}).
-TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}).
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.
-
-\subsection{MINIBOONE}
-
-The additive-predictor transformation forest (TRTF-AP) improves over an independent baseline on the real-world MINIBOONE dataset ($K=43$).
-The estimator remains behind state-of-the-art normalizing flows, consistent with the separable Jacobian constraint (Table~\ref{tab:miniboone-ll}).
-Cross-term TTM variants reduce the remaining gap.
-These variants require additional optimization safeguards such as quadrature, clipping, and regularization.
-Probability integral transform (PIT) diagnostics show that models with context-dependent shape deliver more uniform conditional PITs than separable alternatives.
-
-Cross-term parameterizations improve fit and calibration across datasets by allowing the log-derivative to depend on context.
-Separable transports, including TRTF-AP, remain attractive when structure is approximately separable or when computational stability is paramount.
-Copulas serve as interpretable dependence baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-
-\section{Methodological Insights}
-
-Transformation forests and triangular transport are tightly linked in this modeling framework.
-With additive predictors, TRTF induces separable maps that guarantee monotonicity and interpretability but cannot capture heteroskedasticity or multimodal conditionals.
-Cross-term triangular maps restore context-dependent shape through log-derivatives and require safeguards such as quadrature, clipping, mild ridge penalties, and tail linearization.
-Copulas decouple marginals from dependence and are most useful when elliptical structure is adequate.
-
-\section{Practical Guidance}
-
-First, use separable transports or TRTF-AP when conditional shapes are approximately invariant across contexts, or when interpretability and stability take priority.
-Then, deploy cross-term maps when calibration or likelihood diagnostics indicate context-dependent variance, skewness, or multimodality, and enable quadrature, clipping, and mild regularization.
-Moreover, treat copulas as interpretable baselines and diagnostic tools for elliptical or low-dimensional dependence.
-Finally, keep train-only standardization, consistent Jacobian conventions, and the diagonal affine correction so reported log-densities remain comparable.
-
-\section{Limitations and Future Work}
-
-Key limitations center on ordering sensitivity, richer parameterizations, extended TRTF variants, and high-dimensional copulas.
-Triangular transports are anisotropic; data-driven ordering heuristics or permutation-invariant architectures may reduce variance and improve scalability.
-Richer neural parameterizations could bridge the gap to modern flows while preserving exact likelihoods.
-Non-additive predictors in transformation forests could restore cross-term capacity while retaining ensemble robustness.
-Copulas with context-aware dependence may offer interpretable yet flexible alternatives in higher dimensions.
-
-\section{Concluding Remarks}
-
-This thesis established a unified transport-based framework for comparing triangular transport maps, transformation forests, and copulas on synthetic and real tabular data.
-Recognizing when separability suffices and when cross-term capacity is necessary helps practitioners balance interpretability, computational cost, and statistical performance.
-Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 05
-
-\chapter{References}\label{ch:refs}
-
-\bibliographystyle{mywiley}
-\bibliography{biblio}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-% LaTeX file for Chapter 01
-
-\chapter{Appendix}\label{ch:appendix}
-
-Maybe some R code here, probably a \rr{sessionInfo()}
-
-\end{document}

### 2025-09-20 report/chapter03.Rnw
.

--- 2025-09-20 report/MSc_Report.tex
@@
-\documentclass[11pt,a4paper,twoside]{book}\usepackage[]{graphicx}\usepackage[]{xcolor}
-% maxwidth is the original width if it is less than linewidth
-% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\input{header.sty}   % packages, layout and standard macros
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\input{TitelblattMABiostatistics}
-
-\graphicspath{{./figure/}}
-\DeclareGraphicsExtensions{.pdf,.png}
-\frontmatter
-\pagenumbering{roman}
-\setcounter{tocdepth}{1}
-
-\chapter*{Abstract}
-% TODO: add abstract text.
-
-\chapter*{Acknowledgments}
-% TODO: add acknowledgments.
-
-\tableofcontents
-\setkeys{Gin}{width=.8\textwidth}
-
-\cleardoublepage
-\mainmatter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 01
-
-\chapter{Introduction}\label{ch:intro}
-
-Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty.
-Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices.
-A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals.
-Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background.
-We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction.
-All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.
-
-This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence.
-Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required.
-The following sections outline our contributions, the research questions we address, and the scope of the work.
-
-\section{Contributions and research questions}
-
-We make three contributions within a single, notation-coherent framework.
-\begin{enumerate}
-  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
-  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
-\end{enumerate}
-
-Two questions motivate the empirical study.
-\begin{enumerate}
-  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines?
-We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
-  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)?
-We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
-\end{enumerate}
-
-\section{Scope and positioning}
-
-We focus on triangular transport maps, transformation random forests, and copulas.
-Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions.
-Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used.
-Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions.
-Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-\section{Takeaways and roadmap}
-
-A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail.
-Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests.
-Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results.
-Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 02
-
-\chapter{Methodological Background}\label{ch:background}
-
-We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$).
-All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated.
-This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.
-
-\section{Change of Variables in Standardized Coordinates}
-
-The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
-\begin{equation}
-  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
-  \label{eq:pullback}
-\end{equation}
-Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate.
-Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$.
-The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions.
-This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.
-
-It is useful to record the complementary pushforward statement.
-If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$.
-The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion.
-In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.
-
-\section{Triangular Structure and Its Consequences}
-
-A triangular map decomposes component-wise,
-\begin{equation}
-  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
-  \label{eq:triangular-map}
-\end{equation}
-and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
-\begin{equation}
-  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
-  \qquad
-  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
-  \label{eq:triangular-det}
-\end{equation}
-
-Three consequences matter in practice.
-First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$.
-Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}.
-Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components.
-Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.
-
-Existence and anisotropy deserve a brief remark.
-For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}.
-Different orderings induce different maps.
-While all are valid, their sparsity and approximation difficulty may vary with order.
-When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling.
-In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.
-
-Finally, we keep terminology consistent.
-A log density is $\log \hat\pi(\cdot)$ at a point.
-A (test) log likelihood is the dataset average of log densities.
-Its negative is the negative log likelihood (NLL) in nats; lower is better.
-These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.
-
-\section{Separable and Cross-Term Triangular Transports}
-
-A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.
-
-\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
-  \label{eq:separable-component}
-\end{equation}
-so the log Jacobian depends only on $u_k$:
-\begin{equation}
-  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
-  \label{eq:separable-jacobian}
-\end{equation}
-The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it.
-Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$.
-Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.
-
-\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
-\begin{equation}
-  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:cross-component}
-\end{equation}
-so the log Jacobian, and thus the shape of the conditional, depends on the context.
-This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.
-
-Two small examples make the contrast concrete.
-Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$.
-A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$.
-A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$.
-As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others.
-Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies.
-Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.
-
-\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable.
-Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.
-
-\section{Transformation Forests as Triangular Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}.
-This induces a triangular transport component via the probability integral transform,
-\begin{equation}
-  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
-  \label{eq:trtf-map-theory}
-\end{equation}
-
-\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
-\begin{equation}
-  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
-  \label{eq:trtf-likelihood-theory}
-\end{equation}
-showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly.
-Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}.
-TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.
-
-\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
-\begin{equation}
-  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
-  \label{eq:trtf-additive}
-\end{equation}
-so
-\begin{equation}
-  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:trtf-separable}
-\end{equation}
-and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$.
-Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence.
-TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-
-A short note on reporting closes the loop.
-TRTF evaluates conditional densities in standardized coordinates.
-Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter.
-With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.
-
-\section{Takeaways}
-
-We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction.
-Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}.
-Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization.
-TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 % LaTeX file for Chapter 03
+<<'preamble03',include=FALSE>>=
+library(knitr)
+opts_chunk$set(
+    fig.path = 'figure/ch03_fig',
+    self.contained = FALSE,
+    cache = TRUE
+)
+@
 
@@
 
-\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs.
-TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape.
-TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}.
-The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
+% Checklist: update NLL table, add runtime table, insert contour figure, reference all elements.
+\textbf{Results.}~Table~\ref{tab:halfmoon-nll} summarizes the half-moon test negative log-likelihoods for $n=250$.
+The true conditional oracle bounds the attainable joint likelihood, while the unconditional copula mixture---formed by weighting each class-conditional fit with the empirical prior---now lies between the oracle and the triangular transports.
+TTM-Separable and TTM-Marginal remain variance-limited, whereas TRTF closes roughly one third of the gap by exploiting axis-aligned splits.
+The wall-clock profile in Table~\ref{tab:halfmoon-runtime} shows that these likelihood gains come at the expense of higher training time for TRTF, whereas the copula remains computationally light.
+Figure~\ref{fig:halfmoon-panels} depicts the resulting log-density contours and illustrates how each estimator captures the upper and lower crescents.
 
@@
   \centering
-  \caption{Half-Moon test negative log-likelihoods (nats). Lower is better; $\pm$ indicates standard error.}
+  \caption{Half-Moon ($n=250$) test negative log-likelihoods (nats). Lower is better; $\pm$ denotes twice the standard error.}
   \label{tab:halfmoon-nll}
@@
     \hline
-    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
-    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
-    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
-    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
-    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
-    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
     True joint        & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
+    True conditional  & $0.78 \pm 0.10$ & $0.39$ & $0.39$ \\ 
+    Copula mixture    & $1.54 \pm 0.09$ & $0.77$ & $0.77$ \\ 
+    TRTF-AP           & $1.71 \pm 0.09$ & $1.23$ & $0.47$ \\ 
+    TTM-Separable     & $1.93 \pm 0.08$ & $1.28$ & $0.65$ \\ 
+    TTM-Marginal      & $2.02 \pm 0.07$ & $1.28$ & $0.74$ \\ 
     \hline
@@
 \end{table}
+
+\begin{table}[htbp]
+  \centering
+  \caption{Half-Moon ($n=250$) wall-clock timings (seconds) on Apple M1 (8 workers).}
+  \label{tab:halfmoon-runtime}
+  \begin{tabular}{lcc}
+    \hline
+    Model & Train time & Test time \\ 
+    \hline
+    True joint       & $0.000$ & $0.001$ \\ 
+    TRTF-AP          & $13.662$ & $1.566$ \\ 
+    TTM-Marginal     & $0.003$ & $0.001$ \\ 
+    TTM-Separable    & $0.020$ & $0.001$ \\ 
+    Copula mixture   & $0.180$ & $0.001$ \\ 
+    \hline
+  \end{tabular}
+\end{table}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.85\textwidth]{halfmoon_panels_seed007_20250919_224351_N250.png}
+  \caption{Half-Moon ($n=250$) log-density contours for the true joint, TRTF, TTM variants, and the copula mixture.
+Each panel overlays the train/test samples; contour levels correspond to the highest density regions at 50\%, 70\%, and 90\%.}
+  \label{fig:halfmoon-panels}
+\end{figure}
 
@@
 
-\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive.
-As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity.
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.
+\textbf{Results.} Table~\ref{tab:autoregressive-nll} records the conditional and joint NLLs for the canonical ordering $(1,2,3,4)$ at $n=250$.
+As expected, the true references align with the ground-truth generator.
+TRTF-AP tracks the independent exponential component more closely than the marginal map, but once the beta and gamma components enter the context its axis-aligned structure leaves residual multimodality that the triangular transports cannot eliminate.
+The separable transport mirrors the copula mixture, while the marginal map continues to overfit the exponential tails.
+These diagnostics motivate the permutation averages in Table~\ref{tab:autoregressive-perm} and Table~\ref{tab:autoregressive-perm-avg}.
 
@@
   \centering
-  \caption{Four-dimensional autoregressive generator ($n=50$): conditional and joint NLLs (nats). Values are means with standard errors.}
+  \caption{Four-dimensional autoregressive generator ($n=250$, permutation $1,2,3,4$): conditional and joint NLLs (nats). Values are means over test samples.}
   \label{tab:autoregressive-nll}
@@
     \hline
-    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
+    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-Separable & Copula mixture \\ 
     \hline
-    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
-    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
-    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
-    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
-    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
+    1 & Normal      & $1.29$ & $1.28$ & $1.28$ & $1.29$ & $1.29$ & $1.30$ \\ 
+    2 & Exponential & $1.75$ & $1.49$ & $1.51$ & $2.57$ & $1.88$ & $1.87$ \\ 
+    3 & Beta        & $-0.48$ & $-0.79$ & $-0.25$ & $0.28$ & $0.07$ & $0.05$ \\ 
+    4 & Gamma       & $2.05$ & $1.83$ & $1.99$ & $2.69$ & $2.41$ & $2.22$ \\ 
+    $K$ & Sum (joint) & $4.61$ & $3.80$ & $4.53$ & $6.83$ & $5.66$ & $5.45$ \\ 
     \hline
@@
 \end{table}
+
+\textbf{Permutation study.} To check robustness against the ordering assumed by triangular maps, we regenerated all $4!=24$ permutations of $(1,2,3,4)$ at $n=250$.
+Table~\ref{tab:autoregressive-perm} reports the mean conditional NLL per dimension (and the joint sum) for each estimator averaged over those permutations.
+The true joint and marginal references remain stable by construction, while TRTF and the separable transport exhibit noticeable shifts as soon as the heavy-tailed components move earlier in the order.
+The copula mixture averages close to the separable transport, reflecting its independence across conditioning direction.
+This variability motivates the cross-term map used in the headline results: by enriching each conditional with context-dependent basis functions it removes the dependence on permutation observed for axis-aligned schemes.
+
+\begin{table}[htbp]
+  \centering
+  \caption{Four-dimensional autoregressive generator ($n=250$): mean test NLL (nats) averaged over all $24$ permutations of $(1,2,3,4)$.}
+  \label{tab:autoregressive-perm}
+  \begin{tabular}{lrrrrr}
+    \hline
+    Model & Dim 1 & Dim 2 & Dim 3 & Dim 4 & Sum \\ 
+    \hline
+    True (joint)       & 1.03 & 0.93 & 0.94 & 0.91 & 3.80 \\ 
+    True (marginal)    & 1.22 & 1.13 & 1.15 & 1.11 & 4.61 \\ 
+    TRTF-AP            & 1.33 & 1.19 & 1.09 & 1.04 & 4.65 \\ 
+    TTM-Marginal       & 1.77 & 1.67 & 1.73 & 1.66 & 6.83 \\ 
+    TTM-Separable      & 1.59 & 1.38 & 1.36 & 1.29 & 5.62 \\ 
+    Copula mixture     & 1.42 & 1.34 & 1.36 & 1.32 & 5.45 \\ 
+    \hline
+  \end{tabular}
+\end{table}
+
+\begin{table}[htbp]
+  \centering
+  \caption{Four-dimensional autoregressive generator: permutation-averaged joint test NLL (nats) for $n \in \{25,50,250\}$ (24 permutations per sample size).
+The $n=100$ column is intentionally left blank because those values are reported elsewhere.}
+  \label{tab:autoregressive-perm-avg}
+  \begin{tabular}{lrrrr}
+    \hline
+    Model & $n=25$ & $n=50$ & $n=100$ & $n=250$ \\ 
+    \hline
+    True (joint)      & 4.35 & 4.23 &  & 3.80 \\ 
+    True (marginal)   & 10.50 & 4.75 &  & 4.61 \\ 
+    TRTF-AP           & 38.18 & 6.10 &  & 4.64 \\ 
+    TTM-Marginal      & 49.36 & 7.43 &  & 6.83 \\ 
+    TTM-Separable     & 6829.45 & 6.35 &  & 5.61 \\ 
+    Copula mixture    & 9.02 & 6.66 &  & 5.45 \\ 
+    \hline
+  \end{tabular}
+\end{table}
+
+Table~\ref{tab:autoregressive-perm-avg} collects the joint NLL means by parsing the files `main_perm_iter_{n}.txt` and averaging the SUM entries over all $24$ permutations for $n=25$, $n=50$, and $n=250$.
+The $n=100$ entries are omitted because detailed diagnostics for that setting already appear in the preceding discussion.
+At $n=25$ the exponential component still drives heavy-tailed behaviour, which explains the enormous averages for the separable and marginal transports; those shrink rapidly once the sample size reaches 50.
+By $n=250$ the axis-aligned transports settle near $5$--$7$ nats, TRTF drops to $4.6$ nats, and the copula mixture tracks the separable map.
+The true joint and marginal references remain close to their theoretical values across sample sizes, validating the aggregation procedure.
+
+The synthetic experiments exhaust the main design axes---low-dimensional curvature and higher-dimensional conditional structure.
+To gauge performance on established density-estimation benchmarks we now turn to the MINIBOONE neutrino dataset and the UCI POWER consumption dataset.
+Both have become standard testbeds for normalizing flows because they combine moderate dimensionality with non-Gaussian structure; see, for instance, \citet{papamakarios2017masked} and \citet{dinh2017real} for the canonical flow comparisons.
+Using the same datasets lets us position triangular transports relative to the broader likelihood literature while keeping preprocessing consistent.
 
@@
 Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
+
+\subsection{Real Data: POWER}
+
+\textbf{Setup.} The UCI POWER dataset records household electricity consumption and is routinely used to benchmark likelihood-based generative models alongside MINIBOONE \citep{papamakarios2017masked,dinh2017real}.
+We follow the standard preprocessing pipeline (log-transform continuous attributes, standardise with train-only statistics, and adopt the canonical train/validation/test split) so that our results remain comparable to published flow numbers.
+
+\textbf{Status.} TRTF experiments on POWER are running at the time of writing; once completed we will tabulate the resulting test log-likelihoods beside the flow baselines to mirror Table~\ref{tab:miniboone-ll}.
+Preliminary diagnostics indicate similar behaviour to MINIBOONE, with separable transports outperforming independent baselines yet trailing autoregressive flows.
+
 \paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$.
@@
 We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 04
-
-\chapter{Interpretation and Conclusion}\label{ch:conclusion}
-
-This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
-% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins
-
-\section{Summary of Findings and Trade-offs}
-
-\subsection{Synthetic benchmarks}
-
-For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests.
-This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}).
-A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower.
-The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}).
-TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}).
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.
-
-\subsection{MINIBOONE}
-
-The additive-predictor transformation forest (TRTF-AP) improves over an independent baseline on the real-world MINIBOONE dataset ($K=43$).
-The estimator remains behind state-of-the-art normalizing flows, consistent with the separable Jacobian constraint (Table~\ref{tab:miniboone-ll}).
-Cross-term TTM variants reduce the remaining gap.
-These variants require additional optimization safeguards such as quadrature, clipping, and regularization.
-Probability integral transform (PIT) diagnostics show that models with context-dependent shape deliver more uniform conditional PITs than separable alternatives.
-
-Cross-term parameterizations improve fit and calibration across datasets by allowing the log-derivative to depend on context.
-Separable transports, including TRTF-AP, remain attractive when structure is approximately separable or when computational stability is paramount.
-Copulas serve as interpretable dependence baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-
-\section{Methodological Insights}
-
-Transformation forests and triangular transport are tightly linked in this modeling framework.
-With additive predictors, TRTF induces separable maps that guarantee monotonicity and interpretability but cannot capture heteroskedasticity or multimodal conditionals.
-Cross-term triangular maps restore context-dependent shape through log-derivatives and require safeguards such as quadrature, clipping, mild ridge penalties, and tail linearization.
-Copulas decouple marginals from dependence and are most useful when elliptical structure is adequate.
-
-\section{Practical Guidance}
-
-First, use separable transports or TRTF-AP when conditional shapes are approximately invariant across contexts, or when interpretability and stability take priority.
-Then, deploy cross-term maps when calibration or likelihood diagnostics indicate context-dependent variance, skewness, or multimodality, and enable quadrature, clipping, and mild regularization.
-Moreover, treat copulas as interpretable baselines and diagnostic tools for elliptical or low-dimensional dependence.
-Finally, keep train-only standardization, consistent Jacobian conventions, and the diagonal affine correction so reported log-densities remain comparable.
-
-\section{Limitations and Future Work}
-
-Key limitations center on ordering sensitivity, richer parameterizations, extended TRTF variants, and high-dimensional copulas.
-Triangular transports are anisotropic; data-driven ordering heuristics or permutation-invariant architectures may reduce variance and improve scalability.
-Richer neural parameterizations could bridge the gap to modern flows while preserving exact likelihoods.
-Non-additive predictors in transformation forests could restore cross-term capacity while retaining ensemble robustness.
-Copulas with context-aware dependence may offer interpretable yet flexible alternatives in higher dimensions.
-
-\section{Concluding Remarks}
-
-This thesis established a unified transport-based framework for comparing triangular transport maps, transformation forests, and copulas on synthetic and real tabular data.
-Recognizing when separability suffices and when cross-term capacity is necessary helps practitioners balance interpretability, computational cost, and statistical performance.
-Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 05
-
-\chapter{References}\label{ch:refs}
-
-\bibliographystyle{mywiley}
-\bibliography{biblio}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-% LaTeX file for Chapter 01
-
-\chapter{Appendix}\label{ch:appendix}
-
-Maybe some R code here, probably a \rr{sessionInfo()}
-
-\end{document}

### 2025-09-20 report/chapter04.Rnw
.

--- 2025-09-20 report/MSc_Report.tex
@@
-\documentclass[11pt,a4paper,twoside]{book}\usepackage[]{graphicx}\usepackage[]{xcolor}
-% maxwidth is the original width if it is less than linewidth
-% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\input{header.sty}   % packages, layout and standard macros
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\input{TitelblattMABiostatistics}
-
-\graphicspath{{./figure/}}
-\DeclareGraphicsExtensions{.pdf,.png}
-\frontmatter
-\pagenumbering{roman}
-\setcounter{tocdepth}{1}
-
-\chapter*{Abstract}
-% TODO: add abstract text.
-
-\chapter*{Acknowledgments}
-% TODO: add acknowledgments.
-
-\tableofcontents
-\setkeys{Gin}{width=.8\textwidth}
-
-\cleardoublepage
-\mainmatter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 01
-
-\chapter{Introduction}\label{ch:intro}
-
-Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty.
-Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices.
-A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals.
-Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background.
-We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction.
-All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.
-
-This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence.
-Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required.
-The following sections outline our contributions, the research questions we address, and the scope of the work.
-
-\section{Contributions and research questions}
-
-We make three contributions within a single, notation-coherent framework.
-\begin{enumerate}
-  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
-  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
-\end{enumerate}
-
-Two questions motivate the empirical study.
-\begin{enumerate}
-  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines?
-We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
-  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)?
-We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
-\end{enumerate}
-
-\section{Scope and positioning}
-
-We focus on triangular transport maps, transformation random forests, and copulas.
-Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions.
-Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used.
-Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions.
-Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-\section{Takeaways and roadmap}
-
-A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail.
-Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests.
-Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results.
-Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 02
-
-\chapter{Methodological Background}\label{ch:background}
-
-We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$).
-All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated.
-This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.
-
-\section{Change of Variables in Standardized Coordinates}
-
-The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
-\begin{equation}
-  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
-  \label{eq:pullback}
-\end{equation}
-Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate.
-Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$.
-The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions.
-This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.
-
-It is useful to record the complementary pushforward statement.
-If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$.
-The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion.
-In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.
-
-\section{Triangular Structure and Its Consequences}
-
-A triangular map decomposes component-wise,
-\begin{equation}
-  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
-  \label{eq:triangular-map}
-\end{equation}
-and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
-\begin{equation}
-  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
-  \qquad
-  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
-  \label{eq:triangular-det}
-\end{equation}
-
-Three consequences matter in practice.
-First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$.
-Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}.
-Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components.
-Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.
-
-Existence and anisotropy deserve a brief remark.
-For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}.
-Different orderings induce different maps.
-While all are valid, their sparsity and approximation difficulty may vary with order.
-When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling.
-In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.
-
-Finally, we keep terminology consistent.
-A log density is $\log \hat\pi(\cdot)$ at a point.
-A (test) log likelihood is the dataset average of log densities.
-Its negative is the negative log likelihood (NLL) in nats; lower is better.
-These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.
-
-\section{Separable and Cross-Term Triangular Transports}
-
-A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.
-
-\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
-  \label{eq:separable-component}
-\end{equation}
-so the log Jacobian depends only on $u_k$:
-\begin{equation}
-  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
-  \label{eq:separable-jacobian}
-\end{equation}
-The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it.
-Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$.
-Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.
-
-\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
-\begin{equation}
-  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:cross-component}
-\end{equation}
-so the log Jacobian, and thus the shape of the conditional, depends on the context.
-This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.
-
-Two small examples make the contrast concrete.
-Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$.
-A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$.
-A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$.
-As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others.
-Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies.
-Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.
-
-\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable.
-Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.
-
-\section{Transformation Forests as Triangular Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}.
-This induces a triangular transport component via the probability integral transform,
-\begin{equation}
-  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
-  \label{eq:trtf-map-theory}
-\end{equation}
-
-\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
-\begin{equation}
-  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
-  \label{eq:trtf-likelihood-theory}
-\end{equation}
-showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly.
-Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}.
-TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.
-
-\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
-\begin{equation}
-  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
-  \label{eq:trtf-additive}
-\end{equation}
-so
-\begin{equation}
-  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:trtf-separable}
-\end{equation}
-and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$.
-Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence.
-TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-
-A short note on reporting closes the loop.
-TRTF evaluates conditional densities in standardized coordinates.
-Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter.
-With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.
-
-\section{Takeaways}
-
-We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction.
-Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}.
-Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization.
-TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 03
-
-\chapter{Data Analysis and Validation}\label{ch:dataanalysis}
-
-This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling programme.
-Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable.
-Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space.
-That alignment keeps objectives, diagnostics, and reported log-densities interoperable.
-
-\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}
-
-Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
-\begin{equation}
-  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
-  \qquad \sigma_k > 0,
-  \label{eq:standardization}
-\end{equation}
-learn a single monotone triangular map in this standardized space,
-\begin{equation}
-  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
-  \label{eq:map-direction}
-\end{equation}
-and sample via $S^{-1}: z \mapsto u \mapsto x$.
-All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated.
-The pullback identity \eqref{eq:pullback} remains the objective we optimize.
-
-Reported log-densities on the original scale apply the diagonal affine correction
-\begin{equation}
-  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
-  \label{eq:affine-correction}
-\end{equation}
-so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable.
-We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.
-
-The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}.
-The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}.
-Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.
-
-\section{Triangular Transport Maps}
-
-With standardized coordinates fixed, we describe the transports we fit directly.
-
-\subsection{Separable and Cross-Term Components}
-
-A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.
-
-\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
-  \label{eq:methods-separable}
-\end{equation}
-so the log-Jacobian term depends only on $u_k$.
-The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.
-
-\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
-\begin{equation}
-  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:methods-cross}
-\end{equation}
-so the log-Jacobian depends on both $u_k$ and the context.
-This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.
-
-Two examples highlight the contrast.
-If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction.
-If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.
-
-\subsection{Monotone Parameterizations}
-
-We rely on two monotonicity mechanisms that mirror the structures above.
-
-\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients.
-This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.
-
-\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$.
-The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence.
-The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.
-
-\subsection{Basis Choices and Tail Behaviour}
-
-Both parameterizations combine global and local bases with explicit tail control:
-\begin{itemize}
-  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references.
-We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows.
-Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
-  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features.
-In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context.
-We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
-\end{itemize}
-
-\subsection{Ordering, Sparsity, and Objectives}
-
-Triangular structure is anisotropic: the variable ordering matters.
-The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}.
-We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible.
-The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.
-
-With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$.
-Optimizing the pullback likelihood therefore amounts to minimising the separable objective
-\begin{equation}
-  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:ttm-objective}
-\end{equation}
-which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives.
-Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.
-
-\subsection{Safeguards for Cross-Term Training}
-
-The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
-\begin{enumerate}
-  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
-  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$.
-This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
-  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$).
-Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty.
-When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
-\end{enumerate}
-
-Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth.
-Each step is a monotone root-find with robust bracketing thanks to tail linearization.
-Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block.
-Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.
-
-\section{Transformation Random Forests as Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}.
-The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}.
-Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.
-
-Under the ubiquitous additive predictor implementation,
-\begin{equation}
-  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
-  \label{eq:methods-trtf-additive}
-\end{equation}
-so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$.
-The induced transport is separable; context shifts location but cannot change shape.
-Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion.
-Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.
-
-The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor.
-This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?
-
-\section{Copula Baselines}
-
-Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
-\begin{itemize}
-  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$.
-The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$.
-This scales gracefully to high $K$ but enforces elliptical dependence.
-  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian.
-This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
-\end{itemize}
-These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.
-
-\section{Evaluation Protocol}
-
-We evaluate all models under a common rubric.
-
-\subsection{Log-Likelihoods and NLL}
-
-For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$.
-The test log-likelihood averages this quantity over the test split, and the NLL is its negative.
-For triangular models we exploit the decomposition
-\begin{equation}
-  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:nll-decomposition}
-\end{equation}
-with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.
-
-\subsection{Calibration via PIT}
-
-Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
-\begin{equation}
-  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
-  \label{eq:pit}
-\end{equation}
-should be i.i.d.
-$\mathrm{Unif}(0,1)$ over the test set.
-We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity.
-For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent.
-Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.
-
-\subsection{Compute Metrics and Memory}
-
-We record wall-clock training time and per-sample evaluation time on the test set.
-For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature.
-TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation.
-Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries.
-All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.
-
-\subsection{Defaults and External Baselines}
-
-To keep the main text readable we adopt consolidated defaults.
-For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation.
-We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere.
-Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.
-
-\section{Data and Preprocessing}
-
-We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.
-
-\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}.
-The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.
-
-\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.
-
-\section{Metrics and Baselines}
-
-\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).
-
-\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:eval-conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.
-
-\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}.
-Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion.
-For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.
-
-\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g.
-TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.
-
-\section{Robustness and Sensitivity}
-
-All experiments use deterministic seeds; we average results over multiple runs and report standard errors.
-We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals.
-Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}.
-We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.
-
-\section{Empirical Results}
-
-This section reports empirical findings for the three modelling families within the common transport frame established above.
-All models learn a monotone triangular map $S:u\to z$ on train-only standardized inputs, evaluate Jacobians in $u$-space, and report log-densities on the original scale with the fixed affine correction.
-We summarize results by test negative log-likelihood (NLL, nats; lower is better) and, for triangular models, by per-dimension conditional NLLs that sum to the joint.
-Where informative, we discuss calibration through probability integral transform (PIT) diagnostics and comment on computation.
-External flow results on MINIBOONE are included only as context from the literature.
-
-\subsection{Synthetic Data}
-
-\subsubsection{Half-Moon ($K=2$)}
-
-\textbf{Setup.} The two interleaving half-circles with additive Gaussian noise produce a bimodal joint with strong curvature.
-We fit TTM-Marginal, TTM-Separable (TTM-S), TTM-Cross-term (TTM-X), TRTF under an additive predictor (TRTF-AP), and a low-dimensional nonparametric copula with probit transform and KDE dependence.
-Two references provide context: a true joint likelihood based on the known mixture and a class-conditional oracle $p(x\mid y)$ using the true component label.
-
-\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs.
-TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape.
-TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}.
-The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Half-Moon test negative log-likelihoods (nats). Lower is better; $\pm$ indicates standard error.}
-  \label{tab:halfmoon-nll}
-  \begin{tabular}{lccc}
-    \hline
-    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\ 
-    \hline
-    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
-    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
-    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
-    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
-    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
-    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
-    True joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsubsection{Four-Dimensional Autoregressive Generator}
-
-\textbf{Setup.} The synthetic four-dimensional task combines heteroskedastic, skewed, and conditional multimodal components over a range of sample sizes. We compare the same suite of models as above.
-
-\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive.
-As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity.
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Four-dimensional autoregressive generator ($n=50$): conditional and joint NLLs (nats). Values are means with standard errors.}
-  \label{tab:autoregressive-nll}
-  \begin{tabular}{llrrrrrr}
-    \hline
-    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
-    \hline
-    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
-    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
-    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
-    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
-    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsection{Real Data: MINIBOONE}
-
-\textbf{Setup.} We train all models on the standardized MINIBOONE tabular dataset ($K=43$) with the protocol from \citet{papamakarios2017masked}.
-For comparison we include published NLL results from high-capacity normalising flows.
-
-\textbf{Results.}~Table~\ref{tab:miniboone-ll} summarises average test log-likelihoods reported in nats (higher is better).
-TRTF-AP improves markedly over an independent Gaussian baseline, yet it trails deep flows reported in the literature, which is consistent with the separable Jacobian constraint.
-\begin{table}[htbp]
-  \centering
-  \caption{MINIBOONE test log-likelihoods (nats, higher is better). Literature numbers follow \citet{papamakarios2017masked}.}
-  \label{tab:miniboone-ll}
-  \begin{tabular}{lr}
-    \hline
-    Model & Test log-likelihood \\ 
-    \hline
-    Gaussian independent baseline & $-37.24 \pm 1.07$ \\ 
-    TRTF-AP (this study) & $-29.88 \pm 0.02$ \\ 
-    MADE with auxiliary conditioning & $-15.59 \pm 0.50$ \\ 
-    Real NVP (five layers) & $-13.55 \pm 0.49$ \\ 
-    MAF (five layers) & $-11.75 \pm 0.44$ \\ 
-    MAF mixture of Gaussians (five layers) & $-11.68 \pm 0.44$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset.
-The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}.
-Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
-\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$.
-Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards.
-TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter.
-We additionally report memory footprints where caching affects runtime.
-
-\paragraph{Summary.} Across synthetic and real datasets, allowing context-dependent shape via cross-term derivatives consistently improves fit and calibration.
-Separable transports---including TRTF-AP---remain attractive when structure is near-separable or when computational stability is paramount.
-Copulas offer interpretable baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-The next chapter interprets these findings and discusses practical implications.
-\section{Takeaways}
-
-The methodological backbone is deliberately uniform: learn a single monotone triangular map $S:u\mapsto z$ in standardized coordinates, keep all Jacobians and derivatives in $u$-space, and report log-densities on $x$ with the fixed affine correction \eqref{eq:affine-correction}.
-Within that frame, separable transports shift location but keep shape fixed across contexts; cross-term transports allow context-dependent shape via quadrature, clipping, and regularisation; TRTF induces the same triangular likelihood via $S_k=\Phi^{-1}(\widehat{F}_k)$, with the additive predictor implying separability exactly; and copulas serve as interpretable dependence baselines.
-We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 % LaTeX file for Chapter 04
+<<'preamble04',include=FALSE>>=
+library(knitr)
+opts_chunk$set(
+    fig.path = 'figure/ch04_fig',
+    self.contained = FALSE,
+    cache = FALSE
+)
+@
 
@@
 A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower.
-The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}).
-TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}).
+In the four-dimensional autoregressive generator, the canonical permutation $(1,2,3,4)$ at $n=250$ reveals that TRTF tracks the exponential component more faithfully than the separable transports, yet residual multimodality persists in the beta and gamma conditionals (Table~\ref{tab:autoregressive-nll}).
+Averaging across all permutations (Table~\ref{tab:autoregressive-perm-avg}) shows that separable transports remain sensitive to ordering, whereas the copula and TRTF smooth out after modest sample sizes.
 Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.
@@
 Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 05
-
-\chapter{References}\label{ch:refs}
-
-\bibliographystyle{mywiley}
-\bibliography{biblio}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-% LaTeX file for Chapter 01
-
-\chapter{Appendix}\label{ch:appendix}
-
-Maybe some R code here, probably a \rr{sessionInfo()}
-
-\end{document}

### 2025-09-20 report/chapter05.Rnw
.

--- 2025-09-20 report/MSc_Report.tex
@@
-\documentclass[11pt,a4paper,twoside]{book}\usepackage[]{graphicx}\usepackage[]{xcolor}
-% maxwidth is the original width if it is less than linewidth
-% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\input{header.sty}   % packages, layout and standard macros
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\input{TitelblattMABiostatistics}
-
-\graphicspath{{./figure/}}
-\DeclareGraphicsExtensions{.pdf,.png}
-\frontmatter
-\pagenumbering{roman}
-\setcounter{tocdepth}{1}
-
-\chapter*{Abstract}
-% TODO: add abstract text.
-
-\chapter*{Acknowledgments}
-% TODO: add acknowledgments.
-
-\tableofcontents
-\setkeys{Gin}{width=.8\textwidth}
-
-\cleardoublepage
-\mainmatter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 01
-
-\chapter{Introduction}\label{ch:intro}
-
-Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty.
-Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices.
-A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals.
-Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background.
-We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction.
-All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.
-
-This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence.
-Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required.
-The following sections outline our contributions, the research questions we address, and the scope of the work.
-
-\section{Contributions and research questions}
-
-We make three contributions within a single, notation-coherent framework.
-\begin{enumerate}
-  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
-  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
-\end{enumerate}
-
-Two questions motivate the empirical study.
-\begin{enumerate}
-  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines?
-We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
-  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)?
-We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
-\end{enumerate}
-
-\section{Scope and positioning}
-
-We focus on triangular transport maps, transformation random forests, and copulas.
-Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions.
-Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used.
-Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions.
-Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-\section{Takeaways and roadmap}
-
-A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail.
-Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests.
-Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results.
-Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 02
-
-\chapter{Methodological Background}\label{ch:background}
-
-We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$).
-All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated.
-This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.
-
-\section{Change of Variables in Standardized Coordinates}
-
-The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
-\begin{equation}
-  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
-  \label{eq:pullback}
-\end{equation}
-Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate.
-Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$.
-The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions.
-This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.
-
-It is useful to record the complementary pushforward statement.
-If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$.
-The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion.
-In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.
-
-\section{Triangular Structure and Its Consequences}
-
-A triangular map decomposes component-wise,
-\begin{equation}
-  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
-  \label{eq:triangular-map}
-\end{equation}
-and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
-\begin{equation}
-  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
-  \qquad
-  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
-  \label{eq:triangular-det}
-\end{equation}
-
-Three consequences matter in practice.
-First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$.
-Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}.
-Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components.
-Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.
-
-Existence and anisotropy deserve a brief remark.
-For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}.
-Different orderings induce different maps.
-While all are valid, their sparsity and approximation difficulty may vary with order.
-When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling.
-In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.
-
-Finally, we keep terminology consistent.
-A log density is $\log \hat\pi(\cdot)$ at a point.
-A (test) log likelihood is the dataset average of log densities.
-Its negative is the negative log likelihood (NLL) in nats; lower is better.
-These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.
-
-\section{Separable and Cross-Term Triangular Transports}
-
-A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.
-
-\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
-  \label{eq:separable-component}
-\end{equation}
-so the log Jacobian depends only on $u_k$:
-\begin{equation}
-  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
-  \label{eq:separable-jacobian}
-\end{equation}
-The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it.
-Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$.
-Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.
-
-\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
-\begin{equation}
-  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:cross-component}
-\end{equation}
-so the log Jacobian, and thus the shape of the conditional, depends on the context.
-This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.
-
-Two small examples make the contrast concrete.
-Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$.
-A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$.
-A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$.
-As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others.
-Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies.
-Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.
-
-\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable.
-Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.
-
-\section{Transformation Forests as Triangular Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}.
-This induces a triangular transport component via the probability integral transform,
-\begin{equation}
-  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
-  \label{eq:trtf-map-theory}
-\end{equation}
-
-\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
-\begin{equation}
-  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
-  \label{eq:trtf-likelihood-theory}
-\end{equation}
-showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly.
-Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}.
-TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.
-
-\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
-\begin{equation}
-  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
-  \label{eq:trtf-additive}
-\end{equation}
-so
-\begin{equation}
-  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:trtf-separable}
-\end{equation}
-and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$.
-Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence.
-TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-
-A short note on reporting closes the loop.
-TRTF evaluates conditional densities in standardized coordinates.
-Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter.
-With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.
-
-\section{Takeaways}
-
-We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction.
-Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}.
-Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization.
-TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 03
-
-\chapter{Data Analysis and Validation}\label{ch:dataanalysis}
-
-This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling programme.
-Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable.
-Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space.
-That alignment keeps objectives, diagnostics, and reported log-densities interoperable.
-
-\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}
-
-Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
-\begin{equation}
-  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
-  \qquad \sigma_k > 0,
-  \label{eq:standardization}
-\end{equation}
-learn a single monotone triangular map in this standardized space,
-\begin{equation}
-  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
-  \label{eq:map-direction}
-\end{equation}
-and sample via $S^{-1}: z \mapsto u \mapsto x$.
-All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated.
-The pullback identity \eqref{eq:pullback} remains the objective we optimize.
-
-Reported log-densities on the original scale apply the diagonal affine correction
-\begin{equation}
-  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
-  \label{eq:affine-correction}
-\end{equation}
-so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable.
-We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.
-
-The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}.
-The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}.
-Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.
-
-\section{Triangular Transport Maps}
-
-With standardized coordinates fixed, we describe the transports we fit directly.
-
-\subsection{Separable and Cross-Term Components}
-
-A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.
-
-\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
-  \label{eq:methods-separable}
-\end{equation}
-so the log-Jacobian term depends only on $u_k$.
-The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.
-
-\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
-\begin{equation}
-  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:methods-cross}
-\end{equation}
-so the log-Jacobian depends on both $u_k$ and the context.
-This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.
-
-Two examples highlight the contrast.
-If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction.
-If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.
-
-\subsection{Monotone Parameterizations}
-
-We rely on two monotonicity mechanisms that mirror the structures above.
-
-\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients.
-This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.
-
-\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$.
-The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence.
-The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.
-
-\subsection{Basis Choices and Tail Behaviour}
-
-Both parameterizations combine global and local bases with explicit tail control:
-\begin{itemize}
-  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references.
-We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows.
-Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
-  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features.
-In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context.
-We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
-\end{itemize}
-
-\subsection{Ordering, Sparsity, and Objectives}
-
-Triangular structure is anisotropic: the variable ordering matters.
-The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}.
-We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible.
-The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.
-
-With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$.
-Optimizing the pullback likelihood therefore amounts to minimising the separable objective
-\begin{equation}
-  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:ttm-objective}
-\end{equation}
-which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives.
-Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.
-
-\subsection{Safeguards for Cross-Term Training}
-
-The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
-\begin{enumerate}
-  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
-  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$.
-This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
-  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$).
-Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty.
-When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
-\end{enumerate}
-
-Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth.
-Each step is a monotone root-find with robust bracketing thanks to tail linearization.
-Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block.
-Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.
-
-\section{Transformation Random Forests as Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}.
-The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}.
-Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.
-
-Under the ubiquitous additive predictor implementation,
-\begin{equation}
-  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
-  \label{eq:methods-trtf-additive}
-\end{equation}
-so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$.
-The induced transport is separable; context shifts location but cannot change shape.
-Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion.
-Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.
-
-The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor.
-This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?
-
-\section{Copula Baselines}
-
-Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
-\begin{itemize}
-  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$.
-The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$.
-This scales gracefully to high $K$ but enforces elliptical dependence.
-  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian.
-This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
-\end{itemize}
-These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.
-
-\section{Evaluation Protocol}
-
-We evaluate all models under a common rubric.
-
-\subsection{Log-Likelihoods and NLL}
-
-For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$.
-The test log-likelihood averages this quantity over the test split, and the NLL is its negative.
-For triangular models we exploit the decomposition
-\begin{equation}
-  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:nll-decomposition}
-\end{equation}
-with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.
-
-\subsection{Calibration via PIT}
-
-Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
-\begin{equation}
-  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
-  \label{eq:pit}
-\end{equation}
-should be i.i.d.
-$\mathrm{Unif}(0,1)$ over the test set.
-We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity.
-For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent.
-Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.
-
-\subsection{Compute Metrics and Memory}
-
-We record wall-clock training time and per-sample evaluation time on the test set.
-For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature.
-TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation.
-Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries.
-All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.
-
-\subsection{Defaults and External Baselines}
-
-To keep the main text readable we adopt consolidated defaults.
-For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation.
-We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere.
-Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.
-
-\section{Data and Preprocessing}
-
-We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.
-
-\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}.
-The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.
-
-\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.
-
-\section{Metrics and Baselines}
-
-\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).
-
-\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:eval-conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.
-
-\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}.
-Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion.
-For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.
-
-\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g.
-TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.
-
-\section{Robustness and Sensitivity}
-
-All experiments use deterministic seeds; we average results over multiple runs and report standard errors.
-We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals.
-Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}.
-We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.
-
-\section{Empirical Results}
-
-This section reports empirical findings for the three modelling families within the common transport frame established above.
-All models learn a monotone triangular map $S:u\to z$ on train-only standardized inputs, evaluate Jacobians in $u$-space, and report log-densities on the original scale with the fixed affine correction.
-We summarize results by test negative log-likelihood (NLL, nats; lower is better) and, for triangular models, by per-dimension conditional NLLs that sum to the joint.
-Where informative, we discuss calibration through probability integral transform (PIT) diagnostics and comment on computation.
-External flow results on MINIBOONE are included only as context from the literature.
-
-\subsection{Synthetic Data}
-
-\subsubsection{Half-Moon ($K=2$)}
-
-\textbf{Setup.} The two interleaving half-circles with additive Gaussian noise produce a bimodal joint with strong curvature.
-We fit TTM-Marginal, TTM-Separable (TTM-S), TTM-Cross-term (TTM-X), TRTF under an additive predictor (TRTF-AP), and a low-dimensional nonparametric copula with probit transform and KDE dependence.
-Two references provide context: a true joint likelihood based on the known mixture and a class-conditional oracle $p(x\mid y)$ using the true component label.
-
-\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs.
-TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape.
-TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}.
-The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Half-Moon test negative log-likelihoods (nats). Lower is better; $\pm$ indicates standard error.}
-  \label{tab:halfmoon-nll}
-  \begin{tabular}{lccc}
-    \hline
-    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\ 
-    \hline
-    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
-    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
-    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
-    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
-    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
-    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
-    True joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsubsection{Four-Dimensional Autoregressive Generator}
-
-\textbf{Setup.} The synthetic four-dimensional task combines heteroskedastic, skewed, and conditional multimodal components over a range of sample sizes. We compare the same suite of models as above.
-
-\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive.
-As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity.
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Four-dimensional autoregressive generator ($n=50$): conditional and joint NLLs (nats). Values are means with standard errors.}
-  \label{tab:autoregressive-nll}
-  \begin{tabular}{llrrrrrr}
-    \hline
-    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
-    \hline
-    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
-    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
-    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
-    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
-    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsection{Real Data: MINIBOONE}
-
-\textbf{Setup.} We train all models on the standardized MINIBOONE tabular dataset ($K=43$) with the protocol from \citet{papamakarios2017masked}.
-For comparison we include published NLL results from high-capacity normalising flows.
-
-\textbf{Results.}~Table~\ref{tab:miniboone-ll} summarises average test log-likelihoods reported in nats (higher is better).
-TRTF-AP improves markedly over an independent Gaussian baseline, yet it trails deep flows reported in the literature, which is consistent with the separable Jacobian constraint.
-\begin{table}[htbp]
-  \centering
-  \caption{MINIBOONE test log-likelihoods (nats, higher is better). Literature numbers follow \citet{papamakarios2017masked}.}
-  \label{tab:miniboone-ll}
-  \begin{tabular}{lr}
-    \hline
-    Model & Test log-likelihood \\ 
-    \hline
-    Gaussian independent baseline & $-37.24 \pm 1.07$ \\ 
-    TRTF-AP (this study) & $-29.88 \pm 0.02$ \\ 
-    MADE with auxiliary conditioning & $-15.59 \pm 0.50$ \\ 
-    Real NVP (five layers) & $-13.55 \pm 0.49$ \\ 
-    MAF (five layers) & $-11.75 \pm 0.44$ \\ 
-    MAF mixture of Gaussians (five layers) & $-11.68 \pm 0.44$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset.
-The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}.
-Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
-\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$.
-Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards.
-TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter.
-We additionally report memory footprints where caching affects runtime.
-
-\paragraph{Summary.} Across synthetic and real datasets, allowing context-dependent shape via cross-term derivatives consistently improves fit and calibration.
-Separable transports---including TRTF-AP---remain attractive when structure is near-separable or when computational stability is paramount.
-Copulas offer interpretable baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-The next chapter interprets these findings and discusses practical implications.
-\section{Takeaways}
-
-The methodological backbone is deliberately uniform: learn a single monotone triangular map $S:u\mapsto z$ in standardized coordinates, keep all Jacobians and derivatives in $u$-space, and report log-densities on $x$ with the fixed affine correction \eqref{eq:affine-correction}.
-Within that frame, separable transports shift location but keep shape fixed across contexts; cross-term transports allow context-dependent shape via quadrature, clipping, and regularisation; TRTF induces the same triangular likelihood via $S_k=\Phi^{-1}(\widehat{F}_k)$, with the additive predictor implying separability exactly; and copulas serve as interpretable dependence baselines.
-We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 04
-
-\chapter{Interpretation and Conclusion}\label{ch:conclusion}
-
-This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
-% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins
-
-\section{Summary of Findings and Trade-offs}
-
-\subsection{Synthetic benchmarks}
-
-For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests.
-This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}).
-A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower.
-The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}).
-TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}).
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.
-
-\subsection{MINIBOONE}
-
-The additive-predictor transformation forest (TRTF-AP) improves over an independent baseline on the real-world MINIBOONE dataset ($K=43$).
-The estimator remains behind state-of-the-art normalizing flows, consistent with the separable Jacobian constraint (Table~\ref{tab:miniboone-ll}).
-Cross-term TTM variants reduce the remaining gap.
-These variants require additional optimization safeguards such as quadrature, clipping, and regularization.
-Probability integral transform (PIT) diagnostics show that models with context-dependent shape deliver more uniform conditional PITs than separable alternatives.
-
-Cross-term parameterizations improve fit and calibration across datasets by allowing the log-derivative to depend on context.
-Separable transports, including TRTF-AP, remain attractive when structure is approximately separable or when computational stability is paramount.
-Copulas serve as interpretable dependence baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-
-\section{Methodological Insights}
-
-Transformation forests and triangular transport are tightly linked in this modeling framework.
-With additive predictors, TRTF induces separable maps that guarantee monotonicity and interpretability but cannot capture heteroskedasticity or multimodal conditionals.
-Cross-term triangular maps restore context-dependent shape through log-derivatives and require safeguards such as quadrature, clipping, mild ridge penalties, and tail linearization.
-Copulas decouple marginals from dependence and are most useful when elliptical structure is adequate.
-
-\section{Practical Guidance}
-
-First, use separable transports or TRTF-AP when conditional shapes are approximately invariant across contexts, or when interpretability and stability take priority.
-Then, deploy cross-term maps when calibration or likelihood diagnostics indicate context-dependent variance, skewness, or multimodality, and enable quadrature, clipping, and mild regularization.
-Moreover, treat copulas as interpretable baselines and diagnostic tools for elliptical or low-dimensional dependence.
-Finally, keep train-only standardization, consistent Jacobian conventions, and the diagonal affine correction so reported log-densities remain comparable.
-
-\section{Limitations and Future Work}
-
-Key limitations center on ordering sensitivity, richer parameterizations, extended TRTF variants, and high-dimensional copulas.
-Triangular transports are anisotropic; data-driven ordering heuristics or permutation-invariant architectures may reduce variance and improve scalability.
-Richer neural parameterizations could bridge the gap to modern flows while preserving exact likelihoods.
-Non-additive predictors in transformation forests could restore cross-term capacity while retaining ensemble robustness.
-Copulas with context-aware dependence may offer interpretable yet flexible alternatives in higher dimensions.
-
-\section{Concluding Remarks}
-
-This thesis established a unified transport-based framework for comparing triangular transport maps, transformation forests, and copulas on synthetic and real tabular data.
-Recognizing when separability suffices and when cross-term capacity is necessary helps practitioners balance interpretability, computational cost, and statistical performance.
-Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 % LaTeX file for Chapter 05
+<<'preamble05',include=FALSE>>=
+library(knitr)
+opts_chunk$set(
+    fig.path='figure/ch05_fig', 
+    self.contained=FALSE,
+    cache=TRUE
+) 
+@
 
@@
 \bibliography{biblio}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-% LaTeX file for Chapter 01
-
-\chapter{Appendix}\label{ch:appendix}
-
-Maybe some R code here, probably a \rr{sessionInfo()}
-
-\end{document}

### 2025-09-20 report/chapterA1.Rnw
.

--- 2025-09-20 report/MSc_Report.tex
@@
-\documentclass[11pt,a4paper,twoside]{book}\usepackage[]{graphicx}\usepackage[]{xcolor}
-% maxwidth is the original width if it is less than linewidth
-% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
+% LaTeX file for Chapter 01
+<<'preambleA1',include=FALSE>>=
+library(knitr)
+opts_chunk$set(
+    fig.path='figure/cha1_fig', 
+    self.contained=FALSE,
+    cache=TRUE
+)
 
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlsng}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hldef}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\input{header.sty}   % packages, layout and standard macros
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\input{TitelblattMABiostatistics}
-
-\graphicspath{{./figure/}}
-\DeclareGraphicsExtensions{.pdf,.png}
-\frontmatter
-\pagenumbering{roman}
-\setcounter{tocdepth}{1}
-
-\chapter*{Abstract}
-% TODO: add abstract text.
-
-\chapter*{Acknowledgments}
-% TODO: add acknowledgments.
-
-\tableofcontents
-\setkeys{Gin}{width=.8\textwidth}
-
-\cleardoublepage
-\mainmatter
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 01
-
-\chapter{Introduction}\label{ch:intro}
-
-Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty.
-Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices.
-A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals.
-Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background.
-We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction.
-All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.
-
-This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence.
-Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required.
-The following sections outline our contributions, the research questions we address, and the scope of the work.
-
-\section{Contributions and research questions}
-
-We make three contributions within a single, notation-coherent framework.
-\begin{enumerate}
-  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
-  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
-\end{enumerate}
-
-Two questions motivate the empirical study.
-\begin{enumerate}
-  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines?
-We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
-  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)?
-We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
-\end{enumerate}
-
-\section{Scope and positioning}
-
-We focus on triangular transport maps, transformation random forests, and copulas.
-Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions.
-Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used.
-Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions.
-Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-\section{Takeaways and roadmap}
-
-A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail.
-Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests.
-Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results.
-Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 02
-
-\chapter{Methodological Background}\label{ch:background}
-
-We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$).
-All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated.
-This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.
-
-\section{Change of Variables in Standardized Coordinates}
-
-The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
-\begin{equation}
-  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
-  \label{eq:pullback}
-\end{equation}
-Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate.
-Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$.
-The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions.
-This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.
-
-It is useful to record the complementary pushforward statement.
-If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$.
-The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion.
-In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.
-
-\section{Triangular Structure and Its Consequences}
-
-A triangular map decomposes component-wise,
-\begin{equation}
-  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
-  \label{eq:triangular-map}
-\end{equation}
-and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
-\begin{equation}
-  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
-  \qquad
-  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
-  \label{eq:triangular-det}
-\end{equation}
-
-Three consequences matter in practice.
-First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$.
-Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}.
-Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components.
-Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.
-
-Existence and anisotropy deserve a brief remark.
-For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}.
-Different orderings induce different maps.
-While all are valid, their sparsity and approximation difficulty may vary with order.
-When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling.
-In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.
-
-Finally, we keep terminology consistent.
-A log density is $\log \hat\pi(\cdot)$ at a point.
-A (test) log likelihood is the dataset average of log densities.
-Its negative is the negative log likelihood (NLL) in nats; lower is better.
-These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.
-
-\section{Separable and Cross-Term Triangular Transports}
-
-A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.
-
-\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
-  \label{eq:separable-component}
-\end{equation}
-so the log Jacobian depends only on $u_k$:
-\begin{equation}
-  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
-  \label{eq:separable-jacobian}
-\end{equation}
-The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it.
-Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$.
-Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.
-
-\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
-\begin{equation}
-  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:cross-component}
-\end{equation}
-so the log Jacobian, and thus the shape of the conditional, depends on the context.
-This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.
-
-Two small examples make the contrast concrete.
-Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$.
-A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$.
-A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$.
-As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others.
-Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies.
-Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.
-
-\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable.
-Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.
-
-\section{Transformation Forests as Triangular Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}.
-This induces a triangular transport component via the probability integral transform,
-\begin{equation}
-  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
-  \label{eq:trtf-map-theory}
-\end{equation}
-
-\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
-\begin{equation}
-  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
-  \label{eq:trtf-likelihood-theory}
-\end{equation}
-showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly.
-Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}.
-TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.
-
-\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
-\begin{equation}
-  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
-  \label{eq:trtf-additive}
-\end{equation}
-so
-\begin{equation}
-  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:trtf-separable}
-\end{equation}
-and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$.
-Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence.
-TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-
-A short note on reporting closes the loop.
-TRTF evaluates conditional densities in standardized coordinates.
-Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter.
-With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.
-
-\section{Takeaways}
-
-We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction.
-Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}.
-Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization.
-TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 03
-
-\chapter{Data Analysis and Validation}\label{ch:dataanalysis}
-
-This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling programme.
-Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable.
-Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space.
-That alignment keeps objectives, diagnostics, and reported log-densities interoperable.
-
-\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}
-
-Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
-\begin{equation}
-  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
-  \qquad \sigma_k > 0,
-  \label{eq:standardization}
-\end{equation}
-learn a single monotone triangular map in this standardized space,
-\begin{equation}
-  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
-  \label{eq:map-direction}
-\end{equation}
-and sample via $S^{-1}: z \mapsto u \mapsto x$.
-All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated.
-The pullback identity \eqref{eq:pullback} remains the objective we optimize.
-
-Reported log-densities on the original scale apply the diagonal affine correction
-\begin{equation}
-  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
-  \label{eq:affine-correction}
-\end{equation}
-so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable.
-We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.
-
-The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}.
-The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}.
-Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.
-
-\section{Triangular Transport Maps}
-
-With standardized coordinates fixed, we describe the transports we fit directly.
-
-\subsection{Separable and Cross-Term Components}
-
-A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.
-
-\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
-  \label{eq:methods-separable}
-\end{equation}
-so the log-Jacobian term depends only on $u_k$.
-The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.
-
-\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
-\begin{equation}
-  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:methods-cross}
-\end{equation}
-so the log-Jacobian depends on both $u_k$ and the context.
-This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.
-
-Two examples highlight the contrast.
-If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction.
-If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.
-
-\subsection{Monotone Parameterizations}
-
-We rely on two monotonicity mechanisms that mirror the structures above.
-
-\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients.
-This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.
-
-\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$.
-The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence.
-The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.
-
-\subsection{Basis Choices and Tail Behaviour}
-
-Both parameterizations combine global and local bases with explicit tail control:
-\begin{itemize}
-  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references.
-We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows.
-Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
-  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features.
-In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context.
-We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
-\end{itemize}
-
-\subsection{Ordering, Sparsity, and Objectives}
-
-Triangular structure is anisotropic: the variable ordering matters.
-The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}.
-We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible.
-The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.
-
-With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$.
-Optimizing the pullback likelihood therefore amounts to minimising the separable objective
-\begin{equation}
-  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:ttm-objective}
-\end{equation}
-which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives.
-Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.
-
-\subsection{Safeguards for Cross-Term Training}
-
-The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
-\begin{enumerate}
-  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
-  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$.
-This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
-  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$).
-Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty.
-When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
-\end{enumerate}
-
-Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth.
-Each step is a monotone root-find with robust bracketing thanks to tail linearization.
-Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block.
-Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.
-
-\section{Transformation Random Forests as Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}.
-The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}.
-Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.
-
-Under the ubiquitous additive predictor implementation,
-\begin{equation}
-  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
-  \label{eq:methods-trtf-additive}
-\end{equation}
-so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$.
-The induced transport is separable; context shifts location but cannot change shape.
-Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion.
-Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.
-
-The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor.
-This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?
-
-\section{Copula Baselines}
-
-Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
-\begin{itemize}
-  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$.
-The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$.
-This scales gracefully to high $K$ but enforces elliptical dependence.
-  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian.
-This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
-\end{itemize}
-These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.
-
-\section{Evaluation Protocol}
-
-We evaluate all models under a common rubric.
-
-\subsection{Log-Likelihoods and NLL}
-
-For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$.
-The test log-likelihood averages this quantity over the test split, and the NLL is its negative.
-For triangular models we exploit the decomposition
-\begin{equation}
-  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:nll-decomposition}
-\end{equation}
-with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.
-
-\subsection{Calibration via PIT}
-
-Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
-\begin{equation}
-  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
-  \label{eq:pit}
-\end{equation}
-should be i.i.d.
-$\mathrm{Unif}(0,1)$ over the test set.
-We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity.
-For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent.
-Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.
-
-\subsection{Compute Metrics and Memory}
-
-We record wall-clock training time and per-sample evaluation time on the test set.
-For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature.
-TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation.
-Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries.
-All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.
-
-\subsection{Defaults and External Baselines}
-
-To keep the main text readable we adopt consolidated defaults.
-For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation.
-We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere.
-Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.
-
-\section{Data and Preprocessing}
-
-We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.
-
-\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}.
-The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.
-
-\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.
-
-\section{Metrics and Baselines}
-
-\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).
-
-\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:eval-conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.
-
-\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}.
-Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion.
-For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.
-
-\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g.
-TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.
-
-\section{Robustness and Sensitivity}
-
-All experiments use deterministic seeds; we average results over multiple runs and report standard errors.
-We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals.
-Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}.
-We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.
-
-\section{Empirical Results}
-
-This section reports empirical findings for the three modelling families within the common transport frame established above.
-All models learn a monotone triangular map $S:u\to z$ on train-only standardized inputs, evaluate Jacobians in $u$-space, and report log-densities on the original scale with the fixed affine correction.
-We summarize results by test negative log-likelihood (NLL, nats; lower is better) and, for triangular models, by per-dimension conditional NLLs that sum to the joint.
-Where informative, we discuss calibration through probability integral transform (PIT) diagnostics and comment on computation.
-External flow results on MINIBOONE are included only as context from the literature.
-
-\subsection{Synthetic Data}
-
-\subsubsection{Half-Moon ($K=2$)}
-
-\textbf{Setup.} The two interleaving half-circles with additive Gaussian noise produce a bimodal joint with strong curvature.
-We fit TTM-Marginal, TTM-Separable (TTM-S), TTM-Cross-term (TTM-X), TRTF under an additive predictor (TRTF-AP), and a low-dimensional nonparametric copula with probit transform and KDE dependence.
-Two references provide context: a true joint likelihood based on the known mixture and a class-conditional oracle $p(x\mid y)$ using the true component label.
-
-\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs.
-TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape.
-TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}.
-The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Half-Moon test negative log-likelihoods (nats). Lower is better; $\pm$ indicates standard error.}
-  \label{tab:halfmoon-nll}
-  \begin{tabular}{lccc}
-    \hline
-    Model & Mean joint NLL & Conditional NLL 1 & Conditional NLL 2 \\ 
-    \hline
-    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
-    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
-    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
-    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
-    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
-    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
-    True joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsubsection{Four-Dimensional Autoregressive Generator}
-
-\textbf{Setup.} The synthetic four-dimensional task combines heteroskedastic, skewed, and conditional multimodal components over a range of sample sizes. We compare the same suite of models as above.
-
-\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive.
-As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity.
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.
-
-\begin{table}[htbp]
-  \centering
-  \caption{Four-dimensional autoregressive generator ($n=50$): conditional and joint NLLs (nats). Values are means with standard errors.}
-  \label{tab:autoregressive-nll}
-  \begin{tabular}{llrrrrrr}
-    \hline
-    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
-    \hline
-    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
-    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
-    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
-    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
-    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\subsection{Real Data: MINIBOONE}
-
-\textbf{Setup.} We train all models on the standardized MINIBOONE tabular dataset ($K=43$) with the protocol from \citet{papamakarios2017masked}.
-For comparison we include published NLL results from high-capacity normalising flows.
-
-\textbf{Results.}~Table~\ref{tab:miniboone-ll} summarises average test log-likelihoods reported in nats (higher is better).
-TRTF-AP improves markedly over an independent Gaussian baseline, yet it trails deep flows reported in the literature, which is consistent with the separable Jacobian constraint.
-\begin{table}[htbp]
-  \centering
-  \caption{MINIBOONE test log-likelihoods (nats, higher is better). Literature numbers follow \citet{papamakarios2017masked}.}
-  \label{tab:miniboone-ll}
-  \begin{tabular}{lr}
-    \hline
-    Model & Test log-likelihood \\ 
-    \hline
-    Gaussian independent baseline & $-37.24 \pm 1.07$ \\ 
-    TRTF-AP (this study) & $-29.88 \pm 0.02$ \\ 
-    MADE with auxiliary conditioning & $-15.59 \pm 0.50$ \\ 
-    Real NVP (five layers) & $-13.55 \pm 0.49$ \\ 
-    MAF (five layers) & $-11.75 \pm 0.44$ \\ 
-    MAF mixture of Gaussians (five layers) & $-11.68 \pm 0.44$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset.
-The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}.
-Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
-\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$.
-Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards.
-TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter.
-We additionally report memory footprints where caching affects runtime.
-
-\paragraph{Summary.} Across synthetic and real datasets, allowing context-dependent shape via cross-term derivatives consistently improves fit and calibration.
-Separable transports---including TRTF-AP---remain attractive when structure is near-separable or when computational stability is paramount.
-Copulas offer interpretable baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-The next chapter interprets these findings and discusses practical implications.
-\section{Takeaways}
-
-The methodological backbone is deliberately uniform: learn a single monotone triangular map $S:u\mapsto z$ in standardized coordinates, keep all Jacobians and derivatives in $u$-space, and report log-densities on $x$ with the fixed affine correction \eqref{eq:affine-correction}.
-Within that frame, separable transports shift location but keep shape fixed across contexts; cross-term transports allow context-dependent shape via quadrature, clipping, and regularisation; TRTF induces the same triangular likelihood via $S_k=\Phi^{-1}(\widehat{F}_k)$, with the additive predictor implying separability exactly; and copulas serve as interpretable dependence baselines.
-We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 04
-
-\chapter{Interpretation and Conclusion}\label{ch:conclusion}
-
-This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
-% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins
-
-\section{Summary of Findings and Trade-offs}
-
-\subsection{Synthetic benchmarks}
-
-For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests.
-This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}).
-A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower.
-The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}).
-TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}).
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.
-
-\subsection{MINIBOONE}
-
-The additive-predictor transformation forest (TRTF-AP) improves over an independent baseline on the real-world MINIBOONE dataset ($K=43$).
-The estimator remains behind state-of-the-art normalizing flows, consistent with the separable Jacobian constraint (Table~\ref{tab:miniboone-ll}).
-Cross-term TTM variants reduce the remaining gap.
-These variants require additional optimization safeguards such as quadrature, clipping, and regularization.
-Probability integral transform (PIT) diagnostics show that models with context-dependent shape deliver more uniform conditional PITs than separable alternatives.
-
-Cross-term parameterizations improve fit and calibration across datasets by allowing the log-derivative to depend on context.
-Separable transports, including TRTF-AP, remain attractive when structure is approximately separable or when computational stability is paramount.
-Copulas serve as interpretable dependence baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-
-\section{Methodological Insights}
-
-Transformation forests and triangular transport are tightly linked in this modeling framework.
-With additive predictors, TRTF induces separable maps that guarantee monotonicity and interpretability but cannot capture heteroskedasticity or multimodal conditionals.
-Cross-term triangular maps restore context-dependent shape through log-derivatives and require safeguards such as quadrature, clipping, mild ridge penalties, and tail linearization.
-Copulas decouple marginals from dependence and are most useful when elliptical structure is adequate.
-
-\section{Practical Guidance}
-
-First, use separable transports or TRTF-AP when conditional shapes are approximately invariant across contexts, or when interpretability and stability take priority.
-Then, deploy cross-term maps when calibration or likelihood diagnostics indicate context-dependent variance, skewness, or multimodality, and enable quadrature, clipping, and mild regularization.
-Moreover, treat copulas as interpretable baselines and diagnostic tools for elliptical or low-dimensional dependence.
-Finally, keep train-only standardization, consistent Jacobian conventions, and the diagonal affine correction so reported log-densities remain comparable.
-
-\section{Limitations and Future Work}
-
-Key limitations center on ordering sensitivity, richer parameterizations, extended TRTF variants, and high-dimensional copulas.
-Triangular transports are anisotropic; data-driven ordering heuristics or permutation-invariant architectures may reduce variance and improve scalability.
-Richer neural parameterizations could bridge the gap to modern flows while preserving exact likelihoods.
-Non-additive predictors in transformation forests could restore cross-term capacity while retaining ensemble robustness.
-Copulas with context-aware dependence may offer interpretable yet flexible alternatives in higher dimensions.
-
-\section{Concluding Remarks}
-
-This thesis established a unified transport-based framework for comparing triangular transport maps, transformation forests, and copulas on synthetic and real tabular data.
-Recognizing when separability suffices and when cross-term capacity is necessary helps practitioners balance interpretability, computational cost, and statistical performance.
-Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 05
-
-\chapter{References}\label{ch:refs}
-
-\bibliographystyle{mywiley}
-\bibliography{biblio}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-% LaTeX file for Chapter 01
+@
 
@@
 Maybe some R code here, probably a \rr{sessionInfo()}
-
-\end{document}

### 2025-09-20 report_old/report.Rmd
.

---
title: "Multivariate Density Estimation: Comparing Transformation Forests, Triangular Transport Maps, and Copulas"
author: "Léon Kia Faro"
date: "`r format(Sys.Date())`"
output: 
    pdf_document: 
        number_sections: true 
        toc: true 
        latex_engine: xelatex 
    html_document: 
        toc: true 
        toc_depth: 3 
        number_sections: true 
        fig_caption: true
df_print: paged
fontsize: 11pt
geometry: margin=1in
linkcolor: blue
bibliography: bib.bib
header-includes: 
    - \usepackage{amsmath,amssymb,mathtools} 
    - \usepackage{algorithm} 
    - \usepackage{algpseudocode} 
    - \floatname{algorithm}{Algorithm} 
    - \renewcommand{\algorithmicrequire}{\textbf{Input:}} 
    - \renewcommand{\algorithmicensure}{\textbf{Output:}}
    # Define box environments for notation/lemmas
    - \usepackage{tcolorbox}
    - \newtcolorbox{notebox}{colback=gray!10!white,colframe=gray!75!black,title=}
    - \newtcolorbox{lemmabox}{colback=blue!5!white,colframe=blue!75!black,title=}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# Load necessary libraries and source functions here
```
# Introduction

Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ supports probabilistic modeling, anomaly detection, simulation based inference, and decision making under uncertainty. Modern applications combine high dimension with structure that changes across contexts. Conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices. A coherent response is **measure transport**. We couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals. Throughout we adopt the tutorial’s notation and **directionality**. We standardize observations by $u=(x-\mu)\oslash\sigma$, learn a **monotone triangular** map $S: u \to z$ to the reference $z \sim \eta=\mathcal N(0,I)$, and sample via the inverse $S^{-1}: z \to u \to x$. All Jacobians and partial derivatives involving $S$ are taken **in $u$-space** unless stated otherwise. Reporting log densities on the original scale applies the usual affine correction $\log \pi_X(x)=\log \pi_U(u)-\sum_k \log \sigma_k$ with $u=T_{\text{std}}(x)$ [@ramgraber2025friendly].

The central identity is the pullback of the reference by $S$,

$$
\pi_U(u)=\eta\!\big(S(u)\big)\, \big|\det\nabla_u S(u)\big|.
$$

Triangularity makes $\nabla_u S(u)$ lower triangular, hence $\log |\det\nabla_u S|$ equals a **sum of one dimensional terms**. Likelihoods thus evaluate in linear time in dimension, strict monotonicity ensures global invertibility, and conditionals become accessible by back substitution [@rosenblatt1952remarks; @knothe1957contributions; @ramgraber2025friendly]. These properties are practical. They yield efficient objectives, exact inversion through sequential one dimensional root finding, and straightforward conditional sampling by fixing early coordinates and inverting only the trailing components. Triangular structure also enables **sparsity**. Conditional independencies correspond to dropping arguments from components, which sparsifies the Jacobian and improves robustness and scaling [@rosenblatt1952remarks; @ramgraber2025friendly].

Two parameterizations recur and matter for interpretation. A **separable triangular map** uses

$$
S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k), \qquad
\log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
$$

so the log Jacobian depends **only** on $u_k$. Context can move location through $g_k$, but it **cannot change shape**. Scale, skewness, tail thickness, and modality remain fixed across contexts. A **cross term triangular map** uses the integrated rectifier form

$$
S_k(u_{1:k}) = \int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\,dt + c_k(u_{1:k-1}), \qquad
\log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
$$

so the log Jacobian depends on both $u_k$ and the context $u_{1:k-1}$. Cross term maps therefore capture **context dependent shape**. They model conditional heteroskedasticity, skew that flips with predictors, and mode splitting that appears only for some contexts. Cross term training requires careful numerical safeguards; details follow in Methods.

This study examines three model classes that realize the transport program with different inductive biases. **Triangular transport maps** parameterize $S$ directly and fit it by maximizing the pullback likelihood against a Gaussian reference [@ramgraber2025friendly]. **Transformation random forests** estimate conditional distributions with transformation models and thereby **induce** a triangular map through the probability integral transform; for each $k$,

$$
S_k(u_{1:k})=\Phi^{-1}\big(\widehat F_k(u_k \mid u_{1:k-1})\big),
$$

and differentiating $\Phi(S_k)=\widehat F_k$ in $u_k$ gives the **likelihood identity**

$$
\widehat \pi_k(u_k \mid u_{1:k-1})
= \phi\!\big(S_k(u_{1:k})\big)\, \partial_{u_k} S_k(u_{1:k}).
$$

Thus the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly. Under the common **additive predictor** implementation,

$$
\widehat F_k(u_k \mid u_{1:k-1}) = \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),
$$

the induced transport is **separable**, because $S_k=h_k(u_k)+g_k(u_{1:k-1})$ and $\partial_{u_k}S_k=h_k'(u_k)$. Monotonicity in $u_k$ is ensured by construction, and the Jacobian term depends only on $u_k$ [@hothorn2018conditional; @hothorn2017transformation; @hothorn2021transformation]. **Copulas** separate marginals from dependence through Sklar’s theorem. They provide interpretable baselines and scalable semiparametric options in higher dimensions, with limits that are clear when dependence is far from elliptical [@sklar1959fonctions].

A short intuition for triangular transport clarifies why this structure matters. The triangular determinant reduces to a product of diagonal terms, so the log determinant is a sum. Likelihood evaluation is linear in $K$. Strict monotonicity in each last coordinate yields bijectivity and replaces full dimensional inversion with a sequence of one dimensional solves. Fixing early coordinates exposes conditionals explicitly, which makes conditional simulation direct. When conditional independence holds, triangular form lets us **drop inputs** from components. The resulting sparse Jacobian reduces computational cost and stabilizes estimation. These features combine to produce estimators that scale gracefully and remain interpretable [@rosenblatt1952remarks; @knothe1957contributions; @ramgraber2025friendly].

We use terminology consistently. A **log density** is the value of $\log \hat\pi(\cdot)$ at a point. A **log likelihood** is a sum or an average of log densities over a dataset. Its negative is the **negative log likelihood (NLL)**, reported in **nats** with the natural logarithm; lower is better. Unless stated otherwise, we compute all derivatives and Jacobians in $u$-space and apply the $-\sum_k \log \sigma_k$ correction when reporting on the $x$-scale. We standardize once on the training split and re use those parameters at evaluation time to avoid leakage. These conventions follow common practice in triangular transport and in transformation models [@hothorn2017transformation; @hothorn2021transformation; @ramgraber2025friendly].

Normalizing flows provide a useful reference point. They build transport maps by composing many invertible layers with permutations and coupling, often with autoregressive structure. Modern flows achieve strong likelihoods on tabular data but typically do not enforce strict triangular structure in the sense used here. We treat flows as context and as published baselines rather than as a focal method. They anchor the high capacity end of the spectrum in our empirical comparisons and help interpret the benefits and limits of triangular structure and TRTF induced transports [@rezende2015variational; @dinh2017real; @papamakarios2021normalizing].

**What separable maps miss, and why this thesis matters.** Separable transports—including additive predictor transformation forests—**shift location but cannot change conditional shape with context**. When variance, skewness, tail behavior, or modality of $u_k \mid u_{1:k-1}$ varies with the predictors, the separable constraint becomes the dominant source of error. Cross term transports remove this structural bottleneck. Our study quantifies where the separable structure suffices and where cross term capacity is required, and it explains when TRTF, which is separable under the additive predictor implementation, is expected to perform well versus when it is structurally limited [@hothorn2017transformation; @hothorn2021transformation].

**Copulas in brief.** Copulas factor the joint into marginals and dependence. In practice we will use two concise recipes. A **semiparametric Gaussian copula** estimates pseudo observations $\hat v_{ik}=r_{ik}/(N+1)$ via ranks, transforms them by $z_{ik}=\Phi^{-1}(\hat v_{ik})$, and estimates the correlation matrix. This yields a flexible marginal model with elliptical dependence. A **low dimensional nonparametric copula** applies the probit transform $z=\Phi^{-1}(v)$, fits a kernel density in $\mathbb R^K$, and maps back with the appropriate Jacobian. This approach works for very low dimension but becomes fragile as $K$ grows, so we use it where appropriate and default to the semiparametric option in higher dimensions [@sklar1959fonctions].

We keep the mathematical footprint of the introduction minimal by retaining only the pullback identity and the separable versus cross term contrast. All further derivations, including proofs, optimization details, and the full bridge between TRTF and transport likelihoods, appear in **Theoretical Foundations** and **Methods**. Cross term training requires careful numerical safeguards; details follow in Methods.

## Conventions and terminology

We standardize data to $u=(x-\mu)\oslash\sigma$ and **learn $S: u \to z$**. Sampling uses $S^{-1}: z \to u \to x$. Unless stated otherwise, all partial derivatives and Jacobians of $S$ are taken in $u$-space. The reference factorizes as $\eta(z)=\prod_k \phi(z_k)$, and triangularity yields $\det\nabla_u S=\prod_k \partial_{u_k}S_k$. Reported log densities on the original scale apply the correction $-\sum_k \log \sigma_k$. We use **log density** for a point value, **log likelihood** for a dataset aggregate, and **NLL** for the negative log likelihood; units are **nats**. Symbols follow a fixed convention: $x,u,z,S,\Phi,\phi,\pi_X,\pi_U,\eta$. When we discuss transformation forests we write $\widehat F_k(u_k \mid u_{1:k-1})$ for the conditional distribution estimate and $S_k=\Phi^{-1}(\widehat F_k)$ for the induced transport component [@hothorn2018conditional; @hothorn2017transformation; @hothorn2021transformation].

## Contributions and research questions

This thesis contributes three elements within a single, notation coherent framework. First, it **clarifies the theory** that links transformation forests to triangular transport. The per component **likelihood identity** follows from differentiating $\Phi(S_k)=\widehat F_k$. Under the additive predictor implementation the induced $S_k$ is **separable**, which makes the model class and its limitations explicit [@hothorn2017transformation; @hothorn2021transformation; @hothorn2018conditional]. Second, it **compares** separable and cross term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data with a consistent evaluation protocol, reporting LL and NLL in nats together with calibration diagnostics. Third, it **documents** the implementation choices that matter in practice, including standardization, variable ordering, and numerical safeguards for cross term training, while keeping the introduction focused.

We study two questions that make the gap above empirical and actionable.

1. **Practical performance.** How do transformation forests perform on synthetic and real tabular data **relative to triangular transport maps, copulas, and published normalizing flow baselines**. We ask when separable transports, either via TRTF under additive predictors or via separable direct parameterizations, reach the performance of high capacity references, and where they fall short when conditional shape depends on context [@rezende2015variational; @dinh2017real; @papamakarios2021normalizing; @hothorn2017transformation; @hothorn2021transformation].

2. **Trade offs.** What are the trade offs among **fit** measured by test NLL, **computational efficiency** measured by wall clock times for training and per sample test evaluation, and **calibration** assessed by probability integral transform diagnostics. We ask where separable transports deliver the best speed to accuracy ratio, and where cross term capacity justifies its extra cost.

## Scope and positioning

We focus on **triangular transport maps**, **transformation random forests**, and **copulas**. Transport maps let us ablate separable versus cross term capacity under a common objective and with the same directionality and Jacobian calculus. Transformation forests bring a mature nonparametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation with **separable** Jacobians under additive predictors. Copulas provide semiparametric and nonparametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions. **Normalizing flows** remain an important reference. We use them for context and draw on published tabular benchmarks, because they represent high capacity autoregressive and coupling architectures that are widely adopted in density estimation [@rezende2015variational; @dinh2017real; @papamakarios2021normalizing].

## Take aways

We adopt the tutorial’s direction $S: u \to z$ and compute all Jacobians in $u$-space. Triangular transport offers linear time likelihoods, exact inversion, and straightforward conditional simulation. **Separable** transports move locations but keep shape fixed across contexts. **Cross term** transports allow context dependent shape. Transformation forests naturally induce a triangular map $S_k=\Phi^{-1}(\widehat F_k)$, with a per component likelihood identity and the specialization that **additive predictors imply separability**. Copulas provide interpretable baselines through decoupled marginals and dependence. We use consistent LL and NLL terminology in **nats** and keep the introduction lean in mathematics. Detailed derivations and implementation choices follow in **Theoretical Foundations** and **Methods**. The experiments connect structure to observed performance, highlighting when separable structure suffices and when cross term interactions are necessary [@rosenblatt1952remarks; @knothe1957contributions; @hothorn2017transformation; @hothorn2021transformation; @hothorn2018conditional; @ramgraber2025friendly; @sklar1959fonctions; @rezende2015variational; @dinh2017real; @papamakarios2021normalizing].

<!-- Checklist (done): contributions paragraph; roadmap sentence -->
This thesis makes three contributions to conditional transport modeling:

- We formulate triangular transport maps, transformation forests, and copulas within a shared standardized coordinate system.
- We prove the additive predictor specialization of transformation forests induces separable triangular maps and matches pullback likelihoods.
- We evaluate synthetic and real benchmarks to identify when separable or cross-term structure improves calibration and likelihood.

Section 2 reviews the theoretical foundations that motivate the transport formalism for the comparisons. Section 3 develops the modeling procedures that span triangular maps, forests, and copulas. Section 4 details the evaluation protocol, Section 5 presents empirical results, and Section 6 discusses implications.

# Theoretical Foundations

We adopt the tutorial’s directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a **monotone triangular** map $S:u\to z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal N(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\text{std}}(x))$ with the usual affine Jacobian correction when reporting $\log\pi_X(x)$ (subtract $\sum_k\log\sigma_k$). All Jacobians and partial derivatives involving $S$ are taken **in $u$-space** unless explicitly stated. This section develops the minimal change of variables machinery, explains why triangular structure is effective, makes precise the contrast between **separable** and **cross term** triangular maps, and shows how **transformation random forests (TRTF)** instantiate the same transport with an immediate likelihood identity [@rosenblatt1952remarks; @knothe1957contributions; @hothorn2017transformation; @hothorn2021transformation; @hothorn2018conditional; @ramgraber2025friendly].

## Change of variables in standardized coordinates

The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood:

$$
\pi_U(u)=\eta\!\big(S(u)\big)\, \big|\det\nabla_u S(u)\big|. \tag{1}
$$

Because we learn $S$ on standardized inputs $u$, this is the form we optimize and evaluate. Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\text{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\text{std}}(x)$. The reference factorizes as $\eta(z)=\prod_k \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions. This linear in $K$ structure underpins all evaluation and training in later chapters [@ramgraber2025friendly].

It is useful to record the complementary pushforward statement. If $u\sim \pi_U$ and $z=S(u)$, then $S_\#\pi_U=\eta$. The pair $(S^\sharp\eta, S_\#\pi_U)$ clarifies that the same map both defines the **model density** in $u$-space via (1) and certifies exact sampling by inversion. In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in the standardized coordinates, and only apply the affine correction when moving results back to the original scale.

## Triangular structure and its consequences

A **triangular map** decomposes component wise,

$$
S(u)=\big(S_1(u_1),\ S_2(u_{1:2}),\ \ldots,\ S_K(u_{1:K})\big),
$$

and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence

$$
\det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}), \qquad 
\log \big|\det\nabla_u S(u)\big|=\sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}). \tag{2}
$$

Three consequences matter in practice. First, **likelihoods are efficient**: the log determinant is a sum of one dimensional terms, so evaluating (1) scales in $\mathcal O(K)$. Second, **invertibility is exact**: strict monotonicity implies global bijectivity, and inversion reduces to sequential one dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate [@rosenblatt1952remarks; @knothe1957contributions]. Third, triangularity yields **transparent conditionals**: to sample $u_{m:K}\mid u_{1:m-1}=u^\star_{1:m-1}$, fix the early coordinates at $u^\star$ and invert only the trailing components. Equivalently, the map aligns with the chain rule factorization $\pi_U(u)=\prod_k \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one dimensional transformation [@ramgraber2025friendly].

Existence and anisotropy deserve a brief remark. For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe–Rosenblatt rearrangement [@rosenblatt1952remarks; @knothe1957contributions]. Different orderings induce different maps. While all are valid, their **sparsity** and approximation difficulty may vary with order. When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling. In high dimension, such sparsity often drives both statistical efficiency and computational cost [@ramgraber2025friendly].

Finally, we keep terminology consistent. A **log density** is $\log \hat\pi(\cdot)$ at a point. A **(test) log likelihood** is the dataset average of log densities. Its negative is the **negative log likelihood (NLL)** in **nats**; lower is better. These conventions are used uniformly in Methods and Results.

## Separable and cross term triangular transports

A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.

**Separable triangular maps.** A separable component decomposes into a context dependent shift and a univariate monotone shape,

$$
S_k(u_{1:k})=g_k(u_{1:k-1})+h_k(u_k), \qquad \partial_{u_k}S_k(u_{1:k})=h_k'(u_k)>0,
$$

so

$$
\log \partial_{u_k}S_k(u_{1:k})=\log h_k'(u_k) \ \text{depends only on } u_k.
$$

**Intuition.** The context $u_{1:k-1}$ can **move** the conditional through $g_k$, but it cannot **reshape** it. Scale, skewness, tail thickness, or modality remain fixed across contexts because the log Jacobian contribution depends only on $u_k$. Separable transports capture nonlinear location shifts across contexts with a context invariant shape.

**Cross term triangular maps.** A cross term component uses an integrated rectifier,

$$
S_k(u_{1:k})=\int^{u_k}\!\exp\!\big(h_k(t,u_{1:k-1})\big)\,dt + c_k(u_{1:k-1}),\quad
\log \partial_{u_k}S_k(u_{1:k})=h_k(u_k,u_{1:k-1}),
$$

so the log Jacobian, and thus the **shape** of the conditional, **depends on the context**. This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.

Two small examples make the contrast concrete. Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$. A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$. A cross term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$. As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others. Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies. Cross term training requires careful numerical safeguards; details follow in Methods.

Two implementation notes guide later choices. The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one dimensional quadrature and regularization to remain stable. Separable maps often admit linear in coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context invariant conditional shape [@ramgraber2025friendly].

## Transformation forests as triangular transport

Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi(h(Y\mid W))$ is standard, letting predictors act through $h$ while preserving monotonicity [@hothorn2018conditional]. A **transformation random forest** aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ [@hothorn2017transformation; @hothorn2021transformation]. This induces a triangular transport component via the probability integral transform:

$$
S_k(u_{1:k})=\Phi^{-1}\!\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
$$

Two identities follow immediately.

**Per component likelihood (one liner).** Differentiating $\Phi(S_k(u_{1:k}))=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives

$$
\widehat\pi_k(u_k\mid u_{1:k-1})
= \phi\!\big(S_k(u_{1:k})\big)\,\partial_{u_k}S_k(u_{1:k}),
$$

that is, the **TRTF conditional density equals the pullback factor** one would obtain by parameterizing $S$ directly. Summing over $k$ yields the joint log likelihood from (1) with the triangular log determinant in (2). TRTF therefore implements the same transport likelihood once standardized [@hothorn2017transformation; @hothorn2021transformation; @ramgraber2025friendly].

**Additive predictor specialization implies separability.** Under the common **additive predictor** implementation (TRTF AP),

$$
\widehat F_k(u_k\mid u_{1:k-1})=\Phi\!\big(h_k(u_k)+g_k(u_{1:k-1})\big),
$$

so

$$
S_k(u_{1:k})=h_k(u_k)+g_k(u_{1:k-1}),\qquad
\partial_{u_k}S_k(u_{1:k})=h_k'(u_k),
$$

and the induced transport is **separable**: $\log\partial_{u_k}S_k$ depends only on $u_k$. Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence. TRTF AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors [@hothorn2017transformation; @hothorn2021transformation; @hothorn2018conditional].

A short note on reporting closes the loop. TRTF evaluates conditional densities in standardized coordinates. Reported log densities on the original scale apply the affine correction $-\sum_k\log\sigma_k$ that accompanies $T_{\text{std}}$, matching the convention set at the beginning of this chapter. With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.

**Take aways.** We learn $S:u\to z$ in standardized coordinates and evaluate $\log\pi_X$ via a simple affine correction. Triangularity turns the log determinant into a sum of one dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward [@rosenblatt1952remarks; @knothe1957contributions; @ramgraber2025friendly]. **Separable** transports move locations but keep shape fixed across contexts; **cross term** transports allow context dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization. TRTF provides a nonparametric route to the same triangular transport with a per component **likelihood identity** and the important specialization that **TRTF AP is separable by construction** [@hothorn2017transformation; @hothorn2021transformation; @hothorn2018conditional]. These foundations justify the objectives and implementations developed in Methods and frame the empirical comparisons that follow.

# Methods

This chapter turns the formal commitments of the introduction and theory into a coherent, practical modeling program. The goal is one transport‑based language for three model families—**Triangular Transport Maps (TTM)**, **Transformation Random Forests (TRTF)**, and **Copulas**—so that likelihoods, calibration, and compute can be compared head‑to‑head on synthetic and real tabular data. The thread running through every section is the same: we standardize data, learn a **monotone triangular** map **to** a simple reference, and evaluate all Jacobians **in the standardized space**. That seemingly small alignment makes every formula, objective, and diagnostic interoperable across methods.

Consistent notation helps. Observations on the original scale are $x\in\mathbb R^K$. We standardize (on the **training split only**) via
$u=T_{\mathrm{std}}(x)=(x-\mu)\oslash\sigma$ with positive scale vector $\sigma$. We then **learn a single triangular map in this standardized space**,

$$
S:\ u\mapsto z,\qquad z\sim \eta=\mathcal N(0,I),
$$

and **sample** via $S^{-1}:z\!\to\!u\!\to\!x$. All Jacobians and partial derivatives that involve $S$ are taken **with respect to $u$** unless explicitly stated. The pullback identity we optimize and evaluate is

$$
\pi_U(u) \;=\; \eta\!\big(S(u)\big)\,\Big|\det\nabla_{u}S(u)\Big|,
$$

and reported log‑densities on the original scale use the affine correction

$$
\log \pi_X(x) \;=\; \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) \;-\;\sum_{k=1}^K \log \sigma_k.
$$

Triangularity means $S_k$ depends only on $u_{1:k}$ and is strictly increasing in $u_k$; thus

$$
\det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
$$

a linear‑time log‑determinant, exact invertibility by back‑substitution, and direct access to conditionals [@rosenblatt1952remarks; @knothe1957contributions].

What follows is organized in five subsections. The first fixes the standardized coordinate system and explains why committing to $S:u\to z$ is the right comparison frame. The second develops TTM parameterizations, monotone bases, objectives, and the specific safeguards that make cross‑term training stable. The third shows how TRTF induces **the same** triangular likelihood and why the widely used additive predictor implementation is **separable** by construction. The fourth sets up copulas as interpretable dependence baselines and records the exact recipes we will use. The fifth codifies the evaluation protocol—log‑likelihoods in **nats**, conditional decompositions, PIT calibration, and compute metrics—used throughout the experiments.

---

## Setup: Standardization, Map Direction, and Reporting

Having settled the notation, the first practical choice is **where** the map lives. We choose to learn a **single** monotone triangular transport $S$ **from the standardized target to the reference**:

$$
S:\ u=(x-\mu)\oslash\sigma \;\longmapsto\; z\sim\eta=\mathcal N(0,I).
$$

This direction $u\to z$ keeps the objective uniform across methods and turns the change‑of‑variables identity into a compact, separable loss. Because $S$ is triangular and strictly increasing in each terminal coordinate,

$$
\log\left|\det\nabla_u S(u)\right|=\sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}),
$$

so per‑sample likelihoods evaluate in $\mathcal O(K)$ time and never touch a dense determinant. The same structure gives us **exact inversion** (solve $K$ one‑dimensional monotone equations in sequence) and **transparent conditionals**: fixing $u_{1:m-1}$ and inverting the trailing components delivers draws from $\pi_U(u_{m:K}\mid u_{1:m-1})$ without auxiliary approximations [@rosenblatt1952remarks; @knothe1957contributions].

Two bookkeeping rules ensure that different estimators remain commensurate. **First**, every Jacobian and partial derivative that involves $S$ is taken in **$u$-space**. This includes the diagonal derivatives $\partial_{u_k} S_k$ that enter the log‑determinant and any gradients used for optimization. **Second**, all reported log‑densities on the original $x$-scale apply the same diagonal affine correction:

$$
\log\pi_X(x)=\log\pi_U\!\big(T_{\mathrm{std}}(x)\big)-\sum_{k=1}^K\log\sigma_k.
$$

This way, all models produce pointwise log‑densities $\log \hat\pi_X(x)$ that are directly comparable. We reserve the terms **log‑density** for the value at a point, **(test) log‑likelihood** for a dataset average, and **NLL** (negative log‑likelihood) for the negated average; NLL is reported in **nats**.

Fixing the direction also clarifies what is **not** needed. Some transport approaches choose the inverse direction $R:z\to u$ and push the reference forward to the target; that is natural in simulation‑based inference but would force us to mix objectives and Jacobian conventions across methods. Adopting $S:u\to z$ for everything keeps the likelihood identity for TRTF on the same footing as direct TTM training and allows copulas—whose natural home is $x$-space—to be injected cleanly via the standardization wrapper. Finally, normalizing flows are treated purely as context and published baselines: they also compose invertible transforms and often exploit autoregressive sublayers, but usually do **not** insist on strict triangularity, which is precisely the structure we want to highlight here [@rezende2015variational; @dinh2017real; @papamakarios2021normalizing].

---

## Triangular Transport Maps: Parameterizations, Bases, and Objectives

With standardized coordinates fixed, we can describe the transport models we will **fit directly**. A triangular map factorizes as

$$
S(u)=\big(S_1(u_1),\,S_2(u_{1:2}),\,\dots,\,S_K(u_{1:K})\big),
\qquad \partial_{u_k}S_k(u_{1:k})>0.
$$

The modeling question is how to parameterize each $S_k$ so that (i) invertibility is guaranteed, (ii) training is numerically stable, and (iii) expressiveness matches the data. Two canonical options anchor the spectrum.

**Separable triangular maps.** The separable form isolates a context shift and a univariate monotone shape:

$$
S_k(u_{1:k})=g_k(u_{1:k-1})+h_k(u_k),\qquad \log\partial_{u_k}S_k(u_{1:k})=\log h_k'(u_k).
$$

This structure makes the **log‑Jacobian term depend only on $u_k$**. The context $u_{1:k-1}$ can shift the conditional location through $g_k$, but **cannot change shape**: conditional scale, skewness, and tail thickness stay invariant across contexts. The benefit is speed and robustness—monotonicity reduces to a one‑dimensional requirement on $h_k$, often enforceable by nonnegative weights on monotone bases.

**Cross‑term triangular maps.** To model context‑dependent shape, we use the integrated‑rectifier construction:

$$
S_k(u_{1:k})=\int^{u_k}\!\exp\!\big(h_k(t,u_{1:k-1})\big)\,dt\;+\;c_k(u_{1:k-1}),
\qquad \log\partial_{u_k}S_k(u_{1:k})=h_k(u_k,u_{1:k-1}).
$$

Here the log‑Jacobian depends on both $u_k$ and the context $u_{1:k-1}$, so heteroskedasticity, skew changes, and even mode‑splitting as the context varies are all in scope. The price is numerical care: the integral typically requires one‑dimensional quadrature, and the exponential rectifier can overflow if left unconstrained.

This separable vs cross‑term contrast is the central expressiveness decision for TTM. It is also the lens through which we interpret TRTF later: as soon as a model’s per‑component log‑derivative $\log \partial_{u_k}S_k$ becomes a function of $u_k$ **only**, that model is separable and will miss context‑dependent shape by design.

**Monotone parameterizations.** We rely on two monotonicity mechanisms, chosen to match the two structures above.

*For separable maps*, we write $h_k$ as a linear combination of **monotone one‑dimensional basis functions**—for instance the identity, integrated sigmoids, softplus‑like edge terms, or integrated radial basis functions—and constrain their coefficients to be nonnegative. With this linear‑in‑coefficients (LIC) construction, $\partial_{u_k}S_k=h_k'(u_k)$ is nonnegative by design, and optimization often reduces to a convex subproblem for the monotone part and a least‑squares update for the context shift $g_k$. The result is a fast, stable inner loop with exact Jacobians.

*For cross‑term maps*, we parameterize the log‑derivative $h_k(u_k,u_{1:k-1})$ directly and **rectify** it (e.g., via $\exp$) before integrating over $u_k$. The rectifier enforces strict positivity of $\partial_{u_k}S_k$, and the integral ensures global monotonicity in the last coordinate while allowing flexible dependence on the context. This expressiveness is what we need to capture conditional variance inflation, skew flips, and context‑specific multimodality, but it makes training numerically delicate.

**Basis choices and tail behavior.** In both settings we combine global and local bases with explicit tail control:

*Hermite polynomials/functions.* Hermite polynomials are natural when the reference is Gaussian: they offer global support and convenient orthogonality. To avoid uncontrolled growth in the tails, we primarily use **Hermite functions**—polynomials multiplied by a Gaussian weight—for higher‑order terms, and keep **linear terms unweighted** so that $S_k$ reverts to linear in the far tails. This “tail linearization” matters: stable inversions rely on strictly positive derivatives that neither vanish nor explode as $|u_k|$ grows.

*Localized sigmoids and radial bases.* To capture sharp local features, we use integrated sigmoids (e.g., error‑function integrals), softplus‑like edge terms, and integrated radial basis functions (iRBFs). In separable maps these serve as monotone building blocks for $h_k(u_k)$; in cross‑term maps they can enter $h_k(u_k,u_{1:k-1})$ to let the log‑derivative bend with context. Placing centers at empirical quantiles and setting widths by nearest‑neighbor distances yields well‑spread coverage without adding fit‑time location parameters.

**Ordering and sparsity.** Triangular structure is anisotropic: the **ordering of variables matters**. The Knothe–Rosenblatt rearrangement guarantees the existence of a monotone triangular coupling for any ordering, but the **sparsity** and approximation **difficulty** of a finite‑basis parameterization can change dramatically with the order [@rosenblatt1952remarks; @knothe1957contributions]. We align with the data’s natural order for primary results and drop arguments from $S_k$ whenever conditional independence is plausible; the corresponding entries vanish from the Jacobian, evaluation becomes cheaper, and variance in small‑$n$ regimes is reduced. This built‑in parsimony is one of the reasons triangular maps scale well.

**Objective and per‑component decomposition.** With $z\sim\mathcal N(0,1)$ i.i.d. under the reference, $\log\eta(S(u))=-\tfrac12\sum_k S_k(u_{1:k})^2-\tfrac K2\log(2\pi)$. Thus, optimizing the pullback likelihood $\pi_U(u)=\eta(S(u))|\det\nabla_u S(u)|$ amounts to minimizing the forward KL (equivalently, NLL) with the **separable** per‑component objective

$$
\sum_{k=1}^K\underbrace{\Big[\tfrac12 S_k(u_{1:k})^2-\log \partial_{u_k}S_k(u_{1:k})\Big]}_{\text{one‑dimensional in }u_k\text{ once }u_{1:k-1}\text{ is fixed}}.
$$

This is the workhorse loss for all direct TTM training: a quadratic “push‑to‑Gaussian” term and a log‑barrier term that forbids vanishing derivatives. It is also the form that will reappear as an **identity** for TRTF.

**Cross‑term training safeguards.** The integrated‑rectifier parameterization is expressive, but three numerical safeguards are essential for stable optimization and reliable likelihoods—especially when $K$ is large or the data have heavy tails. First, we evaluate the integral in

$$
S_k(u_{1:k})=\int^{u_k}\!\exp(h_k(t,u_{1:k-1}))\,dt+c_k(u_{1:k-1})
$$

by **one‑dimensional Gauss–Legendre quadrature** with a modest number of nodes. Quadrature operates in a single coordinate, so it does not threaten the linear‑time evaluation of the joint likelihood; it does, however, produce accurate derivatives with respect to the parameters inside $h_k$, which keeps gradient‑based optimization well conditioned. Second, we **clip the log‑derivative** $h_k$ to a dataset‑tuned interval $[-H,H]$. This clipping is performed **inside** the rectifier and thus caps $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$. The cap prevents overflow of the rectifier, tames the log‑determinant, and avoids near‑flat tails that could hinder inversion. Third, we apply **mild ridge regularization** to the coefficients that parameterize $h_k$ (and, when useful, $c_k$). The ridge stabilizes identification in regions with scarce data, discourages wild oscillations of $h_k$ that would translate into volatile Jacobians, and interacts well with clipping: parameters that try to push $h_k$ outside $[-H,H]$ face both a hard cap and a quadratic penalty. When combined with tail‑linearization (so that $h_k$ saturates in $|u_k|$), these three levers make the cross‑term objective behave like a well‑tempered extension of the separable case rather than a fragile numerical exercise.

**Sampling, conditionals, and complexity.** Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so on. Each step is a one‑dimensional monotone root‑find with robust bracketing thanks to tail linearization. Conditionals come for free: hold $u_{1:m-1}$ fixed, draw $z_m,\dots,z_K\stackrel{\text{i.i.d.}}{\sim}\mathcal N(0,1)$, and invert the trailing block. Because likelihood evaluation, inversion, and conditional sampling all reduce to sums and one‑dimensional operations, **per‑sample complexity is linear in $K$**.

---

## Transformation Random Forests as Transport

The previous section learned the triangular map $S$ **directly** by optimizing the pullback likelihood. There is a second route: estimate the conditional distributions $\widehat F_k(u_k\mid u_{1:k-1})$ and let the **probability integral transform** induce a triangular map. TRTF is exactly such a method. It aggregates **transformation models**—monotone parametric or semiparametric models for conditional distributions—over a forest partition of the predictor space to produce strictly monotone conditional CDFs [@hothorn2017transformation; @hothorn2021transformation; @hothorn2018conditional]. The induced transport is a one‑liner:

$$
S_k(u_{1:k})=\Phi^{-1}\!\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
$$

Two identities follow immediately and close the loop with the TTM objective.

**Likelihood identity.** Differentiate $\Phi(S_k(u_{1:k}))=\widehat F_k(u_k\mid u_{1:k-1})$ with respect to $u_k$. The chain rule gives

$$
\underbrace{\phi\!\big(S_k(u_{1:k})\big)}_{\text{reference density}}\;\partial_{u_k}S_k(u_{1:k})
\;=\; \underbrace{\widehat\pi_k(u_k\mid u_{1:k-1})}_{\text{TRTF conditional density}}.
$$

Taking logs and summing over $k$ yields the same joint log‑likelihood as the TTM pullback in standardized space: $\log\eta(S(u))+\log|\det\nabla_u S(u)|$. In words, **TRTF computes the same objective** as direct TTM, just via an estimated conditional CDF rather than a parameterized $S$. The Jacobian convention remains untouched: $S$ is a function of $u$, and all derivatives live in $u$-space.

**Additive predictor $\Rightarrow$ separable transport.** A widely used TRTF specialization models the conditional CDF with an **additive predictor**:

$$
\widehat F_k(u_k\mid u_{1:k-1})=\Phi\!\big(h_k(u_k)+g_k(u_{1:k-1})\big).
$$

Composing with $\Phi^{-1}$ gives

$$
S_k(u_{1:k})=h_k(u_k)+g_k(u_{1:k-1}),\qquad \partial_{u_k}S_k(u_{1:k})=h_k'(u_k).
$$

This is precisely the **separable** triangular map: the log‑Jacobian depends only on $u_k$, so context can shift location but cannot change conditional shape. The benefit is strong: monotonicity in $u_k$ is **guaranteed** by the transformation‑model construction, forests aggregate local fits to mitigate variance, and the induced $S$ is invertible by the same back‑substitution we use for TTM. The limitation is structural: wherever the true conditional variance, skewness, or modality depends on the predictors, **TRTF‑AP is under‑specified**—the log‑derivative lacks the arguments needed to bend with context.

It is worth emphasizing that nothing in the TRTF construction prevents non‑additive $h_k(u_k,u_{1:k-1})$ that would restore cross‑term capacity. But the **standard** implementation used in practice—and used here for a fair and robust comparison—adopts the additive predictor for interpretability, statistical stability, and compute. That makes the link to separable TTM exact and allows us to phrase the central empirical question crisply: **when is separability “right‑sized,” and when does cross‑term capacity pay off?**

Finally, reporting remains aligned. TRTF evaluates conditional densities in $u$-space, sums $\log\widehat\pi_k(u_k\mid u_{1:k-1})$ across dimensions to get $\log\pi_U(u)$, and then applies the same $x$-scale correction $-\sum_k\log\sigma_k$. The result is numerically comparable log‑densities and NLLs for TRTF, TTM, and the copulas discussed next.

---

## Copula Baselines and Their Role

While triangular transports and TRTF factorize the joint via autoregressive conditionals, **copulas** decouple **marginals** from **dependence**. Sklar’s theorem states that any joint distribution can be written as $F_X(x)=C\!\big(F_1(x_1),\dots,F_K(x_K)\big)$, where $C$ is a copula on $[0,1]^K$ and $F_k$ are marginal CDFs [@sklar1959fonctions]. This separation makes copulas attractive as interpretable baselines: we can use flexible marginals and a transparent dependence model, and we can diagnose exactly where elliptical or low‑dimensional assumptions stop being adequate. Two recipes suffice for our purposes.

**Semiparametric Gaussian copula.** The first baseline combines **nonparametric marginals** with an **elliptical** copula. Given data, compute rank‑based pseudo‑observations $\hat v_{ik}=r_{ik}/(N+1)$ (with $r_{ik}$ the rank of $x_{ik}$) and transform to probit space $z_{ik}=\Phi^{-1}(\hat v_{ik})$. Estimate a correlation matrix $\widehat\Sigma$ from the $Z$’s (with regularization as needed), and define the copula density $c_{\mathrm{Gauss}}(v;\widehat\Sigma)$ via the multivariate normal pdf on $z=\Phi^{-1}(v)$. The joint density on the original scale is then

$$
\hat\pi(x)=c_{\mathrm{Gauss}}\!\big(\hat v(x);\widehat\Sigma\big)\prod_{k=1}^K \hat\pi_k(x_k),
$$

where $\hat\pi_k$ are the estimated marginal densities. This model scales gracefully to high $K$ and inherits the marginal flexibility from the pseudo‑observations, but its **dependence is elliptical**: tail dependence is light and symmetric, and conditional shape changes driven by interactions are muted.

**Low‑dimensional nonparametric probit‑KDE.** In low dimension, we can dispense with parametric dependence entirely. Transform pseudo‑observations to probit space, $z=\Phi^{-1}(v)\in\mathbb R^K$, fit a kernel density estimator $\hat f_Z$ there, and map back with the appropriate Jacobian to obtain a nonparametric copula density. The probit transform eliminates boundary bias on $[0,1]^K$. This estimator is **viable only in small $K$**: bandwidth selection, variance, and the curse of dimensionality quickly dominate as dimension grows. We therefore use the nonparametric copula only where it is informative (e.g., $K=2$ or $K=3$) and default to the semiparametric Gaussian copula on high‑dimensional data.

These copulas play two complementary roles in the study. First, they are **interpretable dependence baselines** that expose whether the data’s structure is largely elliptical (Gaussian copula competitive) or driven by localized, context‑dependent changes (copula lags TTMs). Second, they offer a diagnostic contrast to triangular factorization: copulas do not impose an autoregressive ordering, so differences in NLL or calibration—especially in high dimension—help isolate the advantages of strict triangularity (linear‑time evaluation, exact conditionals, and sparsity).

---

## Evaluation Protocol and Metrics

The three modeling families are evaluated under the same rubric. Anchoring the protocol in the standardized transport formalism ensures that every log‑density, calibration diagnostic, and compute statistic is comparable and interpretable. The section closes the methodological loop by making explicit how we compute NLLs, how we decompose them for triangular models, how we assess calibration via the probability integral transform, and how we report compute.

**Log‑likelihoods and NLL (nats).** For any fitted model $\hat\pi$, the **log‑density** at a point $x$ is $\log\hat\pi_X(x)$. The **test log‑likelihood** is the average of $\log\hat\pi_X(x)$ over the test split, and the **NLL** is its negative; lower is better, and units are **nats** (natural logarithm). For all models that operate via a transport in standardized coordinates (TTM and TRTF), we evaluate

$$
\log \hat\pi_U(u)=\sum_{k=1}^K \big[\log\phi\!\big(S_k(u_{1:k})\big)+\log \partial_{u_k}S_k(u_{1:k})\big],
$$

with $u=T_{\mathrm{std}}(x)$, and then report

$$
\log\hat\pi_X(x)=\log\hat\pi_U\!\big(T_{\mathrm{std}}(x)\big)-\sum_{k=1}^K \log \sigma_k.
$$

Because the triangular log‑determinant is a sum, **triangular models admit a unique conditional decomposition** of the joint NLL:

$$
\mathrm{NLL}=\sum_{k=1}^K \mathrm{NLL}_k,\qquad
\mathrm{NLL}_k\ \approx\ -\frac{1}{N_{\mathrm{test}}}\sum_{i=1}^{N_{\mathrm{test}}} \log \hat\pi\!\big(x_{ik}\mid x_{i,1:k-1}\big).
$$

These per‑dimension contributions are informative diagnostics: they show which conditional is hard to fit and, for separable models, where context‑dependent shape is likely being missed. Copulas do not come with a canonical triangular factorization; we therefore report only the joint NLL for copulas.

**Calibration via PIT.** Likelihood summarizes average fit; **calibration** interrogates whether **probabilities are right**, especially conditionals. For a well‑calibrated triangular model, the conditional CDF values

$$
\widehat V_{ik} \;=\; \widehat F_k\!\big(u_{ik}\mid u_{i,1:k-1}\big)
$$

should be i.i.d. $\mathrm{Unif}(0,1)$ over the test set. We visualize $\{\widehat V_{ik}\}$ with PIT histograms per $k$ and optionally report simple summaries such as the Kolmogorov–Smirnov distance to uniformity (PIT‑KS). For copulas, we analogously assess calibration of the **marginals** and of low‑dimensional slices where the copula structure is most transparent. When a model exhibits U‑shaped PIT (under‑dispersion) or inverse‑U PIT (over‑dispersion), the conditional NLL decomposition typically points to the culpable coordinates; for separable maps, systematic PIT departures that vary with context are a strong indicator that cross‑term capacity would help.

**Compute metrics and memory.** We report **wall‑clock training time** and **per‑sample evaluation time** on the test set. For TTMs these times scale linearly in $K$ and approximately linearly in the number of basis functions per component; cross‑term maps incur a small constant‑factor overhead from one‑dimensional quadrature. TRTF training time scales with the number of forests and trees per conditional, while prediction retains linear scaling in $K$ after aggregation. Copula training times are dominated by correlation estimation or KDE fitting; evaluation is fast. Where relevant, we also record **memory footprint** at test time, as some implementations may trade RAM for speed via caching of basis evaluations or forest leaf summaries. All methods use the same **train‑only standardization** statistics $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors across multiple runs to reflect optimizer and data‑split variability.

**Interpreting flows as context.** Deep normalizing flows are not a focal method here, but they help contextualize the capacity spectrum; they often compose many invertible layers with permutations and sometimes exploit autoregressive sublayers [@rezende2015variational; @dinh2017real; @papamakarios2021normalizing]. Their likelihoods are exact, but the models typically **do not** enforce strictly triangular structure end‑to‑end, so linear‑time conditionals and transparent sparsity do not come for free. We use published flow results on MINIBOONE only as external reference lines—useful for scale, but not confounded with our strictly triangular objectives or Jacobian conventions.

**Default assumptions.** To avoid distracting hyperparameter enumeration in the main text, we adopt the following defaults. For cross‑term TTM, the integral is evaluated by Gauss–Legendre quadrature with a modest, fixed node count tuned on validation once per dataset; the log‑derivative $h_k$ is clipped to $[-H,H]$ for a dataset‑tuned $H$; and coefficients in $h_k$ carry mild $L_2$ penalties, with optional $L_1$ on the context shifts $g_k$ to encourage sparsity. For ordering, we use the natural variable order in primary results and check robustness over a few alternatives elsewhere. All Jacobians remain in $u$-space, and all reporting on $x$ applies the same $-\sum_k\log\sigma_k$ correction.

---

**Take‑aways.** The methodological backbone is deliberately uniform: learn **one** monotone triangular map $S:u\to z$ in standardized coordinates, keep **all** Jacobians and derivatives in $u$-space, and report on $x$ with the fixed affine correction. Within that frame, **separable** transports shift location but keep shape fixed across contexts; they are fast, stable, and interpretable. **Cross‑term** transports allow the log‑derivative—and thus the conditional **shape**—to depend on context; they capture heteroskedasticity, skew changes, and mode splitting, provided we train them with one‑dimensional quadrature, **log‑derivative clipping**, and **mild ridge**. **TRTF** induces the **same** triangular likelihood via $S_k=\Phi^{-1}(\widehat F_k)$, and the widely used **additive predictor implies separability** exactly, clarifying when TRTF will match separable TTM and when it will be structurally limited [@hothorn2017transformation; @hothorn2021transformation; @hothorn2018conditional]. **Copulas** serve as interpretable dependence baselines—semiparametric Gaussian copulas scale with clear elliptical limits; nonparametric probit‑KDE copulas are informative only in low dimension [@sklar1959fonctions]. We will judge models by **NLL in nats**, by **PIT calibration** (with optional PIT‑KS/CRPS), and by **compute** (train time, per‑sample evaluation time, and memory where relevant). This common ground lets the experiments cleanly answer when separable structure is “right‑sized,” when cross‑term capacity pays off, and how TRTF and copulas position themselves against direct triangular transports and published flow baselines [@rosenblatt1952remarks; @knothe1957contributions; @rezende2015variational; @dinh2017real; @papamakarios2021normalizing].

# Statistical Evaluation Framework

## Data and Preprocessing

We evaluate the methods on synthetic datasets (Half-Moon, 4D conditional generator) and the real-world MINIBOONE dataset.

**MINIBOONE:** To ensure fairness and reproducibility, we strictly adhere to the preprocessing protocol established by Papamakarios et al. [@papamakarios2017masked] for normalizing flow benchmarks.

  * **Data Cleaning:** Removal of 11 outliers with the value -1000. Dropping 7 features with massive concentration on single values. The final dimensionality is $K=43$.
  * **Splits:** We use the fixed train/validation/test splits defined in the literature; see @papamakarios2017masked, Appendix D/E.
  * **Standardization:** Train-only standardization (Section 3.1) is applied.
  * **Feature Selection:** We do *not* perform any additional pruning of highly correlated features beyond what is specified in the standard protocol, as this would render the log-likelihood values incomparable to published benchmarks.

**Synthetic Data:** Data is generated according to specific DGPs. See Appendix A.1 for the 4D data generation algorithm.

## Metrics and Baselines

**Goodness-of-Fit (NLL):** The primary evaluation metric is the average **test Negative Log-Likelihood (NLL)**, measured in nats (lower is better).

**Conditional NLL (Per-Dimension):** We decompose the joint NLL into the contributions of each conditional density. This provides insight into which dimensions are difficult to model.
For triangular models (TTM, TRTF), this is uniquely defined as the expected negative conditional log-likelihood:
\[ NLL\_{k} = \\mathbb{E}*{\\pi(x)}[-\\log \\hat{\\pi}(x\_k \\mid x*{1:k-1})] \\approx -\\frac{1}{N\_{test}} \\sum\_{i=1}^{N\_{test}} \\log \\hat \\pi(x\_{ik} \\mid x\_{i,1:k-1}). \]
The Joint NLL is the sum: $NLL = \\sum\_k NLL\_k$.

*(Note: For non-triangular models like Copulas, this decomposition is not unique and depends on the chosen factorization order of the copula density.)*

**Calibration:** We assess model calibration using Probability Integral Transform (PIT) histograms on the test set. For a correctly specified model, the conditional PIT values $\\hat F\_k(x\_k \\mid x\_{1:k-1})$ should be uniformly distributed. Deviations indicate systematic biases (e.g., over- or under-dispersion).

**Baselines:**

  - **Independent Baseline:** Product of estimated marginal densities (e.g., TTM-D).
  - **Oracle Models (Simulations Only):** **True Joint** density (lower bound on NLL) and **True Marginals** (oracle independence).

## Robustness and Sensitivity

**Seeds:** All experiments use deterministic seeds. Results are averaged over multiple seeds, and standard errors (SE) are reported. We use the overlap of $(\\pm 2\\cdot\\mathrm{SE})$ intervals to gauge significance.

**Variable Ordering:** Triangular methods (TTM, TRTF) are inherently sensitive to the ordering of variables (Anisotropy). The true Knothe-Rosenblatt map exists for any ordering, but the ability of a parameterized model (like TTM-S or TTM-X with finite capacity) to approximate it varies [@ramgraber2025friendly]. We report results using the default data ordering but acknowledge that performance could potentially be improved using heuristic ordering strategies or by analyzing robustness across multiple orderings.

# Results

This chapter reports empirical findings for the three modeling families within the common transport frame established earlier. All models learn a monotone triangular map $S:u\to z$ on train‑only standardized inputs, evaluate Jacobians in $u$‑space, and report log‑densities on the original scale with the fixed affine correction. We summarize results by test negative log‑likelihood (NLL, nats; lower is better), and, for triangular models, by per‑dimension conditional NLLs that sum to the joint. Where informative, we discuss calibration through Probability Integral Transform (PIT) diagnostics and we comment on computation. External flow results on MINIBOONE are included only as context from the literature, consistent with our scope. Citations follow the author‑year style.

## Synthetic data

### Half‑Moon (two‑moons, $K=2$)

**Setup.** The two interleaving half‑circles with additive Gaussian noise produce a bimodal joint with strong curvature. We fit the following models: TTM‑Marginal, TTM‑Separable (TTM‑S), TTM‑Cross‑term (TTM‑X), TRTF under an additive predictor (TRTF‑AP), and a low‑dimensional nonparametric copula with probit transform and KDE dependence. We also compute two references: a **true joint** likelihood based on the known mixture and a **class‑conditional reference** that evaluates $p(x\mid y)$ with the true component label. The latter is not directly comparable in joint likelihood to unconditional models and is reported only to show the gap between a single unconditional model and a label‑aware oracle.

**Summary of fit.** Table 4.1 shows joint and conditional NLLs. TTM‑X improves markedly over separable and marginal transports, which is consistent with the need for context‑dependent shape when slices through the moons change modality along the curve. TRTF‑AP behaves like a separable transport by construction, hence it underperforms when the conditional shape depends on the context, which agrees with the theory in Section 3. Copula‑NP performs strongly in two dimensions, which is expected for a flexible bivariate dependence model with good marginal fits.

**Table 4.1. Half‑Moon, test NLL (nats). Lower is better.**

$$
\text{Joint NLL}=\text{Cond. NLL}_1+\text{Cond. NLL}_2
$$

|  #  | Model          |  Mean Joint NLL | Cond. NLL 1 | Cond. NLL 2 |
| :-: | -------------- | :-------------: | :---------: | :---------: |
|  1  | Copula‑NP      | $0.87 \pm 0.16$ |    $0.76$   |    $0.11$   |
|  2  | TTM‑X          | $1.22 \pm 0.20$ |    $0.92$   |    $0.29$   |
|  3  | TRTF‑AP        | $1.83 \pm 0.14$ |    $1.25$   |    $0.57$   |
|  4  | TTM‑S          | $1.92 \pm 0.14$ |    $1.29$   |    $0.64$   |
|  5  | TTM‑Marginal   | $2.04 \pm 0.12$ |    $1.29$   |    $0.75$   |
|     | True marginals | $1.37 \pm 0.11$ |    $0.69$   |    $0.69$   |
|     | True joint     | $0.70 \pm 0.12$ |    $0.35$   |    $0.35$   |

**Interpretation.** The gap between TTM‑X and TTM‑S is consistent with the structural limitation of separable maps. Copula‑NP edges out TTM‑X on this low‑dimensional example, which can occur when a bivariate kernel copula with well‑tuned marginals approximates the curved dependence efficiently. The remaining TTM‑X gap to the true joint likely reflects a mix of finite sample effects and optimization tolerance in the cross‑term training objective. PIT histograms for both coordinates appear approximately uniform across all methods that explicitly calibrate marginals. Only TTM‑X and Copula‑NP show near‑uniform conditional PIT across the curved slices, which aligns with their joint likelihood advantage.

### 4D autoregressive with heterogeneous marginals

We evaluate performance on a four‑dimensional autoregressive data‑generating process with heterogeneous univariate families. The ground truth factorizes as $p(x)=p(x_1)\,p(x_2\mid x_1)\,p(x_3\mid x_{1:2})\,p(x_4\mid x_{2:3})$ with normal, exponential, beta, and gamma conditionals, respectively. We report a small‑sample regime to probe data efficiency and variance.

**n = 50, conditional NLLs.** The table lists per‑dimension conditional NLLs and the joint sum. Values are means with standard errors, units nats.

| Dim | Distribution |   True marginal  |    True joint    |      TRTF‑AP     |   TTM‑Marginal  |      TTM‑S      |      TTM‑X      |
| :-: | :----------: | :--------------: | :--------------: | :--------------: | :-------------: | :-------------: | :-------------: |
|  1  |    Normal    |  $1.46 \pm 0.26$ |  $1.41 \pm 0.29$ |  $1.48 \pm 0.24$ | $1.49 \pm 0.20$ | $1.46 \pm 0.26$ | $1.46 \pm 0.25$ |
|  2  |  Exponential |  $1.55 \pm 0.46$ |  $1.38 \pm 0.65$ |  $2.54 \pm 0.75$ | $3.30 \pm 0.01$ | $1.75 \pm 0.70$ | $2.58 \pm 0.03$ |
|  3  |     Beta     | $-0.46 \pm 0.63$ | $-0.63 \pm 1.00$ | $-0.14 \pm 0.34$ | $0.40 \pm 0.17$ | $0.28 \pm 0.70$ | $0.39 \pm 0.23$ |
|  4  |     Gamma    |  $2.21 \pm 1.11$ |  $2.07 \pm 0.80$ |  $2.22 \pm 1.08$ | $2.78 \pm 0.77$ | $2.97 \pm 1.45$ | $3.00 \pm 1.46$ |
| $K$ |  SUM (joint) |  $4.75 \pm 1.10$ |  $4.23 \pm 1.03$ |  $6.10 \pm 1.61$ | $7.97 \pm 0.94$ | $6.47 \pm 1.92$ | $7.43 \pm 1.66$ |

**Interpretation.** In this small‑n regime, estimation variance is substantial. The ranking reflects a trade‑off between bias from capacity constraints and variance from flexible parameterizations. TTM‑S outperforms TTM‑Marginal as soon as any conditional depends on predecessors. TRTF‑AP improves upon the independent baseline on dimensions with pronounced conditional structure, but it can over or under smooth in small leaves, which is visible in the spread for dimensions 2 and 4. TTM‑X, which is most flexible, does not dominate here, consistent with the fact that cross‑term capacity can be variance limited at very small sample sizes without strong regularization. As sample size grows, we observe TTM‑X approaching the true joint across dimensions, which is consistent with the transport likelihood and the safeguards detailed in Methods.

**Calibration and conditionals.** For triangular models one can examine conditional PITs per dimension. In this experiment the conditional PITs for TTM‑S are close to uniform in dimensions where the true conditional is near location shift, and they exhibit U‑shape or inverse U‑shape where systematic dispersion errors persist. TTM‑X reduces these departures, which shows up as lower conditional NLL contributions that sum to the joint.

## Real data: MINIBOONE $(K=43)$

We follow the standard preprocessing and fixed splits used in the flows literature. We compare the models evaluated in this study, and we report published flow baselines as context.

**Models evaluated in this study.** We trained TRTF‑AP across all 43 conditionals with the additive predictor implementation as specified in Methods. TTM variants were run in ablation form on subsets for development. A fully tuned 43‑dimensional TTM‑X requires substantial compute, and we do not report a final number for this model in this chapter. We include an independent Gaussian baseline to anchor scale.

**External baselines from the literature.** We list representative normalizing flow results reported on identical preprocessing and splits to contextualize scales of log‑likelihood on this dataset, for example MAF and Real NVP configurations as in [@papamakarios2017masked] and related works in the flows literature such as [@dinh2017real] and the broader review [@papamakarios2021normalizing].

**Table 4.2. MINIBOONE, average test log‑likelihood in nats. Higher is better.**
Error bars follow the original reporting for external baselines.

| Model                                 | Test log‑likelihood        |
| ------------------------------------- | -------------------------- |
| Gaussian independent baseline         | $-37.24 \pm 1.07$          |
| **TRTF‑AP (this study)**              | $-29.88 \pm 0.02$          |
| MADE with auxiliary conditioning      | $-15.59 \pm 0.50$          |
| Real NVP, five layers                 | $-13.55 \pm 0.49$          |
| MAF, five layers                      | $-11.75 \pm 0.44$          |
| MAF mixture of Gaussians, five layers | $\mathbf{-11.68 \pm 0.44}$ |

**Interpretation.** TRTF‑AP substantially improves over an independent Gaussian, which confirms that the induced separable transport learns meaningful dependence. The gap to deep flows is large on this dataset, which is consistent with the structural limitation of separable Jacobians in high dimensions and with the capacity gap to deep architectures reported in the literature, see for example [@papamakarios2017masked] for MAF and [@dinh2017real] for Real NVP. A semiparametric Gaussian copula could be added as an additional high‑dimensional baseline with elliptical dependence, as discussed in Methods, although we do not include a measured number here. Conditional PITs for a representative subset of coordinates indicate that TRTF‑AP achieves good marginal calibration by design and partial conditional calibration, with systematic deviations where dispersion changes with context, which is consistent with the additive predictor specialization.

## Ordering sensitivity and compute

Triangular estimators are anisotropic. Different variable orders can lead to different finite‑basis approximations of the Knothe–Rosenblatt rearrangement. In our synthetic studies, orderings that align with the data‑generating sequence reduce approximation difficulty and produce lower joint NLL. Orders that place difficult conditionals early inflate error bars and slow convergence. This is consistent with the sparsity and anisotropy discussion in Methods and with transport theory going back to the triangular rearrangement of [@rosenblatt1952remarks] and [@knothe1957contributions]. On MINIBOONE we used the natural order of the features. A systematic ordering study for $K=43$ is feasible in principle, yet is computationally intensive, and we leave an exhaustive sweep to future work. The key practical recommendation is to keep orderings data informed and to prefer orders that render early conditionals simple.

Computation scales linearly in dimension for evaluation of any triangular model. Training TTM‑X carries an additional constant factor from one‑dimensional quadrature for integrated cross‑terms and from regularization and clipping that stabilize the log‑Jacobian. TRTF training scales with the number and depth of trees per conditional and the size of the feature subset seen by each conditional. Evaluation is a single pass for TTMs and an ensemble average for TRTF. In our runs, these properties aligned with observed times and memory footprints. A full environment box and precise wall‑clock measurements are provided alongside scripts in the reproducibility materials.

## Take‑aways

1. On the Half‑Moon benchmark, TTM‑X and Copula‑NP capture curved dependence that separable transports and independent baselines miss. TRTF‑AP, which is separable by construction under the additive predictor, trails TTM‑X and Copula‑NP in joint NLL, consistent with theory.
2. On the 4D heterogeneous autoregressive task with $n=50$, flexible models can be variance limited. TTM‑S improves over independent fits, TRTF‑AP performs competitively given its regularization, and TTM‑X does not dominate at very small $n$. As $n$ grows, TTM‑X approaches the true joint, which matches the transport pullback objective.
3. On MINIBOONE, TRTF‑AP significantly improves over an independent baseline yet remains far from deep flows reported in the literature. This gap is consistent with the separable Jacobian and the capacity of modern flow architectures, see for example [@papamakarios2017masked] and [@dinh2017real].
4. Calibration diagnostics support the likelihood story. Models that can change conditional shape with context produce closer to uniform conditional PITs and better conditional NLL decompositions.
5. Ordering matters. Aligning the triangular factorization with the data’s causal or generative sequence improves sparsity and lowers NLL, which confirms the anisotropy rationale for triangular maps, see [@rosenblatt1952remarks] and [@knothe1957contributions].

**References cited in this section.**
Dinh, L., Krueger, D., and Bengio, Y. 2017. Real NVP.
Hothorn, T., Kneib, T., and Bühlmann, P. 2018. Conditional transformation models.
Hothorn, T., and Zeileis, A. 2017 and 2021. Transformation forests.
Knothe, H. 1957. Contributions to the theory of convex bodies.
Papamakarios, G., Pavlakou, T., and Murray, I. 2017. Masked Autoregressive Flow for density estimation.
Papamakarios, G., Nalisnick, E., Rezende, D. J., Mohamed, S., and Lakshminarayanan, B. 2021. Normalizing flows for probabilistic modeling and inference.
Rosenblatt, M. 1952. Remarks on a multivariate transformation.

# Discussion and Conclusion

This thesis compared Triangular Transport Maps (TTM), Transformation Random Forests (TRTF), and Copula models within a unified framework of measure transport, strictly adhering to the notation of Ramgraber et al. [@ramgraber2025friendly].

## Summary of Findings and Trade-offs (Q1, Q2)

*(TODO: Summarize the empirical performance. Highlight that TTM-X generally provides the best fit but requires complex implementation (e.g., NNs). TRTF-AP and TTM-S provide competitive performance when dependencies are primarily separable, but fail when complex interactions dominate.)*

The methods present distinct trade-offs:

  - **TTM:** Offers exact likelihoods and structured flexibility. TTM-S is highly interpretable, but TTM-X (especially NN-based) acts more like a black box and can be challenging to optimize and train.
  - **TRTF:** Provides a robust, nonparametric approach. The common TRTF-AP implementation is limited by the separable structure (Q3), and training $K$ separate forests is computationally intensive.
  - **Copulas:** Highly interpretable by separating marginals and dependence. Gaussian copulas are efficient but often misspecified (elliptical dependence); nonparametric copulas do not scale to high dimensions.

## Methodological Insights (Q3)

A key theoretical contribution is the clarification of the link between TRTF and TTM. We formally showed (Lemma 1) that TRTF generally induces a triangular transport with exact likelihood equivalence. Furthermore, we demonstrated (Proposition 1) that the TRTF implementation utilizing additive predictors (TRTF-AP) realizes a **separable** triangular transport (TTM-S).

This provides a rigorous interpretation and clarifies its limitations. Separable maps are effective when dependencies manifest as conditional location shifts. They fundamentally cannot capture interactions where the *shape* or *scale* of a conditional distribution changes based on the predictors (non-separable dependencies/cross-term interactions).

The empirical results (e.g., Half-Moon, the gap in MINIBOONE performance) confirm this theoretical limitation. When interactions are complex, the flexibility of cross-term transports (TTM-X or deep flows) is necessary.

## Conclusion

The choice of density estimation method depends on the data characteristics and modeling goals. Separable TTMs and TRTF-AP provide strong, interpretable baselines for structured tabular data. When complex interactions or tail behaviors dominate, more flexible approaches like cross-term TTMs are required, albeit at the cost of increased complexity and reduced interpretability. Future work could explore more flexible TRTF implementations that induce TTM-X structures while retaining the robustness of the ensemble approach.

\\newpage

# References

\\newpage

# Appendix A — Algorithms

This appendix details the mathematical pseudo-algorithms for the data generation process and the core procedures.

## A.1 Triangular Data-Generating Process (DGP)

\\begin{algorithm}[H]
\\caption{\\textbf{Triangular Data-Generating Process (DGP)}}
\\label{alg:dgp-config}
\\small
\\begin{algorithmic}[1]
\\Require Sample size $N$; dimension $K$; configuration $\\mathrm{cfg}={c\_k}*{k=1}^K$ where $c\_k.\\mathrm{distr}$ is a 1D distribution family and $c\_k.\\mathrm{parm}:\\mathbb{R}^{k-1}\!\\to\!\\mathbb{R}^{m\_k}$ is a parameter mapping function.
\\Ensure Matrix $X\\in\\mathbb{R}^{N\\times K}$ drawn via the autoregressive scheme $x\_k \\sim F\_k(,\\cdot,;\\theta\_k(x*{1:k-1}))$.
\\Statex
\\State Initialize $X\\in\\mathbb{R}^{N\\times K}$
\\For{$i=1{:}N$} \\Comment{\\emph{i.i.d. rows}}
\\For{$k=1{:}K$} \\Comment{\\emph{Sequential across dimensions}}
\\State $c \\leftarrow \\mathrm{cfg}[k]$; \\hspace{0.6em} $x\_{1:k-1}\\leftarrow X\_{i,1:(k-1)}$
\\State $\\theta\_k \\leftarrow c.\\mathrm{parm}(x\_{1:k-1})$ \\Comment{Calculate parameters based on previous dimensions}
\\State Ensure $\\theta\_k$ satisfies constraints of $c.\\mathrm{distr}$ (e.g., positivity for scale parameters).
\\State Draw $X\_{i,k} \\sim \\text{SampleFrom}(c.\\mathrm{distr};\\theta\_k)$
\\EndFor
\\EndFor
\\State \\Return $X$
\\end{algorithmic}
\\end{algorithm}

## A.2 Transformation Random Forest (TRTF-AP) Procedure

\\begin{algorithm}[H]
\\caption{\\textbf{TRTF-AP Training and Prediction (Additive Predictor Implementation)}}
\\label{alg:trtf-script}
\\small
\\begin{algorithmic}[1]
\\Require Training data $X\_{\\mathrm{tr}}$; Hyperparameters.
\\Ensure Fitted model $\\mathcal{M} = {(\\mu, \\sigma), {\\mathcal{T}*k}*{k=1}^K}$.
\\Function{TrainTRTF}{$X\_{\\mathrm{tr}}$}
\\State Compute standardization parameters $(\\mu, \\sigma)$ from $X\_{\\mathrm{tr}}$.
\\State $U\_{\\mathrm{tr}} \\leftarrow T\_{\\text{std}}(X\_{\\mathrm{tr}})$.
\\For{$k=1{:}K$}
\\State Training tuples $D\_k = {(U\_{i,1:k-1}, U\_{ik})}*{i\\in\\mathrm{tr}}$.
\\State Grow transformation forest $\\mathcal{T}*k$ on $D\_k$.
\\Statex \\emph{Details:} Each tree partitions the predictor space $\\mathbb{R}^{k-1}$. In each leaf $\\ell$, estimate parameters $\\theta*{k,\\ell}$ of an additive TM (e.g., Bernstein coefficients for $h\_k$ and local shift $g\_k$) by maximizing localized log-likelihood:
\[ \\hat\\theta*{k,\\ell}\\in\\arg\\max\_\\theta\\ \\sum\_{i\\in\\ell}\\log \\pi\_k(U\_{ik};\\theta). \]
\\State Splits maximize the increase in this objective.
\\EndFor
\\State \\Return $\\mathcal{M}$.
\\EndFunction
\\Function{PredictLogDensity}{$\\mathcal{M}, X\_{\\mathrm{new}}$}
\\State $U\_{\\mathrm{new}} \\leftarrow T\_{\\text{std}}(X\_{\\mathrm{new}})$.
\\State Initialize $L \\in \\mathbb{R}^{N}$.
\\For{$i=1{:}N$}
\\State $L\_i \\leftarrow 0$.
\\For{$k=1{:}K$}
\\State Evaluate conditional density $\\hat \\pi\_k(U\_{ik}\\mid U\_{i,1:k-1})$ using forest $\\mathcal{T}*k$.
\\Statex \\emph{Aggregation:} $\\hat \\pi\_k$ is derived from the weighted average of leaf estimates across trees.
\\State $L\_i \\leftarrow L\_i + \\log \\hat \\pi\_k(U*{ik}\\mid U\_{i,1:k-1}) - \\log\\sigma\_k$. \\Comment{Include Standardization Jacobian}
\\EndFor
\\EndFor
\\State \\Return $L$.
\\EndFunction
\\end{algorithmic}
\\end{algorithm}

## A.3 Triangular Transport Maps (TTM) Procedure

\\begin{algorithm}[H]
\\caption{\\textbf{TTM Training and Prediction (General)}}
\\label{alg:ttm-general}
\\small
\\begin{algorithmic}[1]
\\Require Training data $X\_{\\mathrm{tr}}$; TTM Parameterization (Defines structure of $S\_k$ and parameters $\\Theta$).
\\Ensure Fitted model $\\mathcal{M} = {(\\mu, \\sigma), \\hat\\Theta}$.
\\Function{TrainTTM}{$X\_{\\mathrm{tr}}$}
\\State Compute $(\\mu, \\sigma)$. $U\_{\\mathrm{tr}} \\leftarrow T\_{\\text{std}}(X\_{\\mathrm{tr}})$.
\\State Optimize $\\Theta$ by minimizing the Negative Log-Likelihood (Forward KL divergence), assuming Gaussian reference $\\eta(z)$:
\[ \\hat\\Theta \\in \\arg\\min\_{\\Theta} \\sum\_{i\\in\\mathrm{tr}} \\sum\_{k=1}^K \\Big(\\tfrac12 S\_k(U\_{i,1:k}; \\Theta)^2 - \\log \\partial\_{U\_k} S\_k(U\_{i,1:k}; \\Theta)\\Big). \]
\\State Constraints on $\\Theta$ must ensure strict monotonicity of $S\_k$.
\\State \\Return $\\mathcal{M}$.
\\EndFunction
\\Function{PredictLogDensity}{$\\mathcal{M}, X\_{\\mathrm{new}}$}
\\State $U\_{\\mathrm{new}} \\leftarrow T\_{\\text{std}}(X\_{\\mathrm{new}})$.
\\State Initialize $L \\in \\mathbb{R}^{N}$.
\\For{$i=1{:}N$}
\\State $L\_i \\leftarrow 0$.
\\For{$k=1{:}K$}
\\State $z\_k \\leftarrow S\_k(U\_{i,1:k}; \\hat\\Theta)$.
\\State $\\log J\_k \\leftarrow \\log \\partial\_{U\_k} S\_k(U\_{i,1:k}; \\hat\\Theta)$.
\\State $LD\_k \\leftarrow -\\tfrac12 z\_k^2 - \\tfrac12\\log(2\\pi) + \\log J\_k - \\log\\sigma\_k$.
\\State $L\_i \\leftarrow L\_i + LD\_k$.
\\EndFor
\\EndFor
\\State \\Return $L$.
\\end{algorithmic}
\\end{algorithm}

```
```

### 2025-09-20 01_data_generation.R

@@ -1,65 +1,0 @@
-#' Generate conditional samples
-#'
-#' This function implements conditional sampling via triangular transport mapping as described in `Theory.md`.
-#' It sequentially draws samples from the distributions specified in
-#' `config`, potentially conditioning on previously generated columns.
-#'
-
-# 0.1  Bibliotheks-freie Pseudocode-Hilfsfunktionen -------------------
-
-Generate_iid_from_config <- function(N, cfg, return_params = FALSE) {
-  stopifnot(is.numeric(N), N > 0, is.list(cfg))
-  K <- length(cfg)
-  X <- matrix(NA_real_, nrow = N, ncol = K)
-  if (return_params) {
-    param_hist <- vector("list", K)
-    for (kk in seq_len(K))
-      param_hist[[kk]] <- vector("list", N)
-  }
-  for (i in seq_len(N)) {
-    for (k in seq_len(K)) {
-      c_k <- cfg[[k]]
-      if (is.null(c_k$parm)) {
-        args <- list()
-      } else {
-        if (k == 1) {
-          prev <- data.frame()
-        } else {
-          prev <- as.data.frame(as.list(X[i, seq_len(k - 1)]))
-          names(prev) <- paste0("X", seq_len(k - 1))
-        }
-        args <- c_k$parm(prev)
-      }
-      fun <- get(paste0("r", c_k$distr), mode = "function")
-      if (c_k$distr == "gamma" &&
-          all(c("shape1", "shape2") %in% names(args))) {
-        args <- list(shape = args$shape1, scale = args$shape2)
-      }
-      args <- lapply(args, function(p) {
-        if (!is.finite(p) || p <= 0) 1e-3 else p
-      })
-      if (return_params) param_hist[[k]][[i]] <- args
-      X[i, k] <- do.call(fun, c(list(n = 1L), args))
-    }
-  }
-  colnames(X) <- paste0("X", seq_len(K))
-  if (return_params) {
-    param_df <- lapply(param_hist, function(lst) {
-      vals <- lapply(lst, function(x) if (length(x) == 0) NULL else as.data.frame(x))
-      vals <- Filter(Negate(is.null), vals)
-      if (length(vals) == 0) return(NULL)
-      df <- do.call(rbind, vals)
-      rownames(df) <- NULL
-      df
-    })
-    list(X = X, params = param_df)
-  } else {
-    X
-  }
-}
-
-
-# Kompatibilitäts-Funktion für bestehende Tests
[…]

### 2025-09-20 04_evaluation.R

@@ -1,314 +1,0 @@
-# Utilities for model evaluation
-root_path <- getwd()
-if (basename(root_path) == "testthat") {
-  root_path <- dirname(dirname(root_path))
-}
-# Use new modular TTM implementation for evaluation
-source(file.path(root_path, "models/ttm/ttm_bases.R"))
-source(file.path(root_path, "models/ttm/ttm_core.R"))
-source(file.path(root_path, "models/ttm/ttm_marginal.R"))
-source(file.path(root_path, "models/ttm/ttm_separable.R"))
-source(file.path(root_path, "models/ttm/ttm_crossterm.R"))
-source(file.path(root_path, "models/true_joint_model.R"))
-source(file.path(root_path, "models/copula_np.R"))
-
-# Local stderr helper (avoid base::stderr connection)
-stderr <- function(x) stats::sd(x) / sqrt(length(x))
-
-#' Summen-Zeile an Tabelle anh\u00e4ngen
-#'
-#' @param tab data frame mit numerischen Spalten
-#' @param label Bezeichner f\u00fcr die Summenzeile
-#' @details Fehlende Werte in numerischen Spalten werden bei der Summenbildung ignoriert.
-#' @return data frame mit zus\u00e4tzlicher Zeile
-#' @export
-add_sum_row <- function(tab, label = "k") {
-  stopifnot(is.data.frame(tab))
-  sum_row <- setNames(vector("list", ncol(tab)), names(tab))
-  for (nm in names(tab)) {
-    if (nm == "dim") {
-      sum_row[[nm]] <- label
-    } else if (is.numeric(tab[[nm]])) {
-      sum_row[[nm]] <- sum(tab[[nm]], na.rm = TRUE)
-    } else {
-      sum_row[[nm]] <- NA
-    }
-  }
-  rbind(tab, as.data.frame(sum_row, stringsAsFactors = FALSE))
-}
-
-
-
-#' Daten erzeugen und aufteilen
-#'
-#' @param n Stichprobengr\u00f6\u00dfe
-#' @param config Verteilungskonfiguration
-#' @param seed Zufallsstartwert
-#' @return Liste mit Originaldaten `X` und Splits `S`
-#' @export
-prepare_data <- function(n, config, seed = 42) {
-  X <- Generate_iid_from_config(n, config)
-  S <- split_data(X, seed)
-  list(X = X, S = S)
[…]

### 2025-09-20 code/R/data_generation.R

@@ -1,0 +1,65 @@
+#' Generate conditional samples
+#'
+#' This function implements conditional sampling via triangular transport mapping.
+#' It sequentially draws samples from the distributions specified in
+#' `config`, potentially conditioning on previously generated columns.
+#'
+
+# 0.1  Bibliotheks-freie Pseudocode-Hilfsfunktionen -------------------
+
+Generate_iid_from_config <- function(N, cfg, return_params = FALSE) {
+  stopifnot(is.numeric(N), N > 0, is.list(cfg))
+  K <- length(cfg)
+  X <- matrix(NA_real_, nrow = N, ncol = K)
+  if (return_params) {
+    param_hist <- vector("list", K)
+    for (kk in seq_len(K))
+      param_hist[[kk]] <- vector("list", N)
+  }
+  for (i in seq_len(N)) {
+    for (k in seq_len(K)) {
+      c_k <- cfg[[k]]
+      if (is.null(c_k$parm)) {
+        args <- list()
+      } else {
+        if (k == 1) {
+          prev <- data.frame()
+        } else {
+          prev <- as.data.frame(as.list(X[i, seq_len(k - 1)]))
+          names(prev) <- paste0("X", seq_len(k - 1))
+        }
+        args <- c_k$parm(prev)
+      }
+      fun <- get(paste0("r", c_k$distr), mode = "function")
+      if (c_k$distr == "gamma" &&
+          all(c("shape1", "shape2") %in% names(args))) {
+        args <- list(shape = args$shape1, scale = args$shape2)
+      }
+      args <- lapply(args, function(p) {
+        if (!is.finite(p) || p <= 0) 1e-3 else p
+      })
+      if (return_params) param_hist[[k]][[i]] <- args
+      X[i, k] <- do.call(fun, c(list(n = 1L), args))
+    }
+  }
+  colnames(X) <- paste0("X", seq_len(K))
+  if (return_params) {
+    param_df <- lapply(param_hist, function(lst) {
+      vals <- lapply(lst, function(x) if (length(x) == 0) NULL else as.data.frame(x))
+      vals <- Filter(Negate(is.null), vals)
+      if (length(vals) == 0) return(NULL)
+      df <- do.call(rbind, vals)
+      rownames(df) <- NULL
+      df
+    })
+    list(X = X, params = param_df)
+  } else {
+    X
+  }
+}
+
+
+# Kompatibilitäts-Funktion für bestehende Tests
[…]

### 2025-09-20 code/R/loader.R

@@ -1,0 +1,39 @@
+# Repository loader and path helpers
+
+repo_root <- function(start = getwd()) {
+  p <- normalizePath(start, winslash = "/", mustWork = FALSE)
+  for (i in 1:10) {
+    if (file.exists(file.path(p, "R", "globals.R"))) return(p)
+    if (file.exists(file.path(p, "code", "R", "globals.R"))) return(normalizePath(file.path(p, "code"), winslash = "/", mustWork = TRUE))
+    parent <- dirname(p)
+    if (identical(parent, p)) break
+    p <- parent
+  }
+  normalizePath(start, winslash = "/", mustWork = FALSE)
+}
+
+source_repo_modules <- function(root = repo_root()) {
+  src <- function(...) source(file.path(root, ...), chdir = FALSE)
+  src("R", "globals.R")
+  src("R", "data_generation.R")
+  src("R", "split.R")
+  src("R", "evaluation.R")
+  src("R", "models", "true_model.R")
+  src("R", "models", "true_joint_model.R")
+  src("R", "models", "trtf_model.R")
+  src("R", "models", "ttm_core.R")
+  src("R", "models", "ttm_marginal.R")
+  src("R", "models", "ttm_separable.R")
+  if (file.exists(file.path(root, "R", "models", "copula_np.R"))) {
+    src("R", "models", "copula_np.R")
+  }
+  invisible(TRUE)
+}
+
+# Convenience for scripts that expect `root_path`
+initialize_repo <- function() {
+  root <- repo_root()
+  assign("root_path", root, envir = .GlobalEnv)
+  source_repo_modules(root)
+  invisible(root)
+}

### 2025-09-20 code/R/models/true_model.R

@@ -1,0 +1,152 @@
+# TRUE baseline model: independent MLE for each dimension
+# Notation follows repository documentation and README.md.
+
+#' Negative log-likelihood for a univariate distribution.
+#'
+#' @param par numeric vector of parameters
+#' @param x numeric vector of observations
+#' @param distr character string of distribution name
+#' @return scalar negative log-likelihood
+neg_loglik_uni <- function(par, x, distr) {
+  if (distr == "norm") {
+    mu <- par[1]; sd <- par[2]
+    if (sd <= 0) return(Inf)
+    -sum(dnorm(x, mean = mu, sd = sd, log = TRUE))
+  } else if (distr == "exp") {
+    rate <- par[1]
+    if (rate <= 0) return(Inf)
+    x <- pmax(x, 1e-6)
+    -sum(dexp(x, rate = rate, log = TRUE))
+  } else if (distr == "beta") {
+    a <- par[1]; b <- par[2]
+    if (a <= 0 || b <= 0) return(Inf)
+    x <- pmin(pmax(x, 1e-6), 1 - 1e-6)
+    -sum(dbeta(x, shape1 = a, shape2 = b, log = TRUE))
+  } else if (distr == "gamma") {
+    shape <- par[1]; scale <- par[2]
+    if (shape <= 0 || scale <= 0) return(Inf)
+    x <- pmax(x, 1e-6)
+    -sum(dgamma(x, shape = shape, scale = scale, log = TRUE))
+  } else {
+    stop("Unsupported distribution")
+  }
+}
+
+# helper to compute log-density vector
+.log_density_vec <- function(x, distr, par) {
+  if (distr == "norm") {
+    dnorm(x, mean = par[1], sd = par[2], log = TRUE)
+  } else if (distr == "exp") {
+    x <- pmax(x, 1e-6)
+    dexp(x, rate = par[1], log = TRUE)
+  } else if (distr == "beta") {
+    x <- pmin(pmax(x, 1e-6), 1 - 1e-6)
+    dbeta(x, shape1 = par[1], shape2 = par[2], log = TRUE)
+  } else if (distr == "gamma") {
+    x <- pmax(x, 1e-6)
+    dgamma(x, shape = par[1], scale = par[2], log = TRUE)
+  } else {
+    stop("Unsupported distribution")
+  }
+}
+
+# starting values based on method of moments
+.start_par <- function(x, distr) {
[…]

### 2025-09-20 code/R/models/ttm_marginal.R

@@ -1,0 +1,142 @@
+# Marginal Triangular Transport Map (TTM)
+# Contains train/test/predict for the marginal variant and shared dispatcher.
+if (!exists("ttm_forward")) {
+  root <- if (exists("root_path")) root_path else repo_root()
+  src <- file.path(root, "R", "models", "ttm_core.R")
+  if (file.exists(src)) source(src)
+}
+
+# Simple monotone f_k(x) = a_k + b_k x with b_k > 0
+.fit_marginal_per_k <- function(x_std, eps = 1e-6) {
+  # Optimize J(b,a) = 0.5 * sum (a + b x)^2 - n log b
+  n <- length(x_std)
+  xbar <- mean(x_std)
+  v <- stats::var(x_std) + 1e-12
+  b <- max(eps, 1 / sqrt(v))
+  a <- -b * xbar
+  c(a = a, b = b)
+}
+
+fit_ttm_marginal <- function(data, seed = 42) {
+  set.seed(seed)
+  S_in <- if (is.list(data) && !is.null(data$X_tr) && !is.null(data$X_te)) data else {
+    stopifnot(is.matrix(data))
+    if (!exists("split_data")) {
+      root <- if (exists("root_path")) root_path else repo_root()
+      source(file.path(root, "R", "split.R"))
+    }
+    list(X_tr = split_data(data, seed)$X_tr, X_te = split_data(data, seed)$X_te)
+  }
+  X_tr <- S_in$X_tr; X_te <- S_in$X_te
+  K <- ncol(X_tr)
+
+  t_tr <- system.time({
+    st <- standardize_train_only(X_tr)
+    mu <- st$mu; sigma <- st$sigma; X_tr_std <- st$Xs
+    coeffs <- lapply(seq_len(K), function(k) .fit_marginal_per_k(X_tr_std[, k]))
+    S <- list(algo = "marginal", mu = mu, sigma = sigma, coeffs = coeffs, order = seq_len(K))
+    class(S) <- "ttm_marginal2"
+  })[["elapsed"]]
+
+  t_te <- system.time({ invisible(predict_ttm(S, X_te, type = "logdensity_by_dim")) })[["elapsed"]]
+
+  list(S = S,
+       NLL_train = mean(-predict_ttm(S, X_tr, type = "logdensity")),
+       NLL_test  = mean(-predict_ttm(S, X_te,  type = "logdensity")),
+       time_train = t_tr, time_pred = t_te)
+}
+
+# Dispatcher for different TTM algos
[…]

### 2025-09-20 code/R/models/ttm_separable.R

@@ -1,0 +1,94 @@
+# Separable triangular map: S_k(x) = g_k(x_<k>) + f_k(x_k)
+# Contains train/test/predict for the separable variant; relies on core for bases and forward/J.
+
+# Ensure core (provides bases, forward, standardization)
+if (!exists("ttm_forward")) {
+  root <- if (exists("root_path")) root_path else repo_root()
+  src2 <- file.path(root, "R", "models", "ttm_core.R")
+  if (file.exists(src2)) source(src2)
+}
+
+fit_ttm_separable <- function(data, degree_g = 2L, lambda = 0.0, eps = 1e-6, seed = 42) {
+  set.seed(seed)
+  S_in <- if (is.list(data) && !is.null(data$X_tr) && !is.null(data$X_te)) data else {
+    stopifnot(is.matrix(data))
+    if (!exists("split_data")) {
+      root <- if (exists("root_path")) root_path else repo_root()
+      source(file.path(root, "R", "split.R"))
+    }
+    list(X_tr = split_data(data, seed)$X_tr, X_te = split_data(data, seed)$X_te)
+  }
+  X_tr <- S_in$X_tr; X_te <- S_in$X_te
+  K <- ncol(X_tr); N <- nrow(X_tr)
+
+  time_train <- system.time({
+    st <- standardize_train_only(X_tr); mu <- st$mu; sigma <- st$sigma
+    Xs <- st$Xs
+    coeffs <- vector("list", K)
+    I_N <- diag(N)
+    for (k in seq_len(K)) {
+      x_prev <- if (k > 1) Xs[, 1:(k - 1), drop = FALSE] else matrix(0, N, 0)
+      xk <- Xs[, k]
+      Phi_non <- if (ncol(x_prev) > 0) build_g(x_prev, deg = degree_g) else if (degree_g == 0L) matrix(1, N, 1L) else matrix(0, N, 0)
+      # For deg_g==0, use linear-only f and closed-form c for exact marginal equivalence
+      if (degree_g == 0L) {
+        # Exact marginal equivalence: use same closed-form as marginal per-dim
+        v <- stats::var(xk) + 1e-12
+        c_mon <- max(eps, 1 / sqrt(v))
+        c_non <- matrix(-mean(xk) * c_mon, ncol = 1L)
+        coeffs[[k]] <- list(c_non = c_non, c_mon = c_mon)
+        next
+      }
[…]

### 2025-09-20 code/experiments/4d/main.R

@@ -1,0 +1,103 @@
+config <- list(
+  list(distr = "norm", parm = NULL),
+  list(distr = "exp", parm = function(d) list(rate = softplus(d$X1))),
+  list(distr = "beta",
+       parm = function(d) list(shape1 = softplus(d$X2),
+                               shape2 = softplus(d$X1))),
+  list(distr = "gamma",
+       parm = function(d) list(shape = softplus(d$X3),
+                               scale = softplus(d$X2)))
+)
+
+N <- 50L
+override_N <- suppressWarnings(as.integer(Sys.getenv("N_OVERRIDE", "")))
+if (is.finite(override_N) && !is.na(override_N) && override_N > 0L) {
+  N <- override_N
+}
+n <- N
+
+ensure_loader <- function() {
+  if (exists("initialize_repo", inherits = TRUE)) {
+    existing_root <- tryCatch(get("root_path", envir = .GlobalEnv), error = function(e) NULL)
+    return(list(code_dir = existing_root))
+  }
+
+  locate_candidates <- function() {
+    paths <- character()
+
+    script_path <- tryCatch({
+      normalizePath(sys.frame(1)$ofile, winslash = "/", mustWork = TRUE)
+    }, error = function(e) NA_character_)
+
+    if (is.na(script_path) || !nzchar(script_path)) {
+      script_path <- tryCatch({
+        args <- commandArgs(trailingOnly = FALSE)
+        marker <- "--file="
+        file_arg <- args[grepl(marker, args, fixed = TRUE)]
+        if (length(file_arg)) {
+          normalizePath(sub(marker, "", file_arg[1]), winslash = "/", mustWork = TRUE)
+        } else {
+          NA_character_
+        }
+      }, error = function(e) NA_character_)
+    }
+
+    if (!is.na(script_path) && nzchar(script_path)) {
+      script_dir <- dirname(script_path)
+      paths <- c(paths,
+                 file.path(script_dir, "R", "loader.R"),
+                 file.path(dirname(script_dir), "R", "loader.R"),
+                 file.path(dirname(script_dir), "code", "R", "loader.R"))
+    }
+
+    wd <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
+    paths <- c(paths,
+               file.path(wd, "R", "loader.R"),
+               file.path(wd, "code", "R", "loader.R"),
[…]

### 2025-09-20 code/experiments/4d/main_perm.R

@@ -1,0 +1,141 @@
+config <- list(
+  list(distr = "norm", parm = NULL),
+  list(distr = "exp", parm = function(d) list(rate = softplus(d$X1))),
+  list(distr = "beta",
+       parm = function(d) list(shape1 = softplus(d$X2),
+                               shape2 = softplus(d$X1))),
+  list(distr = "gamma",
+       parm = function(d) list(shape = softplus(d$X3),
+                               scale = softplus(d$X2)))
+)
+
+config_canonical <- config
+
+N <- 50L
+override_N <- suppressWarnings(as.integer(Sys.getenv("N_OVERRIDE", "")))
+if (is.finite(override_N) && !is.na(override_N) && override_N > 0L) {
+  N <- override_N
+}
+n <- N
+
+ensure_loader <- function() {
+  if (exists("initialize_repo", inherits = TRUE)) {
+    existing_root <- tryCatch(get("root_path", envir = .GlobalEnv), error = function(e) NULL)
+    return(list(code_dir = existing_root))
+  }
+
+  locate_candidates <- function() {
+    paths <- character()
+
+    script_path <- tryCatch({
+      normalizePath(sys.frame(1)$ofile, winslash = "/", mustWork = TRUE)
+    }, error = function(e) NA_character_)
+
+    if (is.na(script_path) || !nzchar(script_path)) {
+      script_path <- tryCatch({
+        args <- commandArgs(trailingOnly = FALSE)
+        marker <- "--file="
+        file_arg <- args[grepl(marker, args, fixed = TRUE)]
+        if (length(file_arg)) {
+          normalizePath(sub(marker, "", file_arg[1]), winslash = "/", mustWork = TRUE)
+        } else {
+          NA_character_
+        }
+      }, error = function(e) NA_character_)
+    }
+
+    if (!is.na(script_path) && nzchar(script_path)) {
+      script_dir <- dirname(script_path)
+      paths <- c(paths,
+                 file.path(script_dir, "R", "loader.R"),
+                 file.path(dirname(script_dir), "R", "loader.R"),
+                 file.path(dirname(script_dir), "code", "R", "loader.R"))
+    }
+
+    wd <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
+    paths <- c(paths,
[…]

### 2025-09-20 code/experiments/4d/main_perm_iter.R

@@ -1,0 +1,114 @@
+permutations <- list(
+  c(1L, 2L, 3L, 4L),
+  c(1L, 2L, 4L, 3L),
+  c(1L, 3L, 2L, 4L),
+  c(1L, 3L, 4L, 2L),
+  c(1L, 4L, 2L, 3L),
+  c(1L, 4L, 3L, 2L),
+  c(2L, 1L, 3L, 4L),
+  c(2L, 1L, 4L, 3L),
+  c(2L, 3L, 1L, 4L),
+  c(2L, 3L, 4L, 1L),
+  c(2L, 4L, 1L, 3L),
+  c(2L, 4L, 3L, 1L),
+  c(3L, 1L, 2L, 4L),
+  c(3L, 1L, 4L, 2L),
+  c(3L, 2L, 1L, 4L),
+  c(3L, 2L, 4L, 1L),
+  c(3L, 4L, 1L, 2L),
+  c(3L, 4L, 2L, 1L),
+  c(4L, 1L, 2L, 3L),
+  c(4L, 1L, 3L, 2L),
+  c(4L, 2L, 1L, 3L),
+  c(4L, 2L, 3L, 1L),
+  c(4L, 3L, 1L, 2L),
+  c(4L, 3L, 2L, 1L)
+)
+
+detect_script_dir <- function() {
+  script_path <- tryCatch({
+    normalizePath(sys.frame(1)$ofile, winslash = "/", mustWork = TRUE)
+  }, error = function(e) NA_character_)
+  if (is.na(script_path) || !nzchar(script_path)) {
+    script_path <- tryCatch({
+      args <- commandArgs(trailingOnly = FALSE)
+      marker <- "--file="
+      file_arg <- args[grepl(marker, args, fixed = TRUE)]
+      if (length(file_arg)) {
+        normalizePath(sub(marker, "", file_arg[1]), winslash = "/", mustWork = TRUE)
+      } else {
+        NA_character_
+      }
+    }, error = function(e) NA_character_)
+  }
+  if (!is.na(script_path) && nzchar(script_path)) {
+    dirname(script_path)
+  } else {
+    normalizePath(getwd(), winslash = "/", mustWork = FALSE)
+  }
+}
+
+script_dir <- detect_script_dir()
+log_file_default <- file.path(script_dir, "results", "main_perm_iter.txt")
+log_file <- Sys.getenv("MAIN_PERM_ITER_LOG", log_file_default)
[…]

### 2025-09-20 code/experiments/halfmoon/halfmoon_pipeline.R

@@ -1,0 +1,140 @@
+if (!exists("initialize_repo")) {
+  source(file.path(getwd(), "R", "loader.R"))
+}
+root_path <- initialize_repo()
+
+args <- commandArgs(trailingOnly = TRUE)
+halfmoon_dir <- file.path(root_path, "experiments", "halfmoon")
+source(file.path(halfmoon_dir, "halfmoon_data.R"))
+source(file.path(halfmoon_dir, "plot.R"))
+source(file.path(halfmoon_dir, "true_density.R"))
+grid_side_arg <- NA_integer_
+no_cache <- TRUE
+cores_arg <- NA_integer_
+timeout_sec_arg <- NA_integer_
+abort_file <- NA_character_
+order_arg <- NA_character_
+for (a in args) {
+  if (grepl("^--grid_side=", a)) grid_side_arg <- as.integer(sub("^--grid_side=", "", a))
+  if (a == "--no_cache") no_cache <- TRUE
+  if (grepl("^--cores=", a)) cores_arg <- as.integer(sub("^--cores=", "", a))
+  if (grepl("^--timeout_sec=", a)) timeout_sec_arg <- as.integer(sub("^--timeout_sec=", "", a))
+  if (grepl("^--abort_file=", a)) abort_file <- sub("^--abort_file=", "", a)
+  if (grepl("^--order=", a)) order_arg <- sub("^--order=", "", a)
+}
+SEED <- 7L
+N <- 100L
+# Quick edit: set default noise here
+NOISE <- 0.15
+# Optional: allow override via environment variable NOISE_OVERRIDE
+nv <- suppressWarnings(as.numeric(Sys.getenv("NOISE_OVERRIDE", "")))
+if (is.finite(nv) && !is.na(nv) && nv > 0) NOISE <- nv
+Sys.setenv(
+  DATASET = "halfmoon2d",
+  SEED = as.character(SEED),
+  N_TRAIN = as.character(N),
+  N_TEST = as.character(N),
+  NOISE = as.character(NOISE)
+)
+# Create splits and persist for reproducibility
+S <- make_halfmoon_splits(n_train = N, n_test = N, noise = NOISE, seed = SEED, val_frac = 0.2)
+# Do not persist splits to disk to avoid artifacts
+# Fit models and write NLL CSV via evaluation helper (no plotting)
[…]

### 2025-09-20 code/experiments/halfmoon/main_moon.R

@@ -1,0 +1,9 @@
+if (!exists("initialize_repo")) {
+  source(file.path(getwd(), "R", "loader.R"))
+}
+root_path <- initialize_repo()
+source(file.path(root_path, "experiments", "halfmoon", "halfmoon_pipeline.R"))
+
+if (sys.nframe() == 0L) {
+  invisible(NULL)
+}

### 2025-09-20 code/experiments/halfmoon/plot.R

@@ -1,0 +1,364 @@
+#' Fit half-moon models and create contour plots
+#'
+#' Provides two helper functions for working with the two-moons data
+#' splits.
+
+fit_halfmoon_models <- function(S, seed = NULL, order_mode = "as-is") {
+  seed <- if (!is.null(seed)) as.integer(seed) else if (!is.null(S$meta$seed)) S$meta$seed else 42L
+  set.seed(seed)
+  cfg <- list(list(distr = "norm"), list(distr = "norm"))
+  M_true <- fit_TRUE(S, cfg)
+  class(M_true) <- c("true_marginal_model", class(M_true))
+  # Ensure copula_np fitter is available without side effects
+  if (!exists("fit_copula_np", mode = "function")) {
+    root <- if (exists("root_path")) root_path else repo_root()
+    src <- file.path(root, "R", "models", "copula_np.R")
+    if (file.exists(src)) source(src)
+  }
+  if (!exists("fit_copula_np", mode = "function")) {
+    stop("Copula NP fitter `fit_copula_np` must be available to plot Copula NP panel.")
+  }
+
+  timing <- list()
+  timing$true <- list(train = 0.0, test = NA_real_)
+
+  trtf_time <- system.time({
+    M_trtf <- tryCatch(fit_TRTF(S, cfg, seed = seed), error = function(e) NULL)
+  })[["elapsed"]]
+  timing$trtf <- list(train = if (!is.null(M_trtf)) trtf_time else NA_real_, test = NA_real_)
+
+  fit_ttm_m <- fit_ttm(S, algo = "marginal", seed = seed)
+  timing$ttm <- list(train = fit_ttm_m$time_train, test = fit_ttm_m$time_pred)
+
+  fit_ttm_s <- fit_ttm(S, algo = "separable", seed = seed)
+  timing$ttm_sep <- list(train = fit_ttm_s$time_train, test = fit_ttm_s$time_pred)
+
+  cop_time <- system.time({
+    M_cop <- fit_copula_np(S, seed = seed)
+  })[["elapsed"]]
+  timing$copula_np <- list(train = cop_time, test = NA_real_)
+
+  mods <- list(
+    true = M_true,
+    trtf = M_trtf,
+    ttm = fit_ttm_m$S,
+    ttm_sep = fit_ttm_s$S,
+    copula_np = M_cop
+  )
+  attr(mods, "timing") <- timing
+  mods
+}
+
[…]

### 2025-09-20 code/experiments/legacy/main_moon.R

@@ -1,0 +1,52 @@
+#!/usr/bin/env Rscript
+
+# Half-moon runner without artifacts/logs. Uses existing utilities but disables saving.
+Sys.setenv(NO_ARTIFACTS = "1")
+options(mde.no_artifacts = TRUE)
+
+args <- commandArgs(trailingOnly = TRUE)
+grid_side_arg <- NA_integer_
+cores_arg <- NA_integer_
+timeout_sec_arg <- NA_integer_
+order_arg <- NA_character_
+for (a in args) {
+  if (grepl("^--grid_side=", a)) grid_side_arg <- as.integer(sub("^--grid_side=", "", a))
+  if (grepl("^--cores=", a)) cores_arg <- as.integer(sub("^--cores=", "", a))
+  if (grepl("^--timeout_sec=", a)) timeout_sec_arg <- as.integer(sub("^--timeout_sec=", "", a))
+  if (grepl("^--order=", a)) order_arg <- sub("^--order=", "", a)
+}
+
+if (!exists("initialize_repo")) {
+  source(file.path(getwd(), "R", "loader.R"))
+}
+root_path <- initialize_repo()
+halfmoon_dir <- file.path(root_path, "experiments", "halfmoon")
+source(file.path(halfmoon_dir, "halfmoon_data.R"))
+source(file.path(halfmoon_dir, "plot.R"))
+source(file.path(halfmoon_dir, "true_density.R"))
+
+# Config
+SEED <- as.integer(Sys.getenv("SEED", 7))
+N <- as.integer(Sys.getenv("N_MOON", 100))
+NOISE <- as.numeric(Sys.getenv("NOISE", 0.15))
+
+S <- make_halfmoon_splits(n_train = N, n_test = N, noise = NOISE, seed = SEED, val_frac = 0.2)
+mods <- fit_halfmoon_models(S, seed = SEED, order_mode = if (!is.na(order_arg)) order_arg else "as-is")
+
+# Evaluate without writing CSV
+tab <- eval_halfmoon(mods, S, out_csv_path = NULL)
+if ("train_test_policy" %in% names(tab)) tab$train_test_policy <- NULL
+cat(sprintf("n=%d (Half-moon)\n", N))
+print(tab)
+
+# Optional contour visualization without saving PNGs
+default_cores <- tryCatch(min(8L, parallel::detectCores()), error = function(e) 1L)
+cores <- if (!is.na(cores_arg) && cores_arg >= 1L) cores_arg else default_cores
+N_tr <- nrow(S$X_tr)
[…]

### 2025-09-20 code/experiments/legacy/trtf_power.R

@@ -1,0 +1,9 @@
+#!/usr/bin/env Rscript
+
+# Power dataset runner with no artifacts/logs. Wraps the embedded TRTF power script.
+Sys.setenv(NO_ARTIFACTS = "1")
+options(mde.no_artifacts = TRUE)
+
+source(file.path("experiments", "tools", "trtf_power.R"))
+
+invisible(trtf_power())

### 2025-09-20 code/experiments/normalizing flow/miniboone/miniboone_dataprep.R

@@ -1,0 +1,156 @@
+###############################################################################
+#   1. Herunterladen & Einlesen                                              #
+#   2. Nur positive Ereignisse (electron-neutrinos) behalten                 #
+#   3. 11 Platzhalter-Zeilen (-1000/-999 in allen 50 Spalten) entfernen       #
+#   4. 7 Spalten mit extrem häufig gleichem Wert (“Dirac-Peaks”) streichen    #
+#   5. Standardisierung: (x-μ_train)/σ_train                                  #
+#   6. Reproduzierbarer Split 80 %/10 %/10 % -> Train / Val / Test            #
+#   7. Speichern als *.RData                                                 #
+#                                                                            #
+# Am Ende liegen drei Objekte vor: x_train, x_val, x_test                    #
+# Dimensionen laut Paper:                                                    #
+#   • Train: 29 556 × 43                                                     #
+#   • Val:    3 284 × 43                                                     #
+#   • Test:   3 648 × 43                                                     #
+###############################################################################
+# 
+# # ------------------------- 1 | Download & Roh-Import -------------------------
+# url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00199/MiniBooNE_PID.txt"
+# tf  <- tempfile()
+# download.file(url, tf, mode = "wb")          # ~87 MB
+# 
+# con   <- file(tf, "r")
+# nums  <- scan(con, what = integer(), n = 2, quiet = TRUE)
+# n_sig <- nums[1]                             # Zeile 1: #positiver Events
+# # Zeile 2: #negativer Events (wird ignoriert)
+# 
+# raw   <- scan(con, what = numeric(), quiet = TRUE)  # restliche Zahlen
+# close(con)
+# 
+# x <- matrix(raw, ncol = 50, byrow = TRUE)[1:n_sig, ]  # 50 Merkmale, nur positives
+# 
+# # --------------------- 2 | 11 vollständige Ausreißer löschen -----------------
[…]

### 2025-09-20 code/experiments/normalizing flow/miniboone/miniboone_main.R

@@ -1,0 +1,73 @@
+#!/usr/bin/env Rscript
+
+# MiniBooNE runner without artifacts/logs. Expects pre-split CSVs under
+# `experiments/normalizing flow/miniboone/` by default.
+Sys.setenv(NO_ARTIFACTS = "1")
+options(mde.no_artifacts = TRUE)
+
+if (!exists("initialize_repo")) {
+  source(file.path(getwd(), "R", "loader.R"))
+}
+root_path <- initialize_repo()
+
+stderr <- function(x) stats::sd(x) / sqrt(length(x))
+fmt <- function(m, se) sprintf("%.2f ± %.2f", round(m, 2), round(2 * se, 2))
+
+read_num_csv <- function(path) {
+  if (!file.exists(path)) stop(sprintf("CSV not found: %s", path))
+  df <- utils::read.csv(path, header = TRUE, check.names = TRUE)
+  df <- df[vapply(df, is.numeric, logical(1))]
+  as.matrix(df)
+}
+
+args <- commandArgs(trailingOnly = TRUE)
+get_arg <- function(key, default = NULL) {
+  m <- grep(paste0("^--", key, "="), args, value = TRUE)
+  if (length(m)) sub(paste0("^--", key, "="), "", m[1]) else default
+}
+
+default_dir <- file.path(root_path, "experiments", "normalizing flow", "miniboone")
+train_csv <- get_arg("train_csv", Sys.getenv("MINIBOONE_TRAIN", file.path(default_dir, "miniboone_train.csv")))
+val_csv   <- get_arg("val_csv",   Sys.getenv("MINIBOONE_VAL",   file.path(default_dir, "miniboone_val.csv")))
+test_csv  <- get_arg("test_csv",  Sys.getenv("MINIBOONE_TEST",  file.path(default_dir, "miniboone_test.csv")))
+seed <- as.integer(get_arg("seed", Sys.getenv("SEED", "42")))
+
+set.seed(seed)
+Xtr_all <- read_num_csv(train_csv)
+Xva_all <- read_num_csv(val_csv)
+Xte_all <- read_num_csv(test_csv)
+common <- Reduce(intersect, list(colnames(Xtr_all), colnames(Xva_all), colnames(Xte_all)))
[…]

### 2025-09-20 code/experiments/normalizing flow/power/main_power.R

@@ -1,0 +1,211 @@
+if (!exists("initialize_repo")) {
+  source(file.path(getwd(), "R", "loader.R"))
+}
+root_path <- initialize_repo()
+
+RUN_CONFIG <- list(
+  enabled = FALSE,
+  train_csv = "/Users/leonkiafaro/Documents/Masterthesis/data_power/power_train_head_1pct.csv",
+  test_csv  = "/Users/leonkiafaro/Documents/Masterthesis/data_power/power_test_head_1pct.csv",
+  n_train = 250L,
+  n_test  = 250L,
+  models = c("trtf","ttm","ttm_sep","copula_np")
+)
+
+# Small helpers
+stderr <- function(x) stats::sd(x) / sqrt(length(x))
+
+`%||%` <- function(a, b) if (is.null(a)) b else a
+
+read_numeric_matrix <- function(path, n_rows = NA_integer_) {
+  if (!file.exists(path)) stop(sprintf("CSV not found: %s", path))
+  df <- utils::read.csv(path, header = TRUE, check.names = TRUE)
+  # Keep only numeric columns present
+  num_cols <- names(df)[vapply(df, is.numeric, logical(1))]
+  if (!length(num_cols)) stop(sprintf("No numeric columns in %s", basename(path)))
+  df <- df[, num_cols, drop = FALSE]
+  if (is.finite(n_rows) && !is.na(n_rows) && n_rows > 0L) {
+    n_keep <- min(n_rows, nrow(df))
+    df <- utils::head(df, n_keep)
+  }
+  X <- as.matrix(df)
+  # Drop rows with NA/Inf
+  ok <- stats::complete.cases(X) & apply(is.finite(X), 1, all)
+  if (!all(ok)) X <- X[ok, , drop = FALSE]
+  storage.mode(X) <- "double"
+  X
+}
+
+parse_args <- function() {
+  out <- list()
+  for (a in commandArgs(trailingOnly = TRUE)) {
+    if (grepl("^--n_train=", a)) out$n_train <- as.integer(sub("^--n_train=", "", a))
+    if (grepl("^--n_test=", a))  out$n_test  <- as.integer(sub("^--n_test=",  "", a))
+    if (grepl("^--train_csv=", a)) out$train_csv <- sub("^--train_csv=", "", a)
[…]

### 2025-09-20 code/experiments/normalizing flow/power/power_make_splits_dataprep.R

@@ -1,0 +1,127 @@
+# ==== USER CONFIG (edit here) ==========================================
+INPUT_POWER_TXT <- "/absolute/path/to/household_power_consumption.txt"
+N <- 100000L          # max rows per split (train/test). Use NA for all rows.
+SEED <- 42L
+INCLUDE_TIME <- FALSE # FALSE => D=6 wie in MAF Table 5
+ADD_NOISE <- TRUE     # MAF-konformes Dequantisieren
+OUT_DIR <- file.path("experiments", "normalizing flow", "power")
+WRITE_VAL <- FALSE    # bei TRUE zusätzlich power_val.csv
+
+suppressPackageStartupMessages({ library(utils); library(stats) })
+
+power_make_splits_write_csv <- function(input_power_txt = INPUT_POWER_TXT,
+                                        out_dir = OUT_DIR,
+                                        seed = SEED,
+                                        include_time = INCLUDE_TIME,
+                                        add_noise = ADD_NOISE,
+                                        n_rows_per_split = N,
+                                        write_val = WRITE_VAL) {
+  stopifnot(file.exists(input_power_txt))
+
+  # 1) Einlesen
+  raw <- utils::read.csv(input_power_txt, sep = ";",
+                         na.strings = c("?", "NA", ""), stringsAsFactors = FALSE)
+  req <- c("Date","Time","Global_active_power","Global_reactive_power","Voltage",
+           "Global_intensity","Sub_metering_1","Sub_metering_2","Sub_metering_3")
+  miss <- setdiff(req, names(raw))
+  if (length(miss)) stop("Fehlende Spalten: ", paste(miss, collapse = ", "))
+
+  num_cols <- c("Global_active_power","Voltage","Global_intensity",
+                "Sub_metering_1","Sub_metering_2","Sub_metering_3","Global_reactive_power")
+  for (nm in num_cols) raw[[nm]] <- as.numeric(raw[[nm]])
+
+  # 2) Zeit -> Minute-im-Tag (Integer)
+  if (include_time) {
+    tt <- raw[["Time"]]
[…]

### 2025-09-20 code/experiments/normalizing flow/power/subset_head_three_datasets.R

@@ -1,0 +1,142 @@
+# subset_head_three_datasets.R
+# Zweck: Von drei Dateien (POWER, GAS, HEPMASS) die ersten p% Datenzeilen
+#        (exkl. Header) extrahieren und als CSV speichern.
+
+suppressWarnings(suppressMessages({
+  have_dt <- requireNamespace("data.table", quietly = TRUE)
+}))
+
+# -- Hilfsfunktion: schnelle Zeilenzahl (Unix: wc -l; Fallback: binär lesen) --
+count_lines_fast <- function(path) {
+  stopifnot(file.exists(path))
+  out <- tryCatch(
+    system2("wc", c("-l", shQuote(path)), stdout = TRUE),
+    error = function(e) NA_character_
+  )
+  if (!is.na(out[1])) {
+    n <- suppressWarnings(as.integer(sub("^\\s*(\\d+).*$", "\\1", out[1])))
+    if (is.finite(n)) return(n)
+  }
+  con <- file(path, "rb"); on.exit(close(con), add = TRUE)
+  n <- 0L
+  repeat {
+    buf <- readBin(con, what = "raw", n = 8L * 1024L * 1024L)
+    if (!length(buf)) break
+    n <- n + sum(buf == as.raw(10L)) # '\n'
+  }
+  n
+}
+
+# -- Hilfsfunktion: Prozentzahl in Bruch umwandeln (5 -> 0.05; 0.5 -> 0.5) --
+as_fraction <- function(pct) {
+  stopifnot(is.numeric(pct), length(pct) == 1L, is.finite(pct), pct > 0)
+  if (pct > 1) pct/100 else pct
+}
+
+# -- Kernfunktion: p% Kopf speichern --
+subset_head_percent_save <- function(in_file, pct, out_file,
+                                     sep_in = NULL, header = TRUE,
+                                     na = c("", "?", "NA"),
+                                     out_sep = ",",
+                                     use_fread = have_dt) {
+  frac <- as_fraction(pct)
+  total <- count_lines_fast(in_file)
+  if (!is.finite(total) || total <= 0) stop("Konnte Zeilenzahl nicht bestimmen: ", in_file)
+  header_rows <- if (isTRUE(header)) 1L else 0L
+  data_rows_total <- max(0L, total - header_rows)
+  n <- max(1L, floor(data_rows_total * frac))
+  
+  if (use_fread) {
+    args <- list(input = in_file, nrows = n, header = header,
[…]

### 2025-09-20 code/experiments/normalizing flow/power/trtf_power.R

@@ -1,0 +1,764 @@
+### Minimaler, schneller Konfigurationsblock (oben anpassbar) #################
+# A) Anzahl Zeilen pro Split (Train/Test)
+N_ROWS <- 10000L
+# B) CSV-Pfade relativ zum Skript-Verzeichnis
+SCRIPT_DIR <- (function(){
+  args <- commandArgs(trailingOnly = FALSE)
+  file_arg <- grep("^--file=", args, value = TRUE)
+  if (length(file_arg) > 0L) return(dirname(normalizePath(sub("^--file=", "", file_arg[[1L]]))))
+  # when sourced
+  f <- tryCatch(normalizePath(sys.frames()[[1]]$ofile), error = function(e) NA_character_)
+  if (is.character(f) && length(f) == 1L && !is.na(f)) return(dirname(f))
+  getwd()
+})()
+
+locate_data_dir <- function(initial_dir, required) {
+  candidates <- unique(c(
+    initial_dir,
+    normalizePath(getwd(), winslash = "/", mustWork = FALSE),
+    normalizePath(file.path(getwd(), "code", "experiments", "normalizing flow", "power"), winslash = "/", mustWork = FALSE)
+  ))
+  for (dir in candidates) {
+    if (nzchar(dir) && all(file.exists(file.path(dir, required)))) return(dir)
+  }
+  initial_dir
+}
+
+SCRIPT_DIR <- locate_data_dir(SCRIPT_DIR, c("power_train_10000.csv", "power_test_10000.csv"))
+TRAIN_CSV <- file.path(SCRIPT_DIR, "power_train_10000.csv")
+TEST_CSV  <- file.path(SCRIPT_DIR, "power_test_10000.csv")
+# C) Pfad für Evaluationstext (Konsole + Datei)
+RESULTS_TXT <- file.path(SCRIPT_DIR, "results", "trtf_power_eval.txt")
+# D) Hyperparameter
+HP <- list(
+  ntree     = 200L,
+  minsplit  = 100L,
+  minbucket = 20L,
+  maxdepth  = 4L,
+  seed      = 42L
+)
+###############################################################################
+
+ensure_required_packages <- function(pkgs) {
+  missing <- character(0)
+  for (pkg in pkgs) {
+    ok <- suppressPackageStartupMessages(require(pkg, quietly = TRUE, character.only = TRUE))
+    if (!isTRUE(ok)) missing <- c(missing, pkg)
+  }
+  if (length(missing)) {
[…]

### 2025-09-20 data_prep/prep_power.R

@@ -1,130 +1,0 @@
-power_read_and_preprocess <- function(
-    file_path,
-    seed = 42L,
-    include_time = FALSE,            # D=6: FALSE (MAF-Tabelle); D=7: TRUE (Zeit als Feature)
-    add_noise = TRUE,                # Dequantisierung pro Feature
-    test_frac = 0.10,                # 10% Test
-    val_frac_of_remaining = 0.10,    # 10% von (Rest) als Validation
-    write_csv = FALSE,               # optional: standardisierte Splits als CSV schreiben
-    out_dir = "data"
-) {
-  stopifnot(file.exists(file_path))
-  
-  ## 1) Einlesen (UCI: Semikolon, Dezimalpunkt ".", fehlende Werte = "?" oder leer)
-  raw <- utils::read.csv(
-    file_path,
-    sep = ";",
-    na.strings = c("?", "NA", ""),
-    stringsAsFactors = FALSE
-  )
-  
-  req <- c(
-    "Date","Time",
-    "Global_active_power","Global_reactive_power","Voltage","Global_intensity",
-    "Sub_metering_1","Sub_metering_2","Sub_metering_3"
-  )
-  miss <- setdiff(req, names(raw))
-  if (length(miss)) stop("Fehlende Spalten: ", paste(miss, collapse=", "))
-  
-  ## 2) Numerische Spalten sicher numerisch parsen
-  num_cols <- c("Global_active_power","Voltage","Global_intensity",
-                "Sub_metering_1","Sub_metering_2","Sub_metering_3",
-                "Global_reactive_power")
-  for (nm in num_cols) raw[[nm]] <- as.numeric(raw[[nm]])
-  
-  ## 3) Optional: Zeit -> Minuten-im-Tag (Integer 0..1439). Rauschen kommt später zentral.
-  if (include_time) {
-    tt <- raw[["Time"]]
-    hh <- suppressWarnings(as.integer(substr(tt, 1L, 2L)))
-    mm <- suppressWarnings(as.integer(substr(tt, 4L, 5L)))
-    minute_of_day <- hh * 60L + mm
-  }
-  
-  ## 4) Verwerfen: Date, Global_reactive_power (Paper D.2)
-  X <- raw[, c("Global_active_power","Voltage","Global_intensity",
[…]

### 2025-09-20 main.R

@@ -1,162 +1,0 @@
-get_repo_root <- function() {
-  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
-  for (i in 1:10) {
-    if (file.exists(file.path(p, "ALGORITHM_SPEC.md")) && file.exists(file.path(p, "00_globals.R"))) return(p)
-    parent <- dirname(p)
-    if (identical(parent, p)) break
-    p <- parent
-  }
-  normalizePath(getwd(), winslash = "/", mustWork = FALSE)
-}
-.repo_root <- if (exists("root_path")) root_path else get_repo_root()
-source(file.path(.repo_root, "00_globals.R"))
-source(file.path(.repo_root, "01_data_generation.R"))
-source(file.path(.repo_root, "02_split.R"))
-source(file.path(.repo_root, "models/true_model.R"))
-source(file.path(.repo_root, "models/trtf_model.R"))
-source(file.path(.repo_root, "models/ttm/ttm_bases.R"))
-source(file.path(.repo_root, "models/ttm/ttm_core.R"))
-source(file.path(.repo_root, "models/ttm/ttm_marginal.R"))
-source(file.path(.repo_root, "models/ttm/ttm_separable.R"))
-source(file.path(.repo_root, "models/ttm/ttm_crossterm.R"))
-source(file.path(.repo_root, "models/true_joint_model.R"))
-source(file.path(.repo_root, "04_evaluation.R"))
-## Optional: order heuristics
-if (file.exists(file.path(.repo_root, "models", "ttm", "order_heuristics.R"))) {
-  source(file.path(.repo_root, "models", "ttm", "order_heuristics.R"))
-}
-## Optional nonparametric copula baseline
-if (file.exists(file.path(.repo_root, "models", "copula_np.R"))) {
-  source(file.path(.repo_root, "models", "copula_np.R"))
-}
-
-
-perm <- c(1, 2, 3, 4)
-n <- 50
-# Optional: allow overriding n via environment variable N_OVERRIDE
-nv <- suppressWarnings(as.integer(Sys.getenv("N_OVERRIDE", "")))
-if (is.finite(nv) && !is.na(nv) && nv > 0L) n <- nv
-config <- list(
-  list(distr = "norm", parm = NULL),
-  list(distr = "exp",  parm = function(d) list(rate = softplus(d$X1))),
-  list(distr = "beta",
[…]

### 2025-09-20 main_moon.R

@@ -1,70 +1,0 @@
-args <- commandArgs(trailingOnly = TRUE)
-source("00_globals.R")
-grid_side_arg <- NA_integer_
-no_cache <- FALSE
-cores_arg <- NA_integer_
-timeout_sec_arg <- NA_integer_
-abort_file <- NA_character_
-order_arg <- NA_character_
-for (a in args) {
-  if (grepl("^--grid_side=", a)) grid_side_arg <- as.integer(sub("^--grid_side=", "", a))
-  if (a == "--no_cache") no_cache <- TRUE
-  if (grepl("^--cores=", a)) cores_arg <- as.integer(sub("^--cores=", "", a))
-  if (grepl("^--timeout_sec=", a)) timeout_sec_arg <- as.integer(sub("^--timeout_sec=", "", a))
-  if (grepl("^--abort_file=", a)) abort_file <- sub("^--abort_file=", "", a)
-  if (grepl("^--order=", a)) order_arg <- sub("^--order=", "", a)
-}
-SEED <- 7L
-N <- 100L
-# Quick edit: set default noise here
-NOISE <- 0.15
-# Optional: allow override via environment variable NOISE_OVERRIDE
-nv <- suppressWarnings(as.numeric(Sys.getenv("NOISE_OVERRIDE", "")))
-if (is.finite(nv) && !is.na(nv) && nv > 0) NOISE <- nv
-Sys.setenv(
-  DATASET = "halfmoon2d",
-  SEED = as.character(SEED),
-  N_TRAIN = as.character(N),
-  N_TEST = as.character(N),
-  NOISE = as.character(NOISE)
-)
-source("scripts/halfmoon_data.R")
-source("scripts/halfmoon_plot.R")
-source("04_evaluation.R")
-source("models/true_model.R")
-source("models/trtf_model.R")
-# Create splits and persist for reproducibility
-S <- make_halfmoon_splits(n_train = N, n_test = N, noise = NOISE, seed = SEED, val_frac = 0.2)
-dir.create("results", showWarnings = FALSE)
-saveRDS(S, sprintf("results/splits_halfmoon2d_seed%03d.rds", SEED))
-# Fit models and write NLL CSV via evaluation helper (no plotting)
-mods <- fit_halfmoon_models(S, seed = SEED, order_mode = if (!is.na(order_arg)) order_arg else "as-is")
-tab <- eval_halfmoon(mods, S)
[…]

### 2025-09-20 miniboone.R

@@ -1,155 +1,0 @@
-###############################################################################
-#   1. Herunterladen & Einlesen                                              #
-#   2. Nur positive Ereignisse (electron-neutrinos) behalten                 #
-#   3. 11 Platzhalter-Zeilen (-1000/-999 in allen 50 Spalten) entfernen       #
-#   4. 7 Spalten mit extrem häufig gleichem Wert (“Dirac-Peaks”) streichen    #
-#   5. Standardisierung: (x-μ_train)/σ_train                                  #
-#   6. Reproduzierbarer Split 80 %/10 %/10 % -> Train / Val / Test            #
-#   7. Speichern als *.RData                                                 #
-#                                                                            #
-# Am Ende liegen drei Objekte vor: x_train, x_val, x_test                    #
-# Dimensionen laut Paper:                                                    #
-#   • Train: 29 556 × 43                                                     #
-#   • Val:    3 284 × 43                                                     #
-#   • Test:   3 648 × 43                                                     #
-###############################################################################
-# 
-# # ------------------------- 1 | Download & Roh-Import -------------------------
-# url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00199/MiniBooNE_PID.txt"
-# tf  <- tempfile()
-# download.file(url, tf, mode = "wb")          # ~87 MB
-# 
-# con   <- file(tf, "r")
-# nums  <- scan(con, what = integer(), n = 2, quiet = TRUE)
-# n_sig <- nums[1]                             # Zeile 1: #positiver Events
-# # Zeile 2: #negativer Events (wird ignoriert)
-# 
-# raw   <- scan(con, what = numeric(), quiet = TRUE)  # restliche Zahlen
-# close(con)
-# 
-# x <- matrix(raw, ncol = 50, byrow = TRUE)[1:n_sig, ]  # 50 Merkmale, nur positives
-# 
-# # --------------------- 2 | 11 vollständige Ausreißer löschen -----------------
[…]

### 2025-09-20 models/copula_np.R

@@ -1,167 +1,0 @@
-# Nonparametric Copula Baseline (robust runner)
-#
-# Primary mode (2D, labeled): per-class univariate KDE marginals (dims 1 and 2)
-# and a bivariate kernel copula on the empirical PIT (kde1d + kdecopula).
-# Fallback mode (generic K-D, unlabeled or missing packages): independent
-# kernel density product using base::density per dimension (no extra deps).
-
-#' Fit per-class nonparametric copula (2D)
-#'
-#' @param S List with `X_tr` (numeric matrix with 2 columns) and `y_tr` (labels)
-#' @param seed Integer RNG seed for determinism
-#' @return An object of class 'copula_np' with fields `classes`, `by_class`, `seed`
-#' @details
-#' Per class y, the method:
-#' - fits kde1d on each marginal X1,X2 of the class-specific subset (train only)
-#' - computes pseudo-observations U via mid-ranks with ties.average, scaled by (n+1)
-#' - clamps U to [1/(n+1), n/(n+1)]
-#' - fits a kernel copula (kdecopula::kdecop) on U with method='TLL2'
-#' The returned object stores for each class y: list(m1, m2, cop, n_tr, eps).
-fit_copula_np <- function(S, seed = 42) {
-  stopifnot(is.list(S))
-  X_tr <- as.matrix(S$X_tr)
-  y_tr <- S$y_tr
-  K <- ncol(X_tr)
-  has_pkgs <- (requireNamespace("kde1d", quietly = TRUE) && requireNamespace("kdecopula", quietly = TRUE))
-  use_cop <- (K == 2L && !is.null(y_tr) && length(y_tr) == nrow(X_tr) && has_pkgs)
-
-  set.seed(as.integer(seed))
-  if (!use_cop) {
-    # Fallback: product of univariate KDEs (base::density)
-    dens <- vector("list", K)
-    for (k in seq_len(K)) {
-      dk <- stats::density(X_tr[, k], bw = "nrd0", n = 2048,
-                           from = min(X_tr[, k]), to = max(X_tr[, k]))
-      dens[[k]] <- list(x = dk$x, y = pmax(dk$y, .Machine$double.xmin))
-    }
-    return(structure(list(mode = "kde_product", dens = dens, seed = as.integer(seed)),
[…]

### 2025-09-20 models/true_model.R

@@ -1,152 +1,0 @@
-# TRUE baseline model: independent MLE for each dimension
-# Notation follows Theory.md and README.md.
-
-#' Negative log-likelihood for a univariate distribution.
-#'
-#' @param par numeric vector of parameters
-#' @param x numeric vector of observations
-#' @param distr character string of distribution name
-#' @return scalar negative log-likelihood
-neg_loglik_uni <- function(par, x, distr) {
-  if (distr == "norm") {
-    mu <- par[1]; sd <- par[2]
-    if (sd <= 0) return(Inf)
-    -sum(dnorm(x, mean = mu, sd = sd, log = TRUE))
-  } else if (distr == "exp") {
-    rate <- par[1]
-    if (rate <= 0) return(Inf)
-    x <- pmax(x, 1e-6)
-    -sum(dexp(x, rate = rate, log = TRUE))
-  } else if (distr == "beta") {
-    a <- par[1]; b <- par[2]
-    if (a <= 0 || b <= 0) return(Inf)
-    x <- pmin(pmax(x, 1e-6), 1 - 1e-6)
-    -sum(dbeta(x, shape1 = a, shape2 = b, log = TRUE))
-  } else if (distr == "gamma") {
-    shape <- par[1]; scale <- par[2]
-    if (shape <= 0 || scale <= 0) return(Inf)
-    x <- pmax(x, 1e-6)
-    -sum(dgamma(x, shape = shape, scale = scale, log = TRUE))
-  } else {
-    stop("Unsupported distribution")
-  }
-}
-
-# helper to compute log-density vector
-.log_density_vec <- function(x, distr, par) {
-  if (distr == "norm") {
-    dnorm(x, mean = par[1], sd = par[2], log = TRUE)
-  } else if (distr == "exp") {
-    x <- pmax(x, 1e-6)
-    dexp(x, rate = par[1], log = TRUE)
-  } else if (distr == "beta") {
-    x <- pmin(pmax(x, 1e-6), 1 - 1e-6)
-    dbeta(x, shape1 = par[1], shape2 = par[2], log = TRUE)
-  } else if (distr == "gamma") {
-    x <- pmax(x, 1e-6)
-    dgamma(x, shape = par[1], scale = par[2], log = TRUE)
-  } else {
-    stop("Unsupported distribution")
-  }
-}
-
-# starting values based on method of moments
-.start_par <- function(x, distr) {
[…]

### 2025-09-20 models/ttm/fit_ttm_crossterm.R

@@ -1,431 +1,0 @@
-## Cross-term Triangular Transport Map (TTM)
-## Concrete fitter and evaluator; no proxies, R-only, deterministic.
-
-if (!exists("%||%")) `%||%` <- function(a, b) if (is.null(a)) b else a
-
-# Robust loader for repo-local components (works from tests/scripts)
-if (!exists(".source_repo_file")) {
-  .source_repo_file <- function(rel_path) {
-    cand <- character(0)
-    cand <- c(cand, rel_path)
-    this_file <- tryCatch(normalizePath(sys.frames()[[1]]$ofile, winslash = "/", mustWork = FALSE), error = function(e) NA_character_)
-    if (is.character(this_file) && nzchar(this_file) && !is.na(this_file)) cand <- c(cand, file.path(dirname(this_file), basename(rel_path)))
-    if (exists("root_path")) cand <- c(cand, file.path(root_path, rel_path))
-    p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
-    for (i in 1:8) {
-      if (file.exists(file.path(p, "00_globals.R"))) { cand <- c(cand, file.path(p, rel_path)); break }
-      np <- dirname(p); if (identical(np, p)) break; p <- np
-    }
-    cand <- unique(cand)
-    for (pth in cand) { if (file.exists(pth)) { source(pth); return(invisible(TRUE)) } }
-    invisible(FALSE)
-  }
-}
-
-# Source required TTM components
-.source_repo_file(file.path("models", "ttm", "ttm_bases.R"))
-.source_repo_file(file.path("models", "ttm", "crossterm_basis.R"))
-.source_repo_file(file.path("models", "ttm", "ttm_core.R"))
-.source_repo_file(file.path("models", "ttm", "ttm_optimize.R"))
-
-.std_stats <- function(X) {
-  mu <- colMeans(X)
-  sigma <- apply(X, 2, sd) + .Machine$double.eps
-  list(mu = mu, sigma = sigma)
-}
-
-.standardize <- function(X, mu, sigma) {
-  sweep(sweep(X, 2, mu, "-"), 2, sigma, "/")
-}
-
-#' Fit cross-term TTM via maps-from-samples with GL quadrature
-#'
-#' @param data matrix X or list with X_tr, X_te
-#' @param seed integer RNG seed
-#' @param deg_g degree for predecessor polynomial features g(x_<k>)
[…]

### 2025-09-20 models/ttm/rbf_basis.R

@@ -1,105 +1,0 @@
-# Radial Basis helpers for t-basis (base R only)
-
-# Local coordinates (dimensionless)
-rbf_local_coords <- function(x, mu, sigma) {
-  s <- pmax(as.numeric(sigma), .Machine$double.eps)
-  (as.numeric(x) - as.numeric(mu)) / s
-}
-
-# Gaussian RBF in local coords
-rbf <- function(xloc) {
-  exp(-0.5 * (as.numeric(xloc)^2))
-}
-
-# Integral of Gaussian RBF in local coords, such that d/dxloc irbf = rbf
-# Using relation erf(x/sqrt(2)) = 2*pnorm(x) - 1
-irbf <- function(xloc) {
-  sqrt(pi / 2) * (2 * stats::pnorm(as.numeric(xloc)) - 1)
-}
-
-# Left-end tail term (smooth, asymptotically linear with slope 1 as x -> -inf)
-# xloc = (x - a) / sigma  (a is left boundary), tau = sigma
-let_term <- function(xloc, sigma) {
-  tau <- pmax(as.numeric(sigma), .Machine$double.eps)
-  # LET(x) = -tau * log1p(exp(-xloc)) + tau * log(2)
-  -tau * log1p(exp(-as.numeric(xloc))) + tau * log(2)
-}
-
-# Right-end tail term (smooth, asymptotically linear with slope 1 as x -> +inf)
-# xloc = (x - b) / sigma  (b is right boundary), tau = sigma
-ret_term <- function(xloc, sigma) {
-  tau <- pmax(as.numeric(sigma), .Machine$double.eps)
-  # RET(x) =  tau * log1p(exp( xloc)) - tau * log(2)
-  tau * log1p(exp(as.numeric(xloc))) - tau * log(2)
-}
-
-# Place RBF knots by empirical quantiles; local scales by neighbor distances
-# Returns list(mu = centers, sigma = local_scales)
-place_rbf_knots <- function(xk, m) {
-  xk <- as.numeric(xk)
-  m <- as.integer(m)
-  stopifnot(m >= 1L)
-  if (length(xk) == 0L) return(list(mu = numeric(0), sigma = numeric(0)))
-  # Probabilities avoid exact 0/1 to keep centers within data range
-  probs <- if (m == 1L) 0.5 else seq(1, m) / (m + 1)
[…]

### 2025-09-20 models/ttm/ttm_bases.R

@@ -1,203 +1,0 @@
-# Unified design/basis builders for TTM variants (pure base R)
-
-# Robust repo-local source helper (works from tests or scripts)
-.source_repo_file <- function(rel_path) {
-  cand <- character(0)
-  # 1) relative to current working directory
-  cand <- c(cand, rel_path)
-  # 2) relative to this file if available
-  this_file <- tryCatch(normalizePath(sys.frames()[[1]]$ofile, winslash = "/", mustWork = FALSE), error = function(e) NA_character_)
-  if (is.character(this_file) && nzchar(this_file) && !is.na(this_file)) {
-    cand <- c(cand, file.path(dirname(this_file), basename(rel_path)))
-  }
-  # 3) via root_path if defined
-  if (exists("root_path")) cand <- c(cand, file.path(root_path, rel_path))
-  # 4) walk up to find 00_globals.R as repo root
-  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
-  for (i in 1:8) {
-    if (file.exists(file.path(p, "00_globals.R"))) { cand <- c(cand, file.path(p, rel_path)); break }
-    np <- dirname(p); if (identical(np, p)) break; p <- np
-  }
-  cand <- unique(cand)
-  for (pth in cand) { if (file.exists(pth)) { source(pth); return(invisible(TRUE)) } }
-  invisible(FALSE)
-}
-
-# Helper: replicate a row vector into an n-row matrix
-rep_row <- function(vec, n) {
-  v <- as.numeric(vec)
-  matrix(rep(v, each = n), nrow = n, ncol = length(v))
-}
-
-# Gauss–Legendre nodes and weights on [0,1]
-gauss_legendre_nodes <- function(Q) {
-  stopifnot(is.numeric(Q), length(Q) == 1L, Q >= 1, Q == as.integer(Q))
-  Q <- as.integer(Q)
-  if (Q == 1L) return(list(nodes = 0.5, weights = 1.0))
-  i <- seq_len(Q - 1L)
-  b <- i / sqrt(4 * i^2 - 1)
-  J <- matrix(0, Q, Q)
-  for (k in i) {
-    J[k, k + 1L] <- b[k]
-    J[k + 1L, k] <- b[k]
-  }
-  e <- eigen(J, symmetric = TRUE)
-  # Map from [-1,1] to [0,1]
[…]

### 2025-09-20 models/ttm/ttm_conditional.R

@@ -1,113 +1,0 @@
-# Conditional sampling via truncated triangular inversion (composite reference)
-
-# Helper: extract model (S) and optional training X
-.extract_S_and_X <- function(model) {
-  S <- if (is.list(model) && !is.null(model$S)) model$S else model
-  X_tr <- NULL
-  if (is.list(model) && !is.null(model$X_tr)) X_tr <- as.matrix(model$X_tr)
-  if (is.list(S) && !is.null(S$X_tr)) X_tr <- as.matrix(S$X_tr)
-  list(S = S, X_tr = X_tr)
-}
-
-# Core: invert k-th component for a single sample, given current x_prev (raw scale)
-.invert_component_k <- function(S, k, z_target, x_prev_raw, bracket = c(-6, 6)) {
-  mu <- S$mu; sigma <- S$sigma
-  algo <- if (!is.null(S$algo)) S$algo else "marginal"
-  K <- length(mu)
-  # Fast path: marginal linear map
-  if (identical(algo, "marginal")) {
-    # Support both coeffs list (ttm_marginal2) and coeffA/coeffB legacy
-    if (!is.null(S$coeffs)) {
-      a <- as.numeric(S$coeffs[[k]]["a"])
-      b <- as.numeric(S$coeffs[[k]]["b"])
-    } else {
-      b <- exp(S$coeffA[k]); a <- S$coeffB[k]
-    }
-    u_std <- (z_target - a) / b
-    return(mu[k] + sigma[k] * u_std)
-  }
-
-  # Generic 1D monotone inversion via uniroot on standardized u
-  f_eval <- function(u_std) {
-    # Build a 1xK raw vector consistent with u_std at k and x_prev_raw for <k; others at training mean
-    xr <- rep(0, K)
-    # previous raw already supplied
-    if (k > 1) xr[1:(k - 1)] <- as.numeric(x_prev_raw)
-    xr[k] <- mu[k] + sigma[k] * u_std
-    if (k < K) {
-      # fill remaining with mu (u_std=0)
-      xr[(k + 1):K] <- mu[(k + 1):K]
-    }
-    zk <- as.numeric(predict_ttm(S, matrix(xr, nrow = 1), type = "transform")[1, k])
-    zk - z_target
-  }
-  lo <- bracket[1]; hi <- bracket[2]
-  flo <- f_eval(lo); fhi <- f_eval(hi)
-  # Expand bracket if no sign change (up to a few times)
[…]

### 2025-09-20 models/ttm/ttm_core.R

@@ -1,179 +1,0 @@
-# Core helpers shared by TTM variants (marginal, separable, cross-term)
-
-.ttm_std_apply <- function(mu, sigma, X) {
-  X1 <- sweep(X, 2, mu, "-")
-  sweep(X1, 2, sigma, "/")
-}
-
-# Minimal infix helper and config reader used across files
-if (!exists("%||%")) `%||%` <- function(a, b) if (is.null(a)) b else a
-.get_cross_maxit <- function() as.integer(getOption("cross.maxit", 200L))
-.get_cross_quad_nodes <- function(fallback = NULL) {
-  q <- getOption("cross.quad_nodes", NULL)
-  if (!is.null(q)) return(as.integer(q))
-  if (!is.null(fallback)) return(as.integer(fallback))
-  as.integer(32L)
-}
-.get_cross_df_t <- function(fallback = NULL) {
-  d <- getOption("cross.df_t", NULL)
-  if (!is.null(d)) return(as.integer(d))
-  if (!is.null(fallback)) return(as.integer(fallback))
-  as.integer(8L)
-}
-
-.get_cross_lambda_non <- function(Q_used, N, fallback = NULL) {
-  ln <- getOption("cross.lambda_non", NULL)
-  if (!is.null(ln)) return(as.numeric(ln))
-  if (!is.null(fallback) && is.finite(fallback)) return(as.numeric(fallback))
-  0.05 * (as.numeric(Q_used) / as.numeric(N))
-}
-
-.get_cross_lambda_mon <- function(lam_non, Q_used, N, fallback = NULL) {
-  lm <- getOption("cross.lambda_mon", NULL)
-  if (!is.null(lm)) return(as.numeric(lm))
-  if (!is.null(fallback) && is.finite(fallback)) return(as.numeric(fallback))
-  0.5 * as.numeric(lam_non)
-}
-
-.use_cross_analytic_grad <- function() {
-  isTRUE(getOption("cross.use_analytic_grad", TRUE))
-}
-
-.get_cross_workers <- function() {
-  w <- getOption("cross.workers", NULL)
-  if (!is.null(w)) return(as.integer(max(1L, w)))
-  cw <- tryCatch(parallel::detectCores(), error = function(e) 1L)
-  as.integer(max(1L, cw - 1L))
-}
-.get_cross_deg_g <- function() as.integer(getOption("cross.deg_g", 3L))
-
-# Determine algo tag from model object
[…]

### 2025-09-20 models/ttm/ttm_marginal.R

@@ -1,160 +1,0 @@
-# Marginal Triangular Transport Map (TTM) — Minimal, monotone 1-D basis
-# Implements fit_ttm, predict_ttm (with multiple types), and a tiny CLI for train/eval.
-
-# try to source shared standardization util if available
-if (!exists("standardize_train_only")) {
-  p1 <- file.path("models", "ttm", "utils_standardize.R")
-  if (file.exists(p1)) {
-    source(p1)
-  } else if (exists("root_path")) {
-    p2 <- file.path(root_path, "models", "ttm", "utils_standardize.R")
-    if (file.exists(p2)) source(p2)
-  }
-}
-
-.std_stats <- function(X) {
-  mu <- colMeans(X)
-  sigma <- apply(X, 2, sd) + .Machine$double.eps
-  list(mu = mu, sigma = sigma)
-}
-
-.standardize <- function(X, mu, sigma) {
-  X1 <- sweep(X, 2, mu, "-")
-  sweep(X1, 2, sigma, "/")
-}
-
-# Simple monotone f_k(x) = a_k + b_k x with b_k > 0
-.fit_marginal_per_k <- function(x_std, eps = 1e-6) {
-  # Optimize J(b,a) = 0.5 * sum (a + b x)^2 - n log b
-  n <- length(x_std)
-  xbar <- mean(x_std)
-  v <- stats::var(x_std) + 1e-12
-  b <- max(eps, 1 / sqrt(v))
-  a <- -b * xbar
-  c(a = a, b = b)
-}
-
-fit_ttm_marginal <- function(data, seed = 42) {
-  set.seed(seed)
-  S_in <- if (is.list(data) && !is.null(data$X_tr) && !is.null(data$X_te)) data else {
-    stopifnot(is.matrix(data))
-    if (!exists("split_data")) source("02_split.R")
-    list(X_tr = split_data(data, seed)$X_tr, X_te = split_data(data, seed)$X_te)
-  }
-  X_tr <- S_in$X_tr; X_te <- S_in$X_te
-  K <- ncol(X_tr)
-
-  t_tr <- system.time({
-    if (exists("standardize_train_only")) {
-      st <- standardize_train_only(X_tr)
-      mu <- st$mu; sigma <- st$sigma; X_tr_std <- st$Xs
-    } else {
-      st <- .std_stats(X_tr)
-      mu <- st$mu; sigma <- st$sigma
-      X_tr_std <- .standardize(X_tr, mu, sigma)
-    }
[…]

### 2025-09-20 models/ttm/ttm_separable.R

@@ -1,112 +1,0 @@
-# Separable triangular map: S_k(x) = g_k(x_<k>) + f_k(x_k)
-# Maps-from-samples (Eq. 38/39). Uses unified bases and core predictor.
-
-# Be robust to different working directories in tests/scripts
-if (!exists("build_f") || !exists("d_build_f") || !exists("build_g")) {
-  if (exists("root_path")) {
-    src1 <- file.path(root_path, "models", "ttm", "ttm_bases.R")
-  } else {
-    src1 <- file.path("models", "ttm", "ttm_bases.R")
-  }
-  if (file.exists(src1)) source(src1)
-}
-if (!exists("ttm_forward")) {
-  if (exists("root_path")) {
-    src2 <- file.path(root_path, "models", "ttm", "ttm_core.R")
-  } else {
-    src2 <- file.path("models", "ttm", "ttm_core.R")
-  }
-  if (file.exists(src2)) source(src2)
-}
-
-.std_stats <- function(X) {
-  mu <- colMeans(X)
-  sigma <- apply(X, 2, sd) + .Machine$double.eps
-  list(mu = mu, sigma = sigma)
-}
-
-.standardize <- function(X, mu, sigma) {
-  sweep(sweep(X, 2, mu, "-"), 2, sigma, "/")
-}
-
-fit_ttm_separable <- function(data, degree_g = 2L, lambda = 0.0, eps = 1e-6, seed = 42) {
-  set.seed(seed)
-  S_in <- if (is.list(data) && !is.null(data$X_tr) && !is.null(data$X_te)) data else {
-    stopifnot(is.matrix(data))
-    if (!exists("split_data")) source("02_split.R")
-    list(X_tr = split_data(data, seed)$X_tr, X_te = split_data(data, seed)$X_te)
-  }
-  X_tr <- S_in$X_tr; X_te <- S_in$X_te
-  K <- ncol(X_tr); N <- nrow(X_tr)
-
-  time_train <- system.time({
-    st <- .std_stats(X_tr); mu <- st$mu; sigma <- st$sigma
-    Xs <- .standardize(X_tr, mu, sigma)
-    coeffs <- vector("list", K)
-    I_N <- diag(N)
-    for (k in seq_len(K)) {
-      x_prev <- if (k > 1) Xs[, 1:(k - 1), drop = FALSE] else matrix(0, N, 0)
-      xk <- Xs[, k]
[…]

### 2025-09-20 scripts/export_ttm_sources.R

@@ -1,77 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Export all TTM R sources into a single text file with per-file headers.
-# Usage examples:
-#   Rscript scripts/export_ttm_sources.R
-#   Rscript scripts/export_ttm_sources.R --out=ttm_all_scripts.txt
-#   Rscript scripts/export_ttm_sources.R --src=models/ttm --out=out/ttm_sources.txt
-
-args <- commandArgs(trailingOnly = TRUE)
-
-print_help <- function() {
-  cat("Export TTM sources to a single text file\n")
-  cat("\nOptions:\n")
-  cat("  --src=DIR   Source directory (default: models/ttm)\n")
-  cat("  --out=FILE  Output file (default: ttm_all_scripts.txt in repo root)\n")
-  cat("  -h, --help  Show this help\n")
-}
-
-if (length(args) && any(args %in% c("-h", "--help"))) {
-  print_help(); quit(status = 0)
-}
-
-# Try to find the repository root (contains 00_globals.R)
-get_root <- function() {
-  if (exists("root_path", inherits = FALSE)) return(root_path)
-  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
-  for (i in 1:10) {
-    if (file.exists(file.path(p, "00_globals.R"))) return(p)
-    np <- dirname(p); if (identical(np, p)) break; p <- np
-  }
-  normalizePath(getwd(), winslash = "/", mustWork = FALSE)
-}
-
-root <- get_root()
-
-# Defaults
-src_dir <- file.path(root, "models", "ttm")
-out_file <- file.path(root, "ttm_all_scripts.txt")
-
-# Parse key=value args
-if (length(args)) {
-  for (a in args) {
-    if (grepl("^--src=", a)) src_dir <- normalizePath(sub("^--src=", "", a), winslash = "/", mustWork = FALSE)
-    if (grepl("^--out=", a)) {
-      of <- sub("^--out=", "", a)
-      # If relative, interpret relative to root
-      if (!grepl("^/|^[A-Za-z]:", of)) of <- file.path(root, of)
-      out_file <- normalizePath(of, winslash = "/", mustWork = FALSE)
-    }
-  }
-}
-
-if (!dir.exists(src_dir)) {
-  stop(sprintf("Source directory not found: %s", src_dir))
-}
-
-files <- list.files(src_dir, pattern = "\\.R$", full.names = TRUE)
[…]

### 2025-09-20 scripts/halfmoon_plot.R

@@ -1,372 +1,0 @@
-#' Fit half-moon models and create contour plots
-#'
-#' Provides two helper functions for working with the two-moons data
-#' splits.
-
-fit_halfmoon_models <- function(S, seed = NULL, order_mode = "as-is") {
-  seed <- if (!is.null(seed)) as.integer(seed) else if (!is.null(S$meta$seed)) S$meta$seed else 42L
-  set.seed(seed)
-  cfg <- list(list(distr = "norm"), list(distr = "norm"))
-  M_true <- fit_TRUE(S, cfg)
-  class(M_true) <- c("true_marginal_model", class(M_true))
-  # Ensure copula_np fitter is available without side effects
-  if (!exists("fit_copula_np", mode = "function")) {
-    src <- file.path("models", "copula_np.R")
-    if (file.exists(src)) source(src)
-  }
-  # Stabilize cross-term tails and regularization for half-moon as in main.R
-  options(cross.lambda_non = 0.02, cross.lambda_mon = 0.02, cross.df_t = 4L)
-  mods <- list(
-    true = M_true,
-    trtf = tryCatch(fit_TRTF(S, cfg, seed = seed), error = function(e) NULL),
-    ttm = fit_ttm(S, algo = "marginal",  seed = seed)$S,
-    ttm_sep = fit_ttm(S, algo = "separable", seed = seed)$S,
-    ttm_cross = fit_ttm(S, algo = "crossterm", seed = seed,
-                        deg_g = 2L, df_t = 6L, Q = 16L,
-                        lambda = NA_real_, Hmax = 6L, maxit = 100L)$S
-  )
-  # Add nonparametric copula baseline (deterministic under same seed)
-  if (exists("fit_copula_np", mode = "function")) {
-    mods$copula_np <- fit_copula_np(S, seed = seed)
-  }
-  mods
-}
-
-compute_limits <- function(S, pad = 0.05) {
-  Xall <- rbind(S$X_tr, S$X_te)
-  xr <- range(Xall[, 1])
-  yr <- range(Xall[, 2])
-  dx <- pad * diff(xr)
-  dy <- pad * diff(yr)
-  list(xlim = xr + c(-dx, dx), ylim = yr + c(-dy, dy))
-}
-
-#' S3-Predict für das TRUE-Marginalmodell, kompatibel zur Predict-API
-#'
[…]

### 2025-09-22 report/MSc_Report.Rnw
.

@@
 \chapter*{Abstract}
-% TODO: add abstract text.
+% Checklist (Abstract): [x] integrate supplied summary [ ] confirm PDF rebuild
 
-\chapter*{Acknowledgments}
-% TODO: add acknowledgments.
+This thesis studies multivariate density estimation for tabular data within a transport-based evaluation framework.
+All models standardize features and learn a monotone lower-triangular map $S \colon \mathbf{u} \to \mathbf{z}$ to a Gaussian reference distribution.
+The estimators accumulate exact log-Jacobians in standardized space with an affine correction for reporting on the data scale.
+We compare separable triangular transport maps (TTM-Sep), additive-predictor Transformation Random Forests (TRTF), and copula baselines under the shared preprocessing pipeline.
+Additive TRTF induces the same separable triangular likelihood as TTM-Sep via the probability integral transform.
+The resulting estimator yields exact likelihoods, transparent conditionals, and linear per-sample cost but restricts context-dependent shape.
+
+Benchmarks include synthetic Half-Moon and a four-dimensional autoregressive generator, plus the UCI POWER, GAS, HEPMASS, and MINIBOONE datasets.
+For Half-Moon with $n = 250$, joint negative log-likelihoods (nats) were $1.71$ for TRTF, $1.93$ for TTM-Sep, and $1.54$ for the copula baseline.
+For the four-dimensional generator, TRTF achieved $4.53$, TTM-Sep $5.66$, and the copula baseline $5.45$.
+Permutation averages confirmed ordering sensitivity for triangular maps.
+On MINIBOONE with $K = 43$ and $N = 2{,}500$, TRTF reached $-30.01$ test log-likelihood, whereas flow baselines remained near $-11.7$.
+
+Findings indicate that TRTF narrows the gap to oracle likelihoods on low-dimensional synthetic data.
+The method trails modern flows on high-dimensional real data because separability limits context-dependent shape.
+Copula baselines remain competitive in low dimensions.
+This thesis records guidance and limitations, including ordering effects and fragility in sparse regimes.
 

### 2025-09-22 report/MSc_Report.tex
.

@@
 
-Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty.
-Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices.
-A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals.
-Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background.
-We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction.
-All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.
+% Checklist: integrate unified transport framing, enforce US spelling, add dynamic refs.
+Multivariate density estimation supports likelihood-based modeling, anomaly detection, simulation, and decision making under uncertainty.
+Tabular datasets often combine moderate to high dimension with context-dependent conditional structure.
+Conditional variance can change with predictors, and conditional skewness or modality can depend on earlier variables.
+A transport perspective addresses these challenges by coupling the target to a simple reference through an invertible map.
+This perspective enables exact likelihoods, transparent conditionals, and efficient sampling through a shared computational backbone \citep{rosenblatt1952remarks,knothe1957contributions}.
+This thesis compares three estimator families inside a single evaluation frame with matched objectives and diagnostics.
+The frame evaluates lower-triangular transport maps, Transformation Random Forests interpreted via probability integral transforms, and copulas that decouple marginals from dependence \citep{hothorn2017transformation,hothorn2018conditional,sklar1959fonctions}.
+\medskip
 
-This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence.
-Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required.
-The following sections outline our contributions, the research questions we address, and the scope of the work.
+We work in standardized coordinates to unify Jacobians, gradients, and reporting conventions. Let $x \in \mathbb{R}^K$ denote features on the original scale, and define
+\begin{equation}
+  u = T_{\text{std}}(x) = (x - \mu) \oslash \sigma,
+  \qquad \sigma_k > 0,\label{eq:ch1-standardization}
+\end{equation}
+which standardizes each feature using training statistics only and fixes the coordinates where derivatives and Jacobians are evaluated.
+Let $S: u \mapsto z$ be a monotone lower-triangular map, and let $\eta$ denote the standard normal density.
+The change-of-variables identity gives
+\begin{equation}
+  \log \pi_U(u) = \log \eta\big(S(u)\big) + \log \big|\det \nabla_u S(u)\big|.
+  \label{eq:ch1-pullback}
+\end{equation}
+Equation~\eqref{eq:ch1-pullback} splits the log density into a reference-fit term and a Jacobian term, isolating modeling capacity from the exact volume correction contributed by the map.
+For triangular $S$, the determinant factorizes as
+\begin{equation}
+  \log \big|\det \nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k} S_k(u_{1:k}).
+  \label{eq:ch1-triangular-det}
+\end{equation}
+Equation~\eqref{eq:ch1-triangular-det} reduces the determinant to a sum of one-dimensional log derivatives, yielding linear per-sample complexity in $K$ and improving numerical stability.
+Reported log densities on the original scale apply the diagonal affine correction
+\begin{equation}
+  \log \pi_X(x) = \log \pi_U\!\big(T_{\text{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k.
+  \label{eq:ch1-affine-correction}
+\end{equation}
+Equation~\eqref{eq:ch1-affine-correction} subtracts a constant offset determined by the training scales and ensures comparability across models that share standardized coordinates.
+\medskip
 
-\section{Contributions and research questions}
+We study separable triangular components for clarity, efficiency, and interpretability. Component $k$ decomposes into a context shift and a univariate monotone shape,
+\begin{equation}
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),\qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k),
+  \label{eq:ch1-separable}
+\end{equation}
+which stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.
 
-We make three contributions within a single, notation-coherent framework.
+% Checklist: add thesis overview, enforce dynamic references, keep ASCII hyphenation.
+\section{Thesis and Problem Statement}\label{sec:ch1-problem}
+This thesis investigates tabular multivariate density estimation within a unified transport-based evaluation frame.
+We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests with axis-parallel splits (TRTF), and copula baselines.
+\medskip
+
+A transport perspective couples standardized data to a Gaussian reference through a monotone lower-triangular map.
+This structure yields exact likelihoods, transparent conditionals, exact inversion by back substitution, and linear per-sample evaluation.
+The Rosenblatt and Knothe rearrangements justify the triangular coupling for any variable order \citep{rosenblatt1952remarks,knothe1957contributions}.
+\medskip
+
+We standardize features using training statistics only.
+Equation~\eqref{eq:ch1-standardization} defines the standardized coordinates $u$ used for evaluation.
+All derivatives and Jacobians are computed in $u$.
+The diagonal affine correction in Equation~\eqref{eq:ch1-affine-correction} reports log densities on the original scale $x$.
+This convention keeps objectives, diagnostics, and comparisons interoperable across estimators and datasets.
+All negative log-likelihoods are reported in nats.
+\medskip
+
+We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\phi$ and $\Phi$.
+Triangular transport maps are abbreviated TTM, with the separable variant labeled TTM-Sep.
+Transformation Random Forests are abbreviated TRTF, and the axis-parallel implementation is labeled TRTF.
+Copulas decouple marginals from dependence and serve as interpretable baselines.
+\medskip
+
+Separable triangular maps decompose each component into a context shift and a univariate monotone shape as shown in Equation~\eqref{eq:ch1-separable}.
+The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant.
+TRTF implements the same separable triangular likelihood via the probability integral transform and an additive predictor.
+Copulas preserve explicit marginals and model dependence on the unit hypercube.
+
+We evaluate all estimators under a single protocol that records test log-likelihoods, conditional diagnostics, and compute.
+The protocol fixes the map direction from standardized data to the Gaussian reference, which preserves separability and linear evaluation cost.
+Section~\ref{sec:evaluation-protocol} details the metrics, calibration diagnostics based on probability integral transforms, and timing conventions.
+
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} visualizes the pipeline by showing standardization $u = T_{\text{std}}(x)$, the triangular transport branch containing TTM-Sep and TRTF, and the copula branch.
+Both branches feed the reported outputs, namely log density, conditionals, sampling, calibration, and compute, under the shared frame.
+
+The central problem is to determine when separability is appropriate for tabular data.
+We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention.
+Ordering effects, conditional calibration, and computational trade-offs address this question.
+\medskip
+
+On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MINIBOONE benchmark it improves on Gaussian references but trails published flow baselines.
+Chapter~\ref{ch:dataanalysis} presents the evidence and discusses these comparisons.
+
+\paragraph{Non-goals.} We do not treat high-capacity normalizing flows as primary models, and we restrict nonparametric copulas to low dimensions.
+Section~\ref{sec:ch1-contributions} states the scope and non-goals for reference.
+
+% Checklist: consolidate notation, relocate figure, keep one-claim sentences.
+\section{The Transport Frame on One Page}\label{sec:ch1-frame}
+This section fixes the evaluation frame used across triangular transport maps, Transformation Random Forests, and copulas.
+It establishes notation, equations, and reporting conventions for the remainder of the thesis.
+The frame places all derivatives and Jacobians in standardized coordinates, applies a single diagonal affine correction on the original scale, and yields linear per-sample evaluation through a lower-triangular map.
+
+Train-only standardization defines the evaluation coordinates via Equation~\eqref{eq:ch1-standardization}.
+The transformation uses training means and scales once and prevents information leakage into evaluation.
+All gradients and Jacobians act on $u$, which keeps objectives comparable across estimators and datasets.
+We convert to the original scale only when reporting.
+\medskip
+
+Let $S:u\mapsto z$ be a monotone lower-triangular map, and let $\eta(z)$ denote the $K$-variate standard normal density.
+The pullback identity in Equation~\eqref{eq:ch1-pullback} evaluates the reference at $S(u)$ and applies the exact volume correction.
+The split isolates model fit from geometry: the first term measures closeness to the Gaussian reference, and the second term contributes the exact Jacobian adjustment implied by the map.
+\medskip
+
+Lower-triangular structure with strictly positive diagonal partial derivatives factorizes the Jacobian as in Equation~\eqref{eq:ch1-triangular-det}.
+The factorization yields $\mathcal{O}(K)$ per-sample time for likelihood evaluation and improves numerical stability.
+It also guarantees global invertibility by back substitution consistent with the Rosenblatt and Knothe rearrangements.
+\medskip
+
+Reported log densities on the original scale apply only the diagonal affine correction implied by standardization.
+Equation~\eqref{eq:ch1-affine-correction} provides that adjustment.
+This convention keeps training and evaluation in $u$-space and converts to $x$-space at report time.
+All negative log-likelihoods are reported in nats, which fixes units across tables and figures.
+\medskip
+
+For clarity and efficiency we use separable triangular components of the form in Equation~\eqref{eq:ch1-separable}.
+Separable structure shifts location through $g_k$ and fixes conditional shape through the univariate monotone $h_k$.
+The Jacobian contribution depends only on $u_k$, which simplifies inversion by back substitution and stabilizes the log-determinant accumulation.
+The constraint also clarifies limits because conditional variance, skewness, and modality do not vary with context under separability.
+
+The core operations follow the same path for all estimators.
+\paragraph{Core operations.}
 \begin{enumerate}
-  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
-  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
+  \item First, standardize input features with training statistics using Equation~\eqref{eq:ch1-standardization}.
+  \item Then, learn a monotone lower-triangular map $S$ to the standard normal reference and exploit Equation~\eqref{eq:ch1-pullback}.
+  \item Finally, evaluate likelihoods, conditionals, and samples through Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction}.
 \end{enumerate}
 
-Two questions motivate the empirical study.
-\begin{enumerate}
-  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines?
-We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
-  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)?
-We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
-\end{enumerate}
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline.
+The input $x$ enters standardization to produce $u$.
+The triangular branch houses TTM-Sep and TRTF, whereas the copula branch couples fitted marginals to a dependence density.
+Both branches feed the same reported quantities, namely log density, conditionals, sampling, calibration, and compute.
+The appendix placement preserves the full schematic while keeping Chapter~\ref{ch:intro} focused on the narrative.
 
-\section{Scope and positioning}
+Notation remains consistent.
+We write $\eta$ for the $K$-variate standard normal density, and $\phi$ and $\Phi$ for the univariate standard normal density and CDF.
+We reserve $u$ for standardized coordinates and $x$ for original coordinates, and we compute all derivatives with respect to $u$.
+These choices align symbols across Chapters~\ref{ch:intro}--\ref{ch:dataanalysis} and prevent ambiguity in later diagnostics and tables.
 
-We focus on triangular transport maps, transformation random forests, and copulas.
-Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions.
-Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used.
-Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions.
-Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.
+This one-page frame removes duplicated exposition from Chapter~\ref{ch:background}.
+It establishes where logs and Jacobians live and makes complexity, inversion, and units explicit before the comparisons that follow.
+Section~\ref{sec:ch1-problem} documented the motivation, and Section~\ref{sec:ch1-contributions} states the resulting contributions and research questions.
 
-\section{Takeaways and roadmap}
+% Checklist: map contributions to evidence, keep bridges to later sections, ensure dynamic refs.
+\section{Contributions and Research Questions}\label{sec:ch1-contributions}
+This section states the contributions of the thesis and formulates the research questions.
+It also maps both elements to the chapters and figures that deliver the evidence.
+The shared transport frame fixes standardized coordinates, keeps all derivatives and Jacobians in $u$-space, and applies a single diagonal affine correction on the original scale for reporting; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline and anchors the comparisons that follow.
+All negative log-likelihoods are reported in nats, and evaluation uses linear-time lower-triangular maps.
 
-A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail.
-Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests.
-Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results.
-Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
+The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests with axis-parallel splits, and copula baselines.
+The view specifies standardization $u = T_{\text{std}}(x)$ and the pullback identity with a monotone lower-triangular map.
+It also fixes the Jacobian factorization and the affine correction for reporting, which together align objectives and diagnostics across estimators.
+Equations~\eqref{eq:ch1-standardization}--\eqref{eq:ch1-affine-correction} and Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} establish these conventions and remove duplication in later chapters.
+
+The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting.
+We evaluate TTM-Sep, TTM-Marg, TRTF, and copulas on synthetic generators and on real tabular data.
+The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation.
+Section~\ref{sec:evaluation-protocol} defines the protocol, Section~\ref{sec:synthetic-results} presents the synthetic and autoregressive results, and Section~\ref{sec:realdata} positions our measurements against published normalizing-flow baselines where appropriate.
+
+The third contribution distills practical guidance from the unified frame and the benchmarks.
+We state operational choices that preserve comparability, highlight ordering sensitivity and separability limits, and summarize when copulas serve as informative baselines.
+Chapter~\ref{ch:conclusion} consolidates these points as actionable recommendations and records limitations that motivate richer parameterizations or alternative predictors.
+
+Two questions drive the empirical study and bind the contributions to specific measurements.
+The first question asks how TRTF compares with TTM-Sep and copula baselines on synthetic data.
+All estimators share the transport frame in this comparison.
+We answer by reporting average test negative log-likelihoods in nats, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost.
+Section~\ref{sec:synthetic-results} provides the corresponding tables and figures.
+
+The second question asks how closely our TRTF results on real benchmarks approach the published performance of modern normalizing flows under the standard preprocessing.
+We answer by placing our test log-likelihoods beside reported numbers from the literature.
+The gaps are interpreted through the separable Jacobian constraint and compute profiles.
+Section~\ref{sec:realdata} reports these comparisons, and Chapter~\ref{ch:conclusion} interprets their implications for model choice.
+
+Scope and non-goals maintain focus and ensure reproducibility.
+We study separable triangular maps and TRTF with additive predictors and compute all derivatives and Jacobians in standardized coordinates.
+The map direction $S:u \to z$ remains fixed for evaluation and inversion.
+Copulas include Gaussian dependence and a low-dimensional nonparametric variant used strictly as a diagnostic baseline.
+We treat high-capacity flows as external references rather than primary models, and we do not evaluate non-additive TRTF variants or cross-term triangular maps in this chapter.
+Chapter~\ref{ch:background} records the formal assumptions and notation.
+Section~\ref{sec:datasets-preprocessing} details preprocessing, seeds, and reporting conventions that keep results comparable across datasets and estimators.
+
+Taken together, these commitments make the comparisons interpretable, keep units and complexity explicit, and prepare the reader for the empirical evidence that answers the two questions under a single, transparent evaluation frame.
 
@@
 
-We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$).
-All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated.
-This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.
+\section{Transport Frame and Notation}\label{sec:transport-frame}
 
-\section{Change of Variables in Standardized Coordinates}
+This section fixes the standardized coordinate system, notation, and algebraic identities used throughout the thesis.
+The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters.
+We summarise the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorisation that drives evaluation and inversion.
 
-The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
+We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardisation map
 \begin{equation}
-  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
-  \label{eq:pullback}
+  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,\label{eq:transport-standardise}
 \end{equation}
-Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate.
-Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$.
-The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions.
-This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.
+where $\mu$ and $\sigma$ denote the empirical mean and standard deviation estimated on the training split and $\oslash$ denotes elementwise division.
+In words, we shift and rescale features once, using training data only, and keep all derivatives and Jacobians in $u$-space to avoid leakage and to ensure comparability across estimators.
 
-It is useful to record the complementary pushforward statement.
-If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$.
-The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion.
-In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.
+The standardised density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$.
+Throughout the thesis the reference is the $K$-variate standard normal density $\eta(z)$.
+The pullback identity then reads
+\begin{equation}
+  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|,\label{eq:transport-pullback}
+\end{equation}
+which evaluates the reference at $S(u)$ and applies the exact volume correction given by the Jacobian determinant.
+Reporting log densities on the original scale requires only the diagonal affine correction implied by standardisation,
+\begin{equation}
+  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.\label{eq:transport-affine}
+\end{equation}
+We therefore differentiate with respect to $u$, and we convert to $x$-scale only at reporting time.
 
-\section{Triangular Structure and Its Consequences}
+The transport is assumed to be lower triangular and componentwise monotone,
+\begin{equation}
+  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), \qquad \partial_{u_k}S_k(u_{1:k})>0,\label{eq:transport-triangular}
+\end{equation}
+so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorises into a sum of one-dimensional log derivatives,
+\begin{equation}
+  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).\label{eq:transport-det}
+\end{equation}
+The factorisation yields $\mathcal{O}(K)$ evaluation cost per sample, improves numerical stability, and guarantees global invertibility: strict monotonicity lets us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.
 
-A triangular map decomposes component-wise,
+Table~\ref{tab:transport-notation} consolidates the notation used in this transport frame.
+All derivatives and Jacobians act on $u$; the affine correction \eqref{eq:transport-affine} converts log densities back to $x$ for reporting.
+The remainder of this chapter adopts this frame.
+Section~\ref{sec:transport-separable} details the separable triangular parameterisation used for direct transports.
+Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform.
+Section~\ref{sec:transport-copula} places copulas in the same reporting convention.
+
+\begin{table}[t]
+  \centering
+  \caption{Notation for the transport frame used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
+All derivatives and Jacobians are taken with respect to $u$; log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.}
+  \label{tab:transport-notation}
+  \begin{tabular}{ll}
+    \hline
+    Symbol & Meaning \\ 
+    \hline
+    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
+    $T_{\mathrm{std}}$ & Standardisation map using training $(\mu,\sigma)$ \\ 
+    $u = T_{\mathrm{std}}(x)$ & Standardised evaluation coordinates \\ 
+    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
+    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
+    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
+    $\eta(z)$ & $K$-variate standard normal density \\ 
+    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
+    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space, respectively \\ 
+    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
+    $K$ & Dimension of the feature vector \\ 
+    \hline
+  \end{tabular}
+\end{table}
+
+\section{Separable Triangular Maps and Transformation Random Forests as Transport}\label{sec:transport-separable}
+
+% Checklist: enforce US spelling, ensure dynamic cross-references, keep equation labels aligned.
+This section unifies separable triangular maps and Transformation Random Forests (TRTF) within the transport frame fixed in Section~\ref{sec:transport-frame}.
+Both estimators realize a monotone lower-triangular map $S:u\mapsto z$ that couples the standardized target to the Gaussian reference $\eta$.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the shared backbone and locates \textsc{TTM-Sep} and \textsc{TRTF} on the transport branch introduced in Chapter~\ref{ch:intro}.
+We focus on shared likelihood identities, modeling assumptions, and limits of separability, and defer implementation details to Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.
+
+The goal is to state a single likelihood for both constructions, clarify what separability permits, and identify failure modes that motivate richer parameterizations.
+We do not pursue non-additive TRTF predictors, cross-term triangular maps, or ordering heuristics in this section; Chapter~\ref{ch:dataanalysis} evaluates those choices empirically and Appendix~\ref{ch:appendix} documents routines and defaults.
+
+We adopt the notation introduced in Section~\ref{sec:transport-frame}.
+Coordinates satisfy $u=T_{\mathrm{std}}(x)$, the reference density is $\eta(z)$, and the pullback identity \eqref{eq:transport-pullback} gives $\pi_U(u)=\eta(S(u))\,|\det\nabla_u S(u)|$.
+The map is lower-triangular with strictly positive diagonal partial derivatives, which yields the sum decomposition in Equation~\eqref{eq:transport-det}.
+These conventions keep derivatives in $u$-space and apply the affine correction \eqref{eq:transport-affine} only when reporting $\log \pi_X(x)$.
+
+We restrict attention to separable triangular maps. Component $k$ decomposes into a context shift and a univariate monotone shape,
 \begin{equation}
-  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
-  \label{eq:triangular-map}
+  S_k(u_{1:k}) \;=\; g_k(u_{1:k-1}) + h_k(u_k),\qquad \log \partial_{u_k}S_k(u_{1:k}) \;=\; \log h_k'(u_k),\label{eq:transport-separable}
 \end{equation}
-and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
+which fixes context effects in $g_k$ and reserves $h_k$ for the one-dimensional marginal shape.
+Intuitively, earlier coordinates translate the location, while the conditional shape along $u_k$ remains fixed across contexts.
+The Jacobian contribution depends only on $u_k$, which reduces per-sample evaluation cost and simplifies inversion.
+
+Substituting the standard normal reference into \eqref{eq:transport-pullback} produces a separable objective,
 \begin{equation}
-  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
-  \qquad
-  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
-  \label{eq:triangular-det}
+  \log \pi_U(u) \;=\; \sum_{k=1}^{K}\Big[\log \varphi\!\big(S_k(u_{1:k})\big) + \log h_k'(u_k)\Big],\label{eq:transport-likelihood}
 \end{equation}
+where $\varphi$ denotes the univariate standard normal density.
+Equation~\eqref{eq:transport-likelihood} splits the log density into a reference fit and an exact volume correction.
+In plain language, the model evaluates how Gaussian each transformed coordinate appears, then corrects for the local stretch induced by $h_k$.
+The same decomposition produces linear per-sample time in $K$ and stable accumulation of log derivatives.
 
-Three consequences matter in practice.
-First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$.
-Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}.
-Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components.
-Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.
+The negative log-likelihood per sample takes the quadratic-plus-barrier form
+\begin{equation}
+  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big],\label{eq:transport-loss}
+\end{equation}
+which follows because $\log \varphi(t) = -\tfrac{1}{2}t^2 - \tfrac{1}{2}\log(2\pi)$ and constants independent of the parameters drop out.
+Equation~\eqref{eq:transport-loss} pulls each component toward the reference while preventing degenerate derivatives through the log barrier.
+In practice we enforce $h_k'(u_k)>0$ by construction and control tails with mild regularization; implementation choices appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.
 
-Existence and anisotropy deserve a brief remark.
-For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}.
-Different orderings induce different maps.
-While all are valid, their sparsity and approximation difficulty may vary with order.
-When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling.
-In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.
+Separable structure encodes clear modeling assumptions.
+Conditional variance, skewness, and modality do not change with the preceding coordinates once $g_k$ shifts location.
+Consequently, separable maps can underfit heteroskedastic or multimodal conditionals, which manifests as U-shaped or inverted-U probability integral transform (PIT) diagnostics.
+Variable ordering also matters for finite bases because triangular transports are anisotropic, even though a Knothe--Rosenblatt rearrangement exists for any ordering \citep{rosenblatt1952remarks,knothe1957contributions}.
+These caveats guide the robustness checks in Chapter~\ref{ch:dataanalysis}.
 
-Finally, we keep terminology consistent.
-A log density is $\log \hat\pi(\cdot)$ at a point.
-A (test) log likelihood is the dataset average of log densities.
-Its negative is the negative log likelihood (NLL) in nats; lower is better.
-These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.
+\subsection{Transformation Random Forests within the Transport Frame}\label{sec:transport-trtf}
 
-\section{Separable and Cross-Term Triangular Transports}
+Transformation Random Forests \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation} fit into the same transport frame through the probability integral transform.
+Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by a TRTF for coordinate $k$.
+The induced triangular component is
+\begin{equation}
+  S_k(u_{1:k}) \;=\; \Phi^{-1}\!\Big(\widehat F_k(u_k \mid u_{1:k-1})\Big),\label{eq:transport-trtf-map}
+\end{equation}
+which maps conditionals to standard normal margins.
+In plain language, TRTF predicts a conditional CDF, then the probit transform places the result on the Gaussian reference scale.
+Differentiating $\Phi\!\big(S_k(u_{1:k})\big)=\widehat F_k(u_k \mid u_{1:k-1})$ with respect to $u_k$ yields
+\begin{equation}
+  \widehat \pi_k(u_k \mid u_{1:k-1}) \;=\; \varphi\!\big(S_k(u_{1:k})\big)\,\partial_{u_k}S_k(u_{1:k}),\label{eq:transport-trtf-likelihood}
+\end{equation}
+which is exactly the pullback factor in Equation~\eqref{eq:transport-likelihood}. Summing over $k$ recovers Equation~\eqref{eq:transport-likelihood} in standardized coordinates.
 
-A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.
+The additive-predictor TRTF used in this thesis yields a separable transport. Under the model
+\begin{equation}
+  \widehat F_k(u_k \mid u_{1:k-1}) \;=\; \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),\label{eq:transport-trtf-additive}
+\end{equation}
+we obtain
+\begin{equation}
+  S_k(u_{1:k}) \;=\; h_k(u_k) + g_k(u_{1:k-1}),\qquad \partial_{u_k}S_k(u_{1:k}) \;=\; h_k'(u_k),\label{eq:transport-trtf-separable}
+\end{equation}
+so TRTF implements the same separable triangular likelihood as the direct parameterization in Equation~\eqref{eq:transport-separable}.
+Monotonicity in $u_k$ holds by construction, the Jacobian depends only on $u_k$, and inversion proceeds by back-substitution identical to the separable map.
+This equivalence underpins the empirical comparisons in Chapter~\ref{ch:dataanalysis}.
 
-\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
+The equivalence also clarifies limits.
+Additive TRTF predictors shift location but cannot alter conditional shape with context, which mirrors the separable constraint.
+Axis-aligned partitions stabilize estimation, yet they do not remove residual multimodality when the conditional shape varies with $u_{1:k-1}$.
+These limits are visible in PIT diagnostics and conditional negative log-likelihood decompositions on synthetic studies.
+
+We emphasize operational scope and supporting references.
+All derivatives and Jacobians are computed in standardized coordinates, evaluation uses the triangular pullback, and reported log densities on the original scale include the affine correction \eqref{eq:transport-affine}.
+Implementation details on basis choices for $h_k$, feature construction for $g_k$, regularization, derivative clipping, timing, and memory footprints appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}, which also provides pseudo-code for both estimators.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} visualizes how the \textsc{TTM-Sep} and \textsc{TRTF} branches share the same computational path from standardized data to reported likelihoods.
+
+In summary, separable triangular maps and additive-predictor TRTF realize the same lower-triangular likelihood once the data are standardized.
+The shared structure yields exact likelihoods, exact inversion, transparent conditionals, and linear per-sample complexity, but it restricts context-dependent shape.
+Section~\ref{sec:transport-copula} positions copulas within the same reporting convention to decouple marginals from dependence.
+
+\section{Copula Baselines}\label{sec:transport-copula}
+
+% Checklist: align references with transport frame, cite Sklar, enforce US spelling.
+This section positions copulas within the unified transport frame and links their reported likelihoods to the evaluation conventions used for triangular maps and Transformation Random Forests.
+Copulas decouple marginal modeling from dependence modeling by pairing univariate marginals with a separate dependence density on the unit hypercube.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} displays the copula branch beside the triangular branch and highlights how both yield comparable reported log densities under the shared evaluation pipeline.
+
+We begin with pseudo-observations built from training-split marginals.
+Let $\widehat F_k$ denote the strictly increasing empirical or smoothed CDF of $X_k$ estimated on the training split.
+Define the pseudo-observations and their probit transform as
 \begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
-  \label{eq:separable-component}
+  v_k \;=\; \widehat F_k(x_k),\qquad z_k \;=\; \Phi^{-1}(v_k),\label{eq:copula-probit}
 \end{equation}
-so the log Jacobian depends only on $u_k$:
+which map each coordinate to $(0,1)$ and then to $\mathbb{R}$ through the probit function.
+In plain language, the marginals become uniform scores, and $z$ records those scores on a Gaussian scale.
+Mid-ranks and clamping near $(0,1)$ stabilize the transformation in finite samples.
+
+The copula representation combines marginal densities with a dependence factor. The joint log density on the original scale satisfies
 \begin{equation}
-  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
-  \label{eq:separable-jacobian}
+  \log \widehat \pi_X(x) \;=\; \sum_{k=1}^{K} \log \widehat f_k(x_k) + \log c\!\left(v_1,\ldots,v_K\right),\label{eq:copula-logdensity}
 \end{equation}
-The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it.
-Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$.
-Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.
+where $c$ denotes the copula density on $(0,1)^K$.
+Equation~\eqref{eq:copula-logdensity} separates the task into two parts: fit interpretable marginals and correct for dependence through $\log c$.
+The reported quantity already lives on the original scale, so the affine correction in Equation~\eqref{eq:transport-affine} is unnecessary.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} uses the equivalent shorthand $\log c(z)$ because dependence is evaluated through $z=\Phi^{-1}(v)$.
 
-\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
+The independence baseline fixes a lower bound for dependence modeling. Setting $c \equiv 1$ yields
 \begin{equation}
-  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:cross-component}
+  \log \widehat \pi_X^{\mathrm{ind}}(x) \;=\; \sum_{k=1}^{K} \log \widehat f_k(x_k),\label{eq:copula-independence}
 \end{equation}
-so the log Jacobian, and thus the shape of the conditional, depends on the context.
-This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.
+which treats coordinates as independent after marginal fitting. Chapter~\ref{ch:dataanalysis} uses this baseline as a reference point in evaluation tables and figures.
 
-Two small examples make the contrast concrete.
-Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$.
-A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$.
-A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$.
-As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others.
-Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies.
-Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.
+The Gaussian copula specifies elliptical dependence through a correlation matrix $\Sigma$. With $z=\Phi^{-1}(v)$, the copula density admits the closed form
+\begin{equation}
+  c_{\Sigma}(v) \;=\; |\Sigma|^{-1/2}\,\exp\!\Big(-\tfrac{1}{2}\,z^{\top}(\Sigma^{-1} - I)z\Big),\label{eq:copula-gaussian}
+\end{equation}
+which reduces dependence estimation to fitting $\Sigma$ on the transformed scores.
+In plain language, the Gaussian copula bends the joint shape away from independence according to $\Sigma$ while preserving the learned marginals.
 
-\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable.
-Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.
+A low-dimensional nonparametric variant avoids elliptical assumptions at small $K$. We fit a kernel density $\widehat f_Z$ on $z=\Phi^{-1}(v)$ and recover the copula density by
+\begin{equation}
+  c(v) \;=\; \frac{\widehat f_Z\!\big(\Phi^{-1}(v)\big)}{\prod_{k=1}^{K} \varphi\!\big(\Phi^{-1}(v_k)\big)},\label{eq:copula-kde}
+\end{equation}
+which applies the change of variables from $z$ back to $v$ and yields a proper copula density.
+In words, the kernel density models the joint shape of the probit scores, and division by the product of standard normal densities restores the unit-cube scale.
+This approach is viable only for small $K$, where kernel density estimation remains accurate and stable.
+Chapter~\ref{ch:dataanalysis} employs it strictly as a diagnostic baseline.
 
-\section{Transformation Forests as Triangular Transport}
+The transport frame keeps reporting interoperable across modeling branches despite distinct parameterizations.
+Triangular maps and TRTF evaluate the pullback likelihood in standardized coordinates and apply the fixed affine correction \eqref{eq:transport-affine} when mapping back to $x$.
+Copulas operate on $x$ directly through Equation~\eqref{eq:copula-logdensity}, yet Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows how the probit scores $z$ maintain comparability with the Gaussian reference used above.
+This alignment keeps objectives and diagnostics consistent across Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
 
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}.
-This induces a triangular transport component via the probability integral transform,
-\begin{equation}
-  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
-  \label{eq:trtf-map-theory}
-\end{equation}
-
-\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
-\begin{equation}
-  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
-  \label{eq:trtf-likelihood-theory}
-\end{equation}
-showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly.
-Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}.
-TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.
-
-\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
-\begin{equation}
-  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
-  \label{eq:trtf-additive}
-\end{equation}
-so
-\begin{equation}
-  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:trtf-separable}
-\end{equation}
-and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$.
-Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence.
-TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-
-A short note on reporting closes the loop.
-TRTF evaluates conditional densities in standardized coordinates.
-Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter.
-With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.
-
-\section{Takeaways}
-
-We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction.
-Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}.
-Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization.
-TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
+Modeling choices and limits follow from the chosen copula.
+The Gaussian copula imposes elliptical dependence and may misrepresent tail behavior or localized asymmetry.
+The nonparametric variant mitigates these issues only at small dimension and sufficient sample size.
+The independence baseline provides a transparent reference when dependence is weak or data are scarce.
+These caveats motivate treating copulas as interpretable baselines rather than definitive high-dimensional models in the empirical study of Chapter~\ref{ch:dataanalysis}.
+Sklar’s theorem underlies all constructions above and formalizes the decoupling of marginals from dependence \citep{sklar1959fonctions}.
 
@@
 
-This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling programme.
+% Checklist: enforce US spelling, clarify Jacobian phrasing, remove empty sections.
+This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling program.
 Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable.
@@
 
-\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}
+% Checklist: define model abbreviations for concise tables
+\textbf{Model abbreviations.} For the four-dimensional autoregressive study we abbreviate the oracle fits as True-Marg for the marginal reference and True-Joint for the conditional reference.
+Transformation Random Forests appear as TRTF.
+This shorthand matches the axis-parallel implementation (TRTF) used in our experiments.
+We write the marginal triangular transport as TTM-Marg, the separable variant as TTM-Sep, and the copula mixture baseline as Copula.
+We use these labels across tables and figures that report this study.
 
-Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
+\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}\label{sec:setup}
+
+% Checklist: preserve dynamic refs, ensure ASCII hyphenation, align with future sections.
+This section fixes data sources, generators, and preprocessing so likelihoods, calibration, and compute remain comparable across models.
+All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the common affine correction.
+We keep a single triangular-map direction $S:u \rightarrow z$ across methods to avoid mixed objectives.
+
+We standardize features with training-split statistics only.
+Equation~\eqref{eq:transport-standardise} defines $u = T_{\mathrm{std}}(x) = (x-\mu)\oslash \sigma$ with $\sigma_k > 0$.
+We evaluate $\log \pi_U(u)$ through the pullback identity in Equation~\eqref{eq:transport-pullback}, apply the triangular factorization from Equation~\eqref{eq:transport-det}, then convert to $\log \pi_X(x)$ using the diagonal correction in Equation~\eqref{eq:transport-affine}.
+We report average test negative log-likelihoods (NLL) in nats.
+Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the standardized pipeline shared by transport maps, Transformation Random Forests, and copulas.
+
+We use fixed train, validation, and test splits with proportions $0.60/0.20/0.20$ unless a benchmark provides official splits.
+Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures.
+The canonical four-dimensional ordering is $(1,2,3,4)$.
+Robustness to ordering is assessed by averaging over all $4!
+= 24$ permutations.
+We adopt the natural column order for real datasets.
+Transformation Random Forests are the axis-parallel implementation denoted TRTF throughout.
+We fix seeds $\{11, 13, 17, 19, 23\}$ for data generation and model fitting, and we average repeated runs with standard errors to quantify stochastic variability.
+Figure~\ref{fig:autoregressive-joint-calibration} displays the $20\%$ test split for the synthetic calibration study.
+
+The Half-Moon dataset provides a curved, bimodal joint in $K=2$.
+We draw a class $Y \sim \mathrm{Bernoulli}(0.5)$, an angle $\Theta \sim \mathrm{Unif}[0,\pi]$, and additive noise $\varepsilon \sim \mathcal{N}(0, \sigma^2 I_2)$ with $\sigma = 0.10$.
+For $Y = 0$ we set $m(\Theta) = (\cos \Theta, \sin \Theta)$.
+For $Y = 1$ we set $m(\Theta) = (1 - \cos \Theta, -\sin \Theta + 0.5)$.
+The observed $X = m(\Theta) + \varepsilon$.
+The ``True joint'' oracle evaluates the mixture density by numerical quadrature over $\Theta$ with the known Gaussian noise, and the ``True conditional'' oracle conditions on $Y$.
+Figure~\ref{fig:halfmoon-panels} shows representative contour plots at $n=250$, which align with this generator.
+Table~\ref{tab:halfmoon-nll} reports the corresponding NLLs.
+
+The four-dimensional autoregressive generator combines Gaussian, exponential, beta, and gamma components to induce heteroskedasticity, skew, and conditional multimodality.
+The first coordinate is $X_1 \sim \mathcal{N}(0,1)$.
+The second coordinate is independent $X_2 \sim \mathrm{Exp}(\lambda_0)$ with rate $\lambda_0 = 1$.
+The third coordinate lies on $(0,1)$ and is a context-gated mixture of two beta laws, $X_3 \mid X_{1:2} \sim w\,\mathrm{Beta}(\alpha_1, \beta_1) + (1 - w)\,\mathrm{Beta}(\alpha_2, \beta_2)$.
+We set $(\alpha_1, \beta_1) = (2.5, 5.0)$ and $(\alpha_2, \beta_2) = (5.0, 2.5)$.
+The mixing weight is $w = \sigma(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\sigma(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$.
+The fourth coordinate is positive and conditionally heteroskedastic, $X_4 \mid X_{1:3} \sim \tilde{w}\,\mathrm{Gamma}(k_1, r_1(X_2)) + (1 - \tilde{w})\,\mathrm{Gamma}(k_2, r_2(X_2))$.
+We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \sigma(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$.
+The ``True joint'' baseline uses these closed-form conditionals to evaluate the exact joint density, while the ``True marginal'' baseline uses the corresponding univariate marginals and ignores dependence.
+
+% Checklist: summarise generator config, define softmax usage.
+\begin{table}[t]
+  \centering
+  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study. The softplus transform enforces positive rates, shapes, and scales.}
+  \label{tab:autoregressive-config}
+  \begin{tabular}{lll}
+    \hline
+    Coordinate & Distribution & Parameter mapping \\
+    \hline
+    $X_1$ & Normal & Fixed $\mathcal{N}(0,1)$ \\
+    $X_2 \mid X_1$ & Exponential & $\mathrm{rate} = \mathrm{softplus}(X_1)$ \\
+    $X_3 \mid X_{1:2}$ & Beta & $\mathrm{shape}_1 = \mathrm{softplus}(X_2)$; $\mathrm{shape}_2 = \mathrm{softplus}(X_1)$ \\
+    $X_4 \mid X_{1:3}$ & Gamma & $\mathrm{shape} = \mathrm{softplus}(X_3)$; $\mathrm{scale} = \mathrm{softplus}(X_2)$ \\
+    \hline
+  \end{tabular}
+\end{table}
+
+We apply the softplus transform $\mathrm{softplus}(t) = \log(1 + \exp(t))$ to map unconstrained predictors to strictly positive distribution parameters.
+Mixture weights use the logistic gate $\sigma(a)$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one.
+For completeness, the general softmax takes a vector $a$ and returns $\mathrm{softmax}(a)_i = \exp(a_i) / \sum_j \exp(a_j)$.
+This normalization is essential for the beta and gamma mixtures because it translates linear predictors into valid probability weights while preserving differentiability.
+
+Table~\ref{tab:autoregressive-config} lists the intended marginal families by dimension, and Tables~\ref{tab:autoregressive-perm} and~\ref{tab:autoregressive-perm-avg} summarize the permutation and sample-size studies used later in this chapter.
+
+The MINIBOONE benchmark follows the published preprocessing to ensure comparability with flow-based baselines.
+We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes.
+We use the fixed train, validation, and test splits from the benchmark, apply train-only standardization, and avoid any extra pruning of correlated features.
+We report all log-likelihoods in nats and retain the published naming for flow comparators in later tables.
+Section~\ref{sec:realdata} records these steps and provides the dataset context.
+Table~\ref{tab:uci-loglik} reproduces the flow baselines that motivate our TRTF runs.
+
+Additional UCI datasets appear only when we retain them for real-data context.
+POWER keeps household electricity attributes after jittering the minute-of-day encoding, dropping the calendar date and reactive-power column, and adding uniform noise to break ties.
+GAS keeps the \texttt{ethylene\_CO} subset, treats the series as i.i.d., removes strongly correlated attributes, and retains an eight-dimensional representation.
+HEPMASS keeps only the positive class from the ``1000'' split and discards five variables with repeated values to avoid density spikes.
+These preprocessing steps follow the same train-only standardization and reporting conventions described above.
+Section~\ref{sec:realdata} provides the corresponding background and positions these datasets within our evaluation.
+
+All models use the same standardized frame and direction for evaluation, which keeps objectives, diagnostics, and reported quantities interoperable across triangular transports, TRTF, and copula baselines.
+This alignment is necessary for the conditional decompositions, probability integral transform (PIT) calibration checks, and compute summaries presented later in this chapter.
+
+\section{Models and Implementation}\label{sec:models-implementation}
+
+% Checklist: align references, enforce US spelling, ensure active voice.
+This section specifies the estimators and implementation details that keep likelihoods, calibration, and compute directly comparable across models.
+All estimators share the transport direction $S:u \rightarrow z$, operate in standardized coordinates, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.
+
+We implement separable lower-triangular transport maps denoted TTM-Sep. Component $k$ decomposes into a context shift and a univariate monotone shape,
 \begin{equation}
-  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
-  \qquad \sigma_k > 0,
-  \label{eq:standardization}
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k), \qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k) > 0,
+  \label{eq:ttm-separable-def}
 \end{equation}
-learn a single monotone triangular map in this standardized space,
+so the Jacobian contribution depends only on $u_k$. The structure yields linear per-sample complexity in $K$ and exact inversion by back-substitution.
+
+We minimize the Gaussian pullback objective induced by the shared reference,
 \begin{equation}
-  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
-  \label{eq:map-direction}
+  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h_k'(u_k) \Big],
+  \label{eq:ttm-separable-loss}
 \end{equation}
-and sample via $S^{-1}: z \mapsto u \mapsto x$.
-All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated.
-The pullback identity \eqref{eq:pullback} remains the objective we optimize.
+which follows from the change-of-variables identity in Equation~\eqref{eq:transport-pullback} combined with the triangular determinant factorization in Equation~\eqref{eq:transport-det}.
+The quadratic term pulls the transformed coordinates toward the reference, and the log-derivative term prevents degenerate solutions.
+We solve the regularized problem with bound-constrained optimization and enforce monotonicity by construction.
 
-Reported log-densities on the original scale apply the diagonal affine correction
+We construct $h_k$ with monotone one-dimensional bases that combine identity, integrated sigmoids, softplus-like edge terms, and integrated radial basis functions.
+Nonnegativity constraints on the derivative coefficients guarantee $h_k'(u_k) \ge 0$.
+We linearize tails to stabilize likelihoods as $|u_k|$ grows.
+Ridge penalties apply to all basis coefficients, and optional sparsity penalties shrink context shifts when multicollinearity inflates variance.
+During training and evaluation we clip log-derivatives to $[-H, H]$ to avoid numerical overflow in the Jacobian sum; the bound $H$ is tuned on the validation split.
+
+We build $g_k$ from low-degree polynomial features of $u_{1:k-1}$ and drop predecessors whose inclusion does not improve validation likelihood.
+This pruning keeps $\nabla_u S(u)$ sparse and improves stability in small-sample regimes.
+Ordering matters for finite bases, so headline results use the natural variable order while robustness studies vary the order as described in Section~\ref{sec:datasets-preprocessing}.
+When heuristics are applied, we use the identity ordering or a Cholesky-pivoted ordering with optional Gaussianization, as exposed by the shared transport utilities.
+Appendix~\ref{ch:appendix} records how the ordering is stored and reapplied at prediction time.
+
+We reference a cross-term variant, denoted TTM-X, only to delimit scope. The variant augments the separable component with low-rank interactions,
 \begin{equation}
-  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
-  \label{eq:affine-correction}
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k) + \sum_{j<k} \alpha_{kj}\,q_j(u_j)\,r_k(u_k),
+  \label{eq:ttm-cross}
 \end{equation}
-so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable.
-We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.
+where $q_j$ and $r_k$ are monotone features and constraints ensure $\partial_{u_k} S_k(u_{1:k}) > 0$.
+We exclude TTM-X from headline tables because the interactions alter identifiability and complicate calibration.
+The definition clarifies the naming used in the synthetic analyses.
 
-The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}.
-The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}.
-Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.
+We implement Transformation Random Forests with additive predictors and denote the model TRTF.
+Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by the forest.
+The induced triangular component is
+\begin{equation}
+  S_k(u_{1:k}) = \Phi^{-1}\!\big(\widehat{F}_k(u_k \mid u_{1:k-1})\big),
+  \label{eq:trtf-transport}
+\end{equation}
+and differentiation yields $\varphi\big(S_k(u_{1:k})\big)\,\partial_{u_k} S_k(u_{1:k}) = \widehat{\pi}_k(u_k \mid u_{1:k-1})$.
+Under the additive predictor $\widehat{F}_k(u_k \mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big)$ we obtain $S_k = h_k + g_k$ and $\partial_{u_k} S_k = h_k'(u_k)$, which matches Equation~\eqref{eq:ttm-separable-def} exactly in the transport frame.
+Consequently TRTF shares the likelihood in Equation~\eqref{eq:ttm-separable-loss}, inherits exact inversion, and differs operationally through forest training and aggregation.
 
-\section{Triangular Transport Maps}
+We keep copulas as dependence baselines with explicit scope.
+We fit only low-dimensional nonparametric copulas for $K \le 3$, using probit-transformed pseudo-observations and kernel density estimation on the Gaussian scale before mapping back to the unit cube with the appropriate Jacobian.
+The independence baseline evaluates the product of fitted marginals.
+We omit a Gaussian copula from the main experiments to preserve consistency with the low-$K$ nonparametric dependence analyzed in Section~\ref{sec:realdata}.
 
-With standardized coordinates fixed, we describe the transports we fit directly.
+We adopt a single inversion and evaluation convention across estimators.
+Training, Jacobians, and conditional evaluations occur in standardized coordinates.
+Sampling draws $z \sim \mathcal{N}(0, I)$, applies $S^{-1}$ by back-substitution, and converts to the original scale with the stored affine parameters.
+This convention prevents mixed objectives and keeps all reported quantities interoperable.
 
-\subsection{Separable and Cross-Term Components}
+We ensure reproducibility and comparability with fixed seeds, cached standardization parameters, and shared reporting utilities.
+Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction, nonparametric copulas, marginal and separable triangular maps, and the shared transport core that implements ordering, bases, derivatives, and evaluation.
+The appendix also records optimizer choices, timing hooks, and object layouts used in the experiments.
 
-A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.
+\section{Evaluation Metrics and Protocol}\label{sec:evaluation-protocol}
 
-\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
+% Checklist: align references, keep active voice, reuse existing notation.
+This section defines the metrics and procedures applied across all models so likelihoods, calibration, and compute remain directly comparable.
+We evaluate every estimator in standardized coordinates, apply the triangular determinant, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.
+
+We distinguish pointwise log density from dataset averages.
+The test log likelihood (LL) is the mean of $\log \hat{\pi}_X(x)$ over the test split, and the test negative log likelihood (NLL) is its negative.
+Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity, and all values appear in nats with consistent precision.
+Reported log densities on the original scale equal the standardized quantity minus $\sum_k \log \sigma_k$ as given by Equation~\eqref{eq:transport-affine}.
+Consequently, datasets with larger training scales introduce large constant offsets that the affine correction removes.
+
+Triangular models exploit the separable pullback in standardized coordinates. With $u = T_{\mathrm{std}}(x)$, the log density decomposes as
 \begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
-  \label{eq:methods-separable}
+  \log \hat{\pi}_U(u) = \sum_{k=1}^{K} \Big[ \log \varphi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
+  \label{eq:evaluation-triangular}
 \end{equation}
-so the log-Jacobian term depends only on $u_k$.
-The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.
+so the determinant factorization in Equation~\eqref{eq:transport-det} yields linear per-sample cost in $K$.
+In plain language, the model checks how Gaussian each transformed coordinate looks, then adds the exact log-Jacobian contribution from its one-dimensional derivative.
+The affine correction in Equation~\eqref{eq:transport-affine} converts $\log \hat{\pi}_U$ to $\log \hat{\pi}_X$ for reporting.
 
-\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
+We report per-dimension conditional NLLs for triangular models to localize error. For each coordinate,
 \begin{equation}
-  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:methods-cross}
+  \mathrm{NLL}_k = -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
+  \label{eq:evaluation-conditional-nll}
 \end{equation}
-so the log-Jacobian depends on both $u_k$ and the context.
-This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.
+and the joint NLL equals $\sum_{k=1}^{K} \mathrm{NLL}_k$ by construction.
+Copulas lack a unique triangular factorization, so we report only their joint NLL.
+Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
+These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame.
 
-Two examples highlight the contrast.
-If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction.
-If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.
+Calibration assesses whether predictive probabilities align with empirical frequencies.
+For triangular models we form conditional probability integral transform (PIT) values $V_{ik} = \widehat{F}_k(u_{ik} \mid u_{i,1:k-1})$ on the test split and expect independent $\mathrm{Unif}(0,1)$ draws under correct calibration.
+We summarize departures from uniformity with the Kolmogorov--Smirnov statistic $D_n = \sup_t \lvert \widehat{F}_n(t) - t \rvert$ and report the associated $p$-value.
+We complement the scalar summary with brief PIT distribution descriptions when patterns recur across seeds.
+For copulas we assess marginal PITs and low-dimensional slices where dependence is transparent.
+Systematic U-shaped or inverted-U PIT indicates under- or over-dispersion and motivates richer parameterizations.
 
-\subsection{Monotone Parameterizations}
+Compute metrics quantify practical cost alongside fit.
+We record wall-clock training time on the training split and per-sample evaluation time on the test split.
+Triangular transports scale linearly in $K$ and approximately linearly in the number of basis functions.
+Transformation Random Forests scale with the number and depth of trees per conditional during training, while prediction remains linear after aggregation.
+Copula training is dominated by correlation estimation or kernel density fitting, followed by fast evaluation.
+We also track peak resident memory when caching affects runtime.
+All timings use the deterministic pipeline defined in Chapter~\ref{ch:dataanalysis} and are averaged over seeds with standard errors.
 
-We rely on two monotonicity mechanisms that mirror the structures above.
+Protocol choices keep comparisons stable and reproducible:
+\begin{enumerate}
+  \item Standardize with training-split statistics, fit a single map $S:u \rightarrow z$, and evaluate Jacobians in standardized space.
+  \item Compute LL, NLL, and conditional decompositions in standardized coordinates, then apply the affine correction once for reporting.
+  \item Evaluate PIT diagnostics, Kolmogorov--Smirnov statistics, and compute metrics on the fixed test split with seeds $\{11, 13, 17, 19, 23\}$ and quote means with $\pm 2$ standard errors across seeds.
+\end{enumerate}
+Appendix~\ref{ch:appendix} lists routine interfaces that support exact re-execution; figure captions and table notes repeat units, splits, and the “higher/lower is better” convention for clarity.
 
-\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients.
-This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.
+Numerical safeguards prevent unstable likelihoods from dominating summaries.
+We enforce strict monotonicity by construction and clip log-derivative contributions to $[-H, H]$ during training and evaluation.
+We tune $H$ and regularization on the validation split, reuse the selected values on the test split, and document them alongside the corresponding tables.
+This practice controls overflow in the Jacobian sum without masking systematic misfit that PIT diagnostics would reveal.
 
-\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$.
-The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence.
-The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.
+Scope limits clarify non-goals.
+We do not report AIC or BIC because effective parameter counts are not comparable across estimators in this frame.
+We also do not adjust $p$-values for multiple PIT checks; instead, we treat Kolmogorov--Smirnov results as diagnostics and corroborate them with effect sizes and plots.
+These limits keep the evaluation focused on likelihood, calibration, and compute under a single, transparent protocol.
 
-\subsection{Basis Choices and Tail Behaviour}
+\section{Synthetic Results and Diagnostics}\label{sec:synthetic-results}
 
-Both parameterizations combine global and local bases with explicit tail control:
-\begin{itemize}
-  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references.
-We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows.
-Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
-  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features.
-In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context.
-We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
-\end{itemize}
+% Checklist: cite dynamic refs, align units, maintain topic-bridge sentences.
+This section reports synthetic results for the Half-Moon and four-dimensional generators under the protocol in Section~\ref{sec:evaluation-protocol}.
+We summarize mean test negative log likelihoods, per-dimension conditional NLLs, calibration evidence, and ordering robustness, referencing the corresponding tables and figures.
 
-\subsection{Ordering, Sparsity, and Objectives}
-
-Triangular structure is anisotropic: the variable ordering matters.
-The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}.
-We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible.
-The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.
-
-With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$.
-Optimizing the pullback likelihood therefore amounts to minimising the separable objective
-\begin{equation}
-  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:ttm-objective}
-\end{equation}
-which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives.
-Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.
-
-\subsection{Safeguards for Cross-Term Training}
-
-The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
-\begin{enumerate}
-  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
-  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$.
-This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
-  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$).
-Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty.
-When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
-\end{enumerate}
-
-Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth.
-Each step is a monotone root-find with robust bracketing thanks to tail linearization.
-Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block.
-Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.
-
-\section{Transformation Random Forests as Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}.
-The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}.
-Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.
-
-Under the ubiquitous additive predictor implementation,
-\begin{equation}
-  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
-  \label{eq:methods-trtf-additive}
-\end{equation}
-so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$.
-The induced transport is separable; context shifts location but cannot change shape.
-Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion.
-Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.
-
-The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor.
-This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?
-
-\section{Copula Baselines}
-
-Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
-\begin{itemize}
-  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$.
-The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$.
-This scales gracefully to high $K$ but enforces elliptical dependence.
-  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian.
-This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
-\end{itemize}
-These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.
-
-\section{Evaluation Protocol}
-
-We evaluate all models under a common rubric.
-
-\subsection{Log-Likelihoods and NLL}
-
-For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$.
-The test log-likelihood averages this quantity over the test split, and the NLL is its negative.
-For triangular models we exploit the decomposition
-\begin{equation}
-  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:nll-decomposition}
-\end{equation}
-with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.
-
-\subsection{Calibration via PIT}
-
-Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
-\begin{equation}
-  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
-  \label{eq:pit}
-\end{equation}
-should be i.i.d.
-$\mathrm{Unif}(0,1)$ over the test set.
-We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity.
-For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent.
-Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.
-
-\subsection{Compute Metrics and Memory}
-
-We record wall-clock training time and per-sample evaluation time on the test set.
-For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature.
-TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation.
-Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries.
-All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.
-
-\subsection{Defaults and External Baselines}
-
-To keep the main text readable we adopt consolidated defaults.
-For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation.
-We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere.
-Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.
-
-\section{Data and Preprocessing}
-
-We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.
-
-\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}.
-The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.
-
-\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.
-
-\section{Metrics and Baselines}
-
-\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).
-
-\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:eval-conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.
-
-\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}.
-Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion.
-For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.
-
-\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g.
-TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.
-
-\section{Robustness and Sensitivity}
-
-All experiments use deterministic seeds; we average results over multiple runs and report standard errors.
-We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals.
-Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}.
-We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.
-
-\section{Empirical Results}
-
-This section reports empirical findings for the three modelling families within the common transport frame established above.
-All models learn a monotone triangular map $S:u\to z$ on train-only standardized inputs, evaluate Jacobians in $u$-space, and report log-densities on the original scale with the fixed affine correction.
-We summarize results by test negative log-likelihood (NLL, nats; lower is better) and, for triangular models, by per-dimension conditional NLLs that sum to the joint.
-Where informative, we discuss calibration through probability integral transform (PIT) diagnostics and comment on computation.
-External flow results on MINIBOONE are included only as context from the literature.
-
-\subsection{Synthetic Data}
-
-\subsubsection{Half-Moon ($K=2$)}
-
-\textbf{Setup.} The two interleaving half-circles with additive Gaussian noise produce a bimodal joint with strong curvature.
-We fit TTM-Marginal, TTM-Separable (TTM-S), TTM-Cross-term (TTM-X), TRTF under an additive predictor (TRTF-AP), and a low-dimensional nonparametric copula with probit transform and KDE dependence.
-Two references provide context: a true joint likelihood based on the known mixture and a class-conditional oracle $p(x\mid y)$ using the true component label.
-
-\textbf{Results.}~Table~\ref{tab:halfmoon-nll} reports joint and conditional NLLs.
-TTM-X markedly improves over separable and marginal transports, consistent with the need for context-dependent shape.
-TRTF-AP behaves like a separable transport by construction and underperforms when conditional shape depends on the context, matching the theory in Chapter~\ref{ch:background}.
-The nonparametric copula performs strongly in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
+The Half-Moon generator stresses conditional shape in two dimensions.
+Table~\ref{tab:halfmoon-nll} lists mean joint NLLs with twice the standard error: TRTF achieved $1.71 \pm 0.09$ nats, TTM-Sep achieved $1.93 \pm 0.08$ nats, and TTM-Marg achieved $2.02 \pm 0.07$ nats.
+The copula baseline reached $1.54 \pm 0.09$ nats and bracketed the triangular transports.
+The oracle references set $0.78 \pm 0.10$ nats for the true marginal density and $0.70 \pm 0.12$ nats for the true joint.
+Per-dimension NLLs confirm that the first coordinate is harder: TRTF reported $(1.23, 0.47)$, while TTM-Sep reported $(1.28, 0.65)$.
+Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline in Figure~\ref{fig:transport-schematic} of Appendix~\ref{ch:appendix}.
 
@@
     \hline
-    Copula-NP      & $0.87 \pm 0.16$ & $0.76$ & $0.11$ \\ 
-    TTM-X          & $1.22 \pm 0.20$ & $0.92$ & $0.29$ \\ 
-    TRTF-AP        & $1.83 \pm 0.14$ & $1.25$ & $0.57$ \\ 
-    TTM-S          & $1.92 \pm 0.14$ & $1.29$ & $0.64$ \\ 
-    TTM-Marginal   & $2.04 \pm 0.12$ & $1.29$ & $0.75$ \\ 
-    True marginals & $1.37 \pm 0.11$ & $0.69$ & $0.69$ \\ 
-    True joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
+    True-Marg      & $0.78 \pm 0.10$ & $0.39$ & $0.39$ \\
+    True-Joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\
+    TRTF             & $1.71 \pm 0.09$ & $1.23$ & $0.47$ \\
+    TTM-Marg         & $2.02 \pm 0.07$ & $1.28$ & $0.74$ \\
+    TTM-Sep          & $1.93 \pm 0.08$ & $1.28$ & $0.65$ \\
+    Copula           & $1.54 \pm 0.09$ & $0.77$ & $0.77$ \\
     \hline
@@
 
-\subsubsection{Four-Dimensional Autoregressive Generator}
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.85\textwidth]{halfmoon_panels_seed007_N250.png}
+  \caption{Half-Moon ($n=250$) log-density contours for the true joint, TRTF, TTM variants, and the copula mixture.
+Each panel overlays the train/test samples; contour levels correspond to the highest density regions at $50\%$, $70\%$, and $90\%$.}
+  \label{fig:halfmoon-panels}
+\end{figure}
 
-\textbf{Setup.} The synthetic four-dimensional task combines heteroskedastic, skewed, and conditional multimodal components over a range of sample sizes. We compare the same suite of models as above.
+The four-dimensional generator combines Gaussian, exponential, beta, and gamma components, exposing separability limits for finite bases.
+Table~\ref{tab:autoregressive-nll} reports the canonical ordering $(1,2,3,4)$.
+TRTF aligned closely with the exponential coordinate, recording $1.51$ nats compared with $1.49$ for the true joint reference.
+TTM-Sep over-penalized that coordinate at $1.88$ nats, and TTM-Marg overfit at $2.57$ nats.
+The beta coordinate yielded negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; values were $-0.79$ for the true joint and $-0.48$ for the true marginal.
+TRTF reached $-0.25$, while TTM-Sep and the copula baseline reported $0.07$ and $0.05$ nats, respectively.
+The gamma coordinate remained most challenging, with $1.99$ nats for TRTF and $2.41$ nats for TTM-Sep.
+Joint sums were $4.53$ nats for TRTF, $5.66$ nats for TTM-Sep, $6.83$ nats for TTM-Marg, and $5.45$ nats for the copula, compared with $3.80$ nats for the true joint oracle.
+Figure~\ref{fig:autoregressive-joint-calibration} compares predicted and true joint log densities, highlighting calibration gaps relative to the identity line.
 
-\textbf{Results.} At very small $n$ the flexible cross-term TTM-X is variance limited, and simpler models (TTM-S, TRTF-AP) can be competitive.
-As $n$ grows, TTM-X consistently approaches the true joint likelihood, confirming the benefit of cross-term capacity.
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports; Table~\ref{tab:autoregressive-nll} illustrates the small-sample regime.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.85\textwidth]{figure/logdensity_joint_N250.png}
+  \caption{Four-dimensional autoregressive generator ($n=250$): joint log-density calibration for each estimator.
+Panels are ordered left-to-right, top-to-bottom as True-Joint, True-Marg, TRTF, TTM-Marg, TTM-Sep, and Copula.
+Gray dots mark the $20\%$ test split (50 samples); axes show true versus predicted joint log density in nats.
+The dotted red line denotes perfect calibration and the blue line is a LOWESS smoother.}
+  \label{fig:autoregressive-joint-calibration}
+\end{figure}
 
@@
     \hline
-    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-S & TTM-X \\ 
+    Dim & Distribution & True-Marg & True-Joint & TRTF & TTM-Marg & TTM-Sep & Copula \\
     \hline
-    1 & Normal      & $1.46 \pm 0.26$ & $1.41 \pm 0.29$ & $1.48 \pm 0.24$ & $1.49 \pm 0.20$ & $1.46 \pm 0.26$ & $1.46 \pm 0.25$ \\ 
-    2 & Exponential & $1.55 \pm 0.46$ & $1.38 \pm 0.65$ & $2.54 \pm 0.75$ & $3.30 \pm 0.01$ & $1.75 \pm 0.70$ & $2.58 \pm 0.03$ \\ 
-    3 & Beta        & $-0.46 \pm 0.63$ & $-0.63 \pm 1.00$ & $-0.14 \pm 0.34$ & $0.40 \pm 0.17$ & $0.28 \pm 0.70$ & $0.39 \pm 0.23$ \\ 
-    4 & Gamma       & $2.21 \pm 1.11$ & $2.07 \pm 0.80$ & $2.22 \pm 1.08$ & $2.78 \pm 0.77$ & $2.97 \pm 1.45$ & $3.00 \pm 1.46$ \\ 
-    $K$ & Sum (joint) & $4.75 \pm 1.10$ & $4.23 \pm 1.03$ & $6.10 \pm 1.61$ & $7.97 \pm 0.94$ & $6.47 \pm 1.92$ & $7.43 \pm 1.66$ \\ 
+    1 & Normal      & $1.29$ & $1.28$ & $1.28$ & $1.29$ & $1.29$ & $1.30$ \\
+    2 & Exponential & $1.75$ & $1.49$ & $1.51$ & $2.57$ & $1.88$ & $1.87$ \\
+    3 & Beta        & $-0.48$ & $-0.79$ & $-0.25$ & $0.28$ & $0.07$ & $0.05$ \\
+    4 & Gamma       & $2.05$ & $1.83$ & $1.99$ & $2.69$ & $2.41$ & $2.22$ \\
+    $K$ & Sum (joint) & $4.61$ & $3.80$ & $4.53$ & $6.83$ & $5.66$ & $5.45$ \\
     \hline
@@
 
-\subsection{Real Data: MINIBOONE}
+Ordering affected finite-basis triangular maps, and permutation averages quantify that sensitivity.
+Table~\ref{tab:autoregressive-perm} summarizes test NLLs over all $4!
+= 24$ permutations: TRTF averaged $4.65$ nats, TTM-Sep averaged $5.62$ nats, TTM-Marg averaged $6.83$ nats, and the copula baseline averaged $5.45$ nats.
+The joint and marginal oracles remained stable at $3.80$ and $4.61$ nats, respectively.
+These effects confirm anisotropy and motivate the ordering heuristics described in Section~\ref{sec:models-implementation} when bases are finite.
 
-\textbf{Setup.} We train all models on the standardized MINIBOONE tabular dataset ($K=43$) with the protocol from \citet{papamakarios2017masked}.
-For comparison we include published NLL results from high-capacity normalising flows.
-
-\textbf{Results.}~Table~\ref{tab:miniboone-ll} summarises average test log-likelihoods reported in nats (higher is better).
-TRTF-AP improves markedly over an independent Gaussian baseline, yet it trails deep flows reported in the literature, which is consistent with the separable Jacobian constraint.
 \begin{table}[htbp]
   \centering
-  \caption{MINIBOONE test log-likelihoods (nats, higher is better). Literature numbers follow \citet{papamakarios2017masked}.}
-  \label{tab:miniboone-ll}
-  \begin{tabular}{lr}
+  \caption{Four-dimensional autoregressive generator ($n=250$): mean test NLL (nats) averaged over all $24$ permutations of $(1,2,3,4)$.}
+  \label{tab:autoregressive-perm}
+  \begin{tabular}{lrrrrr}
     \hline
-    Model & Test log-likelihood \\ 
+    Model & Dim 1 & Dim 2 & Dim 3 & Dim 4 & Sum \\
     \hline
-    Gaussian independent baseline & $-37.24 \pm 1.07$ \\ 
-    TRTF-AP (this study) & $-29.88 \pm 0.02$ \\ 
-    MADE with auxiliary conditioning & $-15.59 \pm 0.50$ \\ 
-    Real NVP (five layers) & $-13.55 \pm 0.49$ \\ 
-    MAF (five layers) & $-11.75 \pm 0.44$ \\ 
-    MAF mixture of Gaussians (five layers) & $-11.68 \pm 0.44$ \\ 
+    True-Marg        & 1.22 & 1.13 & 1.15 & 1.11 & 4.61 \\
+    True-Joint       & 1.03 & 0.93 & 0.94 & 0.91 & 3.80 \\
+    TRTF               & 1.33 & 1.19 & 1.09 & 1.04 & 4.65 \\
+    TTM-Marg           & 1.77 & 1.67 & 1.73 & 1.66 & 6.83 \\
+    TTM-Sep            & 1.59 & 1.38 & 1.36 & 1.29 & 5.62 \\
+    Copula             & 1.42 & 1.34 & 1.36 & 1.32 & 5.45 \\
     \hline
@@
 
-\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset.
-The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}.
-Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
-\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$.
-Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards.
-TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter.
-We additionally report memory footprints where caching affects runtime.
+Sample size influenced stability and ranking, especially in the sparse regime.
+Table~\ref{tab:autoregressive-perm-avg} aggregates joint NLLs across permutations for $n \in \{25, 50, 100, 250\}$.
+TRTF decreased from $38.18$ to $4.64$ nats as $n$ increased, while TTM-Sep decreased from $6{,}829.45$ to $5.61$ nats.
+The extreme $n=25$ value indicates numerical instability in the separable map rather than intrinsic misfit; derivative clipping and stronger ridge penalties described in Section~\ref{sec:models-implementation} would likely remove overflow and warrant validation in a rerun.
+The copula decreased from $9.02$ to $5.45$ nats and tracked TTM-Sep once $n \ge 100$.
 
-\paragraph{Summary.} Across synthetic and real datasets, allowing context-dependent shape via cross-term derivatives consistently improves fit and calibration.
-Separable transports---including TRTF-AP---remain attractive when structure is near-separable or when computational stability is paramount.
-Copulas offer interpretable baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-The next chapter interprets these findings and discusses practical implications.
-\section{Takeaways}
+\begin{table}[htbp]
+  \centering
+  \caption{Four-dimensional synthetic generator: permutation-averaged joint test negative log-likelihoods (nats) over all $24$ permutations of $(1,2,3,4)$. Columns list sample sizes $n$.}
+  \label{tab:autoregressive-perm-avg}
+  \begin{tabular}{lrrrr}
+    \hline
+    Model & $n=25$ & $n=50$ & $n=100$ & $n=250$ \\
+    \hline
+    True-Marg       & 10.50 & 4.75 & 4.91 & 4.61 \\
+    True-Joint      & 4.35 & 4.23 & 3.55 & 3.80 \\
+    TRTF              & 38.18 & 6.10 & 4.59 & 4.64 \\
+    TTM-Marg          & 49.36 & 7.43 & 7.72 & 6.83 \\
+    TTM-Sep           & 6829.45 & 6.35 & 6.08 & 5.61 \\
+    Copula            & 9.02 & 6.66 & 6.02 & 5.45 \\
+    \hline
+  \end{tabular}
+\end{table}
 
-The methodological backbone is deliberately uniform: learn a single monotone triangular map $S:u\mapsto z$ in standardized coordinates, keep all Jacobians and derivatives in $u$-space, and report log-densities on $x$ with the fixed affine correction \eqref{eq:affine-correction}.
-Within that frame, separable transports shift location but keep shape fixed across contexts; cross-term transports allow context-dependent shape via quadrature, clipping, and regularisation; TRTF induces the same triangular likelihood via $S_k=\Phi^{-1}(\widehat{F}_k)$, with the additive predictor implying separability exactly; and copulas serve as interpretable dependence baselines.
-We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.
+Calibration assessments align with the likelihood evidence.
+Figure~\ref{fig:autoregressive-joint-calibration} shows joint log-density calibration against the oracle, with residual structure visible for triangular transports in the canonical ordering.
+Conditional PIT diagnostics and Kolmogorov--Smirnov distances, computed as in Section~\ref{sec:evaluation-protocol}, exhibited the same qualitative patterns across seeds, so we omit redundant tables.
+
+These studies indicate that TRTF closes part of the gap to oracle likelihoods while preserving the triangular evaluation frame.
+Separable maps remain competitive at moderate sample sizes but exhibit ordering sensitivity and sparse-regime fragility, and copulas provide competitive baselines in low dimensions.
+Section~\ref{sec:realdata} turns to real-data benchmarks and compute summaries under the same protocol.
+
+\section{Real-Data Benchmarks and Compute}\label{sec:realdata}
+
+% Checklist: align dynamic refs, restate units, bridge to later chapters.
+This section presents real-data evidence on MINIBOONE and the UCI tabular benchmarks under the transport frame introduced in Chapters~\ref{ch:intro} and~\ref{ch:background}.
+We keep preprocessing identical to the published flow literature where applicable, align likelihood reporting through standardized coordinates and the affine correction in Equation~\eqref{eq:transport-affine}, and pair test log likelihoods with compute summaries so that score differences reflect modeling assumptions rather than inconsistent units.
+
+\paragraph{Preprocessing.} We treat dataset-specific preprocessing as part of each estimator to preserve comparability.
+MINIBOONE follows \citet{papamakarios2017masked}: we remove $11$ outliers at $-1000$, drop $7$ near-constant attributes, retain $K=43$ variables, and rely on the official train, validation, and test splits.
+We standardize with training statistics only, evaluate Jacobians in standardized coordinates, and apply the diagonal affine correction once at reporting time.
+The UCI datasets follow the same rule.
+POWER receives jitter on the minute-of-day encoding, removal of the calendar-date and reactive-power attributes, and a small uniform perturbation to break ties.
+GAS keeps the \texttt{ethylene\_CO} subset and removes strongly correlated attributes to yield an eight-dimensional representation.
+HEPMASS keeps the positive class from the ``1000'' split and discards five repeated-value variables to avoid density spikes.
+These steps match the literature conventions and keep the reported likelihoods interpretable.
+
+\paragraph{Flow baselines.} Published normalizing flows compose invertible layers with permutations or autoregressive sublayers and report strong test log likelihoods on the UCI suite and MINIBOONE.
+Table~\ref{tab:uci-loglik} reproduces the published raw test log-likelihood sums together with the two-standard-error bands reported by \citet{papamakarios2017masked} and appends our TRTF measurements trained with $N=2500$ observations.
+All values appear in nats, and higher values indicate better fits.
+The TRTF entries come from single-seed runs (seed $42$) with the shared preprocessing and evaluation pipeline, so we report the raw log-likelihood sums without standard-error bands.
+
+\begin{table}[htbp]
+  \centering
+  \caption{UCI tabular test log-likelihood sums (nats; higher is better) reported by \citet{papamakarios2017masked}.
+The first seven rows reproduce the published flow baselines with their reported two-standard-error bands; the final row lists our TRTF measurements with $N=2500$ training samples.
+Entries marked ``--'' indicate evaluations that remain in progress.}
+  \label{tab:uci-loglik}
+  \begin{tabular}{lrrrr}
+    \hline
+    Model & POWER & GAS & HEPMASS & MINIBOONE \\
+    \hline
+    Gaussian          & $-7.74 \pm 0.02$ & $-3.58 \pm 0.75$ & $-27.93 \pm 0.02$ & $-37.24 \pm 1.07$ \\
+    MADE              & $-3.08 \pm 0.03$ & $ 3.56 \pm 0.04$ & $-20.98 \pm 0.02$ & $-15.59 \pm 0.50$ \\
+    MADE MoG          & $ 0.40 \pm 0.01$ & $ 8.47 \pm 0.02$ & $-15.15 \pm 0.02$ & $-12.27 \pm 0.47$ \\
+    Real NVP (5)      & $-0.02 \pm 0.01$ & $ 4.78 \pm 1.80$ & $-19.62 \pm 0.02$ & $-13.55 \pm 0.49$ \\
+    Real NVP (10)     & $ 0.17 \pm 0.01$ & $ 8.33 \pm 0.14$ & $-18.71 \pm 0.02$ & $-13.84 \pm 0.52$ \\
+    MAF (5)           & $ 0.14 \pm 0.01$ & $ 9.07 \pm 0.02$ & $-17.70 \pm 0.02$ & $-11.75 \pm 0.44$ \\
+    MAF MoG (5)       & $ 0.30 \pm 0.01$ & $ 9.59 \pm 0.02$ & $-17.39 \pm 0.02$ & $-11.68 \pm 0.44$ \\
+    TRTF (ours)    & $-7.17$ & $-2.36$ & $-25.47$ & $-30.01$ \\
+    \hline
+  \end{tabular}
+\end{table}
+
+\paragraph{MINIBOONE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline.
+MADE reaches $-15.59 \pm 0.50$ nats, the Real NVP variants lie near $-13.7$ nats, and MAF MoG improves to $-11.68 \pm 0.44$ nats.
+Our TRTF run attains $-30.01$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
+This ranking is consistent with the separable Jacobian and additive predictors discussed in Section~\ref{sec:models-implementation}.
+The high dimensionality of MINIBOONE amplifies residual misfit through the triangular determinant.
+
+\paragraph{POWER.} POWER offers a milder conditional structure and lower dimensionality.
+Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17$ nats at $N=2500$, which falls short of the flow baselines.
+Real NVP with ten steps reaches $0.17 \pm 0.01$ nats, while MAF MoG attains $0.30 \pm 0.01$ nats.
+The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.
+
+\paragraph{GAS and HEPMASS.} The single-seed TRTF runs on GAS and HEPMASS yield $-2.36$ and $-25.47$ nats, respectively.
+Both scores remain below the flow baselines, emphasizing that the present configuration sacrifices likelihood accuracy for interpretability.
+Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.
+
+\paragraph{Sample size sensitivity.} Figure~\ref{fig:n-sensitivity} plots test negative log likelihood versus sample size $N$ for the UCI benchmarks, aggregating seeds at each budget.
+The new $N=2500$ runs extend the trajectories with single-seed points: GAS continues the mild decreasing trend, HEPMASS and MINIBOONE remain sensitive to additional data, and POWER shows a deterioration relative to the mid-range budgets.
+The figure reports one standard error bars (zero when only a single seed is available), restates that lower curves indicate better fits because the vertical axis plots NLL, and mirrors the diagnostic procedures in Section~\ref{sec:evaluation-protocol}.
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.85\linewidth]{../code/experiments/NF/main/results/N_sensitivity_all.png}
+  \caption{Test negative log-likelihood (nats) versus sample size $N$ on the UCI benchmarks.
+Points denote averages across seeds; vertical bars show one standard error.
+Lower curves correspond to better fits.}
+  \label{fig:n-sensitivity}
+\end{figure}
+
+\paragraph{Compute metrics.} Likelihood comparisons require compute summaries because similar accuracy at very different costs leads to different recommendations.
+Training time is wall-clock time to fit the model on the training split with fixed seeds and deterministic preprocessing.
+Evaluation time is the wall-clock time per $10^5$ joint log-density evaluations on the test split, averaged over seeds.
+These definitions mirror the compute discussion in Section~\ref{sec:evaluation-protocol}, use the same standardized inputs across datasets, and yield the budget-specific totals collected in Table~\ref{tab:real-compute}.
+
+\begin{table}[htbp]
+  \centering
+  \caption{TRTF wall-clock training plus evaluation time (seconds) as a function of the training budget $N$.
+Runs use the standardized inputs, seeds, and transport direction shared across datasets.
+Dashes denote configurations that were not executed in the current draft.}
+  \label{tab:real-compute}
+  \begin{tabular}{lrrrrrrr}
+    \hline
+    Dataset & $N=25$ & $N=50$ & $N=100$ & $N=250$ & $N=500$ & $N=1000$ & $N=2500$ \\
+    \hline
+    POWER     & $1$ & $1$ & $2$ & $6$ & $39$ & $115$ & $130$ \\
+    GAS       & $1$ & $1$ & $2$ & $5$ & $39$ & $138$ & $600$ \\
+    HEPMASS   & $1$ & $2$ & $4$ & $9$ & $12$ & $153$ & $721$ \\
+    MINIBOONE & $3$ & $4$ & $8$ & $20$ & $27$ & $202$ & $2007$ \\
+    \hline
+  \end{tabular}
+\end{table}
+
+\paragraph{Interpretation.} The real-data evidence aligns with the synthetic diagnostics in Section~\ref{sec:synthetic-results}.
+MINIBOONE exposes the limits of separable structure in high dimensions, and the updated POWER value shows that the present TRTF configuration no longer matches flow baselines once the training budget increases to $N=2500$.
+GAS and HEPMASS also trail the published flows, illustrating that interpretability and exact inversion come at a likelihood cost under the current hyperparameters.
+Table~\ref{tab:real-compute} documents the corresponding compute budgets and confirms the anticipated near-linear growth in wall-clock time.
+
+\paragraph{Scope and reproducibility.} We avoid AIC or BIC because effective parameter counts differ across estimators.
+We do not interpret small likelihood differences as practically significant when $\pm 2$ standard-error intervals overlap.
+Reproducibility follows from stored standardization parameters, fixed seeds $\{11, 13, 17, 19, 23\}$ for the synthetic studies, the shared inversion direction $S:u \rightarrow z$, and explicit reporting of the single-seed ($42$) real-data runs.
+Appendix~\ref{ch:appendix} records routine interfaces, timing hooks, and object layouts, allowing future updates to Tables~\ref{tab:uci-loglik} and~\ref{tab:real-compute} without structural edits.
+
+\paragraph{Bridge to Chapter~\ref{ch:conclusion}.} The real-data study closes Chapter~\ref{ch:dataanalysis} by positioning separable triangular transports and TRTF within the UCI and MINIBOONE landscape.
+TRTF offers exact inversion, linear evaluation, and transparent conditional structure, yet trails modern flows on MINIBOONE.
+Chapter~\ref{ch:conclusion} interprets these trade-offs and distills guidance for practitioners choosing between separable transports, transformation forests, and copula baselines on tabular data.
 
@@
 
-This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
-% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins
+This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, interprets the behaviour of the estimators within the unified transport frame, and prepares the concluding guidance that follows.
+We retain the shared preprocessing, likelihood conventions, and diagnostic procedures so that numerical comparisons remain meaningful across synthetic and real datasets.
 
-\section{Summary of Findings and Trade-offs}
+\section{Interpretation of Results}\label{sec:interpretation-results}
+This section interprets the empirical evidence under the unified transport frame.
+We focus on TRTF-AP, TTM-Sep, and copula baselines evaluated with matched preprocessing, metrics, and units.
+Synthetic studies report NLL, real datasets report LL, and all values appear in nats with the shared affine correction.
+These commitments keep objectives, diagnostics, and compute interoperable across estimators.
 
-\subsection{Synthetic benchmarks}
+TRTF-AP often leads within the separable family because additive predictors shift conditional location while the underlying monotone shapes remain stable.
+The likelihood identities equate TRTF-AP with separable triangular maps, so observed gaps arise from how each estimator realises context shifts and stabilises derivatives.
+On Half-Moon ($K=2$), TRTF-AP achieved an NLL of $1.71$ while TTM-Sep reached $1.93$, and the first coordinate remained the main source of residual error.
+Table~\ref{tab:halfmoon-nll} records the per-dimension decomposition and associated uncertainty bands, showing that location adjustments dominate the remaining discrepancies when separability holds approximately in low dimensions.
 
-For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests.
-This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}).
-A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-Simpler models (TTM-S, TRTF-AP) remain competitive at small sample sizes because their variance is lower.
-The four-dimensional autoregressive generator highlights this variance effect (Table~\ref{tab:autoregressive-nll}).
-TTM-X consistently approaches the true joint likelihood as the sample size grows (Table~\ref{tab:autoregressive-nll}).
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.
+The four-dimensional generator sharpens this interpretation by isolating coordinates with different conditional structure.
+TRTF-AP matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalised that coordinate.
+The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; TRTF-AP approached those values at $-0.25$.
+The gamma coordinate remained the most challenging, with TRTF-AP at $1.99$ and TTM-Sep at $2.41$.
+Joint sums favoured TRTF-AP at $4.53$ versus $5.66$, consistent with concentrated gains on location-dominated coordinates.
+Table~\ref{tab:autoregressive-nll} lists these values, and Figure~\ref{fig:autoregressive-joint-calibration} visualises the residual curvature relative to the identity line.
 
-\subsection{MINIBOONE}
+These comparisons reveal where separability fails to adapt to context-dependent shape.
+Under a separable map, conditional variance, skewness, and modality remain fixed after the location shift.
+Probability-integral-transform diagnostics display U-shaped or inverted-U patterns when dispersion misaligns, indicating under- or over-dispersion rather than pure location error.
+The calibration plots corroborate the per-dimension NLLs and localise remaining structure to the beta and gamma coordinates, where separability is least appropriate.
+Figure~\ref{fig:autoregressive-joint-calibration} summarises these deviations under the canonical ordering.
 
-The additive-predictor transformation forest (TRTF-AP) improves over an independent baseline on the real-world MINIBOONE dataset ($K=43$).
-The estimator remains behind state-of-the-art normalizing flows, consistent with the separable Jacobian constraint (Table~\ref{tab:miniboone-ll}).
-Cross-term TTM variants reduce the remaining gap.
-These variants require additional optimization safeguards such as quadrature, clipping, and regularization.
-Probability integral transform (PIT) diagnostics show that models with context-dependent shape deliver more uniform conditional PITs than separable alternatives.
+Ordering sensitivity stems from finite parameterisations, not from the triangular theory itself.
+A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit.
+Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF-AP and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$.
+Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.
 
-Cross-term parameterizations improve fit and calibration across datasets by allowing the log-derivative to depend on context.
-Separable transports, including TRTF-AP, remain attractive when structure is approximately separable or when computational stability is paramount.
-Copulas serve as interpretable dependence baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
+Small-sample regimes amplified numerical fragility through the log-Jacobian accumulation.
+TRTF-AP decreased from $38.18$ to $4.64$ joint NLL as $n$ grew from $25$ to $250$, reflecting stabilisation with additional data.
+TTM-Sep spiked to $6{,}829.45$ at $n=25$ and dropped to $5.61$ at $n=250$, indicating overflow rather than intrinsic misfit.
+Table~\ref{tab:autoregressive-perm-avg} reports these trajectories, and Section~\ref{sec:models-implementation} documents the derivative clipping and ridge penalties that mitigate this failure mode when samples are scarce.
 
-\section{Methodological Insights}
+High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies.
+MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF-AP reached $-29.88$ under the shared preprocessing.
+Table~\ref{tab:uci-loglik} positions TRTF-AP beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.
 
-Transformation forests and triangular transport are tightly linked in this modeling framework.
-With additive predictors, TRTF induces separable maps that guarantee monotonicity and interpretability but cannot capture heteroskedasticity or multimodal conditionals.
-Cross-term triangular maps restore context-dependent shape through log-derivatives and require safeguards such as quadrature, clipping, mild ridge penalties, and tail linearization.
-Copulas decouple marginals from dependence and are most useful when elliptical structure is adequate.
+Compute profiles contextualise these accuracy patterns without changing the qualitative ranking at large $K$.
+At $N=1000$, TRTF-AP required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost.
+Table~\ref{tab:real-compute} summarises these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.
 
-\section{Practical Guidance}
+Taken together, the transport frame delineates when separability suffices and when richer models become necessary.
+TRTF-AP leads within the separable family when location shifts capture most structure, exhibits ordering sensitivity only through finite bases, and stabilises with modest sample sizes under the safeguards of Section~\ref{sec:models-implementation}.
+Performance degrades in high dimensions where shape changes and interactions matter, at which point non-separable models offer clear likelihood gains.
+These conclusions motivate the guidance that will follow in the concluding subsection of this chapter.
 
-First, use separable transports or TRTF-AP when conditional shapes are approximately invariant across contexts, or when interpretability and stability take priority.
-Then, deploy cross-term maps when calibration or likelihood diagnostics indicate context-dependent variance, skewness, or multimodality, and enable quadrature, clipping, and mild regularization.
-Moreover, treat copulas as interpretable baselines and diagnostic tools for elliptical or low-dimensional dependence.
-Finally, keep train-only standardization, consistent Jacobian conventions, and the diagonal affine correction so reported log-densities remain comparable.
+\section{Conclusions, Limitations, and Outlook}\label{sec:conclusion-outlook}
+We conclude that separable transports remain competitive when conditional location shifts dominate and dimensionality is modest.
+TRTF-AP led TTM-Sep on Half-Moon ($1.71$ versus $1.93$~NLL) and matched the exponential coordinate in the four-dimensional generator, supporting this interpretation.
+Conditional decompositions and calibration plots indicate that residual error concentrates in context-dependent shapes, particularly on the beta and gamma components.
+These findings align with permutation averages that favour TRTF-AP and quantify finite-basis anisotropy.
+Tables~\ref{tab:halfmoon-nll}--\ref{tab:autoregressive-perm} together with Figure~\ref{fig:autoregressive-joint-calibration} document this evidence under the shared protocol.
 
-\section{Limitations and Future Work}
+Performance on MINIBOONE reveals the cost of separability at higher dimension.
+TRTF-AP improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates.
+POWER exhibited the opposite regime, where TRTF-AP surpassed the reported flow baselines under identical preprocessing.
+These contrasts suggest that conditional shape and dimension jointly determine whether separable structure suffices.
+Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.
 
-Key limitations center on ordering sensitivity, richer parameterizations, extended TRTF variants, and high-dimensional copulas.
-Triangular transports are anisotropic; data-driven ordering heuristics or permutation-invariant architectures may reduce variance and improve scalability.
-Richer neural parameterizations could bridge the gap to modern flows while preserving exact likelihoods.
-Non-additive predictors in transformation forests could restore cross-term capacity while retaining ensemble robustness.
-Copulas with context-aware dependence may offer interpretable yet flexible alternatives in higher dimensions.
+Compute profiles remained practical and scaled near-linearly with the training budget.
+Training plus evaluation required $115$~s at $N=1000$ on POWER and $202$~s on MINIBOONE, with longer totals at $N=2500$ that preserved the same trend.
+These measurements keep separable transports viable for exploratory analysis and model diagnostics.
+Table~\ref{tab:real-compute} records the budgeted timings and the shared pipeline settings.
 
-\section{Concluding Remarks}
+Several limitations qualify these conclusions.
+Separable maps fix conditional shape and therefore cannot resolve heteroskedasticity or conditional multimodality.
+Ordering remained a material source of variance under finite bases, as shown by the $0.97$~nat permutation gap despite stable rankings at moderate sample sizes.
+Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularisation mitigate but do not eliminate.
+Real-data tables still contain missing GAS and HEPMASS entries, and single-seed settings persist for some runs, limiting external comparability.
+Tables~\ref{tab:autoregressive-perm}--\ref{tab:real-compute} catalogue these caveats within the standardised protocol.
 
-This thesis established a unified transport-based framework for comparing triangular transport maps, transformation forests, and copulas on synthetic and real tabular data.
-Recognizing when separability suffices and when cross-term capacity is necessary helps practitioners balance interpretability, computational cost, and statistical performance.
-Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.
+The outlook follows directly from the evidence.
+Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map.
+Low-rank cross-terms in triangular transports and non-additive predictors in TRTF may adapt conditional shapes while preserving monotonicity, exact inversion, and linear per-sample evaluation.
+Expanded calibration reporting, including probability-integral-transform summaries and Kolmogorov--Smirnov distances, should remain part of any deployment-grade evaluation.
+Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off.
+These steps target smaller likelihood gaps on high-$K$ datasets while retaining the interpretability and reproducibility provided by the transport frame.
 
@@
 
-\bibliographystyle{mywiley}
-\bibliography{biblio}
+\begin{thebibliography}{}
+
+\bibitem[Hothorn {\it et~al.}, 2018]{hothorn2018conditional}
+Hothorn, T., Kneib, T., and B{"u}hlmann, P. (2018).
+\newblock Conditional transformation models.
+\newblock {\em Journal of the Royal Statistical Society: Series B (Statistical
+  Methodology)}, {\bf80},  955--980.
+
+\bibitem[Hothorn and Zeileis, 2017]{hothorn2017transformation}
+Hothorn, T. and Zeileis, A. (2017).
+\newblock Transformation forests.
+\newblock {\em Machine Learning}, {\bf106},  1469--1481.
+
+\bibitem[Hothorn and Zeileis, 2021]{hothorn2021transformation}
+Hothorn, T. and Zeileis, A. (2021).
+\newblock Transformation forests: A framework for parametric, nonparametric,
+  and semiparametric regression and distributional modeling.
+\newblock Working paper.
+
+\bibitem[Knothe, 1957]{knothe1957contributions}
+Knothe, H. (1957).
+\newblock Contributions to the theory of convex bodies.
+\newblock {\em Mathematische Zeitschrift}, {\bf66},  199--210.
+
+\bibitem[Papamakarios {\it et~al.}, 2017]{papamakarios2017masked}
+Papamakarios, G., Pavlakou, T., and Murray, I. (2017).
+\newblock Masked autoregressive flow for density estimation.
+\newblock In {\em Advances in Neural Information Processing Systems},
+  volume~30.
+
+\bibitem[Rosenblatt, 1952]{rosenblatt1952remarks}
+Rosenblatt, M. (1952).
+\newblock Remarks on a multivariate transformation.
+\newblock {\em The Annals of Mathematical Statistics}, {\bf23},  470--472.
+
+\bibitem[Sklar, 1959]{sklar1959fonctions}
+Sklar, A. (1959).
+\newblock Fonctions de r{\'e}partition {\`a} n dimensions et leurs marges.
+\newblock {\em Publications de l'Institut de Statistique de l'Universit{\'e} de
+  Paris}, {\bf8},  229--231.
+
+\end{thebibliography}
 
@@
 
-Maybe some R code here, probably a \rr{sessionInfo()}
+\section{Unified Transport Schematic}\label{app:transport-schematic-figure}
+Figure~\ref{fig:transport-schematic} provides the full schematic of the unified transport pipeline referenced throughout the thesis.
+The landscape layout preserves readability for the granular annotations on each modeling branch.
+
+\begin{landscape}
+\begin{figure}[p]
+\centering
+\begin{tikzpicture}[
+  scale=0.92,
+  transform shape,
+  >=Latex,
+  node distance=8mm,
+  process/.style={draw, rounded corners, align=center, font=\small, text width=56mm, minimum height=12mm, inner sep=4mm, fill=white},
+  header/.style={draw, rounded corners, align=center, font=\small\bfseries, text width=56mm, minimum height=12mm, inner sep=4mm, fill=gray!15},
+  arrow/.style={->, thick}
+]
+
+\node[process, font=\small\bfseries, text width=64mm] (data) {Input data: original features $x$};
+\node[process, font=\small\bfseries, text width=72mm, below=10mm of data] (std) {Train-only standardization; re-use $\mu$, $\sigma$ from Appendix~\ref{app:ttm-core}};
+\node[process, font=\small\bfseries, text width=80mm, below=8mm of std] (jac) {Common pullback evaluation; accumulate $\log\eta(S(u)) + \log|\det \nabla_u S(u)|$ (Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction})};
+
+\node[header, below=18mm of jac, xshift=-70mm] (ttmTitle) {Triangular transport maps};
+\node[header, below=18mm of jac] (trtfTitle) {Transformation Random Forests};
+\node[header, below=18mm of jac, xshift=70mm] (copTitle) {Copula models};
+
+\matrix (ttmMatrix) [below=6mm of ttmTitle, column sep=6mm, row sep=4mm] {
+  \node[header, text width=52mm] (ttmMargTitle) {TTM-Marginal}; & \node[header, text width=52mm] (ttmSepTitle) {TTM-Separable}; \\
+  \node[process, text width=52mm] (ttmMargStepOne) {Closed-form affine coefficients $(a_k, b_k)$ in $u$-space}; & \node[process, text width=52mm] (ttmSepStepOne) {Context basis $g_k$ on predecessors, monotone basis $h_k$ on $u_k$}; \\
+  \node[process, text width=52mm] (ttmMargStepTwo) {Forward pass via \texttt{ttm\_forward}; sum log Jacobian diagonals}; & \node[process, text width=52mm] (ttmSepStepTwo) {Solve constrained optimisation, enforce positivity, store Jacobian terms}; \\
+};
+
+\node[process, below=6mm of trtfTitle, text width=62mm] (trtfStepOne) {Estimate monotone marginals (Appendix~\ref{app:trtf})};
+\node[process, below=6mm of trtfStepOne, text width=62mm] (trtfStepTwo) {Fit transformation forests for $X_k \mid X_{1:(k-1)}$ with additive predictor};
+\node[process, below=6mm of trtfStepTwo, text width=62mm] (trtfStepThree) {Recover triangular likelihood via probability integral transform};
+
+\node[process, below=6mm of copTitle, text width=60mm] (copStepOne) {Estimate train marginals $\widehat F_k$, densities $\widehat f_k$};
+\node[process, below=6mm of copStepOne, text width=60mm] (copStepTwo) {Fit dependence on probit scores (KDE copula)};
+\node[process, below=6mm of copStepTwo, text width=60mm] (copStepThree) {Combine marginals with copula-based dependence};
+
+\node[process, font=\small\bfseries, text width=150mm, below=26mm of trtfStepThree] (outputs) {Reported outputs: log density, conditional diagnostics, samples, calibration, compute summaries};
+
+\draw[arrow] (data) -- (std);
+\draw[arrow] (std) -- (jac);
+\draw[arrow] (jac.south west) to[out=215,in=90] (ttmTitle.north);
+\draw[arrow] (jac) -- (trtfTitle);
+\draw[arrow] (jac.south east) to[out=325,in=90] (copTitle.north);
+
+\draw[arrow] (ttmTitle) -- (ttmMargTitle);
+\draw[arrow] (ttmTitle) -- (ttmSepTitle);
+\draw[arrow] (ttmMargTitle) -- (ttmMargStepOne);
+\draw[arrow] (ttmMargStepOne) -- (ttmMargStepTwo);
+\draw[arrow] (ttmSepTitle) -- (ttmSepStepOne);
+\draw[arrow] (ttmSepStepOne) -- (ttmSepStepTwo);
+
+\draw[arrow] (trtfTitle) -- (trtfStepOne);
+\draw[arrow] (trtfStepOne) -- (trtfStepTwo);
+\draw[arrow] (trtfStepTwo) -- (trtfStepThree);
+
+\draw[arrow] (copTitle) -- (copStepOne);
+\draw[arrow] (copStepOne) -- (copStepTwo);
+\draw[arrow] (copStepTwo) -- (copStepThree);
+
+\draw[arrow] (ttmMargStepTwo) to[out=-90,in=195] (outputs.west);
+\draw[arrow] (ttmSepStepTwo) to[out=-90,in=165] (outputs.west);
+\draw[arrow] (trtfStepThree) -- (outputs);
+\draw[arrow] (copStepThree) to[out=-90,in=0] (outputs.east);
+
+\end{tikzpicture}%
+\caption{Unified evaluation pipeline shared by triangular transport maps, Transformation Random Forests, and copulas.
+The diagram shows how standardized features flow through the triangular pullback or copula dependence block before reporting log densities, conditional diagnostics, samples, calibration metrics, and compute summaries.
+The shared preprocessing, Jacobian accumulation, and reporting path define the evaluation protocol used across the thesis.}
+\label{fig:transport-schematic}
+\end{figure}
+\end{landscape}
+
+\section{Pseudo-code Summaries for Model Routines}
+
+This appendix records consolidated pseudo-code for the core R implementations used in the experiments.
+Each summary captures inputs, main processing stages, and outputs so the execution flow is transparent without consulting the source code files.
+
+\subsection{Transformation Random Forest (TRTF)}\label{app:trtf}
+
+\textbf{Routine:}\quad\texttt{fit\_TRTF(S, config, seed, cores)} (calls \texttt{mytrtf}).
+
+\begin{enumerate}
+  \item Validate that the training matrix is numeric, set the RNG seed, and label columns as $X_1,\ldots,X_K$.
+  \item Fit an intercept-only transformation model \texttt{BoxCox} for each $X_k$ to provide baseline monotone transformations.
+  \item For $k = 2,\ldots,K$:
+    \begin{enumerate}
+      \item Build the formula $X_k \sim X_1 + \cdots + X_{k-1}$.
+      \item Choose \texttt{mtry = max(1, floor((k-1)/2))} and standard \texttt{ctree} controls (\texttt{minsplit}, \texttt{minbucket}, \texttt{maxdepth}).
+      \item Fit a transformation forest with \texttt{traforest} and store the conditional model (one forest per $k$).
+    \end{enumerate}
+  \item Return a \texttt{mytrtf} object containing baseline transformations, conditional forests, variable-importance scores, and the seed.
+
+  \item \textbf{Prediction (\texttt{predict.mytrtf}):}
+    \begin{enumerate}
+      \item Convert new data to the same column naming scheme and evaluate $X_1$ through its baseline transformation model to obtain marginal log densities.
+      \item For each conditional forest ($k\geq 2$) evaluate the log density of $X_k$ given $X_{1:(k-1)}$, extracting the diagonal when the forest returns a log-density matrix.
+      \item Stack the per-dimension log densities (\texttt{logdensity\_by\_dim}) or sum them to obtain the joint log likelihood (\texttt{logdensity}).
+    \end{enumerate}
+\end{enumerate}
+
+\subsection{Nonparametric Copula Baseline}\label{app:copula}
+
+\textbf{Routine:}\quad\texttt{fit\_copula\_np(S, seed)}.
+
+\begin{enumerate}
+  \item Inspect the training matrix and optional class labels; detect whether the dedicated copula packages are available.
+  \item If prerequisites fail (dimension $K \neq 2$ or labels missing), fall back to independent univariate kernel density estimates per dimension and store them for later interpolation.
+  \item Otherwise, for each class label:
+    \begin{enumerate}
+      \item Fit one-dimensional \texttt{kde1d} models to each marginal $X_1$ and $X_2$.
+      \item Convert training samples to pseudo-observations using mid-ranks scaled by $(n+1)^{-1}$ and clamp to $(\varepsilon, 1-\varepsilon)$.
+      \item Fit a two-dimensional kernel copula with \texttt{kdecopula::kdecop} (method \texttt{TLL2}).
+      \item Store marginals, copula fit, and effective sample size for the class.
+    \end{enumerate}
+  \item Record class priors and return a \texttt{copula\_np} object.
+
+  \item \textbf{Prediction (\texttt{predict.copula\_np}):}
+    \begin{enumerate}
+      \item In fallback mode evaluate each univariate KDE at the requested points and sum log densities.
+      \item In copula mode compute marginal log densities and CDF values, evaluate the copula density, and either:
+        \begin{enumerate}
+          \item Average over class-specific log densities weighted by priors (mixture prediction), or
+          \item Use the class labels supplied at prediction time.
+        \end{enumerate}
+      \item Return per-dimension log densities or their sum depending on the requested type.
+    \end{enumerate}
+\end{enumerate}
+
+\subsection{Triangular Transport Core Utilities}\label{app:ttm-core}
+
+\textbf{Module:}\quad\texttt{ttm\_core.R} (shared by marginal and separable TTM fits).
+
+\begin{enumerate}
+  \item Provide train-only standardisation helpers that cache feature means and standard deviations and reapply them to new data.
+  \item Define basis builders: polynomial features for predecessor coordinates $g_k$, monotone basis functions $f_k$ for the current coordinate, and their derivatives.
+  \item Implement optional ordering heuristics (identity or Cholesky pivoting with optional Gaussianisation) and persist selected permutations.
+  \item Expose a dispatcher \texttt{ttm\_forward(model, X)} that:
+    \begin{enumerate}
+      \item Standardises inputs using stored parameters.
+      \item For marginal maps applies affine transformations $a_k + b_k x_k$ with precomputed coefficients.
+      \item For separable maps constructs $g_k$ and $f_k$, computes $S_k = g_k + f_k$, and records the Jacobian diagonal $\partial_{x_k} S_k$.
+    \end{enumerate}
+  \item Provide \texttt{ttm\_ld\_by\_dim} to combine the forward map with the Gaussian reference, yielding per-dimension log densities used by all TTM variants.
+\end{enumerate}
+
+\subsection{Marginal Triangular Transport Map}\label{app:ttm-marg}
+
+\textbf{Routine:}\quad\texttt{fit\_ttm\_marginal(data, seed)}.
+
+\begin{enumerate}
+  \item Split data into train/test subsets if only a matrix is provided; otherwise accept a prepared list.
+  \item Standardise training features and, for each dimension $k$, compute closed-form coefficients $(a_k, b_k)$ that minimise the Gaussian pullback objective subject to $b_k > 0$.
+  \item Store model parameters (standardisation, per-dimension coefficients, ordering) and time measurements.
+  \item During prediction call \texttt{ttm\_forward} with the marginal coefficients and convert Jacobian diagonals to log densities via \texttt{ttm\_ld\_by\_dim}; aggregate per-dimension contributions when the joint log density is requested.
+\end{enumerate}
+
+\subsection{Separable Triangular Transport Map}\label{app:ttm-sep}
+
+\textbf{Routine:}\quad\texttt{fit\_ttm\_separable(data, degree\_g, lambda, seed)}.
+
+\begin{enumerate}
+  \item Prepare train/test splits and standardise training features as in the marginal case.
+  \item For each coordinate $k$:
+    \begin{enumerate}
+      \item Build polynomial features $g_k$ on previous coordinates (degree set by \texttt{degree\_g}).
+      \item Build monotone basis functions $f_k$ on the current coordinate and their derivatives.
+      \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
+      \item Otherwise solve the regularised optimisation problem
+            $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
+            using \texttt{optim} with L-BFGS-B while enforcing positivity of the derivative.
+      \item Store coefficients $c_{\text{non}}$ and $c_{\text{mon}}$ for the coordinate.
+    \end{enumerate}
+  \item Assemble the model list with standardisation parameters, coefficients, and metadata; record training/prediction timings.
+  \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
+\end{enumerate}
+
+\subsection{Evaluation Utilities}\label{app:evaluation}
+
+\textbf{Module:}\quad\texttt{evaluation.R} (experiment orchestration).
+
+\begin{enumerate}
+  \item Define convenience helpers such as \texttt{stderr(x)} and \texttt{add\_sum\_row} for table post-processing.
+  \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
+  \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split, times their evaluations, and returns the fitted objects together with per-dimension log-likelihood arrays.
+  \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
+  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artefacts.
+\end{enumerate}
+
+These structured summaries allow reproducing the algorithmic flow of each model without navigating the full R implementation.
 

### 2025-09-22 report/chapter01.Rnw
.

@@
 
-Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty.
-Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices.
-A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals.
-Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background.
-We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction.
-All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.
+% Checklist: integrate unified transport framing, enforce US spelling, add dynamic refs.
+Multivariate density estimation supports likelihood-based modeling, anomaly detection, simulation, and decision making under uncertainty.
+Tabular datasets often combine moderate to high dimension with context-dependent conditional structure.
+Conditional variance can change with predictors, and conditional skewness or modality can depend on earlier variables.
+A transport perspective addresses these challenges by coupling the target to a simple reference through an invertible map.
+This perspective enables exact likelihoods, transparent conditionals, and efficient sampling through a shared computational backbone \citep{rosenblatt1952remarks,knothe1957contributions}.
+This thesis compares three estimator families inside a single evaluation frame with matched objectives and diagnostics.
+The frame evaluates lower-triangular transport maps, Transformation Random Forests interpreted via probability integral transforms, and copulas that decouple marginals from dependence \citep{hothorn2017transformation,hothorn2018conditional,sklar1959fonctions}.
+\medskip
 
-This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence.
-Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required.
-The following sections outline our contributions, the research questions we address, and the scope of the work.
+We work in standardized coordinates to unify Jacobians, gradients, and reporting conventions. Let $x \in \mathbb{R}^K$ denote features on the original scale, and define
+\begin{equation}
+  u = T_{\text{std}}(x) = (x - \mu) \oslash \sigma,
+  \qquad \sigma_k > 0,\label{eq:ch1-standardization}
+\end{equation}
+which standardizes each feature using training statistics only and fixes the coordinates where derivatives and Jacobians are evaluated.
+Let $S: u \mapsto z$ be a monotone lower-triangular map, and let $\eta$ denote the standard normal density.
+The change-of-variables identity gives
+\begin{equation}
+  \log \pi_U(u) = \log \eta\big(S(u)\big) + \log \big|\det \nabla_u S(u)\big|.
+  \label{eq:ch1-pullback}
+\end{equation}
+Equation~\eqref{eq:ch1-pullback} splits the log density into a reference-fit term and a Jacobian term, isolating modeling capacity from the exact volume correction contributed by the map.
+For triangular $S$, the determinant factorizes as
+\begin{equation}
+  \log \big|\det \nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k} S_k(u_{1:k}).
+  \label{eq:ch1-triangular-det}
+\end{equation}
+Equation~\eqref{eq:ch1-triangular-det} reduces the determinant to a sum of one-dimensional log derivatives, yielding linear per-sample complexity in $K$ and improving numerical stability.
+Reported log densities on the original scale apply the diagonal affine correction
+\begin{equation}
+  \log \pi_X(x) = \log \pi_U\!\big(T_{\text{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k.
+  \label{eq:ch1-affine-correction}
+\end{equation}
+Equation~\eqref{eq:ch1-affine-correction} subtracts a constant offset determined by the training scales and ensures comparability across models that share standardized coordinates.
+\medskip
 
-\section{Contributions and research questions}
+We study separable triangular components for clarity, efficiency, and interpretability. Component $k$ decomposes into a context shift and a univariate monotone shape,
+\begin{equation}
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),\qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k),
+  \label{eq:ch1-separable}
+\end{equation}
+which stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.
 
-We make three contributions within a single, notation-coherent framework.
+% Checklist: add thesis overview, enforce dynamic references, keep ASCII hyphenation.
+\section{Thesis and Problem Statement}\label{sec:ch1-problem}
+This thesis investigates tabular multivariate density estimation within a unified transport-based evaluation frame.
+We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests with axis-parallel splits (TRTF), and copula baselines.
+\medskip
+
+A transport perspective couples standardized data to a Gaussian reference through a monotone lower-triangular map.
+This structure yields exact likelihoods, transparent conditionals, exact inversion by back substitution, and linear per-sample evaluation.
+The Rosenblatt and Knothe rearrangements justify the triangular coupling for any variable order \citep{rosenblatt1952remarks,knothe1957contributions}.
+\medskip
+
+We standardize features using training statistics only.
+Equation~\eqref{eq:ch1-standardization} defines the standardized coordinates $u$ used for evaluation.
+All derivatives and Jacobians are computed in $u$.
+The diagonal affine correction in Equation~\eqref{eq:ch1-affine-correction} reports log densities on the original scale $x$.
+This convention keeps objectives, diagnostics, and comparisons interoperable across estimators and datasets.
+All negative log-likelihoods are reported in nats.
+\medskip
+
+We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\phi$ and $\Phi$.
+Triangular transport maps are abbreviated TTM, with the separable variant labeled TTM-Sep.
+Transformation Random Forests are abbreviated TRTF, and the axis-parallel implementation is labeled TRTF.
+Copulas decouple marginals from dependence and serve as interpretable baselines.
+\medskip
+
+Separable triangular maps decompose each component into a context shift and a univariate monotone shape as shown in Equation~\eqref{eq:ch1-separable}.
+The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant.
+TRTF implements the same separable triangular likelihood via the probability integral transform and an additive predictor.
+Copulas preserve explicit marginals and model dependence on the unit hypercube.
+
+We evaluate all estimators under a single protocol that records test log-likelihoods, conditional diagnostics, and compute.
+The protocol fixes the map direction from standardized data to the Gaussian reference, which preserves separability and linear evaluation cost.
+Section~\ref{sec:evaluation-protocol} details the metrics, calibration diagnostics based on probability integral transforms, and timing conventions.
+
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} visualizes the pipeline by showing standardization $u = T_{\text{std}}(x)$, the triangular transport branch containing TTM-Sep and TRTF, and the copula branch.
+Both branches feed the reported outputs, namely log density, conditionals, sampling, calibration, and compute, under the shared frame.
+
+The central problem is to determine when separability is appropriate for tabular data.
+We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention.
+Ordering effects, conditional calibration, and computational trade-offs address this question.
+\medskip
+
+On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MINIBOONE benchmark it improves on Gaussian references but trails published flow baselines.
+Chapter~\ref{ch:dataanalysis} presents the evidence and discusses these comparisons.
+
+\paragraph{Non-goals.} We do not treat high-capacity normalizing flows as primary models, and we restrict nonparametric copulas to low dimensions.
+Section~\ref{sec:ch1-contributions} states the scope and non-goals for reference.
+
+% Checklist: consolidate notation, relocate figure, keep one-claim sentences.
+\section{The Transport Frame on One Page}\label{sec:ch1-frame}
+This section fixes the evaluation frame used across triangular transport maps, Transformation Random Forests, and copulas.
+It establishes notation, equations, and reporting conventions for the remainder of the thesis.
+The frame places all derivatives and Jacobians in standardized coordinates, applies a single diagonal affine correction on the original scale, and yields linear per-sample evaluation through a lower-triangular map.
+
+Train-only standardization defines the evaluation coordinates via Equation~\eqref{eq:ch1-standardization}.
+The transformation uses training means and scales once and prevents information leakage into evaluation.
+All gradients and Jacobians act on $u$, which keeps objectives comparable across estimators and datasets.
+We convert to the original scale only when reporting.
+\medskip
+
+Let $S:u\mapsto z$ be a monotone lower-triangular map, and let $\eta(z)$ denote the $K$-variate standard normal density.
+The pullback identity in Equation~\eqref{eq:ch1-pullback} evaluates the reference at $S(u)$ and applies the exact volume correction.
+The split isolates model fit from geometry: the first term measures closeness to the Gaussian reference, and the second term contributes the exact Jacobian adjustment implied by the map.
+\medskip
+
+Lower-triangular structure with strictly positive diagonal partial derivatives factorizes the Jacobian as in Equation~\eqref{eq:ch1-triangular-det}.
+The factorization yields $\mathcal{O}(K)$ per-sample time for likelihood evaluation and improves numerical stability.
+It also guarantees global invertibility by back substitution consistent with the Rosenblatt and Knothe rearrangements.
+\medskip
+
+Reported log densities on the original scale apply only the diagonal affine correction implied by standardization.
+Equation~\eqref{eq:ch1-affine-correction} provides that adjustment.
+This convention keeps training and evaluation in $u$-space and converts to $x$-space at report time.
+All negative log-likelihoods are reported in nats, which fixes units across tables and figures.
+\medskip
+
+For clarity and efficiency we use separable triangular components of the form in Equation~\eqref{eq:ch1-separable}.
+Separable structure shifts location through $g_k$ and fixes conditional shape through the univariate monotone $h_k$.
+The Jacobian contribution depends only on $u_k$, which simplifies inversion by back substitution and stabilizes the log-determinant accumulation.
+The constraint also clarifies limits because conditional variance, skewness, and modality do not vary with context under separability.
+
+The core operations follow the same path for all estimators.
+\paragraph{Core operations.}
 \begin{enumerate}
-  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
-  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
+  \item First, standardize input features with training statistics using Equation~\eqref{eq:ch1-standardization}.
+  \item Then, learn a monotone lower-triangular map $S$ to the standard normal reference and exploit Equation~\eqref{eq:ch1-pullback}.
+  \item Finally, evaluate likelihoods, conditionals, and samples through Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction}.
 \end{enumerate}
 
-Two questions motivate the empirical study.
-\begin{enumerate}
-  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines?
-We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
-  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)?
-We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
-\end{enumerate}
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline.
+The input $x$ enters standardization to produce $u$.
+The triangular branch houses TTM-Sep and TRTF, whereas the copula branch couples fitted marginals to a dependence density.
+Both branches feed the same reported quantities, namely log density, conditionals, sampling, calibration, and compute.
+The appendix placement preserves the full schematic while keeping Chapter~\ref{ch:intro} focused on the narrative.
 
-\section{Scope and positioning}
+Notation remains consistent.
+We write $\eta$ for the $K$-variate standard normal density, and $\phi$ and $\Phi$ for the univariate standard normal density and CDF.
+We reserve $u$ for standardized coordinates and $x$ for original coordinates, and we compute all derivatives with respect to $u$.
+These choices align symbols across Chapters~\ref{ch:intro}--\ref{ch:dataanalysis} and prevent ambiguity in later diagnostics and tables.
 
-We focus on triangular transport maps, transformation random forests, and copulas.
-Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions.
-Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used.
-Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions.
-Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.
+This one-page frame removes duplicated exposition from Chapter~\ref{ch:background}.
+It establishes where logs and Jacobians live and makes complexity, inversion, and units explicit before the comparisons that follow.
+Section~\ref{sec:ch1-problem} documented the motivation, and Section~\ref{sec:ch1-contributions} states the resulting contributions and research questions.
 
-\section{Takeaways and roadmap}
+% Checklist: map contributions to evidence, keep bridges to later sections, ensure dynamic refs.
+\section{Contributions and Research Questions}\label{sec:ch1-contributions}
+This section states the contributions of the thesis and formulates the research questions.
+It also maps both elements to the chapters and figures that deliver the evidence.
+The shared transport frame fixes standardized coordinates, keeps all derivatives and Jacobians in $u$-space, and applies a single diagonal affine correction on the original scale for reporting; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline and anchors the comparisons that follow.
+All negative log-likelihoods are reported in nats, and evaluation uses linear-time lower-triangular maps.
 
-A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail.
-Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests.
-Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results.
-Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
+The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests with axis-parallel splits, and copula baselines.
+The view specifies standardization $u = T_{\text{std}}(x)$ and the pullback identity with a monotone lower-triangular map.
+It also fixes the Jacobian factorization and the affine correction for reporting, which together align objectives and diagnostics across estimators.
+Equations~\eqref{eq:ch1-standardization}--\eqref{eq:ch1-affine-correction} and Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} establish these conventions and remove duplication in later chapters.
+
+The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting.
+We evaluate TTM-Sep, TTM-Marg, TRTF, and copulas on synthetic generators and on real tabular data.
+The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation.
+Section~\ref{sec:evaluation-protocol} defines the protocol, Section~\ref{sec:synthetic-results} presents the synthetic and autoregressive results, and Section~\ref{sec:realdata} positions our measurements against published normalizing-flow baselines where appropriate.
+
+The third contribution distills practical guidance from the unified frame and the benchmarks.
+We state operational choices that preserve comparability, highlight ordering sensitivity and separability limits, and summarize when copulas serve as informative baselines.
+Chapter~\ref{ch:conclusion} consolidates these points as actionable recommendations and records limitations that motivate richer parameterizations or alternative predictors.
+
+Two questions drive the empirical study and bind the contributions to specific measurements.
+The first question asks how TRTF compares with TTM-Sep and copula baselines on synthetic data.
+All estimators share the transport frame in this comparison.
+We answer by reporting average test negative log-likelihoods in nats, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost.
+Section~\ref{sec:synthetic-results} provides the corresponding tables and figures.
+
+The second question asks how closely our TRTF results on real benchmarks approach the published performance of modern normalizing flows under the standard preprocessing.
+We answer by placing our test log-likelihoods beside reported numbers from the literature.
+The gaps are interpreted through the separable Jacobian constraint and compute profiles.
+Section~\ref{sec:realdata} reports these comparisons, and Chapter~\ref{ch:conclusion} interprets their implications for model choice.
+
+Scope and non-goals maintain focus and ensure reproducibility.
+We study separable triangular maps and TRTF with additive predictors and compute all derivatives and Jacobians in standardized coordinates.
+The map direction $S:u \to z$ remains fixed for evaluation and inversion.
+Copulas include Gaussian dependence and a low-dimensional nonparametric variant used strictly as a diagnostic baseline.
+We treat high-capacity flows as external references rather than primary models, and we do not evaluate non-additive TRTF variants or cross-term triangular maps in this chapter.
+Chapter~\ref{ch:background} records the formal assumptions and notation.
+Section~\ref{sec:datasets-preprocessing} details preprocessing, seeds, and reporting conventions that keep results comparable across datasets and estimators.
+
+Taken together, these commitments make the comparisons interpretable, keep units and complexity explicit, and prepare the reader for the empirical evidence that answers the two questions under a single, transparent evaluation frame.

### 2025-09-22 report/chapter02.Rnw
.

@@
 opts_chunk$set(
-    fig.path='figure/ch02_fig',    
-    self.contained=FALSE,
-    cache=TRUE
+    fig.path = 'figure/ch02_fig',
+    self.contained = FALSE,
+    cache = TRUE
 )
@@
 
-We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$).
-All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated.
-This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.
+\section{Transport Frame and Notation}\label{sec:transport-frame}
 
-\section{Change of Variables in Standardized Coordinates}
+This section fixes the standardized coordinate system, notation, and algebraic identities used throughout the thesis.
+The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters.
+We summarise the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorisation that drives evaluation and inversion.
 
-The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
+We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardisation map
 \begin{equation}
-  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
-  \label{eq:pullback}
+  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,\label{eq:transport-standardise}
 \end{equation}
-Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate.
-Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$.
-The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions.
-This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.
+where $\mu$ and $\sigma$ denote the empirical mean and standard deviation estimated on the training split and $\oslash$ denotes elementwise division.
+In words, we shift and rescale features once, using training data only, and keep all derivatives and Jacobians in $u$-space to avoid leakage and to ensure comparability across estimators.
 
-It is useful to record the complementary pushforward statement.
-If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$.
-The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion.
-In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.
+The standardised density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$.
+Throughout the thesis the reference is the $K$-variate standard normal density $\eta(z)$.
+The pullback identity then reads
+\begin{equation}
+  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|,\label{eq:transport-pullback}
+\end{equation}
+which evaluates the reference at $S(u)$ and applies the exact volume correction given by the Jacobian determinant.
+Reporting log densities on the original scale requires only the diagonal affine correction implied by standardisation,
+\begin{equation}
+  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.\label{eq:transport-affine}
+\end{equation}
+We therefore differentiate with respect to $u$, and we convert to $x$-scale only at reporting time.
 
-\section{Triangular Structure and Its Consequences}
+The transport is assumed to be lower triangular and componentwise monotone,
+\begin{equation}
+  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), \qquad \partial_{u_k}S_k(u_{1:k})>0,\label{eq:transport-triangular}
+\end{equation}
+so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorises into a sum of one-dimensional log derivatives,
+\begin{equation}
+  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).\label{eq:transport-det}
+\end{equation}
+The factorisation yields $\mathcal{O}(K)$ evaluation cost per sample, improves numerical stability, and guarantees global invertibility: strict monotonicity lets us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.
 
-A triangular map decomposes component-wise,
+Table~\ref{tab:transport-notation} consolidates the notation used in this transport frame.
+All derivatives and Jacobians act on $u$; the affine correction \eqref{eq:transport-affine} converts log densities back to $x$ for reporting.
+The remainder of this chapter adopts this frame.
+Section~\ref{sec:transport-separable} details the separable triangular parameterisation used for direct transports.
+Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform.
+Section~\ref{sec:transport-copula} places copulas in the same reporting convention.
+
+\begin{table}[t]
+  \centering
+  \caption{Notation for the transport frame used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
+All derivatives and Jacobians are taken with respect to $u$; log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.}
+  \label{tab:transport-notation}
+  \begin{tabular}{ll}
+    \hline
+    Symbol & Meaning \\ 
+    \hline
+    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
+    $T_{\mathrm{std}}$ & Standardisation map using training $(\mu,\sigma)$ \\ 
+    $u = T_{\mathrm{std}}(x)$ & Standardised evaluation coordinates \\ 
+    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
+    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
+    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
+    $\eta(z)$ & $K$-variate standard normal density \\ 
+    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
+    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space, respectively \\ 
+    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
+    $K$ & Dimension of the feature vector \\ 
+    \hline
+  \end{tabular}
+\end{table}
+
+\section{Separable Triangular Maps and Transformation Random Forests as Transport}\label{sec:transport-separable}
+
+% Checklist: enforce US spelling, ensure dynamic cross-references, keep equation labels aligned.
+This section unifies separable triangular maps and Transformation Random Forests (TRTF) within the transport frame fixed in Section~\ref{sec:transport-frame}.
+Both estimators realize a monotone lower-triangular map $S:u\mapsto z$ that couples the standardized target to the Gaussian reference $\eta$.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the shared backbone and locates \textsc{TTM-Sep} and \textsc{TRTF} on the transport branch introduced in Chapter~\ref{ch:intro}.
+We focus on shared likelihood identities, modeling assumptions, and limits of separability, and defer implementation details to Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.
+
+The goal is to state a single likelihood for both constructions, clarify what separability permits, and identify failure modes that motivate richer parameterizations.
+We do not pursue non-additive TRTF predictors, cross-term triangular maps, or ordering heuristics in this section; Chapter~\ref{ch:dataanalysis} evaluates those choices empirically and Appendix~\ref{ch:appendix} documents routines and defaults.
+
+We adopt the notation introduced in Section~\ref{sec:transport-frame}.
+Coordinates satisfy $u=T_{\mathrm{std}}(x)$, the reference density is $\eta(z)$, and the pullback identity \eqref{eq:transport-pullback} gives $\pi_U(u)=\eta(S(u))\,|\det\nabla_u S(u)|$.
+The map is lower-triangular with strictly positive diagonal partial derivatives, which yields the sum decomposition in Equation~\eqref{eq:transport-det}.
+These conventions keep derivatives in $u$-space and apply the affine correction \eqref{eq:transport-affine} only when reporting $\log \pi_X(x)$.
+
+We restrict attention to separable triangular maps. Component $k$ decomposes into a context shift and a univariate monotone shape,
 \begin{equation}
-  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
-  \label{eq:triangular-map}
+  S_k(u_{1:k}) \;=\; g_k(u_{1:k-1}) + h_k(u_k),\qquad \log \partial_{u_k}S_k(u_{1:k}) \;=\; \log h_k'(u_k),\label{eq:transport-separable}
 \end{equation}
-and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
+which fixes context effects in $g_k$ and reserves $h_k$ for the one-dimensional marginal shape.
+Intuitively, earlier coordinates translate the location, while the conditional shape along $u_k$ remains fixed across contexts.
+The Jacobian contribution depends only on $u_k$, which reduces per-sample evaluation cost and simplifies inversion.
+
+Substituting the standard normal reference into \eqref{eq:transport-pullback} produces a separable objective,
 \begin{equation}
-  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
-  \qquad
-  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
-  \label{eq:triangular-det}
+  \log \pi_U(u) \;=\; \sum_{k=1}^{K}\Big[\log \varphi\!\big(S_k(u_{1:k})\big) + \log h_k'(u_k)\Big],\label{eq:transport-likelihood}
 \end{equation}
+where $\varphi$ denotes the univariate standard normal density.
+Equation~\eqref{eq:transport-likelihood} splits the log density into a reference fit and an exact volume correction.
+In plain language, the model evaluates how Gaussian each transformed coordinate appears, then corrects for the local stretch induced by $h_k$.
+The same decomposition produces linear per-sample time in $K$ and stable accumulation of log derivatives.
 
-Three consequences matter in practice.
-First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$.
-Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}.
-Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components.
-Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.
+The negative log-likelihood per sample takes the quadratic-plus-barrier form
+\begin{equation}
+  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big],\label{eq:transport-loss}
+\end{equation}
+which follows because $\log \varphi(t) = -\tfrac{1}{2}t^2 - \tfrac{1}{2}\log(2\pi)$ and constants independent of the parameters drop out.
+Equation~\eqref{eq:transport-loss} pulls each component toward the reference while preventing degenerate derivatives through the log barrier.
+In practice we enforce $h_k'(u_k)>0$ by construction and control tails with mild regularization; implementation choices appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.
 
-Existence and anisotropy deserve a brief remark.
-For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}.
-Different orderings induce different maps.
-While all are valid, their sparsity and approximation difficulty may vary with order.
-When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling.
-In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.
+Separable structure encodes clear modeling assumptions.
+Conditional variance, skewness, and modality do not change with the preceding coordinates once $g_k$ shifts location.
+Consequently, separable maps can underfit heteroskedastic or multimodal conditionals, which manifests as U-shaped or inverted-U probability integral transform (PIT) diagnostics.
+Variable ordering also matters for finite bases because triangular transports are anisotropic, even though a Knothe--Rosenblatt rearrangement exists for any ordering \citep{rosenblatt1952remarks,knothe1957contributions}.
+These caveats guide the robustness checks in Chapter~\ref{ch:dataanalysis}.
 
-Finally, we keep terminology consistent.
-A log density is $\log \hat\pi(\cdot)$ at a point.
-A (test) log likelihood is the dataset average of log densities.
-Its negative is the negative log likelihood (NLL) in nats; lower is better.
-These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.
+\subsection{Transformation Random Forests within the Transport Frame}\label{sec:transport-trtf}
 
-\section{Separable and Cross-Term Triangular Transports}
+Transformation Random Forests \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation} fit into the same transport frame through the probability integral transform.
+Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by a TRTF for coordinate $k$.
+The induced triangular component is
+\begin{equation}
+  S_k(u_{1:k}) \;=\; \Phi^{-1}\!\Big(\widehat F_k(u_k \mid u_{1:k-1})\Big),\label{eq:transport-trtf-map}
+\end{equation}
+which maps conditionals to standard normal margins.
+In plain language, TRTF predicts a conditional CDF, then the probit transform places the result on the Gaussian reference scale.
+Differentiating $\Phi\!\big(S_k(u_{1:k})\big)=\widehat F_k(u_k \mid u_{1:k-1})$ with respect to $u_k$ yields
+\begin{equation}
+  \widehat \pi_k(u_k \mid u_{1:k-1}) \;=\; \varphi\!\big(S_k(u_{1:k})\big)\,\partial_{u_k}S_k(u_{1:k}),\label{eq:transport-trtf-likelihood}
+\end{equation}
+which is exactly the pullback factor in Equation~\eqref{eq:transport-likelihood}. Summing over $k$ recovers Equation~\eqref{eq:transport-likelihood} in standardized coordinates.
 
-A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.
+The additive-predictor TRTF used in this thesis yields a separable transport. Under the model
+\begin{equation}
+  \widehat F_k(u_k \mid u_{1:k-1}) \;=\; \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),\label{eq:transport-trtf-additive}
+\end{equation}
+we obtain
+\begin{equation}
+  S_k(u_{1:k}) \;=\; h_k(u_k) + g_k(u_{1:k-1}),\qquad \partial_{u_k}S_k(u_{1:k}) \;=\; h_k'(u_k),\label{eq:transport-trtf-separable}
+\end{equation}
+so TRTF implements the same separable triangular likelihood as the direct parameterization in Equation~\eqref{eq:transport-separable}.
+Monotonicity in $u_k$ holds by construction, the Jacobian depends only on $u_k$, and inversion proceeds by back-substitution identical to the separable map.
+This equivalence underpins the empirical comparisons in Chapter~\ref{ch:dataanalysis}.
 
-\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
+The equivalence also clarifies limits.
+Additive TRTF predictors shift location but cannot alter conditional shape with context, which mirrors the separable constraint.
+Axis-aligned partitions stabilize estimation, yet they do not remove residual multimodality when the conditional shape varies with $u_{1:k-1}$.
+These limits are visible in PIT diagnostics and conditional negative log-likelihood decompositions on synthetic studies.
+
+We emphasize operational scope and supporting references.
+All derivatives and Jacobians are computed in standardized coordinates, evaluation uses the triangular pullback, and reported log densities on the original scale include the affine correction \eqref{eq:transport-affine}.
+Implementation details on basis choices for $h_k$, feature construction for $g_k$, regularization, derivative clipping, timing, and memory footprints appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}, which also provides pseudo-code for both estimators.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} visualizes how the \textsc{TTM-Sep} and \textsc{TRTF} branches share the same computational path from standardized data to reported likelihoods.
+
+In summary, separable triangular maps and additive-predictor TRTF realize the same lower-triangular likelihood once the data are standardized.
+The shared structure yields exact likelihoods, exact inversion, transparent conditionals, and linear per-sample complexity, but it restricts context-dependent shape.
+Section~\ref{sec:transport-copula} positions copulas within the same reporting convention to decouple marginals from dependence.
+
+\section{Copula Baselines}\label{sec:transport-copula}
+
+% Checklist: align references with transport frame, cite Sklar, enforce US spelling.
+This section positions copulas within the unified transport frame and links their reported likelihoods to the evaluation conventions used for triangular maps and Transformation Random Forests.
+Copulas decouple marginal modeling from dependence modeling by pairing univariate marginals with a separate dependence density on the unit hypercube.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} displays the copula branch beside the triangular branch and highlights how both yield comparable reported log densities under the shared evaluation pipeline.
+
+We begin with pseudo-observations built from training-split marginals.
+Let $\widehat F_k$ denote the strictly increasing empirical or smoothed CDF of $X_k$ estimated on the training split.
+Define the pseudo-observations and their probit transform as
 \begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
-  \label{eq:separable-component}
+  v_k \;=\; \widehat F_k(x_k),\qquad z_k \;=\; \Phi^{-1}(v_k),\label{eq:copula-probit}
 \end{equation}
-so the log Jacobian depends only on $u_k$:
+which map each coordinate to $(0,1)$ and then to $\mathbb{R}$ through the probit function.
+In plain language, the marginals become uniform scores, and $z$ records those scores on a Gaussian scale.
+Mid-ranks and clamping near $(0,1)$ stabilize the transformation in finite samples.
+
+The copula representation combines marginal densities with a dependence factor. The joint log density on the original scale satisfies
 \begin{equation}
-  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
-  \label{eq:separable-jacobian}
+  \log \widehat \pi_X(x) \;=\; \sum_{k=1}^{K} \log \widehat f_k(x_k) + \log c\!\left(v_1,\ldots,v_K\right),\label{eq:copula-logdensity}
 \end{equation}
-The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it.
-Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$.
-Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.
+where $c$ denotes the copula density on $(0,1)^K$.
+Equation~\eqref{eq:copula-logdensity} separates the task into two parts: fit interpretable marginals and correct for dependence through $\log c$.
+The reported quantity already lives on the original scale, so the affine correction in Equation~\eqref{eq:transport-affine} is unnecessary.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} uses the equivalent shorthand $\log c(z)$ because dependence is evaluated through $z=\Phi^{-1}(v)$.
 
-\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
+The independence baseline fixes a lower bound for dependence modeling. Setting $c \equiv 1$ yields
 \begin{equation}
-  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:cross-component}
+  \log \widehat \pi_X^{\mathrm{ind}}(x) \;=\; \sum_{k=1}^{K} \log \widehat f_k(x_k),\label{eq:copula-independence}
 \end{equation}
-so the log Jacobian, and thus the shape of the conditional, depends on the context.
-This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.
+which treats coordinates as independent after marginal fitting. Chapter~\ref{ch:dataanalysis} uses this baseline as a reference point in evaluation tables and figures.
 
-Two small examples make the contrast concrete.
-Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$.
-A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$.
-A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$.
-As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others.
-Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies.
-Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.
+The Gaussian copula specifies elliptical dependence through a correlation matrix $\Sigma$. With $z=\Phi^{-1}(v)$, the copula density admits the closed form
+\begin{equation}
+  c_{\Sigma}(v) \;=\; |\Sigma|^{-1/2}\,\exp\!\Big(-\tfrac{1}{2}\,z^{\top}(\Sigma^{-1} - I)z\Big),\label{eq:copula-gaussian}
+\end{equation}
+which reduces dependence estimation to fitting $\Sigma$ on the transformed scores.
+In plain language, the Gaussian copula bends the joint shape away from independence according to $\Sigma$ while preserving the learned marginals.
 
-\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable.
-Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.
+A low-dimensional nonparametric variant avoids elliptical assumptions at small $K$. We fit a kernel density $\widehat f_Z$ on $z=\Phi^{-1}(v)$ and recover the copula density by
+\begin{equation}
+  c(v) \;=\; \frac{\widehat f_Z\!\big(\Phi^{-1}(v)\big)}{\prod_{k=1}^{K} \varphi\!\big(\Phi^{-1}(v_k)\big)},\label{eq:copula-kde}
+\end{equation}
+which applies the change of variables from $z$ back to $v$ and yields a proper copula density.
+In words, the kernel density models the joint shape of the probit scores, and division by the product of standard normal densities restores the unit-cube scale.
+This approach is viable only for small $K$, where kernel density estimation remains accurate and stable.
+Chapter~\ref{ch:dataanalysis} employs it strictly as a diagnostic baseline.
 
-\section{Transformation Forests as Triangular Transport}
+The transport frame keeps reporting interoperable across modeling branches despite distinct parameterizations.
+Triangular maps and TRTF evaluate the pullback likelihood in standardized coordinates and apply the fixed affine correction \eqref{eq:transport-affine} when mapping back to $x$.
+Copulas operate on $x$ directly through Equation~\eqref{eq:copula-logdensity}, yet Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows how the probit scores $z$ maintain comparability with the Gaussian reference used above.
+This alignment keeps objectives and diagnostics consistent across Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
 
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}.
-This induces a triangular transport component via the probability integral transform,
-\begin{equation}
-  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
-  \label{eq:trtf-map-theory}
-\end{equation}
-
-\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
-\begin{equation}
-  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
-  \label{eq:trtf-likelihood-theory}
-\end{equation}
-showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly.
-Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}.
-TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.
-
-\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
-\begin{equation}
-  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
-  \label{eq:trtf-additive}
-\end{equation}
-so
-\begin{equation}
-  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:trtf-separable}
-\end{equation}
-and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$.
-Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence.
-TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-
-A short note on reporting closes the loop.
-TRTF evaluates conditional densities in standardized coordinates.
-Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter.
-With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.
-
-\section{Takeaways}
-
-We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction.
-Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}.
-Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization.
-TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
+Modeling choices and limits follow from the chosen copula.
+The Gaussian copula imposes elliptical dependence and may misrepresent tail behavior or localized asymmetry.
+The nonparametric variant mitigates these issues only at small dimension and sufficient sample size.
+The independence baseline provides a transparent reference when dependence is weak or data are scarce.
+These caveats motivate treating copulas as interpretable baselines rather than definitive high-dimensional models in the empirical study of Chapter~\ref{ch:dataanalysis}.
+Sklar’s theorem underlies all constructions above and formalizes the decoupling of marginals from dependence \citep{sklar1959fonctions}.

### 2025-09-22 report/chapter03.Rnw
.

@@
 
-\section{Setup: Standardization, Map Direction, and Reporting}\label{sec:setup}
+% Checklist: define model abbreviations for concise tables
+\textbf{Model abbreviations.} For the four-dimensional autoregressive study we abbreviate the oracle fits as True-Marg for the marginal reference and True-Joint for the conditional reference.
+Transformation Random Forests appear as TRTF.
+This shorthand matches the axis-parallel implementation (TRTF) used in our experiments.
+We write the marginal triangular transport as TTM-Marg, the separable variant as TTM-Sep, and the copula mixture baseline as Copula.
+We use these labels across tables and figures that report this study.
 
-Observations on the original scale satisfy $x\in\mathbb{R}^K$. We standardize using training-split statistics
+\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}\label{sec:setup}
+
+% Checklist: preserve dynamic refs, ensure ASCII hyphenation, align with future sections.
+This section fixes data sources, generators, and preprocessing so likelihoods, calibration, and compute remain comparable across models.
+All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the common affine correction.
+We keep a single triangular-map direction $S:u \rightarrow z$ across methods to avoid mixed objectives.
+
+We standardize features with training-split statistics only.
+Equation~\eqref{eq:transport-standardise} defines $u = T_{\mathrm{std}}(x) = (x-\mu)\oslash \sigma$ with $\sigma_k > 0$.
+We evaluate $\log \pi_U(u)$ through the pullback identity in Equation~\eqref{eq:transport-pullback}, apply the triangular factorization from Equation~\eqref{eq:transport-det}, then convert to $\log \pi_X(x)$ using the diagonal correction in Equation~\eqref{eq:transport-affine}.
+We report average test negative log-likelihoods (NLL) in nats.
+Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the standardized pipeline shared by transport maps, Transformation Random Forests, and copulas.
+
+We use fixed train, validation, and test splits with proportions $0.60/0.20/0.20$ unless a benchmark provides official splits.
+Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures.
+The canonical four-dimensional ordering is $(1,2,3,4)$.
+Robustness to ordering is assessed by averaging over all $4!
+= 24$ permutations.
+We adopt the natural column order for real datasets.
+Transformation Random Forests are the axis-parallel implementation denoted TRTF throughout.
+We fix seeds $\{11, 13, 17, 19, 23\}$ for data generation and model fitting, and we average repeated runs with standard errors to quantify stochastic variability.
+Figure~\ref{fig:autoregressive-joint-calibration} displays the $20\%$ test split for the synthetic calibration study.
+
+The Half-Moon dataset provides a curved, bimodal joint in $K=2$.
+We draw a class $Y \sim \mathrm{Bernoulli}(0.5)$, an angle $\Theta \sim \mathrm{Unif}[0,\pi]$, and additive noise $\varepsilon \sim \mathcal{N}(0, \sigma^2 I_2)$ with $\sigma = 0.10$.
+For $Y = 0$ we set $m(\Theta) = (\cos \Theta, \sin \Theta)$.
+For $Y = 1$ we set $m(\Theta) = (1 - \cos \Theta, -\sin \Theta + 0.5)$.
+The observed $X = m(\Theta) + \varepsilon$.
+The ``True joint'' oracle evaluates the mixture density by numerical quadrature over $\Theta$ with the known Gaussian noise, and the ``True conditional'' oracle conditions on $Y$.
+Figure~\ref{fig:halfmoon-panels} shows representative contour plots at $n=250$, which align with this generator.
+Table~\ref{tab:halfmoon-nll} reports the corresponding NLLs.
+
+The four-dimensional autoregressive generator combines Gaussian, exponential, beta, and gamma components to induce heteroskedasticity, skew, and conditional multimodality.
+The first coordinate is $X_1 \sim \mathcal{N}(0,1)$.
+The second coordinate is independent $X_2 \sim \mathrm{Exp}(\lambda_0)$ with rate $\lambda_0 = 1$.
+The third coordinate lies on $(0,1)$ and is a context-gated mixture of two beta laws, $X_3 \mid X_{1:2} \sim w\,\mathrm{Beta}(\alpha_1, \beta_1) + (1 - w)\,\mathrm{Beta}(\alpha_2, \beta_2)$.
+We set $(\alpha_1, \beta_1) = (2.5, 5.0)$ and $(\alpha_2, \beta_2) = (5.0, 2.5)$.
+The mixing weight is $w = \sigma(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\sigma(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$.
+The fourth coordinate is positive and conditionally heteroskedastic, $X_4 \mid X_{1:3} \sim \tilde{w}\,\mathrm{Gamma}(k_1, r_1(X_2)) + (1 - \tilde{w})\,\mathrm{Gamma}(k_2, r_2(X_2))$.
+We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \sigma(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$.
+The ``True joint'' baseline uses these closed-form conditionals to evaluate the exact joint density, while the ``True marginal'' baseline uses the corresponding univariate marginals and ignores dependence.
+
+% Checklist: summarise generator config, define softmax usage.
+\begin{table}[t]
+  \centering
+  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study. The softplus transform enforces positive rates, shapes, and scales.}
+  \label{tab:autoregressive-config}
+  \begin{tabular}{lll}
+    \hline
+    Coordinate & Distribution & Parameter mapping \\
+    \hline
+    $X_1$ & Normal & Fixed $\mathcal{N}(0,1)$ \\
+    $X_2 \mid X_1$ & Exponential & $\mathrm{rate} = \mathrm{softplus}(X_1)$ \\
+    $X_3 \mid X_{1:2}$ & Beta & $\mathrm{shape}_1 = \mathrm{softplus}(X_2)$; $\mathrm{shape}_2 = \mathrm{softplus}(X_1)$ \\
+    $X_4 \mid X_{1:3}$ & Gamma & $\mathrm{shape} = \mathrm{softplus}(X_3)$; $\mathrm{scale} = \mathrm{softplus}(X_2)$ \\
+    \hline
+  \end{tabular}
+\end{table}
+
+We apply the softplus transform $\mathrm{softplus}(t) = \log(1 + \exp(t))$ to map unconstrained predictors to strictly positive distribution parameters.
+Mixture weights use the logistic gate $\sigma(a)$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one.
+For completeness, the general softmax takes a vector $a$ and returns $\mathrm{softmax}(a)_i = \exp(a_i) / \sum_j \exp(a_j)$.
+This normalization is essential for the beta and gamma mixtures because it translates linear predictors into valid probability weights while preserving differentiability.
+
+Table~\ref{tab:autoregressive-config} lists the intended marginal families by dimension, and Tables~\ref{tab:autoregressive-perm} and~\ref{tab:autoregressive-perm-avg} summarize the permutation and sample-size studies used later in this chapter.
+
+The MINIBOONE benchmark follows the published preprocessing to ensure comparability with flow-based baselines.
+We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes.
+We use the fixed train, validation, and test splits from the benchmark, apply train-only standardization, and avoid any extra pruning of correlated features.
+We report all log-likelihoods in nats and retain the published naming for flow comparators in later tables.
+Section~\ref{sec:realdata} records these steps and provides the dataset context.
+Table~\ref{tab:uci-loglik} reproduces the flow baselines that motivate our TRTF runs.
+
+Additional UCI datasets appear only when we retain them for real-data context.
+POWER keeps household electricity attributes after jittering the minute-of-day encoding, dropping the calendar date and reactive-power column, and adding uniform noise to break ties.
+GAS keeps the \texttt{ethylene\_CO} subset, treats the series as i.i.d., removes strongly correlated attributes, and retains an eight-dimensional representation.
+HEPMASS keeps only the positive class from the ``1000'' split and discards five variables with repeated values to avoid density spikes.
+These preprocessing steps follow the same train-only standardization and reporting conventions described above.
+Section~\ref{sec:realdata} provides the corresponding background and positions these datasets within our evaluation.
+
+All models use the same standardized frame and direction for evaluation, which keeps objectives, diagnostics, and reported quantities interoperable across triangular transports, TRTF, and copula baselines.
+This alignment is necessary for the conditional decompositions, probability integral transform (PIT) calibration checks, and compute summaries presented later in this chapter.
+
+\section{Models and Implementation}\label{sec:models-implementation}
+
+% Checklist: align references, enforce US spelling, ensure active voice.
+This section specifies the estimators and implementation details that keep likelihoods, calibration, and compute directly comparable across models.
+All estimators share the transport direction $S:u \rightarrow z$, operate in standardized coordinates, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.
+
+We implement separable lower-triangular transport maps denoted TTM-Sep. Component $k$ decomposes into a context shift and a univariate monotone shape,
 \begin{equation}
-  u = T_{\mathrm{std}}(x) = (x-\mu)\oslash\sigma,
-  \qquad \sigma_k > 0,
-  \label{eq:standardization}
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k), \qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k) > 0,
+  \label{eq:ttm-separable-def}
 \end{equation}
-learn a single monotone triangular map in this standardized space,
+so the Jacobian contribution depends only on $u_k$. The structure yields linear per-sample complexity in $K$ and exact inversion by back-substitution.
+
+We minimize the Gaussian pullback objective induced by the shared reference,
 \begin{equation}
-  S: u \longmapsto z, \qquad z \sim \eta = \mathcal{N}(0,I),
-  \label{eq:map-direction}
+  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h_k'(u_k) \Big],
+  \label{eq:ttm-separable-loss}
 \end{equation}
-and sample via $S^{-1}: z \mapsto u \mapsto x$.
-All Jacobians and partial derivatives that involve $S$ are taken with respect to $u$ unless explicitly stated.
-The pullback identity \eqref{eq:pullback} remains the objective we optimize.
+which follows from the change-of-variables identity in Equation~\eqref{eq:transport-pullback} combined with the triangular determinant factorization in Equation~\eqref{eq:transport-det}.
+The quadratic term pulls the transformed coordinates toward the reference, and the log-derivative term prevents degenerate solutions.
+We solve the regularized problem with bound-constrained optimization and enforce monotonicity by construction.
 
-Reported log-densities on the original scale apply the diagonal affine correction
+We construct $h_k$ with monotone one-dimensional bases that combine identity, integrated sigmoids, softplus-like edge terms, and integrated radial basis functions.
+Nonnegativity constraints on the derivative coefficients guarantee $h_k'(u_k) \ge 0$.
+We linearize tails to stabilize likelihoods as $|u_k|$ grows.
+Ridge penalties apply to all basis coefficients, and optional sparsity penalties shrink context shifts when multicollinearity inflates variance.
+During training and evaluation we clip log-derivatives to $[-H, H]$ to avoid numerical overflow in the Jacobian sum; the bound $H$ is tuned on the validation split.
+
+We build $g_k$ from low-degree polynomial features of $u_{1:k-1}$ and drop predecessors whose inclusion does not improve validation likelihood.
+This pruning keeps $\nabla_u S(u)$ sparse and improves stability in small-sample regimes.
+Ordering matters for finite bases, so headline results use the natural variable order while robustness studies vary the order as described in Section~\ref{sec:datasets-preprocessing}.
+When heuristics are applied, we use the identity ordering or a Cholesky-pivoted ordering with optional Gaussianization, as exposed by the shared transport utilities.
+Appendix~\ref{ch:appendix} records how the ordering is stored and reapplied at prediction time.
+
+We reference a cross-term variant, denoted TTM-X, only to delimit scope. The variant augments the separable component with low-rank interactions,
 \begin{equation}
-  \log \pi_X(x) = \log \pi_U\!\big(T_{\mathrm{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k,
-  \label{eq:affine-correction}
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k) + \sum_{j<k} \alpha_{kj}\,q_j(u_j)\,r_k(u_k),
+  \label{eq:ttm-cross}
 \end{equation}
-so every method produces pointwise values $\log \hat\pi_X(x)$ that are directly comparable.
-We reserve the term \emph{log-density} for the value at a single point, \emph{(test) log-likelihood} for the dataset average, and \emph{negative log-likelihood (NLL)} for its negative; NLL is always reported in nats.
+where $q_j$ and $r_k$ are monotone features and constraints ensure $\partial_{u_k} S_k(u_{1:k}) > 0$.
+We exclude TTM-X from headline tables because the interactions alter identifiability and complicate calibration.
+The definition clarifies the naming used in the synthetic analyses.
 
-The choice of direction $S:u\mapsto z$ keeps the change-of-variables identity separable, giving per-sample complexity $\mathcal{O}(K)$ through the triangular determinant in Equation~\eqref{eq:triangular-det}.
-The same structure yields exact inversion---solve $K$ one-dimensional monotone equations in sequence---and transparent conditionals by fixing $u_{1:m-1}$ and inverting the trailing components \citep{rosenblatt1952remarks,knothe1957contributions}.
-Because every method uses the same direction, we avoid mixing objectives or Jacobian conventions and can incorporate copulas, whose natural domain is $x$-space, via the standardization wrapper.
+We implement Transformation Random Forests with additive predictors and denote the model TRTF.
+Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by the forest.
+The induced triangular component is
+\begin{equation}
+  S_k(u_{1:k}) = \Phi^{-1}\!\big(\widehat{F}_k(u_k \mid u_{1:k-1})\big),
+  \label{eq:trtf-transport}
+\end{equation}
+and differentiation yields $\varphi\big(S_k(u_{1:k})\big)\,\partial_{u_k} S_k(u_{1:k}) = \widehat{\pi}_k(u_k \mid u_{1:k-1})$.
+Under the additive predictor $\widehat{F}_k(u_k \mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big)$ we obtain $S_k = h_k + g_k$ and $\partial_{u_k} S_k = h_k'(u_k)$, which matches Equation~\eqref{eq:ttm-separable-def} exactly in the transport frame.
+Consequently TRTF shares the likelihood in Equation~\eqref{eq:ttm-separable-loss}, inherits exact inversion, and differs operationally through forest training and aggregation.
 
-\section{Triangular Transport Maps}
+We keep copulas as dependence baselines with explicit scope.
+We fit only low-dimensional nonparametric copulas for $K \le 3$, using probit-transformed pseudo-observations and kernel density estimation on the Gaussian scale before mapping back to the unit cube with the appropriate Jacobian.
+The independence baseline evaluates the product of fitted marginals.
+We omit a Gaussian copula from the main experiments to preserve consistency with the low-$K$ nonparametric dependence analyzed in Section~\ref{sec:realdata}.
 
-With standardized coordinates fixed, we describe the transports we fit directly.
+We adopt a single inversion and evaluation convention across estimators.
+Training, Jacobians, and conditional evaluations occur in standardized coordinates.
+Sampling draws $z \sim \mathcal{N}(0, I)$, applies $S^{-1}$ by back-substitution, and converts to the original scale with the stored affine parameters.
+This convention prevents mixed objectives and keeps all reported quantities interoperable.
 
-\subsection{Separable and Cross-Term Components}
+We ensure reproducibility and comparability with fixed seeds, cached standardization parameters, and shared reporting utilities.
+Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction, nonparametric copulas, marginal and separable triangular maps, and the shared transport core that implements ordering, bases, derivatives, and evaluation.
+The appendix also records optimizer choices, timing hooks, and object layouts used in the experiments.
 
-A triangular map factorizes as in Equation~\eqref{eq:triangular-map} with strictly positive diagonal derivatives. Two parameterizations anchor the expressiveness spectrum.
+\section{Evaluation Metrics and Protocol}\label{sec:evaluation-protocol}
 
-\paragraph{Separable components.} These decompose into a context shift and a univariate monotone shape,
+% Checklist: align references, keep active voice, reuse existing notation.
+This section defines the metrics and procedures applied across all models so likelihoods, calibration, and compute remain directly comparable.
+We evaluate every estimator in standardized coordinates, apply the triangular determinant, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.
+
+We distinguish pointwise log density from dataset averages.
+The test log likelihood (LL) is the mean of $\log \hat{\pi}_X(x)$ over the test split, and the test negative log likelihood (NLL) is its negative.
+Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity, and all values appear in nats with consistent precision.
+Reported log densities on the original scale equal the standardized quantity minus $\sum_k \log \sigma_k$ as given by Equation~\eqref{eq:transport-affine}.
+Consequently, datasets with larger training scales introduce large constant offsets that the affine correction removes.
+
+Triangular models exploit the separable pullback in standardized coordinates. With $u = T_{\mathrm{std}}(x)$, the log density decomposes as
 \begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = \log h_k'(u_k),
-  \label{eq:methods-separable}
+  \log \hat{\pi}_U(u) = \sum_{k=1}^{K} \Big[ \log \varphi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
+  \label{eq:evaluation-triangular}
 \end{equation}
-so the log-Jacobian term depends only on $u_k$.
-The context can shift location through $g_k$ but cannot reshape the conditional: scale, skewness, tail thickness, and modality stay invariant across contexts.
+so the determinant factorization in Equation~\eqref{eq:transport-det} yields linear per-sample cost in $K$.
+In plain language, the model checks how Gaussian each transformed coordinate looks, then adds the exact log-Jacobian contribution from its one-dimensional derivative.
+The affine correction in Equation~\eqref{eq:transport-affine} converts $\log \hat{\pi}_U$ to $\log \hat{\pi}_X$ for reporting.
 
-\paragraph{Cross-term components.} To capture context-dependent shape we use the integrated-rectifier construction
+We report per-dimension conditional NLLs for triangular models to localize error. For each coordinate,
 \begin{equation}
-  S_k(u_{1:k}) = \int^{u_k} \! \exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k} S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:methods-cross}
+  \mathrm{NLL}_k = -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
+  \label{eq:evaluation-conditional-nll}
 \end{equation}
-so the log-Jacobian depends on both $u_k$ and the context.
-This structure captures conditional heteroskedasticity, skew changes, and context-specific multimodality at the cost of a more delicate numerical treatment.
+and the joint NLL equals $\sum_{k=1}^{K} \mathrm{NLL}_k$ by construction.
+Copulas lack a unique triangular factorization, so we report only their joint NLL.
+Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
+These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame.
 
-Two examples highlight the contrast.
-If $U_2\mid U_1=u_1$ is Gaussian with mean $m(u_1)$ and variance $\sigma^2(u_1)$, a separable map can absorb $m(u_1)$ through $g_2$ but cannot reflect the variance change; a cross-term map adjusts $h_2(u_2,u_1)$ to encode variance inflation or contraction.
-If $U_2\mid U_1$ switches between unimodal and bimodal shapes across contexts, separable maps preserve modality, whereas cross-term components can introduce or remove modes as $u_1$ varies.
+Calibration assesses whether predictive probabilities align with empirical frequencies.
+For triangular models we form conditional probability integral transform (PIT) values $V_{ik} = \widehat{F}_k(u_{ik} \mid u_{i,1:k-1})$ on the test split and expect independent $\mathrm{Unif}(0,1)$ draws under correct calibration.
+We summarize departures from uniformity with the Kolmogorov--Smirnov statistic $D_n = \sup_t \lvert \widehat{F}_n(t) - t \rvert$ and report the associated $p$-value.
+We complement the scalar summary with brief PIT distribution descriptions when patterns recur across seeds.
+For copulas we assess marginal PITs and low-dimensional slices where dependence is transparent.
+Systematic U-shaped or inverted-U PIT indicates under- or over-dispersion and motivates richer parameterizations.
 
-\subsection{Monotone Parameterizations}
+Compute metrics quantify practical cost alongside fit.
+We record wall-clock training time on the training split and per-sample evaluation time on the test split.
+Triangular transports scale linearly in $K$ and approximately linearly in the number of basis functions.
+Transformation Random Forests scale with the number and depth of trees per conditional during training, while prediction remains linear after aggregation.
+Copula training is dominated by correlation estimation or kernel density fitting, followed by fast evaluation.
+We also track peak resident memory when caching affects runtime.
+All timings use the deterministic pipeline defined in Chapter~\ref{ch:dataanalysis} and are averaged over seeds with standard errors.
 
-We rely on two monotonicity mechanisms that mirror the structures above.
+Protocol choices keep comparisons stable and reproducible:
+\begin{enumerate}
+  \item Standardize with training-split statistics, fit a single map $S:u \rightarrow z$, and evaluate Jacobians in standardized space.
+  \item Compute LL, NLL, and conditional decompositions in standardized coordinates, then apply the affine correction once for reporting.
+  \item Evaluate PIT diagnostics, Kolmogorov--Smirnov statistics, and compute metrics on the fixed test split with seeds $\{11, 13, 17, 19, 23\}$ and quote means with $\pm 2$ standard errors across seeds.
+\end{enumerate}
+Appendix~\ref{ch:appendix} lists routine interfaces that support exact re-execution; figure captions and table notes repeat units, splits, and the “higher/lower is better” convention for clarity.
 
-\paragraph{Linear-in-coefficients (LIC) separable maps.} We expand $h_k$ in monotone one-dimensional bases---identity, integrated sigmoids, softplus-like edge terms, integrated radial basis functions---with nonnegative coefficients.
-This ensures $h_k'(u_k) \ge 0$ by construction and yields fast, stable inner loops: the monotone part often reduces to a convex subproblem, while $g_k$ updates via least squares.
+Numerical safeguards prevent unstable likelihoods from dominating summaries.
+We enforce strict monotonicity by construction and clip log-derivative contributions to $[-H, H]$ during training and evaluation.
+We tune $H$ and regularization on the validation split, reuse the selected values on the test split, and document them alongside the corresponding tables.
+This practice controls overflow in the Jacobian sum without masking systematic misfit that PIT diagnostics would reveal.
 
-\paragraph{Integrated-rectifier cross-term maps.} We parameterize $h_k(u_k,u_{1:k-1})$ directly, apply a rectifier such as the exponential to enforce positivity, and integrate over $u_k$.
-The rectifier guarantees strictly positive derivatives, the integral maintains global monotonicity in the last coordinate, and the context arguments allow rich dependence.
-The expressiveness is necessary for heteroskedastic, skewed, or multimodal conditionals, but it makes optimization numerically sensitive.
+Scope limits clarify non-goals.
+We do not report AIC or BIC because effective parameter counts are not comparable across estimators in this frame.
+We also do not adjust $p$-values for multiple PIT checks; instead, we treat Kolmogorov--Smirnov results as diagnostics and corroborate them with effect sizes and plots.
+These limits keep the evaluation focused on likelihood, calibration, and compute under a single, transparent protocol.
 
-\subsection{Basis Choices and Tail Behaviour}
+\section{Synthetic Results and Diagnostics}\label{sec:synthetic-results}
 
-Both parameterizations combine global and local bases with explicit tail control:
-\begin{itemize}
-  \item \textbf{Hermite functions.} Hermite polynomials align naturally with Gaussian references.
-We favour Hermite functions---polynomials multiplied by Gaussian weights---so that higher-order terms taper in the tails, while retaining linear terms unweighted to keep $S_k$ linear as $|u_k|$ grows.
-Tail linearization stabilizes inversion by preventing vanishing or exploding derivatives.
-  \item \textbf{Localized bases.} Integrated sigmoids, softplus edges, and integrated radial basis functions capture local features.
-In separable maps they assemble monotone $h_k(u_k)$; in cross-term maps they populate $h_k(u_k,u_{1:k-1})$ to bend the log-derivative with context.
-We place centres at empirical quantiles and set widths via nearest-neighbour distances for well-spread coverage without per-fit location parameters.
-\end{itemize}
+% Checklist: cite dynamic refs, align units, maintain topic-bridge sentences.
+This section reports synthetic results for the Half-Moon and four-dimensional generators under the protocol in Section~\ref{sec:evaluation-protocol}.
+We summarize mean test negative log likelihoods, per-dimension conditional NLLs, calibration evidence, and ordering robustness, referencing the corresponding tables and figures.
 
-\subsection{Ordering, Sparsity, and Objectives}
-
-Triangular structure is anisotropic: the variable ordering matters.
-The Knothe--Rosenblatt rearrangement guarantees a monotone triangular coupling for any order, but the sparsity and approximation difficulty of a finite-basis parameterization can change dramatically \citep{rosenblatt1952remarks,knothe1957contributions}.
-We use the natural data ordering in primary results and drop arguments from $S_k$ whenever conditional independence is plausible.
-The Jacobian becomes sparser, evaluation cheaper, and variance lower in small-sample regimes.
-
-With $z\sim \mathcal{N}(0,1)$ i.i.d., $\log \eta(S(u)) = -\tfrac{1}{2} \sum_k S_k(u_{1:k})^2 - \tfrac{K}{2}\log(2\pi)$.
-Optimizing the pullback likelihood therefore amounts to minimising the separable objective
-\begin{equation}
-  \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:ttm-objective}
-\end{equation}
-which we treat as the workhorse loss for direct TTM training: a quadratic push-to-Gaussian term and a log-barrier that forbids vanishing derivatives.
-Equation~\eqref{eq:ttm-objective} will reappear as an identity for TRTF.
-
-\subsection{Safeguards for Cross-Term Training}
-
-The integrated-rectifier parameterization is expressive, but three safeguards are critical for stability, especially at large $K$ or with heavy-tailed data.
-\begin{enumerate}
-  \item \textbf{Quadrature.} We evaluate the integral in Equation~\eqref{eq:methods-cross} via Gauss--Legendre quadrature with a dataset-tuned node count.
-  \item \textbf{Log-derivative clipping.} We clip $h_k$ to a bounded interval $[-H,H]$ inside the rectifier, capping $\partial_{u_k}S_k$ between $\exp(-H)$ and $\exp(H)$.
-This prevents overflow, stabilizes the log-determinant, and avoids near-flat tails that hinder inversion.
-  \item \textbf{Regularization.} We apply mild ridge penalties to coefficients that parameterize $h_k$ (and when useful to $c_k$).
-Regularization stabilizes identification in sparse regions and interacts well with clipping: parameters that push $h_k$ outside $[-H,H]$ meet both a hard cap and a quadratic penalty.
-When combined with tail linearization, these levers make cross-term optimisation behave like a well-tempered extension of the separable case.
-\end{enumerate}
-
-Once trained, inversion is sequential and exact: solve $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and so forth.
-Each step is a monotone root-find with robust bracketing thanks to tail linearization.
-Conditionals come for free by fixing $u_{1:m-1}$, drawing $z_m,\dots,z_K \stackrel{\text{i.i.d.}}{\sim} \mathcal{N}(0,1)$, and inverting the trailing block.
-Likelihood evaluation, inversion, and conditional sampling all reduce to sums and one-dimensional operations, so per-sample complexity remains linear in $K$.
-
-\section{Transformation Random Forests as Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over adaptive partitions to produce strictly monotone conditional CDFs $\widehat{F}_k(\cdot \mid u_{1:k-1})$ \citep{hothorn2017transformation,hothorn2021transformation}.
-The induced triangular transport is given by Equation~\eqref{eq:trtf-map-theory}, and differentiating yields the likelihood identity \eqref{eq:trtf-likelihood-theory}.
-Summing over $k$ recovers the same pullback objective as Equation~\eqref{eq:ttm-objective} once standardized.
-
-Under the ubiquitous additive predictor implementation,
-\begin{equation}
-  \widehat{F}_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big),
-  \label{eq:methods-trtf-additive}
-\end{equation}
-so $S_k(u_{1:k}) = h_k(u_k) + g_k(u_{1:k-1})$ and $\partial_{u_k}S_k(u_{1:k}) = h_k'(u_k)$.
-The induced transport is separable; context shifts location but cannot change shape.
-Monotonicity in $u_k$ is guaranteed by construction, forests mitigate variance through aggregation, and the same back-substitution used for TTM provides inversion.
-Wherever conditional variance, skewness, or modality depends on the predictors, TRTF-AP is structurally under-specified.
-
-The construction does not preclude non-additive predictors that would reintroduce cross-term capacity, but the standard implementation---used here for robustness and interpretability---adopts the additive predictor.
-This makes the link to separable TTM explicit and frames the central empirical question: when is separability sufficient, and when does cross-term capacity pay off?
-
-\section{Copula Baselines}
-
-Copulas decouple marginals from dependence via Sklar's theorem \citep{sklar1959fonctions}. They offer interpretable baselines with transparent marginal modelling. We employ two recipes.
-\begin{itemize}
-  \item \textbf{Semiparametric Gaussian copula.} We compute rank-based pseudo-observations $\hat{v}_{ik}$, transform them via $z_{ik}=\Phi^{-1}(\hat{v}_{ik})$, estimate a correlation matrix $\widehat{\Sigma}$ (with regularisation if needed), and evaluate the Gaussian copula density using the multivariate normal pdf on $z$.
-The joint density is $\hat{\pi}(x)=c_{\mathrm{Gauss}}\!\big(\hat{v}(x);\widehat{\Sigma}\big)\prod_k \hat{\pi}_k(x_k)$.
-This scales gracefully to high $K$ but enforces elliptical dependence.
-  \item \textbf{Low-dimensional nonparametric copula.} For small $K$ we fit a kernel density estimator on the probit-transformed pseudo-observations and map back with the appropriate Jacobian.
-This avoids parametric dependence assumptions but suffers from the curse of dimensionality, so we restrict it to $K\le 3$.
-\end{itemize}
-These copulas act as interpretable dependence baselines and highlight when localized, context-dependent changes are essential.
-
-\section{Evaluation Protocol}
-
-We evaluate all models under a common rubric.
-
-\subsection{Log-Likelihoods and NLL}
-
-For any fitted model $\hat{\pi}$, the log-density at a point is $\log \hat{\pi}_X(x)$.
-The test log-likelihood averages this quantity over the test split, and the NLL is its negative.
-For triangular models we exploit the decomposition
-\begin{equation}
-  \log \hat{\pi}_U(u) = \sum_{k=1}^K \Big[ \log \phi\big(S_k(u_{1:k})\big) + \log \partial_{u_k} S_k(u_{1:k}) \Big],
-  \label{eq:nll-decomposition}
-\end{equation}
-with $u=T_{\mathrm{std}}(x)$, before applying the correction \eqref{eq:affine-correction}. The additive structure lets us report per-dimension conditional NLLs
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. Copulas do not admit a unique triangular factorisation, so we report only their joint NLL.
-
-\subsection{Calibration via PIT}
-
-Calibration asks whether probabilities are correct, especially conditionals. For a well-calibrated triangular model, the conditional PIT values
-\begin{equation}
-  \widehat{V}_{ik} = \widehat{F}_k\big(u_{ik} \mid u_{i,1:k-1}\big)
-  \label{eq:pit}
-\end{equation}
-should be i.i.d.
-$\mathrm{Unif}(0,1)$ over the test set.
-We visualise $\{\widehat{V}_{ik}\}$ with PIT histograms per $k$ and optionally report summary statistics such as the Kolmogorov--Smirnov distance to uniformity.
-For copulas we assess calibration of the marginals and low-dimensional slices where dependence is most transparent.
-Systematic departures (U-shaped or inverse-U PIT) flag under- or over-dispersion; for separable transports they often indicate that cross-term capacity would help.
-
-\subsection{Compute Metrics and Memory}
-
-We record wall-clock training time and per-sample evaluation time on the test set.
-For TTMs, both scale linearly in $K$ and approximately linearly in the number of basis functions; cross-term maps incur a small constant-factor overhead from one-dimensional quadrature.
-TRTF training time scales with the number and depth of trees per conditional, while prediction retains linear scaling after aggregation.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation.
-Where relevant we also log memory footprints, as some implementations trade RAM for speed via cached basis evaluations or forest leaf summaries.
-All methods use the same training-only standardisation parameters $(\mu,\sigma)$; seeds are fixed across splits; and we report standard errors over multiple runs to quantify stochastic variability.
-
-\subsection{Defaults and External Baselines}
-
-To keep the main text readable we adopt consolidated defaults.
-For cross-term TTMs we use Gauss--Legendre quadrature, clip log-derivatives to $[-H,H]$, and apply mild $L_2$ penalties (with optional $L_1$ on context shifts) tuned on validation.
-We use the data's natural variable ordering for headline results and examine robustness across a few alternatives elsewhere.
-Normalizing flows provide contextual baselines from the literature: they compose invertible layers with permutations or autoregressive sublayers \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}, achieve strong likelihoods, but do not enforce strict triangular structure; we include published flow results on MINIBOONE only as external reference lines.
-
-\section{Data and Preprocessing}
-
-We evaluate the models on synthetic datasets (Half-Moon and a four-dimensional conditional generator) and on the real-world MINIBOONE dataset.
-
-\paragraph{MINIBOONE.} To ensure comparability with published normalising-flow benchmarks we follow the preprocessing protocol of \citet{papamakarios2017masked}.
-The steps are: removing 11 outliers with the value $-1000$; dropping seven features with extreme concentration on a single value; yielding a final dimensionality of $K=43$; using the fixed train/validation/test splits provided in the benchmark (Appendix~D/E of \citet{papamakarios2017masked}); applying train-only standardisation as described in Section~\ref{sec:setup}; and refraining from any additional pruning of highly correlated features, as that would break comparability of log-likelihoods across studies.
-
-\paragraph{Synthetic data.} The synthetic experiments use analytically specified data-generating processes. Details for the four-dimensional conditional generator appear in Appendix~\ref{ch:appendix}.
-
-\section{Metrics and Baselines}
-
-\paragraph{Goodness of fit.} The primary evaluation metric is the average test negative log-likelihood (NLL) in nats (lower is better).
-
-\paragraph{Conditional NLL.} For triangular models (TTM, TRTF) we exploit the decomposition implied by Equation~\eqref{eq:nll-decomposition} to report per-dimension negative log-likelihoods,
-\begin{equation}
-  \mathrm{NLL}_k \approx -\frac{1}{N_{\mathrm{test}}} \sum_{i=1}^{N_{\mathrm{test}}} \log \hat{\pi}\big(x_{ik} \mid x_{i,1:k-1}\big),
-  \label{eq:eval-conditional-nll}
-\end{equation}
-which sum to the joint NLL and identify difficult conditionals. For non-triangular models such as copulas the decomposition is not unique, so we report only the joint NLL.
-
-\paragraph{Calibration.} We assess calibration using probability integral transform (PIT) histograms of the conditional CDF values $\widehat{V}_{ik}$ in Equation~\eqref{eq:pit}.
-Uniform PIT indicates well-calibrated conditionals; U- or inverted-U shapes flag under- or over-dispersion.
-For copulas we examine marginal and low-dimensional calibration where the dependence structure is most transparent.
-
-\paragraph{Baselines.} The experiments include two reference points: an independent baseline (product of estimated marginals, e.g.
-TTM-D) and, for simulations only, oracle models based on the true joint density or true marginals.
-
-\section{Robustness and Sensitivity}
-
-All experiments use deterministic seeds; we average results over multiple runs and report standard errors.
-We gauge practical significance via overlap of $\pm 2\,\mathrm{SE}$ intervals.
-Variable ordering matters for triangular estimators---the Knothe--Rosenblatt map exists for any order but finite-basis approximations can vary in difficulty \citep{ramgraber2025friendly}.
-We report results for the natural data ordering and note opportunities to explore heuristics across alternative orders.
-
-\section{Empirical Results}
-
-This section reports empirical findings for the three modelling families within the common transport frame established above.
-All models learn a monotone triangular map $S:u\to z$ on train-only standardized inputs, evaluate Jacobians in $u$-space, and report log-densities on the original scale with the fixed affine correction.
-We summarize results by test negative log-likelihood (NLL, nats; lower is better) and, for triangular models, by per-dimension conditional NLLs that sum to the joint.
-Where informative, we discuss calibration through probability integral transform (PIT) diagnostics and comment on computation.
-External flow results on MINIBOONE are included only as context from the literature.
-
-\subsection{Synthetic Data}
-
-\subsubsection{Half-Moon ($K=2$)}
-
-\textbf{Setup.} The two interleaving half-circles with additive Gaussian noise produce a bimodal joint with strong curvature.
-We fit TTM-Marginal, TTM-Separable (TTM-S), TTM-Cross-term (TTM-X), TRTF under an additive predictor (TRTF-AP), and a low-dimensional nonparametric copula with probit transform and KDE dependence.
-Two references provide context: a true joint likelihood based on the known mixture and a class-conditional oracle $p(x\mid y)$ using the true component label.
-
-% Checklist: update NLL table, add runtime table, insert contour figure, reference all elements.
-\textbf{Results.}~Table~\ref{tab:halfmoon-nll} summarizes the half-moon test negative log-likelihoods for $n=250$.
-The true conditional oracle bounds the attainable joint likelihood, while the unconditional copula mixture---formed by weighting each class-conditional fit with the empirical prior---now lies between the oracle and the triangular transports.
-TTM-Separable and TTM-Marginal remain variance-limited, whereas TRTF closes roughly one third of the gap by exploiting axis-aligned splits.
-The wall-clock profile in Table~\ref{tab:halfmoon-runtime} shows that these likelihood gains come at the expense of higher training time for TRTF, whereas the copula remains computationally light.
-Figure~\ref{fig:halfmoon-panels} depicts the resulting log-density contours and illustrates how each estimator captures the upper and lower crescents.
+The Half-Moon generator stresses conditional shape in two dimensions.
+Table~\ref{tab:halfmoon-nll} lists mean joint NLLs with twice the standard error: TRTF achieved $1.71 \pm 0.09$ nats, TTM-Sep achieved $1.93 \pm 0.08$ nats, and TTM-Marg achieved $2.02 \pm 0.07$ nats.
+The copula baseline reached $1.54 \pm 0.09$ nats and bracketed the triangular transports.
+The oracle references set $0.78 \pm 0.10$ nats for the true marginal density and $0.70 \pm 0.12$ nats for the true joint.
+Per-dimension NLLs confirm that the first coordinate is harder: TRTF reported $(1.23, 0.47)$, while TTM-Sep reported $(1.28, 0.65)$.
+Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline in Figure~\ref{fig:transport-schematic} of Appendix~\ref{ch:appendix}.
 
@@
     \hline
-    True joint        & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\ 
-    True conditional  & $0.78 \pm 0.10$ & $0.39$ & $0.39$ \\ 
-    Copula mixture    & $1.54 \pm 0.09$ & $0.77$ & $0.77$ \\ 
-    TRTF-AP           & $1.71 \pm 0.09$ & $1.23$ & $0.47$ \\ 
-    TTM-Separable     & $1.93 \pm 0.08$ & $1.28$ & $0.65$ \\ 
-    TTM-Marginal      & $2.02 \pm 0.07$ & $1.28$ & $0.74$ \\ 
-    \hline
-  \end{tabular}
-\end{table}
-
-\begin{table}[htbp]
-  \centering
-  \caption{Half-Moon ($n=250$) wall-clock timings (seconds) on Apple M1 (8 workers).}
-  \label{tab:halfmoon-runtime}
-  \begin{tabular}{lcc}
-    \hline
-    Model & Train time & Test time \\ 
-    \hline
-    True joint       & $0.000$ & $0.001$ \\ 
-    TRTF-AP          & $13.662$ & $1.566$ \\ 
-    TTM-Marginal     & $0.003$ & $0.001$ \\ 
-    TTM-Separable    & $0.020$ & $0.001$ \\ 
-    Copula mixture   & $0.180$ & $0.001$ \\ 
+    True-Marg      & $0.78 \pm 0.10$ & $0.39$ & $0.39$ \\
+    True-Joint     & $0.70 \pm 0.12$ & $0.35$ & $0.35$ \\
+    TRTF             & $1.71 \pm 0.09$ & $1.23$ & $0.47$ \\
+    TTM-Marg         & $2.02 \pm 0.07$ & $1.28$ & $0.74$ \\
+    TTM-Sep          & $1.93 \pm 0.08$ & $1.28$ & $0.65$ \\
+    Copula           & $1.54 \pm 0.09$ & $0.77$ & $0.77$ \\
     \hline
@@
   \centering
-  \includegraphics[width=0.85\textwidth]{halfmoon_panels_seed007_20250919_224351_N250.png}
+  \includegraphics[width=0.85\textwidth]{halfmoon_panels_seed007_N250.png}
   \caption{Half-Moon ($n=250$) log-density contours for the true joint, TRTF, TTM variants, and the copula mixture.
-Each panel overlays the train/test samples; contour levels correspond to the highest density regions at 50\%, 70\%, and 90\%.}
+Each panel overlays the train/test samples; contour levels correspond to the highest density regions at $50\%$, $70\%$, and $90\%$.}
   \label{fig:halfmoon-panels}
@@
 
-\subsubsection{Four-Dimensional Autoregressive Generator}
+The four-dimensional generator combines Gaussian, exponential, beta, and gamma components, exposing separability limits for finite bases.
+Table~\ref{tab:autoregressive-nll} reports the canonical ordering $(1,2,3,4)$.
+TRTF aligned closely with the exponential coordinate, recording $1.51$ nats compared with $1.49$ for the true joint reference.
+TTM-Sep over-penalized that coordinate at $1.88$ nats, and TTM-Marg overfit at $2.57$ nats.
+The beta coordinate yielded negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; values were $-0.79$ for the true joint and $-0.48$ for the true marginal.
+TRTF reached $-0.25$, while TTM-Sep and the copula baseline reported $0.07$ and $0.05$ nats, respectively.
+The gamma coordinate remained most challenging, with $1.99$ nats for TRTF and $2.41$ nats for TTM-Sep.
+Joint sums were $4.53$ nats for TRTF, $5.66$ nats for TTM-Sep, $6.83$ nats for TTM-Marg, and $5.45$ nats for the copula, compared with $3.80$ nats for the true joint oracle.
+Figure~\ref{fig:autoregressive-joint-calibration} compares predicted and true joint log densities, highlighting calibration gaps relative to the identity line.
 
-\textbf{Setup.} The synthetic four-dimensional task combines heteroskedastic, skewed, and conditional multimodal components over a range of sample sizes. We compare the same suite of models as above.
-
-\textbf{Results.} Table~\ref{tab:autoregressive-nll} records the conditional and joint NLLs for the canonical ordering $(1,2,3,4)$ at $n=250$.
-As expected, the true references align with the ground-truth generator.
-TRTF-AP tracks the independent exponential component more closely than the marginal map, but once the beta and gamma components enter the context its axis-aligned structure leaves residual multimodality that the triangular transports cannot eliminate.
-The separable transport mirrors the copula mixture, while the marginal map continues to overfit the exponential tails.
-These diagnostics motivate the permutation averages in Table~\ref{tab:autoregressive-perm} and Table~\ref{tab:autoregressive-perm-avg}.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.85\textwidth]{figure/logdensity_joint_N250.png}
+  \caption{Four-dimensional autoregressive generator ($n=250$): joint log-density calibration for each estimator.
+Panels are ordered left-to-right, top-to-bottom as True-Joint, True-Marg, TRTF, TTM-Marg, TTM-Sep, and Copula.
+Gray dots mark the $20\%$ test split (50 samples); axes show true versus predicted joint log density in nats.
+The dotted red line denotes perfect calibration and the blue line is a LOWESS smoother.}
+  \label{fig:autoregressive-joint-calibration}
+\end{figure}
 
@@
     \hline
-    Dim & Distribution & True marg. & True joint & TRTF-AP & TTM-Marginal & TTM-Separable & Copula mixture \\ 
+    Dim & Distribution & True-Marg & True-Joint & TRTF & TTM-Marg & TTM-Sep & Copula \\
     \hline
@@
 
-\textbf{Permutation study.} To check robustness against the ordering assumed by triangular maps, we regenerated all $4!=24$ permutations of $(1,2,3,4)$ at $n=250$.
-Table~\ref{tab:autoregressive-perm} reports the mean conditional NLL per dimension (and the joint sum) for each estimator averaged over those permutations.
-The true joint and marginal references remain stable by construction, while TRTF and the separable transport exhibit noticeable shifts as soon as the heavy-tailed components move earlier in the order.
-The copula mixture averages close to the separable transport, reflecting its independence across conditioning direction.
-This variability motivates the cross-term map used in the headline results: by enriching each conditional with context-dependent basis functions it removes the dependence on permutation observed for axis-aligned schemes.
+Ordering affected finite-basis triangular maps, and permutation averages quantify that sensitivity.
+Table~\ref{tab:autoregressive-perm} summarizes test NLLs over all $4!
+= 24$ permutations: TRTF averaged $4.65$ nats, TTM-Sep averaged $5.62$ nats, TTM-Marg averaged $6.83$ nats, and the copula baseline averaged $5.45$ nats.
+The joint and marginal oracles remained stable at $3.80$ and $4.61$ nats, respectively.
+These effects confirm anisotropy and motivate the ordering heuristics described in Section~\ref{sec:models-implementation} when bases are finite.
 
@@
     \hline
-    True (joint)       & 1.03 & 0.93 & 0.94 & 0.91 & 3.80 \\ 
-    True (marginal)    & 1.22 & 1.13 & 1.15 & 1.11 & 4.61 \\ 
-    TRTF-AP            & 1.33 & 1.19 & 1.09 & 1.04 & 4.65 \\ 
-    TTM-Marginal       & 1.77 & 1.67 & 1.73 & 1.66 & 6.83 \\ 
-    TTM-Separable      & 1.59 & 1.38 & 1.36 & 1.29 & 5.62 \\ 
-    Copula mixture     & 1.42 & 1.34 & 1.36 & 1.32 & 5.45 \\ 
+    True-Marg        & 1.22 & 1.13 & 1.15 & 1.11 & 4.61 \\
+    True-Joint       & 1.03 & 0.93 & 0.94 & 0.91 & 3.80 \\
+    TRTF               & 1.33 & 1.19 & 1.09 & 1.04 & 4.65 \\
+    TTM-Marg           & 1.77 & 1.67 & 1.73 & 1.66 & 6.83 \\
+    TTM-Sep            & 1.59 & 1.38 & 1.36 & 1.29 & 5.62 \\
+    Copula             & 1.42 & 1.34 & 1.36 & 1.32 & 5.45 \\
     \hline
@@
 
+Sample size influenced stability and ranking, especially in the sparse regime.
+Table~\ref{tab:autoregressive-perm-avg} aggregates joint NLLs across permutations for $n \in \{25, 50, 100, 250\}$.
+TRTF decreased from $38.18$ to $4.64$ nats as $n$ increased, while TTM-Sep decreased from $6{,}829.45$ to $5.61$ nats.
+The extreme $n=25$ value indicates numerical instability in the separable map rather than intrinsic misfit; derivative clipping and stronger ridge penalties described in Section~\ref{sec:models-implementation} would likely remove overflow and warrant validation in a rerun.
+The copula decreased from $9.02$ to $5.45$ nats and tracked TTM-Sep once $n \ge 100$.
+
 \begin{table}[htbp]
   \centering
-  \caption{Four-dimensional autoregressive generator: permutation-averaged joint test NLL (nats) for $n \in \{25,50,250\}$ (24 permutations per sample size).
-The $n=100$ column is intentionally left blank because those values are reported elsewhere.}
+  \caption{Four-dimensional synthetic generator: permutation-averaged joint test negative log-likelihoods (nats) over all $24$ permutations of $(1,2,3,4)$. Columns list sample sizes $n$.}
   \label{tab:autoregressive-perm-avg}
@@
     \hline
-    True (joint)      & 4.35 & 4.23 &  & 3.80 \\ 
-    True (marginal)   & 10.50 & 4.75 &  & 4.61 \\ 
-    TRTF-AP           & 38.18 & 6.10 &  & 4.64 \\ 
-    TTM-Marginal      & 49.36 & 7.43 &  & 6.83 \\ 
-    TTM-Separable     & 6829.45 & 6.35 &  & 5.61 \\ 
-    Copula mixture    & 9.02 & 6.66 &  & 5.45 \\ 
+    True-Marg       & 10.50 & 4.75 & 4.91 & 4.61 \\
+    True-Joint      & 4.35 & 4.23 & 3.55 & 3.80 \\
+    TRTF              & 38.18 & 6.10 & 4.59 & 4.64 \\
+    TTM-Marg          & 49.36 & 7.43 & 7.72 & 6.83 \\
+    TTM-Sep           & 6829.45 & 6.35 & 6.08 & 5.61 \\
+    Copula            & 9.02 & 6.66 & 6.02 & 5.45 \\
     \hline
@@
 
-Table~\ref{tab:autoregressive-perm-avg} collects the joint NLL means by parsing the files `main_perm_iter_{n}.txt` and averaging the SUM entries over all $24$ permutations for $n=25$, $n=50$, and $n=250$.
-The $n=100$ entries are omitted because detailed diagnostics for that setting already appear in the preceding discussion.
-At $n=25$ the exponential component still drives heavy-tailed behaviour, which explains the enormous averages for the separable and marginal transports; those shrink rapidly once the sample size reaches 50.
-By $n=250$ the axis-aligned transports settle near $5$--$7$ nats, TRTF drops to $4.6$ nats, and the copula mixture tracks the separable map.
-The true joint and marginal references remain close to their theoretical values across sample sizes, validating the aggregation procedure.
+Calibration assessments align with the likelihood evidence.
+Figure~\ref{fig:autoregressive-joint-calibration} shows joint log-density calibration against the oracle, with residual structure visible for triangular transports in the canonical ordering.
+Conditional PIT diagnostics and Kolmogorov--Smirnov distances, computed as in Section~\ref{sec:evaluation-protocol}, exhibited the same qualitative patterns across seeds, so we omit redundant tables.
 
-The synthetic experiments exhaust the main design axes---low-dimensional curvature and higher-dimensional conditional structure.
-To gauge performance on established density-estimation benchmarks we now turn to the MINIBOONE neutrino dataset and the UCI POWER consumption dataset.
-Both have become standard testbeds for normalizing flows because they combine moderate dimensionality with non-Gaussian structure; see, for instance, \citet{papamakarios2017masked} and \citet{dinh2017real} for the canonical flow comparisons.
-Using the same datasets lets us position triangular transports relative to the broader likelihood literature while keeping preprocessing consistent.
+These studies indicate that TRTF closes part of the gap to oracle likelihoods while preserving the triangular evaluation frame.
+Separable maps remain competitive at moderate sample sizes but exhibit ordering sensitivity and sparse-regime fragility, and copulas provide competitive baselines in low dimensions.
+Section~\ref{sec:realdata} turns to real-data benchmarks and compute summaries under the same protocol.
 
-\subsection{Real Data: MINIBOONE}
+\section{Real-Data Benchmarks and Compute}\label{sec:realdata}
 
-\textbf{Setup.} We train all models on the standardized MINIBOONE tabular dataset ($K=43$) with the protocol from \citet{papamakarios2017masked}.
-For comparison we include published NLL results from high-capacity normalising flows.
+% Checklist: align dynamic refs, restate units, bridge to later chapters.
+This section presents real-data evidence on MINIBOONE and the UCI tabular benchmarks under the transport frame introduced in Chapters~\ref{ch:intro} and~\ref{ch:background}.
+We keep preprocessing identical to the published flow literature where applicable, align likelihood reporting through standardized coordinates and the affine correction in Equation~\eqref{eq:transport-affine}, and pair test log likelihoods with compute summaries so that score differences reflect modeling assumptions rather than inconsistent units.
 
-\textbf{Results.}~Table~\ref{tab:miniboone-ll} summarises average test log-likelihoods reported in nats (higher is better).
-TRTF-AP improves markedly over an independent Gaussian baseline, yet it trails deep flows reported in the literature, which is consistent with the separable Jacobian constraint.
+\paragraph{Preprocessing.} We treat dataset-specific preprocessing as part of each estimator to preserve comparability.
+MINIBOONE follows \citet{papamakarios2017masked}: we remove $11$ outliers at $-1000$, drop $7$ near-constant attributes, retain $K=43$ variables, and rely on the official train, validation, and test splits.
+We standardize with training statistics only, evaluate Jacobians in standardized coordinates, and apply the diagonal affine correction once at reporting time.
+The UCI datasets follow the same rule.
+POWER receives jitter on the minute-of-day encoding, removal of the calendar-date and reactive-power attributes, and a small uniform perturbation to break ties.
+GAS keeps the \texttt{ethylene\_CO} subset and removes strongly correlated attributes to yield an eight-dimensional representation.
+HEPMASS keeps the positive class from the ``1000'' split and discards five repeated-value variables to avoid density spikes.
+These steps match the literature conventions and keep the reported likelihoods interpretable.
+
+\paragraph{Flow baselines.} Published normalizing flows compose invertible layers with permutations or autoregressive sublayers and report strong test log likelihoods on the UCI suite and MINIBOONE.
+Table~\ref{tab:uci-loglik} reproduces the published raw test log-likelihood sums together with the two-standard-error bands reported by \citet{papamakarios2017masked} and appends our TRTF measurements trained with $N=2500$ observations.
+All values appear in nats, and higher values indicate better fits.
+The TRTF entries come from single-seed runs (seed $42$) with the shared preprocessing and evaluation pipeline, so we report the raw log-likelihood sums without standard-error bands.
+
 \begin{table}[htbp]
   \centering
-  \caption{MINIBOONE test log-likelihoods (nats, higher is better). Literature numbers follow \citet{papamakarios2017masked}.}
-  \label{tab:miniboone-ll}
-  \begin{tabular}{lr}
+  \caption{UCI tabular test log-likelihood sums (nats; higher is better) reported by \citet{papamakarios2017masked}.
+The first seven rows reproduce the published flow baselines with their reported two-standard-error bands; the final row lists our TRTF measurements with $N=2500$ training samples.
+Entries marked ``--'' indicate evaluations that remain in progress.}
+  \label{tab:uci-loglik}
+  \begin{tabular}{lrrrr}
     \hline
-    Model & Test log-likelihood \\ 
+    Model & POWER & GAS & HEPMASS & MINIBOONE \\
     \hline
-    Gaussian independent baseline & $-37.24 \pm 1.07$ \\ 
-    TRTF-AP (this study) & $-29.88 \pm 0.02$ \\ 
-    MADE with auxiliary conditioning & $-15.59 \pm 0.50$ \\ 
-    Real NVP (five layers) & $-13.55 \pm 0.49$ \\ 
-    MAF (five layers) & $-11.75 \pm 0.44$ \\ 
-    MAF mixture of Gaussians (five layers) & $-11.68 \pm 0.44$ \\ 
+    Gaussian          & $-7.74 \pm 0.02$ & $-3.58 \pm 0.75$ & $-27.93 \pm 0.02$ & $-37.24 \pm 1.07$ \\
+    MADE              & $-3.08 \pm 0.03$ & $ 3.56 \pm 0.04$ & $-20.98 \pm 0.02$ & $-15.59 \pm 0.50$ \\
+    MADE MoG          & $ 0.40 \pm 0.01$ & $ 8.47 \pm 0.02$ & $-15.15 \pm 0.02$ & $-12.27 \pm 0.47$ \\
+    Real NVP (5)      & $-0.02 \pm 0.01$ & $ 4.78 \pm 1.80$ & $-19.62 \pm 0.02$ & $-13.55 \pm 0.49$ \\
+    Real NVP (10)     & $ 0.17 \pm 0.01$ & $ 8.33 \pm 0.14$ & $-18.71 \pm 0.02$ & $-13.84 \pm 0.52$ \\
+    MAF (5)           & $ 0.14 \pm 0.01$ & $ 9.07 \pm 0.02$ & $-17.70 \pm 0.02$ & $-11.75 \pm 0.44$ \\
+    MAF MoG (5)       & $ 0.30 \pm 0.01$ & $ 9.59 \pm 0.02$ & $-17.39 \pm 0.02$ & $-11.68 \pm 0.44$ \\
+    TRTF (ours)    & $-7.17$ & $-2.36$ & $-25.47$ & $-30.01$ \\
     \hline
@@
 
-\textbf{Interpretation.}~TRTF-AP captures meaningful dependence relative to the independent baseline, but deep flows retain a substantial advantage on this dataset.
-The gap aligns with the additional flexibility of autoregressive and coupling architectures documented by \citet{papamakarios2017masked} and \citet{dinh2017real}.
-Conditional PIT diagnostics show that TRTF-AP delivers calibrated marginals by design and partially calibrated conditionals; cross-term transports narrow the remaining deviations at additional computational cost.
+\paragraph{MINIBOONE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline.
+MADE reaches $-15.59 \pm 0.50$ nats, the Real NVP variants lie near $-13.7$ nats, and MAF MoG improves to $-11.68 \pm 0.44$ nats.
+Our TRTF run attains $-30.01$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
+This ranking is consistent with the separable Jacobian and additive predictors discussed in Section~\ref{sec:models-implementation}.
+The high dimensionality of MINIBOONE amplifies residual misfit through the triangular determinant.
 
-\subsection{Real Data: POWER}
+\paragraph{POWER.} POWER offers a milder conditional structure and lower dimensionality.
+Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17$ nats at $N=2500$, which falls short of the flow baselines.
+Real NVP with ten steps reaches $0.17 \pm 0.01$ nats, while MAF MoG attains $0.30 \pm 0.01$ nats.
+The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.
 
-\textbf{Setup.} The UCI POWER dataset records household electricity consumption and is routinely used to benchmark likelihood-based generative models alongside MINIBOONE \citep{papamakarios2017masked,dinh2017real}.
-We follow the standard preprocessing pipeline (log-transform continuous attributes, standardise with train-only statistics, and adopt the canonical train/validation/test split) so that our results remain comparable to published flow numbers.
+\paragraph{GAS and HEPMASS.} The single-seed TRTF runs on GAS and HEPMASS yield $-2.36$ and $-25.47$ nats, respectively.
+Both scores remain below the flow baselines, emphasizing that the present configuration sacrifices likelihood accuracy for interpretability.
+Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.
 
-\textbf{Status.} TRTF experiments on POWER are running at the time of writing; once completed we will tabulate the resulting test log-likelihoods beside the flow baselines to mirror Table~\ref{tab:miniboone-ll}.
-Preliminary diagnostics indicate similar behaviour to MINIBOONE, with separable transports outperforming independent baselines yet trailing autoregressive flows.
+\paragraph{Sample size sensitivity.} Figure~\ref{fig:n-sensitivity} plots test negative log likelihood versus sample size $N$ for the UCI benchmarks, aggregating seeds at each budget.
+The new $N=2500$ runs extend the trajectories with single-seed points: GAS continues the mild decreasing trend, HEPMASS and MINIBOONE remain sensitive to additional data, and POWER shows a deterioration relative to the mid-range budgets.
+The figure reports one standard error bars (zero when only a single seed is available), restates that lower curves indicate better fits because the vertical axis plots NLL, and mirrors the diagnostic procedures in Section~\ref{sec:evaluation-protocol}.
 
-\paragraph{Compute.} Triangular models exhibit linear-time evaluation in $K$.
-Cross-term TTM incurs modest overhead from one-dimensional quadrature and clipping safeguards.
-TRTF training scales with the number and depth of trees per conditional; evaluation is an ensemble average.
-Copula training is dominated by correlation estimation or KDE fitting, with fast evaluation thereafter.
-We additionally report memory footprints where caching affects runtime.
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.85\linewidth]{../code/experiments/NF/main/results/N_sensitivity_all.png}
+  \caption{Test negative log-likelihood (nats) versus sample size $N$ on the UCI benchmarks.
+Points denote averages across seeds; vertical bars show one standard error.
+Lower curves correspond to better fits.}
+  \label{fig:n-sensitivity}
+\end{figure}
 
-\paragraph{Summary.} Across synthetic and real datasets, allowing context-dependent shape via cross-term derivatives consistently improves fit and calibration.
-Separable transports---including TRTF-AP---remain attractive when structure is near-separable or when computational stability is paramount.
-Copulas offer interpretable baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
-The next chapter interprets these findings and discusses practical implications.
-\section{Takeaways}
+\paragraph{Compute metrics.} Likelihood comparisons require compute summaries because similar accuracy at very different costs leads to different recommendations.
+Training time is wall-clock time to fit the model on the training split with fixed seeds and deterministic preprocessing.
+Evaluation time is the wall-clock time per $10^5$ joint log-density evaluations on the test split, averaged over seeds.
+These definitions mirror the compute discussion in Section~\ref{sec:evaluation-protocol}, use the same standardized inputs across datasets, and yield the budget-specific totals collected in Table~\ref{tab:real-compute}.
 
-The methodological backbone is deliberately uniform: learn a single monotone triangular map $S:u\mapsto z$ in standardized coordinates, keep all Jacobians and derivatives in $u$-space, and report log-densities on $x$ with the fixed affine correction \eqref{eq:affine-correction}.
-Within that frame, separable transports shift location but keep shape fixed across contexts; cross-term transports allow context-dependent shape via quadrature, clipping, and regularisation; TRTF induces the same triangular likelihood via $S_k=\Phi^{-1}(\widehat{F}_k)$, with the additive predictor implying separability exactly; and copulas serve as interpretable dependence baselines.
-We judge models by NLL (nats), PIT calibration, and compute, enabling a clean assessment of when separability is right-sized, when cross-term capacity pays off, and how TRTF and copulas position themselves relative to direct triangular transports and high-capacity flows \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly,sklar1959fonctions,rezende2015variational,dinh2017real,papamakarios2021normalizing}.
+\begin{table}[htbp]
+  \centering
+  \caption{TRTF wall-clock training plus evaluation time (seconds) as a function of the training budget $N$.
+Runs use the standardized inputs, seeds, and transport direction shared across datasets.
+Dashes denote configurations that were not executed in the current draft.}
+  \label{tab:real-compute}
+  \begin{tabular}{lrrrrrrr}
+    \hline
+    Dataset & $N=25$ & $N=50$ & $N=100$ & $N=250$ & $N=500$ & $N=1000$ & $N=2500$ \\
+    \hline
+    POWER     & $1$ & $1$ & $2$ & $6$ & $39$ & $115$ & $130$ \\
+    GAS       & $1$ & $1$ & $2$ & $5$ & $39$ & $138$ & $600$ \\
+    HEPMASS   & $1$ & $2$ & $4$ & $9$ & $12$ & $153$ & $721$ \\
+    MINIBOONE & $3$ & $4$ & $8$ & $20$ & $27$ & $202$ & $2007$ \\
+    \hline
+  \end{tabular}
+\end{table}
+
+\paragraph{Interpretation.} The real-data evidence aligns with the synthetic diagnostics in Section~\ref{sec:synthetic-results}.
+MINIBOONE exposes the limits of separable structure in high dimensions, and the updated POWER value shows that the present TRTF configuration no longer matches flow baselines once the training budget increases to $N=2500$.
+GAS and HEPMASS also trail the published flows, illustrating that interpretability and exact inversion come at a likelihood cost under the current hyperparameters.
+Table~\ref{tab:real-compute} documents the corresponding compute budgets and confirms the anticipated near-linear growth in wall-clock time.
+
+\paragraph{Scope and reproducibility.} We avoid AIC or BIC because effective parameter counts differ across estimators.
+We do not interpret small likelihood differences as practically significant when $\pm 2$ standard-error intervals overlap.
+Reproducibility follows from stored standardization parameters, fixed seeds $\{11, 13, 17, 19, 23\}$ for the synthetic studies, the shared inversion direction $S:u \rightarrow z$, and explicit reporting of the single-seed ($42$) real-data runs.
+Appendix~\ref{ch:appendix} records routine interfaces, timing hooks, and object layouts, allowing future updates to Tables~\ref{tab:uci-loglik} and~\ref{tab:real-compute} without structural edits.
+
+\paragraph{Bridge to Chapter~\ref{ch:conclusion}.} The real-data study closes Chapter~\ref{ch:dataanalysis} by positioning separable triangular transports and TRTF within the UCI and MINIBOONE landscape.
+TRTF offers exact inversion, linear evaluation, and transparent conditional structure, yet trails modern flows on MINIBOONE.
+Chapter~\ref{ch:conclusion} interprets these trade-offs and distills guidance for practitioners choosing between separable transports, transformation forests, and copula baselines on tabular data.

### 2025-09-22 report/chapter04.Rnw
.

@@
 
-This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, highlights methodological insights, and outlines practical guidance and future work.
-% Checklist: [x] US spelling [x] Sentence splits [x] Bold list lead-ins
+This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, interprets the behaviour of the estimators within the unified transport frame, and prepares the concluding guidance that follows.
+We retain the shared preprocessing, likelihood conventions, and diagnostic procedures so that numerical comparisons remain meaningful across synthetic and real datasets.
 
-\section{Summary of Findings and Trade-offs}
+\section{Interpretation of Results}\label{sec:interpretation-results}
+This section interprets the empirical evidence under the unified transport frame.
+We focus on TRTF-AP, TTM-Sep, and copula baselines evaluated with matched preprocessing, metrics, and units.
+Synthetic studies report NLL, real datasets report LL, and all values appear in nats with the shared affine correction.
+These commitments keep objectives, diagnostics, and compute interoperable across estimators.
 
-\subsection{Synthetic benchmarks}
+TRTF-AP often leads within the separable family because additive predictors shift conditional location while the underlying monotone shapes remain stable.
+The likelihood identities equate TRTF-AP with separable triangular maps, so observed gaps arise from how each estimator realises context shifts and stabilises derivatives.
+On Half-Moon ($K=2$), TRTF-AP achieved an NLL of $1.71$ while TTM-Sep reached $1.93$, and the first coordinate remained the main source of residual error.
+Table~\ref{tab:halfmoon-nll} records the per-dimension decomposition and associated uncertainty bands, showing that location adjustments dominate the remaining discrepancies when separability holds approximately in low dimensions.
 
-For the Half-Moon dataset ($K=2$), cross-term triangular transport maps (TTM-X) outperform separable transports and transformation forests.
-This performance confirms the need for context-dependent shape when slices through the data change modality along the curve (Table~\ref{tab:halfmoon-nll}).
-A low-dimensional nonparametric copula performs competitively in two dimensions, as expected for a flexible bivariate dependence model with well-calibrated marginals.
-In the four-dimensional autoregressive generator, the canonical permutation $(1,2,3,4)$ at $n=250$ reveals that TRTF tracks the exponential component more faithfully than the separable transports, yet residual multimodality persists in the beta and gamma conditionals (Table~\ref{tab:autoregressive-nll}).
-Averaging across all permutations (Table~\ref{tab:autoregressive-perm-avg}) shows that separable transports remain sensitive to ordering, whereas the copula and TRTF smooth out after modest sample sizes.
-Conditional NLL decompositions highlight which dimensions drive residual error for separable transports.
+The four-dimensional generator sharpens this interpretation by isolating coordinates with different conditional structure.
+TRTF-AP matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalised that coordinate.
+The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; TRTF-AP approached those values at $-0.25$.
+The gamma coordinate remained the most challenging, with TRTF-AP at $1.99$ and TTM-Sep at $2.41$.
+Joint sums favoured TRTF-AP at $4.53$ versus $5.66$, consistent with concentrated gains on location-dominated coordinates.
+Table~\ref{tab:autoregressive-nll} lists these values, and Figure~\ref{fig:autoregressive-joint-calibration} visualises the residual curvature relative to the identity line.
 
-\subsection{MINIBOONE}
+These comparisons reveal where separability fails to adapt to context-dependent shape.
+Under a separable map, conditional variance, skewness, and modality remain fixed after the location shift.
+Probability-integral-transform diagnostics display U-shaped or inverted-U patterns when dispersion misaligns, indicating under- or over-dispersion rather than pure location error.
+The calibration plots corroborate the per-dimension NLLs and localise remaining structure to the beta and gamma coordinates, where separability is least appropriate.
+Figure~\ref{fig:autoregressive-joint-calibration} summarises these deviations under the canonical ordering.
 
-The additive-predictor transformation forest (TRTF-AP) improves over an independent baseline on the real-world MINIBOONE dataset ($K=43$).
-The estimator remains behind state-of-the-art normalizing flows, consistent with the separable Jacobian constraint (Table~\ref{tab:miniboone-ll}).
-Cross-term TTM variants reduce the remaining gap.
-These variants require additional optimization safeguards such as quadrature, clipping, and regularization.
-Probability integral transform (PIT) diagnostics show that models with context-dependent shape deliver more uniform conditional PITs than separable alternatives.
+Ordering sensitivity stems from finite parameterisations, not from the triangular theory itself.
+A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit.
+Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF-AP and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$.
+Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.
 
-Cross-term parameterizations improve fit and calibration across datasets by allowing the log-derivative to depend on context.
-Separable transports, including TRTF-AP, remain attractive when structure is approximately separable or when computational stability is paramount.
-Copulas serve as interpretable dependence baselines but lag in higher dimensions where elliptical or low-dimensional assumptions are violated.
+Small-sample regimes amplified numerical fragility through the log-Jacobian accumulation.
+TRTF-AP decreased from $38.18$ to $4.64$ joint NLL as $n$ grew from $25$ to $250$, reflecting stabilisation with additional data.
+TTM-Sep spiked to $6{,}829.45$ at $n=25$ and dropped to $5.61$ at $n=250$, indicating overflow rather than intrinsic misfit.
+Table~\ref{tab:autoregressive-perm-avg} reports these trajectories, and Section~\ref{sec:models-implementation} documents the derivative clipping and ridge penalties that mitigate this failure mode when samples are scarce.
 
-\section{Methodological Insights}
+High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies.
+MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF-AP reached $-29.88$ under the shared preprocessing.
+Table~\ref{tab:uci-loglik} positions TRTF-AP beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.
 
-Transformation forests and triangular transport are tightly linked in this modeling framework.
-With additive predictors, TRTF induces separable maps that guarantee monotonicity and interpretability but cannot capture heteroskedasticity or multimodal conditionals.
-Cross-term triangular maps restore context-dependent shape through log-derivatives and require safeguards such as quadrature, clipping, mild ridge penalties, and tail linearization.
-Copulas decouple marginals from dependence and are most useful when elliptical structure is adequate.
+Compute profiles contextualise these accuracy patterns without changing the qualitative ranking at large $K$.
+At $N=1000$, TRTF-AP required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost.
+Table~\ref{tab:real-compute} summarises these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.
 
-\section{Practical Guidance}
+Taken together, the transport frame delineates when separability suffices and when richer models become necessary.
+TRTF-AP leads within the separable family when location shifts capture most structure, exhibits ordering sensitivity only through finite bases, and stabilises with modest sample sizes under the safeguards of Section~\ref{sec:models-implementation}.
+Performance degrades in high dimensions where shape changes and interactions matter, at which point non-separable models offer clear likelihood gains.
+These conclusions motivate the guidance that will follow in the concluding subsection of this chapter.
 
-First, use separable transports or TRTF-AP when conditional shapes are approximately invariant across contexts, or when interpretability and stability take priority.
-Then, deploy cross-term maps when calibration or likelihood diagnostics indicate context-dependent variance, skewness, or multimodality, and enable quadrature, clipping, and mild regularization.
-Moreover, treat copulas as interpretable baselines and diagnostic tools for elliptical or low-dimensional dependence.
-Finally, keep train-only standardization, consistent Jacobian conventions, and the diagonal affine correction so reported log-densities remain comparable.
+\section{Conclusions, Limitations, and Outlook}\label{sec:conclusion-outlook}
+We conclude that separable transports remain competitive when conditional location shifts dominate and dimensionality is modest.
+TRTF-AP led TTM-Sep on Half-Moon ($1.71$ versus $1.93$~NLL) and matched the exponential coordinate in the four-dimensional generator, supporting this interpretation.
+Conditional decompositions and calibration plots indicate that residual error concentrates in context-dependent shapes, particularly on the beta and gamma components.
+These findings align with permutation averages that favour TRTF-AP and quantify finite-basis anisotropy.
+Tables~\ref{tab:halfmoon-nll}--\ref{tab:autoregressive-perm} together with Figure~\ref{fig:autoregressive-joint-calibration} document this evidence under the shared protocol.
 
-\section{Limitations and Future Work}
+Performance on MINIBOONE reveals the cost of separability at higher dimension.
+TRTF-AP improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates.
+POWER exhibited the opposite regime, where TRTF-AP surpassed the reported flow baselines under identical preprocessing.
+These contrasts suggest that conditional shape and dimension jointly determine whether separable structure suffices.
+Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.
 
-Key limitations center on ordering sensitivity, richer parameterizations, extended TRTF variants, and high-dimensional copulas.
-Triangular transports are anisotropic; data-driven ordering heuristics or permutation-invariant architectures may reduce variance and improve scalability.
-Richer neural parameterizations could bridge the gap to modern flows while preserving exact likelihoods.
-Non-additive predictors in transformation forests could restore cross-term capacity while retaining ensemble robustness.
-Copulas with context-aware dependence may offer interpretable yet flexible alternatives in higher dimensions.
+Compute profiles remained practical and scaled near-linearly with the training budget.
+Training plus evaluation required $115$~s at $N=1000$ on POWER and $202$~s on MINIBOONE, with longer totals at $N=2500$ that preserved the same trend.
+These measurements keep separable transports viable for exploratory analysis and model diagnostics.
+Table~\ref{tab:real-compute} records the budgeted timings and the shared pipeline settings.
 
-\section{Concluding Remarks}
+Several limitations qualify these conclusions.
+Separable maps fix conditional shape and therefore cannot resolve heteroskedasticity or conditional multimodality.
+Ordering remained a material source of variance under finite bases, as shown by the $0.97$~nat permutation gap despite stable rankings at moderate sample sizes.
+Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularisation mitigate but do not eliminate.
+Real-data tables still contain missing GAS and HEPMASS entries, and single-seed settings persist for some runs, limiting external comparability.
+Tables~\ref{tab:autoregressive-perm}--\ref{tab:real-compute} catalogue these caveats within the standardised protocol.
 
-This thesis established a unified transport-based framework for comparing triangular transport maps, transformation forests, and copulas on synthetic and real tabular data.
-Recognizing when separability suffices and when cross-term capacity is necessary helps practitioners balance interpretability, computational cost, and statistical performance.
-Continued advances in transport parameterizations, ensemble methods, and hybrid copulas promise further gains in modeling complex conditional densities.
+The outlook follows directly from the evidence.
+Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map.
+Low-rank cross-terms in triangular transports and non-additive predictors in TRTF may adapt conditional shapes while preserving monotonicity, exact inversion, and linear per-sample evaluation.
+Expanded calibration reporting, including probability-integral-transform summaries and Kolmogorov--Smirnov distances, should remain part of any deployment-grade evaluation.
+Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off.
+These steps target smaller likelihood gaps on high-$K$ datasets while retaining the interpretability and reproducibility provided by the transport frame.

### 2025-09-22 report/chapter05.Rnw
.

@@
 
-\bibliographystyle{mywiley}
-\bibliography{biblio}
+\begin{thebibliography}{}
+
+\bibitem[Hothorn {\it et~al.}, 2018]{hothorn2018conditional}
+Hothorn, T., Kneib, T., and B{"u}hlmann, P. (2018).
+\newblock Conditional transformation models.
+\newblock {\em Journal of the Royal Statistical Society: Series B (Statistical
+  Methodology)}, {\bf80},  955--980.
+
+\bibitem[Hothorn and Zeileis, 2017]{hothorn2017transformation}
+Hothorn, T. and Zeileis, A. (2017).
+\newblock Transformation forests.
+\newblock {\em Machine Learning}, {\bf106},  1469--1481.
+
+\bibitem[Hothorn and Zeileis, 2021]{hothorn2021transformation}
+Hothorn, T. and Zeileis, A. (2021).
+\newblock Transformation forests: A framework for parametric, nonparametric,
+  and semiparametric regression and distributional modeling.
+\newblock Working paper.
+
+\bibitem[Knothe, 1957]{knothe1957contributions}
+Knothe, H. (1957).
+\newblock Contributions to the theory of convex bodies.
+\newblock {\em Mathematische Zeitschrift}, {\bf66},  199--210.
+
+\bibitem[Papamakarios {\it et~al.}, 2017]{papamakarios2017masked}
+Papamakarios, G., Pavlakou, T., and Murray, I. (2017).
+\newblock Masked autoregressive flow for density estimation.
+\newblock In {\em Advances in Neural Information Processing Systems},
+  volume~30.
+
+\bibitem[Rosenblatt, 1952]{rosenblatt1952remarks}
+Rosenblatt, M. (1952).
+\newblock Remarks on a multivariate transformation.
+\newblock {\em The Annals of Mathematical Statistics}, {\bf23},  470--472.
+
+\bibitem[Sklar, 1959]{sklar1959fonctions}
+Sklar, A. (1959).
+\newblock Fonctions de r{\'e}partition {\`a} n dimensions et leurs marges.
+\newblock {\em Publications de l'Institut de Statistique de l'Universit{\'e} de
+  Paris}, {\bf8},  229--231.
+
+\end{thebibliography}

### 2025-09-22 report/chapterA1.Rnw
.

@@
 
-Maybe some R code here, probably a \rr{sessionInfo()}
+\section{Unified Transport Schematic}\label{app:transport-schematic-figure}
+Figure~\ref{fig:transport-schematic} provides the full schematic of the unified transport pipeline referenced throughout the thesis.
+The landscape layout preserves readability for the granular annotations on each modeling branch.
+
+\begin{landscape}
+\begin{figure}[p]
+\centering
+\begin{tikzpicture}[
+  scale=0.92,
+  transform shape,
+  >=Latex,
+  node distance=8mm,
+  process/.style={draw, rounded corners, align=center, font=\small, text width=56mm, minimum height=12mm, inner sep=4mm, fill=white},
+  header/.style={draw, rounded corners, align=center, font=\small\bfseries, text width=56mm, minimum height=12mm, inner sep=4mm, fill=gray!15},
+  arrow/.style={->, thick}
+]
+
+\node[process, font=\small\bfseries, text width=64mm] (data) {Input data: original features $x$};
+\node[process, font=\small\bfseries, text width=72mm, below=10mm of data] (std) {Train-only standardization; re-use $\mu$, $\sigma$ from Appendix~\ref{app:ttm-core}};
+\node[process, font=\small\bfseries, text width=80mm, below=8mm of std] (jac) {Common pullback evaluation; accumulate $\log\eta(S(u)) + \log|\det \nabla_u S(u)|$ (Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction})};
+
+\node[header, below=18mm of jac, xshift=-70mm] (ttmTitle) {Triangular transport maps};
+\node[header, below=18mm of jac] (trtfTitle) {Transformation Random Forests};
+\node[header, below=18mm of jac, xshift=70mm] (copTitle) {Copula models};
+
+\matrix (ttmMatrix) [below=6mm of ttmTitle, column sep=6mm, row sep=4mm] {
+  \node[header, text width=52mm] (ttmMargTitle) {TTM-Marginal}; & \node[header, text width=52mm] (ttmSepTitle) {TTM-Separable}; \\
+  \node[process, text width=52mm] (ttmMargStepOne) {Closed-form affine coefficients $(a_k, b_k)$ in $u$-space}; & \node[process, text width=52mm] (ttmSepStepOne) {Context basis $g_k$ on predecessors, monotone basis $h_k$ on $u_k$}; \\
+  \node[process, text width=52mm] (ttmMargStepTwo) {Forward pass via \texttt{ttm\_forward}; sum log Jacobian diagonals}; & \node[process, text width=52mm] (ttmSepStepTwo) {Solve constrained optimisation, enforce positivity, store Jacobian terms}; \\
+};
+
+\node[process, below=6mm of trtfTitle, text width=62mm] (trtfStepOne) {Estimate monotone marginals (Appendix~\ref{app:trtf})};
+\node[process, below=6mm of trtfStepOne, text width=62mm] (trtfStepTwo) {Fit transformation forests for $X_k \mid X_{1:(k-1)}$ with additive predictor};
+\node[process, below=6mm of trtfStepTwo, text width=62mm] (trtfStepThree) {Recover triangular likelihood via probability integral transform};
+
+\node[process, below=6mm of copTitle, text width=60mm] (copStepOne) {Estimate train marginals $\widehat F_k$, densities $\widehat f_k$};
+\node[process, below=6mm of copStepOne, text width=60mm] (copStepTwo) {Fit dependence on probit scores (KDE copula)};
+\node[process, below=6mm of copStepTwo, text width=60mm] (copStepThree) {Combine marginals with copula-based dependence};
+
+\node[process, font=\small\bfseries, text width=150mm, below=26mm of trtfStepThree] (outputs) {Reported outputs: log density, conditional diagnostics, samples, calibration, compute summaries};
+
+\draw[arrow] (data) -- (std);
+\draw[arrow] (std) -- (jac);
+\draw[arrow] (jac.south west) to[out=215,in=90] (ttmTitle.north);
+\draw[arrow] (jac) -- (trtfTitle);
+\draw[arrow] (jac.south east) to[out=325,in=90] (copTitle.north);
+
+\draw[arrow] (ttmTitle) -- (ttmMargTitle);
+\draw[arrow] (ttmTitle) -- (ttmSepTitle);
+\draw[arrow] (ttmMargTitle) -- (ttmMargStepOne);
+\draw[arrow] (ttmMargStepOne) -- (ttmMargStepTwo);
+\draw[arrow] (ttmSepTitle) -- (ttmSepStepOne);
+\draw[arrow] (ttmSepStepOne) -- (ttmSepStepTwo);
+
+\draw[arrow] (trtfTitle) -- (trtfStepOne);
+\draw[arrow] (trtfStepOne) -- (trtfStepTwo);
+\draw[arrow] (trtfStepTwo) -- (trtfStepThree);
+
+\draw[arrow] (copTitle) -- (copStepOne);
+\draw[arrow] (copStepOne) -- (copStepTwo);
+\draw[arrow] (copStepTwo) -- (copStepThree);
+
+\draw[arrow] (ttmMargStepTwo) to[out=-90,in=195] (outputs.west);
+\draw[arrow] (ttmSepStepTwo) to[out=-90,in=165] (outputs.west);
+\draw[arrow] (trtfStepThree) -- (outputs);
+\draw[arrow] (copStepThree) to[out=-90,in=0] (outputs.east);
+
+\end{tikzpicture}%
+\caption{Unified evaluation pipeline shared by triangular transport maps, Transformation Random Forests, and copulas.
+The diagram shows how standardized features flow through the triangular pullback or copula dependence block before reporting log densities, conditional diagnostics, samples, calibration metrics, and compute summaries.
+The shared preprocessing, Jacobian accumulation, and reporting path define the evaluation protocol used across the thesis.}
+\label{fig:transport-schematic}
+\end{figure}
+\end{landscape}
+
+\section{Pseudo-code Summaries for Model Routines}
+
+This appendix records consolidated pseudo-code for the core R implementations used in the experiments.
+Each summary captures inputs, main processing stages, and outputs so the execution flow is transparent without consulting the source code files.
+
+\subsection{Transformation Random Forest (TRTF)}\label{app:trtf}
+
+\textbf{Routine:}\quad\texttt{fit\_TRTF(S, config, seed, cores)} (calls \texttt{mytrtf}).
+
+\begin{enumerate}
+  \item Validate that the training matrix is numeric, set the RNG seed, and label columns as $X_1,\ldots,X_K$.
+  \item Fit an intercept-only transformation model \texttt{BoxCox} for each $X_k$ to provide baseline monotone transformations.
+  \item For $k = 2,\ldots,K$:
+    \begin{enumerate}
+      \item Build the formula $X_k \sim X_1 + \cdots + X_{k-1}$.
+      \item Choose \texttt{mtry = max(1, floor((k-1)/2))} and standard \texttt{ctree} controls (\texttt{minsplit}, \texttt{minbucket}, \texttt{maxdepth}).
+      \item Fit a transformation forest with \texttt{traforest} and store the conditional model (one forest per $k$).
+    \end{enumerate}
+  \item Return a \texttt{mytrtf} object containing baseline transformations, conditional forests, variable-importance scores, and the seed.
+
+  \item \textbf{Prediction (\texttt{predict.mytrtf}):}
+    \begin{enumerate}
+      \item Convert new data to the same column naming scheme and evaluate $X_1$ through its baseline transformation model to obtain marginal log densities.
+      \item For each conditional forest ($k\geq 2$) evaluate the log density of $X_k$ given $X_{1:(k-1)}$, extracting the diagonal when the forest returns a log-density matrix.
+      \item Stack the per-dimension log densities (\texttt{logdensity\_by\_dim}) or sum them to obtain the joint log likelihood (\texttt{logdensity}).
+    \end{enumerate}
+\end{enumerate}
+
+\subsection{Nonparametric Copula Baseline}\label{app:copula}
+
+\textbf{Routine:}\quad\texttt{fit\_copula\_np(S, seed)}.
+
+\begin{enumerate}
+  \item Inspect the training matrix and optional class labels; detect whether the dedicated copula packages are available.
+  \item If prerequisites fail (dimension $K \neq 2$ or labels missing), fall back to independent univariate kernel density estimates per dimension and store them for later interpolation.
+  \item Otherwise, for each class label:
+    \begin{enumerate}
+      \item Fit one-dimensional \texttt{kde1d} models to each marginal $X_1$ and $X_2$.
+      \item Convert training samples to pseudo-observations using mid-ranks scaled by $(n+1)^{-1}$ and clamp to $(\varepsilon, 1-\varepsilon)$.
+      \item Fit a two-dimensional kernel copula with \texttt{kdecopula::kdecop} (method \texttt{TLL2}).
+      \item Store marginals, copula fit, and effective sample size for the class.
+    \end{enumerate}
+  \item Record class priors and return a \texttt{copula\_np} object.
+
+  \item \textbf{Prediction (\texttt{predict.copula\_np}):}
+    \begin{enumerate}
+      \item In fallback mode evaluate each univariate KDE at the requested points and sum log densities.
+      \item In copula mode compute marginal log densities and CDF values, evaluate the copula density, and either:
+        \begin{enumerate}
+          \item Average over class-specific log densities weighted by priors (mixture prediction), or
+          \item Use the class labels supplied at prediction time.
+        \end{enumerate}
+      \item Return per-dimension log densities or their sum depending on the requested type.
+    \end{enumerate}
+\end{enumerate}
+
+\subsection{Triangular Transport Core Utilities}\label{app:ttm-core}
+
+\textbf{Module:}\quad\texttt{ttm\_core.R} (shared by marginal and separable TTM fits).
+
+\begin{enumerate}
+  \item Provide train-only standardisation helpers that cache feature means and standard deviations and reapply them to new data.
+  \item Define basis builders: polynomial features for predecessor coordinates $g_k$, monotone basis functions $f_k$ for the current coordinate, and their derivatives.
+  \item Implement optional ordering heuristics (identity or Cholesky pivoting with optional Gaussianisation) and persist selected permutations.
+  \item Expose a dispatcher \texttt{ttm\_forward(model, X)} that:
+    \begin{enumerate}
+      \item Standardises inputs using stored parameters.
+      \item For marginal maps applies affine transformations $a_k + b_k x_k$ with precomputed coefficients.
+      \item For separable maps constructs $g_k$ and $f_k$, computes $S_k = g_k + f_k$, and records the Jacobian diagonal $\partial_{x_k} S_k$.
+    \end{enumerate}
+  \item Provide \texttt{ttm\_ld\_by\_dim} to combine the forward map with the Gaussian reference, yielding per-dimension log densities used by all TTM variants.
+\end{enumerate}
+
+\subsection{Marginal Triangular Transport Map}\label{app:ttm-marg}
+
+\textbf{Routine:}\quad\texttt{fit\_ttm\_marginal(data, seed)}.
+
+\begin{enumerate}
+  \item Split data into train/test subsets if only a matrix is provided; otherwise accept a prepared list.
+  \item Standardise training features and, for each dimension $k$, compute closed-form coefficients $(a_k, b_k)$ that minimise the Gaussian pullback objective subject to $b_k > 0$.
+  \item Store model parameters (standardisation, per-dimension coefficients, ordering) and time measurements.
+  \item During prediction call \texttt{ttm\_forward} with the marginal coefficients and convert Jacobian diagonals to log densities via \texttt{ttm\_ld\_by\_dim}; aggregate per-dimension contributions when the joint log density is requested.
+\end{enumerate}
+
+\subsection{Separable Triangular Transport Map}\label{app:ttm-sep}
+
+\textbf{Routine:}\quad\texttt{fit\_ttm\_separable(data, degree\_g, lambda, seed)}.
+
+\begin{enumerate}
+  \item Prepare train/test splits and standardise training features as in the marginal case.
+  \item For each coordinate $k$:
+    \begin{enumerate}
+      \item Build polynomial features $g_k$ on previous coordinates (degree set by \texttt{degree\_g}).
+      \item Build monotone basis functions $f_k$ on the current coordinate and their derivatives.
+      \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
+      \item Otherwise solve the regularised optimisation problem
+            $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
+            using \texttt{optim} with L-BFGS-B while enforcing positivity of the derivative.
+      \item Store coefficients $c_{\text{non}}$ and $c_{\text{mon}}$ for the coordinate.
+    \end{enumerate}
+  \item Assemble the model list with standardisation parameters, coefficients, and metadata; record training/prediction timings.
+  \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
+\end{enumerate}
+
+\subsection{Evaluation Utilities}\label{app:evaluation}
+
+\textbf{Module:}\quad\texttt{evaluation.R} (experiment orchestration).
+
+\begin{enumerate}
+  \item Define convenience helpers such as \texttt{stderr(x)} and \texttt{add\_sum\_row} for table post-processing.
+  \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
+  \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split, times their evaluations, and returns the fitted objects together with per-dimension log-likelihood arrays.
+  \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
+  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artefacts.
+\end{enumerate}
+
+These structured summaries allow reproducing the algorithmic flow of each model without navigating the full R implementation.

### 2025-09-22 code/experiments/4d/plot_logdensity_scatter.R

@@ -1,0 +1,167 @@
+#!/usr/bin/env Rscript
+
+options(stringsAsFactors = FALSE)
+
+# Locate this script and work from its directory so relative paths match
+locate_script_dir <- function() {
+  frame_files <- vapply(sys.frames(), function(fr) {
+    if (!is.null(fr$ofile)) return(fr$ofile)
+    NA_character_
+  }, character(1), USE.NAMES = FALSE)
+  cand <- frame_files[!is.na(frame_files)][1]
+  if (!is.na(cand)) {
+    return(dirname(normalizePath(cand, winslash = "/", mustWork = TRUE)))
+  }
+  args <- commandArgs(trailingOnly = FALSE)
+  marker <- "--file="
+  file_arg <- args[grepl(marker, args, fixed = TRUE)]
+  if (length(file_arg)) {
+    return(dirname(normalizePath(sub(marker, "", file_arg[1]), winslash = "/", mustWork = TRUE)))
+  }
+  normalizePath(getwd(), winslash = "/", mustWork = FALSE)
+}
+
+script_dir <- locate_script_dir()
+code_root <- normalizePath(file.path(script_dir, "..", ".."), winslash = "/", mustWork = TRUE)
+old_wd <- getwd()
+setwd(code_root)
+on.exit(setwd(old_wd), add = TRUE)
+
+Sys.setenv(N_OVERRIDE = "250")
+
+source(file.path(script_dir, "main.R"), local = FALSE)
+
+if (!exists("perm", envir = .GlobalEnv) || !is.numeric(get("perm", envir = .GlobalEnv))) {
+  assign("perm", seq_len(length(config)), envir = .GlobalEnv)
+}
+
+seed <- as.integer(Sys.getenv("SEED", "42"))
+set.seed(seed)
+
+prep <- prepare_data(n, config, seed = seed)
+S0 <- prep$S
+K <- ncol(S0$X_tr)
+perm_use <- get("perm", envir = .GlobalEnv)
+if (!is.numeric(perm_use) || length(perm_use) != K) {
+  perm_use <- seq_len(K)
+  assign("perm", perm_use, envir = .GlobalEnv)
+}
+S <- list(
+  X_tr = S0$X_tr[, perm_use, drop = FALSE],
+  X_te = S0$X_te[, perm_use, drop = FALSE]
+)
+cfg <- config[perm_use]
+
[…]

### 2025-09-22 code/experiments/NF/gas/gas_preprocess.R

@@ -1,0 +1,125 @@
+# GAS (UCI) -> MAF-kompatible CSVs mit 10k Zeilen pro Split
+# Erwartet: ethylene_CO.txt im Arbeitsordner (oder Pfad unten anpassen)
+
+co_file <- "ethylene_CO.txt"
+if (!file.exists(co_file)) stop("Datei nicht gefunden: ", co_file)
+
+# --- Laden (schnell mit data.table::fread, sonst Base R) ---
+use_fread <- requireNamespace("data.table", quietly = TRUE)
+df <- if (use_fread) {
+  data.table::fread(co_file, fill = TRUE)
+} else {
+  read.table(co_file, header = TRUE, sep = "", stringsAsFactors = FALSE)
+}
+
+if (ncol(df) < 19) stop("Unerwartete Spaltenanzahl. Erwartet 19, gefunden: ", ncol(df))
+
+# Laut UCI: Spalten 1..3 = Zeit & Setpoints; 4..19 = 16 Sensoren
+X <- as.matrix(df[, 4:19])
+
+# --- Stark korrelierte Attribute entfernen (|Pearson| > 0.98), deterministisch von links nach rechts ---
+thr <- 0.98
+target_features <- 8L
+CM <- cor(X)  # 16x16
+CM[is.na(CM)] <- 0
+keep <- rep(TRUE, ncol(X))
+for (i in seq_len(ncol(X) - 1)) {
+  if (!keep[i]) next
+  for (j in (i + 1):ncol(X)) {
+    if (keep[j] && abs(CM[i, j]) > thr) keep[j] <- FALSE
+  }
+}
+
+kept_idx <- which(keep)
+if (length(kept_idx) > target_features) {
+  drop_order <- rev(kept_idx)
+  for (idx_drop in drop_order) {
+    if (length(kept_idx) <= target_features) break
+    keep[idx_drop] <- FALSE
+    kept_idx <- which(keep)
+  }
+}
+
+if (length(kept_idx) < target_features) {
+  combos <- utils::combn(seq_len(ncol(X)), target_features)
+  found <- NULL
+  for (col_idx in seq_len(ncol(combos))) {
+    cand <- combos[, col_idx]
+    sub_cor <- abs(CM[cand, cand])
+    diag(sub_cor) <- 0
+    if (all(sub_cor <= thr)) {
+      found <- cand
+      break
+    }
+  }
+  if (is.null(found)) {
[…]

### 2025-09-22 code/experiments/NF/hepmass/prep_hepmass_10k_preprocess.R

@@ -1,0 +1,111 @@
+# prep_hepmass_10k_preprocess.R
+# Zweck: 1) 1000_train/test laden  2) je TOP_N=10k Zeilen nehmen  3) Preprocessing  4) 2 CSV speichern
+TOP_N <- 10000
+OUT_TRAIN <- "hepmass_train_10k_preprocessed.csv"
+OUT_TEST  <- "hepmass_test_10k_preprocessed.csv"
+
+options(stringsAsFactors = FALSE)
+
+install_if_missing <- function(pkgs){
+  to_get <- pkgs[!pkgs %in% rownames(installed.packages())]
+  if(length(to_get)) install.packages(to_get, repos = "https://cloud.r-project.org")
+}
+install_if_missing(c("data.table"))
+library(data.table)
+
+# bevorzugt lokale .csv, dann .csv.gz; sonst Download von UCI
+resolve_path <- function(fname_csv, base="https://archive.ics.uci.edu/ml/machine-learning-databases/00347"){
+  f_csv <- file.path(".", fname_csv)
+  f_gz  <- paste0(f_csv, ".gz")
+  if (file.exists(f_csv)) return(f_csv)
+  if (file.exists(f_gz))  return(f_gz)
+  url <- file.path(base, basename(f_gz))
+  message("Lade: ", url)
+  download.file(url, f_gz, mode = "wb", quiet = TRUE)
+  return(f_gz)
+}
+
+# 1) Pfade ermitteln
+path_train <- resolve_path("1000_train.csv")
+path_test  <- resolve_path("1000_test.csv")
+
+# 2) Jeweils nur TOP_N Zeilen einlesen (schnell, deterministisch)
+tr_raw <- fread(path_train, nrows = TOP_N)
+te_raw <- fread(path_test,  nrows = TOP_N)
+
+# 3) Preprocessing
+# 3.1 nur positive Klasse (Label==1 in erster Spalte), dann Label-Spalte entfernen
+tr <- tr_raw[tr_raw[[1]] == 1][ , -1]
+te <- te_raw[te_raw[[1]] == 1][ , -1]
+
+# Trainings- und Testspalten auf konsistente Namen bringen
+if (ncol(tr) != ncol(te)) {
[…]

### 2025-09-22 code/experiments/NF/main/plotsensitivity.R

@@ -1,0 +1,154 @@
+# scripts/plot_n_sensitivity.R
+# Base R script to plot N-sensitivity (NLL vs N) with error bars per dataset.
+# Reads TRTF/NF-style result *.txt files from GAS/POWER/HEPMASS/MINIBOONE folders
+# and extracts "sum:" as mean log-likelihood. Error bars = standard error over
+# seeds for each (dataset, N). Output PNG is written to the results root.
+# Usage:
+#   Rscript plotsensitivity.R [RESULTS_ROOT]
+# Default RESULTS_ROOT:
+#   /Users/leonkiafaro/Documents/Masterthesis/code/experiments/NF/main/results
+
+stderr <- function(x) {
+  if (!length(x)) return(NA_real_)
+  if (length(x) == 1L) return(0)
+  s <- stats::sd(x)
+  if (is.na(s) || !is.finite(s)) return(0)
+  s / sqrt(length(x))
+}
+
+extract_nll_from_file <- function(path) {
+  bn <- basename(path)
+  base <- sub("\\.txt$", "", bn)
+  toks <- strsplit(base, "_", fixed = TRUE)[[1]]
+  seed_idx <- which(grepl("^seed[0-9]+$", toks))
+  n_idx <- which(grepl("^N[0-9]+$", toks))
+  if (length(seed_idx)) {
+    seed <- suppressWarnings(as.integer(sub("^seed", "", toks[seed_idx[1L]])))
+  } else {
+    seed <- NA_integer_
+  }
+  if (length(n_idx)) {
+    N <- suppressWarnings(as.integer(sub("^N", "", toks[n_idx[1L]])))
+  } else {
+    N <- suppressWarnings(as.integer(sub(".*_N([0-9]+).*", "\\1", bn)))
+  }
+  lines <- tryCatch(readLines(path, warn = FALSE), error = function(e) character(0))
+  sum_line <- if (length(lines)) grep("^\\s*sum\\s*:", lines, value = TRUE, ignore.case = TRUE) else character(0)
+  ll <- NA_real_
+  if (length(sum_line)) {
+    val <- sub("^.*sum\\s*:\\s*", "", sum_line[1L])
+    ll <- suppressWarnings(as.numeric(val))
+  }
+  nll <- if (is.finite(ll)) -ll else NA_real_
+  list(N = N, seed = seed, nll = nll)
+}
+
+args <- commandArgs(trailingOnly = TRUE)
[…]

### 2025-09-22 code/experiments/NF/miniboone/miniboone_dataprep.R

@@ -1,0 +1,90 @@
+# MiniBooNE -> 2x10k CSVs (MAF-style preprocessing on first 20k positives)
+# Expected input: MiniBooNE_PID.txt in working dir
+
+suppressPackageStartupMessages({
+  library(utils)
+  library(stats)
+})
+
+txt <- "MiniBooNE_PID.txt"
+if (!file.exists(txt)) stop("File not found: ", txt)
+
+# --- Read header (supports 'n_signal n_background' on line 1 OR split across 2 lines) ---
+hdr_line <- readLines(txt, n = 1L, warn = FALSE)
+p1 <- scan(text = hdr_line, what = integer(), quiet = TRUE)
+if (length(p1) == 2L) {
+  n_signal <- p1[1]; n_background <- p1[2]
+  skip <- 1L
+} else {
+  stop("Unrecognized header format: expecting two integers for counts")
+}
+
+message(sprintf("Header: n_signal=%d | n_background=%d", n_signal, n_background))
+
+# --- Load only positive examples (signal block) ---
+use_fread <- requireNamespace("data.table", quietly = TRUE)
+raw_mat <- if (use_fread) {
+  data.table::fread(txt, skip = skip, header = FALSE, data.table = FALSE,
+                    fill = TRUE, showProgress = FALSE)
+} else {
+  utils::read.table(txt, skip = skip, header = FALSE, fill = TRUE)
+}
+
+if (nrow(raw_mat) < n_signal) {
+  stop("File shorter than expected: have ", nrow(raw_mat), " rows, need ", n_signal)
+}
+if (ncol(raw_mat) < 50) stop("Expected >=50 columns, found: ", ncol(raw_mat))
+if (ncol(raw_mat) > 50) raw_mat <- raw_mat[, seq_len(50)]
+raw_mat[] <- lapply(raw_mat, function(x) as.numeric(as.character(x)))
+
+pos <- raw_mat[seq_len(n_signal), , drop = FALSE]
+ok <- stats::complete.cases(pos)
+pos <- pos[ok, , drop = FALSE]
+
+# --- Remove rows where all features equal -1000 (documented outliers) ---
+is_out <- rowSums(pos == -1000) == ncol(pos)
+pos <- pos[!is_out, , drop = FALSE]
+
+# --- Drop rows containing sentinel placeholders (-1000 or -999) in any column ---
[…]

### 2025-09-22 code/experiments/halfmoon/halfmoon_pipeline.R

@@ -55,6 +55,7 @@ eval_time <- system.time({
 
 # Record approximate prediction times per model (seconds)
 timing$true$test <- system.time({ true_logdensity(S$X_te, S, Q = 32L) })[["elapsed"]]
+timing$true_marg$test <- system.time({ predict(mods$true_marg, S$X_te, type = "logdensity_by_dim") })[["elapsed"]]
 timing$trtf$test <- if (!is.null(mods$trtf)) {
   system.time({ predict(mods$trtf, S$X_te, type = "logdensity_by_dim") })[["elapsed"]]
 } else NA_real_
@@ -110,7 +111,7 @@ timing$plot_png <- if (!is.null(res$png_elapsed)) res$png_elapsed else NA_real_
 # Compose run summary text
 summary_path <- file.path(run_dir, sprintf("halfmoon_summary_seed%03d.txt", SEED))
 fmt_time <- function(x) ifelse(is.na(x) || !is.finite(x), "NA", sprintf("%.3f", x))
-model_labels <- c(true = "True (Joint)", trtf = "TRTF", ttm = "TTM Marginal",
+model_labels <- c(true = "True (Joint)", true_marg = "True (Marg)", trtf = "TRTF", ttm = "TTM Marginal",
                   ttm_sep = "TTM Separable", copula_np = "Copula NP")
 timing_lines <- vapply(names(model_labels), function(nm) {
   train_t <- if (!is.null(timing[[nm]]$train)) timing[[nm]]$train else NA_real_

### 2025-09-22 code/experiments/halfmoon/plot.R

@@ -21,6 +21,7 @@ fit_halfmoon_models <- function(S, seed = NULL, order_mode = "as-is") {
 
   timing <- list()
   timing$true <- list(train = 0.0, test = NA_real_)
+  timing$true_marg <- list(train = 0.0, test = NA_real_)
 
   trtf_time <- system.time({
     M_trtf <- tryCatch(fit_TRTF(S, cfg, seed = seed), error = function(e) NULL)
@@ -40,6 +41,7 @@ fit_halfmoon_models <- function(S, seed = NULL, order_mode = "as-is") {
 
   mods <- list(
     true = M_true,
+    true_marg = M_true,
     trtf = M_trtf,
     ttm = fit_ttm_m$S,
     ttm_sep = fit_ttm_s$S,
@@ -272,7 +274,7 @@ plot_halfmoon_models <- function(mods, S, grid_side,
   xseq <- seq(xlim[1], xlim[2], length.out = grid_side)
   yseq <- seq(ylim[1], ylim[2], length.out = grid_side)
   G <- as.matrix(expand.grid(xseq, yseq))
-  panels <- c("true", "trtf", "ttm", "ttm_sep", "copula_np")
+  panels <- c("true", "true_marg", "trtf", "ttm", "ttm_sep", "copula_np")
   eval_timer_start <- proc.time()
   # TRUE-Panel: exakte Log-Dichte gem. Generator
   source(file.path(root_path, "experiments", "halfmoon", "true_density.R"))
@@ -311,8 +313,8 @@ plot_halfmoon_models <- function(mods, S, grid_side,
   ncol <- min(3L, n_panels)
   nrow <- ceiling(n_panels / ncol)
   op <- par(mfrow = c(nrow, ncol), mar = c(3, 3, 2, 1))
-  titles <- c(true = "True (Joint)", trtf = "TRTF", ttm = "TTM Marginal",
-              ttm_sep = "TTM Separable", copula_np = "Copula NP")
+  titles <- c(true = "True (Joint)", true_marg = "True (Marg)", trtf = "TRTF",
+              ttm = "TTM Marginal", ttm_sep = "TTM Separable", copula_np = "Copula NP")
   for (nm in panels) {
     Z <- matrix(LD_list[[nm]], nrow = grid_side, ncol = grid_side)
     plot(NA, xlim = xlim, ylim = ylim, xlab = "x1", ylab = "x2",
[…]

### 2025-09-22 code/experiments/normalizing flow/miniboone/miniboone_dataprep.R

@@ -1,156 +1,0 @@
-###############################################################################
-#   1. Herunterladen & Einlesen                                              #
-#   2. Nur positive Ereignisse (electron-neutrinos) behalten                 #
-#   3. 11 Platzhalter-Zeilen (-1000/-999 in allen 50 Spalten) entfernen       #
-#   4. 7 Spalten mit extrem häufig gleichem Wert (“Dirac-Peaks”) streichen    #
-#   5. Standardisierung: (x-μ_train)/σ_train                                  #
-#   6. Reproduzierbarer Split 80 %/10 %/10 % -> Train / Val / Test            #
-#   7. Speichern als *.RData                                                 #
-#                                                                            #
-# Am Ende liegen drei Objekte vor: x_train, x_val, x_test                    #
-# Dimensionen laut Paper:                                                    #
-#   • Train: 29 556 × 43                                                     #
-#   • Val:    3 284 × 43                                                     #
-#   • Test:   3 648 × 43                                                     #
-###############################################################################
-# 
-# # ------------------------- 1 | Download & Roh-Import -------------------------
-# url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00199/MiniBooNE_PID.txt"
-# tf  <- tempfile()
-# download.file(url, tf, mode = "wb")          # ~87 MB
-# 
-# con   <- file(tf, "r")
-# nums  <- scan(con, what = integer(), n = 2, quiet = TRUE)
-# n_sig <- nums[1]                             # Zeile 1: #positiver Events
-# # Zeile 2: #negativer Events (wird ignoriert)
-# 
-# raw   <- scan(con, what = numeric(), quiet = TRUE)  # restliche Zahlen
-# close(con)
-# 
-# x <- matrix(raw, ncol = 50, byrow = TRUE)[1:n_sig, ]  # 50 Merkmale, nur positives
-# 
-# # --------------------- 2 | 11 vollständige Ausreißer löschen -----------------
[…]

### 2025-09-22 code/experiments/normalizing flow/power/main_power.R

@@ -1,211 +1,0 @@
-if (!exists("initialize_repo")) {
-  source(file.path(getwd(), "R", "loader.R"))
-}
-root_path <- initialize_repo()
-
-RUN_CONFIG <- list(
-  enabled = FALSE,
-  train_csv = "/Users/leonkiafaro/Documents/Masterthesis/data_power/power_train_head_1pct.csv",
-  test_csv  = "/Users/leonkiafaro/Documents/Masterthesis/data_power/power_test_head_1pct.csv",
-  n_train = 250L,
-  n_test  = 250L,
-  models = c("trtf","ttm","ttm_sep","copula_np")
-)
-
-# Small helpers
-stderr <- function(x) stats::sd(x) / sqrt(length(x))
-
-`%||%` <- function(a, b) if (is.null(a)) b else a
-
-read_numeric_matrix <- function(path, n_rows = NA_integer_) {
-  if (!file.exists(path)) stop(sprintf("CSV not found: %s", path))
-  df <- utils::read.csv(path, header = TRUE, check.names = TRUE)
-  # Keep only numeric columns present
-  num_cols <- names(df)[vapply(df, is.numeric, logical(1))]
-  if (!length(num_cols)) stop(sprintf("No numeric columns in %s", basename(path)))
-  df <- df[, num_cols, drop = FALSE]
-  if (is.finite(n_rows) && !is.na(n_rows) && n_rows > 0L) {
-    n_keep <- min(n_rows, nrow(df))
-    df <- utils::head(df, n_keep)
-  }
-  X <- as.matrix(df)
-  # Drop rows with NA/Inf
-  ok <- stats::complete.cases(X) & apply(is.finite(X), 1, all)
-  if (!all(ok)) X <- X[ok, , drop = FALSE]
-  storage.mode(X) <- "double"
-  X
-}
-
-parse_args <- function() {
-  out <- list()
-  for (a in commandArgs(trailingOnly = TRUE)) {
-    if (grepl("^--n_train=", a)) out$n_train <- as.integer(sub("^--n_train=", "", a))
-    if (grepl("^--n_test=", a))  out$n_test  <- as.integer(sub("^--n_test=",  "", a))
-    if (grepl("^--train_csv=", a)) out$train_csv <- sub("^--train_csv=", "", a)
[…]

### 2025-09-22 code/experiments/normalizing flow/power/subset_head_three_datasets.R

@@ -1,142 +1,0 @@
-# subset_head_three_datasets.R
-# Zweck: Von drei Dateien (POWER, GAS, HEPMASS) die ersten p% Datenzeilen
-#        (exkl. Header) extrahieren und als CSV speichern.
-
-suppressWarnings(suppressMessages({
-  have_dt <- requireNamespace("data.table", quietly = TRUE)
-}))
-
-# -- Hilfsfunktion: schnelle Zeilenzahl (Unix: wc -l; Fallback: binär lesen) --
-count_lines_fast <- function(path) {
-  stopifnot(file.exists(path))
-  out <- tryCatch(
-    system2("wc", c("-l", shQuote(path)), stdout = TRUE),
-    error = function(e) NA_character_
-  )
-  if (!is.na(out[1])) {
-    n <- suppressWarnings(as.integer(sub("^\\s*(\\d+).*$", "\\1", out[1])))
-    if (is.finite(n)) return(n)
-  }
-  con <- file(path, "rb"); on.exit(close(con), add = TRUE)
-  n <- 0L
-  repeat {
-    buf <- readBin(con, what = "raw", n = 8L * 1024L * 1024L)
-    if (!length(buf)) break
-    n <- n + sum(buf == as.raw(10L)) # '\n'
-  }
-  n
-}
-
-# -- Hilfsfunktion: Prozentzahl in Bruch umwandeln (5 -> 0.05; 0.5 -> 0.5) --
-as_fraction <- function(pct) {
-  stopifnot(is.numeric(pct), length(pct) == 1L, is.finite(pct), pct > 0)
-  if (pct > 1) pct/100 else pct
-}
-
-# -- Kernfunktion: p% Kopf speichern --
-subset_head_percent_save <- function(in_file, pct, out_file,
-                                     sep_in = NULL, header = TRUE,
-                                     na = c("", "?", "NA"),
-                                     out_sep = ",",
-                                     use_fread = have_dt) {
-  frac <- as_fraction(pct)
-  total <- count_lines_fast(in_file)
-  if (!is.finite(total) || total <= 0) stop("Konnte Zeilenzahl nicht bestimmen: ", in_file)
-  header_rows <- if (isTRUE(header)) 1L else 0L
-  data_rows_total <- max(0L, total - header_rows)
-  n <- max(1L, floor(data_rows_total * frac))
-  
-  if (use_fread) {
-    args <- list(input = in_file, nrows = n, header = header,
[…]

### 2025-09-22 code/experiments/normalizing flow/power/trtf_power.R

@@ -1,764 +1,0 @@
-### Minimaler, schneller Konfigurationsblock (oben anpassbar) #################
-# A) Anzahl Zeilen pro Split (Train/Test)
-N_ROWS <- 10000L
-# B) CSV-Pfade relativ zum Skript-Verzeichnis
-SCRIPT_DIR <- (function(){
-  args <- commandArgs(trailingOnly = FALSE)
-  file_arg <- grep("^--file=", args, value = TRUE)
-  if (length(file_arg) > 0L) return(dirname(normalizePath(sub("^--file=", "", file_arg[[1L]]))))
-  # when sourced
-  f <- tryCatch(normalizePath(sys.frames()[[1]]$ofile), error = function(e) NA_character_)
-  if (is.character(f) && length(f) == 1L && !is.na(f)) return(dirname(f))
-  getwd()
-})()
-
-locate_data_dir <- function(initial_dir, required) {
-  candidates <- unique(c(
-    initial_dir,
-    normalizePath(getwd(), winslash = "/", mustWork = FALSE),
-    normalizePath(file.path(getwd(), "code", "experiments", "normalizing flow", "power"), winslash = "/", mustWork = FALSE)
-  ))
-  for (dir in candidates) {
-    if (nzchar(dir) && all(file.exists(file.path(dir, required)))) return(dir)
-  }
-  initial_dir
-}
-
-SCRIPT_DIR <- locate_data_dir(SCRIPT_DIR, c("power_train_10000.csv", "power_test_10000.csv"))
-TRAIN_CSV <- file.path(SCRIPT_DIR, "power_train_10000.csv")
-TEST_CSV  <- file.path(SCRIPT_DIR, "power_test_10000.csv")
-# C) Pfad für Evaluationstext (Konsole + Datei)
-RESULTS_TXT <- file.path(SCRIPT_DIR, "results", "trtf_power_eval.txt")
-# D) Hyperparameter
-HP <- list(
-  ntree     = 200L,
-  minsplit  = 100L,
-  minbucket = 20L,
-  maxdepth  = 4L,
-  seed      = 42L
-)
-###############################################################################
-
-ensure_required_packages <- function(pkgs) {
-  missing <- character(0)
-  for (pkg in pkgs) {
-    ok <- suppressPackageStartupMessages(require(pkg, quietly = TRUE, character.only = TRUE))
-    if (!isTRUE(ok)) missing <- c(missing, pkg)
-  }
-  if (length(missing)) {
[…]

### 2025-09-29 report/MSc_Report.Rnw
.

--- 2025-09-20 report/MSc_Report.Rnw
@@
 \chapter*{Abstract}
-% TODO: add abstract text.
+% Checklist (Abstract): [x] integrate supplied summary [x] confirm PDF rebuild
 
-\chapter*{Acknowledgments}
-% TODO: add acknowledgments.
+This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF; additive predictor), and copulas (used only for low-dimensional diagnostics, $K\!\le\!3$).
+All methods use standardized inputs and a common evaluation protocol so that likelihoods, diagnostics, and compute are directly comparable.
+In the configuration studied (additive predictor and monotone CDF smoothing), TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation.
+
+On Half-Moon ($n=250$), mean joint negative log-likelihoods (NLL; lower is better) were $1.71$ (TRTF), $1.93$ (TTM-Sep), and $1.54$ (copula).
+On a four-dimensional autoregressive generator they were $4.53$, $5.66$, and $5.45$, respectively; permutation averages confirm order sensitivity for triangular maps.
+On MiniBooNE ($K=43$; sum test log-likelihood), TRTF reached $-30.01$ under the standard preprocessing and training budget used here; published flow models report values around $-12$ to $-16$ under their settings.
+These numbers are not strictly comparable but indicate the relative accuracy of this configuration.
+
+Overall, TRTF tends to lead within the separable family at low dimension, while higher-dimensional datasets expose the limits of separable structure.
+We report robustness checks (ordering), calibration diagnostics, and the numerical safeguards used, and we outline directions toward richer parameterizations within the same evaluation frame.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-<<'child-chapter05', child='chapter05.Rnw'>>=
-@
+% References
+\bibliographystyle{plainnat}
+\bibliography{biblio}
 

### 2025-09-29 report/MSc_Report.tex
.

@@
 \chapter*{Abstract}
-% Checklist (Abstract): [x] integrate supplied summary [ ] confirm PDF rebuild
-
-This thesis studies multivariate density estimation for tabular data within a transport-based evaluation framework.
-All models standardize features and learn a monotone lower-triangular map $S \colon \mathbf{u} \to \mathbf{z}$ to a Gaussian reference distribution.
-The estimators accumulate exact log-Jacobians in standardized space with an affine correction for reporting on the data scale.
-We compare separable triangular transport maps (TTM-Sep), additive-predictor Transformation Random Forests (TRTF), and copula baselines under the shared preprocessing pipeline.
-Additive TRTF induces the same separable triangular likelihood as TTM-Sep via the probability integral transform.
-The resulting estimator yields exact likelihoods, transparent conditionals, and linear per-sample cost but restricts context-dependent shape.
-
-Benchmarks include synthetic Half-Moon and a four-dimensional autoregressive generator, plus the UCI POWER, GAS, HEPMASS, and MINIBOONE datasets.
-For Half-Moon with $n = 250$, joint negative log-likelihoods (nats) were $1.71$ for TRTF, $1.93$ for TTM-Sep, and $1.54$ for the copula baseline.
-For the four-dimensional generator, TRTF achieved $4.53$, TTM-Sep $5.66$, and the copula baseline $5.45$.
-Permutation averages confirmed ordering sensitivity for triangular maps.
-On MINIBOONE with $K = 43$ and $N = 2{,}500$, TRTF reached $-30.01$ test log-likelihood, whereas flow baselines remained near $-11.7$.
-
-Findings indicate that TRTF narrows the gap to oracle likelihoods on low-dimensional synthetic data.
-The method trails modern flows on high-dimensional real data because separability limits context-dependent shape.
-Copula baselines remain competitive in low dimensions.
-This thesis records guidance and limitations, including ordering effects and fragility in sparse regimes.
+% Checklist (Abstract): [x] integrate supplied summary [x] confirm PDF rebuild
+
+This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF; additive predictor), and copulas (used only for low-dimensional diagnostics, $K\!\le\!3$).
+All methods use standardized inputs and a common evaluation protocol so that likelihoods, diagnostics, and compute are directly comparable.
+In the configuration studied (additive predictor and monotone CDF smoothing), TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation.
+
+On Half-Moon ($n=250$), mean joint negative log-likelihoods (NLL; lower is better) were $1.71$ (TRTF), $1.93$ (TTM-Sep), and $1.54$ (copula).
+On a four-dimensional autoregressive generator they were $4.53$, $5.66$, and $5.45$, respectively; permutation averages confirm order sensitivity for triangular maps.
+On MiniBooNE ($K=43$; sum test log-likelihood), TRTF reached $-30.01$ under the standard preprocessing and training budget used here; published flow models report values around $-12$ to $-16$ under their settings.
+These numbers are not strictly comparable but indicate the relative accuracy of this configuration.
+
+Overall, TRTF tends to lead within the separable family at low dimension, while higher-dimensional datasets expose the limits of separable structure.
+We report robustness checks (ordering), calibration diagnostics, and the numerical safeguards used, and we outline directions toward richer parameterizations within the same evaluation frame.
 
@@
 
-We work in standardized coordinates to unify Jacobians, gradients, and reporting conventions. Let $x \in \mathbb{R}^K$ denote features on the original scale, and define
-\begin{equation}
-  u = T_{\text{std}}(x) = (x - \mu) \oslash \sigma,
-  \qquad \sigma_k > 0,\label{eq:ch1-standardization}
-\end{equation}
-which standardizes each feature using training statistics only and fixes the coordinates where derivatives and Jacobians are evaluated.
-Let $S: u \mapsto z$ be a monotone lower-triangular map, and let $\eta$ denote the standard normal density.
-The change-of-variables identity gives
-\begin{equation}
-  \log \pi_U(u) = \log \eta\big(S(u)\big) + \log \big|\det \nabla_u S(u)\big|.
-  \label{eq:ch1-pullback}
-\end{equation}
-Equation~\eqref{eq:ch1-pullback} splits the log density into a reference-fit term and a Jacobian term, isolating modeling capacity from the exact volume correction contributed by the map.
-For triangular $S$, the determinant factorizes as
-\begin{equation}
-  \log \big|\det \nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k} S_k(u_{1:k}).
-  \label{eq:ch1-triangular-det}
-\end{equation}
-Equation~\eqref{eq:ch1-triangular-det} reduces the determinant to a sum of one-dimensional log derivatives, yielding linear per-sample complexity in $K$ and improving numerical stability.
-Reported log densities on the original scale apply the diagonal affine correction
-\begin{equation}
-  \log \pi_X(x) = \log \pi_U\!\big(T_{\text{std}}(x)\big) - \sum_{k=1}^K \log \sigma_k.
-  \label{eq:ch1-affine-correction}
-\end{equation}
-Equation~\eqref{eq:ch1-affine-correction} subtracts a constant offset determined by the training scales and ensures comparability across models that share standardized coordinates.
+We adopt a shared evaluation protocol summarized in Section~\ref{sec:ch1-frame} and detailed in Chapter~\ref{ch:background}; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the pipeline at a glance.
 \medskip
 
-We study separable triangular components for clarity, efficiency, and interpretability. Component $k$ decomposes into a context shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),\qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:ch1-separable}
-\end{equation}
-which stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.
+For clarity, efficiency, and interpretability we use separable triangular components as in Equation~\eqref{eq:transport-separable}.
+This structure stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.
 
@@
 This thesis investigates tabular multivariate density estimation within a unified transport-based evaluation frame.
-We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests with axis-parallel splits (TRTF), and copula baselines.
+We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and copula baselines.
 \medskip
@@
 We standardize features using training statistics only.
-Equation~\eqref{eq:ch1-standardization} defines the standardized coordinates $u$ used for evaluation.
+Equation~\eqref{eq:transport-standardise} defines the standardized coordinates $u$ used for evaluation.
 All derivatives and Jacobians are computed in $u$.
-The diagonal affine correction in Equation~\eqref{eq:ch1-affine-correction} reports log densities on the original scale $x$.
+The diagonal affine correction in Equation~\eqref{eq:transport-affine} reports log densities on the original scale $x$.
 This convention keeps objectives, diagnostics, and comparisons interoperable across estimators and datasets.
-All negative log-likelihoods are reported in nats.
+All log quantities are reported in nats.
 \medskip
 
-We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\phi$ and $\Phi$.
-Triangular transport maps are abbreviated TTM, with the separable variant labeled TTM-Sep.
-Transformation Random Forests are abbreviated TRTF, and the axis-parallel implementation is labeled TRTF.
+We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\varphi$ and $\Phi$.
+Abbreviations for models and references appear in Table~\ref{tab:model-abbrev}.
 Copulas decouple marginals from dependence and serve as interpretable baselines.
@@
 
-Separable triangular maps decompose each component into a context shift and a univariate monotone shape as shown in Equation~\eqref{eq:ch1-separable}.
+Separable triangular maps decompose each component into a context shift and a univariate monotone shape as in Equation~\eqref{eq:transport-separable}.
 The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant.
-TRTF implements the same separable triangular likelihood via the probability integral transform and an additive predictor.
-Copulas preserve explicit marginals and model dependence on the unit hypercube.
-
-We evaluate all estimators under a single protocol that records test log-likelihoods, conditional diagnostics, and compute.
-The protocol fixes the map direction from standardized data to the Gaussian reference, which preserves separability and linear evaluation cost.
-Section~\ref{sec:evaluation-protocol} details the metrics, calibration diagnostics based on probability integral transforms, and timing conventions.
+Under strictly increasing conditional CDFs after standard monotone smoothing and with an additive predictor, TRTF implements the same separable triangular likelihood via the probability integral transform.
+Copulas preserve explicit marginals and model dependence on the unit hypercube; in this thesis they serve strictly as low-dimensional ($K\!\le\!3$) diagnostic baselines and are not evaluated on high-$K$ datasets.
+
+We evaluate all estimators under the single protocol referenced above, with matched preprocessing and reporting to keep results comparable.
+Section~\ref{sec:evaluation-protocol} defines metrics and timing conventions.
 
@@
 The central problem is to determine when separability is appropriate for tabular data.
-We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention.
+We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention (reported log densities on $x$ apply the affine correction in Eq.~\eqref{eq:transport-affine}).
 Ordering effects, conditional calibration, and computational trade-offs address this question.
@@
 
-On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MINIBOONE benchmark it improves on Gaussian references but trails published flow baselines.
+On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MiniBooNE benchmark it improves on Gaussian references but trails published flow baselines.
 Chapter~\ref{ch:dataanalysis} presents the evidence and discusses these comparisons.
@@
 \paragraph{Non-goals.} We do not treat high-capacity normalizing flows as primary models, and we restrict nonparametric copulas to low dimensions.
+We also exclude non-separable TRTF predictors and cross-term triangular maps because they raise compute and tuning costs substantially and complicate calibration/identifiability in our setting; the present scope focuses on separable structure for transparency and exact inversion.
 Section~\ref{sec:ch1-contributions} states the scope and non-goals for reference.
+A brief pointer to expected changes if these variants were enabled—how cross-terms or non-additive predictors would likely affect calibration and NLL—is given in Chapter~\ref{ch:conclusion}, Section~\ref{sec:conclusion-outlook}.
 
@@
 \section{The Transport Frame on One Page}\label{sec:ch1-frame}
-This section fixes the evaluation frame used across triangular transport maps, Transformation Random Forests, and copulas.
-It establishes notation, equations, and reporting conventions for the remainder of the thesis.
-The frame places all derivatives and Jacobians in standardized coordinates, applies a single diagonal affine correction on the original scale, and yields linear per-sample evaluation through a lower-triangular map.
-
-Train-only standardization defines the evaluation coordinates via Equation~\eqref{eq:ch1-standardization}.
-The transformation uses training means and scales once and prevents information leakage into evaluation.
-All gradients and Jacobians act on $u$, which keeps objectives comparable across estimators and datasets.
-We convert to the original scale only when reporting.
-\medskip
-
-Let $S:u\mapsto z$ be a monotone lower-triangular map, and let $\eta(z)$ denote the $K$-variate standard normal density.
-The pullback identity in Equation~\eqref{eq:ch1-pullback} evaluates the reference at $S(u)$ and applies the exact volume correction.
-The split isolates model fit from geometry: the first term measures closeness to the Gaussian reference, and the second term contributes the exact Jacobian adjustment implied by the map.
-\medskip
-
-Lower-triangular structure with strictly positive diagonal partial derivatives factorizes the Jacobian as in Equation~\eqref{eq:ch1-triangular-det}.
-The factorization yields $\mathcal{O}(K)$ per-sample time for likelihood evaluation and improves numerical stability.
-It also guarantees global invertibility by back substitution consistent with the Rosenblatt and Knothe rearrangements.
-\medskip
-
-Reported log densities on the original scale apply only the diagonal affine correction implied by standardization.
-Equation~\eqref{eq:ch1-affine-correction} provides that adjustment.
-This convention keeps training and evaluation in $u$-space and converts to $x$-space at report time.
-All negative log-likelihoods are reported in nats, which fixes units across tables and figures.
-\medskip
-
-For clarity and efficiency we use separable triangular components of the form in Equation~\eqref{eq:ch1-separable}.
-Separable structure shifts location through $g_k$ and fixes conditional shape through the univariate monotone $h_k$.
-The Jacobian contribution depends only on $u_k$, which simplifies inversion by back substitution and stabilizes the log-determinant accumulation.
-The constraint also clarifies limits because conditional variance, skewness, and modality do not vary with context under separability.
-
-The core operations follow the same path for all estimators.
-\paragraph{Core operations.}
-\begin{enumerate}
-  \item First, standardize input features with training statistics using Equation~\eqref{eq:ch1-standardization}.
-  \item Then, learn a monotone lower-triangular map $S$ to the standard normal reference and exploit Equation~\eqref{eq:ch1-pullback}.
-  \item Finally, evaluate likelihoods, conditionals, and samples through Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction}.
-\end{enumerate}
-
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline.
-The input $x$ enters standardization to produce $u$.
-The triangular branch houses TTM-Sep and TRTF, whereas the copula branch couples fitted marginals to a dependence density.
-Both branches feed the same reported quantities, namely log density, conditionals, sampling, calibration, and compute.
-The appendix placement preserves the full schematic while keeping Chapter~\ref{ch:intro} focused on the narrative.
+To avoid duplication, the canonical derivations and notation live in Chapter~\ref{ch:background}.
+This section serves only as a map: we standardize with train-only statistics (Eq.~\eqref{eq:transport-standardise}), evaluate likelihoods via the pullback (Eq.~\eqref{eq:transport-pullback}), exploit the triangular determinant factorization (Eq.~\eqref{eq:transport-det}), and apply the affine correction for reporting (Eq.~\eqref{eq:transport-affine}).
+Separable components are defined in Eq.~\eqref{eq:transport-separable}.
+Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the pipeline.
 
 Notation remains consistent.
-We write $\eta$ for the $K$-variate standard normal density, and $\phi$ and $\Phi$ for the univariate standard normal density and CDF.
+We write $\eta$ for the $K$-variate standard normal density, and $\varphi$ and $\Phi$ for the univariate standard normal density and CDF.
 We reserve $u$ for standardized coordinates and $x$ for original coordinates, and we compute all derivatives with respect to $u$.
@@
 \section{Contributions and Research Questions}\label{sec:ch1-contributions}
-This section states the contributions of the thesis and formulates the research questions.
-It also maps both elements to the chapters and figures that deliver the evidence.
-The shared transport frame fixes standardized coordinates, keeps all derivatives and Jacobians in $u$-space, and applies a single diagonal affine correction on the original scale for reporting; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline and anchors the comparisons that follow.
-All negative log-likelihoods are reported in nats, and evaluation uses linear-time lower-triangular maps.
-
-The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests with axis-parallel splits, and copula baselines.
-The view specifies standardization $u = T_{\text{std}}(x)$ and the pullback identity with a monotone lower-triangular map.
-It also fixes the Jacobian factorization and the affine correction for reporting, which together align objectives and diagnostics across estimators.
-Equations~\eqref{eq:ch1-standardization}--\eqref{eq:ch1-affine-correction} and Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} establish these conventions and remove duplication in later chapters.
+This section states the contributions and the research questions, and maps them to the chapters and figures that deliver the evidence.
+We adopt the shared transport frame summarized in Section~\ref{sec:ch1-frame}; Chapter~\ref{ch:background} records notation and assumptions, and Figure~\ref{fig:transport-schematic} anchors the comparisons.
+
+The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests, and copula baselines.
+Where we claim an equivalence between TRTF and separable triangular transports, it holds under the conditions made explicit in Section~\ref{sec:transport-trtf} (strictly increasing conditional CDFs after monotone smoothing and an additive predictor).
+Chapter~\ref{ch:background} and Figure~\ref{fig:transport-schematic} establish the conventions and remove duplication in later chapters.
 
 The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting.
-We evaluate TTM-Sep, TTM-Marg, TRTF, and copulas on synthetic generators and on real tabular data.
+We evaluate TTM-Sep, TTM-Marg, and TRTF on synthetic generators and real tabular data; copulas are included only as low-dimensional ($K\!\le\!3$) diagnostic baselines (e.g., Half-Moon, 4D) and are not used in high-$K$ studies.
 The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation.
@@
 All estimators share the transport frame in this comparison.
-We answer by reporting average test negative log-likelihoods in nats, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost.
+We answer by reporting average test negative log-likelihoods, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost.
 Section~\ref{sec:synthetic-results} provides the corresponding tables and figures.
@@
 
+% Abbreviation table for reuse across chapters
+\begin{table}[t]
+  \centering
+  \caption{Model abbreviations used throughout the thesis.}
+  \label{tab:model-abbrev}
+  \begin{tabular}{ll}
+    \hline
+    Label & Meaning \\
+    \hline
+    TTM-Marg & Marginal triangular transport map (per-dimension; no context) \\
+    TTM-Sep & Separable triangular transport map (additive: $g_k$ shift + monotone $h_k$) \\
+    TRTF & Transformation Random Forests (axis-parallel splits) \\
+    True-Marg & Oracle marginal density \\
+    True-Joint & Oracle conditional joint density \\
+    Copula & Copula baseline (Gaussian or nonparametric) \\
+    \hline
+  \end{tabular}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters.
-We summarise the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorisation that drives evaluation and inversion.
-
-We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardisation map
+We summarize the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorization that drives evaluation and inversion.
+
+We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardization map
 \begin{equation}
@@
 
-The standardised density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$.
+The standardized density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$.
 Throughout the thesis the reference is the $K$-variate standard normal density $\eta(z)$.
@@
 which evaluates the reference at $S(u)$ and applies the exact volume correction given by the Jacobian determinant.
-Reporting log densities on the original scale requires only the diagonal affine correction implied by standardisation,
+Reporting log densities on the original scale requires only the diagonal affine correction implied by standardization,
 \begin{equation}
@@
 \end{equation}
-so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorises into a sum of one-dimensional log derivatives,
+so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorizes into a sum of one-dimensional log derivatives,
 \begin{equation}
@@
 \end{equation}
-The factorisation yields $\mathcal{O}(K)$ evaluation cost per sample, improves numerical stability, and guarantees global invertibility: strict monotonicity lets us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.
+The factorization yields $\mathcal{O}(K)$ evaluation cost per-sample, improves numerical stability, and guarantees global invertibility: strictly monotone diagonal derivatives let us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.
 
@@
 The remainder of this chapter adopts this frame.
-Section~\ref{sec:transport-separable} details the separable triangular parameterisation used for direct transports.
-Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform.
+Section~\ref{sec:transport-separable} details the separable triangular parameterization used for direct transports.
+Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform under the conditions stated there (strictly increasing conditional CDFs after monotone smoothing and an additive predictor).
 Section~\ref{sec:transport-copula} places copulas in the same reporting convention.
@@
     $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
-    $T_{\mathrm{std}}$ & Standardisation map using training $(\mu,\sigma)$ \\ 
-    $u = T_{\mathrm{std}}(x)$ & Standardised evaluation coordinates \\ 
+    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ \\ 
+    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates \\ 
     $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
@@
 Both estimators realize a monotone lower-triangular map $S:u\mapsto z$ that couples the standardized target to the Gaussian reference $\eta$.
+The use of triangular transports builds on modern measure-transport literature; see, for instance, triangular transformations and their properties in \citet{bogachev2005triangular}.
 Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the shared backbone and locates \textsc{TTM-Sep} and \textsc{TRTF} on the transport branch introduced in Chapter~\ref{ch:intro}.
@@
 
+\begin{shaded}
+\textbf{Assumptions.} Unless stated otherwise, we assume:
+\begin{itemize}
+  \item \emph{Lower-triangularity:} $S$ has the structure in Eq.~\eqref{eq:transport-triangular}.
+  \item \emph{Strict monotone coordinates:} $\partial_{u_k} S_k(u_{1:k}) > 0$ for all $k$ and all arguments.
+  \item \emph{Separable component:} Eq.~\eqref{eq:transport-separable} holds, so conditional shape along $u_k$ is fixed across contexts.
+\end{itemize}
+\end{shaded}
+
 Substituting the standard normal reference into \eqref{eq:transport-pullback} produces a separable objective,
@@
 
-The negative log-likelihood per sample takes the quadratic-plus-barrier form
+The negative log-likelihood per-sample takes the quadratic-plus-barrier form
 \begin{equation}
@@
 Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by a TRTF for coordinate $k$.
-The induced triangular component is
+(In practice, forest CDFs can be stepwise; we assume a measurable, strictly increasing version after standard monotone smoothing so that inversion and derivatives are well-defined.) The induced triangular component is
 \begin{equation}
@@
 so TRTF implements the same separable triangular likelihood as the direct parameterization in Equation~\eqref{eq:transport-separable}.
-Monotonicity in $u_k$ holds by construction, the Jacobian depends only on $u_k$, and inversion proceeds by back-substitution identical to the separable map.
+The map is monotone in $u_k$ by construction, the Jacobian depends only on $u_k$, and inversion proceeds by back-substitution identical to the separable map.
 This equivalence underpins the empirical comparisons in Chapter~\ref{ch:dataanalysis}.
@@
 
-In summary, separable triangular maps and additive-predictor TRTF realize the same lower-triangular likelihood once the data are standardized.
+In summary, separable triangular maps and additive-predictor TRTF realize the same lower-triangular likelihood once the data are standardized and the conditional CDFs are strictly increasing after monotone smoothing.
 The shared structure yields exact likelihoods, exact inversion, transparent conditionals, and linear per-sample complexity, but it restricts context-dependent shape.
@@
 This section positions copulas within the unified transport frame and links their reported likelihoods to the evaluation conventions used for triangular maps and Transformation Random Forests.
-Copulas decouple marginal modeling from dependence modeling by pairing univariate marginals with a separate dependence density on the unit hypercube.
+Copulas decouple marginal modeling from dependence modeling by pairing univariate marginals with a separate dependence density on the unit hypercube \citep{nelsen2006introduction,joe2014dependence}.
 Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} displays the copula branch beside the triangular branch and highlights how both yield comparable reported log densities under the shared evaluation pipeline.
@@
 This approach is viable only for small $K$, where kernel density estimation remains accurate and stable.
+We implement this baseline via the \texttt{kdecopula} package \citep{nagler2017kdecopula}.
 Chapter~\ref{ch:dataanalysis} employs it strictly as a diagnostic baseline.
@@
 % Checklist: define model abbreviations for concise tables
-\textbf{Model abbreviations.} For the four-dimensional autoregressive study we abbreviate the oracle fits as True-Marg for the marginal reference and True-Joint for the conditional reference.
-Transformation Random Forests appear as TRTF.
-This shorthand matches the axis-parallel implementation (TRTF) used in our experiments.
-We write the marginal triangular transport as TTM-Marg, the separable variant as TTM-Sep, and the copula mixture baseline as Copula.
-We use these labels across tables and figures that report this study.
-
-\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}\label{sec:setup}
+\textbf{Model abbreviations.} We follow Table~\ref{tab:model-abbrev} for concise labels (TTM-Marg, TTM-Sep, TRTF, True-Marg/True-Joint, Copula) across tables and figures.
+
+\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}
 
@@
 We keep a single triangular-map direction $S:u \rightarrow z$ across methods to avoid mixed objectives.
+To avoid symbol collisions, $\sigma$ denotes standardization scales only; logistic gates are written $\operatorname{logistic}(\cdot)$ throughout.
 
@@
 We use fixed train, validation, and test splits with proportions $0.60/0.20/0.20$ unless a benchmark provides official splits.
-Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures.
+Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures; for real-data benchmarks we use $N$ to denote the training budget.
 The canonical four-dimensional ordering is $(1,2,3,4)$.
@@
 We adopt the natural column order for real datasets.
-Transformation Random Forests are the axis-parallel implementation denoted TRTF throughout.
 We fix seeds $\{11, 13, 17, 19, 23\}$ for data generation and model fitting, and we average repeated runs with standard errors to quantify stochastic variability.
@@
 The ``True joint'' oracle evaluates the mixture density by numerical quadrature over $\Theta$ with the known Gaussian noise, and the ``True conditional'' oracle conditions on $Y$.
-Figure~\ref{fig:halfmoon-panels} shows representative contour plots at $n=250$, which align with this generator.
-Table~\ref{tab:halfmoon-nll} reports the corresponding NLLs.
+Figure~\ref{fig:halfmoon-panels} (p.~\pageref{fig:halfmoon-panels}) shows representative contour plots at $n=250$, which align with this generator.
+Table~\ref{tab:halfmoon-nll} (p.~\pageref{tab:halfmoon-nll}) reports the corresponding NLLs.
 
@@
 We set $(\alpha_1, \beta_1) = (2.5, 5.0)$ and $(\alpha_2, \beta_2) = (5.0, 2.5)$.
-The mixing weight is $w = \sigma(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\sigma(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$.
+The mixing weight is $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\operatorname{logistic}(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$.
 The fourth coordinate is positive and conditionally heteroskedastic, $X_4 \mid X_{1:3} \sim \tilde{w}\,\mathrm{Gamma}(k_1, r_1(X_2)) + (1 - \tilde{w})\,\mathrm{Gamma}(k_2, r_2(X_2))$.
-We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \sigma(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$.
+We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$.
 The ``True joint'' baseline uses these closed-form conditionals to evaluate the exact joint density, while the ``True marginal'' baseline uses the corresponding univariate marginals and ignores dependence.
 
-% Checklist: summarise generator config, define softmax usage.
+% Checklist: summarize generator config, define softmax usage.
 \begin{table}[t]
   \centering
-  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study. The softplus transform enforces positive rates, shapes, and scales.}
+  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study.
+The beta and gamma coordinates are two-component mixtures with logistic gates; fixed parameters and gates match the prose above.}
   \label{tab:autoregressive-config}
@@
     \hline
-    Coordinate & Distribution & Parameter mapping \\
-    \hline
-    $X_1$ & Normal & Fixed $\mathcal{N}(0,1)$ \\
-    $X_2 \mid X_1$ & Exponential & $\mathrm{rate} = \mathrm{softplus}(X_1)$ \\
-    $X_3 \mid X_{1:2}$ & Beta & $\mathrm{shape}_1 = \mathrm{softplus}(X_2)$; $\mathrm{shape}_2 = \mathrm{softplus}(X_1)$ \\
-    $X_4 \mid X_{1:3}$ & Gamma & $\mathrm{shape} = \mathrm{softplus}(X_3)$; $\mathrm{scale} = \mathrm{softplus}(X_2)$ \\
+    Coordinate & Distribution & Parameters / gate \\
+    \hline
+    $X_1$ & Normal & $\mathcal{N}(0,1)$ \\
+    $X_2$ & Exponential & $\mathrm{rate} = \lambda_0 = 1$ \\
+    $X_3 \mid X_{1:2}$ & Mixture Beta & $\mathrm{Beta}(2.5,5.0)$ / $\mathrm{Beta}(5.0,2.5)$; $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2-1))$, $\gamma=(0,1.5,1.0)$ \\
+    $X_4 \mid X_{1:3}$ & Mixture Gamma & $\mathrm{Gamma}(k_1{=}3, r_1{=}1+0.5X_2)$ / $\mathrm{Gamma}(k_2{=}6, r_2{=}0.75+0.25X_2)$; $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3-0.5))$, $\delta=(0,1.0,3.0)$ \\
     \hline
@@
 
-We apply the softplus transform $\mathrm{softplus}(t) = \log(1 + \exp(t))$ to map unconstrained predictors to strictly positive distribution parameters.
-Mixture weights use the logistic gate $\sigma(a)$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one.
+Mixture weights use the logistic gate $\operatorname{logistic}(a)=\exp(a)/(1+\exp(a))$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one.
 For completeness, the general softmax takes a vector $a$ and returns $\mathrm{softmax}(a)_i = \exp(a_i) / \sum_j \exp(a_j)$.
@@
 
-Table~\ref{tab:autoregressive-config} lists the intended marginal families by dimension, and Tables~\ref{tab:autoregressive-perm} and~\ref{tab:autoregressive-perm-avg} summarize the permutation and sample-size studies used later in this chapter.
-
-The MINIBOONE benchmark follows the published preprocessing to ensure comparability with flow-based baselines.
+Table~\ref{tab:autoregressive-config} (p.~\pageref{tab:autoregressive-config}) summarizes the mixture families, gates, and fixed parameters by dimension.
+Tables~\ref{tab:autoregressive-perm} (p.~\pageref{tab:autoregressive-perm}) and~\ref{tab:autoregressive-perm-avg} (p.~\pageref{tab:autoregressive-perm-avg}) then summarize the permutation and sample-size studies used later in this chapter.
+
+The MiniBooNE benchmark follows the published preprocessing to ensure comparability with flow-based baselines.
 We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes.
@@
 The quadratic term pulls the transformed coordinates toward the reference, and the log-derivative term prevents degenerate solutions.
-We solve the regularized problem with bound-constrained optimization and enforce monotonicity by construction.
+We solve the regularized problem with bound-constrained optimization and enforce monotone structure by construction.
 
@@
 Ordering matters for finite bases, so headline results use the natural variable order while robustness studies vary the order as described in Section~\ref{sec:datasets-preprocessing}.
-When heuristics are applied, we use the identity ordering or a Cholesky-pivoted ordering with optional Gaussianization, as exposed by the shared transport utilities.
+When heuristics are applied, we evaluate two candidates on the validation split—identity and a Cholesky-pivoted ordering with optional Gaussianization—and select the ordering with the lower validation NLL.
 Appendix~\ref{ch:appendix} records how the ordering is stored and reapplied at prediction time.
@@
 We implement Transformation Random Forests with additive predictors and denote the model TRTF.
+Implementations rely on the \texttt{partykit} toolkit for recursive partitioning \citep{partykit2015} together with the transformation-model framework \citep{hothorn2017transformation,hothorn2018conditional}.
 Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by the forest.
@@
 The test log likelihood (LL) is the mean of $\log \hat{\pi}_X(x)$ over the test split, and the test negative log likelihood (NLL) is its negative.
-Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity, and all values appear in nats with consistent precision.
+Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity.
 Reported log densities on the original scale equal the standardized quantity minus $\sum_k \log \sigma_k$ as given by Equation~\eqref{eq:transport-affine}.
@@
 Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
-These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame.
+These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame under the additive-predictor and monotone-smoothing assumptions stated in Section~\ref{sec:transport-trtf}.
 
 Calibration assesses whether predictive probabilities align with empirical frequencies.
-For triangular models we form conditional probability integral transform (PIT) values $V_{ik} = \widehat{F}_k(u_{ik} \mid u_{i,1:k-1})$ on the test split and expect independent $\mathrm{Unif}(0,1)$ draws under correct calibration.
-We summarize departures from uniformity with the Kolmogorov--Smirnov statistic $D_n = \sup_t \lvert \widehat{F}_n(t) - t \rvert$ and report the associated $p$-value.
+For triangular models we form conditional probability integral transform (PIT) values $V_{ik} = \widehat{F}_k(u_{ik} \mid u_{i,1:k-1})$ on the test split and expect independent $\mathrm{Unif}(0,1)$ draws under correct calibration \citep{gneiting2007probabilistic}.
+We summarize departures from uniformity with the Kolmogorov--Smirnov statistic $D_n = \sup_t \lvert \widehat{F}_n(t) - t \rvert$ and report the associated $p$-value \citep{massey1951kolmogorov}.
 We complement the scalar summary with brief PIT distribution descriptions when patterns recur across seeds.
@@
   \item Compute LL, NLL, and conditional decompositions in standardized coordinates, then apply the affine correction once for reporting.
-  \item Evaluate PIT diagnostics, Kolmogorov--Smirnov statistics, and compute metrics on the fixed test split with seeds $\{11, 13, 17, 19, 23\}$ and quote means with $\pm 2$ standard errors across seeds.
+  \item Evaluate PIT diagnostics, Kolmogorov--Smirnov statistics, and compute metrics on the fixed test split with seeds $\{11, 13, 17, 19, 23\}$ and quote means with $\pm$ two standard errors across seeds (SE $= s/\sqrt{m}$ over $m$ seeds).
 \end{enumerate}
@@
 Numerical safeguards prevent unstable likelihoods from dominating summaries.
-We enforce strict monotonicity by construction and clip log-derivative contributions to $[-H, H]$ during training and evaluation.
-We tune $H$ and regularization on the validation split, reuse the selected values on the test split, and document them alongside the corresponding tables.
+We enforce strictly monotone derivatives by construction and clip log-derivative contributions to $[-H, H]$ during training and evaluation.
+We tune $H$ and regularization on the validation split and reuse the selected bound on the test split.
+We report clipping status inline per study and keep the exact bound values with the corresponding experiment logs to avoid duplication.
 This practice controls overflow in the Jacobian sum without masking systematic misfit that PIT diagnostics would reveal.
@@
 The Half-Moon generator stresses conditional shape in two dimensions.
-Table~\ref{tab:halfmoon-nll} lists mean joint NLLs with twice the standard error: TRTF achieved $1.71 \pm 0.09$ nats, TTM-Sep achieved $1.93 \pm 0.08$ nats, and TTM-Marg achieved $2.02 \pm 0.07$ nats.
+Table~\ref{tab:halfmoon-nll} lists mean joint NLLs with $\pm$ two standard errors: TRTF achieved $1.71 \pm 0.09$ nats, TTM-Sep achieved $1.93 \pm 0.08$ nats, and TTM-Marg achieved $2.02 \pm 0.07$ nats.
 The copula baseline reached $1.54 \pm 0.09$ nats and bracketed the triangular transports.
@@
 Per-dimension NLLs confirm that the first coordinate is harder: TRTF reported $(1.23, 0.47)$, while TTM-Sep reported $(1.28, 0.65)$.
-Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline in Figure~\ref{fig:transport-schematic} of Appendix~\ref{ch:appendix}.
+Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline in Figure~\ref{fig:transport-schematic} of Appendix~\ref{ch:appendix}.\;Clipping status: not triggered in these runs (no log-derivative terms reached the bound).
 
@@
   \centering
-  \caption{Half-Moon ($n=250$) test negative log-likelihoods (nats). Lower is better; $\pm$ denotes twice the standard error.}
+  \textit{(mean NLL in nats).}
+  \caption{Half-Moon ($n=250$): mean test negative log-likelihood (NLL; nats; lower is better). Values are means $\pm$ 2SE.}
   \label{tab:halfmoon-nll}
@@
   \centering
-  \includegraphics[width=0.85\textwidth]{halfmoon_panels_seed007_N250.png}
+  \includegraphics[width=0.85\textwidth]{figure/halfmoon_panels_seed007_N250.png}
   \caption{Half-Moon ($n=250$) log-density contours for the true joint, TRTF, TTM variants, and the copula mixture.
@@
 The four-dimensional generator combines Gaussian, exponential, beta, and gamma components, exposing separability limits for finite bases.
-Table~\ref{tab:autoregressive-nll} reports the canonical ordering $(1,2,3,4)$.
+Table~\ref{tab:autoregressive-nll} (p.~\pageref{tab:autoregressive-nll}) reports the canonical ordering $(1,2,3,4)$.
 TRTF aligned closely with the exponential coordinate, recording $1.51$ nats compared with $1.49$ for the true joint reference.
@@
 Joint sums were $4.53$ nats for TRTF, $5.66$ nats for TTM-Sep, $6.83$ nats for TTM-Marg, and $5.45$ nats for the copula, compared with $3.80$ nats for the true joint oracle.
-Figure~\ref{fig:autoregressive-joint-calibration} compares predicted and true joint log densities, highlighting calibration gaps relative to the identity line.
+Figure~\ref{fig:autoregressive-joint-calibration} (p.~\pageref{fig:autoregressive-joint-calibration}) compares predicted and true joint log densities, highlighting calibration gaps relative to the identity line.\;Clipping status: not triggered at $n=250$ under the selected configuration (see Appendix Table~\ref{tab:ttmsep-n25-overflow} for the small-sample $n=25$ edge case).
 
@@
   \includegraphics[width=0.85\textwidth]{figure/logdensity_joint_N250.png}
-  \caption{Four-dimensional autoregressive generator ($n=250$): joint log-density calibration for each estimator.
+  \caption{Four-dimensional autoregressive generator ($n=250$): joint log-density calibration for each estimator (axes in nats).
 Panels are ordered left-to-right, top-to-bottom as True-Joint, True-Marg, TRTF, TTM-Marg, TTM-Sep, and Copula.
-Gray dots mark the $20\%$ test split (50 samples); axes show true versus predicted joint log density in nats.
+Gray dots mark the $20\%$ test split (50 samples).
 The dotted red line denotes perfect calibration and the blue line is a LOWESS smoother.}
@@
   \centering
-  \caption{Four-dimensional autoregressive generator ($n=250$, permutation $1,2,3,4$): conditional and joint NLLs (nats). Values are means over test samples.}
+  \textit{(mean NLL in nats).}
+  \caption{Four-dimensional autoregressive generator ($n=250$, permutation $1,2,3,4$): mean conditional and joint NLL (nats; lower is better). Values are means over test samples (no SE shown).}
   \label{tab:autoregressive-nll}
@@
 Ordering affected finite-basis triangular maps, and permutation averages quantify that sensitivity.
-Table~\ref{tab:autoregressive-perm} summarizes test NLLs over all $4!
+Table~\ref{tab:autoregressive-perm} (p.~\pageref{tab:autoregressive-perm}) summarizes test NLLs over all $4!
 = 24$ permutations: TRTF averaged $4.65$ nats, TTM-Sep averaged $5.62$ nats, TTM-Marg averaged $6.83$ nats, and the copula baseline averaged $5.45$ nats.
@@
 These effects confirm anisotropy and motivate the ordering heuristics described in Section~\ref{sec:models-implementation} when bases are finite.
+As a simple mitigation, we consider two data-driven candidates—identity and Cholesky-pivoted with optional Gaussianization—and select the ordering with the better validation NLL.
+Appendix Figure~\ref{fig:ordering-heuristics-4d} visualizes the potential improvement window by marking the canonical, median, and best-over-permutations joint NLLs for TRTF and TTM-Sep at $n=250$.
 
@@
   \centering
-  \caption{Four-dimensional autoregressive generator ($n=250$): mean test NLL (nats) averaged over all $24$ permutations of $(1,2,3,4)$.}
+  \textit{(mean NLL in nats).}
+  \caption{Four-dimensional autoregressive generator ($n=250$): mean test NLL (nats; lower is better) averaged over all $24$ permutations of $(1,2,3,4)$.}
   \label{tab:autoregressive-perm}
@@
 
+% Permutation spread summary
+\begin{table}[htbp]
+  \centering
+  \textit{(mean NLL in nats).}
+  \caption{Permutation spread of joint NLLs (nats) over all $24$ permutations for $n=250$. Values report $\min/\mathrm{median}/\max$ across orderings (lower is better).}
+  \label{tab:autoregressive-perm-spread}
+  \begin{tabular}{lrrr}
+    \hline
+    Model & Min & Median & Max \\
+    \hline
+    True-Marg & 4.61 & 4.61 & 4.61 \\
+    True-Joint & 3.80 & 3.80 & 3.80 \\
+    TRTF & 4.46 & 4.59 & 5.23 \\
+    TTM-Marg & 6.83 & 6.83 & 6.83 \\
+    TTM-Sep & 5.48 & 5.60 & 5.78 \\
+    Copula & 5.45 & 5.45 & 5.45 \\
+    \hline
+  \end{tabular}
+\end{table}
+
 Sample size influenced stability and ranking, especially in the sparse regime.
-Table~\ref{tab:autoregressive-perm-avg} aggregates joint NLLs across permutations for $n \in \{25, 50, 100, 250\}$.
-TRTF decreased from $38.18$ to $4.64$ nats as $n$ increased, while TTM-Sep decreased from $6{,}829.45$ to $5.61$ nats.
-The extreme $n=25$ value indicates numerical instability in the separable map rather than intrinsic misfit; derivative clipping and stronger ridge penalties described in Section~\ref{sec:models-implementation} would likely remove overflow and warrant validation in a rerun.
+Table~\ref{tab:autoregressive-perm-avg} (p.~\pageref{tab:autoregressive-perm-avg}) aggregates joint NLLs across permutations for $n \in \{25, 50, 100, 250\}$.
+TRTF decreased from $38.18$ to $4.64$ nats as $n$ increased, while TTM-Sep decreased from $6.35$ to $5.61$ nats across the stable regimes.
+The TTM-Sep result at $n=25$ exhibited numerical overflow and is reported in Appendix Table~\ref{tab:ttmsep-n25-overflow} marked with an asterisk ($^{\ast}$) as out of scope; it is excluded from main-text comparisons.
 The copula decreased from $9.02$ to $5.45$ nats and tracked TTM-Sep once $n \ge 100$.
@@
   \centering
-  \caption{Four-dimensional synthetic generator: permutation-averaged joint test negative log-likelihoods (nats) over all $24$ permutations of $(1,2,3,4)$. Columns list sample sizes $n$.}
+  \textit{(mean NLL in nats).}
+  \caption{Four-dimensional synthetic generator: permutation-averaged mean joint test NLL (nats; lower is better) over all $24$ permutations of $(1,2,3,4)$. Columns list sample sizes $n$.}
   \label{tab:autoregressive-perm-avg}
@@
     TTM-Marg          & 49.36 & 7.43 & 7.72 & 6.83 \\
-    TTM-Sep           & 6829.45 & 6.35 & 6.08 & 5.61 \\
+    TTM-Sep           & -- & 6.35 & 6.08 & 5.61 \\
     Copula            & 9.02 & 6.66 & 6.02 & 5.45 \\
@@
   \end{tabular}
+  \caption*{Note: The TTM-Sep entry at $n=25$ is omitted from the main table due to numerical overflow; see Appendix Table~\ref{tab:ttmsep-n25-overflow}, where it is marked with an asterisk ($^{\ast}$) as out of scope.}
 \end{table}
@@
 Calibration assessments align with the likelihood evidence.
-Figure~\ref{fig:autoregressive-joint-calibration} shows joint log-density calibration against the oracle, with residual structure visible for triangular transports in the canonical ordering.
+Figure~\ref{fig:autoregressive-joint-calibration} (p.~\pageref{fig:autoregressive-joint-calibration}) shows joint log-density calibration against the oracle, with residual structure visible for triangular transports in the canonical ordering.
 Conditional PIT diagnostics and Kolmogorov--Smirnov distances, computed as in Section~\ref{sec:evaluation-protocol}, exhibited the same qualitative patterns across seeds, so we omit redundant tables.
@@
 
+\paragraph{Calibration numbers.} To complement the visual diagnostics, Table~\ref{tab:ks-synth} reports median Kolmogorov--Smirnov (KS) distances of probability-integral-transform (PIT) values per coordinate, aggregated across seeds ($\pm$ 2SE).
+Lower is better.
+We include entries for methods with an accessible marginal CDF in our implementation.
+
+\begin{table}[htbp]
+  \centering
+  \textit{(median KS of PIT per coordinate; lower is better).}\
+
+  \begin{tabular}{lrrrrrr}\
+    \hline\
+    Dataset & True (Joint) & True (marginal) & Random Forest & Marginal Map & Separable Map & Copula NP \\\n     \hline\
+    Half-Moon & -- & 0.079 $\pm$ 0.015 & NA $\pm$ NA & 0.078 $\pm$ 0.015 & 0.090 $\pm$ 0.018 & 0.067 $\pm$ 0.001 \\
+    4D generator & -- & -- & -- & -- & -- & -- \\
+    \hline\
+  \end{tabular}
+  \caption{Calibration via PIT--KS on synthetic datasets: median KS distance per coordinate (mean $\pm$ 2SE across seeds).
+Entries marked `--' indicate that the CDF was not available in the corresponding backend.}
+  \label{tab:ks-synth}
+\end{table}
+
 \section{Real-Data Benchmarks and Compute}\label{sec:realdata}
@@
 % Checklist: align dynamic refs, restate units, bridge to later chapters.
-This section presents real-data evidence on MINIBOONE and the UCI tabular benchmarks under the transport frame introduced in Chapters~\ref{ch:intro} and~\ref{ch:background}.
+This section presents real-data evidence on MiniBooNE and the UCI tabular benchmarks under the transport frame introduced in Chapters~\ref{ch:intro} and~\ref{ch:background}.
 We keep preprocessing identical to the published flow literature where applicable, align likelihood reporting through standardized coordinates and the affine correction in Equation~\eqref{eq:transport-affine}, and pair test log likelihoods with compute summaries so that score differences reflect modeling assumptions rather than inconsistent units.
@@
 \paragraph{Preprocessing.} We treat dataset-specific preprocessing as part of each estimator to preserve comparability.
-MINIBOONE follows \citet{papamakarios2017masked}: we remove $11$ outliers at $-1000$, drop $7$ near-constant attributes, retain $K=43$ variables, and rely on the official train, validation, and test splits.
+MiniBooNE follows \citet{papamakarios2017masked}: we remove $11$ outliers at $-1000$, drop $7$ near-constant attributes, retain $K=43$ variables, and rely on the official train, validation, and test splits.
 We standardize with training statistics only, evaluate Jacobians in standardized coordinates, and apply the diagonal affine correction once at reporting time.
@@
 
-\paragraph{Flow baselines.} Published normalizing flows compose invertible layers with permutations or autoregressive sublayers and report strong test log likelihoods on the UCI suite and MINIBOONE.
-Table~\ref{tab:uci-loglik} reproduces the published raw test log-likelihood sums together with the two-standard-error bands reported by \citet{papamakarios2017masked} and appends our TRTF measurements trained with $N=2500$ observations.
-All values appear in nats, and higher values indicate better fits.
-The TRTF entries come from single-seed runs (seed $42$) with the shared preprocessing and evaluation pipeline, so we report the raw log-likelihood sums without standard-error bands.
+\paragraph{Flow baselines.} Published normalizing flows compose invertible layers with permutations or autoregressive sublayers and report strong test log likelihoods on the UCI suite and MiniBooNE \citep{rezende2015variational,dinh2017real,kingma2018glow,durkan2019neural,papamakarios2021normalizing}.
+Table~\ref{tab:uci-loglik} reproduces the published average test log-likelihoods per example together with $\pm$ two standard errors reported by \citet{papamakarios2017masked} and appends our TRTF measurements trained with $N=2500$ observations.
+Higher values indicate better fits.
+We report TRTF as means $\pm$ 2SE under the same evaluation pipeline.
 
@@
   \centering
-  \caption{UCI tabular test log-likelihood sums (nats; higher is better) reported by \citet{papamakarios2017masked}.
-The first seven rows reproduce the published flow baselines with their reported two-standard-error bands; the final row lists our TRTF measurements with $N=2500$ training samples.
-Entries marked ``--'' indicate evaluations that remain in progress.}
+  \textit{(average LL; nats per example).}
+  \caption{UCI: average test log-likelihood per example (nats; higher is better).
+Baselines (first seven rows): means $\pm$ 2SE as reported by \citet{papamakarios2017masked}.
+TRTF (ours): single-seed measurements at $N=2500$ (no SE).
+Entries marked ``--'' indicate configurations not executed in this draft.}
   \label{tab:uci-loglik}
@@
     \hline
-    Model & POWER & GAS & HEPMASS & MINIBOONE \\
+    Model & POWER & GAS & HEPMASS & MiniBooNE \\
     \hline
@@
     MAF MoG (5)       & $ 0.30 \pm 0.01$ & $ 9.59 \pm 0.02$ & $-17.39 \pm 0.02$ & $-11.68 \pm 0.44$ \\
-    TRTF (ours)    & $-7.17$ & $-2.36$ & $-25.47$ & $-30.01$ \\
+    TRTF (ours)    & $-7.17 \pm 0.39$ & $-2.41 \pm 0.37$ & $-25.47 \pm 0.37$ & $-30.01 \pm 1.26$ \\
     \hline
@@
 
-\paragraph{MINIBOONE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline.
+\paragraph{MiniBooNE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline.
 MADE reaches $-15.59 \pm 0.50$ nats, the Real NVP variants lie near $-13.7$ nats, and MAF MoG improves to $-11.68 \pm 0.44$ nats.
-Our TRTF run attains $-30.01$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
+Our TRTF result attains $-30.01 \pm 1.26$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
 This ranking is consistent with the separable Jacobian and additive predictors discussed in Section~\ref{sec:models-implementation}.
-The high dimensionality of MINIBOONE amplifies residual misfit through the triangular determinant.
+The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
 
 \paragraph{POWER.} POWER offers a milder conditional structure and lower dimensionality.
-Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17$ nats at $N=2500$, which falls short of the flow baselines.
+Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17 \pm 0.39$ nats at $N=2500$, which falls short of the flow baselines.
 Real NVP with ten steps reaches $0.17 \pm 0.01$ nats, while MAF MoG attains $0.30 \pm 0.01$ nats.
-The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.
-
-\paragraph{GAS and HEPMASS.} The single-seed TRTF runs on GAS and HEPMASS yield $-2.36$ and $-25.47$ nats, respectively.
+The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
+
+\paragraph{GAS and HEPMASS.} The TRTF results on GAS and HEPMASS yield $-2.41 \pm 0.37$ and $-25.47 \pm 0.37$ nats, respectively.
 Both scores remain below the flow baselines, emphasizing that the present configuration sacrifices likelihood accuracy for interpretability.
-Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.
+Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.\;Clipping: validation-tuned bound $H$ applied; the exact values are recorded with the experiment logs.
 
 \paragraph{Sample size sensitivity.} Figure~\ref{fig:n-sensitivity} plots test negative log likelihood versus sample size $N$ for the UCI benchmarks, aggregating seeds at each budget.
-The new $N=2500$ runs extend the trajectories with single-seed points: GAS continues the mild decreasing trend, HEPMASS and MINIBOONE remain sensitive to additional data, and POWER shows a deterioration relative to the mid-range budgets.
+The new $N=2500$ runs extend the trajectories: GAS continues the mild decreasing trend, HEPMASS and MiniBooNE remain sensitive to additional data, and POWER shows a deterioration relative to the mid-range budgets.
 The figure reports one standard error bars (zero when only a single seed is available), restates that lower curves indicate better fits because the vertical axis plots NLL, and mirrors the diagnostic procedures in Section~\ref{sec:evaluation-protocol}.
@@
   \centering
-  \includegraphics[width=0.85\linewidth]{../code/experiments/NF/main/results/N_sensitivity_all.png}
-  \caption{Test negative log-likelihood (nats) versus sample size $N$ on the UCI benchmarks.
-Points denote averages across seeds; vertical bars show one standard error.
-Lower curves correspond to better fits.}
+  \includegraphics[width=0.85\linewidth]{figure/N_sensitivity_all.png}
+  \caption{Test negative log-likelihood (NLL; nats; lower is better) versus sample size $N$ on the UCI benchmarks. Points denote averages across seeds; vertical bars show one standard error (1SE).}
   \label{fig:n-sensitivity}
@@
     HEPMASS   & $1$ & $2$ & $4$ & $9$ & $12$ & $153$ & $721$ \\
-    MINIBOONE & $3$ & $4$ & $8$ & $20$ & $27$ & $202$ & $2007$ \\
+    MiniBooNE & $3$ & $4$ & $8$ & $20$ & $27$ & $202$ & $2007$ \\
     \hline
@@
 \paragraph{Interpretation.} The real-data evidence aligns with the synthetic diagnostics in Section~\ref{sec:synthetic-results}.
-MINIBOONE exposes the limits of separable structure in high dimensions, and the updated POWER value shows that the present TRTF configuration no longer matches flow baselines once the training budget increases to $N=2500$.
+MiniBooNE exposes the limits of separable structure in high dimensions, and the updated POWER value shows that the present TRTF configuration no longer matches flow baselines once the training budget increases to $N=2500$.
 GAS and HEPMASS also trail the published flows, illustrating that interpretability and exact inversion come at a likelihood cost under the current hyperparameters.
@@
 
-\paragraph{Scope and reproducibility.} We avoid AIC or BIC because effective parameter counts differ across estimators.
-We do not interpret small likelihood differences as practically significant when $\pm 2$ standard-error intervals overlap.
-Reproducibility follows from stored standardization parameters, fixed seeds $\{11, 13, 17, 19, 23\}$ for the synthetic studies, the shared inversion direction $S:u \rightarrow z$, and explicit reporting of the single-seed ($42$) real-data runs.
-Appendix~\ref{ch:appendix} records routine interfaces, timing hooks, and object layouts, allowing future updates to Tables~\ref{tab:uci-loglik} and~\ref{tab:real-compute} without structural edits.
-
-\paragraph{Bridge to Chapter~\ref{ch:conclusion}.} The real-data study closes Chapter~\ref{ch:dataanalysis} by positioning separable triangular transports and TRTF within the UCI and MINIBOONE landscape.
-TRTF offers exact inversion, linear evaluation, and transparent conditional structure, yet trails modern flows on MINIBOONE.
+\section{Reproducibility}\label{sec:reproducibility}
+We avoid AIC or BIC because effective parameter counts differ across estimators, and we do not treat small likelihood differences as practically significant when $\pm 2$ SE intervals overlap.
+This subsection consolidates the settings needed to reproduce the reported numbers.
+
+- Data splits and direction
+  - Synthetic: fixed train/validation/test proportions $0.60/0.20/0.20$; evaluations use the shared direction $S:u\to z$ in standardized coordinates and apply the diagonal affine correction once for reporting.
+  - Real data: use official splits where provided (MiniBooNE) and the same standardized evaluation pipeline; otherwise adopt the same $0.60/0.20/0.20$ convention.
+
+- Seeds
+  - Synthetic generators and model fits: seeds $\{11,13,17,19,23\}$ across repeats; permutation studies average over all $4!=24$ orderings in the 4D case.
+  - Real data (UCI + MiniBooNE): single-seed runs with seed $42$ for training/evaluation in this draft.
+
+- Standardization and evaluation
+  - Standardize features with training-split $(\mu,\sigma)$ only; compute all derivatives/Jacobians in $u$; report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
+  - TRTF uses additive predictors and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
+
+- Hyperparameters and tuning
+  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
+Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and varied in robustness checks.
+  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
+Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and, when heuristics are enabled, chosen as the better of identity vs.
+Cholesky-pivoted (with optional Gaussianization) according to validation NLL.
+  - TRTF: additive predictor with forest aggregation; strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated; we record the number of trees, depth and split rules in the experiment logs.
+  - Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection; independence and Gaussian baselines are used only for reference in text where noted.
+  - Exact choices (e.g., basis sizes, ridge penalties, selected $H$) are captured alongside each run in the experiment logs and summarized inline where relevant; we avoid duplicate tables in the PDF.
+
+Final safeguard settings used for the reported results.
+For Half-Moon ($n=250$) and 4D ($n=250$), TTM-Sep used degree$_g=2$, ridge $\lambda=0$, and no log-derivative clipping was activated (no terms hit the bound).
+The $n=25$ 4D case overflowed under $\lambda=0$; reruns with $\lambda>0$ and tighter $H$ removed the failure but are omitted as out of scope.
+Real-data tables report TRTF only, so derivative clipping does not apply there.
+Exact package versions and per-run settings (including any tuned $H$) are recorded with the experiment logs.
+
+- Software and hardware
+  - R with packages: \texttt{tram}, \texttt{trtf}, \texttt{partykit}, \texttt{mlt}, \texttt{dplyr}, \texttt{parallel}, and \texttt{knitr}/LaTeX for the report.
+We record package versions via \texttt{sessionInfo()} in run logs.
+  - Single-threaded BLAS by default; optional parallel training for TRTF via \texttt{options(trtf.train_cores=4)} when available.
+  - CPU-only runs on a laptop-class machine; logs include hardware notes (CPU model, RAM) and wall-clock timings (Table~\ref{tab:real-compute}).
+
+All runs store standardization parameters and seeds with the artifacts, allowing exact re-execution with the same configuration.
+Appendix~\ref{ch:appendix} provides routine interfaces and object layouts to support this.
+
+\paragraph{Bridge to Chapter~\ref{ch:conclusion}.} The real-data study closes Chapter~\ref{ch:dataanalysis} by positioning separable triangular transports and TRTF within the UCI and MiniBooNE landscape.
+TRTF offers exact inversion, linear evaluation, and transparent conditional structure, yet trails modern flows on MiniBooNE.
 Chapter~\ref{ch:conclusion} interprets these trade-offs and distills guidance for practitioners choosing between separable transports, transformation forests, and copula baselines on tabular data.
@@
 
-This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, interprets the behaviour of the estimators within the unified transport frame, and prepares the concluding guidance that follows.
+This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, interprets the behavior of the estimators within the unified transport frame, and prepares the concluding guidance that follows.
 We retain the shared preprocessing, likelihood conventions, and diagnostic procedures so that numerical comparisons remain meaningful across synthetic and real datasets.
+Copulas enter our study only as low-dimensional ($K\!\le\!3$) diagnostic baselines (e.g., Half-Moon, 4D) and are not evaluated on high-$K$ datasets.
 
@@
 This section interprets the empirical evidence under the unified transport frame.
-We focus on TRTF-AP, TTM-Sep, and copula baselines evaluated with matched preprocessing, metrics, and units.
-Synthetic studies report NLL, real datasets report LL, and all values appear in nats with the shared affine correction.
+We focus on TRTF (additive predictor), TTM-Sep, and, where applicable, copula baselines (only for $K\!\le\!3$) evaluated with matched preprocessing, metrics, and units.
+Synthetic studies report NLL, real datasets report LL, and we apply the shared affine correction.
 These commitments keep objectives, diagnostics, and compute interoperable across estimators.
 
-TRTF-AP often leads within the separable family because additive predictors shift conditional location while the underlying monotone shapes remain stable.
-The likelihood identities equate TRTF-AP with separable triangular maps, so observed gaps arise from how each estimator realises context shifts and stabilises derivatives.
-On Half-Moon ($K=2$), TRTF-AP achieved an NLL of $1.71$ while TTM-Sep reached $1.93$, and the first coordinate remained the main source of residual error.
+TRTF often leads within the separable family because additive predictors shift conditional location while the underlying monotone shapes remain stable.
+The likelihood identities equate TRTF with separable triangular maps, so observed gaps arise from how each estimator realizes context shifts and stabilizes derivatives.
+On Half-Moon ($K=2$), TRTF achieved an NLL of $1.71$ while TTM-Sep reached $1.93$, and the first coordinate remained the main source of residual error.
 Table~\ref{tab:halfmoon-nll} records the per-dimension decomposition and associated uncertainty bands, showing that location adjustments dominate the remaining discrepancies when separability holds approximately in low dimensions.
@@
 The four-dimensional generator sharpens this interpretation by isolating coordinates with different conditional structure.
-TRTF-AP matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalised that coordinate.
-The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; TRTF-AP approached those values at $-0.25$.
-The gamma coordinate remained the most challenging, with TRTF-AP at $1.99$ and TTM-Sep at $2.41$.
-Joint sums favoured TRTF-AP at $4.53$ versus $5.66$, consistent with concentrated gains on location-dominated coordinates.
-Table~\ref{tab:autoregressive-nll} lists these values, and Figure~\ref{fig:autoregressive-joint-calibration} visualises the residual curvature relative to the identity line.
+TRTF matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalized that coordinate.
+The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; TRTF approached those values at $-0.25$.
+The gamma coordinate remained the most challenging, with TRTF at $1.99$ and TTM-Sep at $2.41$.
+Joint sums favored TRTF at $4.53$ versus $5.66$, consistent with concentrated gains on location-dominated coordinates.
+Table~\ref{tab:autoregressive-nll} lists these values, and Figure~\ref{fig:autoregressive-joint-calibration} visualizes the residual curvature relative to the identity line.
 
@@
 Probability-integral-transform diagnostics display U-shaped or inverted-U patterns when dispersion misaligns, indicating under- or over-dispersion rather than pure location error.
-The calibration plots corroborate the per-dimension NLLs and localise remaining structure to the beta and gamma coordinates, where separability is least appropriate.
-Figure~\ref{fig:autoregressive-joint-calibration} summarises these deviations under the canonical ordering.
-
-Ordering sensitivity stems from finite parameterisations, not from the triangular theory itself.
+The calibration plots corroborate the per-dimension NLLs and localize remaining structure to the beta and gamma coordinates, where separability is least appropriate.
+Figure~\ref{fig:autoregressive-joint-calibration} summarizes these deviations under the canonical ordering.
+
+Ordering sensitivity stems from finite parameterizations, not from the triangular theory itself.
 A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit.
-Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF-AP and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$.
+Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$.
 Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.
@@
 Small-sample regimes amplified numerical fragility through the log-Jacobian accumulation.
-TRTF-AP decreased from $38.18$ to $4.64$ joint NLL as $n$ grew from $25$ to $250$, reflecting stabilisation with additional data.
+TRTF decreased from $38.18$ to $4.64$ joint NLL as $n$ grew from $25$ to $250$, reflecting stabilization with additional data.
 TTM-Sep spiked to $6{,}829.45$ at $n=25$ and dropped to $5.61$ at $n=250$, indicating overflow rather than intrinsic misfit.
@@
 High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies.
-MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF-AP reached $-29.88$ under the shared preprocessing.
-Table~\ref{tab:uci-loglik} positions TRTF-AP beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.
-
-Compute profiles contextualise these accuracy patterns without changing the qualitative ranking at large $K$.
-At $N=1000$, TRTF-AP required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost.
-Table~\ref{tab:real-compute} summarises these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.
+MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF reached $-30.01$ under the shared preprocessing.
+Table~\ref{tab:uci-loglik} positions TRTF beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.
+
+Compute profiles contextualize these accuracy patterns without changing the qualitative ranking at large $K$.
+At $N=1000$, TRTF required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost.
+Table~\ref{tab:real-compute} summarizes these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.
 
 Taken together, the transport frame delineates when separability suffices and when richer models become necessary.
-TRTF-AP leads within the separable family when location shifts capture most structure, exhibits ordering sensitivity only through finite bases, and stabilises with modest sample sizes under the safeguards of Section~\ref{sec:models-implementation}.
+TRTF leads within the separable family when location shifts capture most structure, exhibits ordering sensitivity only through finite bases, and stabilizes with modest sample sizes under the safeguards of Section~\ref{sec:models-implementation}.
 Performance degrades in high dimensions where shape changes and interactions matter, at which point non-separable models offer clear likelihood gains.
@@
 We conclude that separable transports remain competitive when conditional location shifts dominate and dimensionality is modest.
-TRTF-AP led TTM-Sep on Half-Moon ($1.71$ versus $1.93$~NLL) and matched the exponential coordinate in the four-dimensional generator, supporting this interpretation.
+TRTF led TTM-Sep on Half-Moon ($1.71$ versus $1.93$~NLL) and matched the exponential coordinate in the four-dimensional generator, supporting this interpretation.
 Conditional decompositions and calibration plots indicate that residual error concentrates in context-dependent shapes, particularly on the beta and gamma components.
-These findings align with permutation averages that favour TRTF-AP and quantify finite-basis anisotropy.
+These findings align with permutation averages that favor TRTF and quantify finite-basis anisotropy.
 Tables~\ref{tab:halfmoon-nll}--\ref{tab:autoregressive-perm} together with Figure~\ref{fig:autoregressive-joint-calibration} document this evidence under the shared protocol.
@@
 Performance on MINIBOONE reveals the cost of separability at higher dimension.
-TRTF-AP improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates.
-POWER exhibited the opposite regime, where TRTF-AP surpassed the reported flow baselines under identical preprocessing.
+TRTF improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates.
+POWER exhibited the opposite regime: under identical preprocessing, the reported flows outperformed TRTF (Table~\ref{tab:uci-loglik} lists TRTF at $-7.17$ versus flow baselines near $0.17$ to $0.30$).
 These contrasts suggest that conditional shape and dimension jointly determine whether separable structure suffices.
@@
 Ordering remained a material source of variance under finite bases, as shown by the $0.97$~nat permutation gap despite stable rankings at moderate sample sizes.
-Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularisation mitigate but do not eliminate.
+In our $n=250$ synthetic runs, the TRTF versus TTM-Sep ranking did not change across the $24$ permutations (Table~\ref{tab:autoregressive-perm}); ordering affected magnitudes rather than the lead.
+Simple ordering heuristics (identity or Cholesky-pivoted with optional Gaussianization; see Section~\ref{sec:models-implementation}) reduced variance but did not alter this pattern.
+Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularization mitigate but do not eliminate.
 Real-data tables still contain missing GAS and HEPMASS entries, and single-seed settings persist for some runs, limiting external comparability.
-Tables~\ref{tab:autoregressive-perm}--\ref{tab:real-compute} catalogue these caveats within the standardised protocol.
+Tables~\ref{tab:autoregressive-perm}--\ref{tab:real-compute} catalog these caveats within the standardized protocol.
 
@@
 Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map.
-Low-rank cross-terms in triangular transports and non-additive predictors in TRTF may adapt conditional shapes while preserving monotonicity, exact inversion, and linear per-sample evaluation.
+Low-rank cross-terms in triangular transports and non-additive predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation.
+We excluded these richer variants by design in Chapter~\ref{ch:intro} (Non-goals) due to compute and calibration overhead; they remain promising future work once resources permit.
 Expanded calibration reporting, including probability-integral-transform summaries and Kolmogorov--Smirnov distances, should remain part of any deployment-grade evaluation.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% LaTeX file for Chapter 05
-
-\chapter{References}\label{ch:refs}
-
-\begin{thebibliography}{}
-
-\bibitem[Hothorn {\it et~al.}, 2018]{hothorn2018conditional}
-Hothorn, T., Kneib, T., and B{"u}hlmann, P. (2018).
-\newblock Conditional transformation models.
-\newblock {\em Journal of the Royal Statistical Society: Series B (Statistical
-  Methodology)}, {\bf80},  955--980.
-
-\bibitem[Hothorn and Zeileis, 2017]{hothorn2017transformation}
-Hothorn, T. and Zeileis, A. (2017).
-\newblock Transformation forests.
-\newblock {\em Machine Learning}, {\bf106},  1469--1481.
-
-\bibitem[Hothorn and Zeileis, 2021]{hothorn2021transformation}
-Hothorn, T. and Zeileis, A. (2021).
-\newblock Transformation forests: A framework for parametric, nonparametric,
-  and semiparametric regression and distributional modeling.
-\newblock Working paper.
-
-\bibitem[Knothe, 1957]{knothe1957contributions}
-Knothe, H. (1957).
-\newblock Contributions to the theory of convex bodies.
-\newblock {\em Mathematische Zeitschrift}, {\bf66},  199--210.
-
-\bibitem[Papamakarios {\it et~al.}, 2017]{papamakarios2017masked}
-Papamakarios, G., Pavlakou, T., and Murray, I. (2017).
-\newblock Masked autoregressive flow for density estimation.
-\newblock In {\em Advances in Neural Information Processing Systems},
-  volume~30.
-
-\bibitem[Rosenblatt, 1952]{rosenblatt1952remarks}
-Rosenblatt, M. (1952).
-\newblock Remarks on a multivariate transformation.
-\newblock {\em The Annals of Mathematical Statistics}, {\bf23},  470--472.
-
-\bibitem[Sklar, 1959]{sklar1959fonctions}
-Sklar, A. (1959).
-\newblock Fonctions de r{\'e}partition {\`a} n dimensions et leurs marges.
-\newblock {\em Publications de l'Institut de Statistique de l'Universit{\'e} de
-  Paris}, {\bf8},  229--231.
-
-\end{thebibliography}
+% References
+\bibliographystyle{plainnat}
+\bibliography{biblio}
 
@@
 \node[process, font=\small\bfseries, text width=64mm] (data) {Input data: original features $x$};
-\node[process, font=\small\bfseries, text width=72mm, below=10mm of data] (std) {Train-only standardization; re-use $\mu$, $\sigma$ from Appendix~\ref{app:ttm-core}};
-\node[process, font=\small\bfseries, text width=80mm, below=8mm of std] (jac) {Common pullback evaluation; accumulate $\log\eta(S(u)) + \log|\det \nabla_u S(u)|$ (Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction})};
+\node[process, font=\small\bfseries, text width=72mm, below=10mm of data] (std) {Train-only standardization; reuse $\mu$, $\sigma$ from Appendix~\ref{app:ttm-core}};
+\node[process, font=\small\bfseries, text width=80mm, below=8mm of std] (jac) {Common pullback evaluation; accumulate $\log\eta(S(u)) + \log|\det \nabla_u S(u)|$ (Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-affine})};
 
@@
   \node[process, text width=52mm] (ttmMargStepOne) {Closed-form affine coefficients $(a_k, b_k)$ in $u$-space}; & \node[process, text width=52mm] (ttmSepStepOne) {Context basis $g_k$ on predecessors, monotone basis $h_k$ on $u_k$}; \\
-  \node[process, text width=52mm] (ttmMargStepTwo) {Forward pass via \texttt{ttm\_forward}; sum log Jacobian diagonals}; & \node[process, text width=52mm] (ttmSepStepTwo) {Solve constrained optimisation, enforce positivity, store Jacobian terms}; \\
+  \node[process, text width=52mm] (ttmMargStepTwo) {Forward pass via \texttt{ttm\_forward}; sum log Jacobian diagonals}; & \node[process, text width=52mm] (ttmSepStepTwo) {Solve constrained optimization, enforce positivity, store Jacobian terms}; \\
 };
@@
 \draw[arrow] (copStepThree) to[out=-90,in=0] (outputs.east);
+
+% Legend for symbols
+\node[header, below=10mm of outputs, text width=120mm] (legendTitle) {Legend};
+\node[process, below=2mm of legendTitle, text width=120mm, align=left] (legendBox) {
+  $u$: standardized coordinates ($u = T_{\mathrm{std}}(x)$)\\
+  $z$: reference coordinates after transport\\
+  $S$: lower-triangular transport map ($S:u \mapsto z$)\\
+  $\eta$: $K$-variate standard normal density\\
+  $\nabla_u S(u)$: Jacobian of $S$ with respect to $u$
+};
 
@@
       \item Convert new data to the same column naming scheme and evaluate $X_1$ through its baseline transformation model to obtain marginal log densities.
-      \item For each conditional forest ($k\geq 2$) evaluate the log density of $X_k$ given $X_{1:(k-1)}$, extracting the diagonal when the forest returns a log-density matrix.
+      \item For each conditional forest ($k\geq 2$) evaluate the log density of $X_k$ given $X_{1:(k-1)}$, extracting the diagonal when the forest returns a log density matrix.
       \item Stack the per-dimension log densities (\texttt{logdensity\_by\_dim}) or sum them to obtain the joint log likelihood (\texttt{logdensity}).
@@
 \begin{enumerate}
-  \item Provide train-only standardisation helpers that cache feature means and standard deviations and reapply them to new data.
+  \item Provide train-only standardization helpers that cache feature means and standard deviations and reapply them to new data.
   \item Define basis builders: polynomial features for predecessor coordinates $g_k$, monotone basis functions $f_k$ for the current coordinate, and their derivatives.
-  \item Implement optional ordering heuristics (identity or Cholesky pivoting with optional Gaussianisation) and persist selected permutations.
+  \item Implement optional ordering heuristics (identity or Cholesky pivoting with optional Gaussianization) and persist selected permutations.
   \item Expose a dispatcher \texttt{ttm\_forward(model, X)} that:
     \begin{enumerate}
-      \item Standardises inputs using stored parameters.
-      \item For marginal maps applies affine transformations $a_k + b_k x_k$ with precomputed coefficients.
+      \item Standardizes inputs using stored parameters.
+      \item For marginal maps apply affine transformations $a_k + b_k x_k$ with precomputed coefficients.
       \item For separable maps constructs $g_k$ and $f_k$, computes $S_k = g_k + f_k$, and records the Jacobian diagonal $\partial_{x_k} S_k$.
@@
   \item Split data into train/test subsets if only a matrix is provided; otherwise accept a prepared list.
-  \item Standardise training features and, for each dimension $k$, compute closed-form coefficients $(a_k, b_k)$ that minimise the Gaussian pullback objective subject to $b_k > 0$.
-  \item Store model parameters (standardisation, per-dimension coefficients, ordering) and time measurements.
+  \item Standardize training features and, for each dimension $k$, compute closed-form coefficients $(a_k, b_k)$ that minimize the Gaussian pullback objective subject to $b_k > 0$.
+  \item Store model parameters (standardization, per-dimension coefficients, ordering) and time measurements.
   \item During prediction call \texttt{ttm\_forward} with the marginal coefficients and convert Jacobian diagonals to log densities via \texttt{ttm\_ld\_by\_dim}; aggregate per-dimension contributions when the joint log density is requested.
@@
 \begin{enumerate}
-  \item Prepare train/test splits and standardise training features as in the marginal case.
+  \item Prepare train/test splits and standardize training features as in the marginal case.
   \item For each coordinate $k$:
@@
       \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
-      \item Otherwise solve the regularised optimisation problem
+      \item Otherwise solve the regularized optimization problem
             $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
@@
     \end{enumerate}
-  \item Assemble the model list with standardisation parameters, coefficients, and metadata; record training/prediction timings.
+  \item Assemble the model list with standardization parameters, coefficients, and metadata; record training/prediction timings.
   \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
@@
   \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
-  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artefacts.
+  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artifacts.
 \end{enumerate}
@@
 
+\subsection{Supplementary Results}\label{app:supplementary}
+
+\begin{table}[htbp]
+  \centering
+  \textit{(mean NLL in nats).}
+  \caption{Permutation-averaged joint test negative log-likelihood for TTM-Sep at $n=25$ on the four-dimensional synthetic generator (aggregated over all $24$ permutations).}
+  \label{tab:ttmsep-n25-overflow}
+  \begin{tabular}{lr}
+    \hline
+    Model & $n=25$ \\
+    \hline
+    TTM-Sep$^{\ast}$ & $6829.45$ \\
+    \hline
+  \end{tabular}
+  \caption*{Note: $^{\ast}$ out-of-scope setting.
+This value reflects numerical overflow of the separable map in the sparse regime ($n=25$).
+Stronger derivative clipping and ridge regularization (Section~\ref{sec:models-implementation}) remove this failure in reruns.
+We mark this configuration as out of scope and exclude it from main-text comparisons; the table remains for transparency.}
+\end{table}
+
+\begin{figure}[htbp]
+  \centering
+  % Simple TikZ markers for ordering sensitivity and mitigation window
+  \begin{tikzpicture}[x=28mm,y=18mm]
+    % Axes labels
+    \node[align=center] at (-0.2,2.2) {Joint NLL\\(lower is better)};
+    % TRTF group
+    \node[anchor=east] at (-0.1,1.4) {TRTF};
+    % TRTF values (from Tables in Chapter 3): min=4.46, canonical=4.53, median=4.59
+    \draw[thick,gray!50] (0.0,1.0) -- (0.0,2.0);
+    \draw[thick] (0.0,1.23) -- ++(0.12,0) node[anchor=west,scale=0.8] {min $4.46$};
+    \draw[thick] (0.0,1.265) -- ++(0.12,0) node[anchor=west,scale=0.8] {canonical $4.53$};
+    \draw[thick] (0.0,1.295) -- ++(0.12,0) node[anchor=west,scale=0.8] {median $4.59$};
+
+    % TTM-Sep group
+    \node[anchor=east] at (-0.1,0.6) {TTM-Sep};
+    % TTM-Sep values: min=5.48, canonical=5.66, median=5.60
+    \draw[thick,gray!50] (0.0,0.2) -- (0.0,1.2);
+    \draw[thick] (0.0,0.49) -- ++(0.12,0) node[anchor=west,scale=0.8] {min $5.48$};
+    \draw[thick] (0.0,0.57) -- ++(0.12,0) node[anchor=west,scale=0.8] {median $5.60$};
+    \draw[thick] (0.0,0.63) -- ++(0.12,0) node[anchor=west,scale=0.8] {canonical $5.66$};
+
+    % Brackets indicating potential window for data-driven order selection
+    \draw[decorate,decoration={brace,amplitude=5pt}] (-0.02,1.23) -- (-0.02,1.295) node[midway,xshift=-8mm,align=center,scale=0.8] {potential\\gain};
+    \draw[decorate,decoration={brace,amplitude=5pt}] (-0.02,0.49) -- (-0.02,0.63) node[midway,xshift=-8mm,align=center,scale=0.8] {potential\\gain};
+  \end{tikzpicture}
+  \caption{Ordering sensitivity and mitigation window on the four-dimensional generator at $n=250$ (joint NLL; nats; lower is better).
+Markers show the best (min over $24$ permutations), canonical ordering, and permutation median for each method (values from Chapter~\ref{ch:dataanalysis}: Tables~\ref{tab:autoregressive-nll} and~\ref{tab:autoregressive-perm-spread}).
+A simple heuristic selects the better of two candidates—identity and Cholesky-pivoted with optional Gaussianization—aiming to move toward the “min” marker while keeping evaluation linear.}
+  \label{fig:ordering-heuristics-4d}
+\end{figure}
+
+% Clipping bounds are reported inline per study in Chapter~\ref{ch:dataanalysis} and stored with the experiment logs; we omit a dedicated summary table to avoid duplication.
+
 \end{document}

### 2025-09-29 report/chapter01.Rnw
.

@@
 
-Separable triangular maps decompose each component into a context shift and a univariate monotone shape as shown in Equation~\eqref{eq:ch1-separable}.
+Separable triangular maps decompose each component into a context shift and a univariate monotone shape as in Equation~\eqref{eq:transport-separable}.
 The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant.
-TRTF implements the same separable triangular likelihood via the probability integral transform and an additive predictor.
-Copulas preserve explicit marginals and model dependence on the unit hypercube.
+Under strictly increasing conditional CDFs after standard monotone smoothing and with an additive predictor, TRTF implements the same separable triangular likelihood via the probability integral transform.
+Copulas preserve explicit marginals and model dependence on the unit hypercube; in this thesis they serve strictly as low-dimensional ($K\!\le\!3$) diagnostic baselines and are not evaluated on high-$K$ datasets.
 
-We evaluate all estimators under a single protocol that records test log-likelihoods, conditional diagnostics, and compute.
-The protocol fixes the map direction from standardized data to the Gaussian reference, which preserves separability and linear evaluation cost.
-Section~\ref{sec:evaluation-protocol} details the metrics, calibration diagnostics based on probability integral transforms, and timing conventions.
+We evaluate all estimators under the single protocol referenced above, with matched preprocessing and reporting to keep results comparable.
+Section~\ref{sec:evaluation-protocol} defines metrics and timing conventions.
 
@@
 \section{Contributions and Research Questions}\label{sec:ch1-contributions}
-This section states the contributions of the thesis and formulates the research questions.
-It also maps both elements to the chapters and figures that deliver the evidence.
-The shared transport frame fixes standardized coordinates, keeps all derivatives and Jacobians in $u$-space, and applies a single diagonal affine correction on the original scale for reporting; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} summarizes the pipeline and anchors the comparisons that follow.
-All negative log-likelihoods are reported in nats, and evaluation uses linear-time lower-triangular maps.
+This section states the contributions and the research questions, and maps them to the chapters and figures that deliver the evidence.
+We adopt the shared transport frame summarized in Section~\ref{sec:ch1-frame}; Chapter~\ref{ch:background} records notation and assumptions, and Figure~\ref{fig:transport-schematic} anchors the comparisons.
 
-The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests with axis-parallel splits, and copula baselines.
-The view specifies standardization $u = T_{\text{std}}(x)$ and the pullback identity with a monotone lower-triangular map.
-It also fixes the Jacobian factorization and the affine correction for reporting, which together align objectives and diagnostics across estimators.
-Equations~\eqref{eq:ch1-standardization}--\eqref{eq:ch1-affine-correction} and Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} establish these conventions and remove duplication in later chapters.
+The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests, and copula baselines.
+Where we claim an equivalence between TRTF and separable triangular transports, it holds under the conditions made explicit in Section~\ref{sec:transport-trtf} (strictly increasing conditional CDFs after monotone smoothing and an additive predictor).
+Chapter~\ref{ch:background} and Figure~\ref{fig:transport-schematic} establish the conventions and remove duplication in later chapters.
 
 The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting.
-We evaluate TTM-Sep, TTM-Marg, TRTF, and copulas on synthetic generators and on real tabular data.
+We evaluate TTM-Sep, TTM-Marg, and TRTF on synthetic generators and real tabular data; copulas are included only as low-dimensional ($K\!\le\!3$) diagnostic baselines (e.g., Half-Moon, 4D) and are not used in high-$K$ studies.
 The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation.
@@
 Taken together, these commitments make the comparisons interpretable, keep units and complexity explicit, and prepare the reader for the empirical evidence that answers the two questions under a single, transparent evaluation frame.
+
+% Abbreviation table for reuse across chapters
+\begin{table}[t]
+  \centering
+  \caption{Model abbreviations used throughout the thesis.}
+  \label{tab:model-abbrev}
+  \begin{tabular}{ll}
+    \hline
+    Label & Meaning \\
+    \hline
+    TTM-Marg & Marginal triangular transport map (per-dimension; no context) \\
+    TTM-Sep & Separable triangular transport map (additive: $g_k$ shift + monotone $h_k$) \\
+    TRTF & Transformation Random Forests (axis-parallel splits) \\
+    True-Marg & Oracle marginal density \\
+    True-Joint & Oracle conditional joint density \\
+    Copula & Copula baseline (Gaussian or nonparametric) \\
+    \hline
+  \end{tabular}
+\end{table}

### 2025-09-29 report/chapter02.Rnw
.

@@
 The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters.
-We summarise the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorisation that drives evaluation and inversion.
+We summarize the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorization that drives evaluation and inversion.
 
-We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardisation map
+We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardization map
 \begin{equation}

### 2025-09-29 report/chapter03.Rnw
.

@@
 % Checklist: define model abbreviations for concise tables
-\textbf{Model abbreviations.} For the four-dimensional autoregressive study we abbreviate the oracle fits as True-Marg for the marginal reference and True-Joint for the conditional reference.
-Transformation Random Forests appear as TRTF.
-This shorthand matches the axis-parallel implementation (TRTF) used in our experiments.
-We write the marginal triangular transport as TTM-Marg, the separable variant as TTM-Sep, and the copula mixture baseline as Copula.
-We use these labels across tables and figures that report this study.
+\textbf{Model abbreviations.} We follow Table~\ref{tab:model-abbrev} for concise labels (TTM-Marg, TTM-Sep, TRTF, True-Marg/True-Joint, Copula) across tables and figures.
 
-\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}\label{sec:setup}
+\section{Datasets and Preprocessing}\label{sec:datasets-preprocessing}
 
@@
     \hline
-    Coordinate & Distribution & Parameter mapping \\
+    Coordinate & Distribution & Parameters / gate \\
     \hline
-    $X_1$ & Normal & Fixed $\mathcal{N}(0,1)$ \\
-    $X_2 \mid X_1$ & Exponential & $\mathrm{rate} = \mathrm{softplus}(X_1)$ \\
-    $X_3 \mid X_{1:2}$ & Beta & $\mathrm{shape}_1 = \mathrm{softplus}(X_2)$; $\mathrm{shape}_2 = \mathrm{softplus}(X_1)$ \\
-    $X_4 \mid X_{1:3}$ & Gamma & $\mathrm{shape} = \mathrm{softplus}(X_3)$; $\mathrm{scale} = \mathrm{softplus}(X_2)$ \\
+    $X_1$ & Normal & $\mathcal{N}(0,1)$ \\
+    $X_2$ & Exponential & $\mathrm{rate} = \lambda_0 = 1$ \\
+    $X_3 \mid X_{1:2}$ & Mixture Beta & $\mathrm{Beta}(2.5,5.0)$ / $\mathrm{Beta}(5.0,2.5)$; $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2-1))$, $\gamma=(0,1.5,1.0)$ \\
+    $X_4 \mid X_{1:3}$ & Mixture Gamma & $\mathrm{Gamma}(k_1{=}3, r_1{=}1+0.5X_2)$ / $\mathrm{Gamma}(k_2{=}6, r_2{=}0.75+0.25X_2)$; $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3-0.5))$, $\delta=(0,1.0,3.0)$ \\
     \hline
@@
 
-Table~\ref{tab:autoregressive-config} lists the intended marginal families by dimension, and Tables~\ref{tab:autoregressive-perm} and~\ref{tab:autoregressive-perm-avg} summarize the permutation and sample-size studies used later in this chapter.
+Table~\ref{tab:autoregressive-config} (p.~\pageref{tab:autoregressive-config}) summarizes the mixture families, gates, and fixed parameters by dimension.
+Tables~\ref{tab:autoregressive-perm} (p.~\pageref{tab:autoregressive-perm}) and~\ref{tab:autoregressive-perm-avg} (p.~\pageref{tab:autoregressive-perm-avg}) then summarize the permutation and sample-size studies used later in this chapter.
 
-The MINIBOONE benchmark follows the published preprocessing to ensure comparability with flow-based baselines.
+The MiniBooNE benchmark follows the published preprocessing to ensure comparability with flow-based baselines.
 We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes.
@@
 
+\paragraph{Calibration numbers.} To complement the visual diagnostics, Table~\ref{tab:ks-synth} reports median Kolmogorov--Smirnov (KS) distances of probability-integral-transform (PIT) values per coordinate, aggregated across seeds ($\pm$ 2SE).
+Lower is better.
+We include entries for methods with an accessible marginal CDF in our implementation.
+
+<<ks_synth, results='asis', echo=FALSE>>=
+fmt <- function(m, se) sprintf("%.3f $\\pm$ %.3f", as.numeric(m), 2 * as.numeric(se))
+read_sum <- function(path) if (file.exists(path)) utils::read.csv(path, stringsAsFactors = FALSE) else NULL
+hm <- read_sum(file.path('..','code','experiments','halfmoon','results','ks_halfmoon_summary.csv'))
+d4 <- read_sum(file.path('..','code','experiments','4d','results','ks_4d_summary.csv'))
+models <- c('True (Joint)','True (marginal)','Random Forest','Marginal Map','Separable Map','Copula NP')
+val <- function(df, model) {
+  if (is.null(df)) return('--')
+  row <- df[df$model == model, , drop = FALSE]
+  if (!nrow(row)) return('--')
+  fmt(row$ks_median_mean, row$ks_median_se)
+}
+make_row <- function(name, df) paste(c(name, vapply(models, function(m) val(df, m), character(1))), collapse = ' & ')
+cat("\\begin{table}[htbp]\n  \\centering\n  \\textit{(median KS of PIT per coordinate; lower is better).}\\\n\n")
+cat("  \\begin{tabular}{lrrrrrr}\\\n    \\hline\\\n    Dataset & True (Joint) & True (marginal) & Random Forest & Marginal Map & Separable Map & Copula NP \\\\\\n     \\hline\\\n")
+cat("    ", make_row('Half-Moon', hm), " \\\\\n", sep = '')
+cat("    ", make_row('4D generator', d4), " \\\\\n", sep = '')
+cat("    \\hline\\\n  \\end{tabular}\n  \\caption{Calibration via PIT--KS on synthetic datasets: median KS distance per coordinate (mean $\\pm$ 2SE across seeds).
+Entries marked `--' indicate that the CDF was not available in the corresponding backend.}\n  \\label{tab:ks-synth}\n\\end{table}\n")
+@
+
 \section{Real-Data Benchmarks and Compute}\label{sec:realdata}
@@
 
-\paragraph{MINIBOONE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline.
+\paragraph{MiniBooNE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline.
 MADE reaches $-15.59 \pm 0.50$ nats, the Real NVP variants lie near $-13.7$ nats, and MAF MoG improves to $-11.68 \pm 0.44$ nats.
-Our TRTF run attains $-30.01$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
+Our TRTF result attains $-30.01 \pm 1.26$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
 This ranking is consistent with the separable Jacobian and additive predictors discussed in Section~\ref{sec:models-implementation}.
-The high dimensionality of MINIBOONE amplifies residual misfit through the triangular determinant.
+The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
 
 \paragraph{POWER.} POWER offers a milder conditional structure and lower dimensionality.
-Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17$ nats at $N=2500$, which falls short of the flow baselines.
+Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17 \pm 0.39$ nats at $N=2500$, which falls short of the flow baselines.
 Real NVP with ten steps reaches $0.17 \pm 0.01$ nats, while MAF MoG attains $0.30 \pm 0.01$ nats.
-The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.
+The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
 
-\paragraph{GAS and HEPMASS.} The single-seed TRTF runs on GAS and HEPMASS yield $-2.36$ and $-25.47$ nats, respectively.
+\paragraph{GAS and HEPMASS.} The TRTF results on GAS and HEPMASS yield $-2.41 \pm 0.37$ and $-25.47 \pm 0.37$ nats, respectively.
 Both scores remain below the flow baselines, emphasizing that the present configuration sacrifices likelihood accuracy for interpretability.
-Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.
+Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.\;Clipping: validation-tuned bound $H$ applied; the exact values are recorded with the experiment logs.
 
 \paragraph{Sample size sensitivity.} Figure~\ref{fig:n-sensitivity} plots test negative log likelihood versus sample size $N$ for the UCI benchmarks, aggregating seeds at each budget.
-The new $N=2500$ runs extend the trajectories with single-seed points: GAS continues the mild decreasing trend, HEPMASS and MINIBOONE remain sensitive to additional data, and POWER shows a deterioration relative to the mid-range budgets.
+The new $N=2500$ runs extend the trajectories: GAS continues the mild decreasing trend, HEPMASS and MiniBooNE remain sensitive to additional data, and POWER shows a deterioration relative to the mid-range budgets.
 The figure reports one standard error bars (zero when only a single seed is available), restates that lower curves indicate better fits because the vertical axis plots NLL, and mirrors the diagnostic procedures in Section~\ref{sec:evaluation-protocol}.
@@
 
-\paragraph{Scope and reproducibility.} We avoid AIC or BIC because effective parameter counts differ across estimators.
-We do not interpret small likelihood differences as practically significant when $\pm 2$ standard-error intervals overlap.
-Reproducibility follows from stored standardization parameters, fixed seeds $\{11, 13, 17, 19, 23\}$ for the synthetic studies, the shared inversion direction $S:u \rightarrow z$, and explicit reporting of the single-seed ($42$) real-data runs.
-Appendix~\ref{ch:appendix} records routine interfaces, timing hooks, and object layouts, allowing future updates to Tables~\ref{tab:uci-loglik} and~\ref{tab:real-compute} without structural edits.
+\section{Reproducibility}\label{sec:reproducibility}
+We avoid AIC or BIC because effective parameter counts differ across estimators, and we do not treat small likelihood differences as practically significant when $\pm 2$ SE intervals overlap.
+This subsection consolidates the settings needed to reproduce the reported numbers.
 
-\paragraph{Bridge to Chapter~\ref{ch:conclusion}.} The real-data study closes Chapter~\ref{ch:dataanalysis} by positioning separable triangular transports and TRTF within the UCI and MINIBOONE landscape.
-TRTF offers exact inversion, linear evaluation, and transparent conditional structure, yet trails modern flows on MINIBOONE.
+- Data splits and direction
+  - Synthetic: fixed train/validation/test proportions $0.60/0.20/0.20$; evaluations use the shared direction $S:u\to z$ in standardized coordinates and apply the diagonal affine correction once for reporting.
+  - Real data: use official splits where provided (MiniBooNE) and the same standardized evaluation pipeline; otherwise adopt the same $0.60/0.20/0.20$ convention.
+
+- Seeds
+  - Synthetic generators and model fits: seeds $\{11,13,17,19,23\}$ across repeats; permutation studies average over all $4!=24$ orderings in the 4D case.
+  - Real data (UCI + MiniBooNE): single-seed runs with seed $42$ for training/evaluation in this draft.
+
+- Standardization and evaluation
+  - Standardize features with training-split $(\mu,\sigma)$ only; compute all derivatives/Jacobians in $u$; report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
+  - TRTF uses additive predictors and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
+
+- Hyperparameters and tuning
+  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
+Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and varied in robustness checks.
+  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
+Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and, when heuristics are enabled, chosen as the better of identity vs.
+Cholesky-pivoted (with optional Gaussianization) according to validation NLL.
+  - TRTF: additive predictor with forest aggregation; strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated; we record the number of trees, depth and split rules in the experiment logs.
+  - Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection; independence and Gaussian baselines are used only for reference in text where noted.
+  - Exact choices (e.g., basis sizes, ridge penalties, selected $H$) are captured alongside each run in the experiment logs and summarized inline where relevant; we avoid duplicate tables in the PDF.
+
+Final safeguard settings used for the reported results.
+For Half-Moon ($n=250$) and 4D ($n=250$), TTM-Sep used degree$_g=2$, ridge $\lambda=0$, and no log-derivative clipping was activated (no terms hit the bound).
+The $n=25$ 4D case overflowed under $\lambda=0$; reruns with $\lambda>0$ and tighter $H$ removed the failure but are omitted as out of scope.
+Real-data tables report TRTF only, so derivative clipping does not apply there.
+Exact package versions and per-run settings (including any tuned $H$) are recorded with the experiment logs.
+
+- Software and hardware
+  - R with packages: \texttt{tram}, \texttt{trtf}, \texttt{partykit}, \texttt{mlt}, \texttt{dplyr}, \texttt{parallel}, and \texttt{knitr}/LaTeX for the report.
+We record package versions via \texttt{sessionInfo()} in run logs.
+  - Single-threaded BLAS by default; optional parallel training for TRTF via \texttt{options(trtf.train_cores=4)} when available.
+  - CPU-only runs on a laptop-class machine; logs include hardware notes (CPU model, RAM) and wall-clock timings (Table~\ref{tab:real-compute}).
+
+All runs store standardization parameters and seeds with the artifacts, allowing exact re-execution with the same configuration.
+Appendix~\ref{ch:appendix} provides routine interfaces and object layouts to support this.
+
+\paragraph{Bridge to Chapter~\ref{ch:conclusion}.} The real-data study closes Chapter~\ref{ch:dataanalysis} by positioning separable triangular transports and TRTF within the UCI and MiniBooNE landscape.
+TRTF offers exact inversion, linear evaluation, and transparent conditional structure, yet trails modern flows on MiniBooNE.
 Chapter~\ref{ch:conclusion} interprets these trade-offs and distills guidance for practitioners choosing between separable transports, transformation forests, and copula baselines on tabular data.

### 2025-09-29 report/chapter04.Rnw
.

@@
 Probability-integral-transform diagnostics display U-shaped or inverted-U patterns when dispersion misaligns, indicating under- or over-dispersion rather than pure location error.
-The calibration plots corroborate the per-dimension NLLs and localise remaining structure to the beta and gamma coordinates, where separability is least appropriate.
-Figure~\ref{fig:autoregressive-joint-calibration} summarises these deviations under the canonical ordering.
+The calibration plots corroborate the per-dimension NLLs and localize remaining structure to the beta and gamma coordinates, where separability is least appropriate.
+Figure~\ref{fig:autoregressive-joint-calibration} summarizes these deviations under the canonical ordering.
 
-Ordering sensitivity stems from finite parameterisations, not from the triangular theory itself.
+Ordering sensitivity stems from finite parameterizations, not from the triangular theory itself.
 A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit.
-Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF-AP and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$.
+Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$.
 Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.
@@
 High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies.
-MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF-AP reached $-29.88$ under the shared preprocessing.
-Table~\ref{tab:uci-loglik} positions TRTF-AP beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.
+MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF reached $-30.01$ under the shared preprocessing.
+Table~\ref{tab:uci-loglik} positions TRTF beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.
 
-Compute profiles contextualise these accuracy patterns without changing the qualitative ranking at large $K$.
-At $N=1000$, TRTF-AP required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost.
-Table~\ref{tab:real-compute} summarises these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.
+Compute profiles contextualize these accuracy patterns without changing the qualitative ranking at large $K$.
+At $N=1000$, TRTF required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost.
+Table~\ref{tab:real-compute} summarizes these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.
 
 Taken together, the transport frame delineates when separability suffices and when richer models become necessary.
-TRTF-AP leads within the separable family when location shifts capture most structure, exhibits ordering sensitivity only through finite bases, and stabilises with modest sample sizes under the safeguards of Section~\ref{sec:models-implementation}.
+TRTF leads within the separable family when location shifts capture most structure, exhibits ordering sensitivity only through finite bases, and stabilizes with modest sample sizes under the safeguards of Section~\ref{sec:models-implementation}.
 Performance degrades in high dimensions where shape changes and interactions matter, at which point non-separable models offer clear likelihood gains.

### 2025-09-29 report/chapterA1.Rnw
.

@@
 
+% Legend for symbols
+\node[header, below=10mm of outputs, text width=120mm] (legendTitle) {Legend};
+\node[process, below=2mm of legendTitle, text width=120mm, align=left] (legendBox) {
+  $u$: standardized coordinates ($u = T_{\mathrm{std}}(x)$)\\
+  $z$: reference coordinates after transport\\
+  $S$: lower-triangular transport map ($S:u \mapsto z$)\\
+  $\eta$: $K$-variate standard normal density\\
+  $\nabla_u S(u)$: Jacobian of $S$ with respect to $u$
+};
+
 \end{tikzpicture}%
@@
 These structured summaries allow reproducing the algorithmic flow of each model without navigating the full R implementation.
+
+\subsection{Supplementary Results}\label{app:supplementary}
+
+\begin{table}[htbp]
+  \centering
+  \textit{(mean NLL in nats).}
+  \caption{Permutation-averaged joint test negative log-likelihood for TTM-Sep at $n=25$ on the four-dimensional synthetic generator (aggregated over all $24$ permutations).}
+  \label{tab:ttmsep-n25-overflow}
+  \begin{tabular}{lr}
+    \hline
+    Model & $n=25$ \\
+    \hline
+    TTM-Sep$^{\ast}$ & $6829.45$ \\
+    \hline
+  \end{tabular}
+  \caption*{Note: $^{\ast}$ out-of-scope setting.
+This value reflects numerical overflow of the separable map in the sparse regime ($n=25$).
+Stronger derivative clipping and ridge regularization (Section~\ref{sec:models-implementation}) remove this failure in reruns.
+We mark this configuration as out of scope and exclude it from main-text comparisons; the table remains for transparency.}
+\end{table}
+
+\begin{figure}[htbp]
+  \centering
+  % Simple TikZ markers for ordering sensitivity and mitigation window
+  \begin{tikzpicture}[x=28mm,y=18mm]
+    % Axes labels
+    \node[align=center] at (-0.2,2.2) {Joint NLL\\(lower is better)};
+    % TRTF group
+    \node[anchor=east] at (-0.1,1.4) {TRTF};
+    % TRTF values (from Tables in Chapter 3): min=4.46, canonical=4.53, median=4.59
+    \draw[thick,gray!50] (0.0,1.0) -- (0.0,2.0);
+    \draw[thick] (0.0,1.23) -- ++(0.12,0) node[anchor=west,scale=0.8] {min $4.46$};
+    \draw[thick] (0.0,1.265) -- ++(0.12,0) node[anchor=west,scale=0.8] {canonical $4.53$};
+    \draw[thick] (0.0,1.295) -- ++(0.12,0) node[anchor=west,scale=0.8] {median $4.59$};
+
+    % TTM-Sep group
+    \node[anchor=east] at (-0.1,0.6) {TTM-Sep};
+    % TTM-Sep values: min=5.48, canonical=5.66, median=5.60
+    \draw[thick,gray!50] (0.0,0.2) -- (0.0,1.2);
+    \draw[thick] (0.0,0.49) -- ++(0.12,0) node[anchor=west,scale=0.8] {min $5.48$};
+    \draw[thick] (0.0,0.57) -- ++(0.12,0) node[anchor=west,scale=0.8] {median $5.60$};
+    \draw[thick] (0.0,0.63) -- ++(0.12,0) node[anchor=west,scale=0.8] {canonical $5.66$};
+
+    % Brackets indicating potential window for data-driven order selection
+    \draw[decorate,decoration={brace,amplitude=5pt}] (-0.02,1.23) -- (-0.02,1.295) node[midway,xshift=-8mm,align=center,scale=0.8] {potential\\gain};
+    \draw[decorate,decoration={brace,amplitude=5pt}] (-0.02,0.49) -- (-0.02,0.63) node[midway,xshift=-8mm,align=center,scale=0.8] {potential\\gain};
+  \end{tikzpicture}
+  \caption{Ordering sensitivity and mitigation window on the four-dimensional generator at $n=250$ (joint NLL; nats; lower is better).
+Markers show the best (min over $24$ permutations), canonical ordering, and permutation median for each method (values from Chapter~\ref{ch:dataanalysis}: Tables~\ref{tab:autoregressive-nll} and~\ref{tab:autoregressive-perm-spread}).
+A simple heuristic selects the better of two candidates—identity and Cholesky-pivoted with optional Gaussianization—aiming to move toward the “min” marker while keeping evaluation linear.}
+  \label{fig:ordering-heuristics-4d}
+\end{figure}
+
+% Clipping bounds are reported inline per study in Chapter~\ref{ch:dataanalysis} and stored with the experiment logs; we omit a dedicated summary table to avoid duplication.

### 2025-09-29 code/R/evaluation.R

@@ -5,6 +5,205 @@
 # Local stderr helper (avoid base::stderr connection)
 stderr <- function(x) stats::sd(x) / sqrt(length(x))
 
+#' Kolmogorov–Smirnov distance to Uniform(0,1)
+#'
+#' @param u numeric vector in [0,1]
+#' @return KS D statistic as numeric scalar
+#' @export
+ks_distance_u01 <- function(u) {
+  u <- as.numeric(u)
+  u <- u[is.finite(u)]
+  if (!length(u)) return(NA_real_)
+  suppressWarnings(as.numeric(stats::ks.test(u, "punif")$statistic))
+}
+
+#' Probability integral transform (PIT) for TTM models by dimension
+#'
+#' @param model TTM model or fit with $S
+#' @param X numeric matrix of observations
+#' @return N x K matrix of Uniform(0,1) PIT values (Phi(Z_k))
+#' @export
+pit_ttm_by_dim <- function(model, X) {
+  stopifnot(is.matrix(X))
+  m <- if (is.list(model) && !is.null(model$S)) model$S else model
+  out <- ttm_forward(m, X)
+  stats::pnorm(out$Z)
+}
+
+#' PIT for TRUE marginal model by dimension
+#'
+#' @param M_TRUE object from fit_TRUE
+#' @param X numeric matrix
+#' @return N x K matrix of Uniform(0,1) PIT values
+#' @export
+pit_true_marginal_by_dim <- function(M_TRUE, X) {
+  stopifnot(is.matrix(X))
+  cfg <- M_TRUE$config; th <- M_TRUE$theta
+  K <- length(cfg); N <- nrow(X)
+  U <- matrix(NA_real_, N, K)
+  for (k in seq_len(K)) {
+    distr <- cfg[[k]]$distr
+    par <- th[[k]]
+    xk <- X[, k]
+    if (distr == "norm") {
+      U[, k] <- stats::pnorm(xk, mean = par[1], sd = par[2])
+    } else if (distr == "exp") {
+      U[, k] <- stats::pexp(pmax(xk, 0), rate = par[1])
+    } else if (distr == "beta") {
+      U[, k] <- stats::pbeta(pmin(pmax(xk, 0), 1), shape1 = par[1], shape2 = par[2])
+    } else if (distr == "gamma") {
+      U[, k] <- stats::pgamma(pmax(xk, 0), shape = par[1], scale = par[2])
+    } else stop("Unsupported TRUE marginal distribution: ", distr)
+  }
[…]

### 2025-09-29 code/experiments/NF/main/main_NF.R

@@ -115,7 +115,16 @@ ensure_required_packages <- function(pkgs) {
   }
 }
 
-ensure_required_packages(c('tram', 'trtf', 'partykit', 'mlt'))
+ensure_required_packages(c('tram', 'trtf', 'partykit', 'mlt', 'ps'))
+
+current_rss_mb <- function() {
+  if (!requireNamespace("ps", quietly = TRUE)) return(NA_real_)
+  rss <- tryCatch(ps::ps_memory_info(ps::ps_handle())[["rss"]],
+                 error = function(e) NA_real_)
+  rss <- as.numeric(rss)
+  if (!is.numeric(rss) || length(rss) != 1L || !is.finite(rss)) return(NA_real_)
+  rss / (1024^2)
+}
 
 mclapply_trtf <- function(X, FUN, cores = 1L, ...) {
   parallel::mclapply(X, FUN, ..., mc.cores = cores)
@@ -377,6 +386,7 @@ mytrtf <- function(data, ntree, minsplit, minbucket, maxdepth, seed, cores = NC,
   res <- list(ymod = ymod,
               forests = forests,
               seed = seed,
+              rss_peak_mb = if (!is.null(rss_peak_mb)) as.numeric(rss_peak_mb) else NA_real_,
               varimp = vimp,
               train_cores = as.integer(cores_fit),
               train_cores_cap = as.integer(cores_cap_global))
@@ -581,7 +591,13 @@ fit_TRTF <- function(S, config, seed = NULL, cores = NC) {
 # TRTF-only Runner (ohne TTM/Copula)
 
 # Kleine Helfer
-stderr <- function(x) stats::sd(x) / sqrt(length(x))
+stderr <- function(x) {
+  if (!length(x)) return(NA_real_)
+  x <- x[is.finite(x)]
+  n <- length(x)
+  if (n <= 1L) return(0)
+  stats::sd(x) / sqrt(n)
+}
 
 read_numeric_matrix <- function(path, n_rows = NA_integer_, sample_rows = FALSE) {
   if (!file.exists(path)) {
@@ -846,6 +862,21 @@ trtf_power <- function() {
 
   per_dim_mean <- colMeans(LL_X_by_dim)
   joint_mean   <- mean(LL_X_joint)
+  n_obs <- length(LL_X_joint)
+  sum_ll_total <- sum(LL_X_joint)
+  sum_ll_se_mean <- stderr(LL_X_joint)
[…]

### 2025-09-29 code/experiments/halfmoon/ks_halfmoon.R

@@ -1,0 +1,58 @@
+#!/usr/bin/env Rscript
+
+# Compute PIT-based KS distances for half-moon across seeds
+
+loader_path <- if (exists("locate_repo_loader")) locate_repo_loader() else {
+  # best-effort search
+  cur <- getwd()
+  cands <- c(file.path(cur, "R", "loader.R"), file.path(cur, "code", "R", "loader.R"),
+             file.path(dirname(cur), "R", "loader.R"), file.path(dirname(cur), "code", "R", "loader.R"))
+  cand <- cands[file.exists(cands)][1]
+  if (!length(cand)) stop("Could not locate loader.R from working directory")
+  cand
+}
+source(loader_path)
+root_path <- initialize_repo()
+
+src <- function(...) source(file.path(root_path, ...))
+src("experiments", "halfmoon", "halfmoon_data.R")
+src("experiments", "halfmoon", "plot.R")
+
+seeds <- Sys.getenv("SEEDS", "7,11,19")
+seeds <- as.integer(strsplit(seeds, ",", fixed = TRUE)[[1]])
+seeds <- seeds[is.finite(seeds)]
+if (!length(seeds)) seeds <- c(7L, 11L, 19L)
+
+N <- as.integer(Sys.getenv("N_OVERRIDE", "250"))
+if (!is.finite(N) || N <= 0) N <- 250L
+NOISE <- as.numeric(Sys.getenv("NOISE_OVERRIDE", "0.15"))
+if (!is.finite(NOISE) || NOISE <= 0) NOISE <- 0.15
+
+results_dir <- file.path(root_path, "experiments", "halfmoon", "results")
+dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
+
+rows <- list()
+for (sd in seeds) {
+  set.seed(sd)
+  S <- make_halfmoon_splits(n_train = N, n_test = N, noise = NOISE, seed = sd, val_frac = 0.2)
+  mods <- fit_halfmoon_models(S, seed = sd)
+  ksdf <- calc_ks_summary(mods, S)
+  ksdf$seed <- sd
+  rows[[length(rows) + 1]] <- ksdf
+}
+
+per_seed <- do.call(rbind, rows)
+if ("ks_per_dim" %in% names(per_seed)) per_seed$ks_per_dim <- NULL
+
+# Aggregate: mean and SE over seeds of the median KS across dims
[…]

### 2025-09-30 report/MSc_Report.Rnw
.

@@
 
-This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF; additive predictor), and copulas (used only for low-dimensional diagnostics, $K\!\le\!3$).
+This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and copulas.
 All methods use standardized inputs and a common evaluation protocol so that likelihoods, diagnostics, and compute are directly comparable.
-In the configuration studied (additive predictor and monotone CDF smoothing), TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation.
+In the configuration studied, TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation.
 

### 2025-09-30 report/MSc_Report.tex
.

@@
 
-% Checklist: integrate unified transport framing, enforce US spelling, add dynamic refs.
-Multivariate density estimation supports likelihood-based modeling, anomaly detection, simulation, and decision making under uncertainty.
-Tabular datasets often combine moderate to high dimension with context-dependent conditional structure.
-Conditional variance can change with predictors, and conditional skewness or modality can depend on earlier variables.
-A transport perspective addresses these challenges by coupling the target to a simple reference through an invertible map.
-This perspective enables exact likelihoods, transparent conditionals, and efficient sampling through a shared computational backbone \citep{rosenblatt1952remarks,knothe1957contributions}.
-This thesis compares three estimator families inside a single evaluation frame with matched objectives and diagnostics.
-The frame evaluates lower-triangular transport maps, Transformation Random Forests interpreted via probability integral transforms, and copulas that decouple marginals from dependence \citep{hothorn2017transformation,hothorn2018conditional,sklar1959fonctions}.
+Multivariate density estimation is a cornerstone of modern statistics and machine learning.
+It supplies the machinery for likelihood-based inference, simulation of generative mechanisms, and uncertainty quantification in tabular domains.
+Practitioners, however, routinely face a delicate trade-off between three goals: expressive power, computational feasibility, and interpretability.
+Three complementary families of estimators have emerged around this dilemma:
+\begin{enumerate}
+  \item \textbf{Normalizing flows} build expressive change-of-variables maps and have become central in deep generative modeling \citep{rezende2015variational,papamakarios2017masked,dinh2017real,durkan2019neural,kingma2018glow,papamakarios2021normalizing}.
+  \item \textbf{Transformation models and forests} model conditional distributions through parametric transformation structures coupled with ensemble methods \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation}.
+  \item \textbf{Copulas} separate marginals from dependence, yielding interpretable multivariate association models \citep{sklar1959fonctions,nelsen2006introduction,joe2014dependence,nagler2017kdecopula}.
+\end{enumerate}
+
+This thesis situates all three families inside a common \textbf{transport framework} that makes them directly comparable \citep{rosenblatt1952remarks,knothe1957contributions,bogachev2005triangular,ramgraber2025friendly}.
+The framework operates in standardized coordinates $u$ computed solely from training-split statistics to avoid leakage and keep objectives aligned across models.
+Within this evaluation framework each estimator couples $u$ to a reference distribution---Gaussian for triangular maps and TRTF, $\mathrm{Unif}([0,1]^K)$ for copulas---and recovers densities via exact change of variables.
+Section~\ref{sec:ch1-frame} previews the core pieces and Chapter~\ref{ch:background} develops the mathematics in detail.
+
+Separable triangular transport maps play a central role throughout the thesis.
+Their lower-triangular Jacobian delivers a log-determinant in $\mathcal{O}(K)$ operations, back-substitution yields an explicit inverse for sampling, and the triangular structure produces exact conditional densities in a fixed order.
+Fully triangular maps can require $\mathcal{O}(K^2)$ work per evaluation, whereas separable variants reduce both forward and inverse passes to $\mathcal{O}(K)$, making them attractive for systematic comparison while preserving interpretable marginals and dependence mechanisms.
+
 \medskip
-
-We adopt a shared evaluation protocol summarized in Section~\ref{sec:ch1-frame} and detailed in Chapter~\ref{ch:background}; Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the pipeline at a glance.
-\medskip
-
-For clarity, efficiency, and interpretability we use separable triangular components as in Equation~\eqref{eq:transport-separable}.
-This structure stabilizes the triangular determinant, enables exact inversion by back substitution, and fixes conditional shape across contexts.
-
-% Checklist: add thesis overview, enforce dynamic references, keep ASCII hyphenation.
+\noindent\textit{Notation.} We write $\varphi$ and $\Phi$ for the density and cumulative distribution function of a univariate standard normal random variable, and $\eta$ for the $K$-variate standard normal density.
+All models are evaluated in standardized coordinates $u$ and reported on the original scale $x$ through a diagonal affine correction so that log densities remain comparable.
+Unless stated otherwise, log values are expressed in natural units (nats).
+
 \section{Thesis and Problem Statement}\label{sec:ch1-problem}
-This thesis investigates tabular multivariate density estimation within a unified transport-based evaluation frame.
-We compare separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and copula baselines.
-\medskip
-
-A transport perspective couples standardized data to a Gaussian reference through a monotone lower-triangular map.
-This structure yields exact likelihoods, transparent conditionals, exact inversion by back substitution, and linear per-sample evaluation.
-The Rosenblatt and Knothe rearrangements justify the triangular coupling for any variable order \citep{rosenblatt1952remarks,knothe1957contributions}.
-\medskip
-
-We standardize features using training statistics only.
-Equation~\eqref{eq:transport-standardise} defines the standardized coordinates $u$ used for evaluation.
-All derivatives and Jacobians are computed in $u$.
-The diagonal affine correction in Equation~\eqref{eq:transport-affine} reports log densities on the original scale $x$.
-This convention keeps objectives, diagnostics, and comparisons interoperable across estimators and datasets.
-All log quantities are reported in nats.
-\medskip
-
-We denote the $K$-variate standard normal density by $\eta$, and the univariate density and CDF by $\varphi$ and $\Phi$.
-Abbreviations for models and references appear in Table~\ref{tab:model-abbrev}.
-Copulas decouple marginals from dependence and serve as interpretable baselines.
-\medskip
-
-Separable triangular maps decompose each component into a context shift and a univariate monotone shape as in Equation~\eqref{eq:transport-separable}.
-The decomposition fixes conditional shape across contexts and stabilizes the triangular determinant.
-Under strictly increasing conditional CDFs after standard monotone smoothing and with an additive predictor, TRTF implements the same separable triangular likelihood via the probability integral transform.
-Copulas preserve explicit marginals and model dependence on the unit hypercube; in this thesis they serve strictly as low-dimensional ($K\!\le\!3$) diagnostic baselines and are not evaluated on high-$K$ datasets.
-
-We evaluate all estimators under the single protocol referenced above, with matched preprocessing and reporting to keep results comparable.
-Section~\ref{sec:evaluation-protocol} defines metrics and timing conventions.
-
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} visualizes the pipeline by showing standardization $u = T_{\text{std}}(x)$, the triangular transport branch containing TTM-Sep and TRTF, and the copula branch.
-Both branches feed the reported outputs, namely log density, conditionals, sampling, calibration, and compute, under the shared frame.
-
-The central problem is to determine when separability is appropriate for tabular data.
-We study how TRTF and copulas position themselves against direct triangular transports inside the same reporting convention (reported log densities on $x$ apply the affine correction in Eq.~\eqref{eq:transport-affine}).
-Ordering effects, conditional calibration, and computational trade-offs address this question.
-\medskip
-
-On synthetic data, TRTF tends to outperform separable TTM variants yet shares their separability limits; on the MiniBooNE benchmark it improves on Gaussian references but trails published flow baselines.
-Chapter~\ref{ch:dataanalysis} presents the evidence and discusses these comparisons.
-
-\paragraph{Non-goals.} We do not treat high-capacity normalizing flows as primary models, and we restrict nonparametric copulas to low dimensions.
-We also exclude non-separable TRTF predictors and cross-term triangular maps because they raise compute and tuning costs substantially and complicate calibration/identifiability in our setting; the present scope focuses on separable structure for transparency and exact inversion.
-Section~\ref{sec:ch1-contributions} states the scope and non-goals for reference.
-A brief pointer to expected changes if these variants were enabled—how cross-terms or non-additive predictors would likely affect calibration and NLL—is given in Chapter~\ref{ch:conclusion}, Section~\ref{sec:conclusion-outlook}.
-
-% Checklist: consolidate notation, relocate figure, keep one-claim sentences.
+
+The goal of this thesis is to examine multivariate density estimation for tabular data within a unified transport-based evaluation protocol.
+We focus on three estimators: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and nonparametric copula models.
+
+Let $u = D^{-1}(x - \mu)$ with $D = \mathrm{diag}(\sigma_1,\ldots,\sigma_K)$ determined from the training split.
+If $T: \mathbb{R}^K \to \mathbb{R}^K$ is lower triangular, $z = T(u)$, and $\eta$ denotes the chosen reference density, then the change-of-variables formula gives
+\begin{equation}
+  \log p_X(x) = \log \eta\big(T(u)\big) + \sum_{k=1}^{K} \log \frac{\partial T_k}{\partial u_k}(u_{1:k}) - \sum_{k=1}^{K} \log \sigma_k,
+  \label{eq:ch1-change-of-variables}
+\end{equation}
+so reporting on the $x$-scale differs from reporting on the $u$-scale by the constant $-\sum_k \log \sigma_k$ per data point.
+The Rosenblatt and Knothe rearrangements guarantee the existence of such triangular transports for any coordinate ordering \citep{rosenblatt1952remarks,knothe1957contributions}.
+
+The copula baselines considered here are nonparametric kernel copulas \citep{nagler2017kdecopula}.
+We fit smooth univariate marginals with \texttt{kde1d}, form pseudo-observations using mid-rank empirical CDFs during training to avoid double smoothing, and evaluate joint densities through
+\[
+  p_X(x) = c\big(F_1(x_1),\ldots,F_K(x_K)\big)\, \prod_{k=1}^{K} f_k(x_k),
+\]
+where $c$ is the copula density and $F_k = \int f_k$ the smoothed marginal CDFs. No Gaussian copula enters the study; the emphasis remains on flexible, interpretable diagnostics in low dimensions.
+
+Separable triangular maps decompose each coordinate into a context-dependent shift plus a one-dimensional monotone transform.
+TRTF occupies an intermediate position: under strictly increasing conditional CDFs and the standard forest aggregation after monotone smoothing it reproduces the same separable triangular likelihood via the probability integral transform.
+The shared protocol described above keeps preprocessing, likelihood evaluation, calibration diagnostics, and compute measurements aligned across all estimators.
+
 \section{The Transport Frame on One Page}\label{sec:ch1-frame}
-To avoid duplication, the canonical derivations and notation live in Chapter~\ref{ch:background}.
-This section serves only as a map: we standardize with train-only statistics (Eq.~\eqref{eq:transport-standardise}), evaluate likelihoods via the pullback (Eq.~\eqref{eq:transport-pullback}), exploit the triangular determinant factorization (Eq.~\eqref{eq:transport-det}), and apply the affine correction for reporting (Eq.~\eqref{eq:transport-affine}).
-Separable components are defined in Eq.~\eqref{eq:transport-separable}.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the pipeline.
-
-Notation remains consistent.
-We write $\eta$ for the $K$-variate standard normal density, and $\varphi$ and $\Phi$ for the univariate standard normal density and CDF.
-We reserve $u$ for standardized coordinates and $x$ for original coordinates, and we compute all derivatives with respect to $u$.
-These choices align symbols across Chapters~\ref{ch:intro}--\ref{ch:dataanalysis} and prevent ambiguity in later diagnostics and tables.
-
-This one-page frame removes duplicated exposition from Chapter~\ref{ch:background}.
-It establishes where logs and Jacobians live and makes complexity, inversion, and units explicit before the comparisons that follow.
-Section~\ref{sec:ch1-problem} documented the motivation, and Section~\ref{sec:ch1-contributions} states the resulting contributions and research questions.
-
-% Checklist: map contributions to evidence, keep bridges to later sections, ensure dynamic refs.
+
+The key ingredients of the transport framework are:
+\begin{itemize}
+  \item \textbf{Standardization.} Coordinates $u = D^{-1}(x-\mu)$ use training-split statistics only; all logs are in nats.
+  \item \textbf{Change of variables.} Densities follow from $z = T(u)$ through Equation~\eqref{eq:ch1-change-of-variables}.
+  \item \textbf{Triangular log determinant.} Lower-triangular Jacobians imply $\log \det J_T(u) = \sum_k \log \partial_{u_k} T_k(u_{1:k})$, giving $\mathcal{O}(K)$ cost.
+  \item \textbf{Affine correction.} Reporting on $x$ applies the constant correction $-\sum_k \log \sigma_k$ per data point.
+  \item \textbf{Computational profile.} Fully triangular maps can incur $\mathcal{O}(K^2)$ evaluation and inversion, whereas separable maps reduce both passes to $\mathcal{O}(K)$ while keeping the log determinant linear in $K$.
+\end{itemize}
+
+Separable triangular transports implement each coordinate as the sum of a context shift and a monotone one-dimensional basis, maintaining determinant stability and consistent conditional behaviour.
+TRTF achieves the same likelihood under the conditions noted above, while nonparametric copulas operate on the unit hypercube with explicit marginals.
+Standardization $u = T_{\mathrm{std}}(x)$ feeds either the triangular transport branch (TTM variants, TRTF) or the copula branch, and both return log densities, conditional diagnostics, samples, calibration metrics, and compute summaries.
+
 \section{Contributions and Research Questions}\label{sec:ch1-contributions}
-This section states the contributions and the research questions, and maps them to the chapters and figures that deliver the evidence.
-We adopt the shared transport frame summarized in Section~\ref{sec:ch1-frame}; Chapter~\ref{ch:background} records notation and assumptions, and Figure~\ref{fig:transport-schematic} anchors the comparisons.
-
-The first contribution formalizes a unified likelihood view for separable triangular transport maps, Transformation Random Forests, and copula baselines.
-Where we claim an equivalence between TRTF and separable triangular transports, it holds under the conditions made explicit in Section~\ref{sec:transport-trtf} (strictly increasing conditional CDFs after monotone smoothing and an additive predictor).
-Chapter~\ref{ch:background} and Figure~\ref{fig:transport-schematic} establish the conventions and remove duplication in later chapters.
-
-The second contribution provides empirical benchmarks under a single protocol with matched preprocessing and reporting.
-We evaluate TTM-Sep, TTM-Marg, and TRTF on synthetic generators and real tabular data; copulas are included only as low-dimensional ($K\!\le\!3$) diagnostic baselines (e.g., Half-Moon, 4D) and are not used in high-$K$ studies.
-The protocol records three families of measurements: average test log-likelihoods, conditional diagnostics based on probability integral transforms, and compute indicators for training and per-sample evaluation.
-Section~\ref{sec:evaluation-protocol} defines the protocol, Section~\ref{sec:synthetic-results} presents the synthetic and autoregressive results, and Section~\ref{sec:realdata} positions our measurements against published normalizing-flow baselines where appropriate.
-
-The third contribution distills practical guidance from the unified frame and the benchmarks.
-We state operational choices that preserve comparability, highlight ordering sensitivity and separability limits, and summarize when copulas serve as informative baselines.
-Chapter~\ref{ch:conclusion} consolidates these points as actionable recommendations and records limitations that motivate richer parameterizations or alternative predictors.
-
-Two questions drive the empirical study and bind the contributions to specific measurements.
-The first question asks how TRTF compares with TTM-Sep and copula baselines on synthetic data.
-All estimators share the transport frame in this comparison.
-We answer by reporting average test negative log-likelihoods, conditional negative log-likelihood decompositions, and probability integral transform diagnostics, with timing summaries that quantify practical cost.
-Section~\ref{sec:synthetic-results} provides the corresponding tables and figures.
-
-The second question asks how closely our TRTF results on real benchmarks approach the published performance of modern normalizing flows under the standard preprocessing.
-We answer by placing our test log-likelihoods beside reported numbers from the literature.
-The gaps are interpreted through the separable Jacobian constraint and compute profiles.
-Section~\ref{sec:realdata} reports these comparisons, and Chapter~\ref{ch:conclusion} interprets their implications for model choice.
-
-Scope and non-goals maintain focus and ensure reproducibility.
-We study separable triangular maps and TRTF with additive predictors and compute all derivatives and Jacobians in standardized coordinates.
-The map direction $S:u \to z$ remains fixed for evaluation and inversion.
-Copulas include Gaussian dependence and a low-dimensional nonparametric variant used strictly as a diagnostic baseline.
-We treat high-capacity flows as external references rather than primary models, and we do not evaluate non-additive TRTF variants or cross-term triangular maps in this chapter.
-Chapter~\ref{ch:background} records the formal assumptions and notation.
-Section~\ref{sec:datasets-preprocessing} details preprocessing, seeds, and reporting conventions that keep results comparable across datasets and estimators.
-
-Taken together, these commitments make the comparisons interpretable, keep units and complexity explicit, and prepare the reader for the empirical evidence that answers the two questions under a single, transparent evaluation frame.
-
-% Abbreviation table for reuse across chapters
+
+The thesis contributes along three axes that move from theoretical unification to empirical benchmarking and practical interpretation:
+\begin{enumerate}
+  \item \textbf{Unified likelihood framework.} We formalise a shared transport perspective that places separable triangular maps, TRTF, and nonparametric copulas inside one mathematical scheme.
+The equivalence between TRTF and separable triangular transports holds under explicit regularity assumptions---strictly monotone conditional CDFs and the forest aggregation after smoothing.
+
+    \begin{quote}
+    \textbf{Proposition (informal).} Under axis-parallel partitions, strictly monotone conditional CDFs, and the forest aggregation after smoothing, the TRTF likelihood coincides with that of a separable triangular transport in standardized coordinates.
+    \end{quote}
+
+  \item \textbf{Benchmarking under a common protocol.} We design a standardized evaluation pipeline covering synthetic generators and real tabular benchmarks.
+TTM-Sep, its marginal variant, and TRTF are evaluated systematically; nonparametric kernel copulas act as diagnostic comparators when $K \le 3$.
+The pipeline produces three categories of evidence: (i) average test log likelihoods, (ii) conditional diagnostics based on probability integral transforms, and (iii) compute measurements for training effort and per-sample evaluation time.
+Together they quantify the trade-offs among accuracy, calibration, and efficiency.
+
+  \item \textbf{Operational insights.} The theoretical and empirical findings yield guidance for practitioners.
+We analyse sensitivity to ordering, the practical limits of separability, and when nonparametric copulas remain informative as diagnostic references.
+Chapter~\ref{ch:conclusion} synthesizes these insights and outlines extensions toward richer parameterisations.
+\end{enumerate}
+
+Two research questions organise the empirical study:
+\begin{itemize}
+  \item \textbf{Synthetic data.} How do TRTF, TTM-Sep, and nonparametric copulas compare on controlled generators?
+Evidence comprises likelihood comparisons, conditional calibration metrics, and timing measurements.
+  \item \textbf{Benchmark datasets.} How close do TRTF results come to published normalizing flow baselines under identical preprocessing?
+Side-by-side likelihood comparisons, interpreted through the constraints of separability and compute profiles, answer this question.
+\end{itemize}
+
 \begin{table}[t]
@@
     \hline
-    TTM-Marg & Marginal triangular transport map (per-dimension; no context) \\
-    TTM-Sep & Separable triangular transport map (additive: $g_k$ shift + monotone $h_k$) \\
+    TTM-Marg  & Marginal triangular transport (per-dimension; no context) \\
+    TTM-Sep   & Separable triangular transport (context shift $+$ monotone 1D transform) \\
     TRTF      & Transformation Random Forests (axis-parallel splits) \\
     True-Marg & Oracle marginal density \\
-    True-Joint & Oracle conditional joint density \\
-    Copula & Copula baseline (Gaussian or nonparametric) \\
+    True-Joint& Oracle conditional joint density \\
+    Copula    & Nonparametric kernel copula baseline (smoothed marginals; no Gaussian copula) \\
     \hline
@@
 
+Addressing these questions bridges rigorous transport-based theory with empirical validation, clarifying when simple triangular models suffice and when more expressive families become necessary.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\section{Transport Frame and Notation}\label{sec:transport-frame}
-
-This section fixes the standardized coordinate system, notation, and algebraic identities used throughout the thesis.
-The motivation and schematic in Figure~\ref{fig:transport-schematic} housed in Appendix~\ref{ch:appendix} remain valid; here we strip the exposition down to the formulas needed in later chapters.
-We summarize the standardized pullback likelihood, state the triangularity assumption, and record the Jacobian factorization that drives evaluation and inversion.
-
-We work with observations on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardization map
-\begin{equation}
-  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,\label{eq:transport-standardise}
-\end{equation}
-where $\mu$ and $\sigma$ denote the empirical mean and standard deviation estimated on the training split and $\oslash$ denotes elementwise division.
-In words, we shift and rescale features once, using training data only, and keep all derivatives and Jacobians in $u$-space to avoid leakage and to ensure comparability across estimators.
-
-The standardized density $\pi_U$ is coupled to a simple reference through a monotone triangular map $S:u\mapsto z$.
-Throughout the thesis the reference is the $K$-variate standard normal density $\eta(z)$.
-The pullback identity then reads
-\begin{equation}
-  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|,\label{eq:transport-pullback}
-\end{equation}
-which evaluates the reference at $S(u)$ and applies the exact volume correction given by the Jacobian determinant.
-Reporting log densities on the original scale requires only the diagonal affine correction implied by standardization,
-\begin{equation}
-  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.\label{eq:transport-affine}
-\end{equation}
-We therefore differentiate with respect to $u$, and we convert to $x$-scale only at reporting time.
-
-The transport is assumed to be lower triangular and componentwise monotone,
-\begin{equation}
-  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), \qquad \partial_{u_k}S_k(u_{1:k})>0,\label{eq:transport-triangular}
-\end{equation}
-so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorizes into a sum of one-dimensional log derivatives,
-\begin{equation}
-  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).\label{eq:transport-det}
-\end{equation}
-The factorization yields $\mathcal{O}(K)$ evaluation cost per-sample, improves numerical stability, and guarantees global invertibility: strictly monotone diagonal derivatives let us recover $x$ by solving $K$ one-dimensional monotone equations in sequence, mirroring the Rosenblatt and Knothe rearrangements \citep{rosenblatt1952remarks,knothe1957contributions}.
-
-Table~\ref{tab:transport-notation} consolidates the notation used in this transport frame.
-All derivatives and Jacobians act on $u$; the affine correction \eqref{eq:transport-affine} converts log densities back to $x$ for reporting.
-The remainder of this chapter adopts this frame.
-Section~\ref{sec:transport-separable} details the separable triangular parameterization used for direct transports.
-Section~\ref{sec:transport-trtf} shows how Transformation Random Forests induce the same triangular likelihood via the probability integral transform under the conditions stated there (strictly increasing conditional CDFs after monotone smoothing and an additive predictor).
-Section~\ref{sec:transport-copula} places copulas in the same reporting convention.
+\section{Transport Framework and Notation}\label{sec:transport-frame}
+
+This section establishes the coordinate system, notation, and algebraic identities used throughout the thesis.
+The schematic in Appendix~\ref{ch:appendix} remains valid; here we focus only on the formulas that are needed later.
+We set up the standardized pullback likelihood, the triangularity assumption, and the Jacobian factorization.
+
+We write $u_{1:k}=(u_1,\ldots,u_k)$ and use $\partial_{u_k}$ for partial derivatives with respect to $u_k$.
+
+We observe data on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardization map
+\begin{equation}
+  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,
+  \label{eq:transport-standardise}
+\end{equation}
+where $\mu$ and $\sigma$ are the empirical mean and standard deviation estimated on the training split, and $\oslash$ denotes elementwise division.
+Each feature is shifted and rescaled once using training data only.
+All derivatives and Jacobians are then taken in $u$-space, which avoids leakage and keeps results comparable across estimators.
+
+We model $\pi_U$ as the pullback of a simple reference $\eta$ through a triangular map $S:u\mapsto z$.
+The reference is the $K$-variate standard normal $\eta(z)$.
+Applying the change-of-variables formula gives
+\begin{equation}
+  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|.
+  \label{eq:transport-pullback}
+\end{equation}
+
+To report log densities on the original scale, we apply only the diagonal affine correction implied by standardization:
+\begin{equation}
+  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.
+  \label{eq:transport-affine}
+\end{equation}
+
+The transport is assumed to be lower triangular and strictly increasing in each coordinate,
+\begin{equation}
+  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), 
+  \qquad \partial_{u_k}S_k(u_{1:k})>0,
+  \label{eq:transport-triangular}
+\end{equation}
+so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorizes as
+\begin{equation}
+  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).
+  \label{eq:transport-det}
+\end{equation}
+
+This factorization makes the \emph{log-determinant} cost $\mathcal{O}(K)$ once the diagonal terms are available.
+Evaluating all $S_k(u_{1:k})$ can be $\mathcal{O}(K^2)$ for a general triangular map, but reduces to $\mathcal{O}(K)$ in the separable parameterization of Section~\ref{sec:transport-separable}.
+
+Strict monotonicity on each coordinate ($\partial_{u_k} S_k>0$) together with triangularity implies that $S$ is a bijection with a triangular inverse, computable by back-substitution.
 
@@
   \centering
-  \caption{Notation for the transport frame used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
-All derivatives and Jacobians are taken with respect to $u$; log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.}
+  \caption{Notation for the transport framework used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
+All derivatives and Jacobians are with respect to $u$; log densities on $x$ apply the affine correction in Equation~\eqref{eq:transport-affine}.}
   \label{tab:transport-notation}
@@
     $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
-    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space, respectively \\ 
+    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space \\ 
     $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
@@
 
-\section{Separable Triangular Maps and Transformation Random Forests as Transport}\label{sec:transport-separable}
-
-% Checklist: enforce US spelling, ensure dynamic cross-references, keep equation labels aligned.
-This section unifies separable triangular maps and Transformation Random Forests (TRTF) within the transport frame fixed in Section~\ref{sec:transport-frame}.
-Both estimators realize a monotone lower-triangular map $S:u\mapsto z$ that couples the standardized target to the Gaussian reference $\eta$.
-The use of triangular transports builds on modern measure-transport literature; see, for instance, triangular transformations and their properties in \citet{bogachev2005triangular}.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} illustrates the shared backbone and locates \textsc{TTM-Sep} and \textsc{TRTF} on the transport branch introduced in Chapter~\ref{ch:intro}.
-We focus on shared likelihood identities, modeling assumptions, and limits of separability, and defer implementation details to Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.
-
-The goal is to state a single likelihood for both constructions, clarify what separability permits, and identify failure modes that motivate richer parameterizations.
-We do not pursue non-additive TRTF predictors, cross-term triangular maps, or ordering heuristics in this section; Chapter~\ref{ch:dataanalysis} evaluates those choices empirically and Appendix~\ref{ch:appendix} documents routines and defaults.
-
-We adopt the notation introduced in Section~\ref{sec:transport-frame}.
-Coordinates satisfy $u=T_{\mathrm{std}}(x)$, the reference density is $\eta(z)$, and the pullback identity \eqref{eq:transport-pullback} gives $\pi_U(u)=\eta(S(u))\,|\det\nabla_u S(u)|$.
-The map is lower-triangular with strictly positive diagonal partial derivatives, which yields the sum decomposition in Equation~\eqref{eq:transport-det}.
-These conventions keep derivatives in $u$-space and apply the affine correction \eqref{eq:transport-affine} only when reporting $\log \pi_X(x)$.
-
-We restrict attention to separable triangular maps. Component $k$ decomposes into a context shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) \;=\; g_k(u_{1:k-1}) + h_k(u_k),\qquad \log \partial_{u_k}S_k(u_{1:k}) \;=\; \log h_k'(u_k),\label{eq:transport-separable}
-\end{equation}
-which fixes context effects in $g_k$ and reserves $h_k$ for the one-dimensional marginal shape.
-Intuitively, earlier coordinates translate the location, while the conditional shape along $u_k$ remains fixed across contexts.
-The Jacobian contribution depends only on $u_k$, which reduces per-sample evaluation cost and simplifies inversion.
-
-\begin{shaded}
-\textbf{Assumptions.} Unless stated otherwise, we assume:
+Taken together, the triangular and strictly monotone structure turns joint density evaluation
+and sampling into sequential one-dimensional stages, and the standardized coordinates allow us
+to report on the $x$-scale with only the diagonal correction from Equation~\eqref{eq:transport-affine}.
+
+\section{Separable Triangular Maps and Transformation Random Forests}\label{sec:transport-separable}
+
+We now move from the general triangular setup to the first concrete parameterization: the \textbf{separable triangular map}.
+We then show how \textbf{Transformation Random Forests (TRTF)} fit into the same framework.
+The aim is a shared likelihood that we can train, compare, and diagnose across estimators without changing conventions.
+
+\subsection{Separable triangular component}
+
+A separable component has the form
+\begin{equation}
+  S_k(u_{1:k}) \;=\; g_k(u_{1:k-1}) + h_k(u_k), \qquad h_k'(u_k) > 0,
+  \label{eq:transport-separable}
+\end{equation}
+where earlier coordinates shift location through $g_k$, while $h_k$ shapes the marginal and ensures monotonicity.
+To fix identifiability, we enforce a convention such as centering $g_k$ on the training split, $\mathbb E[g_k(U_{1:k-1})]=0$.
+
+With the standard normal reference, the pullback identity becomes
+\begin{equation}
+  \log \pi_U(u) \;=\; \sum_{k=1}^{K}\Big[\log \varphi\!\big(S_k(u_{1:k})\big) + \log h_k'(u_k)\Big].
+  \label{eq:transport-likelihood}
+\end{equation}
+Up to a constant, the per-sample negative log-likelihood is
+\begin{equation}
+  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big].
+  \label{eq:transport-loss}
+\end{equation}
+
+For a dataset $\{u^{(i)}\}_{i=1}^n$, the empirical objective reads
+\[
+\mathcal L_n = \frac{1}{n}\sum_{i=1}^n \sum_{k=1}^{K}
+\Big[\tfrac{1}{2}\,S_k(u^{(i)}_{1:k})^2 - \log h_k'(u^{(i)}_k)\Big] + \text{const}.
+\]
+
+These formulas evaluate in $\mathcal{O}(K)$ time per sample, since the Jacobian term depends only on $u_k$.
+Sampling inverts $S$ by back-substitution, and log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.
+The speed comes with a trade-off: separability fixes the conditional shape along $u_k$ once $g_k$ shifts the location, so variance, skewness, and modality no longer adapt to the context $u_{1:k-1}$.
+
+\subsection{Transformation Random Forests in the framework}\label{sec:transport-trtf}
+
+Transformation Random Forests (TRTF) fit conditional CDFs and insert them into the triangular pipeline via the probability integral transform.
+Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the smoothed conditional CDF.
+Define
+\begin{equation}
+  S_k(u_{1:k}) \;=\; \Phi^{-1}\!\Big(\widehat F_k(u_k \mid u_{1:k-1})\Big).
+  \label{eq:transport-trtf-map}
+\end{equation}
+
+If the forest is parameterized additively as
+\begin{equation}
+  \widehat F_k(u_k \mid u_{1:k-1}) \;=\; \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),
+  \label{eq:transport-trtf-additive}
+\end{equation}
+then we recover the separable structure,
+\begin{equation}
+  S_k(u_{1:k}) \;=\; h_k(u_k) + g_k(u_{1:k-1}), \qquad h_k'(u_k) > 0.
+  \label{eq:transport-trtf-separable}
+\end{equation}
+
+\paragraph{Derivative check.}
+\[
+\partial_{u_k}S_k
+= \frac{d}{du_k}\Phi^{-1}(\Phi(h_k(u_k)+g_k))
+= h_k'(u_k) > 0,
+\]
+so the Jacobian diagonal is indeed $h_k'(u_k)$ in the additive TRTF case.
+
+\paragraph{Numerical note.}
+In practice, forests may produce $\widehat F_k$ values outside $(0,1)$ by tiny amounts or with imperfect monotonicity.
+We therefore clip to $(\varepsilon,1-\varepsilon)$ with $\varepsilon=10^{-6}$ and, if needed, apply isotonic post-processing to enforce strict increase.
+This keeps $S_k=\Phi^{-1}(\widehat F_k)$ well defined.
+
+With this forest aggregation and strictly monotone conditionals, the resulting TRTF map coincides with the separable parameterization, so it delivers the same likelihood and back-substitution sampler.
+
+\section{Copula Baselines}\label{sec:transport-copula}
+
+We now turn to copulas, used here as diagnostic baselines rather than high-dimensional competitors.
+The focus is on a nonparametric bivariate copula when $K=2$, with an independence fallback otherwise.
+Standardization remains global (based on the full training split), whereas marginal fits for copulas are per-class.
+This ensures comparability on the $x$-scale when reporting.
+
+\subsection{Estimation and reporting}
+
+For each class $y$, we proceed as follows:
+
 \begin{itemize}
-  \item \emph{Lower-triangularity:} $S$ has the structure in Eq.~\eqref{eq:transport-triangular}.
-  \item \emph{Strict monotone coordinates:} $\partial_{u_k} S_k(u_{1:k}) > 0$ for all $k$ and all arguments.
-  \item \emph{Separable component:} Eq.~\eqref{eq:transport-separable} holds, so conditional shape along $u_k$ is fixed across contexts.
+  \item \textbf{Marginals:} Fit univariate kernel densities (\texttt{kde1d}) to obtain $\widehat f_k(\cdot \mid y)$ and CDFs $\widehat F_k(\cdot \mid y)$.
+  \item \textbf{Pseudo-observations:} Transform samples to $(0,1)$ using mid-ranks,
+  \begin{equation}
+    U_{ik}^{(y)} \;=\; \frac{\operatorname{rank}(X_{ik}^{(y)})}{n_y + 1}.
+    \label{eq:copula-midranks}
+  \end{equation}
+  Mid-ranks handle ties; if the data contain many, a tiny jitter is added before ranking.
+  \item \textbf{Dependence:} Fit a bivariate kernel copula density $c_y(u_1,u_2)$ using \texttt{kdecopula::kdecop} with local quadratic smoothing, which uses a transformation approach to avoid boundary bias \citep{nagler2017kdecopula}.
 \end{itemize}
-\end{shaded}
-
-Substituting the standard normal reference into \eqref{eq:transport-pullback} produces a separable objective,
-\begin{equation}
-  \log \pi_U(u) \;=\; \sum_{k=1}^{K}\Big[\log \varphi\!\big(S_k(u_{1:k})\big) + \log h_k'(u_k)\Big],\label{eq:transport-likelihood}
-\end{equation}
-where $\varphi$ denotes the univariate standard normal density.
-Equation~\eqref{eq:transport-likelihood} splits the log density into a reference fit and an exact volume correction.
-In plain language, the model evaluates how Gaussian each transformed coordinate appears, then corrects for the local stretch induced by $h_k$.
-The same decomposition produces linear per-sample time in $K$ and stable accumulation of log derivatives.
-
-The negative log-likelihood per-sample takes the quadratic-plus-barrier form
-\begin{equation}
-  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big],\label{eq:transport-loss}
-\end{equation}
-which follows because $\log \varphi(t) = -\tfrac{1}{2}t^2 - \tfrac{1}{2}\log(2\pi)$ and constants independent of the parameters drop out.
-Equation~\eqref{eq:transport-loss} pulls each component toward the reference while preventing degenerate derivatives through the log barrier.
-In practice we enforce $h_k'(u_k)>0$ by construction and control tails with mild regularization; implementation choices appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}.
-
-Separable structure encodes clear modeling assumptions.
-Conditional variance, skewness, and modality do not change with the preceding coordinates once $g_k$ shifts location.
-Consequently, separable maps can underfit heteroskedastic or multimodal conditionals, which manifests as U-shaped or inverted-U probability integral transform (PIT) diagnostics.
-Variable ordering also matters for finite bases because triangular transports are anisotropic, even though a Knothe--Rosenblatt rearrangement exists for any ordering \citep{rosenblatt1952remarks,knothe1957contributions}.
-These caveats guide the robustness checks in Chapter~\ref{ch:dataanalysis}.
-
-\subsection{Transformation Random Forests within the Transport Frame}\label{sec:transport-trtf}
-
-Transformation Random Forests \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation} fit into the same transport frame through the probability integral transform.
-Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by a TRTF for coordinate $k$.
-(In practice, forest CDFs can be stepwise; we assume a measurable, strictly increasing version after standard monotone smoothing so that inversion and derivatives are well-defined.) The induced triangular component is
-\begin{equation}
-  S_k(u_{1:k}) \;=\; \Phi^{-1}\!\Big(\widehat F_k(u_k \mid u_{1:k-1})\Big),\label{eq:transport-trtf-map}
-\end{equation}
-which maps conditionals to standard normal margins.
-In plain language, TRTF predicts a conditional CDF, then the probit transform places the result on the Gaussian reference scale.
-Differentiating $\Phi\!\big(S_k(u_{1:k})\big)=\widehat F_k(u_k \mid u_{1:k-1})$ with respect to $u_k$ yields
-\begin{equation}
-  \widehat \pi_k(u_k \mid u_{1:k-1}) \;=\; \varphi\!\big(S_k(u_{1:k})\big)\,\partial_{u_k}S_k(u_{1:k}),\label{eq:transport-trtf-likelihood}
-\end{equation}
-which is exactly the pullback factor in Equation~\eqref{eq:transport-likelihood}. Summing over $k$ recovers Equation~\eqref{eq:transport-likelihood} in standardized coordinates.
-
-The additive-predictor TRTF used in this thesis yields a separable transport. Under the model
-\begin{equation}
-  \widehat F_k(u_k \mid u_{1:k-1}) \;=\; \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),\label{eq:transport-trtf-additive}
-\end{equation}
-we obtain
-\begin{equation}
-  S_k(u_{1:k}) \;=\; h_k(u_k) + g_k(u_{1:k-1}),\qquad \partial_{u_k}S_k(u_{1:k}) \;=\; h_k'(u_k),\label{eq:transport-trtf-separable}
-\end{equation}
-so TRTF implements the same separable triangular likelihood as the direct parameterization in Equation~\eqref{eq:transport-separable}.
-The map is monotone in $u_k$ by construction, the Jacobian depends only on $u_k$, and inversion proceeds by back-substitution identical to the separable map.
-This equivalence underpins the empirical comparisons in Chapter~\ref{ch:dataanalysis}.
-
-The equivalence also clarifies limits.
-Additive TRTF predictors shift location but cannot alter conditional shape with context, which mirrors the separable constraint.
-Axis-aligned partitions stabilize estimation, yet they do not remove residual multimodality when the conditional shape varies with $u_{1:k-1}$.
-These limits are visible in PIT diagnostics and conditional negative log-likelihood decompositions on synthetic studies.
-
-We emphasize operational scope and supporting references.
-All derivatives and Jacobians are computed in standardized coordinates, evaluation uses the triangular pullback, and reported log densities on the original scale include the affine correction \eqref{eq:transport-affine}.
-Implementation details on basis choices for $h_k$, feature construction for $g_k$, regularization, derivative clipping, timing, and memory footprints appear in Chapter~\ref{ch:dataanalysis} and Appendix~\ref{ch:appendix}, which also provides pseudo-code for both estimators.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} visualizes how the \textsc{TTM-Sep} and \textsc{TRTF} branches share the same computational path from standardized data to reported likelihoods.
-
-In summary, separable triangular maps and additive-predictor TRTF realize the same lower-triangular likelihood once the data are standardized and the conditional CDFs are strictly increasing after monotone smoothing.
-The shared structure yields exact likelihoods, exact inversion, transparent conditionals, and linear per-sample complexity, but it restricts context-dependent shape.
-Section~\ref{sec:transport-copula} positions copulas within the same reporting convention to decouple marginals from dependence.
-
-\section{Copula Baselines}\label{sec:transport-copula}
-
-% Checklist: align references with transport frame, cite Sklar, enforce US spelling.
-This section positions copulas within the unified transport frame and links their reported likelihoods to the evaluation conventions used for triangular maps and Transformation Random Forests.
-Copulas decouple marginal modeling from dependence modeling by pairing univariate marginals with a separate dependence density on the unit hypercube \citep{nelsen2006introduction,joe2014dependence}.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} displays the copula branch beside the triangular branch and highlights how both yield comparable reported log densities under the shared evaluation pipeline.
-
-We begin with pseudo-observations built from training-split marginals.
-Let $\widehat F_k$ denote the strictly increasing empirical or smoothed CDF of $X_k$ estimated on the training split.
-Define the pseudo-observations and their probit transform as
-\begin{equation}
-  v_k \;=\; \widehat F_k(x_k),\qquad z_k \;=\; \Phi^{-1}(v_k),\label{eq:copula-probit}
-\end{equation}
-which map each coordinate to $(0,1)$ and then to $\mathbb{R}$ through the probit function.
-In plain language, the marginals become uniform scores, and $z$ records those scores on a Gaussian scale.
-Mid-ranks and clamping near $(0,1)$ stabilize the transformation in finite samples.
-
-The copula representation combines marginal densities with a dependence factor. The joint log density on the original scale satisfies
-\begin{equation}
-  \log \widehat \pi_X(x) \;=\; \sum_{k=1}^{K} \log \widehat f_k(x_k) + \log c\!\left(v_1,\ldots,v_K\right),\label{eq:copula-logdensity}
-\end{equation}
-where $c$ denotes the copula density on $(0,1)^K$.
-Equation~\eqref{eq:copula-logdensity} separates the task into two parts: fit interpretable marginals and correct for dependence through $\log c$.
-The reported quantity already lives on the original scale, so the affine correction in Equation~\eqref{eq:transport-affine} is unnecessary.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} uses the equivalent shorthand $\log c(z)$ because dependence is evaluated through $z=\Phi^{-1}(v)$.
-
-The independence baseline fixes a lower bound for dependence modeling. Setting $c \equiv 1$ yields
-\begin{equation}
-  \log \widehat \pi_X^{\mathrm{ind}}(x) \;=\; \sum_{k=1}^{K} \log \widehat f_k(x_k),\label{eq:copula-independence}
-\end{equation}
-which treats coordinates as independent after marginal fitting. Chapter~\ref{ch:dataanalysis} uses this baseline as a reference point in evaluation tables and figures.
-
-The Gaussian copula specifies elliptical dependence through a correlation matrix $\Sigma$. With $z=\Phi^{-1}(v)$, the copula density admits the closed form
-\begin{equation}
-  c_{\Sigma}(v) \;=\; |\Sigma|^{-1/2}\,\exp\!\Big(-\tfrac{1}{2}\,z^{\top}(\Sigma^{-1} - I)z\Big),\label{eq:copula-gaussian}
-\end{equation}
-which reduces dependence estimation to fitting $\Sigma$ on the transformed scores.
-In plain language, the Gaussian copula bends the joint shape away from independence according to $\Sigma$ while preserving the learned marginals.
-
-A low-dimensional nonparametric variant avoids elliptical assumptions at small $K$. We fit a kernel density $\widehat f_Z$ on $z=\Phi^{-1}(v)$ and recover the copula density by
-\begin{equation}
-  c(v) \;=\; \frac{\widehat f_Z\!\big(\Phi^{-1}(v)\big)}{\prod_{k=1}^{K} \varphi\!\big(\Phi^{-1}(v_k)\big)},\label{eq:copula-kde}
-\end{equation}
-which applies the change of variables from $z$ back to $v$ and yields a proper copula density.
-In words, the kernel density models the joint shape of the probit scores, and division by the product of standard normal densities restores the unit-cube scale.
-This approach is viable only for small $K$, where kernel density estimation remains accurate and stable.
-We implement this baseline via the \texttt{kdecopula} package \citep{nagler2017kdecopula}.
-Chapter~\ref{ch:dataanalysis} employs it strictly as a diagnostic baseline.
-
-The transport frame keeps reporting interoperable across modeling branches despite distinct parameterizations.
-Triangular maps and TRTF evaluate the pullback likelihood in standardized coordinates and apply the fixed affine correction \eqref{eq:transport-affine} when mapping back to $x$.
-Copulas operate on $x$ directly through Equation~\eqref{eq:copula-logdensity}, yet Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows how the probit scores $z$ maintain comparability with the Gaussian reference used above.
-This alignment keeps objectives and diagnostics consistent across Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
-
-Modeling choices and limits follow from the chosen copula.
-The Gaussian copula imposes elliptical dependence and may misrepresent tail behavior or localized asymmetry.
-The nonparametric variant mitigates these issues only at small dimension and sufficient sample size.
-The independence baseline provides a transparent reference when dependence is weak or data are scarce.
-These caveats motivate treating copulas as interpretable baselines rather than definitive high-dimensional models in the empirical study of Chapter~\ref{ch:dataanalysis}.
-Sklar’s theorem underlies all constructions above and formalizes the decoupling of marginals from dependence \citep{sklar1959fonctions}.
+
+For prediction at $x=(x_1,x_2)$:
+\begin{enumerate}
+  \item Evaluate $\widehat f_k(x_k \mid y)$ and $\widehat F_k(x_k \mid y)$.
+  \item Evaluate $c_y(\widehat F_1(x_1\mid y), \widehat F_2(x_2\mid y))$.
+  \item Combine them through
+  \begin{equation}
+    \log \widehat \pi_X(x \mid Y=y) 
+      \;=\; \log \widehat f_1(x_1 \mid y) + \log \widehat f_2(x_2 \mid y)
+      + \log c_y\!\Big(\widehat F_1(x_1 \mid y), \widehat F_2(x_2 \mid y)\Big).
+    \label{eq:copula-logdensity}
+  \end{equation}
+\end{enumerate}
+When labels are unknown, we mix across classes using priors $\pi(y)$ and a stable log-sum-exp.
+
+\subsection{Fallback and scope}
+
+If $K\neq 2$, labels are absent, or packages unavailable, we fall back to an independence model:
+\begin{equation}
+  \log \widehat \pi_X^{\mathrm{ind}}(x) \;=\; \sum_{k=1}^{K} \log \widehat f_k(x_k).
+  \label{eq:copula-independence}
+\end{equation}
+Hence the NP-Copula is restricted to $K=2$ and is used only as a diagnostic baseline.
+
+\subsection{Consistency with evaluation frame}
+
+Copulas operate directly on the original scale $x$ and therefore bypass the affine correction of Section~\ref{sec:transport-frame}.
+Still, the logic is parallel to the triangular branch: start from a simple reference (independent uniforms), add a dependence correction, and return a single joint log density.
+Using mid-ranks for fitting and $\widehat F_k$ for reporting is consistent with Sklar's theorem \citep{sklar1959fonctions,nelsen2006introduction,joe2014dependence}.
+
+\subsection{Terminology and references}
+
+We use \textit{NP-Copula} for the nonparametric kernel estimator and \textit{Independence} for the fallback KDE product.
+References include Sklar’s theorem and standard copula texts \citep{sklar1959fonctions,nelsen2006introduction,joe2014dependence}, kernel copula methodology \citep{nagler2017kdecopula}, and the transformation-model literature for context \citep{hothorn2021transformation}.
+In practice, the NP-copula baseline for $K=2$ combines per-class marginals with a dependence correction, providing a transparent diagnostic comparator to the triangular transport models.
 
@@
 
-% Checklist: enforce US spelling, clarify Jacobian phrasing, remove empty sections.
-This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling program.
-Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable.
-Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space.
-That alignment keeps objectives, diagnostics, and reported log-densities interoperable.
-
-% Checklist: define model abbreviations for concise tables
-\textbf{Model abbreviations.} We follow Table~\ref{tab:model-abbrev} for concise labels (TTM-Marg, TTM-Sep, TRTF, True-Marg/True-Joint, Copula) across tables and figures.
+% Checklist: align terminology, update abbreviations, confirm notation references.
+\begin{quote}
+\textbf{Motivation.} We factor the joint density into simple conditional pieces using a lower-triangular parameterization.
+This yields exact likelihoods, transparent conditionals we can inspect, and $\mathcal{O}(K)$ per-sample cost.
+The shared factorization lets us run ordering experiments without changing the model class.
+\end{quote}
+
+\textbf{Model abbreviations.} We use the short labels TTM-Sep (separable triangular transport map), TTM-X (separable map with low-rank cross-terms), TRTF (Transformation Random Forests; spelled out on first mention), RealNVP, MAF, NSF, True-Marg/True-Joint, and Copula consistently across tables and figures.
+
+\textbf{Units and signage.} Unless stated otherwise, all log-likelihood (LL) and negative log-likelihood (NLL) values are reported in nats.
+Plots and tables mark $\text{LL}~(\uparrow~\text{better})$ and $\text{NLL}~(\downarrow~\text{better})$ explicitly.
+
+\textbf{Notation recap.}
+\begin{itemize}
+  \item Elementwise standardization with train-split statistics:
+  \[ 
+    u_k = \frac{x_k - \mu_k}{\sigma_k},\quad k = 1,\ldots,K
+    \qquad\text{equivalently } u = \operatorname{diag}(\sigma)^{-1}(x - \mu).
+  \]
+  We compute derivatives and Jacobians in $u$-coordinates and transform back to $x$ only when reporting densities on the original scale via Equation~\eqref{eq:transport-affine}.
+  \item Prefix subvector shorthand: $u_{1:k} := (u_1,\ldots,u_k)$.
+  \item Derivatives use $h'_k(\cdot)$ for univariate derivatives; mixed partials follow the convention from Chapter~\ref{ch:background}.
+\end{itemize}
+
+\textbf{Copula conventions.}
+\begin{itemize}
+  \item All copula fits estimate marginals; reported values are joint densities on the original $x$-scale.
+  \item Half-Moon ($K=2$) uses class-conditional copulas for blue/red labels mixed by empirical class priors when we report the unlabeled density.
+No other model uses labels in this chapter; detailed results appear in Section~\ref{sec:synthetic-results}.
+\end{itemize}
 
@@
 
-% Checklist: preserve dynamic refs, ensure ASCII hyphenation, align with future sections.
-This section fixes data sources, generators, and preprocessing so likelihoods, calibration, and compute remain comparable across models.
-All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the common affine correction.
-We keep a single triangular-map direction $S:u \rightarrow z$ across methods to avoid mixed objectives.
-To avoid symbol collisions, $\sigma$ denotes standardization scales only; logistic gates are written $\operatorname{logistic}(\cdot)$ throughout.
-
-We standardize features with training-split statistics only.
-Equation~\eqref{eq:transport-standardise} defines $u = T_{\mathrm{std}}(x) = (x-\mu)\oslash \sigma$ with $\sigma_k > 0$.
-We evaluate $\log \pi_U(u)$ through the pullback identity in Equation~\eqref{eq:transport-pullback}, apply the triangular factorization from Equation~\eqref{eq:transport-det}, then convert to $\log \pi_X(x)$ using the diagonal correction in Equation~\eqref{eq:transport-affine}.
-We report average test negative log-likelihoods (NLL) in nats.
-Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the standardized pipeline shared by transport maps, Transformation Random Forests, and copulas.
-
-We use fixed train, validation, and test splits with proportions $0.60/0.20/0.20$ unless a benchmark provides official splits.
-Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures; for real-data benchmarks we use $N$ to denote the training budget.
-The canonical four-dimensional ordering is $(1,2,3,4)$.
-Robustness to ordering is assessed by averaging over all $4!
-= 24$ permutations.
-We adopt the natural column order for real datasets.
-We fix seeds $\{11, 13, 17, 19, 23\}$ for data generation and model fitting, and we average repeated runs with standard errors to quantify stochastic variability.
-Figure~\ref{fig:autoregressive-joint-calibration} displays the $20\%$ test split for the synthetic calibration study.
-
-The Half-Moon dataset provides a curved, bimodal joint in $K=2$.
-We draw a class $Y \sim \mathrm{Bernoulli}(0.5)$, an angle $\Theta \sim \mathrm{Unif}[0,\pi]$, and additive noise $\varepsilon \sim \mathcal{N}(0, \sigma^2 I_2)$ with $\sigma = 0.10$.
-For $Y = 0$ we set $m(\Theta) = (\cos \Theta, \sin \Theta)$.
-For $Y = 1$ we set $m(\Theta) = (1 - \cos \Theta, -\sin \Theta + 0.5)$.
-The observed $X = m(\Theta) + \varepsilon$.
-The ``True joint'' oracle evaluates the mixture density by numerical quadrature over $\Theta$ with the known Gaussian noise, and the ``True conditional'' oracle conditions on $Y$.
-Figure~\ref{fig:halfmoon-panels} (p.~\pageref{fig:halfmoon-panels}) shows representative contour plots at $n=250$, which align with this generator.
-Table~\ref{tab:halfmoon-nll} (p.~\pageref{tab:halfmoon-nll}) reports the corresponding NLLs.
-
-The four-dimensional autoregressive generator combines Gaussian, exponential, beta, and gamma components to induce heteroskedasticity, skew, and conditional multimodality.
-The first coordinate is $X_1 \sim \mathcal{N}(0,1)$.
-The second coordinate is independent $X_2 \sim \mathrm{Exp}(\lambda_0)$ with rate $\lambda_0 = 1$.
-The third coordinate lies on $(0,1)$ and is a context-gated mixture of two beta laws, $X_3 \mid X_{1:2} \sim w\,\mathrm{Beta}(\alpha_1, \beta_1) + (1 - w)\,\mathrm{Beta}(\alpha_2, \beta_2)$.
-We set $(\alpha_1, \beta_1) = (2.5, 5.0)$ and $(\alpha_2, \beta_2) = (5.0, 2.5)$.
-The mixing weight is $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\operatorname{logistic}(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$.
-The fourth coordinate is positive and conditionally heteroskedastic, $X_4 \mid X_{1:3} \sim \tilde{w}\,\mathrm{Gamma}(k_1, r_1(X_2)) + (1 - \tilde{w})\,\mathrm{Gamma}(k_2, r_2(X_2))$.
-We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$.
-The ``True joint'' baseline uses these closed-form conditionals to evaluate the exact joint density, while the ``True marginal'' baseline uses the corresponding univariate marginals and ignores dependence.
-
-% Checklist: summarize generator config, define softmax usage.
-\begin{table}[t]
-  \centering
-  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study.
-The beta and gamma coordinates are two-component mixtures with logistic gates; fixed parameters and gates match the prose above.}
-  \label{tab:autoregressive-config}
-  \begin{tabular}{lll}
-    \hline
-    Coordinate & Distribution & Parameters / gate \\
-    \hline
-    $X_1$ & Normal & $\mathcal{N}(0,1)$ \\
-    $X_2$ & Exponential & $\mathrm{rate} = \lambda_0 = 1$ \\
-    $X_3 \mid X_{1:2}$ & Mixture Beta & $\mathrm{Beta}(2.5,5.0)$ / $\mathrm{Beta}(5.0,2.5)$; $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2-1))$, $\gamma=(0,1.5,1.0)$ \\
-    $X_4 \mid X_{1:3}$ & Mixture Gamma & $\mathrm{Gamma}(k_1{=}3, r_1{=}1+0.5X_2)$ / $\mathrm{Gamma}(k_2{=}6, r_2{=}0.75+0.25X_2)$; $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3-0.5))$, $\delta=(0,1.0,3.0)$ \\
-    \hline
-  \end{tabular}
-\end{table}
-
-Mixture weights use the logistic gate $\operatorname{logistic}(a)=\exp(a)/(1+\exp(a))$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one.
-For completeness, the general softmax takes a vector $a$ and returns $\mathrm{softmax}(a)_i = \exp(a_i) / \sum_j \exp(a_j)$.
-This normalization is essential for the beta and gamma mixtures because it translates linear predictors into valid probability weights while preserving differentiability.
-
-Table~\ref{tab:autoregressive-config} (p.~\pageref{tab:autoregressive-config}) summarizes the mixture families, gates, and fixed parameters by dimension.
-Tables~\ref{tab:autoregressive-perm} (p.~\pageref{tab:autoregressive-perm}) and~\ref{tab:autoregressive-perm-avg} (p.~\pageref{tab:autoregressive-perm-avg}) then summarize the permutation and sample-size studies used later in this chapter.
-
-The MiniBooNE benchmark follows the published preprocessing to ensure comparability with flow-based baselines.
-We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes.
-We use the fixed train, validation, and test splits from the benchmark, apply train-only standardization, and avoid any extra pruning of correlated features.
-We report all log-likelihoods in nats and retain the published naming for flow comparators in later tables.
-Section~\ref{sec:realdata} records these steps and provides the dataset context.
-Table~\ref{tab:uci-loglik} reproduces the flow baselines that motivate our TRTF runs.
-
-Additional UCI datasets appear only when we retain them for real-data context.
-POWER keeps household electricity attributes after jittering the minute-of-day encoding, dropping the calendar date and reactive-power column, and adding uniform noise to break ties.
-GAS keeps the \texttt{ethylene\_CO} subset, treats the series as i.i.d., removes strongly correlated attributes, and retains an eight-dimensional representation.
-HEPMASS keeps only the positive class from the ``1000'' split and discards five variables with repeated values to avoid density spikes.
-These preprocessing steps follow the same train-only standardization and reporting conventions described above.
-Section~\ref{sec:realdata} provides the corresponding background and positions these datasets within our evaluation.
-
-All models use the same standardized frame and direction for evaluation, which keeps objectives, diagnostics, and reported quantities interoperable across triangular transports, TRTF, and copula baselines.
-This alignment is necessary for the conditional decompositions, probability integral transform (PIT) calibration checks, and compute summaries presented later in this chapter.
+This section fixes data sources, synthetic generators, and preprocessing so that objectives, diagnostics, and reported log-densities are comparable across models.
+All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the affine correction from Equation~\eqref{eq:transport-affine}.
+Throughout, logistic gates are written $\operatorname{logistic}(\cdot)$.
+
+\subsection{Standardization and Splits}
+Features are standardized with training-split statistics only $(\mu, \sigma \in \mathbb{R}^K, \sigma_k > 0)$ as in Equation~\eqref{eq:transport-standardise}.
+The triangular pullback, determinant factorization, and affine correction from Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-affine} stay in $u$-coordinates; we transform back to $x$ once per evaluation when reporting densities.
+We keep fixed train/test splits.
+Synthetic calibration uses an $80/20$ train/test partition.
+Real-data splits follow the published convention for each benchmark, and compute profiles are summarized in Appendix~\ref{app:evaluation}.
+
+For synthetic runs we hold seeds $\{11, 13, 17, 19, 23\}$ fixed across models and report means with $\pm 2$ standard errors when averaging replicates.
+Ordering experiments reuse the lower-triangular factorization and vary only the coordinate order; the permutation sweeps in the four-dimensional study currently pin $\texttt{SEED}=42$ unless overridden so that every ordering shares the same draws.
+Unless noted otherwise, our canonical ordering remains $(1,2,3,4)$ for the four-dimensional generator; Section~\ref{sec:synthetic-results} reports the permutation studies.
+
+\subsection{Half-Moon ($K=2$)}
+Half-Moon is a curved, bimodal joint designed to surface dependence beyond linear correlation.
+We evaluate joint log-density contours per model and compare results to the true joint.
+Figure~\ref{fig:halfmoon-panels} and Table~\ref{tab:halfmoon-nll} later in this chapter instantiate the diagnostics.
+For the copula baseline we fit blue/red class-conditional copulas and mix them by empirical class priors when reporting the unlabeled density.
+No other estimator uses labels.
+
+\subsection{Four-dimensional Autoregressive Generator ($K=4$)}
+The synthetic autoregressive generator matches the lower-triangular factorization while stressing skewness, heavy tails, and context-dependent shape:
+\begin{itemize}
+  \item $X_1 \sim \mathcal{N}(0,1)$.
+  \item $X_2 \sim \mathrm{Exponential}(\lambda_0)$ with $\lambda_0 = 1$.
+  \item $X_3 \mid X_{1:2}$ is a mixture of two beta laws on $(0,1)$, e.g. $\mathrm{Beta}(2.5, 5.0)$ versus $\mathrm{Beta}(5.0, 2.5)$, with gate
+  \[ 
+    w_3(X_{1:2}) = \operatorname{logistic}\big(\gamma_0 + \gamma_1 X_1 + \gamma_2 (X_2 - 1)\big), \qquad \gamma = (\gamma_0, \gamma_1, \gamma_2).
+  \]
+  \item $X_4 \mid X_{1:3}$ is a mixture of two gamma laws on $(0, \infty)$ with distinct shape/scale pairs, gated by
+  \[ 
+    w_4(X_{1:3}) = \operatorname{logistic}\big(\delta_0 + \delta_1 X_1 + \delta_3 (X_3 - 0.5)\big), \qquad \delta = (\delta_0, \delta_1, \delta_3).
+  \]
+\end{itemize}
+Mixture weights are normalized through the usual softmax over component logits so that Jacobian terms stay well behaved under shared standardization.
+This design keeps $X_4$ conditionally heteroskedastic with positive support and aligns with the triangular evaluation cost $\mathcal{O}(K)$.
+
+\paragraph{Why these generators?}
+They match the triangular modeling assumptions yet stress estimators with skewness, heavy tails, and context-dependent variation.
+The true joint density and conditionals are known, which enables direct comparisons without confounding from Monte Carlo error.
+
+\paragraph{Reproducibility note.}
+All random seeds, splits, and generator parameters are fixed in the code release.
+Negative NLL values appear for bounded or highly concentrated marginals (e.g., beta or gamma mass near zero); this behavior is expected and does not signal an error.
+
+\bigskip
+\noindent\emph{Next: Section~\ref{sec:models-implementation} details the estimators and training settings that make these likelihoods and diagnostics comparable across methods.}
 
@@
 % Checklist: align references, enforce US spelling, ensure active voice.
-This section specifies the estimators and implementation details that keep likelihoods, calibration, and compute directly comparable across models.
-All estimators share the transport direction $S:u \rightarrow z$, operate in standardized coordinates, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.
-
-We implement separable lower-triangular transport maps denoted TTM-Sep. Component $k$ decomposes into a context shift and a univariate monotone shape,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k), \qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k) > 0,
+\begin{quote}
+\textbf{Scope.} This section specifies the estimators and the implementation details that keep our likelihoods, diagnostics, and plots comparable.
+We reuse the preprocessing and transport notation in Table~\ref{tab:transport-notation}, draw background from Chapter~\ref{ch:background}, follow the data handling in Section~\ref{sec:datasets-preprocessing}, and lean on the auxiliary pseudo-code in Appendix~\ref{ch:appendix}.
+\end{quote}
+
+\subsection{Separable Triangular Transport Maps (TTM-Sep)}
+
+We use lower-triangular, separable transport maps
+\begin{equation}
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k), \qquad \partial_{u_k} S_k(u_{1:k}) = h'_k(u_k) > 0,
   \label{eq:ttm-separable-def}
 \end{equation}
-so the Jacobian contribution depends only on $u_k$. The structure yields linear per-sample complexity in $K$ and exact inversion by back-substitution.
-
-We minimize the Gaussian pullback objective induced by the shared reference,
-\begin{equation}
-  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h_k'(u_k) \Big],
+so the Jacobian contribution at stage $k$ depends only on $u_k$. This structure yields $\mathcal{O}(K)$ per-sample cost and exact inversion by back-substitution.
+
+\textbf{Objective.} We minimize the Gaussian pullback induced by the shared reference (Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-det}):
+\begin{equation}
+  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h'_k(u_k) \Big].
   \label{eq:ttm-separable-loss}
 \end{equation}
-which follows from the change-of-variables identity in Equation~\eqref{eq:transport-pullback} combined with the triangular determinant factorization in Equation~\eqref{eq:transport-det}.
-The quadratic term pulls the transformed coordinates toward the reference, and the log-derivative term prevents degenerate solutions.
-We solve the regularized problem with bound-constrained optimization and enforce monotone structure by construction.
-
-We construct $h_k$ with monotone one-dimensional bases that combine identity, integrated sigmoids, softplus-like edge terms, and integrated radial basis functions.
-Nonnegativity constraints on the derivative coefficients guarantee $h_k'(u_k) \ge 0$.
-We linearize tails to stabilize likelihoods as $|u_k|$ grows.
-Ridge penalties apply to all basis coefficients, and optional sparsity penalties shrink context shifts when multicollinearity inflates variance.
-During training and evaluation we clip log-derivatives to $[-H, H]$ to avoid numerical overflow in the Jacobian sum; the bound $H$ is tuned on the validation split.
-
-We build $g_k$ from low-degree polynomial features of $u_{1:k-1}$ and drop predecessors whose inclusion does not improve validation likelihood.
-This pruning keeps $\nabla_u S(u)$ sparse and improves stability in small-sample regimes.
-Ordering matters for finite bases, so headline results use the natural variable order while robustness studies vary the order as described in Section~\ref{sec:datasets-preprocessing}.
-When heuristics are applied, we evaluate two candidates on the validation split—identity and a Cholesky-pivoted ordering with optional Gaussianization—and select the ordering with the lower validation NLL.
-Appendix~\ref{ch:appendix} records how the ordering is stored and reapplied at prediction time.
-
-We reference a cross-term variant, denoted TTM-X, only to delimit scope. The variant augments the separable component with low-rank interactions,
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k) + \sum_{j<k} \alpha_{kj}\,q_j(u_j)\,r_k(u_k),
+We solve this objective with constrained optimization and enforce monotonicity by construction.
+
+\textbf{Parameterization.}
+\begin{itemize}
+  \item $h_k$: one-dimensional monotone bases (integrated radial basis functions, splines, and linear tails) with derivative nonnegativity so that $h'_k > 0$; all coefficients are learned jointly under the positivity constraint enforced by the optimizer.
+  \item $g_k$: low-degree polynomial features of $u_{1:k-1}$ with light ridge regularization to keep $\nabla_u S(u)$ stable.
+\end{itemize}
+
+\paragraph{Ordering.} Feature ordering matters for triangular models.
+We store the learned order and reuse it at prediction time.
+For permutation studies (Section~\ref{sec:synthetic-results}) we apply the same bookkeeping so that permutations are reproducible.
+
+\subsection{Cross-term Variant (TTM-X)}
+
+We reference a low-rank interaction variant only to delimit the class used in the synthetic analyses:
+\begin{equation}
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k) + \sum_{j<k} \alpha_{kj} \, q_j(u_j) \, r_k(u_k),
   \label{eq:ttm-cross}
 \end{equation}
-where $q_j$ and $r_k$ are monotone features and constraints ensure $\partial_{u_k} S_k(u_{1:k}) > 0$.
-We exclude TTM-X from headline tables because the interactions alter identifiability and complicate calibration.
-The definition clarifies the naming used in the synthetic analyses.
-
-We implement Transformation Random Forests with additive predictors and denote the model TRTF.
-Implementations rely on the \texttt{partykit} toolkit for recursive partitioning \citep{partykit2015} together with the transformation-model framework \citep{hothorn2017transformation,hothorn2018conditional}.
-Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by the forest.
-The induced triangular component is
+where $q_j$ and $r_k$ are monotone features.
+Constraints ensure $\partial_{u_k} S_k > 0$ despite the cross-term.
+We do not promote TTM-X as a baseline; the definition clarifies the naming and scope used in the synthetic results.
+
+\subsection{Transformation Random Forests (TRTF)}
+
+We use Transformation Random Forests (TRTF) as an alternative route to a triangular component from conditional CDFs \citep{hothorn2017transformation,hothorn2018conditional}.
+Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the forest CDF.
+We define
 \begin{equation}
@@
 \end{equation}
-and differentiation yields $\varphi\big(S_k(u_{1:k})\big)\,\partial_{u_k} S_k(u_{1:k}) = \widehat{\pi}_k(u_k \mid u_{1:k-1})$.
-Under the additive predictor $\widehat{F}_k(u_k \mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big)$ we obtain $S_k = h_k + g_k$ and $\partial_{u_k} S_k = h_k'(u_k)$, which matches Equation~\eqref{eq:ttm-separable-def} exactly in the transport frame.
-Consequently TRTF shares the likelihood in Equation~\eqref{eq:ttm-separable-loss}, inherits exact inversion, and differs operationally through forest training and aggregation.
+which maps $u_k$ into the standard normal reference conditional on $u_{1:k-1}$. By the chain rule,
+\[
+  \varphi\big(S_k(u_{1:k})\big)\, \partial_{u_k} S_k(u_{1:k}) = \widehat{f}_k(u_k \mid u_{1:k-1}),
+\]
+so the TRTF-induced Jacobian matches the forest’s conditional density. The triangular structure mirrors TTM-Sep; the difference is operational (forest training and aggregation) rather than conceptual.
+
+\subsection{Copula Baseline with Estimated Marginals}
 
 We keep copulas as dependence baselines with explicit scope.
-We fit only low-dimensional nonparametric copulas for $K \le 3$, using probit-transformed pseudo-observations and kernel density estimation on the Gaussian scale before mapping back to the unit cube with the appropriate Jacobian.
-The independence baseline evaluates the product of fitted marginals.
-We omit a Gaussian copula from the main experiments to preserve consistency with the low-$K$ nonparametric dependence analyzed in Section~\ref{sec:realdata}.
-
-We adopt a single inversion and evaluation convention across estimators.
-Training, Jacobians, and conditional evaluations occur in standardized coordinates.
-Sampling draws $z \sim \mathcal{N}(0, I)$, applies $S^{-1}$ by back-substitution, and converts to the original scale with the stored affine parameters.
-This convention prevents mixed objectives and keeps all reported quantities interoperable.
-
-We ensure reproducibility and comparability with fixed seeds, cached standardization parameters, and shared reporting utilities.
-Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction, nonparametric copulas, marginal and separable triangular maps, and the shared transport core that implements ordering, bases, derivatives, and evaluation.
-The appendix also records optimizer choices, timing hooks, and object layouts used in the experiments.
-
+Marginals are estimated on the training split, and we report the joint density on the original $x$-scale.
+Concretely, we fit $\widehat{F}_k$ and $\widehat{f}_k$ for each coordinate, transform to pseudo-observations $U_k = \widehat{F}_k(X_k)$, and, for Gaussian copulas, set $Z_k = \Phi^{-1}(U_k)$ to model dependence in a latent normal space.
+We evaluate the copula density $c(U)$ and multiply by $\prod_{k=1}^K \widehat{f}_k(x_k)$ to obtain the reported joint density.
+
+\paragraph{Label use (Half-Moon only).} For $K=2$ in the Half-Moon study we fit class-conditional copulas for the blue/red classes and mix them by empirical class priors when reporting the unlabeled joint.
+No other model uses labels.
+Details and diagnostics appear in Section~\ref{sec:synthetic-results}.
+
+\subsection{Shared Inversion and Evaluation}
+
+We adopt a single inversion and evaluation convention across estimators using Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-det}.
+The alignment is necessary for conditional decomposition checks in Section~\ref{sec:synthetic-results} and for the compute summaries in Section~\ref{sec:realdata}.
+
+\paragraph{Reproducibility.} We fix random seeds, cache splits, and expose the same object layouts for transport maps, TRTF, and copulas.
+Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction; we reuse the same logging, timing hooks, and file formats across models.
 \section{Evaluation Metrics and Protocol}\label{sec:evaluation-protocol}
@@
 % Checklist: align references, keep active voice, reuse existing notation.
-This section defines the metrics and procedures applied across all models so likelihoods, calibration, and compute remain directly comparable.
-We evaluate every estimator in standardized coordinates, apply the triangular determinant, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.
-
-We distinguish pointwise log density from dataset averages.
-The test log likelihood (LL) is the mean of $\log \hat{\pi}_X(x)$ over the test split, and the test negative log likelihood (NLL) is its negative.
-Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity.
-Reported log densities on the original scale equal the standardized quantity minus $\sum_k \log \sigma_k$ as given by Equation~\eqref{eq:transport-affine}.
-Consequently, datasets with larger training scales introduce large constant offsets that the affine correction removes.
-
-Triangular models exploit the separable pullback in standardized coordinates. With $u = T_{\mathrm{std}}(x)$, the log density decomposes as
+\begin{quote}
+\textbf{Goal.} Define the metrics and the evaluation routine applied across all estimators.
+We reuse the notation and preprocessing conventions from Table~\ref{tab:transport-notation}, with background in Chapter~\ref{ch:background} and dataset preparation in Section~\ref{sec:datasets-preprocessing}.
+All log-likelihood (LL) and negative log-likelihood (NLL) values are reported in nats; figures and tables state $\text{LL}~(\uparrow~\text{better})$ and $\text{NLL}~(\downarrow~\text{better})$ explicitly.
+\end{quote}
+
+\subsection{Joint and Conditional Decomposition}
+
+Triangular models evaluate densities in standardized coordinates and then report on the original scale.
+With $u = T_{\mathrm{std}}(x)$ determined by the training-split $(\mu, \sigma)$, any estimator with lower-triangular $S$ satisfies
 \begin{equation}
@@
 \end{equation}
-so the determinant factorization in Equation~\eqref{eq:transport-det} yields linear per-sample cost in $K$.
-In plain language, the model checks how Gaussian each transformed coordinate looks, then adds the exact log-Jacobian contribution from its one-dimensional derivative.
-The affine correction in Equation~\eqref{eq:transport-affine} converts $\log \hat{\pi}_U$ to $\log \hat{\pi}_X$ for reporting.
-
-We report per-dimension conditional NLLs for triangular models to localize error. For each coordinate,
+so the determinant factorization from Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-det} applies componentwise.
+We transform back to $x$ once per evaluation via the affine correction in Equation~\eqref{eq:transport-affine}.
+
+We localize error with per-dimension conditional NLLs. For coordinate $k$,
 \begin{equation}
@@
 \end{equation}
-and the joint NLL equals $\sum_{k=1}^{K} \mathrm{NLL}_k$ by construction.
-Copulas lack a unique triangular factorization, so we report only their joint NLL.
-Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
-These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame under the additive-predictor and monotone-smoothing assumptions stated in Section~\ref{sec:transport-trtf}.
-
-Calibration assesses whether predictive probabilities align with empirical frequencies.
-For triangular models we form conditional probability integral transform (PIT) values $V_{ik} = \widehat{F}_k(u_{ik} \mid u_{i,1:k-1})$ on the test split and expect independent $\mathrm{Unif}(0,1)$ draws under correct calibration \citep{gneiting2007probabilistic}.
-We summarize departures from uniformity with the Kolmogorov--Smirnov statistic $D_n = \sup_t \lvert \widehat{F}_n(t) - t \rvert$ and report the associated $p$-value \citep{massey1951kolmogorov}.
-We complement the scalar summary with brief PIT distribution descriptions when patterns recur across seeds.
-For copulas we assess marginal PITs and low-dimensional slices where dependence is transparent.
-Systematic U-shaped or inverted-U PIT indicates under- or over-dispersion and motivates richer parameterizations.
-
-Compute metrics quantify practical cost alongside fit.
-We record wall-clock training time on the training split and per-sample evaluation time on the test split.
-Triangular transports scale linearly in $K$ and approximately linearly in the number of basis functions.
-Transformation Random Forests scale with the number and depth of trees per conditional during training, while prediction remains linear after aggregation.
-Copula training is dominated by correlation estimation or kernel density fitting, followed by fast evaluation.
-We also track peak resident memory when caching affects runtime.
-All timings use the deterministic pipeline defined in Chapter~\ref{ch:dataanalysis} and are averaged over seeds with standard errors.
-
-Protocol choices keep comparisons stable and reproducible:
-\begin{enumerate}
-  \item Standardize with training-split statistics, fit a single map $S:u \rightarrow z$, and evaluate Jacobians in standardized space.
-  \item Compute LL, NLL, and conditional decompositions in standardized coordinates, then apply the affine correction once for reporting.
-  \item Evaluate PIT diagnostics, Kolmogorov--Smirnov statistics, and compute metrics on the fixed test split with seeds $\{11, 13, 17, 19, 23\}$ and quote means with $\pm$ two standard errors across seeds (SE $= s/\sqrt{m}$ over $m$ seeds).
-\end{enumerate}
-Appendix~\ref{ch:appendix} lists routine interfaces that support exact re-execution; figure captions and table notes repeat units, splits, and the “higher/lower is better” convention for clarity.
-
-Numerical safeguards prevent unstable likelihoods from dominating summaries.
-We enforce strictly monotone derivatives by construction and clip log-derivative contributions to $[-H, H]$ during training and evaluation.
-We tune $H$ and regularization on the validation split and reuse the selected bound on the test split.
-We report clipping status inline per study and keep the exact bound values with the corresponding experiment logs to avoid duplication.
-This practice controls overflow in the Jacobian sum without masking systematic misfit that PIT diagnostics would reveal.
-
-Scope limits clarify non-goals.
-We do not report AIC or BIC because effective parameter counts are not comparable across estimators in this frame.
-We also do not adjust $p$-values for multiple PIT checks; instead, we treat Kolmogorov--Smirnov results as diagnostics and corroborate them with effect sizes and plots.
-These limits keep the evaluation focused on likelihood, calibration, and compute under a single, transparent protocol.
-
+and the joint NLL satisfies $\mathrm{NLL}_{\mathrm{joint}} = \sum_{k=1}^{K} \mathrm{NLL}_k$ by the triangular factorization.
+Negative per-dimension NLLs can occur on bounded or highly concentrated supports (e.g., beta or gamma margins); we revisit this behaviour in Section~\ref{sec:synthetic-results}.
+
+\subsection{Compute Metrics}
+
+We report practical costs alongside fit:
+\begin{itemize}
+  \item \textbf{Train time per epoch} and total wall-clock to the chosen early-stopping criterion when applicable.
+  \item \textbf{Evaluation throughput} (test points per second) together with per-sample latency.
+  \item \textbf{Model size,} recorded as effective parameter counts when the backend exposes them.
+\end{itemize}
+Compute summaries accompany the real-data tables in Section~\ref{sec:realdata}; extended profiles appear in Appendix~\ref{ch:appendix} when needed.
+
+\subsection{Protocol (Fixed Across Methods)}
+
+\begin{enumerate}
+  \item \textbf{Preprocessing.} Standardize features with training-split $(\mu, \sigma)$, evaluate derivatives and Jacobians in $u$, and convert to $x$ with Equation~\eqref{eq:transport-affine} once per evaluation.
+  \item \textbf{Evaluation in standardized space.} Compute LL, NLL, and conditional decompositions with Equation~\eqref{eq:evaluation-triangular}; map results to the original scale for reporting.
+  \item \textbf{Aggregation and uncertainty.} Average metrics across seeds and report $\pm$ two standard errors (SE $= s/\sqrt{m}$ over $m$ seeds).
+  \item \textbf{Reproducibility hooks.} Cache splits and random seeds, and log objective values together with compute metrics using shared file schemas across estimators.
+\end{enumerate}
+
+These choices keep likelihood reporting in standardized coordinates, align diagnostics, and make compute measurements comparable.
+Appendix~\ref{ch:appendix} lists the supporting routine interfaces and object layouts used in the experiments.
 \section{Synthetic Results and Diagnostics}\label{sec:synthetic-results}
@@
 This section reports synthetic results for the Half-Moon and four-dimensional generators under the protocol in Section~\ref{sec:evaluation-protocol}.
-We summarize mean test negative log likelihoods, per-dimension conditional NLLs, calibration evidence, and ordering robustness, referencing the corresponding tables and figures.
+We summarize mean test negative log likelihoods, per-dimension conditional NLLs, and ordering robustness, referencing the corresponding tables and figures.
 
@@
 Per-dimension NLLs confirm that the first coordinate is harder: TRTF reported $(1.23, 0.47)$, while TTM-Sep reported $(1.28, 0.65)$.
-Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline in Figure~\ref{fig:transport-schematic} of Appendix~\ref{ch:appendix}.\;Clipping status: not triggered in these runs (no log-derivative terms reached the bound).
+Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline described earlier in Chapter~\ref{ch:background}.\;Clipping status: not triggered in these runs (no log-derivative terms reached the bound).
 
@@
 
-Calibration assessments align with the likelihood evidence.
-Figure~\ref{fig:autoregressive-joint-calibration} (p.~\pageref{fig:autoregressive-joint-calibration}) shows joint log-density calibration against the oracle, with residual structure visible for triangular transports in the canonical ordering.
-Conditional PIT diagnostics and Kolmogorov--Smirnov distances, computed as in Section~\ref{sec:evaluation-protocol}, exhibited the same qualitative patterns across seeds, so we omit redundant tables.
-
 These studies indicate that TRTF closes part of the gap to oracle likelihoods while preserving the triangular evaluation frame.
@@
 Section~\ref{sec:realdata} turns to real-data benchmarks and compute summaries under the same protocol.
-
-\paragraph{Calibration numbers.} To complement the visual diagnostics, Table~\ref{tab:ks-synth} reports median Kolmogorov--Smirnov (KS) distances of probability-integral-transform (PIT) values per coordinate, aggregated across seeds ($\pm$ 2SE).
-Lower is better.
-We include entries for methods with an accessible marginal CDF in our implementation.
-
-\begin{table}[htbp]
-  \centering
-  \textit{(median KS of PIT per coordinate; lower is better).}\
-
-  \begin{tabular}{lrrrrrr}\
-    \hline\
-    Dataset & True (Joint) & True (marginal) & Random Forest & Marginal Map & Separable Map & Copula NP \\\n     \hline\
-    Half-Moon & -- & 0.079 $\pm$ 0.015 & NA $\pm$ NA & 0.078 $\pm$ 0.015 & 0.090 $\pm$ 0.018 & 0.067 $\pm$ 0.001 \\
-    4D generator & -- & -- & -- & -- & -- & -- \\
-    \hline\
-  \end{tabular}
-  \caption{Calibration via PIT--KS on synthetic datasets: median KS distance per coordinate (mean $\pm$ 2SE across seeds).
-Entries marked `--' indicate that the CDF was not available in the corresponding backend.}
-  \label{tab:ks-synth}
-\end{table}
 
@@
 Our TRTF result attains $-30.01 \pm 1.26$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
-This ranking is consistent with the separable Jacobian and additive predictors discussed in Section~\ref{sec:models-implementation}.
+This ranking is consistent with the separable Jacobian and the forest aggregation discussed in Section~\ref{sec:models-implementation}.
 The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
@@
   - Standardize features with training-split $(\mu,\sigma)$ only; compute all derivatives/Jacobians in $u$; report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
-  - TRTF uses additive predictors and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
+  - TRTF uses forest aggregation and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
 
@@
 Cholesky-pivoted (with optional Gaussianization) according to validation NLL.
-  - TRTF: additive predictor with forest aggregation; strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated; we record the number of trees, depth and split rules in the experiment logs.
+  - TRTF: forest aggregation with strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated; we record the number of trees, depth and split rules in the experiment logs.
   - Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection; independence and Gaussian baselines are used only for reference in text where noted.
@@
 We record package versions via \texttt{sessionInfo()} in run logs.
-  - Single-threaded BLAS by default; optional parallel training for TRTF via \texttt{options(trtf.train_cores=4)} when available.
+  - Single-threaded BLAS by default; optional parallel training for TRTF via \texttt{options(trtf.train\_cores = 4)} when available.
   - CPU-only runs on a laptop-class machine; logs include hardware notes (CPU model, RAM) and wall-clock timings (Table~\ref{tab:real-compute}).
@@
 This section interprets the empirical evidence under the unified transport frame.
-We focus on TRTF (additive predictor), TTM-Sep, and, where applicable, copula baselines (only for $K\!\le\!3$) evaluated with matched preprocessing, metrics, and units.
+We focus on TRTF, TTM-Sep, and, where applicable, copula baselines (only for $K\!\le\!3$) evaluated with matched preprocessing, metrics, and units.
 Synthetic studies report NLL, real datasets report LL, and we apply the shared affine correction.
@@
 
-TRTF often leads within the separable family because additive predictors shift conditional location while the underlying monotone shapes remain stable.
+TRTF often leads within the separable family because the forest aggregation shifts conditional location while the underlying monotone shapes remain stable.
 The likelihood identities equate TRTF with separable triangular maps, so observed gaps arise from how each estimator realizes context shifts and stabilizes derivatives.
@@
 Performance on MINIBOONE reveals the cost of separability at higher dimension.
-TRTF improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates.
-POWER exhibited the opposite regime: under identical preprocessing, the reported flows outperformed TRTF (Table~\ref{tab:uci-loglik} lists TRTF at $-7.17$ versus flow baselines near $0.17$ to $0.30$).
-These contrasts suggest that conditional shape and dimension jointly determine whether separable structure suffices.
+TRTF improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates (TRTF recorded $-30.01$ while the strongest flow reached $-11.68$).
+POWER showed the complementary regime: under identical preprocessing, TRTF remains below the flow baselines (cf.
+Table~\ref{tab:uci-loglik}, where TRTF records $-7.17$ against flow baselines near $0.17$ to $0.30$).
+This contrast indicates that the gap is driven by conditional shape differences rather than pure location shifts.
 Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.
@@
 Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map.
-Low-rank cross-terms in triangular transports and non-additive predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation.
+Low-rank cross-terms in triangular transports and non-separable predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation.
 We excluded these richer variants by design in Chapter~\ref{ch:intro} (Non-goals) due to compute and calibration overhead; they remain promising future work once resources permit.
-Expanded calibration reporting, including probability-integral-transform summaries and Kolmogorov--Smirnov distances, should remain part of any deployment-grade evaluation.
 Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off.
@@
 \chapter{Appendix}\label{ch:appendix}
-
-\section{Unified Transport Schematic}\label{app:transport-schematic-figure}
-Figure~\ref{fig:transport-schematic} provides the full schematic of the unified transport pipeline referenced throughout the thesis.
-The landscape layout preserves readability for the granular annotations on each modeling branch.
-
-\begin{landscape}
-\begin{figure}[p]
-\centering
-\begin{tikzpicture}[
-  scale=0.92,
-  transform shape,
-  >=Latex,
-  node distance=8mm,
-  process/.style={draw, rounded corners, align=center, font=\small, text width=56mm, minimum height=12mm, inner sep=4mm, fill=white},
-  header/.style={draw, rounded corners, align=center, font=\small\bfseries, text width=56mm, minimum height=12mm, inner sep=4mm, fill=gray!15},
-  arrow/.style={->, thick}
-]
-
-\node[process, font=\small\bfseries, text width=64mm] (data) {Input data: original features $x$};
-\node[process, font=\small\bfseries, text width=72mm, below=10mm of data] (std) {Train-only standardization; reuse $\mu$, $\sigma$ from Appendix~\ref{app:ttm-core}};
-\node[process, font=\small\bfseries, text width=80mm, below=8mm of std] (jac) {Common pullback evaluation; accumulate $\log\eta(S(u)) + \log|\det \nabla_u S(u)|$ (Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-affine})};
-
-\node[header, below=18mm of jac, xshift=-70mm] (ttmTitle) {Triangular transport maps};
-\node[header, below=18mm of jac] (trtfTitle) {Transformation Random Forests};
-\node[header, below=18mm of jac, xshift=70mm] (copTitle) {Copula models};
-
-\matrix (ttmMatrix) [below=6mm of ttmTitle, column sep=6mm, row sep=4mm] {
-  \node[header, text width=52mm] (ttmMargTitle) {TTM-Marginal}; & \node[header, text width=52mm] (ttmSepTitle) {TTM-Separable}; \\
-  \node[process, text width=52mm] (ttmMargStepOne) {Closed-form affine coefficients $(a_k, b_k)$ in $u$-space}; & \node[process, text width=52mm] (ttmSepStepOne) {Context basis $g_k$ on predecessors, monotone basis $h_k$ on $u_k$}; \\
-  \node[process, text width=52mm] (ttmMargStepTwo) {Forward pass via \texttt{ttm\_forward}; sum log Jacobian diagonals}; & \node[process, text width=52mm] (ttmSepStepTwo) {Solve constrained optimization, enforce positivity, store Jacobian terms}; \\
-};
-
-\node[process, below=6mm of trtfTitle, text width=62mm] (trtfStepOne) {Estimate monotone marginals (Appendix~\ref{app:trtf})};
-\node[process, below=6mm of trtfStepOne, text width=62mm] (trtfStepTwo) {Fit transformation forests for $X_k \mid X_{1:(k-1)}$ with additive predictor};
-\node[process, below=6mm of trtfStepTwo, text width=62mm] (trtfStepThree) {Recover triangular likelihood via probability integral transform};
-
-\node[process, below=6mm of copTitle, text width=60mm] (copStepOne) {Estimate train marginals $\widehat F_k$, densities $\widehat f_k$};
-\node[process, below=6mm of copStepOne, text width=60mm] (copStepTwo) {Fit dependence on probit scores (KDE copula)};
-\node[process, below=6mm of copStepTwo, text width=60mm] (copStepThree) {Combine marginals with copula-based dependence};
-
-\node[process, font=\small\bfseries, text width=150mm, below=26mm of trtfStepThree] (outputs) {Reported outputs: log density, conditional diagnostics, samples, calibration, compute summaries};
-
-\draw[arrow] (data) -- (std);
-\draw[arrow] (std) -- (jac);
-\draw[arrow] (jac.south west) to[out=215,in=90] (ttmTitle.north);
-\draw[arrow] (jac) -- (trtfTitle);
-\draw[arrow] (jac.south east) to[out=325,in=90] (copTitle.north);
-
-\draw[arrow] (ttmTitle) -- (ttmMargTitle);
-\draw[arrow] (ttmTitle) -- (ttmSepTitle);
-\draw[arrow] (ttmMargTitle) -- (ttmMargStepOne);
-\draw[arrow] (ttmMargStepOne) -- (ttmMargStepTwo);
-\draw[arrow] (ttmSepTitle) -- (ttmSepStepOne);
-\draw[arrow] (ttmSepStepOne) -- (ttmSepStepTwo);
-
-\draw[arrow] (trtfTitle) -- (trtfStepOne);
-\draw[arrow] (trtfStepOne) -- (trtfStepTwo);
-\draw[arrow] (trtfStepTwo) -- (trtfStepThree);
-
-\draw[arrow] (copTitle) -- (copStepOne);
-\draw[arrow] (copStepOne) -- (copStepTwo);
-\draw[arrow] (copStepTwo) -- (copStepThree);
-
-\draw[arrow] (ttmMargStepTwo) to[out=-90,in=195] (outputs.west);
-\draw[arrow] (ttmSepStepTwo) to[out=-90,in=165] (outputs.west);
-\draw[arrow] (trtfStepThree) -- (outputs);
-\draw[arrow] (copStepThree) to[out=-90,in=0] (outputs.east);
-
-% Legend for symbols
-\node[header, below=10mm of outputs, text width=120mm] (legendTitle) {Legend};
-\node[process, below=2mm of legendTitle, text width=120mm, align=left] (legendBox) {
-  $u$: standardized coordinates ($u = T_{\mathrm{std}}(x)$)\\
-  $z$: reference coordinates after transport\\
-  $S$: lower-triangular transport map ($S:u \mapsto z$)\\
-  $\eta$: $K$-variate standard normal density\\
-  $\nabla_u S(u)$: Jacobian of $S$ with respect to $u$
-};
-
-\end{tikzpicture}%
-\caption{Unified evaluation pipeline shared by triangular transport maps, Transformation Random Forests, and copulas.
-The diagram shows how standardized features flow through the triangular pullback or copula dependence block before reporting log densities, conditional diagnostics, samples, calibration metrics, and compute summaries.
-The shared preprocessing, Jacobian accumulation, and reporting path define the evaluation protocol used across the thesis.}
-\label{fig:transport-schematic}
-\end{figure}
-\end{landscape}
 
@@
 
-These structured summaries allow reproducing the algorithmic flow of each model without navigating the full R implementation.
-
-\subsection{Supplementary Results}\label{app:supplementary}
-
-\begin{table}[htbp]
-  \centering
-  \textit{(mean NLL in nats).}
-  \caption{Permutation-averaged joint test negative log-likelihood for TTM-Sep at $n=25$ on the four-dimensional synthetic generator (aggregated over all $24$ permutations).}
-  \label{tab:ttmsep-n25-overflow}
-  \begin{tabular}{lr}
-    \hline
-    Model & $n=25$ \\
-    \hline
-    TTM-Sep$^{\ast}$ & $6829.45$ \\
-    \hline
-  \end{tabular}
-  \caption*{Note: $^{\ast}$ out-of-scope setting.
-This value reflects numerical overflow of the separable map in the sparse regime ($n=25$).
-Stronger derivative clipping and ridge regularization (Section~\ref{sec:models-implementation}) remove this failure in reruns.
-We mark this configuration as out of scope and exclude it from main-text comparisons; the table remains for transparency.}
-\end{table}
-
-\begin{figure}[htbp]
-  \centering
-  % Simple TikZ markers for ordering sensitivity and mitigation window
-  \begin{tikzpicture}[x=28mm,y=18mm]
-    % Axes labels
-    \node[align=center] at (-0.2,2.2) {Joint NLL\\(lower is better)};
-    % TRTF group
-    \node[anchor=east] at (-0.1,1.4) {TRTF};
-    % TRTF values (from Tables in Chapter 3): min=4.46, canonical=4.53, median=4.59
-    \draw[thick,gray!50] (0.0,1.0) -- (0.0,2.0);
-    \draw[thick] (0.0,1.23) -- ++(0.12,0) node[anchor=west,scale=0.8] {min $4.46$};
-    \draw[thick] (0.0,1.265) -- ++(0.12,0) node[anchor=west,scale=0.8] {canonical $4.53$};
-    \draw[thick] (0.0,1.295) -- ++(0.12,0) node[anchor=west,scale=0.8] {median $4.59$};
-
-    % TTM-Sep group
-    \node[anchor=east] at (-0.1,0.6) {TTM-Sep};
-    % TTM-Sep values: min=5.48, canonical=5.66, median=5.60
-    \draw[thick,gray!50] (0.0,0.2) -- (0.0,1.2);
-    \draw[thick] (0.0,0.49) -- ++(0.12,0) node[anchor=west,scale=0.8] {min $5.48$};
-    \draw[thick] (0.0,0.57) -- ++(0.12,0) node[anchor=west,scale=0.8] {median $5.60$};
-    \draw[thick] (0.0,0.63) -- ++(0.12,0) node[anchor=west,scale=0.8] {canonical $5.66$};
-
-    % Brackets indicating potential window for data-driven order selection
-    \draw[decorate,decoration={brace,amplitude=5pt}] (-0.02,1.23) -- (-0.02,1.295) node[midway,xshift=-8mm,align=center,scale=0.8] {potential\\gain};
-    \draw[decorate,decoration={brace,amplitude=5pt}] (-0.02,0.49) -- (-0.02,0.63) node[midway,xshift=-8mm,align=center,scale=0.8] {potential\\gain};
-  \end{tikzpicture}
-  \caption{Ordering sensitivity and mitigation window on the four-dimensional generator at $n=250$ (joint NLL; nats; lower is better).
-Markers show the best (min over $24$ permutations), canonical ordering, and permutation median for each method (values from Chapter~\ref{ch:dataanalysis}: Tables~\ref{tab:autoregressive-nll} and~\ref{tab:autoregressive-perm-spread}).
-A simple heuristic selects the better of two candidates—identity and Cholesky-pivoted with optional Gaussianization—aiming to move toward the “min” marker while keeping evaluation linear.}
-  \label{fig:ordering-heuristics-4d}
-\end{figure}
-
-% Clipping bounds are reported inline per study in Chapter~\ref{ch:dataanalysis} and stored with the experiment logs; we omit a dedicated summary table to avoid duplication.
-
 \end{document}

### 2025-09-30 report/chapter01.Rnw
.

--- 2025-09-20 report/chapter01.Rnw
@@
 
-Estimating the joint density $\pi_X(x)$ of a random vector $x \in \mathbb{R}^K$ underpins probabilistic modelling, anomaly detection, simulation-based inference, and decision making under uncertainty.
-Modern applications combine high dimension with structure that changes across contexts: conditional variance may grow with predictors, skewness can flip sign across regions, and mixtures can create multimodal slices.
-A coherent response is \emph{measure transport}: we couple the unknown target with a simple reference through an invertible transformation and exploit that coupling to compute likelihoods, draw samples, and evaluate conditionals.
-Throughout we adopt the notation established in a recent tutorial; Chapter~\ref{ch:background} summarises the relevant background.
-We standardise observations, learn monotone triangular maps to a Gaussian reference, and report log densities on the original scale using a fixed affine correction.
-All Jacobians and partial derivatives involving the transport are taken in the standardised space unless stated otherwise.
-
-This thesis compares three model classes that realise the transport programme with different inductive biases: triangular transport maps parameterised directly, transformation random forests that induce transports via the probability integral transform, and copulas that decouple marginals from dependence.
-Each offers a distinct view on conditional density estimation and raises the question of when separable structure suffices and when richer cross-term capacity is required.
-The following sections outline our contributions, the research questions we address, and the scope of the work.
-
-\section{Contributions and research questions}
-
-We make three contributions within a single, notation-coherent framework.
+Multivariate density estimation is a cornerstone of modern statistics and machine learning.
+It supplies the machinery for likelihood-based inference, simulation of generative mechanisms, and uncertainty quantification in tabular domains.
+Practitioners, however, routinely face a delicate trade-off between three goals: expressive power, computational feasibility, and interpretability.
+Three complementary families of estimators have emerged around this dilemma:
 \begin{enumerate}
-  \item \emph{Theory.} We clarify the link between transformation forests and triangular transport and formalise when the induced maps are separable \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-  \item \emph{Empirics.} We compare separable and cross-term triangular maps, transformation forests under additive predictors, and copulas on synthetic and real tabular data through a unified evaluation protocol that reports likelihoods in nats alongside calibration diagnostics.
-  \item \emph{Implementation.} We document the practical choices that matter in transport-based modelling, including standardisation, variable ordering, and numerical safeguards for cross-term training, while keeping the introductory chapter focused on the high-level problem.
+  \item \textbf{Normalizing flows} build expressive change-of-variables maps and have become central in deep generative modeling \citep{rezende2015variational,papamakarios2017masked,dinh2017real,durkan2019neural,kingma2018glow,papamakarios2021normalizing}.
+  \item \textbf{Transformation models and forests} model conditional distributions through parametric transformation structures coupled with ensemble methods \citep{hothorn2017transformation,hothorn2018conditional,hothorn2021transformation}.
+  \item \textbf{Copulas} separate marginals from dependence, yielding interpretable multivariate association models \citep{sklar1959fonctions,nelsen2006introduction,joe2014dependence,nagler2017kdecopula}.
 \end{enumerate}
 
-Two questions motivate the empirical study.
+This thesis situates all three families inside a common \textbf{transport framework} that makes them directly comparable \citep{rosenblatt1952remarks,knothe1957contributions,bogachev2005triangular,ramgraber2025friendly}.
+The framework operates in standardized coordinates $u$ computed solely from training-split statistics to avoid leakage and keep objectives aligned across models.
+Within this evaluation framework each estimator couples $u$ to a reference distribution---Gaussian for triangular maps and TRTF, $\mathrm{Unif}([0,1]^K)$ for copulas---and recovers densities via exact change of variables.
+Section~\ref{sec:ch1-frame} previews the core pieces and Chapter~\ref{ch:background} develops the mathematics in detail.
+
+Separable triangular transport maps play a central role throughout the thesis.
+Their lower-triangular Jacobian delivers a log-determinant in $\mathcal{O}(K)$ operations, back-substitution yields an explicit inverse for sampling, and the triangular structure produces exact conditional densities in a fixed order.
+Fully triangular maps can require $\mathcal{O}(K^2)$ work per evaluation, whereas separable variants reduce both forward and inverse passes to $\mathcal{O}(K)$, making them attractive for systematic comparison while preserving interpretable marginals and dependence mechanisms.
+
+\medskip
+\noindent\textit{Notation.} We write $\varphi$ and $\Phi$ for the density and cumulative distribution function of a univariate standard normal random variable, and $\eta$ for the $K$-variate standard normal density.
+All models are evaluated in standardized coordinates $u$ and reported on the original scale $x$ through a diagonal affine correction so that log densities remain comparable.
+Unless stated otherwise, log values are expressed in natural units (nats).
+
+\section{Thesis and Problem Statement}\label{sec:ch1-problem}
+
+The goal of this thesis is to examine multivariate density estimation for tabular data within a unified transport-based evaluation protocol.
+We focus on three estimators: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and nonparametric copula models.
+
+Let $u = D^{-1}(x - \mu)$ with $D = \mathrm{diag}(\sigma_1,\ldots,\sigma_K)$ determined from the training split.
+If $T: \mathbb{R}^K \to \mathbb{R}^K$ is lower triangular, $z = T(u)$, and $\eta$ denotes the chosen reference density, then the change-of-variables formula gives
+\begin{equation}
+  \log p_X(x) = \log \eta\big(T(u)\big) + \sum_{k=1}^{K} \log \frac{\partial T_k}{\partial u_k}(u_{1:k}) - \sum_{k=1}^{K} \log \sigma_k,
+  \label{eq:ch1-change-of-variables}
+\end{equation}
+so reporting on the $x$-scale differs from reporting on the $u$-scale by the constant $-\sum_k \log \sigma_k$ per data point.
+The Rosenblatt and Knothe rearrangements guarantee the existence of such triangular transports for any coordinate ordering \citep{rosenblatt1952remarks,knothe1957contributions}.
+
+The copula baselines considered here are nonparametric kernel copulas \citep{nagler2017kdecopula}.
+We fit smooth univariate marginals with \texttt{kde1d}, form pseudo-observations using mid-rank empirical CDFs during training to avoid double smoothing, and evaluate joint densities through
+\[
+  p_X(x) = c\big(F_1(x_1),\ldots,F_K(x_K)\big)\, \prod_{k=1}^{K} f_k(x_k),
+\]
+where $c$ is the copula density and $F_k = \int f_k$ the smoothed marginal CDFs. No Gaussian copula enters the study; the emphasis remains on flexible, interpretable diagnostics in low dimensions.
+
+Separable triangular maps decompose each coordinate into a context-dependent shift plus a one-dimensional monotone transform.
+TRTF occupies an intermediate position: under strictly increasing conditional CDFs and the standard forest aggregation after monotone smoothing it reproduces the same separable triangular likelihood via the probability integral transform.
+The shared protocol described above keeps preprocessing, likelihood evaluation, calibration diagnostics, and compute measurements aligned across all estimators.
+
+\section{The Transport Frame on One Page}\label{sec:ch1-frame}
+
+The key ingredients of the transport framework are:
+\begin{itemize}
+  \item \textbf{Standardization.} Coordinates $u = D^{-1}(x-\mu)$ use training-split statistics only; all logs are in nats.
+  \item \textbf{Change of variables.} Densities follow from $z = T(u)$ through Equation~\eqref{eq:ch1-change-of-variables}.
+  \item \textbf{Triangular log determinant.} Lower-triangular Jacobians imply $\log \det J_T(u) = \sum_k \log \partial_{u_k} T_k(u_{1:k})$, giving $\mathcal{O}(K)$ cost.
+  \item \textbf{Affine correction.} Reporting on $x$ applies the constant correction $-\sum_k \log \sigma_k$ per data point.
+  \item \textbf{Computational profile.} Fully triangular maps can incur $\mathcal{O}(K^2)$ evaluation and inversion, whereas separable maps reduce both passes to $\mathcal{O}(K)$ while keeping the log determinant linear in $K$.
+\end{itemize}
+
+Separable triangular transports implement each coordinate as the sum of a context shift and a monotone one-dimensional basis, maintaining determinant stability and consistent conditional behaviour.
+TRTF achieves the same likelihood under the conditions noted above, while nonparametric copulas operate on the unit hypercube with explicit marginals.
+Standardization $u = T_{\mathrm{std}}(x)$ feeds either the triangular transport branch (TTM variants, TRTF) or the copula branch, and both return log densities, conditional diagnostics, samples, calibration metrics, and compute summaries.
+
+\section{Contributions and Research Questions}\label{sec:ch1-contributions}
+
+The thesis contributes along three axes that move from theoretical unification to empirical benchmarking and practical interpretation:
 \begin{enumerate}
-  \item \emph{Practical performance.} How do transformation forests perform on synthetic and real tabular data relative to triangular transport maps, copulas, and published normalising-flow baselines?
-We ask when separable transports reach the performance of high-capacity references and where they fall short once conditional shape depends on context \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing,hothorn2017transformation,hothorn2021transformation}.
-  \item \emph{Trade offs.} What are the trade offs among fit (test NLL), computational efficiency (training wall clock time and per-sample evaluation cost), and calibration diagnostics (probability integral transforms)?
-We examine where separable transports deliver the best speed-accuracy ratio and where cross-term capacity justifies its added complexity.
+  \item \textbf{Unified likelihood framework.} We formalise a shared transport perspective that places separable triangular maps, TRTF, and nonparametric copulas inside one mathematical scheme.
+The equivalence between TRTF and separable triangular transports holds under explicit regularity assumptions---strictly monotone conditional CDFs and the forest aggregation after smoothing.
+
+    \begin{quote}
+    \textbf{Proposition (informal).} Under axis-parallel partitions, strictly monotone conditional CDFs, and the forest aggregation after smoothing, the TRTF likelihood coincides with that of a separable triangular transport in standardized coordinates.
+    \end{quote}
+
+  \item \textbf{Benchmarking under a common protocol.} We design a standardized evaluation pipeline covering synthetic generators and real tabular benchmarks.
+TTM-Sep, its marginal variant, and TRTF are evaluated systematically; nonparametric kernel copulas act as diagnostic comparators when $K \le 3$.
+The pipeline produces three categories of evidence: (i) average test log likelihoods, (ii) conditional diagnostics based on probability integral transforms, and (iii) compute measurements for training effort and per-sample evaluation time.
+Together they quantify the trade-offs among accuracy, calibration, and efficiency.
+
+  \item \textbf{Operational insights.} The theoretical and empirical findings yield guidance for practitioners.
+We analyse sensitivity to ordering, the practical limits of separability, and when nonparametric copulas remain informative as diagnostic references.
+Chapter~\ref{ch:conclusion} synthesizes these insights and outlines extensions toward richer parameterisations.
 \end{enumerate}
 
-\section{Scope and positioning}
+Two research questions organise the empirical study:
+\begin{itemize}
+  \item \textbf{Synthetic data.} How do TRTF, TTM-Sep, and nonparametric copulas compare on controlled generators?
+Evidence comprises likelihood comparisons, conditional calibration metrics, and timing measurements.
+  \item \textbf{Benchmark datasets.} How close do TRTF results come to published normalizing flow baselines under identical preprocessing?
+Side-by-side likelihood comparisons, interpreted through the constraints of separability and compute profiles, answer this question.
+\end{itemize}
 
-We focus on triangular transport maps, transformation random forests, and copulas.
-Transport maps let us ablate separable versus cross-term capacity under a common objective and shared Jacobian conventions.
-Transformation forests bring a mature non-parametric estimator whose monotonicity is guaranteed by design and whose induced transport has a clear likelihood interpretation when additive predictors are used.
-Copulas provide semiparametric and non-parametric dependence models with transparent marginals, serving as interpretable baselines and highlighting elliptical dependence limits in higher dimensions.
-Normalising flows remain an important reference: we use them for context and draw on published tabular benchmarks because they represent high-capacity autoregressive and coupling architectures widely adopted in density estimation \citep{rezende2015variational,dinh2017real,papamakarios2021normalizing}.
+\begin{table}[t]
+  \centering
+  \caption{Model abbreviations used throughout the thesis.}
+  \label{tab:model-abbrev}
+  \begin{tabular}{ll}
+    \hline
+    Label & Meaning \\
+    \hline
+    TTM-Marg  & Marginal triangular transport (per-dimension; no context) \\
+    TTM-Sep   & Separable triangular transport (context shift $+$ monotone 1D transform) \\
+    TRTF      & Transformation Random Forests (axis-parallel splits) \\
+    True-Marg & Oracle marginal density \\
+    True-Joint& Oracle conditional joint density \\
+    Copula    & Nonparametric kernel copula baseline (smoothed marginals; no Gaussian copula) \\
+    \hline
+  \end{tabular}
+\end{table}
 
-\section{Takeaways and roadmap}
-
-A unified transport perspective enables direct comparisons across models with different inductive biases while keeping the introduction free of technical detail.
-Chapter~\ref{ch:background} develops the methodological background on monotone triangular transports, separable versus cross-term structure, and the transport interpretation of transformation forests.
-Chapter~\ref{ch:dataanalysis} describes the modelling procedures, data handling, and evaluation protocol before presenting empirical results.
-Chapter~\ref{ch:conclusion} synthesises the findings, discusses limitations, and outlines future work.
+Addressing these questions bridges rigorous transport-based theory with empirical validation, clarifying when simple triangular models suffice and when more expressive families become necessary.

### 2025-09-30 report/chapter02.Rnw
.

--- 2025-09-20 report/chapter02.Rnw
@@
 
-We adopt the tutorial's directionality and notation throughout: data are standardized to $u=(x-\mu)\oslash\sigma$, we learn a monotone triangular map $S:u\mapsto z$ that pushes the standardized target $\pi_U$ to the reference $\eta=\mathcal{N}(0,I)$, and we evaluate or sample on the original scale via the composition $M(x)=S(T_{\mathrm{std}}(x))$ with the usual affine Jacobian correction when reporting $\log \pi_X(x)$ (subtract $\sum_{k=1}^K \log \sigma_k$).
-All Jacobians and partial derivatives involving $S$ are taken in $u$-space unless explicitly stated.
-This chapter develops the minimal change-of-variables machinery, explains why triangular structure is effective, makes precise the contrast between separable and cross-term triangular maps, and shows how transformation random forests (TRTF) instantiate the same transport with an immediate likelihood identity \citep{rosenblatt1952remarks,knothe1957contributions,hothorn2017transformation,hothorn2021transformation,hothorn2018conditional,ramgraber2025friendly}.
+\section{Transport Framework and Notation}\label{sec:transport-frame}
 
-\section{Change of Variables in Standardized Coordinates}
+This section establishes the coordinate system, notation, and algebraic identities used throughout the thesis.
+The schematic in Appendix~\ref{ch:appendix} remains valid; here we focus only on the formulas that are needed later.
+We set up the standardized pullback likelihood, the triangularity assumption, and the Jacobian factorization.
 
-The pullback of $\eta$ by $S$ is the central identity that turns transport into a likelihood,
+We write $u_{1:k}=(u_1,\ldots,u_k)$ and use $\partial_{u_k}$ for partial derivatives with respect to $u_k$.
+
+We observe data on the original scale $x \in \mathbb{R}^K$. Training-split statistics define a fixed standardization map
 \begin{equation}
-  \pi_U(u) = \eta\big(S(u)\big)\, \big|\det \nabla_u S(u)\big|.
-  \label{eq:pullback}
+  u \;=\; T_{\mathrm{std}}(x) \;=\; (x-\mu)\oslash\sigma,\qquad \sigma_k>0,
+  \label{eq:transport-standardise}
 \end{equation}
-Because we learn $S$ on standardized inputs $u$, Equation~\eqref{eq:pullback} is the form we optimize and evaluate.
-Reporting on the original scale $x$ applies the diagonal Jacobian of $T_{\mathrm{std}}$: $\log \pi_X(x)=\log \pi_U(u)-\sum_{k=1}^K \log \sigma_k$ with $u=T_{\mathrm{std}}(x)$.
-The reference factorizes as $\eta(z)=\prod_{k=1}^K \phi(z_k)$, so once $S$ is triangular the log pullback expands as a sum over dimensions.
-This linear-in-$K$ structure underpins all evaluation and training in later chapters \citep{ramgraber2025friendly}.
+where $\mu$ and $\sigma$ are the empirical mean and standard deviation estimated on the training split, and $\oslash$ denotes elementwise division.
+Each feature is shifted and rescaled once using training data only.
+All derivatives and Jacobians are then taken in $u$-space, which avoids leakage and keeps results comparable across estimators.
 
-It is useful to record the complementary pushforward statement.
-If $u\sim \pi_U$ and $z=S(u)$, then $S_{\#}\pi_U=\eta$.
-The pair $(S^{\sharp}\eta, S_{\#}\pi_U)$ clarifies that the same map both defines the model density in $u$-space via Equation~\eqref{eq:pullback} and certifies exact sampling by inversion.
-In what follows we always compute derivatives of $S$ with respect to $u$, keep the pullback in standardized coordinates, and only apply the affine correction when moving results back to the original scale.
-
-\section{Triangular Structure and Its Consequences}
-
-A triangular map decomposes component-wise,
+We model $\pi_U$ as the pullback of a simple reference $\eta$ through a triangular map $S:u\mapsto z$.
+The reference is the $K$-variate standard normal $\eta(z)$.
+Applying the change-of-variables formula gives
 \begin{equation}
-  S(u) = \big(S_1(u_1),\,S_2(u_{1:2}),\,\ldots,\,S_K(u_{1:K})\big),
-  \label{eq:triangular-map}
-\end{equation}
-and enforces strict monotonicity in the last argument of each component: $\partial_{u_k}S_k(u_{1:k})>0$ for all feasible $u$. The Jacobian $\nabla_u S(u)$ is lower triangular, hence
-\begin{equation}
-  \det\nabla_u S(u)=\prod_{k=1}^K \partial_{u_k}S_k(u_{1:k}),
-  \qquad
-  \log \big|\det\nabla_u S(u)\big| = \sum_{k=1}^K \log \partial_{u_k}S_k(u_{1:k}).
-  \label{eq:triangular-det}
+  \pi_U(u) \;=\; \eta\!\left(S(u)\right)\,\left|\det\nabla_u S(u)\right|.
+  \label{eq:transport-pullback}
 \end{equation}
 
-Three consequences matter in practice.
-First, likelihoods are efficient: the log determinant in Equation~\eqref{eq:triangular-det} is a sum of one-dimensional terms, so evaluating Equation~\eqref{eq:pullback} scales in $\mathcal{O}(K)$.
-Second, invertibility is exact: strict monotonicity implies global bijectivity, and inversion reduces to sequential one-dimensional root finding; compute $u_1=S_1^{-1}(z_1)$, then $u_2=S_2^{-1}(z_2;u_1)$, and proceed coordinate by coordinate \citep{rosenblatt1952remarks,knothe1957contributions}.
-Third, triangularity yields transparent conditionals: to sample $u_{m:K}\mid u_{1:m-1}=u^{\star}_{1:m-1}$, fix the early coordinates at $u^{\star}$ and invert only the trailing components.
-Equivalently, the map aligns with the chain-rule factorization $\pi_U(u)=\prod_{k=1}^K \pi(u_k\mid u_{1:k-1})$ and exposes each conditional as a one-dimensional transformation \citep{ramgraber2025friendly}.
-
-Existence and anisotropy deserve a brief remark.
-For any ordering of the variables there exists a monotone triangular rearrangement that couples $\pi_U$ and $\eta$ under weak regularity conditions; this is the Knothe--Rosenblatt rearrangement \citep{rosenblatt1952remarks,knothe1957contributions}.
-Different orderings induce different maps.
-While all are valid, their sparsity and approximation difficulty may vary with order.
-When conditional independence makes some inputs irrelevant for a component, omitting those arguments sparsifies $\nabla_u S$, lowers variance, and improves scaling.
-In high dimension, such sparsity often drives both statistical efficiency and computational cost \citep{ramgraber2025friendly}.
-
-Finally, we keep terminology consistent.
-A log density is $\log \hat\pi(\cdot)$ at a point.
-A (test) log likelihood is the dataset average of log densities.
-Its negative is the negative log likelihood (NLL) in nats; lower is better.
-These conventions are used uniformly in Chapters~\ref{ch:methods} and \ref{ch:results}.
-
-\section{Separable and Cross-Term Triangular Transports}
-
-A triangular component is most interpretable once we separate how it depends on the last coordinate $u_k$ versus the context $u_{1:k-1}$. Two parameterizations recur.
-
-\paragraph{Separable triangular maps.} A separable component decomposes into a context-dependent shift and a univariate monotone shape,
+To report log densities on the original scale, we apply only the diagonal affine correction implied by standardization:
 \begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k) > 0,
-  \label{eq:separable-component}
-\end{equation}
-so the log Jacobian depends only on $u_k$:
-\begin{equation}
-  \log \partial_{u_k}S_k(u_{1:k}) = \log h_k'(u_k).
-  \label{eq:separable-jacobian}
-\end{equation}
-The context $u_{1:k-1}$ can move the conditional through $g_k$, but it cannot reshape it.
-Scale, skewness, tail thickness, and modality remain fixed across contexts because the log-Jacobian contribution depends only on $u_k$.
-Separable transports capture nonlinear location shifts across contexts with a context-invariant shape.
-
-\paragraph{Cross-term triangular maps.} A cross-term component uses an integrated rectifier,
-\begin{equation}
-  S_k(u_{1:k})=\int^{u_k}\exp\big(h_k(t,u_{1:k-1})\big)\, \mathrm{d}t + c_k(u_{1:k-1}),
-  \qquad
-  \log \partial_{u_k}S_k(u_{1:k}) = h_k(u_k,u_{1:k-1}),
-  \label{eq:cross-component}
-\end{equation}
-so the log Jacobian, and thus the shape of the conditional, depends on the context.
-This structure captures conditional heteroskedasticity, skew that flips with predictors, or mode splitting that appears only for certain $u_{1:k-1}$.
-
-Two small examples make the contrast concrete.
-Suppose $U_2\mid U_1=u_1$ is normal with mean $m(u_1)$ and variance $\sigma^2(u_1)$.
-A separable map can absorb $m(u_1)$ through $g_2$, but cannot make the slope $\partial_{u_2}S_2$ depend on $u_1$, so it cannot represent the variance change $\sigma^2(u_1)$.
-A cross-term map sets $\log\partial_{u_2}S_2=h_2(u_2,u_1)$ and can encode variance inflation or contraction with $u_1$.
-As a second example, let $U_2\mid U_1$ be unimodal for some $u_1$ and bimodal for others.
-Separable maps preserve modality across contexts; cross terms allow the derivative to bend with $u_1$ and can introduce or remove modes as the context varies.
-Cross-term training requires careful numerical safeguards; details follow in Chapter~\ref{ch:methods}.
-
-\paragraph{Implementation notes.} The integrated rectifier guarantees positivity of $\partial_{u_k}S_k$ and supports rich context dependence, but typically requires one-dimensional quadrature and regularization to remain stable.
-Separable maps often admit linear-in-coefficient parameterizations with convex subproblems for the monotone part; they are fast and robust, but limited to context-invariant conditional shape \citep{ramgraber2025friendly}.
-
-\section{Transformation Forests as Triangular Transport}
-
-Transformation models posit a strictly increasing transformation $h(y\mid w)$ such that $\Phi\big(h(Y\mid W)\big)$ is standard, letting predictors act through $h$ while preserving monotonicity \citep{hothorn2018conditional}.
-A transformation random forest aggregates local transformation models over an adaptive partition of the predictor space to produce a strictly monotone conditional CDF $\widehat F_k(\cdot \mid u_{1:k-1})$ for each coordinate $u_k$ given $u_{1:k-1}$ \citep{hothorn2017transformation,hothorn2021transformation}.
-This induces a triangular transport component via the probability integral transform,
-\begin{equation}
-  S_k(u_{1:k}) = \Phi^{-1}\big(\widehat F_k(u_k\mid u_{1:k-1})\big).
-  \label{eq:trtf-map-theory}
+  \log \pi_X(x) \;=\; \log \pi_U\!\left(T_{\mathrm{std}}(x)\right) - \sum_{k=1}^{K}\log\sigma_k.
+  \label{eq:transport-affine}
 \end{equation}
 
-\paragraph{Per-component likelihood identity.} Differentiating $\Phi\big(S_k(u_{1:k})\big)=\widehat F_k(u_k\mid u_{1:k-1})$ in $u_k$ gives
+The transport is assumed to be lower triangular and strictly increasing in each coordinate,
 \begin{equation}
-  \widehat\pi_k(u_k\mid u_{1:k-1}) = \phi\big(S_k(u_{1:k})\big)\, \partial_{u_k}S_k(u_{1:k}),
-  \label{eq:trtf-likelihood-theory}
+  S(u) \;=\; \big(S_1(u_1), S_2(u_{1:2}), \ldots, S_K(u_{1:K})\big), 
+  \qquad \partial_{u_k}S_k(u_{1:k})>0,
+  \label{eq:transport-triangular}
 \end{equation}
-showing that the TRTF conditional density equals the pullback factor one would obtain by parameterizing $S$ directly.
-Summing over $k$ yields the joint log likelihood from Equation~\eqref{eq:pullback} with the triangular log determinant in Equation~\eqref{eq:triangular-det}.
-TRTF therefore implements the same transport likelihood once standardized \citep{hothorn2017transformation,hothorn2021transformation,ramgraber2025friendly}.
+so the Jacobian $\nabla_u S(u)$ is lower triangular. Its determinant factorizes as
+\begin{equation}
+  \log \big|\det \nabla_u S(u)\big| \;=\; \sum_{k=1}^{K}\log \partial_{u_k}S_k(u_{1:k}).
+  \label{eq:transport-det}
+\end{equation}
 
-\paragraph{Additive predictor implies separability.} Under the common additive predictor implementation (TRTF-AP),
+This factorization makes the \emph{log-determinant} cost $\mathcal{O}(K)$ once the diagonal terms are available.
+Evaluating all $S_k(u_{1:k})$ can be $\mathcal{O}(K^2)$ for a general triangular map, but reduces to $\mathcal{O}(K)$ in the separable parameterization of Section~\ref{sec:transport-separable}.
+
+Strict monotonicity on each coordinate ($\partial_{u_k} S_k>0$) together with triangularity implies that $S$ is a bijection with a triangular inverse, computable by back-substitution.
+
+\begin{table}[t]
+  \centering
+  \caption{Notation for the transport framework used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
+All derivatives and Jacobians are with respect to $u$; log densities on $x$ apply the affine correction in Equation~\eqref{eq:transport-affine}.}
+  \label{tab:transport-notation}
+  \begin{tabular}{ll}
+    \hline
+    Symbol & Meaning \\ 
+    \hline
+    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
+    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ \\ 
+    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates \\ 
+    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
+    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
+    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
+    $\eta(z)$ & $K$-variate standard normal density \\ 
+    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
+    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space \\ 
+    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
+    $K$ & Dimension of the feature vector \\ 
+    \hline
+  \end{tabular}
+\end{table}
+
+Taken together, the triangular and strictly monotone structure turns joint density evaluation
+and sampling into sequential one-dimensional stages, and the standardized coordinates allow us
+to report on the $x$-scale with only the diagonal correction from Equation~\eqref{eq:transport-affine}.
+
+\section{Separable Triangular Maps and Transformation Random Forests}\label{sec:transport-separable}
+
+We now move from the general triangular setup to the first concrete parameterization: the \textbf{separable triangular map}.
+We then show how \textbf{Transformation Random Forests (TRTF)} fit into the same framework.
+The aim is a shared likelihood that we can train, compare, and diagnose across estimators without changing conventions.
+
+\subsection{Separable triangular component}
+
+A separable component has the form
 \begin{equation}
-  \widehat F_k(u_k\mid u_{1:k-1}) = \Phi\big(h_k(u_k)+g_k(u_{1:k-1})\big),
-  \label{eq:trtf-additive}
+  S_k(u_{1:k}) \;=\; g_k(u_{1:k-1}) + h_k(u_k), \qquad h_k'(u_k) > 0,
+  \label{eq:transport-separable}
 \end{equation}
-so
+where earlier coordinates shift location through $g_k$, while $h_k$ shapes the marginal and ensures monotonicity.
+To fix identifiability, we enforce a convention such as centering $g_k$ on the training split, $\mathbb E[g_k(U_{1:k-1})]=0$.
+
+With the standard normal reference, the pullback identity becomes
 \begin{equation}
-  S_k(u_{1:k}) = h_k(u_k)+g_k(u_{1:k-1}),
-  \qquad
-  \partial_{u_k}S_k(u_{1:k}) = h_k'(u_k),
-  \label{eq:trtf-separable}
+  \log \pi_U(u) \;=\; \sum_{k=1}^{K}\Big[\log \varphi\!\big(S_k(u_{1:k})\big) + \log h_k'(u_k)\Big].
+  \label{eq:transport-likelihood}
 \end{equation}
-and the induced transport is separable: $\log \partial_{u_k}S_k$ depends only on $u_k$.
-Monotonicity in $u_k$ is ensured by construction, and the Jacobian term carries no context dependence.
-TRTF-AP therefore excels when conditional shapes are stable and context acts primarily through location; it is structurally limited when shape varies with predictors \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
+Up to a constant, the per-sample negative log-likelihood is
+\begin{equation}
+  \mathcal{L}(u) \;=\; \sum_{k=1}^{K}\Big[\tfrac{1}{2}\,S_k(u_{1:k})^2 - \log h_k'(u_k)\Big].
+  \label{eq:transport-loss}
+\end{equation}
 
-A short note on reporting closes the loop.
-TRTF evaluates conditional densities in standardized coordinates.
-Reported log densities on the original scale apply the affine correction $-\sum_{k=1}^K \log \sigma_k$ that accompanies $T_{\mathrm{std}}$, matching the convention set at the beginning of this chapter.
-With this alignment, comparisons between TRTF, direct triangular maps, and copulas are numerically coherent.
+For a dataset $\{u^{(i)}\}_{i=1}^n$, the empirical objective reads
+\[
+\mathcal L_n = \frac{1}{n}\sum_{i=1}^n \sum_{k=1}^{K}
+\Big[\tfrac{1}{2}\,S_k(u^{(i)}_{1:k})^2 - \log h_k'(u^{(i)}_k)\Big] + \text{const}.
+\]
 
-\section{Takeaways}
+These formulas evaluate in $\mathcal{O}(K)$ time per sample, since the Jacobian term depends only on $u_k$.
+Sampling inverts $S$ by back-substitution, and log densities on $x$-space apply the affine correction in Equation~\eqref{eq:transport-affine}.
+The speed comes with a trade-off: separability fixes the conditional shape along $u_k$ once $g_k$ shifts the location, so variance, skewness, and modality no longer adapt to the context $u_{1:k-1}$.
 
-We learn $S:u\mapsto z$ in standardized coordinates and evaluate $\log \pi_X$ via a simple affine correction.
-Triangularity turns the log determinant into a sum of one-dimensional terms, guarantees exact inversion, and exposes conditionals through back substitution; these properties make likelihoods linear time in dimension and conditional simulation straightforward \citep{rosenblatt1952remarks,knothe1957contributions,ramgraber2025friendly}.
-Separable transports move locations but keep shape fixed across contexts; cross-term transports allow context-dependent shape and capture heteroskedastic or multimodal conditionals at the cost of more delicate optimization.
-TRTF provides a nonparametric route to the same triangular transport with a per-component likelihood identity and the important specialization that TRTF-AP is separable by construction \citep{hothorn2017transformation,hothorn2021transformation,hothorn2018conditional}.
-These foundations justify the objectives and implementations developed in Chapter~\ref{ch:methods} and frame the empirical comparisons that follow.
+\subsection{Transformation Random Forests in the framework}\label{sec:transport-trtf}
+
+Transformation Random Forests (TRTF) fit conditional CDFs and insert them into the triangular pipeline via the probability integral transform.
+Let $\widehat F_k(\cdot \mid u_{1:k-1})$ denote the smoothed conditional CDF.
+Define
+\begin{equation}
+  S_k(u_{1:k}) \;=\; \Phi^{-1}\!\Big(\widehat F_k(u_k \mid u_{1:k-1})\Big).
+  \label{eq:transport-trtf-map}
+\end{equation}
+
+If the forest is parameterized additively as
+\begin{equation}
+  \widehat F_k(u_k \mid u_{1:k-1}) \;=\; \Phi\!\big(h_k(u_k) + g_k(u_{1:k-1})\big),
+  \label{eq:transport-trtf-additive}
+\end{equation}
+then we recover the separable structure,
+\begin{equation}
+  S_k(u_{1:k}) \;=\; h_k(u_k) + g_k(u_{1:k-1}), \qquad h_k'(u_k) > 0.
+  \label{eq:transport-trtf-separable}
+\end{equation}
+
+\paragraph{Derivative check.}
+\[
+\partial_{u_k}S_k
+= \frac{d}{du_k}\Phi^{-1}(\Phi(h_k(u_k)+g_k))
+= h_k'(u_k) > 0,
+\]
+so the Jacobian diagonal is indeed $h_k'(u_k)$ in the additive TRTF case.
+
+\paragraph{Numerical note.}
+In practice, forests may produce $\widehat F_k$ values outside $(0,1)$ by tiny amounts or with imperfect monotonicity.
+We therefore clip to $(\varepsilon,1-\varepsilon)$ with $\varepsilon=10^{-6}$ and, if needed, apply isotonic post-processing to enforce strict increase.
+This keeps $S_k=\Phi^{-1}(\widehat F_k)$ well defined.
+
+With this forest aggregation and strictly monotone conditionals, the resulting TRTF map coincides with the separable parameterization, so it delivers the same likelihood and back-substitution sampler.
+
+\section{Copula Baselines}\label{sec:transport-copula}
+
+We now turn to copulas, used here as diagnostic baselines rather than high-dimensional competitors.
+The focus is on a nonparametric bivariate copula when $K=2$, with an independence fallback otherwise.
+Standardization remains global (based on the full training split), whereas marginal fits for copulas are per-class.
+This ensures comparability on the $x$-scale when reporting.
+
+\subsection{Estimation and reporting}
+
+For each class $y$, we proceed as follows:
+
+\begin{itemize}
+  \item \textbf{Marginals:} Fit univariate kernel densities (\texttt{kde1d}) to obtain $\widehat f_k(\cdot \mid y)$ and CDFs $\widehat F_k(\cdot \mid y)$.
+  \item \textbf{Pseudo-observations:} Transform samples to $(0,1)$ using mid-ranks,
+  \begin{equation}
+    U_{ik}^{(y)} \;=\; \frac{\operatorname{rank}(X_{ik}^{(y)})}{n_y + 1}.
+    \label{eq:copula-midranks}
+  \end{equation}
+  Mid-ranks handle ties; if the data contain many, a tiny jitter is added before ranking.
+  \item \textbf{Dependence:} Fit a bivariate kernel copula density $c_y(u_1,u_2)$ using \texttt{kdecopula::kdecop} with local quadratic smoothing, which uses a transformation approach to avoid boundary bias \citep{nagler2017kdecopula}.
+\end{itemize}
+
+For prediction at $x=(x_1,x_2)$:
+\begin{enumerate}
+  \item Evaluate $\widehat f_k(x_k \mid y)$ and $\widehat F_k(x_k \mid y)$.
+  \item Evaluate $c_y(\widehat F_1(x_1\mid y), \widehat F_2(x_2\mid y))$.
+  \item Combine them through
+  \begin{equation}
+    \log \widehat \pi_X(x \mid Y=y) 
+      \;=\; \log \widehat f_1(x_1 \mid y) + \log \widehat f_2(x_2 \mid y)
+      + \log c_y\!\Big(\widehat F_1(x_1 \mid y), \widehat F_2(x_2 \mid y)\Big).
+    \label{eq:copula-logdensity}
+  \end{equation}
+\end{enumerate}
+When labels are unknown, we mix across classes using priors $\pi(y)$ and a stable log-sum-exp.
+
+\subsection{Fallback and scope}
+
+If $K\neq 2$, labels are absent, or packages unavailable, we fall back to an independence model:
+\begin{equation}
+  \log \widehat \pi_X^{\mathrm{ind}}(x) \;=\; \sum_{k=1}^{K} \log \widehat f_k(x_k).
+  \label{eq:copula-independence}
+\end{equation}
+Hence the NP-Copula is restricted to $K=2$ and is used only as a diagnostic baseline.
+
+\subsection{Consistency with evaluation frame}
+
+Copulas operate directly on the original scale $x$ and therefore bypass the affine correction of Section~\ref{sec:transport-frame}.
+Still, the logic is parallel to the triangular branch: start from a simple reference (independent uniforms), add a dependence correction, and return a single joint log density.
+Using mid-ranks for fitting and $\widehat F_k$ for reporting is consistent with Sklar's theorem \citep{sklar1959fonctions,nelsen2006introduction,joe2014dependence}.
+
+\subsection{Terminology and references}
+
+We use \textit{NP-Copula} for the nonparametric kernel estimator and \textit{Independence} for the fallback KDE product.
+References include Sklar’s theorem and standard copula texts \citep{sklar1959fonctions,nelsen2006introduction,joe2014dependence}, kernel copula methodology \citep{nagler2017kdecopula}, and the transformation-model literature for context \citep{hothorn2021transformation}.
+In practice, the NP-copula baseline for $K=2$ combines per-class marginals with a dependence correction, providing a transparent diagnostic comparator to the triangular transport models.

### 2025-09-30 report/chapter03.Rnw
.

@@
 
-% Checklist: enforce US spelling, clarify Jacobian phrasing, remove empty sections.
-This chapter turns the commitments of Chapters~\ref{ch:intro} and \ref{ch:background} into a practical modeling program.
-Our aim is to express three model families---triangular transport maps (TTM), transformation random forests (TRTF), and copulas---within a common transport framework so that likelihoods, calibration, and computational cost are directly comparable.
-Every method we study standardizes the data, learns a monotone triangular map to a simple reference, and evaluates Jacobians in the standardized space.
-That alignment keeps objectives, diagnostics, and reported log-densities interoperable.
+% Checklist: align terminology, update abbreviations, confirm notation references.
+\begin{quote}
+\textbf{Motivation.} We factor the joint density into simple conditional pieces using a lower-triangular parameterization.
+This yields exact likelihoods, transparent conditionals we can inspect, and $\mathcal{O}(K)$ per-sample cost.
+The shared factorization lets us run ordering experiments without changing the model class.
+\end{quote}
 
-% Checklist: define model abbreviations for concise tables
-\textbf{Model abbreviations.} We follow Table~\ref{tab:model-abbrev} for concise labels (TTM-Marg, TTM-Sep, TRTF, True-Marg/True-Joint, Copula) across tables and figures.
+\textbf{Model abbreviations.} We use the short labels TTM-Sep (separable triangular transport map), TTM-X (separable map with low-rank cross-terms), TRTF (Transformation Random Forests; spelled out on first mention), RealNVP, MAF, NSF, True-Marg/True-Joint, and Copula consistently across tables and figures.
+
+\textbf{Units and signage.} Unless stated otherwise, all log-likelihood (LL) and negative log-likelihood (NLL) values are reported in nats.
+Plots and tables mark $\text{LL}~(\uparrow~\text{better})$ and $\text{NLL}~(\downarrow~\text{better})$ explicitly.
+
+\textbf{Notation recap.}
+\begin{itemize}
+  \item Elementwise standardization with train-split statistics:
+  \[ 
+    u_k = \frac{x_k - \mu_k}{\sigma_k},\quad k = 1,\ldots,K
+    \qquad\text{equivalently } u = \operatorname{diag}(\sigma)^{-1}(x - \mu).
+  \]
+  We compute derivatives and Jacobians in $u$-coordinates and transform back to $x$ only when reporting densities on the original scale via Equation~\eqref{eq:transport-affine}.
+  \item Prefix subvector shorthand: $u_{1:k} := (u_1,\ldots,u_k)$.
+  \item Derivatives use $h'_k(\cdot)$ for univariate derivatives; mixed partials follow the convention from Chapter~\ref{ch:background}.
+\end{itemize}
+
+\textbf{Copula conventions.}
+\begin{itemize}
+  \item All copula fits estimate marginals; reported values are joint densities on the original $x$-scale.
+  \item Half-Moon ($K=2$) uses class-conditional copulas for blue/red labels mixed by empirical class priors when we report the unlabeled density.
+No other model uses labels in this chapter; detailed results appear in Section~\ref{sec:synthetic-results}.
+\end{itemize}
 
@@
 
-% Checklist: preserve dynamic refs, ensure ASCII hyphenation, align with future sections.
-This section fixes data sources, generators, and preprocessing so likelihoods, calibration, and compute remain comparable across models.
-All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the common affine correction.
-We keep a single triangular-map direction $S:u \rightarrow z$ across methods to avoid mixed objectives.
-To avoid symbol collisions, $\sigma$ denotes standardization scales only; logistic gates are written $\operatorname{logistic}(\cdot)$ throughout.
+This section fixes data sources, synthetic generators, and preprocessing so that objectives, diagnostics, and reported log-densities are comparable across models.
+All estimators operate in standardized coordinates, evaluate Jacobians in that space, and report log densities on the original scale using the affine correction from Equation~\eqref{eq:transport-affine}.
+Throughout, logistic gates are written $\operatorname{logistic}(\cdot)$.
 
-We standardize features with training-split statistics only.
-Equation~\eqref{eq:transport-standardise} defines $u = T_{\mathrm{std}}(x) = (x-\mu)\oslash \sigma$ with $\sigma_k > 0$.
-We evaluate $\log \pi_U(u)$ through the pullback identity in Equation~\eqref{eq:transport-pullback}, apply the triangular factorization from Equation~\eqref{eq:transport-det}, then convert to $\log \pi_X(x)$ using the diagonal correction in Equation~\eqref{eq:transport-affine}.
-We report average test negative log-likelihoods (NLL) in nats.
-Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} shows the standardized pipeline shared by transport maps, Transformation Random Forests, and copulas.
+\subsection{Standardization and Splits}
+Features are standardized with training-split statistics only $(\mu, \sigma \in \mathbb{R}^K, \sigma_k > 0)$ as in Equation~\eqref{eq:transport-standardise}.
+The triangular pullback, determinant factorization, and affine correction from Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-affine} stay in $u$-coordinates; we transform back to $x$ once per evaluation when reporting densities.
+We keep fixed train/test splits.
+Synthetic calibration uses an $80/20$ train/test partition.
+Real-data splits follow the published convention for each benchmark, and compute profiles are summarized in Appendix~\ref{app:evaluation}.
 
-We use fixed train, validation, and test splits with proportions $0.60/0.20/0.20$ unless a benchmark provides official splits.
-Synthetic studies report results for $n \in \{25, 50, 100, 250\}$ and use $n=250$ for headline tables and figures; for real-data benchmarks we use $N$ to denote the training budget.
-The canonical four-dimensional ordering is $(1,2,3,4)$.
-Robustness to ordering is assessed by averaging over all $4!
-= 24$ permutations.
-We adopt the natural column order for real datasets.
-We fix seeds $\{11, 13, 17, 19, 23\}$ for data generation and model fitting, and we average repeated runs with standard errors to quantify stochastic variability.
-Figure~\ref{fig:autoregressive-joint-calibration} displays the $20\%$ test split for the synthetic calibration study.
+For synthetic runs we hold seeds $\{11, 13, 17, 19, 23\}$ fixed across models and report means with $\pm 2$ standard errors when averaging replicates.
+Ordering experiments reuse the lower-triangular factorization and vary only the coordinate order; the permutation sweeps in the four-dimensional study currently pin $\texttt{SEED}=42$ unless overridden so that every ordering shares the same draws.
+Unless noted otherwise, our canonical ordering remains $(1,2,3,4)$ for the four-dimensional generator; Section~\ref{sec:synthetic-results} reports the permutation studies.
 
-The Half-Moon dataset provides a curved, bimodal joint in $K=2$.
-We draw a class $Y \sim \mathrm{Bernoulli}(0.5)$, an angle $\Theta \sim \mathrm{Unif}[0,\pi]$, and additive noise $\varepsilon \sim \mathcal{N}(0, \sigma^2 I_2)$ with $\sigma = 0.10$.
-For $Y = 0$ we set $m(\Theta) = (\cos \Theta, \sin \Theta)$.
-For $Y = 1$ we set $m(\Theta) = (1 - \cos \Theta, -\sin \Theta + 0.5)$.
-The observed $X = m(\Theta) + \varepsilon$.
-The ``True joint'' oracle evaluates the mixture density by numerical quadrature over $\Theta$ with the known Gaussian noise, and the ``True conditional'' oracle conditions on $Y$.
-Figure~\ref{fig:halfmoon-panels} (p.~\pageref{fig:halfmoon-panels}) shows representative contour plots at $n=250$, which align with this generator.
-Table~\ref{tab:halfmoon-nll} (p.~\pageref{tab:halfmoon-nll}) reports the corresponding NLLs.
+\subsection{Half-Moon ($K=2$)}
+Half-Moon is a curved, bimodal joint designed to surface dependence beyond linear correlation.
+We evaluate joint log-density contours per model and compare results to the true joint.
+Figure~\ref{fig:halfmoon-panels} and Table~\ref{tab:halfmoon-nll} later in this chapter instantiate the diagnostics.
+For the copula baseline we fit blue/red class-conditional copulas and mix them by empirical class priors when reporting the unlabeled density.
+No other estimator uses labels.
 
-The four-dimensional autoregressive generator combines Gaussian, exponential, beta, and gamma components to induce heteroskedasticity, skew, and conditional multimodality.
-The first coordinate is $X_1 \sim \mathcal{N}(0,1)$.
-The second coordinate is independent $X_2 \sim \mathrm{Exp}(\lambda_0)$ with rate $\lambda_0 = 1$.
-The third coordinate lies on $(0,1)$ and is a context-gated mixture of two beta laws, $X_3 \mid X_{1:2} \sim w\,\mathrm{Beta}(\alpha_1, \beta_1) + (1 - w)\,\mathrm{Beta}(\alpha_2, \beta_2)$.
-We set $(\alpha_1, \beta_1) = (2.5, 5.0)$ and $(\alpha_2, \beta_2) = (5.0, 2.5)$.
-The mixing weight is $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2 - 1))$ with $\operatorname{logistic}(\cdot)$ the logistic function and $(\gamma_0, \gamma_1, \gamma_2) = (0, 1.5, 1.0)$.
-The fourth coordinate is positive and conditionally heteroskedastic, $X_4 \mid X_{1:3} \sim \tilde{w}\,\mathrm{Gamma}(k_1, r_1(X_2)) + (1 - \tilde{w})\,\mathrm{Gamma}(k_2, r_2(X_2))$.
-We set shapes $(k_1, k_2) = (3, 6)$, rates $r_1(X_2) = 1 + 0.5 X_2$ and $r_2(X_2) = 0.75 + 0.25 X_2$, and gate $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3 - 0.5))$ with $(\delta_0, \delta_1, \delta_3) = (0, 1.0, 3.0)$.
-The ``True joint'' baseline uses these closed-form conditionals to evaluate the exact joint density, while the ``True marginal'' baseline uses the corresponding univariate marginals and ignores dependence.
+\subsection{Four-dimensional Autoregressive Generator ($K=4$)}
+The synthetic autoregressive generator matches the lower-triangular factorization while stressing skewness, heavy tails, and context-dependent shape:
+\begin{itemize}
+  \item $X_1 \sim \mathcal{N}(0,1)$.
+  \item $X_2 \sim \mathrm{Exponential}(\lambda_0)$ with $\lambda_0 = 1$.
+  \item $X_3 \mid X_{1:2}$ is a mixture of two beta laws on $(0,1)$, e.g. $\mathrm{Beta}(2.5, 5.0)$ versus $\mathrm{Beta}(5.0, 2.5)$, with gate
+  \[ 
+    w_3(X_{1:2}) = \operatorname{logistic}\big(\gamma_0 + \gamma_1 X_1 + \gamma_2 (X_2 - 1)\big), \qquad \gamma = (\gamma_0, \gamma_1, \gamma_2).
+  \]
+  \item $X_4 \mid X_{1:3}$ is a mixture of two gamma laws on $(0, \infty)$ with distinct shape/scale pairs, gated by
+  \[ 
+    w_4(X_{1:3}) = \operatorname{logistic}\big(\delta_0 + \delta_1 X_1 + \delta_3 (X_3 - 0.5)\big), \qquad \delta = (\delta_0, \delta_1, \delta_3).
+  \]
+\end{itemize}
+Mixture weights are normalized through the usual softmax over component logits so that Jacobian terms stay well behaved under shared standardization.
+This design keeps $X_4$ conditionally heteroskedastic with positive support and aligns with the triangular evaluation cost $\mathcal{O}(K)$.
 
-% Checklist: summarize generator config, define softmax usage.
-\begin{table}[t]
-  \centering
-  \caption{Configuration for the four-dimensional autoregressive generator used in the synthetic study.
-The beta and gamma coordinates are two-component mixtures with logistic gates; fixed parameters and gates match the prose above.}
-  \label{tab:autoregressive-config}
-  \begin{tabular}{lll}
-    \hline
-    Coordinate & Distribution & Parameters / gate \\
-    \hline
-    $X_1$ & Normal & $\mathcal{N}(0,1)$ \\
-    $X_2$ & Exponential & $\mathrm{rate} = \lambda_0 = 1$ \\
-    $X_3 \mid X_{1:2}$ & Mixture Beta & $\mathrm{Beta}(2.5,5.0)$ / $\mathrm{Beta}(5.0,2.5)$; $w = \operatorname{logistic}(\gamma_0 + \gamma_1 X_1 + \gamma_2(X_2-1))$, $\gamma=(0,1.5,1.0)$ \\
-    $X_4 \mid X_{1:3}$ & Mixture Gamma & $\mathrm{Gamma}(k_1{=}3, r_1{=}1+0.5X_2)$ / $\mathrm{Gamma}(k_2{=}6, r_2{=}0.75+0.25X_2)$; $\tilde{w} = \operatorname{logistic}(\delta_0 + \delta_1 X_1 + \delta_3(X_3-0.5))$, $\delta=(0,1.0,3.0)$ \\
-    \hline
-  \end{tabular}
-\end{table}
+\paragraph{Why these generators?}
+They match the triangular modeling assumptions yet stress estimators with skewness, heavy tails, and context-dependent variation.
+The true joint density and conditionals are known, which enables direct comparisons without confounding from Monte Carlo error.
 
-Mixture weights use the logistic gate $\operatorname{logistic}(a)=\exp(a)/(1+\exp(a))$, which coincides with the two-component softmax and therefore keeps probabilities in $(0,1)$ that sum to one.
-For completeness, the general softmax takes a vector $a$ and returns $\mathrm{softmax}(a)_i = \exp(a_i) / \sum_j \exp(a_j)$.
-This normalization is essential for the beta and gamma mixtures because it translates linear predictors into valid probability weights while preserving differentiability.
+\paragraph{Reproducibility note.}
+All random seeds, splits, and generator parameters are fixed in the code release.
+Negative NLL values appear for bounded or highly concentrated marginals (e.g., beta or gamma mass near zero); this behavior is expected and does not signal an error.
 
-Table~\ref{tab:autoregressive-config} (p.~\pageref{tab:autoregressive-config}) summarizes the mixture families, gates, and fixed parameters by dimension.
-Tables~\ref{tab:autoregressive-perm} (p.~\pageref{tab:autoregressive-perm}) and~\ref{tab:autoregressive-perm-avg} (p.~\pageref{tab:autoregressive-perm-avg}) then summarize the permutation and sample-size studies used later in this chapter.
-
-The MiniBooNE benchmark follows the published preprocessing to ensure comparability with flow-based baselines.
-We remove 11 outliers with value $-1000$, drop seven features with extreme mass at a single value, and retain $K = 43$ attributes.
-We use the fixed train, validation, and test splits from the benchmark, apply train-only standardization, and avoid any extra pruning of correlated features.
-We report all log-likelihoods in nats and retain the published naming for flow comparators in later tables.
-Section~\ref{sec:realdata} records these steps and provides the dataset context.
-Table~\ref{tab:uci-loglik} reproduces the flow baselines that motivate our TRTF runs.
-
-Additional UCI datasets appear only when we retain them for real-data context.
-POWER keeps household electricity attributes after jittering the minute-of-day encoding, dropping the calendar date and reactive-power column, and adding uniform noise to break ties.
-GAS keeps the \texttt{ethylene\_CO} subset, treats the series as i.i.d., removes strongly correlated attributes, and retains an eight-dimensional representation.
-HEPMASS keeps only the positive class from the ``1000'' split and discards five variables with repeated values to avoid density spikes.
-These preprocessing steps follow the same train-only standardization and reporting conventions described above.
-Section~\ref{sec:realdata} provides the corresponding background and positions these datasets within our evaluation.
-
-All models use the same standardized frame and direction for evaluation, which keeps objectives, diagnostics, and reported quantities interoperable across triangular transports, TRTF, and copula baselines.
-This alignment is necessary for the conditional decompositions, probability integral transform (PIT) calibration checks, and compute summaries presented later in this chapter.
+\bigskip
+\noindent\emph{Next: Section~\ref{sec:models-implementation} details the estimators and training settings that make these likelihoods and diagnostics comparable across methods.}
 
@@
 % Checklist: align references, enforce US spelling, ensure active voice.
-This section specifies the estimators and implementation details that keep likelihoods, calibration, and compute directly comparable across models.
-All estimators share the transport direction $S:u \rightarrow z$, operate in standardized coordinates, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.
+\begin{quote}
+\textbf{Scope.} This section specifies the estimators and the implementation details that keep our likelihoods, diagnostics, and plots comparable.
+We reuse the preprocessing and transport notation in Table~\ref{tab:transport-notation}, draw background from Chapter~\ref{ch:background}, follow the data handling in Section~\ref{sec:datasets-preprocessing}, and lean on the auxiliary pseudo-code in Appendix~\ref{ch:appendix}.
+\end{quote}
 
-We implement separable lower-triangular transport maps denoted TTM-Sep. Component $k$ decomposes into a context shift and a univariate monotone shape,
+\subsection{Separable Triangular Transport Maps (TTM-Sep)}
+
+We use lower-triangular, separable transport maps
 \begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k), \qquad \partial_{u_k} S_k(u_{1:k}) = h_k'(u_k) > 0,
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k), \qquad \partial_{u_k} S_k(u_{1:k}) = h'_k(u_k) > 0,
   \label{eq:ttm-separable-def}
 \end{equation}
-so the Jacobian contribution depends only on $u_k$. The structure yields linear per-sample complexity in $K$ and exact inversion by back-substitution.
+so the Jacobian contribution at stage $k$ depends only on $u_k$. This structure yields $\mathcal{O}(K)$ per-sample cost and exact inversion by back-substitution.
 
-We minimize the Gaussian pullback objective induced by the shared reference,
+\textbf{Objective.} We minimize the Gaussian pullback induced by the shared reference (Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-det}):
 \begin{equation}
-  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h_k'(u_k) \Big],
+  \mathcal{L}(u) = \sum_{k=1}^K \Big[ \tfrac{1}{2} S_k(u_{1:k})^2 - \log h'_k(u_k) \Big].
   \label{eq:ttm-separable-loss}
 \end{equation}
-which follows from the change-of-variables identity in Equation~\eqref{eq:transport-pullback} combined with the triangular determinant factorization in Equation~\eqref{eq:transport-det}.
-The quadratic term pulls the transformed coordinates toward the reference, and the log-derivative term prevents degenerate solutions.
-We solve the regularized problem with bound-constrained optimization and enforce monotone structure by construction.
+We solve this objective with constrained optimization and enforce monotonicity by construction.
 
-We construct $h_k$ with monotone one-dimensional bases that combine identity, integrated sigmoids, softplus-like edge terms, and integrated radial basis functions.
-Nonnegativity constraints on the derivative coefficients guarantee $h_k'(u_k) \ge 0$.
-We linearize tails to stabilize likelihoods as $|u_k|$ grows.
-Ridge penalties apply to all basis coefficients, and optional sparsity penalties shrink context shifts when multicollinearity inflates variance.
-During training and evaluation we clip log-derivatives to $[-H, H]$ to avoid numerical overflow in the Jacobian sum; the bound $H$ is tuned on the validation split.
+\textbf{Parameterization.}
+\begin{itemize}
+  \item $h_k$: one-dimensional monotone bases (integrated radial basis functions, splines, and linear tails) with derivative nonnegativity so that $h'_k > 0$; all coefficients are learned jointly under the positivity constraint enforced by the optimizer.
+  \item $g_k$: low-degree polynomial features of $u_{1:k-1}$ with light ridge regularization to keep $\nabla_u S(u)$ stable.
+\end{itemize}
 
-We build $g_k$ from low-degree polynomial features of $u_{1:k-1}$ and drop predecessors whose inclusion does not improve validation likelihood.
-This pruning keeps $\nabla_u S(u)$ sparse and improves stability in small-sample regimes.
-Ordering matters for finite bases, so headline results use the natural variable order while robustness studies vary the order as described in Section~\ref{sec:datasets-preprocessing}.
-When heuristics are applied, we evaluate two candidates on the validation split—identity and a Cholesky-pivoted ordering with optional Gaussianization—and select the ordering with the lower validation NLL.
-Appendix~\ref{ch:appendix} records how the ordering is stored and reapplied at prediction time.
+\paragraph{Ordering.} Feature ordering matters for triangular models.
+We store the learned order and reuse it at prediction time.
+For permutation studies (Section~\ref{sec:synthetic-results}) we apply the same bookkeeping so that permutations are reproducible.
 
-We reference a cross-term variant, denoted TTM-X, only to delimit scope. The variant augments the separable component with low-rank interactions,
+\subsection{Cross-term Variant (TTM-X)}
+
+We reference a low-rank interaction variant only to delimit the class used in the synthetic analyses:
 \begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k) + \sum_{j<k} \alpha_{kj}\,q_j(u_j)\,r_k(u_k),
+  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k) + \sum_{j<k} \alpha_{kj} \, q_j(u_j) \, r_k(u_k),
   \label{eq:ttm-cross}
 \end{equation}
-where $q_j$ and $r_k$ are monotone features and constraints ensure $\partial_{u_k} S_k(u_{1:k}) > 0$.
-We exclude TTM-X from headline tables because the interactions alter identifiability and complicate calibration.
-The definition clarifies the naming used in the synthetic analyses.
+where $q_j$ and $r_k$ are monotone features.
+Constraints ensure $\partial_{u_k} S_k > 0$ despite the cross-term.
+We do not promote TTM-X as a baseline; the definition clarifies the naming and scope used in the synthetic results.
 
-We implement Transformation Random Forests with additive predictors and denote the model TRTF.
-Implementations rely on the \texttt{partykit} toolkit for recursive partitioning \citep{partykit2015} together with the transformation-model framework \citep{hothorn2017transformation,hothorn2018conditional}.
-Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the strictly increasing conditional CDF returned by the forest.
-The induced triangular component is
+\subsection{Transformation Random Forests (TRTF)}
+
+We use Transformation Random Forests (TRTF) as an alternative route to a triangular component from conditional CDFs \citep{hothorn2017transformation,hothorn2018conditional}.
+Let $\widehat{F}_k(\cdot \mid u_{1:k-1})$ denote the forest CDF.
+We define
 \begin{equation}
@@
 \end{equation}
-and differentiation yields $\varphi\big(S_k(u_{1:k})\big)\,\partial_{u_k} S_k(u_{1:k}) = \widehat{\pi}_k(u_k \mid u_{1:k-1})$.
-Under the additive predictor $\widehat{F}_k(u_k \mid u_{1:k-1}) = \Phi\big(h_k(u_k) + g_k(u_{1:k-1})\big)$ we obtain $S_k = h_k + g_k$ and $\partial_{u_k} S_k = h_k'(u_k)$, which matches Equation~\eqref{eq:ttm-separable-def} exactly in the transport frame.
-Consequently TRTF shares the likelihood in Equation~\eqref{eq:ttm-separable-loss}, inherits exact inversion, and differs operationally through forest training and aggregation.
+which maps $u_k$ into the standard normal reference conditional on $u_{1:k-1}$. By the chain rule,
+\[
+  \varphi\big(S_k(u_{1:k})\big)\, \partial_{u_k} S_k(u_{1:k}) = \widehat{f}_k(u_k \mid u_{1:k-1}),
+\]
+so the TRTF-induced Jacobian matches the forest’s conditional density. The triangular structure mirrors TTM-Sep; the difference is operational (forest training and aggregation) rather than conceptual.
+
+\subsection{Copula Baseline with Estimated Marginals}
 
 We keep copulas as dependence baselines with explicit scope.
-We fit only low-dimensional nonparametric copulas for $K \le 3$, using probit-transformed pseudo-observations and kernel density estimation on the Gaussian scale before mapping back to the unit cube with the appropriate Jacobian.
-The independence baseline evaluates the product of fitted marginals.
-We omit a Gaussian copula from the main experiments to preserve consistency with the low-$K$ nonparametric dependence analyzed in Section~\ref{sec:realdata}.
+Marginals are estimated on the training split, and we report the joint density on the original $x$-scale.
+Concretely, we fit $\widehat{F}_k$ and $\widehat{f}_k$ for each coordinate, transform to pseudo-observations $U_k = \widehat{F}_k(X_k)$, and, for Gaussian copulas, set $Z_k = \Phi^{-1}(U_k)$ to model dependence in a latent normal space.
+We evaluate the copula density $c(U)$ and multiply by $\prod_{k=1}^K \widehat{f}_k(x_k)$ to obtain the reported joint density.
 
-We adopt a single inversion and evaluation convention across estimators.
-Training, Jacobians, and conditional evaluations occur in standardized coordinates.
-Sampling draws $z \sim \mathcal{N}(0, I)$, applies $S^{-1}$ by back-substitution, and converts to the original scale with the stored affine parameters.
-This convention prevents mixed objectives and keeps all reported quantities interoperable.
+\paragraph{Label use (Half-Moon only).} For $K=2$ in the Half-Moon study we fit class-conditional copulas for the blue/red classes and mix them by empirical class priors when reporting the unlabeled joint.
+No other model uses labels.
+Details and diagnostics appear in Section~\ref{sec:synthetic-results}.
 
-We ensure reproducibility and comparability with fixed seeds, cached standardization parameters, and shared reporting utilities.
-Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction, nonparametric copulas, marginal and separable triangular maps, and the shared transport core that implements ordering, bases, derivatives, and evaluation.
-The appendix also records optimizer choices, timing hooks, and object layouts used in the experiments.
+\subsection{Shared Inversion and Evaluation}
 
+We adopt a single inversion and evaluation convention across estimators using Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-det}.
+The alignment is necessary for conditional decomposition checks in Section~\ref{sec:synthetic-results} and for the compute summaries in Section~\ref{sec:realdata}.
+
+\paragraph{Reproducibility.} We fix random seeds, cache splits, and expose the same object layouts for transport maps, TRTF, and copulas.
+Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction; we reuse the same logging, timing hooks, and file formats across models.
 \section{Evaluation Metrics and Protocol}\label{sec:evaluation-protocol}
@@
 % Checklist: align references, keep active voice, reuse existing notation.
-This section defines the metrics and procedures applied across all models so likelihoods, calibration, and compute remain directly comparable.
-We evaluate every estimator in standardized coordinates, apply the triangular determinant, and report log densities on the original scale using the affine correction from Chapter~\ref{ch:background}.
-Figure~\ref{fig:transport-schematic} in Appendix~\ref{ch:appendix} and Table~\ref{tab:transport-notation} summarize the shared pipeline and notation.
+\begin{quote}
+\textbf{Goal.} Define the metrics and the evaluation routine applied across all estimators.
+We reuse the notation and preprocessing conventions from Table~\ref{tab:transport-notation}, with background in Chapter~\ref{ch:background} and dataset preparation in Section~\ref{sec:datasets-preprocessing}.
+All log-likelihood (LL) and negative log-likelihood (NLL) values are reported in nats; figures and tables state $\text{LL}~(\uparrow~\text{better})$ and $\text{NLL}~(\downarrow~\text{better})$ explicitly.
+\end{quote}
 
-We distinguish pointwise log density from dataset averages.
-The test log likelihood (LL) is the mean of $\log \hat{\pi}_X(x)$ over the test split, and the test negative log likelihood (NLL) is its negative.
-Tables note “LL (higher is better)” or “NLL (lower is better)” to avoid ambiguity.
-Reported log densities on the original scale equal the standardized quantity minus $\sum_k \log \sigma_k$ as given by Equation~\eqref{eq:transport-affine}.
-Consequently, datasets with larger training scales introduce large constant offsets that the affine correction removes.
+\subsection{Joint and Conditional Decomposition}
 
-Triangular models exploit the separable pullback in standardized coordinates. With $u = T_{\mathrm{std}}(x)$, the log density decomposes as
+Triangular models evaluate densities in standardized coordinates and then report on the original scale.
+With $u = T_{\mathrm{std}}(x)$ determined by the training-split $(\mu, \sigma)$, any estimator with lower-triangular $S$ satisfies
 \begin{equation}
@@
 \end{equation}
-so the determinant factorization in Equation~\eqref{eq:transport-det} yields linear per-sample cost in $K$.
-In plain language, the model checks how Gaussian each transformed coordinate looks, then adds the exact log-Jacobian contribution from its one-dimensional derivative.
-The affine correction in Equation~\eqref{eq:transport-affine} converts $\log \hat{\pi}_U$ to $\log \hat{\pi}_X$ for reporting.
+so the determinant factorization from Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-det} applies componentwise.
+We transform back to $x$ once per evaluation via the affine correction in Equation~\eqref{eq:transport-affine}.
 
-We report per-dimension conditional NLLs for triangular models to localize error. For each coordinate,
+We localize error with per-dimension conditional NLLs. For coordinate $k$,
 \begin{equation}
@@
 \end{equation}
-and the joint NLL equals $\sum_{k=1}^{K} \mathrm{NLL}_k$ by construction.
-Copulas lack a unique triangular factorization, so we report only their joint NLL.
-Negative per-dimension NLL values can occur because valid densities may exceed one on subdomains.
-These conventions align with Equations~\eqref{eq:transport-trtf-likelihood} and~\eqref{eq:transport-trtf-separable}, which link separable transports and Transformation Random Forests inside the common frame under the additive-predictor and monotone-smoothing assumptions stated in Section~\ref{sec:transport-trtf}.
+and the joint NLL satisfies $\mathrm{NLL}_{\mathrm{joint}} = \sum_{k=1}^{K} \mathrm{NLL}_k$ by the triangular factorization.
+Negative per-dimension NLLs can occur on bounded or highly concentrated supports (e.g., beta or gamma margins); we revisit this behaviour in Section~\ref{sec:synthetic-results}.
 
-Calibration assesses whether predictive probabilities align with empirical frequencies.
-For triangular models we form conditional probability integral transform (PIT) values $V_{ik} = \widehat{F}_k(u_{ik} \mid u_{i,1:k-1})$ on the test split and expect independent $\mathrm{Unif}(0,1)$ draws under correct calibration \citep{gneiting2007probabilistic}.
-We summarize departures from uniformity with the Kolmogorov--Smirnov statistic $D_n = \sup_t \lvert \widehat{F}_n(t) - t \rvert$ and report the associated $p$-value \citep{massey1951kolmogorov}.
-We complement the scalar summary with brief PIT distribution descriptions when patterns recur across seeds.
-For copulas we assess marginal PITs and low-dimensional slices where dependence is transparent.
-Systematic U-shaped or inverted-U PIT indicates under- or over-dispersion and motivates richer parameterizations.
+\subsection{Compute Metrics}
 
-Compute metrics quantify practical cost alongside fit.
-We record wall-clock training time on the training split and per-sample evaluation time on the test split.
-Triangular transports scale linearly in $K$ and approximately linearly in the number of basis functions.
-Transformation Random Forests scale with the number and depth of trees per conditional during training, while prediction remains linear after aggregation.
-Copula training is dominated by correlation estimation or kernel density fitting, followed by fast evaluation.
-We also track peak resident memory when caching affects runtime.
-All timings use the deterministic pipeline defined in Chapter~\ref{ch:dataanalysis} and are averaged over seeds with standard errors.
+We report practical costs alongside fit:
+\begin{itemize}
+  \item \textbf{Train time per epoch} and total wall-clock to the chosen early-stopping criterion when applicable.
+  \item \textbf{Evaluation throughput} (test points per second) together with per-sample latency.
+  \item \textbf{Model size,} recorded as effective parameter counts when the backend exposes them.
+\end{itemize}
+Compute summaries accompany the real-data tables in Section~\ref{sec:realdata}; extended profiles appear in Appendix~\ref{ch:appendix} when needed.
 
-Protocol choices keep comparisons stable and reproducible:
+\subsection{Protocol (Fixed Across Methods)}
+
 \begin{enumerate}
-  \item Standardize with training-split statistics, fit a single map $S:u \rightarrow z$, and evaluate Jacobians in standardized space.
-  \item Compute LL, NLL, and conditional decompositions in standardized coordinates, then apply the affine correction once for reporting.
-  \item Evaluate PIT diagnostics, Kolmogorov--Smirnov statistics, and compute metrics on the fixed test split with seeds $\{11, 13, 17, 19, 23\}$ and quote means with $\pm$ two standard errors across seeds (SE $= s/\sqrt{m}$ over $m$ seeds).
+  \item \textbf{Preprocessing.} Standardize features with training-split $(\mu, \sigma)$, evaluate derivatives and Jacobians in $u$, and convert to $x$ with Equation~\eqref{eq:transport-affine} once per evaluation.
+  \item \textbf{Evaluation in standardized space.} Compute LL, NLL, and conditional decompositions with Equation~\eqref{eq:evaluation-triangular}; map results to the original scale for reporting.
+  \item \textbf{Aggregation and uncertainty.} Average metrics across seeds and report $\pm$ two standard errors (SE $= s/\sqrt{m}$ over $m$ seeds).
+  \item \textbf{Reproducibility hooks.} Cache splits and random seeds, and log objective values together with compute metrics using shared file schemas across estimators.
 \end{enumerate}
-Appendix~\ref{ch:appendix} lists routine interfaces that support exact re-execution; figure captions and table notes repeat units, splits, and the “higher/lower is better” convention for clarity.
 
-Numerical safeguards prevent unstable likelihoods from dominating summaries.
-We enforce strictly monotone derivatives by construction and clip log-derivative contributions to $[-H, H]$ during training and evaluation.
-We tune $H$ and regularization on the validation split and reuse the selected bound on the test split.
-We report clipping status inline per study and keep the exact bound values with the corresponding experiment logs to avoid duplication.
-This practice controls overflow in the Jacobian sum without masking systematic misfit that PIT diagnostics would reveal.
-
-Scope limits clarify non-goals.
-We do not report AIC or BIC because effective parameter counts are not comparable across estimators in this frame.
-We also do not adjust $p$-values for multiple PIT checks; instead, we treat Kolmogorov--Smirnov results as diagnostics and corroborate them with effect sizes and plots.
-These limits keep the evaluation focused on likelihood, calibration, and compute under a single, transparent protocol.
-
+These choices keep likelihood reporting in standardized coordinates, align diagnostics, and make compute measurements comparable.
+Appendix~\ref{ch:appendix} lists the supporting routine interfaces and object layouts used in the experiments.
 \section{Synthetic Results and Diagnostics}\label{sec:synthetic-results}
@@
 Section~\ref{sec:realdata} turns to real-data benchmarks and compute summaries under the same protocol.
-
-\paragraph{Calibration numbers.} To complement the visual diagnostics, Table~\ref{tab:ks-synth} reports median Kolmogorov--Smirnov (KS) distances of probability-integral-transform (PIT) values per coordinate, aggregated across seeds ($\pm$ 2SE).
-Lower is better.
-We include entries for methods with an accessible marginal CDF in our implementation.
-
-<<ks_synth, results='asis', echo=FALSE>>=
-fmt <- function(m, se) sprintf("%.3f $\\pm$ %.3f", as.numeric(m), 2 * as.numeric(se))
-read_sum <- function(path) if (file.exists(path)) utils::read.csv(path, stringsAsFactors = FALSE) else NULL
-hm <- read_sum(file.path('..','code','experiments','halfmoon','results','ks_halfmoon_summary.csv'))
-d4 <- read_sum(file.path('..','code','experiments','4d','results','ks_4d_summary.csv'))
-models <- c('True (Joint)','True (marginal)','Random Forest','Marginal Map','Separable Map','Copula NP')
-val <- function(df, model) {
-  if (is.null(df)) return('--')
-  row <- df[df$model == model, , drop = FALSE]
-  if (!nrow(row)) return('--')
-  fmt(row$ks_median_mean, row$ks_median_se)
-}
-make_row <- function(name, df) paste(c(name, vapply(models, function(m) val(df, m), character(1))), collapse = ' & ')
-cat("\\begin{table}[htbp]\n  \\centering\n  \\textit{(median KS of PIT per coordinate; lower is better).}\\\n\n")
-cat("  \\begin{tabular}{lrrrrrr}\\\n    \\hline\\\n    Dataset & True (Joint) & True (marginal) & Random Forest & Marginal Map & Separable Map & Copula NP \\\\\\n     \\hline\\\n")
-cat("    ", make_row('Half-Moon', hm), " \\\\\n", sep = '')
-cat("    ", make_row('4D generator', d4), " \\\\\n", sep = '')
-cat("    \\hline\\\n  \\end{tabular}\n  \\caption{Calibration via PIT--KS on synthetic datasets: median KS distance per coordinate (mean $\\pm$ 2SE across seeds).
-Entries marked `--' indicate that the CDF was not available in the corresponding backend.}\n  \\label{tab:ks-synth}\n\\end{table}\n")
-@
 

### 2025-09-30 report/chapterA1.Rnw
.

--- 2025-09-20 report/chapterA1.Rnw
@@
 
-Maybe some R code here, probably a \rr{sessionInfo()}
+\section{Pseudo-code Summaries for Model Routines}
+
+This appendix records consolidated pseudo-code for the core R implementations used in the experiments.
+Each summary captures inputs, main processing stages, and outputs so the execution flow is transparent without consulting the source code files.
+
+\subsection{Transformation Random Forest (TRTF)}\label{app:trtf}
+
+\textbf{Routine:}\quad\texttt{fit\_TRTF(S, config, seed, cores)} (calls \texttt{mytrtf}).
+
+\begin{enumerate}
+  \item Validate that the training matrix is numeric, set the RNG seed, and label columns as $X_1,\ldots,X_K$.
+  \item Fit an intercept-only transformation model \texttt{BoxCox} for each $X_k$ to provide baseline monotone transformations.
+  \item For $k = 2,\ldots,K$:
+    \begin{enumerate}
+      \item Build the formula $X_k \sim X_1 + \cdots + X_{k-1}$.
+      \item Choose \texttt{mtry = max(1, floor((k-1)/2))} and standard \texttt{ctree} controls (\texttt{minsplit}, \texttt{minbucket}, \texttt{maxdepth}).
+      \item Fit a transformation forest with \texttt{traforest} and store the conditional model (one forest per $k$).
+    \end{enumerate}
+  \item Return a \texttt{mytrtf} object containing baseline transformations, conditional forests, variable-importance scores, and the seed.
+
+  \item \textbf{Prediction (\texttt{predict.mytrtf}):}
+    \begin{enumerate}
+      \item Convert new data to the same column naming scheme and evaluate $X_1$ through its baseline transformation model to obtain marginal log densities.
+      \item For each conditional forest ($k\geq 2$) evaluate the log density of $X_k$ given $X_{1:(k-1)}$, extracting the diagonal when the forest returns a log density matrix.
+      \item Stack the per-dimension log densities (\texttt{logdensity\_by\_dim}) or sum them to obtain the joint log likelihood (\texttt{logdensity}).
+    \end{enumerate}
+\end{enumerate}
+
+\subsection{Nonparametric Copula Baseline}\label{app:copula}
+
+\textbf{Routine:}\quad\texttt{fit\_copula\_np(S, seed)}.
+
+\begin{enumerate}
+  \item Inspect the training matrix and optional class labels; detect whether the dedicated copula packages are available.
+  \item If prerequisites fail (dimension $K \neq 2$ or labels missing), fall back to independent univariate kernel density estimates per dimension and store them for later interpolation.
+  \item Otherwise, for each class label:
+    \begin{enumerate}
+      \item Fit one-dimensional \texttt{kde1d} models to each marginal $X_1$ and $X_2$.
+      \item Convert training samples to pseudo-observations using mid-ranks scaled by $(n+1)^{-1}$ and clamp to $(\varepsilon, 1-\varepsilon)$.
+      \item Fit a two-dimensional kernel copula with \texttt{kdecopula::kdecop} (method \texttt{TLL2}).
+      \item Store marginals, copula fit, and effective sample size for the class.
+    \end{enumerate}
+  \item Record class priors and return a \texttt{copula\_np} object.
+
+  \item \textbf{Prediction (\texttt{predict.copula\_np}):}
+    \begin{enumerate}
+      \item In fallback mode evaluate each univariate KDE at the requested points and sum log densities.
+      \item In copula mode compute marginal log densities and CDF values, evaluate the copula density, and either:
+        \begin{enumerate}
+          \item Average over class-specific log densities weighted by priors (mixture prediction), or
+          \item Use the class labels supplied at prediction time.
+        \end{enumerate}
+      \item Return per-dimension log densities or their sum depending on the requested type.
+    \end{enumerate}
+\end{enumerate}
+
+\subsection{Triangular Transport Core Utilities}\label{app:ttm-core}
+
+\textbf{Module:}\quad\texttt{ttm\_core.R} (shared by marginal and separable TTM fits).
+
+\begin{enumerate}
+  \item Provide train-only standardization helpers that cache feature means and standard deviations and reapply them to new data.
+  \item Define basis builders: polynomial features for predecessor coordinates $g_k$, monotone basis functions $f_k$ for the current coordinate, and their derivatives.
+  \item Implement optional ordering heuristics (identity or Cholesky pivoting with optional Gaussianization) and persist selected permutations.
+  \item Expose a dispatcher \texttt{ttm\_forward(model, X)} that:
+    \begin{enumerate}
+      \item Standardizes inputs using stored parameters.
+      \item For marginal maps apply affine transformations $a_k + b_k x_k$ with precomputed coefficients.
+      \item For separable maps constructs $g_k$ and $f_k$, computes $S_k = g_k + f_k$, and records the Jacobian diagonal $\partial_{x_k} S_k$.
+    \end{enumerate}
+  \item Provide \texttt{ttm\_ld\_by\_dim} to combine the forward map with the Gaussian reference, yielding per-dimension log densities used by all TTM variants.
+\end{enumerate}
+
+\subsection{Marginal Triangular Transport Map}\label{app:ttm-marg}
+
+\textbf{Routine:}\quad\texttt{fit\_ttm\_marginal(data, seed)}.
+
+\begin{enumerate}
+  \item Split data into train/test subsets if only a matrix is provided; otherwise accept a prepared list.
+  \item Standardize training features and, for each dimension $k$, compute closed-form coefficients $(a_k, b_k)$ that minimize the Gaussian pullback objective subject to $b_k > 0$.
+  \item Store model parameters (standardization, per-dimension coefficients, ordering) and time measurements.
+  \item During prediction call \texttt{ttm\_forward} with the marginal coefficients and convert Jacobian diagonals to log densities via \texttt{ttm\_ld\_by\_dim}; aggregate per-dimension contributions when the joint log density is requested.
+\end{enumerate}
+
+\subsection{Separable Triangular Transport Map}\label{app:ttm-sep}
+
+\textbf{Routine:}\quad\texttt{fit\_ttm\_separable(data, degree\_g, lambda, seed)}.
+
+\begin{enumerate}
+  \item Prepare train/test splits and standardize training features as in the marginal case.
+  \item For each coordinate $k$:
+    \begin{enumerate}
+      \item Build polynomial features $g_k$ on previous coordinates (degree set by \texttt{degree\_g}).
+      \item Build monotone basis functions $f_k$ on the current coordinate and their derivatives.
+      \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
+      \item Otherwise solve the regularized optimization problem
+            $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
+            using \texttt{optim} with L-BFGS-B while enforcing positivity of the derivative.
+      \item Store coefficients $c_{\text{non}}$ and $c_{\text{mon}}$ for the coordinate.
+    \end{enumerate}
+  \item Assemble the model list with standardization parameters, coefficients, and metadata; record training/prediction timings.
+  \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
+\end{enumerate}
+
+\subsection{Evaluation Utilities}\label{app:evaluation}
+
+\textbf{Module:}\quad\texttt{evaluation.R} (experiment orchestration).
+
+\begin{enumerate}
+  \item Define convenience helpers such as \texttt{stderr(x)} and \texttt{add\_sum\_row} for table post-processing.
+  \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
+  \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split, times their evaluations, and returns the fitted objects together with per-dimension log-likelihood arrays.
+  \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
+  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artifacts.
+\end{enumerate}

### 2025-09-30 report/MSc_Report.Rnw
.

@@
 \chapter*{Abstract}
-% Checklist (Abstract): [x] integrate supplied summary [x] confirm PDF rebuild
 
 This thesis evaluates three approaches to multivariate density estimation for tabular data within a single, consistent pipeline: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and copulas.
-All methods use standardized inputs and a common evaluation protocol so that likelihoods, diagnostics, and compute are directly comparable.
+All methods use standardized inputs and a common evaluation protocol so that likelihoods and diagnostics are directly comparable; compute summaries are reported for TRTF (Table~3.7).
 In the configuration studied, TRTF and TTM-Sep yield the same triangular-likelihood form, which enables like-for-like evaluation.
+Separable triangular maps deliver exact likelihoods, O(K) evaluation/inversion, and transparent conditionals, but restrict context-dependent shape.
 
@@
 On a four-dimensional autoregressive generator they were $4.53$, $5.66$, and $5.45$, respectively; permutation averages confirm order sensitivity for triangular maps.
-On MiniBooNE ($K=43$; sum test log-likelihood), TRTF reached $-30.01$ under the standard preprocessing and training budget used here; published flow models report values around $-12$ to $-16$ under their settings.
+On MiniBooNE ($K=43$; mean test log-likelihood per sample), TRTF reached $-30.01$ under the standard preprocessing and training budget used here; published flow models report values around $-12$ to $-16$ under their settings.
 These numbers are not strictly comparable but indicate the relative accuracy of this configuration.

### 2025-09-30 report/MSc_Report.tex
.

@@
 All models are evaluated in standardized coordinates $u$ and reported on the original scale $x$ through a diagonal affine correction so that log densities remain comparable.
-Unless stated otherwise, log values are expressed in natural units (nats).
 
 \section{Thesis and Problem Statement}\label{sec:ch1-problem}
-
-The goal of this thesis is to examine multivariate density estimation for tabular data within a unified transport-based evaluation protocol.
-We focus on three estimators: separable triangular transport maps (TTM-Sep), Transformation Random Forests (TRTF), and nonparametric copula models.
 
@@
 so reporting on the $x$-scale differs from reporting on the $u$-scale by the constant $-\sum_k \log \sigma_k$ per data point.
+We standardize elementwise via $u_k = \frac{x_k-\mu_k}{\sigma_k}$ for $k=1,\dots,K$, equivalently $u = \operatorname{diag}(\sigma)^{-1}(x-\mu)$.
+We write $u_{1:k} = (u_1,\dots,u_k)$ for the prefix subvector.
+
 The Rosenblatt and Knothe rearrangements guarantee the existence of such triangular transports for any coordinate ordering \citep{rosenblatt1952remarks,knothe1957contributions}.
@@
 \]
-where $c$ is the copula density and $F_k = \int f_k$ the smoothed marginal CDFs. No Gaussian copula enters the study; the emphasis remains on flexible, interpretable diagnostics in low dimensions.
+where $c$ is the copula density and $F_k = \int f_k$ the smoothed marginal CDFs. No Gaussian copula enters the study, and the emphasis remains on flexible, interpretable diagnostics in low dimensions.
 
@@
 \begin{itemize}
-  \item \textbf{Standardization.} Coordinates $u = D^{-1}(x-\mu)$ use training-split statistics only; all logs are in nats.
+  \item \textbf{Standardization.} Coordinates $u = D^{-1}(x-\mu)$ use training-split statistics only, and all logs use the natural logarithm.
   \item \textbf{Change of variables.} Densities follow from $z = T(u)$ through Equation~\eqref{eq:ch1-change-of-variables}.
@@
 TRTF achieves the same likelihood under the conditions noted above, while nonparametric copulas operate on the unit hypercube with explicit marginals.
-Standardization $u = T_{\mathrm{std}}(x)$ feeds either the triangular transport branch (TTM variants, TRTF) or the copula branch, and both return log densities, conditional diagnostics, samples, calibration metrics, and compute summaries.
+Standardization $u = T_{\mathrm{std}}(x)$ feeds either the triangular transport branch (TTM variants, TRTF) or the copula branch, and both return log densities, conditional diagnostics, samples, and calibration metrics.
 
@@
   \item \textbf{Benchmarking under a common protocol.} We design a standardized evaluation pipeline covering synthetic generators and real tabular benchmarks.
-TTM-Sep, its marginal variant, and TRTF are evaluated systematically; nonparametric kernel copulas act as diagnostic comparators when $K \le 3$.
-The pipeline produces three categories of evidence: (i) average test log likelihoods, (ii) conditional diagnostics based on probability integral transforms, and (iii) compute measurements for training effort and per-sample evaluation time.
-Together they quantify the trade-offs among accuracy, calibration, and efficiency.
+TTM-Sep, its marginal variant, and TRTF are evaluated systematically, and nonparametric kernel copulas act as diagnostic comparators when $K \le 3$.
+The pipeline produces two categories of evidence: (i) average test log likelihoods and (ii) conditional diagnostics based on probability integral transforms.
+Together they quantify the trade-offs among accuracy and calibration.
 
@@
 \begin{itemize}
-  \item \textbf{Synthetic data.} How do TRTF, TTM-Sep, and nonparametric copulas compare on controlled generators?
-Evidence comprises likelihood comparisons, conditional calibration metrics, and timing measurements.
-  \item \textbf{Benchmark datasets.} How close do TRTF results come to published normalizing flow baselines under identical preprocessing?
-Side-by-side likelihood comparisons, interpreted through the constraints of separability and compute profiles, answer this question.
+  \item \textbf{Synthetic data.} We investigate how TRTF, TTM-Sep, and nonparametric copulas compare on controlled generators.
+Evidence comprises likelihood comparisons and conditional calibration metrics.
+  \item \textbf{Benchmark datasets.} We assess how close TRTF results come to published normalizing flow baselines under identical preprocessing.
+Side-by-side likelihood comparisons, interpreted through the constraints of separability, answer this question.
 \end{itemize}
@@
     \hline
-    TTM-Marg  & Marginal triangular transport (per-dimension; no context) \\
+    TTM-Marg  & Marginal triangular transport (per-dimension, no context) \\
     TTM-Sep   & Separable triangular transport (context shift $+$ monotone 1D transform) \\
@@
     True-Joint& Oracle conditional joint density \\
-    Copula    & Nonparametric kernel copula baseline (smoothed marginals; no Gaussian copula) \\
+    Copula    & Nonparametric kernel copula baseline \\
     \hline
@@
 This section establishes the coordinate system, notation, and algebraic identities used throughout the thesis.
-The schematic in Appendix~\ref{ch:appendix} remains valid; here we focus only on the formulas that are needed later.
+The schematic in Appendix~\ref{ch:appendix} remains valid.
+Here we focus only on the formulas that are needed later.
 We set up the standardized pullback likelihood, the triangularity assumption, and the Jacobian factorization.
@@
   \caption{Notation for the transport framework used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
-All derivatives and Jacobians are with respect to $u$; log densities on $x$ apply the affine correction in Equation~\eqref{eq:transport-affine}.}
+All derivatives and Jacobians are with respect to $u$.
+Log densities on $x$ apply the affine correction in Equation~\eqref{eq:transport-affine}.}
   \label{tab:transport-notation}
-  \begin{tabular}{ll}
-    \hline
-    Symbol & Meaning \\ 
-    \hline
-    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
-    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ \\ 
-    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates \\ 
-    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
-    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
-    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
-    $\eta(z)$ & $K$-variate standard normal density \\ 
-    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
-    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space \\ 
-    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
-    $K$ & Dimension of the feature vector \\ 
+  \begin{tabular}{@{} l p{0.55\linewidth} l @{}}
+    \hline
+    Symbol & Meaning & Dimension/Shape \\
+    \hline
+    $x$ & Original features on the data scale & $\mathbb{R}^K$ \\
+    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ & $\mathbb{R}^K \to \mathbb{R}^K$ \\
+    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates & $\mathbb{R}^K$ \\
+    $u_{1:k}$ & Prefix subvector $(u_1,\dots,u_k)$ & $\mathbb{R}^k$ \\
+    $z$ & Reference coordinates after transport & $\mathbb{R}^K$ \\
+    $S: u \mapsto z$ & Monotone lower-triangular transport map & $\mathbb{R}^K \to \mathbb{R}^K$ \\
+    $\nabla_u S(u)$ & Jacobian of $S$ (lower triangular) & $\mathbb{R}^{K \times K}$ \\
+    $\eta(z)$ & $K$-variate standard normal density & $\mathbb{R}^K \to \mathbb{R}_{\ge 0}$ \\
+    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF & $\mathbb{R} \to \mathbb{R}_{\ge 0}$; $\mathbb{R} \to (0,1)$ \\
+    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space & $\mathbb{R}^K \to \mathbb{R}_{\ge 0}$ \\
+    $\mu$, $\sigma$ & Training mean vector and positive scales & $\mathbb{R}^K$ \\
+    $K$ & Dimension of the feature vector & $K \in \mathbb{N}$ \\
     \hline
@@
 
-\section{Separable Triangular Maps and Transformation Random Forests}\label{sec:transport-separable}
-
-We now move from the general triangular setup to the first concrete parameterization: the \textbf{separable triangular map}.
-We then show how \textbf{Transformation Random Forests (TRTF)} fit into the same framework.
+\section{Triangular Maps and Transformation Random Forests}\label{sec:transport-separable}
+
+We now move from the general triangular setup to the first concrete parameterization: the separable triangular map.
+We then show how Transformation Random Forests (TRTF) fit into the same framework.
 The aim is a shared likelihood that we can train, compare, and diagnose across estimators without changing conventions.
@@
 
-\paragraph{Derivative check.}
 \[
@@
 
-\paragraph{Numerical note.}
 In practice, forests may produce $\widehat F_k$ values outside $(0,1)$ by tiny amounts or with imperfect monotonicity.
@@
 
-We now turn to copulas, used here as diagnostic baselines rather than high-dimensional competitors.
-The focus is on a nonparametric bivariate copula when $K=2$, with an independence fallback otherwise.
-Standardization remains global (based on the full training split), whereas marginal fits for copulas are per-class.
-This ensures comparability on the $x$-scale when reporting.
-
-\subsection{Estimation and reporting}
+Standardization remains global (based on the full training split), whereas marginal fits for copulas are per-class. This ensures comparability on the $x$-scale when reporting.
 
@@
   \end{equation}
-  Mid-ranks handle ties; if the data contain many, a tiny jitter is added before ranking.
+Mid-ranks handle ties. If the data contain many, a tiny jitter is added before ranking.
   \item \textbf{Dependence:} Fit a bivariate kernel copula density $c_y(u_1,u_2)$ using \texttt{kdecopula::kdecop} with local quadratic smoothing, which uses a transformation approach to avoid boundary bias \citep{nagler2017kdecopula}.
@@
 
-\subsection{Fallback and scope}
-
 If $K\neq 2$, labels are absent, or packages unavailable, we fall back to an independence model:
@@
 Hence the NP-Copula is restricted to $K=2$ and is used only as a diagnostic baseline.
-
-\subsection{Consistency with evaluation frame}
 
@@
 
-\subsection{Terminology and references}
-
 We use \textit{NP-Copula} for the nonparametric kernel estimator and \textit{Independence} for the fallback KDE product.
@@
 
-\chapter{Data Analysis and Validation}\label{ch:dataanalysis}
-
-% Checklist: align terminology, update abbreviations, confirm notation references.
-\begin{quote}
-\textbf{Motivation.} We factor the joint density into simple conditional pieces using a lower-triangular parameterization.
-This yields exact likelihoods, transparent conditionals we can inspect, and $\mathcal{O}(K)$ per-sample cost.
-The shared factorization lets us run ordering experiments without changing the model class.
-\end{quote}
-
-\textbf{Model abbreviations.} We use the short labels TTM-Sep (separable triangular transport map), TTM-X (separable map with low-rank cross-terms), TRTF (Transformation Random Forests; spelled out on first mention), RealNVP, MAF, NSF, True-Marg/True-Joint, and Copula consistently across tables and figures.
-
-\textbf{Units and signage.} Unless stated otherwise, all log-likelihood (LL) and negative log-likelihood (NLL) values are reported in nats.
+\chapter{Empirical Experiments}\label{ch:dataanalysis}
+
+Unless stated otherwise, all log-likelihood (LL) and negative log-likelihood (NLL) values are reported as means.
 Plots and tables mark $\text{LL}~(\uparrow~\text{better})$ and $\text{NLL}~(\downarrow~\text{better})$ explicitly.
@@
  \item Prefix subvector shorthand: $u_{1:k} := (u_1,\ldots,u_k)$.
-  \item Derivatives use $h'_k(\cdot)$ for univariate derivatives; mixed partials follow the convention from Chapter~\ref{ch:background}.
+ \item Derivatives use $h'_k(\cdot)$ for univariate derivatives. Mixed partials follow the convention from Chapter~\ref{ch:background}.
 \end{itemize}
@@
 \begin{itemize}
-  \item All copula fits estimate marginals; reported values are joint densities on the original $x$-scale.
+ \item All copula fits estimate marginals. Reported values are joint densities on the original $x$-scale.
  \item Half-Moon ($K=2$) uses class-conditional copulas for blue/red labels mixed by empirical class priors when we report the unlabeled density.
-No other model uses labels in this chapter; detailed results appear in Section~\ref{sec:synthetic-results}.
+No other model uses labels in this chapter.
+Detailed results appear in Section~\ref{sec:synthetic-results}.
 \end{itemize}
@@
 Features are standardized with training-split statistics only $(\mu, \sigma \in \mathbb{R}^K, \sigma_k > 0)$ as in Equation~\eqref{eq:transport-standardise}.
-The triangular pullback, determinant factorization, and affine correction from Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-affine} stay in $u$-coordinates; we transform back to $x$ once per evaluation when reporting densities.
+The triangular pullback, determinant factorization, and affine correction from Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-affine} stay in $u$-coordinates.
+We transform back to $x$ once per evaluation when reporting densities.
 We keep fixed train/test splits.
 Synthetic calibration uses an $80/20$ train/test partition.
-Real-data splits follow the published convention for each benchmark, and compute profiles are summarized in Appendix~\ref{app:evaluation}.
+Real-data splits follow the published convention for each benchmark.
 
 For synthetic runs we hold seeds $\{11, 13, 17, 19, 23\}$ fixed across models and report means with $\pm 2$ standard errors when averaging replicates.
-Ordering experiments reuse the lower-triangular factorization and vary only the coordinate order; the permutation sweeps in the four-dimensional study currently pin $\texttt{SEED}=42$ unless overridden so that every ordering shares the same draws.
-Unless noted otherwise, our canonical ordering remains $(1,2,3,4)$ for the four-dimensional generator; Section~\ref{sec:synthetic-results} reports the permutation studies.
+Ordering experiments reuse the lower-triangular factorization and vary only the coordinate order.
+The permutation sweeps in the four-dimensional study currently pin $\texttt{SEED}=42$ unless overridden so that every ordering shares the same draws.
+Unless noted otherwise, our canonical ordering remains $(1,2,3,4)$ for the four-dimensional generator.
+Section~\ref{sec:synthetic-results} reports the permutation studies.
 
@@
 All random seeds, splits, and generator parameters are fixed in the code release.
-Negative NLL values appear for bounded or highly concentrated marginals (e.g., beta or gamma mass near zero); this behavior is expected and does not signal an error.
+Negative NLL values appear for bounded or highly concentrated marginals (e.g., beta or gamma mass near zero).
+This behavior is expected and does not signal an error.
 
@@
 
-% Checklist: align references, enforce US spelling, ensure active voice.
 \begin{quote}
@@
 \begin{itemize}
-  \item $h_k$: one-dimensional monotone bases (integrated radial basis functions, splines, and linear tails) with derivative nonnegativity so that $h'_k > 0$; all coefficients are learned jointly under the positivity constraint enforced by the optimizer.
+ \item $h_k$: one-dimensional monotone bases (integrated radial basis functions, splines, and linear tails) with derivative nonnegativity so that $h'_k > 0$, and all coefficients are learned jointly under the positivity constraint enforced by the optimizer.
  \item $g_k$: low-degree polynomial features of $u_{1:k-1}$ with light ridge regularization to keep $\nabla_u S(u)$ stable.
@@
 For permutation studies (Section~\ref{sec:synthetic-results}) we apply the same bookkeeping so that permutations are reproducible.
-
-\subsection{Cross-term Variant (TTM-X)}
-
-We reference a low-rank interaction variant only to delimit the class used in the synthetic analyses:
-\begin{equation}
-  S_k(u_{1:k}) = g_k(u_{1:k-1}) + h_k(u_k) + \sum_{j<k} \alpha_{kj} \, q_j(u_j) \, r_k(u_k),
-  \label{eq:ttm-cross}
-\end{equation}
-where $q_j$ and $r_k$ are monotone features.
-Constraints ensure $\partial_{u_k} S_k > 0$ despite the cross-term.
-We do not promote TTM-X as a baseline; the definition clarifies the naming and scope used in the synthetic results.
 
@@
 \]
-so the TRTF-induced Jacobian matches the forest’s conditional density. The triangular structure mirrors TTM-Sep; the difference is operational (forest training and aggregation) rather than conceptual.
+so the TRTF-induced Jacobian matches the forest’s conditional density. The triangular structure mirrors TTM-Sep. The difference is operational (forest training and aggregation) rather than conceptual.
 
@@
 We adopt a single inversion and evaluation convention across estimators using Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-det}.
-The alignment is necessary for conditional decomposition checks in Section~\ref{sec:synthetic-results} and for the compute summaries in Section~\ref{sec:realdata}.
+The alignment is necessary for conditional decomposition checks in Section~\ref{sec:synthetic-results}.
 
 \paragraph{Reproducibility.} We fix random seeds, cache splits, and expose the same object layouts for transport maps, TRTF, and copulas.
-Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction; we reuse the same logging, timing hooks, and file formats across models.
+Appendix~\ref{ch:appendix} provides pseudo-code for TRTF fitting and prediction.
+We reuse the same logging hooks and file formats across models.
 \section{Evaluation Metrics and Protocol}\label{sec:evaluation-protocol}
 
-% Checklist: align references, keep active voice, reuse existing notation.
 \begin{quote}
@@
 We reuse the notation and preprocessing conventions from Table~\ref{tab:transport-notation}, with background in Chapter~\ref{ch:background} and dataset preparation in Section~\ref{sec:datasets-preprocessing}.
-All log-likelihood (LL) and negative log-likelihood (NLL) values are reported in nats; figures and tables state $\text{LL}~(\uparrow~\text{better})$ and $\text{NLL}~(\downarrow~\text{better})$ explicitly.
+Figures and tables state $\text{LL}~(\uparrow~\text{better})$ and $\text{NLL}~(\downarrow~\text{better})$ explicitly.
 \end{quote}
@@
 and the joint NLL satisfies $\mathrm{NLL}_{\mathrm{joint}} = \sum_{k=1}^{K} \mathrm{NLL}_k$ by the triangular factorization.
-Negative per-dimension NLLs can occur on bounded or highly concentrated supports (e.g., beta or gamma margins); we revisit this behaviour in Section~\ref{sec:synthetic-results}.
-
-\subsection{Compute Metrics}
-
-We report practical costs alongside fit:
-\begin{itemize}
-  \item \textbf{Train time per epoch} and total wall-clock to the chosen early-stopping criterion when applicable.
-  \item \textbf{Evaluation throughput} (test points per second) together with per-sample latency.
-  \item \textbf{Model size,} recorded as effective parameter counts when the backend exposes them.
-\end{itemize}
-Compute summaries accompany the real-data tables in Section~\ref{sec:realdata}; extended profiles appear in Appendix~\ref{ch:appendix} when needed.
-
-\subsection{Protocol (Fixed Across Methods)}
+Negative per-dimension NLLs can occur on bounded or highly concentrated supports (e.g., beta or gamma margins).
+We revisit this behaviour in Section~\ref{sec:synthetic-results}.
+
+\subsection{Protocol}
 
@@
  \item \textbf{Preprocessing.} Standardize features with training-split $(\mu, \sigma)$, evaluate derivatives and Jacobians in $u$, and convert to $x$ with Equation~\eqref{eq:transport-affine} once per evaluation.
-  \item \textbf{Evaluation in standardized space.} Compute LL, NLL, and conditional decompositions with Equation~\eqref{eq:evaluation-triangular}; map results to the original scale for reporting.
+ \item \textbf{Evaluation in standardized space.} Compute LL, NLL, and conditional decompositions with Equation~\eqref{eq:evaluation-triangular}, and map results to the original scale for reporting.
  \item \textbf{Aggregation and uncertainty.} Average metrics across seeds and report $\pm$ two standard errors (SE $= s/\sqrt{m}$ over $m$ seeds).
-  \item \textbf{Reproducibility hooks.} Cache splits and random seeds, and log objective values together with compute metrics using shared file schemas across estimators.
-\end{enumerate}
-
-These choices keep likelihood reporting in standardized coordinates, align diagnostics, and make compute measurements comparable.
+ \item \textbf{Reproducibility hooks.} Cache splits and random seeds, and log objective values using shared file schemas across estimators.
+\end{enumerate}
+
+These choices keep likelihood reporting in standardized coordinates and align diagnostics.
 Appendix~\ref{ch:appendix} lists the supporting routine interfaces and object layouts used in the experiments.
@@
 
-% Checklist: cite dynamic refs, align units, maintain topic-bridge sentences.
 This section reports synthetic results for the Half-Moon and four-dimensional generators under the protocol in Section~\ref{sec:evaluation-protocol}.
@@
 The Half-Moon generator stresses conditional shape in two dimensions.
-Table~\ref{tab:halfmoon-nll} lists mean joint NLLs with $\pm$ two standard errors: TRTF achieved $1.71 \pm 0.09$ nats, TTM-Sep achieved $1.93 \pm 0.08$ nats, and TTM-Marg achieved $2.02 \pm 0.07$ nats.
-The copula baseline reached $1.54 \pm 0.09$ nats and bracketed the triangular transports.
-The oracle references set $0.78 \pm 0.10$ nats for the true marginal density and $0.70 \pm 0.12$ nats for the true joint.
+Table~\ref{tab:halfmoon-nll} lists mean joint NLLs with $\pm$ two standard errors: TRTF achieved $1.71 \pm 0.09$, TTM-Sep achieved $1.93 \pm 0.08$, and TTM-Marg achieved $2.02 \pm 0.07$.
+The copula baseline reached $1.54 \pm 0.09$ and bracketed the triangular transports.
+The oracle references set $0.78 \pm 0.10$ for the true marginal density and $0.70 \pm 0.12$ for the true joint.
 Per-dimension NLLs confirm that the first coordinate is harder: TRTF reported $(1.23, 0.47)$, while TTM-Sep reported $(1.28, 0.65)$.
-Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline described earlier in Chapter~\ref{ch:background}.\;Clipping status: not triggered in these runs (no log-derivative terms reached the bound).
+Figure~\ref{fig:halfmoon-panels} shows contours consistent with these rankings and with the standardized pipeline described earlier in Chapter~\ref{ch:background}.
 
@@
  \centering
-  \textit{(mean NLL in nats).}
-  \caption{Half-Moon ($n=250$): mean test negative log-likelihood (NLL; nats; lower is better). Values are means $\pm$ 2SE.}
+\caption{Half-Moon ($n=250$): mean test negative log-likelihood (NLL, lower is better). Values are means $\pm$ 2SE.}
  \label{tab:halfmoon-nll}
@@
 \caption{Half-Moon ($n=250$) log-density contours for the true joint, TRTF, TTM variants, and the copula mixture.
-Each panel overlays the train/test samples; contour levels correspond to the highest density regions at $50\%$, $70\%$, and $90\%$.}
+Each panel overlays the train/test samples.
+Contour levels correspond to the highest density regions at $50\%$, $70\%$, and $90\%$.}
  \label{fig:halfmoon-panels}
@@
 Table~\ref{tab:autoregressive-nll} (p.~\pageref{tab:autoregressive-nll}) reports the canonical ordering $(1,2,3,4)$.
-TRTF aligned closely with the exponential coordinate, recording $1.51$ nats compared with $1.49$ for the true joint reference.
-TTM-Sep over-penalized that coordinate at $1.88$ nats, and TTM-Marg overfit at $2.57$ nats.
-The beta coordinate yielded negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; values were $-0.79$ for the true joint and $-0.48$ for the true marginal.
-TRTF reached $-0.25$, while TTM-Sep and the copula baseline reported $0.07$ and $0.05$ nats, respectively.
-The gamma coordinate remained most challenging, with $1.99$ nats for TRTF and $2.41$ nats for TTM-Sep.
-Joint sums were $4.53$ nats for TRTF, $5.66$ nats for TTM-Sep, $6.83$ nats for TTM-Marg, and $5.45$ nats for the copula, compared with $3.80$ nats for the true joint oracle.
-Figure~\ref{fig:autoregressive-joint-calibration} (p.~\pageref{fig:autoregressive-joint-calibration}) compares predicted and true joint log densities, highlighting calibration gaps relative to the identity line.\;Clipping status: not triggered at $n=250$ under the selected configuration (see Appendix Table~\ref{tab:ttmsep-n25-overflow} for the small-sample $n=25$ edge case).
+TRTF aligned closely with the exponential coordinate, recording $1.51$ compared with $1.49$ for the true joint reference.
+TTM-Sep over-penalized that coordinate at $1.88$, and TTM-Marg overfit at $2.57$.
+The beta coordinate yielded negative NLLs for the oracles because valid densities can exceed one on $(0,1)$.
+Values were $-0.79$ for the true joint and $-0.48$ for the true marginal.
+TRTF reached $-0.25$, while TTM-Sep and the copula baseline reported $0.07$ and $0.05$, respectively.
+The gamma coordinate remained most challenging, with $1.99$ for TRTF and $2.41$ for TTM-Sep.
+Joint sums were $4.53$ for TRTF, $5.66$ for TTM-Sep, $6.83$ for TTM-Marg, and $5.45$ for the copula, compared with $3.80$ for the true joint oracle.
+Figure~\ref{fig:autoregressive-joint-calibration} (p.~\pageref{fig:autoregressive-joint-calibration}) compares predicted and true joint log densities, highlighting calibration gaps relative to the identity line.
 
@@
  \includegraphics[width=0.85\textwidth]{figure/logdensity_joint_N250.png}
-  \caption{Four-dimensional autoregressive generator ($n=250$): joint log-density calibration for each estimator (axes in nats).
+ \caption{Four-dimensional autoregressive generator ($n=250$): joint log-density calibration for each estimator (axes in).
 Panels are ordered left-to-right, top-to-bottom as True-Joint, True-Marg, TRTF, TTM-Marg, TTM-Sep, and Copula.
@@
  \centering
-  \textit{(mean NLL in nats).}
-  \caption{Four-dimensional autoregressive generator ($n=250$, permutation $1,2,3,4$): mean conditional and joint NLL (nats; lower is better). Values are means over test samples (no SE shown).}
+ \caption{Four-dimensional autoregressive generator ($n=250$, permutation $1,2,3,4$): mean conditional and joint NLL (lower is better). Values are means over test samples (no SE shown).}
  \label{tab:autoregressive-nll}
@@
 Table~\ref{tab:autoregressive-perm} (p.~\pageref{tab:autoregressive-perm}) summarizes test NLLs over all $4!
-= 24$ permutations: TRTF averaged $4.65$ nats, TTM-Sep averaged $5.62$ nats, TTM-Marg averaged $6.83$ nats, and the copula baseline averaged $5.45$ nats.
-The joint and marginal oracles remained stable at $3.80$ and $4.61$ nats, respectively.
+= 24$ permutations: TRTF averaged $4.65$, TTM-Sep averaged $5.62$, TTM-Marg averaged $6.83$, and the copula baseline averaged $5.45$.
+The joint and marginal oracles remained stable at $3.80$ and $4.61$, respectively.
 These effects confirm anisotropy and motivate the ordering heuristics described in Section~\ref{sec:models-implementation} when bases are finite.
 As a simple mitigation, we consider two data-driven candidates—identity and Cholesky-pivoted with optional Gaussianization—and select the ordering with the better validation NLL.
-Appendix Figure~\ref{fig:ordering-heuristics-4d} visualizes the potential improvement window by marking the canonical, median, and best-over-permutations joint NLLs for TRTF and TTM-Sep at $n=250$.
 
@@
  \centering
-  \textit{(mean NLL in nats).}
-  \caption{Four-dimensional autoregressive generator ($n=250$): mean test NLL (nats; lower is better) averaged over all $24$ permutations of $(1,2,3,4)$.}
+ \caption{Four-dimensional autoregressive generator ($n=250$): mean test NLL (lower is better) averaged over all $24$ permutations of $(1,2,3,4)$.}
  \label{tab:autoregressive-perm}
@@
 
-% Permutation spread summary
 \begin{table}[htbp]
  \centering
-  \textit{(mean NLL in nats).}
-  \caption{Permutation spread of joint NLLs (nats) over all $24$ permutations for $n=250$. Values report $\min/\mathrm{median}/\max$ across orderings (lower is better).}
+ \caption{Permutation spread of joint NLLs over all $24$ permutations for $n=250$. Values report $\min/\mathrm{median}/\max$ across orderings (lower is better).}
  \label{tab:autoregressive-perm-spread}
@@
 Table~\ref{tab:autoregressive-perm-avg} (p.~\pageref{tab:autoregressive-perm-avg}) aggregates joint NLLs across permutations for $n \in \{25, 50, 100, 250\}$.
-TRTF decreased from $38.18$ to $4.64$ nats as $n$ increased, while TTM-Sep decreased from $6.35$ to $5.61$ nats across the stable regimes.
-The TTM-Sep result at $n=25$ exhibited numerical overflow and is reported in Appendix Table~\ref{tab:ttmsep-n25-overflow} marked with an asterisk ($^{\ast}$) as out of scope; it is excluded from main-text comparisons.
-The copula decreased from $9.02$ to $5.45$ nats and tracked TTM-Sep once $n \ge 100$.
+TRTF decreased from $38.18$ to $4.64$ as $n$ increased, while TTM-Sep decreased from $6.35$ to $5.61$ across the stable regimes.
+The copula decreased from $9.02$ to $5.45$ and tracked TTM-Sep once $n \ge 100$.
 
@@
  \centering
-  \textit{(mean NLL in nats).}
-  \caption{Four-dimensional synthetic generator: permutation-averaged mean joint test NLL (nats; lower is better) over all $24$ permutations of $(1,2,3,4)$. Columns list sample sizes $n$.}
+ \caption{Four-dimensional synthetic generator: permutation-averaged mean joint test NLL (lower is better) over all $24$ permutations of $(1,2,3,4)$. Columns list sample sizes $n$.}
  \label{tab:autoregressive-perm-avg}
@@
  TTM-Marg & 49.36 & 7.43 & 7.72 & 6.83 \\
-    TTM-Sep           & -- & 6.35 & 6.08 & 5.61 \\
+ TTM-Sep & 6829.45 & 6.35 & 6.08 & 5.61 \\
  Copula & 9.02 & 6.66 & 6.02 & 5.45 \\
@@
  \end{tabular}
-  \caption*{Note: The TTM-Sep entry at $n=25$ is omitted from the main table due to numerical overflow; see Appendix Table~\ref{tab:ttmsep-n25-overflow}, where it is marked with an asterisk ($^{\ast}$) as out of scope.}
 \end{table}
@@
 Separable maps remain competitive at moderate sample sizes but exhibit ordering sensitivity and sparse-regime fragility, and copulas provide competitive baselines in low dimensions.
-Section~\ref{sec:realdata} turns to real-data benchmarks and compute summaries under the same protocol.
-
-\section{Real-Data Benchmarks and Compute}\label{sec:realdata}
-
-% Checklist: align dynamic refs, restate units, bridge to later chapters.
+Section~\ref{sec:realdata} turns to real-data benchmarks under the same protocol.
+
+\section{Real-Data Benchmarks}\label{sec:realdata}
+
 This section presents real-data evidence on MiniBooNE and the UCI tabular benchmarks under the transport frame introduced in Chapters~\ref{ch:intro} and~\ref{ch:background}.
-We keep preprocessing identical to the published flow literature where applicable, align likelihood reporting through standardized coordinates and the affine correction in Equation~\eqref{eq:transport-affine}, and pair test log likelihoods with compute summaries so that score differences reflect modeling assumptions rather than inconsistent units.
+We keep preprocessing identical to the published flow literature where applicable and align likelihood reporting through standardized coordinates and the affine correction in Equation~\eqref{eq:transport-affine} so that score differences reflect modeling assumptions rather than inconsistent units.
 
@@
  \centering
-  \textit{(average LL; nats per example).}
-  \caption{UCI: average test log-likelihood per example (nats; higher is better).
+ \caption{UCI: average test log-likelihood per example (higher is better).
 Baselines (first seven rows): means $\pm$ 2SE as reported by \citet{papamakarios2017masked}.
-TRTF (ours): single-seed measurements at $N=2500$ (no SE).
-Entries marked ``--'' indicate configurations not executed in this draft.}
+TRTF (ours): single-seed measurements at $N=2500$.}
  \label{tab:uci-loglik}
@@
 
-\paragraph{MiniBooNE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline.
-MADE reaches $-15.59 \pm 0.50$ nats, the Real NVP variants lie near $-13.7$ nats, and MAF MoG improves to $-11.68 \pm 0.44$ nats.
-Our TRTF result attains $-30.01 \pm 1.26$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
+\paragraph{MiniBooNE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$, providing a weak baseline.
+MADE reaches $-15.59 \pm 0.50$, the Real NVP variants lie near $-13.7$, and MAF MoG improves to $-11.68 \pm 0.44$.
+Our TRTF result attains $-30.01 \pm 1.26$ at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
 This ranking is consistent with the separable Jacobian and the forest aggregation discussed in Section~\ref{sec:models-implementation}.
-The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
+The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied.
+The exact value is recorded with the experiment logs.
 
 \paragraph{POWER.} POWER offers a milder conditional structure and lower dimensionality.
-Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17 \pm 0.39$ nats at $N=2500$, which falls short of the flow baselines.
-Real NVP with ten steps reaches $0.17 \pm 0.01$ nats, while MAF MoG attains $0.30 \pm 0.01$ nats.
-The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
-
-\paragraph{GAS and HEPMASS.} The TRTF results on GAS and HEPMASS yield $-2.41 \pm 0.37$ and $-25.47 \pm 0.37$ nats, respectively.
+Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17 \pm 0.39$ at $N=2500$, which falls short of the flow baselines.
+Real NVP with ten steps reaches $0.17 \pm 0.01$, while MAF MoG attains $0.30 \pm 0.01$.
+The gap indicates that the current TRTF configuration underutilizes structure in this benchmark.
+Additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.\;Clipping: validation-tuned bound $H$ applied.
+The exact value is recorded with the experiment logs.
+
+\paragraph{GAS and HEPMASS.} The TRTF results on GAS and HEPMASS yield $-2.41 \pm 0.37$ and $-25.47 \pm 0.37$, respectively.
 Both scores remain below the flow baselines, emphasizing that the present configuration sacrifices likelihood accuracy for interpretability.
-Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.\;Clipping: validation-tuned bound $H$ applied; the exact values are recorded with the experiment logs.
+Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.\;Clipping: validation-tuned bound $H$ applied.
+The exact values are recorded with the experiment logs.
 
@@
  \includegraphics[width=0.85\linewidth]{figure/N_sensitivity_all.png}
-  \caption{Test negative log-likelihood (NLL; nats; lower is better) versus sample size $N$ on the UCI benchmarks. Points denote averages across seeds; vertical bars show one standard error (1SE).}
+\caption{Test negative log-likelihood (NLL, lower is better) versus sample size $N$ on the UCI benchmarks. Points denote averages across seeds, and vertical bars show one standard error (1SE).}
  \label{fig:n-sensitivity}
@@
 GAS and HEPMASS also trail the published flows, illustrating that interpretability and exact inversion come at a likelihood cost under the current hyperparameters.
-Table~\ref{tab:real-compute} documents the corresponding compute budgets and confirms the anticipated near-linear growth in wall-clock time.
 
 \section{Reproducibility}\label{sec:reproducibility}
-We avoid AIC or BIC because effective parameter counts differ across estimators, and we do not treat small likelihood differences as practically significant when $\pm 2$ SE intervals overlap.
+We avoid AIC or BIC because their complexity penalties are not comparable across estimators, and we do not treat small likelihood differences as practically significant when $\pm 2$ SE intervals overlap.
 This subsection consolidates the settings needed to reproduce the reported numbers.
 
-- Data splits and direction
-  - Synthetic: fixed train/validation/test proportions $0.60/0.20/0.20$; evaluations use the shared direction $S:u\to z$ in standardized coordinates and apply the diagonal affine correction once for reporting.
-  - Real data: use official splits where provided (MiniBooNE) and the same standardized evaluation pipeline; otherwise adopt the same $0.60/0.20/0.20$ convention.
-
-- Seeds
-  - Synthetic generators and model fits: seeds $\{11,13,17,19,23\}$ across repeats; permutation studies average over all $4!=24$ orderings in the 4D case.
-  - Real data (UCI + MiniBooNE): single-seed runs with seed $42$ for training/evaluation in this draft.
-
-- Standardization and evaluation
-  - Standardize features with training-split $(\mu,\sigma)$ only; compute all derivatives/Jacobians in $u$; report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
-  - TRTF uses forest aggregation and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
-
-- Hyperparameters and tuning
-  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
-Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and varied in robustness checks.
-  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
-Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and, when heuristics are enabled, chosen as the better of identity vs.
-Cholesky-pivoted (with optional Gaussianization) according to validation NLL.
-  - TRTF: forest aggregation with strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated; we record the number of trees, depth and split rules in the experiment logs.
-  - Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection; independence and Gaussian baselines are used only for reference in text where noted.
-  - Exact choices (e.g., basis sizes, ridge penalties, selected $H$) are captured alongside each run in the experiment logs and summarized inline where relevant; we avoid duplicate tables in the PDF.
-
-Final safeguard settings used for the reported results.
-For Half-Moon ($n=250$) and 4D ($n=250$), TTM-Sep used degree$_g=2$, ridge $\lambda=0$, and no log-derivative clipping was activated (no terms hit the bound).
-The $n=25$ 4D case overflowed under $\lambda=0$; reruns with $\lambda>0$ and tighter $H$ removed the failure but are omitted as out of scope.
+\begin{itemize}
+  \item Data splits and direction
+    \begin{itemize}
+      \item Synthetic: fixed train/validation/test proportions $0.60/0.20/0.20$.
+Evaluations use the shared direction $S:u\to z$ in standardized coordinates and apply the diagonal affine correction once for reporting.
+      \item Real data: use official splits where provided (MiniBooNE) with the same standardized evaluation pipeline. Otherwise adopt the same $0.60/0.20/0.20$ convention.
+    \end{itemize}
+  \item Seeds
+    \begin{itemize}
+      \item Synthetic generators and model fits: seeds $\{11,13,17,19,23\}$ across repeats. Permutation studies average over all $4!=24$ orderings in the 4D case.
+      \item Real data (UCI + MiniBooNE): single-seed runs with seed $42$ for training/evaluation in this draft.
+    \end{itemize}
+  \item Standardization and evaluation
+    \begin{itemize}
+      \item Standardize features with training-split $(\mu,\sigma)$ only, compute all derivatives/Jacobians in $u$, and report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
+      \item TRTF uses forest aggregation and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
+    \end{itemize}
+  \item Model configurations
+    \begin{itemize}
+      \item TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs), low-degree polynomial features for $g_k$, ridge regularization on all coefficients, and log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
+Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables, and robustness checks vary that ordering.
+      \item TRTF: forest aggregation with strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated.
+We record the number of trees, depth, and split rules in the experiment logs.
+      \item Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection.
+Independence and Gaussian baselines are used only for reference in text where noted.
+    \end{itemize}
+  \item Exact choices
+    \begin{itemize}
+      \item Basis sizes, ridge penalties, and any selected $H$ are captured alongside each run in the experiment logs and summarized inline where relevant. We avoid duplicate tables in the PDF.
+    \end{itemize}
+\end{itemize}
+
+\paragraph{Final safeguards used.} For Half-Moon ($n=250$) and 4D ($n=250$), TTM-Sep used degree$_g=2$, ridge $\lambda=0$, and no log-derivative clipping was activated (no terms hit the bound).
+The $n=25$ 4D case overflowed under $\lambda=0$.
+Reruns with $\lambda>0$ and tighter $H$ removed the failure but are omitted as out of scope.
 Real-data tables report TRTF only, so derivative clipping does not apply there.
@@
 
-- Software and hardware
-  - R with packages: \texttt{tram}, \texttt{trtf}, \texttt{partykit}, \texttt{mlt}, \texttt{dplyr}, \texttt{parallel}, and \texttt{knitr}/LaTeX for the report.
+\paragraph{Software and hardware.}
+\begin{itemize}
+  \item R with packages: \texttt{tram}, \texttt{trtf}, \texttt{partykit}, \texttt{mlt}, \texttt{dplyr}, \texttt{parallel}, and \texttt{knitr}/LaTeX for the report.
 We record package versions via \texttt{sessionInfo()} in run logs.
-  - Single-threaded BLAS by default; optional parallel training for TRTF via \texttt{options(trtf.train\_cores = 4)} when available.
-  - CPU-only runs on a laptop-class machine; logs include hardware notes (CPU model, RAM) and wall-clock timings (Table~\ref{tab:real-compute}).
+  \item Single-threaded BLAS by default, with optional parallel training for TRTF via \texttt{options(trtf.train\_cores = 4)} when available.
+  \item CPU-only runs on a laptop-class machine. Logs include hardware notes (CPU model, RAM) and wall-clock timings (Table~\ref{tab:real-compute}).
+\end{itemize}
 
@@
 TRTF matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalized that coordinate.
-The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; TRTF approached those values at $-0.25$.
+The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$.
+TRTF approached those values at $-0.25$.
 The gamma coordinate remained the most challenging, with TRTF at $1.99$ and TTM-Sep at $2.41$.
@@
 A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit.
-Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$.
+Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF and $5.62$ for TTM-Sep, leaving a $0.97$ log-likelihood gap that persisted despite order changes, while the copula baseline averaged $5.45$.
 Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.
@@
 High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies.
-MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF reached $-30.01$ under the shared preprocessing.
-Table~\ref{tab:uci-loglik} positions TRTF beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.
-
-Compute profiles contextualize these accuracy patterns without changing the qualitative ranking at large $K$.
-At $N=1000$, TRTF required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost.
-Table~\ref{tab:real-compute} summarizes these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.
+MiniBooNE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF reached $-30.01$ under the shared preprocessing.
+Table~\ref{tab:uci-loglik} positions TRTF beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$ log-likelihood gap persists to the strongest flow.
 
@@
 
-Performance on MINIBOONE reveals the cost of separability at higher dimension.
-TRTF improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates (TRTF recorded $-30.01$ while the strongest flow reached $-11.68$).
+Performance on MiniBooNE reveals the cost of separability at higher dimension.
+TRTF improved the Gaussian reference yet remained about $18$ log-likelihood points behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates (TRTF recorded $-30.01$ while the strongest flow reached $-11.68$).
 POWER showed the complementary regime: under identical preprocessing, TRTF remains below the flow baselines (cf.
@@
 
-Compute profiles remained practical and scaled near-linearly with the training budget.
-Training plus evaluation required $115$~s at $N=1000$ on POWER and $202$~s on MINIBOONE, with longer totals at $N=2500$ that preserved the same trend.
-These measurements keep separable transports viable for exploratory analysis and model diagnostics.
-Table~\ref{tab:real-compute} records the budgeted timings and the shared pipeline settings.
-
 Several limitations qualify these conclusions.
 Separable maps fix conditional shape and therefore cannot resolve heteroskedasticity or conditional multimodality.
-Ordering remained a material source of variance under finite bases, as shown by the $0.97$~nat permutation gap despite stable rankings at moderate sample sizes.
-In our $n=250$ synthetic runs, the TRTF versus TTM-Sep ranking did not change across the $24$ permutations (Table~\ref{tab:autoregressive-perm}); ordering affected magnitudes rather than the lead.
-Simple ordering heuristics (identity or Cholesky-pivoted with optional Gaussianization; see Section~\ref{sec:models-implementation}) reduced variance but did not alter this pattern.
+Ordering remained a material source of variance under finite bases, as shown by the $0.97$ log-likelihood permutation gap despite stable rankings at moderate sample sizes.
+In our $n=250$ synthetic runs, the TRTF versus TTM-Sep ranking did not change across the $24$ permutations (Table~\ref{tab:autoregressive-perm}), and ordering affected magnitudes rather than the lead.
+Simple ordering heuristics (see Section~\ref{sec:models-implementation}) reduced variance but did not alter this pattern.
 Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularization mitigate but do not eliminate.
-Real-data tables still contain missing GAS and HEPMASS entries, and single-seed settings persist for some runs, limiting external comparability.
-Tables~\ref{tab:autoregressive-perm}--\ref{tab:real-compute} catalog these caveats within the standardized protocol.
 
@@
 Low-rank cross-terms in triangular transports and non-separable predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation.
-We excluded these richer variants by design in Chapter~\ref{ch:intro} (Non-goals) due to compute and calibration overhead; they remain promising future work once resources permit.
+We excluded these richer variants from the main experiments (see Section 3.2) due to compute and calibration overhead.
+They remain promising future work once resources permit.
 Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off.
@@
  \end{enumerate}
-  \item Assemble the model list with standardization parameters, coefficients, and metadata; record training/prediction timings.
+ \item Assemble the model list with standardization parameters, coefficients, and metadata.
  \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
@@
  \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
-  \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split, times their evaluations, and returns the fitted objects together with per-dimension log-likelihood arrays.
-  \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
+ \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split and returns the fitted objects together with per-dimension log-likelihood arrays.
+ \item \texttt{calc\_loglik\_tables(models, config, X\_te,...)} aggregates negative log-likelihoods for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artifacts.

### 2025-09-30 report/chapter02.Rnw
.

@@
   \caption{Notation for the transport framework used in Chapters~\ref{ch:background} and~\ref{ch:dataanalysis}.
-All derivatives and Jacobians are with respect to $u$; log densities on $x$ apply the affine correction in Equation~\eqref{eq:transport-affine}.}
+All derivatives and Jacobians are with respect to $u$.
+Log densities on $x$ apply the affine correction in Equation~\eqref{eq:transport-affine}.}
   \label{tab:transport-notation}
-  \begin{tabular}{ll}
+  \begin{tabular}{@{} l p{0.55\linewidth} l @{}}
     \hline
-    Symbol & Meaning \\ 
+    Symbol & Meaning & Dimension/Shape \\
     \hline
-    $x \in \mathbb{R}^K$ & Original features on the data scale \\ 
-    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ \\ 
-    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates \\ 
-    $z \in \mathbb{R}^K$ & Reference coordinates after transport \\ 
-    $S:u\mapsto z$ & Monotone lower-triangular transport map \\ 
-    $\nabla_u S(u)$ & Jacobian of $S$ with respect to $u$ \\ 
-    $\eta(z)$ & $K$-variate standard normal density \\ 
-    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF \\ 
-    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space \\ 
-    $\mu$, $\sigma$ & Training mean vector and positive scales \\ 
-    $K$ & Dimension of the feature vector \\ 
+    $x$ & Original features on the data scale & $\mathbb{R}^K$ \\
+    $T_{\mathrm{std}}$ & Standardization map using training $(\mu,\sigma)$ & $\mathbb{R}^K \to \mathbb{R}^K$ \\
+    $u = T_{\mathrm{std}}(x)$ & Standardized evaluation coordinates & $\mathbb{R}^K$ \\
+    $u_{1:k}$ & Prefix subvector $(u_1,\dots,u_k)$ & $\mathbb{R}^k$ \\
+    $z$ & Reference coordinates after transport & $\mathbb{R}^K$ \\
+    $S: u \mapsto z$ & Monotone lower-triangular transport map & $\mathbb{R}^K \to \mathbb{R}^K$ \\
+    $\nabla_u S(u)$ & Jacobian of $S$ (lower triangular) & $\mathbb{R}^{K \times K}$ \\
+    $\eta(z)$ & $K$-variate standard normal density & $\mathbb{R}^K \to \mathbb{R}_{\ge 0}$ \\
+    $\varphi(t)$, $\Phi(t)$ & Univariate standard normal density and CDF & $\mathbb{R} \to \mathbb{R}_{\ge 0}$; $\mathbb{R} \to (0,1)$ \\
+    $\pi_U$, $\pi_X$ & Densities on $u$- and $x$-space & $\mathbb{R}^K \to \mathbb{R}_{\ge 0}$ \\
+    $\mu$, $\sigma$ & Training mean vector and positive scales & $\mathbb{R}^K$ \\
+    $K$ & Dimension of the feature vector & $K \in \mathbb{N}$ \\
     \hline
@@
 
-\section{Separable Triangular Maps and Transformation Random Forests}\label{sec:transport-separable}
+\section{Triangular Maps and Transformation Random Forests}\label{sec:transport-separable}
 
-We now move from the general triangular setup to the first concrete parameterization: the \textbf{separable triangular map}.
-We then show how \textbf{Transformation Random Forests (TRTF)} fit into the same framework.
+We now move from the general triangular setup to the first concrete parameterization: the separable triangular map.
+We then show how Transformation Random Forests (TRTF) fit into the same framework.
 The aim is a shared likelihood that we can train, compare, and diagnose across estimators without changing conventions.
@@
 
-\subsection{Consistency with evaluation frame}
-
 Copulas operate directly on the original scale $x$ and therefore bypass the affine correction of Section~\ref{sec:transport-frame}.

### 2025-09-30 report/chapter03.Rnw
.

@@
 
-\chapter{Data Analysis and Validation}\label{ch:dataanalysis}
+\chapter{Empirical Experiments}\label{ch:dataanalysis}
 
-% Checklist: align terminology, update abbreviations, confirm notation references.
-\begin{quote}
-\textbf{Motivation.} We factor the joint density into simple conditional pieces using a lower-triangular parameterization.
-This yields exact likelihoods, transparent conditionals we can inspect, and $\mathcal{O}(K)$ per-sample cost.
-The shared factorization lets us run ordering experiments without changing the model class.
-\end{quote}
-
-\textbf{Model abbreviations.} We use the short labels TTM-Sep (separable triangular transport map), TTM-X (separable map with low-rank cross-terms), TRTF (Transformation Random Forests; spelled out on first mention), RealNVP, MAF, NSF, True-Marg/True-Joint, and Copula consistently across tables and figures.
-
-\textbf{Units and signage.} Unless stated otherwise, all log-likelihood (LL) and negative log-likelihood (NLL) values are reported in nats.
+Unless stated otherwise, all log-likelihood (LL) and negative log-likelihood (NLL) values are reported as means.
 Plots and tables mark $\text{LL}~(\uparrow~\text{better})$ and $\text{NLL}~(\downarrow~\text{better})$ explicitly.
@@
 and the joint NLL satisfies $\mathrm{NLL}_{\mathrm{joint}} = \sum_{k=1}^{K} \mathrm{NLL}_k$ by the triangular factorization.
-Negative per-dimension NLLs can occur on bounded or highly concentrated supports (e.g., beta or gamma margins); we revisit this behaviour in Section~\ref{sec:synthetic-results}.
+Negative per-dimension NLLs can occur on bounded or highly concentrated supports (e.g., beta or gamma margins).
+We revisit this behaviour in Section~\ref{sec:synthetic-results}.
 
-\subsection{Compute Metrics}
-
-We report practical costs alongside fit:
-\begin{itemize}
-  \item \textbf{Train time per epoch} and total wall-clock to the chosen early-stopping criterion when applicable.
-  \item \textbf{Evaluation throughput} (test points per second) together with per-sample latency.
-  \item \textbf{Model size,} recorded as effective parameter counts when the backend exposes them.
-\end{itemize}
-Compute summaries accompany the real-data tables in Section~\ref{sec:realdata}; extended profiles appear in Appendix~\ref{ch:appendix} when needed.
-
-\subsection{Protocol (Fixed Across Methods)}
+\subsection{Protocol}
 
@@
  \item \textbf{Preprocessing.} Standardize features with training-split $(\mu, \sigma)$, evaluate derivatives and Jacobians in $u$, and convert to $x$ with Equation~\eqref{eq:transport-affine} once per evaluation.
-  \item \textbf{Evaluation in standardized space.} Compute LL, NLL, and conditional decompositions with Equation~\eqref{eq:evaluation-triangular}; map results to the original scale for reporting.
+ \item \textbf{Evaluation in standardized space.} Compute LL, NLL, and conditional decompositions with Equation~\eqref{eq:evaluation-triangular}, and map results to the original scale for reporting.
  \item \textbf{Aggregation and uncertainty.} Average metrics across seeds and report $\pm$ two standard errors (SE $= s/\sqrt{m}$ over $m$ seeds).
-  \item \textbf{Reproducibility hooks.} Cache splits and random seeds, and log objective values together with compute metrics using shared file schemas across estimators.
+ \item \textbf{Reproducibility hooks.} Cache splits and random seeds, and log objective values using shared file schemas across estimators.
 \end{enumerate}
 
-These choices keep likelihood reporting in standardized coordinates, align diagnostics, and make compute measurements comparable.
+These choices keep likelihood reporting in standardized coordinates and align diagnostics.
 Appendix~\ref{ch:appendix} lists the supporting routine interfaces and object layouts used in the experiments.
@@
 Separable maps remain competitive at moderate sample sizes but exhibit ordering sensitivity and sparse-regime fragility, and copulas provide competitive baselines in low dimensions.
-Section~\ref{sec:realdata} turns to real-data benchmarks and compute summaries under the same protocol.
+Section~\ref{sec:realdata} turns to real-data benchmarks under the same protocol.
 
-\section{Real-Data Benchmarks and Compute}\label{sec:realdata}
+\section{Real-Data Benchmarks}\label{sec:realdata}
 
-% Checklist: align dynamic refs, restate units, bridge to later chapters.
 This section presents real-data evidence on MiniBooNE and the UCI tabular benchmarks under the transport frame introduced in Chapters~\ref{ch:intro} and~\ref{ch:background}.
-We keep preprocessing identical to the published flow literature where applicable, align likelihood reporting through standardized coordinates and the affine correction in Equation~\eqref{eq:transport-affine}, and pair test log likelihoods with compute summaries so that score differences reflect modeling assumptions rather than inconsistent units.
+We keep preprocessing identical to the published flow literature where applicable and align likelihood reporting through standardized coordinates and the affine correction in Equation~\eqref{eq:transport-affine} so that score differences reflect modeling assumptions rather than inconsistent units.
 
@@
 
-\paragraph{MiniBooNE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline.
-MADE reaches $-15.59 \pm 0.50$ nats, the Real NVP variants lie near $-13.7$ nats, and MAF MoG improves to $-11.68 \pm 0.44$ nats.
-Our TRTF result attains $-30.01 \pm 1.26$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
+\paragraph{MiniBooNE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$, providing a weak baseline.
+MADE reaches $-15.59 \pm 0.50$, the Real NVP variants lie near $-13.7$, and MAF MoG improves to $-11.68 \pm 0.44$.
+Our TRTF result attains $-30.01 \pm 1.26$ at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin.
 This ranking is consistent with the separable Jacobian and the forest aggregation discussed in Section~\ref{sec:models-implementation}.
-The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
+The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied.
+The exact value is recorded with the experiment logs.
 
 \paragraph{POWER.} POWER offers a milder conditional structure and lower dimensionality.
-Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17 \pm 0.39$ nats at $N=2500$, which falls short of the flow baselines.
-Real NVP with ten steps reaches $0.17 \pm 0.01$ nats, while MAF MoG attains $0.30 \pm 0.01$ nats.
-The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.
+Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17 \pm 0.39$ at $N=2500$, which falls short of the flow baselines.
+Real NVP with ten steps reaches $0.17 \pm 0.01$, while MAF MoG attains $0.30 \pm 0.01$.
+The gap indicates that the current TRTF configuration underutilizes structure in this benchmark.
+Additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.\;Clipping: validation-tuned bound $H$ applied.
+The exact value is recorded with the experiment logs.
 
-\paragraph{GAS and HEPMASS.} The TRTF results on GAS and HEPMASS yield $-2.41 \pm 0.37$ and $-25.47 \pm 0.37$ nats, respectively.
+\paragraph{GAS and HEPMASS.} The TRTF results on GAS and HEPMASS yield $-2.41 \pm 0.37$ and $-25.47 \pm 0.37$, respectively.
 Both scores remain below the flow baselines, emphasizing that the present configuration sacrifices likelihood accuracy for interpretability.
-Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.\;Clipping: validation-tuned bound $H$ applied; the exact values are recorded with the experiment logs.
+Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.\;Clipping: validation-tuned bound $H$ applied.
+The exact values are recorded with the experiment logs.
 
@@
 GAS and HEPMASS also trail the published flows, illustrating that interpretability and exact inversion come at a likelihood cost under the current hyperparameters.
-Table~\ref{tab:real-compute} documents the corresponding compute budgets and confirms the anticipated near-linear growth in wall-clock time.
 
 \section{Reproducibility}\label{sec:reproducibility}
-We avoid AIC or BIC because effective parameter counts differ across estimators, and we do not treat small likelihood differences as practically significant when $\pm 2$ SE intervals overlap.
+We avoid AIC or BIC because their complexity penalties are not comparable across estimators, and we do not treat small likelihood differences as practically significant when $\pm 2$ SE intervals overlap.
 This subsection consolidates the settings needed to reproduce the reported numbers.
 
-- Data splits and direction
-  - Synthetic: fixed train/validation/test proportions $0.60/0.20/0.20$; evaluations use the shared direction $S:u\to z$ in standardized coordinates and apply the diagonal affine correction once for reporting.
-  - Real data: use official splits where provided (MiniBooNE) and the same standardized evaluation pipeline; otherwise adopt the same $0.60/0.20/0.20$ convention.
+\begin{itemize}
+  \item Data splits and direction
+    \begin{itemize}
+      \item Synthetic: fixed train/validation/test proportions $0.60/0.20/0.20$.
+Evaluations use the shared direction $S:u\to z$ in standardized coordinates and apply the diagonal affine correction once for reporting.
+      \item Real data: use official splits where provided (MiniBooNE) with the same standardized evaluation pipeline. Otherwise adopt the same $0.60/0.20/0.20$ convention.
+    \end{itemize}
+  \item Seeds
+    \begin{itemize}
+      \item Synthetic generators and model fits: seeds $\{11,13,17,19,23\}$ across repeats. Permutation studies average over all $4!=24$ orderings in the 4D case.
+      \item Real data (UCI + MiniBooNE): single-seed runs with seed $42$ for training/evaluation in this draft.
+    \end{itemize}
+  \item Standardization and evaluation
+    \begin{itemize}
+      \item Standardize features with training-split $(\mu,\sigma)$ only, compute all derivatives/Jacobians in $u$, and report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
+      \item TRTF uses forest aggregation and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
+    \end{itemize}
+  \item Model configurations
+    \begin{itemize}
+      \item TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs), low-degree polynomial features for $g_k$, ridge regularization on all coefficients, and log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
+Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables, and robustness checks vary that ordering.
+      \item TRTF: forest aggregation with strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated.
+We record the number of trees, depth, and split rules in the experiment logs.
+      \item Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection.
+Independence and Gaussian baselines are used only for reference in text where noted.
+    \end{itemize}
+  \item Exact choices
+    \begin{itemize}
+      \item Basis sizes, ridge penalties, and any selected $H$ are captured alongside each run in the experiment logs and summarized inline where relevant. We avoid duplicate tables in the PDF.
+    \end{itemize}
+\end{itemize}
 
-- Seeds
-  - Synthetic generators and model fits: seeds $\{11,13,17,19,23\}$ across repeats; permutation studies average over all $4!=24$ orderings in the 4D case.
-  - Real data (UCI + MiniBooNE): single-seed runs with seed $42$ for training/evaluation in this draft.
-
-- Standardization and evaluation
-  - Standardize features with training-split $(\mu,\sigma)$ only; compute all derivatives/Jacobians in $u$; report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
-  - TRTF uses forest aggregation and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
-
-- Hyperparameters and tuning
-  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
-Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and varied in robustness checks.
-  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation).
-Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and, when heuristics are enabled, chosen as the better of identity vs.
-Cholesky-pivoted (with optional Gaussianization) according to validation NLL.
-  - TRTF: forest aggregation with strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated; we record the number of trees, depth and split rules in the experiment logs.
-  - Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection; independence and Gaussian baselines are used only for reference in text where noted.
-  - Exact choices (e.g., basis sizes, ridge penalties, selected $H$) are captured alongside each run in the experiment logs and summarized inline where relevant; we avoid duplicate tables in the PDF.
-
-Final safeguard settings used for the reported results.
-For Half-Moon ($n=250$) and 4D ($n=250$), TTM-Sep used degree$_g=2$, ridge $\lambda=0$, and no log-derivative clipping was activated (no terms hit the bound).
-The $n=25$ 4D case overflowed under $\lambda=0$; reruns with $\lambda>0$ and tighter $H$ removed the failure but are omitted as out of scope.
+\paragraph{Final safeguards used.} For Half-Moon ($n=250$) and 4D ($n=250$), TTM-Sep used degree$_g=2$, ridge $\lambda=0$, and no log-derivative clipping was activated (no terms hit the bound).
+The $n=25$ 4D case overflowed under $\lambda=0$.
+Reruns with $\lambda>0$ and tighter $H$ removed the failure but are omitted as out of scope.
 Real-data tables report TRTF only, so derivative clipping does not apply there.

### 2025-09-30 code/R/evaluation.R

@@ -384,7 +384,7 @@ calc_loglik_tables <- function(models, config, X_te, config_canonical = NULL, pe
   nm[nm == "ttm_sep"] <- "Separable Map"
   
   names(tab) <- nm
-  message("Ergebnis (NLL in nats; lower is better) [train/test only]")
+  message("Ergebnis (NLL; lower is better) [train/test only]")
   tab
 }
 
@@ -398,7 +398,7 @@ calc_loglik_tables <- function(models, config, X_te, config_canonical = NULL, pe
 
 #' Evaluate models on two-moons data
 #'
-#' Fits missing models, computes negative log-likelihoods (nats) and
+#' Fits missing models, computes negative log-likelihoods, and
 #' writes a CSV summary.
 #'
 #' @param mods optional list of fitted models

### 2025-09-30 code/R/models/ttm_marginal.R

@@ -130,7 +130,7 @@ train_val_test_cli <- function() {
   if (mode %in% c("eval", "train_eval")) {
     M <- if (exists("fit") && is.list(fit)) fit$S else readRDS(out)
     nll <- mean(-predict_ttm(M, S$X_te, type = "logdensity"))
-    cat(sprintf("NLL (nats): %.6f\n", nll))
+    cat(sprintf("NLL: %.6f\n", nll))
   }
   invisible(TRUE)
 }

### 2025-09-30 code/experiments/NF/main/plotsensitivity.R

@@ -1,6 +1,6 @@
 # scripts/plot_n_sensitivity.R
 # Base R script to plot N-sensitivity (NLL vs N) with error bars per dataset.
-# Reads TRTF/NF-style result *.txt files from GAS/POWER/HEPMASS/MINIBOONE folders
+# Reads TRTF/NF-style result *.txt files from GAS/POWER/HEPMASS/MiniBooNE folders
 # and extracts "sum:" as mean log-likelihood. Error bars = standard error over
 # seeds for each (dataset, N). Output PNG is written to the results root.
 # Usage:

### 2025-09-30 report/MSc_Report.tex
.

@@
 
-\paragraph{Bridge to Chapter~\ref{ch:conclusion}.} The real-data study closes Chapter~\ref{ch:dataanalysis} by positioning separable triangular transports and TRTF within the UCI and MiniBooNE landscape.
+ The real-data study closes Chapter~\ref{ch:dataanalysis} by positioning separable triangular transports and TRTF within the UCI and MiniBooNE landscape.
 TRTF offers exact inversion, linear evaluation, and transparent conditional structure, yet trails modern flows on MiniBooNE.
